### 2021-02-21 newvariant.tex
simple paper

\documentclass[12pt]{article}
\usepackage[top=1in,left=1in, right = 1in, footskip=1in]{geometry}

\usepackage{graphicx}
\usepackage{xspace}
%\usepackage{adjustbox}

\newcommand{\comment}{\showcomment}
%% \newcommand{\comment}{\nocomment}

\newcommand{\showcomment}[3]{\textcolor{#1}{\textbf{[#2: }\textsl{#3}\textbf{]}}}
\newcommand{\nocomment}[3]{}

\newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}
\newcommand{\bmb}[1]{\comment{blue}{BMB}{#1}}
\newcommand{\djde}[1]{\comment{red}{DJDE}{#1}}

\newcommand{\eref}[1]{Eq.~\ref{eq:#1}}
\newcommand{\fref}[1]{Fig.~\ref{fig:#1}}
\newcommand{\Fref}[1]{Fig.~\ref{fig:#1}}
\newcommand{\sref}[1]{Sec.~\ref{#1}}
\newcommand{\frange}[2]{Fig.~\ref{fig:#1}--\ref{fig:#2}}
\newcommand{\tref}[1]{Table~\ref{tab:#1}}
\newcommand{\tlab}[1]{\label{tab:#1}}
\newcommand{\seminar}{SE\mbox{$^m$}I\mbox{$^n$}R}

\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}

\usepackage{lineno}
\linenumbers

\usepackage[pdfencoding=auto, psdextra]{hyperref}

\usepackage{natbib}
\bibliographystyle{chicago}
\date{\today}

\usepackage{xspace}
\newcommand*{\ie}{i.e.\@\xspace}

\usepackage{color}

\newcommand{\Rx}[1]{\ensuremath{{\mathcal R}_{#1}}\xspace} 
\newcommand{\Ro}{\Rx{0}}
\newcommand{\Rc}{\Rx{\mathrm{c}}}
\newcommand{\Ri}{\Rx{\mathrm{i}}}
\newcommand{\RR}{\ensuremath{{\mathcal R}}\xspace}
\newcommand{\Rhat}{\ensuremath{{\hat\RR}}}
\newcommand{\Rnaive}{\ensuremath{{\mathcal R}_{\textrm{\tiny naive}}}\xspace}
\newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}
\newcommand{\dd}[1]{\ensuremath{\, \mathrm{d}#1}}
\newcommand{\dtau}{\dd{\tau}}
\newcommand{\dx}{\dd{x}}
\newcommand{\dsigma}{\dd{\sigma}}

\newcommand{\tstart}{\ensuremath{\tsub{t}{start}}\xspace}
\newcommand{\tend}{\ensuremath{\tsub{t}{end}}\xspace}

\newcommand{\betaeff}{\ensuremath{\tsub{\beta}{eff}}\xspace}
\newcommand{\Keff}{\ensuremath{\tsub{K}{eff}}\xspace}

\newcommand{\pt}{p} %% primary time
\newcommand{\st}{s} %% secondary time

\newcommand{\psize}{{\mathcal P}} %% primary cohort size
\newcommand{\ssize}{{\mathcal S}} %% secondary cohort size

\newcommand{\gtime}{\sigma} %% generation interval
\newcommand{\gdist}{g} %% generation-interval distribution

\newcommand{\geff}{g_{\textrm{eff}}} %% generation-interval distribution

\newcommand{\total}{{\mathcal T}} %% total number of serial intervals

\newcommand{\PP}{{\mathcal P}}
\newcommand{\II}{{\mathcal I}}

\begin{document}

\begin{flushleft}{
	\Large
	\textbf\newline{
		Characterizing epidemic strength and speed of new SARS-CoV-2 variant of concern
	}
}
\end{flushleft}

\section{Introduction}

\begin{itemize}
  \item People want to estimate strength of new variant from speed
  \item Typically assume same GI distribution
  \item Recent evidence that GI of the new variant might be longer
  \item How does that affect our estimates?
\end{itemize}

Since the emergence of the new SARS-CoV-2 variant of concern (VoC), estimating its epidemic strength and speed remains one of the key questions in controlling the spread.
Epidemic strength, characterized by the reproduction number $\RR$, provides information about the final size of the epidemic as well as the amount of intervention required to eliminate the disease.
Epidemic speed, characterized by the growth rate $r$ provides information about the population-level time scale of the epidemic and, therefore, when the new variant will replace the existing variant.
These two quantities are linked by generation intervals---defined as time between infection and transmission---which define individual-level time scale of the epidemic.

Some studies have tried to estimate the strength of the novel variant from the observed speed.
Other studies have relied on assumed values of the strength of the novel variant and tried to predict the spread (therefore, the speed) of the novel variant.
While both approaches are reasonable, they can lead to different conclusions.
In many contexts, previous studies have shown that inferences about disease dynamics (e.g., effectiveness of intervention strategy) can changes depending on whether $\RR$ or $r$ is held constant.
Such inference also depends on assumptions about the underlying generation-interval distribution.
While many studies assume that the generation-interval distributions do not differ between the existing strain and the variant, recent evidence suggests that the new variant has longer infectious period, leading to longer generation intervals.

Here, we explore how different assumptions about the new variant affects estimates of its strength and speed.
We ask how the estimates of relative strength (i.e., the ratio between the strength of the existing strain and the variant) and speed (i.e., the difference between the speed of the existing strain and the variant) depend on each other.
We then show how neglecting differences in the generation-interval distributions can lead to biased estimates.

\section{Renewal equation framework}

We use the renewal equation framework to characterize the spread of two pathogen strains---in this case, a wild type SARS-CoV-2 and the variant of concern.
Assuming that the effects mutations are negligible, the current incidence of infection $i_x(t)$ from a wild type ($x=w$) and a variant ($x=v$) can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary cases are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
\begin{equation}
i_x = \int_0^\infty i_x(t-\tau) K_x(t, \tau) \dtau.
\end{equation}
This framework provides a flexible way of modeling disease dynamics and generalizes compartmental models, such as the SEIR model.

The integral of the kernel $\RR_x(t) = \int K_x(t, \tau) \dtau$---referred to as the instantaneous reproduction number---describes the average infectiousness of previously infected individuals at time $t$.
The normalized kernel $g_x(t, \tau) = K_x(t, \tau)/\RR_x(t)$---referred to as the instantaneous generation-interval distribution---describes their relative contribution to current incidence.
Both the reproduction number and the generation-interval distribution can depend on several factors, including intrinsic infectiousness of an infected individual, non-pharmaceutical interventions, awareness-driven behavior, and population-level susceptibility (which can be shared between both the wild type and the variant).
For simplicity, we do not explicitly model changes in the kernel.

Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$.
In this case, the incidence of both strains exhibit exponential growth or decay at rate $r_x$, satisfying the Euler-Lotka equation:
\begin{equation}
\frac{1}{\RR_x} = \int_0^\infty \exp(- r_x \tau) g_x(\tau) \dtau.
\end{equation}
Assuming that the generation-interval distribution follows a gamma distribution, we can write the $r$--$\RR$ relationship in terms of the mean generation interval $\bar{G}_x$ and the effective dispersion $\kappa_x$, which captures the amount of variability in generation intervals:
\begin{equation}
\RR_x \approx (1 + \kappa_x r_x \bar{G}_x)^{1/\kappa_x}.
\end{equation}
Here, the effective dispersion parameter $\kappa_x$ also corresponds to the squared coefficient of variation of the gamma distribution.
This gamma approximation framework has been used widely to model epidemic dynamics, including that of SARS-CoV-2.
We use this framework to understand how the inference of strength and speed of the variant depends on our assumptions about the underlying generation-interval distribution as well as the dynamics of the wild type.

\section{Inferring relative strength from relative speed}

Typically, the speed of an epidemic $r_x$ can be measured directly from case data.
But when incidence of new variant is low (due to recent introduction) or unreliable (due to limited testing), estimating growth rate can be challenging.
During the current pandemic, many studies have focused on the characterizing the proportion of the new variant from genetic sequencing data---
when incidence is changing exponentially, the proportion of the new variant $p(t)$ follows a logistic equation:
\begin{align}
p(t) &= \frac{i_v(0) \exp(r_v t)}{i_w(0) \exp(r_w t) + i_v(0) \exp(r_v t)},\\
&= \frac{1}{1 + \left(i_w(0)/i_v(0)\right) \exp(-\delta t)},
\end{align}
where the growth rate of the logistic equation $\delta = r_v - r_w$ corresponds to the relative speed of the epidemic.

\begin{figure}[!th]
\includegraphics[width=\textwidth]{relstrength.pdf}
\caption{
\textbf{Relative strength of the new variant given observed relative speed.}
}
(A) True relative strength of the new variant $\theta$ given speed of the wild type $r_w$ and the relative speed $\delta$.
The dashed contour line corresponds to $\theta = 1$.
True relative strength of the new variant $\theta$ is calculated using $\bar{G}_w = 5\,\textrm{days}$, $\bar{G}_v = 8\,\textrm{days}$, and $\kappa = 1/5$. 
(B) Estimated relative strength of the new variant $\hat{\theta}$ given speed of the wild type
$r_w$ and the relative speed $\delta$ assuming identical generation-interval distributions.
(C) Biases in the estimates of the relative strength of the new variant $\hat{\theta} - \theta$ given speed of the wild type $r_w$ and the relative speed $\delta$.
The dashed contour line corresponds to $\hat{\theta} = \theta$.
\label{fig:relstrength}
\end{figure}

Given an observed value of the relative speed $\delta$, the true relative strength $\theta = \RR_v/\RR_w$ is given by:
\begin{equation}
\theta = \frac{(1 + \kappa_v (r_w + \delta) \bar{G}_v)^{1/\kappa_v}}{(1 + \kappa_w r_w \bar{G}_w)^{1/\kappa_w}}.
\end{equation}
Previous studies have assumed that the generation-interval distributions do not differ between the wild type and the variant:$\kappa_v = \kappa_w$ and $\bar{G}_v = \bar{G}_w$.
In this case, the estimate of the relative strength is given by:
\begin{equation}
\hat{\theta} = \left(1 + \frac{\kappa_w \delta \bar{G}_w}{1 + \kappa_w r_w \bar{G}_w}\right)^{1/\kappa_w}.
\end{equation}
We assume that the true mean generation interval of the new variant is longer that of the wild type and ask how ignoring differences in generation intervals affect estimates of relative strength.

\fref{relstrength}A illustrates the impact of $r_w$ and $\delta$ on the relative strength $\theta$.
Given the observed value of relative speed $\delta$,
the relative strength of the new variant $\theta$ is going to be higher than we thought if the wild type is spreading fast (higher $r_w$).
However, the variant need not have higher reproduction number in order to spread faster: longer generation interval means that the incidence decays at a slower rate given $\RR$. 
Therefore, when both $\RR_w < 1$ and $\RR_v < 1$, this allows the variant to persist better than the wild type even if it has lower strength under some conditions.

\fref{relstrength}B shows estimates of the relative strength $\hat{\theta}$ assuming equal generation-interval distributions.
In this case, the estimate of relative strength $\hat{\theta}$ is relatively insensitive to the underlying speed of the wild type.
However, when the observed relative speed $\delta$ is high, the estimate of relative strength $\hat{\theta}$ decreases as $r_w$ increases---an opposite pattern from the true relationship.
These differences typically lead to underestimation of the relative strength $\theta$ (\fref{relstrength}C);
however, when both the wild type and the variant are exponentially decaying ($r_w < 0$), neglecting differences in the generation-interval distribution leads to overestimation of the relative strength $\theta$.

\section{Inferring relative speed from relative strength}

It is often more common to make assumptions to about the relative strength of the new variant (e.g., assuming that it is more infectious) and try to predict its dynamics.
Then, given assumptions about relative strength $\theta$, relative speed $\delta$ is given by:
\begin{equation}
\delta = \frac{(\theta \RR_w)^{\kappa_v} - 1}{\kappa_v \bar{G}_v} - \frac{\RR_w^{\kappa_w} - 1}{\kappa_w \bar{G}_w}.
\end{equation}
Assuming $\kappa_v = \kappa_w$ and $\bar{G}_v = \bar{G}_w$, the estimate of the relative speed is given by:
\begin{equation}
\hat{\delta} = \frac{(\theta \RR_w)^{\kappa_w} - \RR_w^{\kappa_w}}{\kappa_w \bar{G}_w}.
\end{equation}

\begin{figure}[!th]
\includegraphics[width=\textwidth]{relspeed.pdf}
\caption{
\textbf{Relative speed of the new variant given assumed relative strength.}
}
(A) True relative speed of the new variant $\delta$ given strength of the wild type $\RR_w$ and the relative strength $\theta$.
The dashed contour line corresponds to $\delta = 0$.
True relative speed of the new variant $\delta$ is calculated using $\bar{G}_w = 5\,\textrm{days}$, $\bar{G}_v = 8\,\textrm{days}$, and $\kappa = 1/5$. 
(B) Estimated relative speed of the new variant $\hat{\delta}$ given strength of the wild type $\RR_w$ and the relative strength $\theta$ assuming identical generation-interval distributions.
(C) Biases in the estimates of the relative speed of the new variant $\hat{\delta} - \delta$ given strength of the wild type $\RR_w$ and the relative strength $\theta$.
The dashed contour line corresponds to $\hat{\delta} = \delta$.
\label{fig:relspeed}
\end{figure}

\fref{relspeed}A shows the impact of $\RR_w$ and $\theta$ on the relative speed $\delta$.
As $\theta$ increases, $\delta$ also increases.
However, $\delta$ need not be positive---since longer generation intervals lead to slower speed given epidemic strength, the new variant can spread at a slower rate even if it is more infectious.
In reality, this is unlikely to be case as we are consistently observing increases in the proportion of new variant across multiple countries, suggesting that the relative strength of the new variant is likely high.
When differences in the generation-interval distributions are ignored, our conclusions change: for a given value of $\theta$, higher $\RR_w$ leads to higher $\delta$ (\fref{relspeed}B).
This generally leads to overestimation of $\delta$, except when both $\RR_w < 1$ and $\RR_v < 1$---
longer generation interval means that the variant is going to decline slower than we thought (i.e., $r_v$ closer to 0), leading to underestimation of $\delta$ (\fref{relspeed}C).

\section{Discussion}

Here, we explored how assumptions the relative strength and speed of the new variant affects conclusions about each other when the new variant has longer generation intervals than the existing strain.
When relative speed $\delta$ is held constant, faster growth of the existing strain leads to larger relative strength.
When relative strength $\theta$ is held constant, higher strength of the existing strain leads to smaller relative speed.
Failing to account for the differences in the generation-interval distributions lead to opposite conclusions, leading to biased estimates of $\delta$ and $\theta$.

\end{document}

### 2021-02-21 relspeed.R

@@ -1,0 +1,64 @@
+library(dplyr)
+library(ggplot2); theme_set(theme_bw(base_size=14, base_family = "Times"))
+library(egg)
+library(tikzDevice)
+
+thetavec <- seq(1, 2, length.out=101)
+Rwvec <- seq(0.5, 1.5, length.out=101)
+
+pardata <- expand.grid(thetavec, Rwvec)
+
+Gw <- 5
+Gv <- 8
+
+kappa <- 1/5
+
+deltadata <- apply(pardata, 1, function(x) {
+  theta <- x[[1]]
+  Rw <- x[[2]]
+  
+  delta <- ((theta * Rw)^kappa - 1)/(kappa*Gv) -  ((Rw)^kappa - 1)/(kappa*Gw)
+  
+  deltahat <- ((theta * Rw)^kappa - 1)/(kappa*Gw) -  ((Rw)^kappa - 1)/(kappa*Gw)
+  
+  data.frame(
+    theta=theta,
+    Rw=Rw,
+    delta=delta,
+    deltahat=deltahat,
+    bias=deltahat-delta
+  )
+}) %>%
+  bind_rows
+
+g1 <- ggplot(deltadata) +
+  geom_tile(aes(Rw, theta, fill=delta)) +
+  geom_contour(aes(Rw, theta, z=delta), breaks=c(-0.03, 0.03, 0.06, 0.09), col="white", lty=1) +
+  geom_contour(aes(Rw, theta, z=delta), breaks=0, col="white", lty=2) +
+  annotate("text", x=1.25, y=1.25, label=expression(delta==0), col="white") +
+  scale_x_continuous(expression(Strength~of~the~wild~"types,"~R[w]), expand=c(0, 0)) +
+  scale_y_continuous(expression(Relative~"strength,"~theta), expand=c(0, 0)) +
+  scale_fill_viridis_c(expression(delta))
+
+g2 <- ggplot(deltadata) +
+  geom_tile(aes(Rw, theta, fill=deltahat)) +
+  geom_contour(aes(Rw, theta, z=deltahat), breaks=c(0.03, 0.06, 0.09, 0.12, 0.15), col="white", lty=1) +
+  scale_x_continuous(expression(Strength~of~the~wild~"types,"~R[w]), expand=c(0, 0)) +
+  scale_y_continuous(expression(Relative~"strength,"~theta), expand=c(0, 0)) +
+  scale_fill_viridis_c(expression(hat(delta)), option="A")
+
+g3 <- ggplot(deltadata) +
[…]

### 2021-02-21 relstrength.R

@@ -1,0 +1,66 @@
+library(dplyr)
+library(ggplot2); theme_set(theme_bw(base_size=14, base_family = "Times"))
+library(egg)
+library(tikzDevice)
+
+deltavec <- seq(0, 0.15, length.out=101)
+rwvec <- seq(-0.1, 0.1, length.out=101)
+
+pardata <- expand.grid(deltavec, rwvec)
+
+Gw <- 5
+Gv <- 8
+
+kappa <- 1/5
+
+thetadata <- apply(pardata, 1, function(x) {
+  delta <- x[[1]]
+  rw <- x[[2]]
+  
+  theta <- ((1 + kappa * (rw + delta) * Gv)^(1/kappa))/
+    ((1 + kappa * rw * Gw)^(1/kappa))
+  
+  thetahat <- ((1 + kappa * (rw + delta) * Gw)^(1/kappa))/
+    ((1 + kappa * rw * Gw)^(1/kappa))
+  
+  
+  data.frame(
+    delta=delta,
+    rw=rw,
+    theta=theta,
+    thetahat=thetahat,
+    bias=thetahat-theta
+  )
+}) %>%
+  bind_rows
+
+g1 <- ggplot(thetadata) +
+  geom_tile(aes(rw, delta, fill=theta)) +
+  geom_contour(aes(rw, delta, z=theta), breaks=c(0.8, 1.2, 1.4, 1.6, 1.8, 2, 2.2, 2.4), col="white", lty=1) +
+  geom_contour(aes(rw, delta, z=theta), breaks=c(1), col="white", lty=2) +
+  annotate("text", x=-0.05, y=0.03, label=expression(theta==1), col="white") +
+  scale_x_continuous(expression(Speed~of~the~wild~"types,"~r[w]), expand=c(0, 0)) +
+  scale_y_continuous(expression(Relative~"speed,"~delta), expand=c(0, 0)) +
+  scale_fill_viridis_c(expression(theta))
+
+g2 <- ggplot(thetadata) +
+  geom_tile(aes(rw, delta, fill=thetahat)) +
+  geom_contour(aes(rw, delta, z=thetahat), col="white", lty=1) +
+  scale_x_continuous(expression(Speed~of~the~wild~"types,"~r[w]), expand=c(0, 0)) +
+  scale_y_continuous(expression(Relative~"speed,"~delta), expand=c(0, 0)) +
+  scale_fill_viridis_c(expression(hat(theta)), option="A")
+
[…]

### 2021-02-21 newvariant.tex
fix
moving caption

@@
 
-\begin{itemize}
-  \item People want to estimate strength of new variant from speed
-  \item Typically assume same GI distribution
-  \item Recent evidence that GI of the new variant might be longer
-  \item How does that affect our estimates?
-\end{itemize}
-
 Since the emergence of the new SARS-CoV-2 variant of concern (VoC), estimating its epidemic strength and speed remains one of the key questions in controlling the spread.
@@
 \textbf{Relative strength of the new variant given observed relative speed.}
-}
 (A) True relative strength of the new variant $\theta$ given speed of the wild type $r_w$ and the relative speed $\delta$.
@@
 The dashed contour line corresponds to $\hat{\theta} = \theta$.
+}
 \label{fig:relstrength}
@@
 \textbf{Relative speed of the new variant given assumed relative strength.}
-}
 (A) True relative speed of the new variant $\delta$ given strength of the wild type $\RR_w$ and the relative strength $\theta$.
@@
 The dashed contour line corresponds to $\hat{\delta} = \delta$.
+}
 \label{fig:relspeed}

### 2021-02-21 newvariant.tex
added a bib file but going to take a break

@@
 
-Since the emergence of the new SARS-CoV-2 variant of concern (VoC), estimating its epidemic strength and speed remains one of the key questions in controlling the spread.
+Since the emergence of the new SARS-CoV-2 variant of concern (VoC), estimating its epidemic strength and speed remains one of the key questions in controlling its spread \citep{davies2021estimated, leung2021early, volz2021transmission}.
 Epidemic strength, characterized by the reproduction number $\RR$, provides information about the final size of the epidemic as well as the amount of intervention required to eliminate the disease.

### 2021-02-21 newvariant.tex
fixing errors

--- 2021-02-21 newvariant.tex
@@
 
-Since the emergence of the new SARS-CoV-2 variant of concern (VoC), estimating its epidemic strength and speed remains one of the key questions in controlling the spread.
-Epidemic strength, characterized by the reproduction number $\RR$, provides information about the final size of the epidemic as well as the amount of intervention required to eliminate the disease.
+Since the emergence of the new SARS-CoV-2 variant of concern (VoC), estimating its epidemic strength and speed remains one of the key questions in controlling its spread \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021}.
+Epidemic strength, characterized by the reproduction number $\RR$, provides information about the final size of the epidemic as well as the amount of intervention required to eliminate the disease \citep{anderson1991infectious}.
 Epidemic speed, characterized by the growth rate $r$ provides information about the population-level time scale of the epidemic and, therefore, when the new variant will replace the existing variant.
-These two quantities are linked by generation intervals---defined as time between infection and transmission---which define individual-level time scale of the epidemic.
+These two quantities are linked by generation intervals---defined as time between infection and transmission---which define individual-level time scale of the epidemic \citep{roberts2007model,svensson2007note,wallinga2007generation}.
 
@@
 Here, we explored how assumptions the relative strength and speed of the new variant affects conclusions about each other when the new variant has longer generation intervals than the existing strain.
+We define relative strength as the ratio between reproduction numbers ($\theta=\RR_v/\RR_w$) and relative speed as the difference between growth rates ($\delta=r_v-r_w$).
 When relative speed $\delta$ is held constant, faster growth of the existing strain leads to larger relative strength.
@@
 
+\bibliography{newvariant}
+
 \end{document}

### 2021-02-22 newvariant.tex
discussion

@@
 
+We argue that both perspectives (i.e., whether we hold relative strength or speed constant) are useful in understanding the dynamics of the new variant.
+Early in the spread, we typically want to fix the relative speed because we can directly observe it.
+As more information about the transmission and immunity profiles of the new variant becomes available, it is better to fix the relative strength as it better mathces biological assumptions.
+We suggest researchers to be mindful about which quantity they hold constant and how their assumptions lead to their conclusions.
+
+Our simple framework further reveals counterintuitive, but important, insight.
+In general, one would expect that stronger intervention strategies is required in order to slow down the replacement of the original strain by the new variant.
+However, this is only true if both have the same mean generation intervals (\fref{relspeed}B).
+If the new variant has longer generation intervals than the original strain, imposing strong intervention (therefore reducing both $\RR_w$ and $\RR_v$ by a constant amount) results in a higher relative speed $\delta$, and therefore faster replacement of the original strain (\fref{relspeed}A).
+
+Since the beginning of the SARS-CoV-2 pandemic, the importance of understanding the role of generation-interval distributions in its spread has been constantly emphasized.
+Yet, we still feel that generation intervals are under-appreciated, often viewed as a leeway to estimating $\RR$.
+Future studies should consider the possibility that the new variant might have longer generation intervals and evaluate how assumptions about generation intervals affect their conclusions.
+
 \bibliography{newvariant}

### 2021-02-22 newvariant.tex
more

@@
 	\textbf\newline{
-		Characterizing epidemic strength and speed of new SARS-CoV-2 variant of concern
+		Characterizing relative epidemic strength and speed of new SARS-CoV-2 variants
 	}
@@
 \end{flushleft}
+
+\section*{Abstract}
+
+Characterizing epidemic dynamics of new SARS-CoV-2 variants is critical to predicting the future course of the pandemic.
+Here, we explore how assumptions about the relative strength (i.e., ratio of reproduction numbers $\RR$) and relative speed (i.e., differences in growth rates $r$) of the new variant affect conclusions about each other, given the possibility that the new variant has longer generation intervals.
+When relative speed---which corresponds to the growth rate of the logistic curve that models the proportion of the new variant---is held constant, higher growth rate of the original strain leads to higher relative strength.
+When relative strength is held constant, however, higher reproduction number of the original strain leads to smaller relative speed.
+Therefore, intervention strategies that reduce $\RR$ can cause the variant to replace the original strain faster.
+Neglecting differences in generation intervals lead to opposite conclusions about the relationship between relative strength and speed, biasing their estimates.
+We suggest that estimates of increased transmissibility of new variants should be re-assessed in the light of these findings.
 
@@
 
-Some studies have tried to estimate the strength of the novel variant from the observed speed.
-Other studies have relied on assumed values of the strength of the novel variant and tried to predict the spread (therefore, the speed) of the novel variant.
+Many studies have tried to estimate the strength of the novel variant from the observed speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021}.
+On the other hand, some studies have assumed a value of the strength of the novel variant and tried to predict future spread (therefore, the speed) of the variant \citep{davies2021estimated}.
 While both approaches are reasonable, they can lead to different conclusions.
-In many contexts, previous studies have shown that inferences about disease dynamics (e.g., effectiveness of intervention strategy) can changes depending on whether $\RR$ or $r$ is held constant.
-Such inference also depends on assumptions about the underlying generation-interval distribution.
+In many contexts, previous studies have shown that inferences about disease dynamics (e.g., effectiveness of intervention strategy) can changes depending on whether $\RR$ or $r$ is held constant \citep{dushoff2020speed}.
+Such inference also depends on assumptions about the underlying generation-interval distribution \citep{roberts2007model,svensson2007note,wallinga2007generation}.
 While many studies assume that the generation-interval distributions do not differ between the existing strain and the variant, recent evidence suggests that the new variant has longer infectious period, leading to longer generation intervals.
@@
 If the new variant has longer generation intervals than the original strain, imposing strong intervention (therefore reducing both $\RR_w$ and $\RR_v$ by a constant amount) results in a higher relative speed $\delta$, and therefore faster replacement of the original strain (\fref{relspeed}A).
+While it is still necessary to impose stronger intervention to control the epidemic if the new variant has higher $\RR$, it may have unexpected consequences.
 

### 2021-02-27 newvariant.tex
Trying again to commit

@@
 The normalized kernel $g_x(t, \tau) = K_x(t, \tau)/\RR_x(t)$---referred to as the instantaneous generation-interval distribution---describes their relative contribution to current incidence.
-Both the reproduction number and the generation-interval distribution can depend on several factors, including intrinsic infectiousness of an infected individual, non-pharmaceutical interventions, awareness-driven behavior, and population-level susceptibility (which can be shared between both the wild type and the variant).
-For simplicity, we do not explicitly model changes in the kernel.
+Both the reproduction number and the generation-interval distribution can depend on several factors, including intrinsic infectiousness of an infected individual, non-pharmaceutical interventions, awareness-driven behavior, and population-level susceptibility.
+Many of these factors can affect between both the wild type and the variant in similar ways.
 
@@
 \end{equation}
-Assuming that the generation-interval distribution follows a gamma distribution, we can write the $r$--$\RR$ relationship in terms of the mean generation interval $\bar{G}_x$ and the effective dispersion $\kappa_x$, which captures the amount of variability in generation intervals:
+We can approximate the $r$--$\RR$ relationship by assuming that the generation-interval distribution is approximately gamma-distributed, and summarizing it using the mean generation interval $\bar{G}_x$ and the squared coefficient of variation $\kappa_x$:
 \begin{equation}
@@
 \end{equation}
-Here, the effective dispersion parameter $\kappa_x$ also corresponds to the squared coefficient of variation of the gamma distribution.
-This gamma approximation framework has been used widely to model epidemic dynamics, including that of SARS-CoV-2.
-We use this framework to understand how the inference of strength and speed of the variant depends on our assumptions about the underlying generation-interval distribution as well as the dynamics of the wild type.
+The gamma assumption covers models that assume exponentially distributed generation intervals (when $\kappa=1$); various gamma assumptions and approximations are widely used in epidemic modeling, including for modelsthat of SARS-CoV-2.
+We use this framework to investigate how inferences about strength and speed of the variant depend on our assumptions about the underlying generation-interval distributions.
+
+\jd{I like to talk about `effective dispersion' when we're getting into the details of distributions and how good the gamma assumption is.
+I think at this level we're OK with saying we approximate by a gamma and kappa is the squared CV.}
 
@@
 
-Typically, the speed of an epidemic $r_x$ can be measured directly from case data.
-But when incidence of new variant is low (due to recent introduction) or unreliable (due to limited testing), estimating growth rate can be challenging.
-During the current pandemic, many studies have focused on the characterizing the proportion of the new variant from genetic sequencing data---
-when incidence is changing exponentially, the proportion of the new variant $p(t)$ follows a logistic equation:
+Epidemic speed $r_x$ is often estimated directly from case data.
+Studies of new SARS-CoV-2 variants, however, have mostly focused on characterizing changes in \emph{proportion} of a new variant.
+This has the advantage of correcting for testing patterns and intensity, and possibly for other transient effects that might affect variants and wildtype viruses similarly.
+When incidence is changing exponentially, the proportion of the new variant $p(t)$ follows a logistic equation:
 \begin{align}
-p(t) &= \frac{i_v(0) \exp(r_v t)}{i_w(0) \exp(r_w t) + i_v(0) \exp(r_v t)},\\
-&= \frac{1}{1 + \left(i_w(0)/i_v(0)\right) \exp(-\delta t)},
+p(t) &= \frac{i_v(0) \exp(r_v t)}{i_w(0) \exp(r_w t) + i_v(0) \exp(r_v t)},
+\\ &= \frac{1}{1 + \left(i_w(0)/i_v(0)\right) \exp(-\delta t)},
 \end{align}
-where the growth rate of the logistic equation $\delta = r_v - r_w$ corresponds to the relative speed of the epidemic.
+where the logistic growth rate $\delta = r_v - r_w$ corresponds to the relative speed of the epidemic. \jd{I now think it's easier to leave the second line in than to explain it in words.}
 

### 2021-02-27 newvariant.tex
edits

@@
 
-Since the emergence of the new SARS-CoV-2 variant of concern (VoC), estimating its epidemic strength and speed remains one of the key questions in controlling its spread \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021}.
+Since the emergence of the new SARS-CoV-2 variant of concern (VoC), estimating its epidemic strength and speed remains one of the key questions in controlling its spread \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
 Epidemic strength, characterized by the reproduction number $\RR$, provides information about the final size of the epidemic as well as the amount of intervention required to eliminate the disease \citep{anderson1991infectious}.
@@
 
-We use the renewal equation framework to characterize the spread of two pathogen strains---in this case, a wild type SARS-CoV-2 and the variant of concern.
-Assuming that the effects mutations are negligible, the current incidence of infection $i_x(t)$ from a wild type ($x=w$) and a variant ($x=v$) can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary cases are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
+We use the renewal equation framework to characterize the spread of two pathogen strains---in this case, the wild type SARS-CoV-2 virus and the variant of concern.
+Assuming that the effects mutations are negligible, the current incidence of infection $i_x(t)$ caused by the wild type ($x=w$) and the variant ($x=v$) can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary cases are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
 \begin{equation}
-i_x = \int_0^\infty i_x(t-\tau) K_x(t, \tau) \dtau.
+i_x(t) = \int_0^\infty i_x(t-\tau) K_x(t, \tau) \dtau.
 \end{equation}
-This framework provides a flexible way of modeling disease dynamics and generalizes compartmental models, such as the SEIR model.
+This framework provides a flexible way of modeling disease dynamics and generalizes compartmental models, such as the SEIR model \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, roberts2007model, champredon2018equivalence}.
 
-The integral of the kernel $\RR_x(t) = \int K_x(t, \tau) \dtau$---referred to as the instantaneous reproduction number---describes the average infectiousness of previously infected individuals at time $t$.
-The normalized kernel $g_x(t, \tau) = K_x(t, \tau)/\RR_x(t)$---referred to as the instantaneous generation-interval distribution---describes their relative contribution to current incidence.
+The integral of the kernel $\RR_x(t) = \int K_x(t, \tau) \dtau$---referred to as the instantaneous reproduction number---describes the average infectiousness of previously infected individuals at time $t$ \citep{fraser2007estimating}.
+The normalized kernel $g_x(t, \tau) = K_x(t, \tau)/\RR_x(t)$---which we refer to as the instantaneous generation-interval distribution---describes their relative contribution to current incidence $i_x(t)$ and provides information about the time scale of disease transmission.
 Both the reproduction number and the generation-interval distribution can depend on several factors, including intrinsic infectiousness of an infected individual, non-pharmaceutical interventions, awareness-driven behavior, and population-level susceptibility.
@@
 Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$.
-In this case, the incidence of both strains exhibit exponential growth or decay at rate $r_x$, satisfying the Euler-Lotka equation:
+In this case, the incidence of both strains exhibit exponential growth or decay at rate $r_x$, satisfying the Euler-Lotka equation \citep{wallinga2007generation}:
 \begin{equation}
@@
 \end{equation}
-We can approximate the $r$--$\RR$ relationship by assuming that the generation-interval distribution is approximately gamma-distributed, and summarizing it using the mean generation interval $\bar{G}_x$ and the squared coefficient of variation $\kappa_x$:
+We can approximate the $r$--$\RR$ relationship by assuming that the generation-interval distribution is approximately gamma-distributed, and summarizing it using the mean generation interval $\bar{G}_x$ and the squared coefficient of variation $\kappa_x$ \citep{park2019practical}:
 \begin{equation}
@@
 \end{equation}
-The gamma assumption covers models that assume exponentially distributed generation intervals (when $\kappa=1$); various gamma assumptions and approximations are widely used in epidemic modeling, including for modelsthat of SARS-CoV-2.
+The gamma assumption covers models that assume exponentially distributed generation intervals (when $\kappa=1$), corresponding to the SIR model \citep{anderson1991infectious}; various gamma assumptions and approximations are widely used in epidemic modeling, including for models of SARS-CoV-2 \citep{doi:10.1098/rsif.2020.0144}.
 We use this framework to investigate how inferences about strength and speed of the variant depend on our assumptions about the underlying generation-interval distributions.
@@
 
-Epidemic speed $r_x$ is often estimated directly from case data.
-Studies of new SARS-CoV-2 variants, however, have mostly focused on characterizing changes in \emph{proportion} of a new variant.
+Epidemic speed $r_x$ is often estimated directly from case data \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}.
+Studies of new SARS-CoV-2 variants, however, have mostly focused on characterizing changes in \emph{proportion} of a new variant \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
 This has the advantage of correcting for testing patterns and intensity, and possibly for other transient effects that might affect variants and wildtype viruses similarly.
@@
 \end{align}
-where the logistic growth rate $\delta = r_v - r_w$ corresponds to the relative speed of the epidemic. \jd{I now think it's easier to leave the second line in than to explain it in words.}
+where the logistic growth rate $\delta = r_v - r_w$ corresponds to the relative speed of the epidemic.
 
@@
 
+We ask how the inference of relative strength $\theta = \RR_v/\RR_w$ of the new variant from its relative speed $\delta$ depends on assumptions about the generation-interval distribution.
+As our baseline scenario, we assume $\delta = 0.1/\textrm{days}$ to match the observed relative speed of the new variant in the UK \citep{davies2021estimated}.
+
+If we assume that the both the wild type and the variant have identical generation-interval distributions, the estimate of the relative strength is given by:
+\begin{equation}
+\hat{\theta} = \left(1 + \frac{\kappa \delta \bar{G}}{1 + \kappa r_w \bar{G}}\right)^{1/\kappa}.
+\end{equation}
+
 Given an observed value of the relative speed $\delta$, the true relative strength $\theta = \RR_v/\RR_w$ is given by:
@@
 \end{equation}
-Previous studies have assumed that the generation-interval distributions do not differ between the wild type and the variant:$\kappa_v = \kappa_w$ and $\bar{G}_v = \bar{G}_w$.
+Previous studies have assumed that the generation-interval distributions do not differ between the wild type and the variant: $\kappa_v = \kappa_w = \kappa$ and $\bar{G}_v = \bar{G}_w = \bar{G}$.
 In this case, the estimate of the relative strength is given by:

### 2021-02-27 newvariant.tex
more

@@
 We use this framework to investigate how inferences about strength and speed of the variant depend on our assumptions about the underlying generation-interval distributions.
+For simplicity we neglect differences in the squared coefficient of variation and assume $\kappa_w = \kappa_v = \kappa$; instead, we focus on the effect of potential differences in the mean generation intervals.
 
@@
 
-We ask how the inference of relative strength $\theta = \RR_v/\RR_w$ of the new variant from its relative speed $\delta$ depends on assumptions about the generation-interval distribution.
-As our baseline scenario, we assume $\delta = 0.1/\textrm{days}$ to match the observed relative speed of the new variant in the UK \citep{davies2021estimated}.
+As our baseline scenario, we assume $\delta = 0.1/\textrm{days}$ based on the observed relative speed of the new variant in the UK \citep{davies2021estimated}.
+We then ask how the inference of relative strength $\theta = \RR_v/\RR_w$ of the new variant from its relative speed $\delta$ depends on assumptions about the generation-interval distribution.
+The mean generation interval $\bar{G}_w$ of the wild type is assumed to equal 5 days \citep{ferretti2020quantifying}.
+The squared coefficient of variation $\kappa$ of the generation-interval distribution is assumed to equal 0.2 \citep{ferretti2020quantifying};
+however, we expect there to be greater uncertainty in $\kappa$ than in $\bar{G}$, and so we evaluate the estimates of relative strength $\theta$ across a wide range of $\kappa$, ranging from 0 (delta distribution) to 1 (exponential distribution).
 
-If we assume that the both the wild type and the variant have identical generation-interval distributions, the estimate of the relative strength is given by:
+Previous studies have assumed that the generation-interval distributions do not differ between the wild type and the variant: $\bar{G}_v = \bar{G}_w = \bar{G}$.
+In this case, the estimate of the relative strength is given by:
 \begin{equation}

### 2021-02-27 newvariant.tex
new figure 1 and off to bed maybe

@@
 The mean generation interval $\bar{G}_w$ of the wild type is assumed to equal 5 days \citep{ferretti2020quantifying}.
+As we expect there to be greater uncertainty in $\kappa$ than in $\bar{G}$, we evaluate the estimates of relative strength $\theta$ across a wide range of $\kappa$, ranging from 0 (delta distribution) to 1 (exponential distribution);
+estimates of $\kappa$ range
+
 The squared coefficient of variation $\kappa$ of the generation-interval distribution is assumed to equal 0.2 \citep{ferretti2020quantifying};
-however, we expect there to be greater uncertainty in $\kappa$ than in $\bar{G}$, and so we evaluate the estimates of relative strength $\theta$ across a wide range of $\kappa$, ranging from 0 (delta distribution) to 1 (exponential distribution).
+however, we
 
@@
 \begin{equation}
-\hat{\theta} = \left(1 + \frac{\kappa \delta \bar{G}}{1 + \kappa r_w \bar{G}}\right)^{1/\kappa}.
+\theta = \left(1 + \frac{\kappa \delta \bar{G}}{1 + \kappa r_w \bar{G}}\right)^{1/\kappa}.
 \end{equation}

### 2021-02-27 relstrength.R

@@ -1,66 +1,86 @@
 library(dplyr)
 library(ggplot2); theme_set(theme_bw(base_size=14, base_family = "Times"))
+library(ggthemes)
 library(egg)
 library(tikzDevice)
 
-deltavec <- seq(0, 0.15, length.out=101)
-rwvec <- seq(-0.1, 0.1, length.out=101)
+delta <- 0.1
 
-pardata <- expand.grid(deltavec, rwvec)
+rwvec <- seq(-0.13, 0.13, length.out=21)
+kappavec <- c(0, 0.2, 0.5, 1)
+
+pardata <- expand.grid(kappavec, rwvec)
 
 Gw <- 5
-Gv <- 8
-
-kappa <- 1/5
+Gv <- 7
 
 thetadata <- apply(pardata, 1, function(x) {
-  delta <- x[[1]]
+  kappa <- x[[1]]
   rw <- x[[2]]
   
-  theta <- ((1 + kappa * (rw + delta) * Gv)^(1/kappa))/
-    ((1 + kappa * rw * Gw)^(1/kappa))
-  
-  thetahat <- ((1 + kappa * (rw + delta) * Gw)^(1/kappa))/
-    ((1 + kappa * rw * Gw)^(1/kappa))
-  
+  if (kappa==0) {
+    
+    thetahat <- exp(Gv * (rw + delta))/exp(Gw * rw)
+    
+    theta <- exp(Gw * (rw + delta))/exp(Gw * rw)
+  } else {
+    thetahat <- ((1 + kappa * (rw + delta) * Gv)^(1/kappa))/
+      ((1 + kappa * rw * Gw)^(1/kappa))
+    
+    theta <- ((1 + kappa * (rw + delta) * Gw)^(1/kappa))/
+      ((1 + kappa * rw * Gw)^(1/kappa))
+  }
   
   data.frame(
     delta=delta,
     rw=rw,
     theta=theta,
     thetahat=thetahat,
-    bias=thetahat-theta
+    bias=thetahat/theta,
+    kappa=kappa
   )
 }) %>%
   bind_rows
 
 g1 <- ggplot(thetadata) +
-  geom_tile(aes(rw, delta, fill=theta)) +
-  geom_contour(aes(rw, delta, z=theta), breaks=c(0.8, 1.2, 1.4, 1.6, 1.8, 2, 2.2, 2.4), col="white", lty=1) +
[…]

### 2021-02-27 relstrength.R
change

@@ -51,7 +51,7 @@ g1 <- ggplot(thetadata) +
   ggtitle("A. Equal generation intervals") +
   theme(
     panel.grid = element_blank(),
-    legend.position = c(0.8, 0.7)
+    legend.position = c(0.8, 0.75)
   )
 
 g2 <- ggplot(thetadata) +

### 2021-02-27 newvariant.tex
tweaks

@@
 The mean generation interval $\bar{G}_w$ of the wild type is assumed to equal 5 days \citep{ferretti2020quantifying}.
-As we expect there to be greater uncertainty in $\kappa$ than in $\bar{G}$, we evaluate the estimates of relative strength $\theta$ across a wide range of $\kappa$, ranging from 0 (delta distribution) to 1 (exponential distribution);
-estimates of $\kappa$ range
-
-The squared coefficient of variation $\kappa$ of the generation-interval distribution is assumed to equal 0.2 \citep{ferretti2020quantifying};
-however, we
+Previous estimates of squared coefficient of variation $\kappa$ in generation intervals are roughly to equal 0.2 \citep{ferretti2020quantifying}.
+However, we expect there to be greater uncertainty in $\kappa$ than in $\bar{G}$, and so we evaluate the estimates of relative strength $\theta$ across a wide range of $\kappa$, ranging from 0 (delta distribution) to 1 (exponential distribution).
 
@@
 \end{equation}
+Therefore, if the wild type is spreading faster than we previously thought (higher $r_w$), we expect our estimate of the relative strength $\theta$ to go down (\fref{relstrength}A).
+The strength
 

### 2021-02-27 newvariant.tex
really need to go to bed

@@
 \textbf{Relative strength of the new variant given observed relative speed.}
-(A) True relative strength of the new variant $\theta$ given speed of the wild type $r_w$ and the relative speed $\delta$.
-The dashed contour line corresponds to $\theta = 1$.
-True relative strength of the new variant $\theta$ is calculated using $\bar{G}_w = 5\,\textrm{days}$, $\bar{G}_v = 8\,\textrm{days}$, and $\kappa = 1/5$. 
-(B) Estimated relative strength of the new variant $\hat{\theta}$ given speed of the wild type
-$r_w$ and the relative speed $\delta$ assuming identical generation-interval distributions.
-(C) Biases in the estimates of the relative strength of the new variant $\hat{\theta} - \theta$ given speed of the wild type $r_w$ and the relative speed $\delta$.
-The dashed contour line corresponds to $\hat{\theta} = \theta$.
+(A) Estimated relative strength of the new variant $\hat{\theta}$ given speed of the wild type
+$r_w$ assuming identical generation-interval distributions.
+(B) True relative strength of the new variant $\theta$ given speed of the wild type $r_w$ accounting for longer generation intervals of the new variant.
+(C) Ratio between the true and the estimated relative strength of the new variant $\theta/\hat{\theta}$.
+True relative strength of the new variant $\theta$ is calculated using $\bar{G}_w = 5\,\textrm{days}$, and $\bar{G}_v = 7\,\textrm{days}$.
 }
@@
 \begin{equation}
-\theta = \left(1 + \frac{\kappa \delta \bar{G}}{1 + \kappa r_w \bar{G}}\right)^{1/\kappa}.
+\hat{\theta} = \left(1 + \frac{\kappa \delta \bar{G}}{1 + \kappa r_w \bar{G}}\right)^{1/\kappa}.
 \end{equation}
 Therefore, if the wild type is spreading faster than we previously thought (higher $r_w$), we expect our estimate of the relative strength $\theta$ to go down (\fref{relstrength}A).
-The strength
+The relative strength also depends on the \emph{effective dispersion} in generation intervals (i.e., the amount of variability in generation intervals, as measured by $\kappa$)---using narrower distributions makes $\hat{\theta}$ less sensitive to $r_w$.
 
-Given an observed value of the relative speed $\delta$, the true relative strength $\theta = \RR_v/\RR_w$ is given by:
+However, if the mean generation interval of the new variant is different from that of the wild type, the true relative strength $\theta = \RR_v/\RR_w$ is given by:
 \begin{equation}
-\theta = \frac{(1 + \kappa_v (r_w + \delta) \bar{G}_v)^{1/\kappa_v}}{(1 + \kappa_w r_w \bar{G}_w)^{1/\kappa_w}}.
+\theta = \frac{(1 + \kappa (r_w + \delta) \bar{G}_v)^{1/\kappa}}{(1 + \kappa r_w \bar{G}_w)^{1/\kappa}}.
 \end{equation}
-Previous studies have assumed that the generation-interval distributions do not differ between the wild type and the variant: $\kappa_v = \kappa_w = \kappa$ and $\bar{G}_v = \bar{G}_w = \bar{G}$.
-In this case, the estimate of the relative strength is given by:
-\begin{equation}
-\hat{\theta} = \left(1 + \frac{\kappa_w \delta \bar{G}_w}{1 + \kappa_w r_w \bar{G}_w}\right)^{1/\kappa_w}.
-\end{equation}
-We assume that the true mean generation interval of the new variant is longer that of the wild type and ask how ignoring differences in generation intervals affect estimates of relative strength.
+\fref{relstrength}B illustrates the impact of $r_w$ on the ``true'' relative strength $\theta$ given $\bar{G}_v=7\,\textrm{days}$.
 
-\fref{relstrength}A illustrates the impact of $r_w$ and $\delta$ on the relative strength $\theta$.
 Given the observed value of relative speed $\delta$,

### 2021-02-27 relstrength.R

@@ -20,14 +20,14 @@ thetadata <- apply(pardata, 1, function(x) {
   
   if (kappa==0) {
     
-    thetahat <- exp(Gv * (rw + delta))/exp(Gw * rw)
+    theta <- exp(Gv * (rw + delta))/exp(Gw * rw)
     
-    theta <- exp(Gw * (rw + delta))/exp(Gw * rw)
+    thetahat <- exp(Gw * (rw + delta))/exp(Gw * rw)
   } else {
-    thetahat <- ((1 + kappa * (rw + delta) * Gv)^(1/kappa))/
+    theta <- ((1 + kappa * (rw + delta) * Gv)^(1/kappa))/
       ((1 + kappa * rw * Gw)^(1/kappa))
     
-    theta <- ((1 + kappa * (rw + delta) * Gw)^(1/kappa))/
+    thetahat <- ((1 + kappa * (rw + delta) * Gw)^(1/kappa))/
       ((1 + kappa * rw * Gw)^(1/kappa))
   }
   
@@ -36,16 +36,16 @@ thetadata <- apply(pardata, 1, function(x) {
     rw=rw,
     theta=theta,
     thetahat=thetahat,
-    bias=thetahat/theta,
+    bias=theta/thetahat,
     kappa=kappa
   )
 }) %>%
   bind_rows
 
 g1 <- ggplot(thetadata) +
-  geom_line(aes(rw, theta, lty=as.factor(kappa), col=as.factor(kappa))) +
+  geom_line(aes(rw, thetahat, lty=as.factor(kappa), col=as.factor(kappa))) +
   scale_x_continuous("Relative speed, $r_w$ (1/days)", expand=c(0, 0)) +
-  scale_y_continuous("Relative strength, $\\theta$", limits=c(1, 2.7), expand=c(0, 0)) +
+  scale_y_continuous("Estimated relative strength, $\\hat{\\theta}$", limits=c(1, 2.7), expand=c(0, 0)) +
   scale_color_colorblind("$\\kappa$") +
   scale_linetype_discrete("$\\kappa$") +
   ggtitle("A. Equal generation intervals") +
@@ -55,9 +55,9 @@ g1 <- ggplot(thetadata) +
   )
 
 g2 <- ggplot(thetadata) +
-  geom_line(aes(rw, thetahat, lty=as.factor(kappa), col=as.factor(kappa))) +
+  geom_line(aes(rw, theta, lty=as.factor(kappa), col=as.factor(kappa))) +
[…]

### 2021-02-27 newvariant.tex
fig 1 done

@@
 
-However, if the mean generation interval of the new variant is different from that of the wild type, the true relative strength $\theta = \RR_v/\RR_w$ is given by:
+However, if the mean generation interval of the new variant is longer than that of the wild type ($\bar{G}_v > \bar{G}_w$), the true relative strength $\theta = \RR_v/\RR_w$ is given by:
 \begin{equation}
-\theta = \frac{(1 + \kappa (r_w + \delta) \bar{G}_v)^{1/\kappa}}{(1 + \kappa r_w \bar{G}_w)^{1/\kappa}}.
+\theta = \left(\frac{1 + \kappa (r_w + \delta) \bar{G}_v}{1 + \kappa r_w \bar{G}_w}\right)^{1/\kappa}.
 \end{equation}
-\fref{relstrength}B illustrates the impact of $r_w$ on the ``true'' relative strength $\theta$ given $\bar{G}_v=7\,\textrm{days}$.
+In this case, the effect of the underlying speed $r_w$ on the relative strength $\theta$ can change directions---\fref{relstrength}B illustrates this effect assuming $\bar{G}_v=7\,\textrm{days}$.
+As $\kappa \to 0$, this approaches
+\begin{equation}
+\theta \to \exp(r_w (\bar{G}_v-\bar{G}_w)+\delta G_v), 
+\end{equation}
+meaning that $\theta$ increases with $r_w$ (\fref{relstrength}B).
+On the other hand, as $\kappa \to 1$, we have
+\begin{equation}
+\theta \to \frac{1 + (r_w + \delta) \bar{G}_v}{1 + r_w \bar{G}_w} = \frac{\bar{G}_v}{\bar{G}_w} + \frac{1+ \delta \bar{G}_v  - \bar{G}_v/\bar{G}_w}{1 + r_w \bar{G}_w},
+\end{equation}
+which decreases with $r_w$ (\fref{relstrength}B).
 
-Given the observed value of relative speed $\delta$,
-the relative strength of the new variant $\theta$ is going to be higher than we thought if the wild type is spreading fast (higher $r_w$).
-However, the variant need not have higher reproduction number in order to spread faster: longer generation interval means that the incidence decays at a slower rate given $\RR$. 
-Therefore, when both $\RR_w < 1$ and $\RR_v < 1$, this allows the variant to persist better than the wild type even if it has lower strength under some conditions.
-
-\fref{relstrength}B shows estimates of the relative strength $\hat{\theta}$ assuming equal generation-interval distributions.
-In this case, the estimate of relative strength $\hat{\theta}$ is relatively insensitive to the underlying speed of the wild type.
-However, when the observed relative speed $\delta$ is high, the estimate of relative strength $\hat{\theta}$ decreases as $r_w$ increases---an opposite pattern from the true relationship.
-These differences typically lead to underestimation of the relative strength $\theta$ (\fref{relstrength}C);
-however, when both the wild type and the variant are exponentially decaying ($r_w < 0$), neglecting differences in the generation-interval distribution leads to overestimation of the relative strength $\theta$.
+Therefore, neglecting potential differences in generation intervals generally leads to under-estimation of the relative strength (\fref{relstrength}C).
+In some cases, this can lead to over-estimation (see $r_w < -0.1$ in \fref{relstrength}C):
+Longer generation intervals lead to slower exponential decay when $\RR < 1$, meaning that the new variant can still increase in proportion ($\delta > 0$) even when $\RR_v < \RR_w$.
+However, this is unlikely to be the case as the new variant spread quickly under mixed conditions.
+Instead, it seems more plausible that current estimates have under-estimated the true relative strength by 1.1---1.4 fold.
 

### 2021-02-27 newvariant.tex
new fig 1 and 2

@@
 
+While relative speed is easier to measure, we often do not expect $\delta$ to be constant---this can depend on epidemic and demographic factors and can vary across time and space.
+Instead, assuming a constant relative strength $\theta$ better matches biological hypotheses about the new variant (e.g., increased transmission and immune escape).
+Several studies have tried to predict the future spread of the new variant, including when it will replace the wild type (therefore, its relative speed $\delta$).
+As our baseline scenario we assume $\theta = 1.8$ and ask how assumptions about the generation-interval distribution affect predictions of $\delta$.
+
+Like others, we first ignore possible differences between the generation-interval distributions of the wild type and the new variant and try to estimate $\delta$:
+\begin{equation}
+\hat{\delta} = \frac{(\theta \RR_w)^{\kappa_w} - \RR_w^{\kappa_w}}{\kappa_w \bar{G}_w}.
+\end{equation}
+
+Given an assumed value of the relative strength $\theta$,
+
 It is often more common to make assumptions to about the relative strength of the new variant (e.g., assuming that it is more infectious) and try to predict its dynamics.

### 2021-02-27 relspeed.R

@@ -1,64 +1,85 @@
 library(dplyr)
 library(ggplot2); theme_set(theme_bw(base_size=14, base_family = "Times"))
+library(ggthemes)
 library(egg)
 library(tikzDevice)
 
-thetavec <- seq(1, 2, length.out=101)
-Rwvec <- seq(0.5, 1.5, length.out=101)
+theta <- 1.8
 
-pardata <- expand.grid(thetavec, Rwvec)
+Rwvec <- seq(0.3, 1.7, length.out=21)
+kappavec <- c(0, 0.2, 0.5, 1)
+
+pardata <- expand.grid(kappavec, Rwvec)
 
 Gw <- 5
-Gv <- 8
-
-kappa <- 1/5
+Gv <- 7
 
 deltadata <- apply(pardata, 1, function(x) {
-  theta <- x[[1]]
+  kappa <- x[[1]]
   Rw <- x[[2]]
   
-  delta <- ((theta * Rw)^kappa - 1)/(kappa*Gv) -  ((Rw)^kappa - 1)/(kappa*Gw)
-  
-  deltahat <- ((theta * Rw)^kappa - 1)/(kappa*Gw) -  ((Rw)^kappa - 1)/(kappa*Gw)
+  if (kappa == 0) {
+    delta <- log(theta * Rw)/Gv - log(Rw)/Gw
+    
+    deltahat <- log(theta * Rw)/Gw - log(Rw)/Gw
+  } else {
+    delta <- ((theta * Rw)^kappa - 1)/(kappa*Gv) -  ((Rw)^kappa - 1)/(kappa*Gw)
+    
+    deltahat <- ((theta * Rw)^kappa - 1)/(kappa*Gw) -  ((Rw)^kappa - 1)/(kappa*Gw)
+    
+  }
   
   data.frame(
     theta=theta,
     Rw=Rw,
     delta=delta,
     deltahat=deltahat,
-    bias=deltahat-delta
+    bias=delta-deltahat,
+    kappa=kappa
   )
 }) %>%
   bind_rows
 
 g1 <- ggplot(deltadata) +
-  geom_tile(aes(Rw, theta, fill=delta)) +
-  geom_contour(aes(Rw, theta, z=delta), breaks=c(-0.03, 0.03, 0.06, 0.09), col="white", lty=1) +
-  geom_contour(aes(Rw, theta, z=delta), breaks=0, col="white", lty=2) +
[…]

### 2021-02-27 relstrength.R

@@ -44,7 +44,7 @@ thetadata <- apply(pardata, 1, function(x) {
 
 g1 <- ggplot(thetadata) +
   geom_line(aes(rw, thetahat, lty=as.factor(kappa), col=as.factor(kappa))) +
-  scale_x_continuous("Relative speed, $r_w$ (1/days)", expand=c(0, 0)) +
+  scale_x_continuous("Wild type speed, $r_w$ (1/days)", expand=c(0, 0)) +
   scale_y_continuous("Estimated relative strength, $\\hat{\\theta}$", limits=c(1, 2.7), expand=c(0, 0)) +
   scale_color_colorblind("$\\kappa$") +
   scale_linetype_discrete("$\\kappa$") +
@@ -56,7 +56,7 @@ g1 <- ggplot(thetadata) +
 
 g2 <- ggplot(thetadata) +
   geom_line(aes(rw, theta, lty=as.factor(kappa), col=as.factor(kappa))) +
-  scale_x_continuous("Relative speed, $r_w$ (1/days)", expand=c(0, 0)) +
+  scale_x_continuous("Wild type speed, $r_w$ (1/days)", expand=c(0, 0)) +
   scale_y_continuous("True relative strength, $\\theta$", limits=c(1, 2.7), expand=c(0, 0)) +
   scale_color_colorblind() +
   ggtitle("B. Longer generation intervals") +
@@ -68,8 +68,8 @@ g2 <- ggplot(thetadata) +
 
 g3 <- ggplot(thetadata) +
   geom_line(aes(rw, bias, lty=as.factor(kappa), col=as.factor(kappa))) +
-  scale_x_continuous("Relative speed, $r_w$ (1/days)", expand=c(0, 0)) +
-  scale_y_continuous("Bias, $\\theta/\\hat{\\theta}$", limits=c(0.9, 1.61), expand=c(0, 0)) +
+  scale_x_continuous("Wild type speed, $r_w$ (1/days)", expand=c(0, 0)) +
+  scale_y_continuous("Changes in estimates, $\\theta/\\hat{\\theta}$", limits=c(0.9, 1.61), expand=c(0, 0)) +
   scale_color_colorblind() +
   ggtitle("C. Bias") +
   theme(

### 2021-02-27 newvariant.tex
discussion

@@
 (C) Ratio between the true and the estimated relative strength of the new variant $\theta/\hat{\theta}$.
-True relative strength of the new variant $\theta$ is calculated using $\bar{G}_w = 5\,\textrm{days}$, and $\bar{G}_v = 7\,\textrm{days}$.
+True relative strength of the new variant $\theta$ is calculated using $\delta=0.1\,\textrm{days}$, $\bar{G}_w = 5\,\textrm{days}$, and $\bar{G}_v = 7\,\textrm{days}$.
 }
@@
 
-Like others, we first ignore possible differences between the generation-interval distributions of the wild type and the new variant and try to estimate $\delta$:
-\begin{equation}
-\hat{\delta} = \frac{(\theta \RR_w)^{\kappa_w} - \RR_w^{\kappa_w}}{\kappa_w \bar{G}_w}.
-\end{equation}
-
-Given an assumed value of the relative strength $\theta$,
-
-It is often more common to make assumptions to about the relative strength of the new variant (e.g., assuming that it is more infectious) and try to predict its dynamics.
-Then, given assumptions about relative strength $\theta$, relative speed $\delta$ is given by:
-\begin{equation}
-\delta = \frac{(\theta \RR_w)^{\kappa_v} - 1}{\kappa_v \bar{G}_v} - \frac{\RR_w^{\kappa_w} - 1}{\kappa_w \bar{G}_w}.
-\end{equation}
-Assuming $\kappa_v = \kappa_w$ and $\bar{G}_v = \bar{G}_w$, the estimate of the relative speed is given by:
-\begin{equation}
-\hat{\delta} = \frac{(\theta \RR_w)^{\kappa_w} - \RR_w^{\kappa_w}}{\kappa_w \bar{G}_w}.
-\end{equation}
-
 \begin{figure}[!th]
@@
 
-\fref{relspeed}A shows the impact of $\RR_w$ and $\theta$ on the relative speed $\delta$.
-As $\theta$ increases, $\delta$ also increases.
-However, $\delta$ need not be positive---since longer generation intervals lead to slower speed given epidemic strength, the new variant can spread at a slower rate even if it is more infectious.
-In reality, this is unlikely to be case as we are consistently observing increases in the proportion of new variant across multiple countries, suggesting that the relative strength of the new variant is likely high.
-When differences in the generation-interval distributions are ignored, our conclusions change: for a given value of $\theta$, higher $\RR_w$ leads to higher $\delta$ (\fref{relspeed}B).
-This generally leads to overestimation of $\delta$, except when both $\RR_w < 1$ and $\RR_v < 1$---
-longer generation interval means that the variant is going to decline slower than we thought (i.e., $r_v$ closer to 0), leading to underestimation of $\delta$ (\fref{relspeed}C).
+Like others, we first ignore possible differences between the generation-interval distributions of the wild type and the new variant and try to estimate $\delta$:
+\begin{equation}
+\hat{\delta} = \frac{(\theta \RR_w)^{\kappa_w} - \RR_w^{\kappa_w}}{\kappa_w \bar{G}_w}.
+\end{equation}
+In this case, the estimated $\hat{\delta}$ increases as the underlying strength of the wild type $\RR_w$ increases (\fref{relspeed}A).
+Assuming a narrower distribution causes $\hat{\delta}$ to be less sensitive to $\RR_w$;
+when $\kappa = 0$ (delta-distributed generation intervals), $\hat{\delta}$ remains constant.
+
+When we allow the mean generation interval of the new variant to be long ($\bar{G}_v > \bar{G}_w$), the true $\theta$--$\delta$ relationship is given by:
+\begin{equation}
+\delta = \frac{(\theta \RR_w)^{\kappa} - 1}{\kappa \bar{G}_v} - \frac{\RR_w^{\kappa} - 1}{\kappa \bar{G}_w}.
+\end{equation}
+In this case, the direction of the effect of $\RR_w$ on $\delta$ can change depending on the value of $\kappa$.
+As $\kappa \to 1$, we have
+\begin{equation}
+\delta \to \frac{(\bar{G}_w \theta - \bar{G}_v) \RR_w + \bar{G}_v - \bar{G}_w}{\bar{G}_v \bar{G}_w},
+\end{equation}
+which changes linearly with $\RR_w$ with a slope of $(\bar{G}_w \theta - \bar{G}_v)/\bar{G}_v \bar{G}_w$ (which can be either negative or positive).
+For the values we consider, this increases with $\RR_w$ (\fref{relspeed}B).
+On the other hand, as $\kappa \to 0$, we have
+\begin{equation}
+\delta \to \frac{\log(\theta \RR_w)}{\bar{G}_v} - \frac{\log(\RR_w)}{\bar{G}_w},
+\end{equation}
+which always decreases with $\RR_w$ provided that $\bar{G}_v > \bar{G}_w$ (\fref{relspeed}B).
+Overall, estimates of $\delta$ become less sensitive to values of $\RR_w$ and range between 0.05/day and 0.15/day, which is consistent with the observed values of $\delta$ from data.
+Therefore, assuming equal generation intervals generally leads to over-estimation of $\delta$ (\fref{relspeed}C).
 
@@
 
-Here, we explored how assumptions the relative strength and speed of the new variant affects conclusions about each other when the new variant has longer generation intervals than the existing strain.
-We define relative strength as the ratio between reproduction numbers ($\theta=\RR_v/\RR_w$) and relative speed as the difference between growth rates ($\delta=r_v-r_w$).
+Here, we explored how the generation-interval distribution shapes the link between relative strength and speed of the new variant.
+We define relative strength as the ratio between reproduction numbers ($\theta=\RR_v/\RR_w$) and relative speed as the difference between growth rates ($\delta=r_v-r_w$) of the variant and the wild type.
+When the generation-interval distributions are asumed to be identical, we find that $\hat{\theta}$ decreases with $$
+
 When relative speed $\delta$ is held constant, faster growth of the existing strain leads to larger relative strength.

### 2021-02-28 newvariant.tex
done for now?

@@
 Characterizing epidemic dynamics of new SARS-CoV-2 variants is critical to predicting the future course of the pandemic.
-Here, we explore how assumptions about the relative strength (i.e., ratio of reproduction numbers $\RR$) and relative speed (i.e., differences in growth rates $r$) of the new variant affect conclusions about each other, given the possibility that the new variant has longer generation intervals.
-When relative speed---which corresponds to the growth rate of the logistic curve that models the proportion of the new variant---is held constant, higher growth rate of the original strain leads to higher relative strength.
-When relative strength is held constant, however, higher reproduction number of the original strain leads to smaller relative speed.
-Therefore, intervention strategies that reduce $\RR$ can cause the variant to replace the original strain faster.
-Neglecting differences in generation intervals lead to opposite conclusions about the relationship between relative strength and speed, biasing their estimates.
+Here, we explore how assumptions about the relative strength (i.e., ratio of reproduction numbers $\RR$) and relative speed (i.e., differences in growth rates $r$) of the new variant affect conclusions about each other.
+We show that failing to account for the possibility that the new variant has longer generation intervals can lead to considerable biases in the inference of the spread of the new variant.
+For example, assuming a mean generation interval of 7 days for the new variant and 5 days for the wild type suggest that 1.1--1.4 fold biases in the degree of increased transmissibility of the new variant is plausible.
 We suggest that estimates of increased transmissibility of new variants should be re-assessed in the light of these findings.
@@
 In many contexts, previous studies have shown that inferences about disease dynamics (e.g., effectiveness of intervention strategy) can changes depending on whether $\RR$ or $r$ is held constant \citep{dushoff2020speed}.
+
 Such inference also depends on assumptions about the underlying generation-interval distribution \citep{roberts2007model,svensson2007note,wallinga2007generation}.
-While many studies assume that the generation-interval distributions do not differ between the existing strain and the variant, recent evidence suggests that the new variant has longer infectious period, leading to longer generation intervals.
+Most studies have assumed that the generation-interval distributions do not differ between the existing strain and the variant, but recent evidence suggests that the new variant may have longer duration of infection: 13.3 days (90\% CI: 10.1--16.5) for the new variant and 8.2 days (90\% CI: 6.5--9.7) for the wild type \citep{kissler2021densely}.
+Longer duration of infection provides opportunity to transmit for a longer period of time, meaning that the generation intervals of the new variant is likely to be longer than that of the wild type.
+However, there is much uncertainty in how much longer generation intervals might be because generation intervals also depend on other factors that affect transmission such as awareness-driven behavior (e.g., faster self-isolation after symptom onset leading to shorter generation intervals).
 
-Here, we explore how different assumptions about the new variant affects estimates of its strength and speed.
+Here, we explore how different assumptions about the new variant affects estimates of its strength and speed given the possibility that it might have longer generation intervals.
 We ask how the estimates of relative strength (i.e., the ratio between the strength of the existing strain and the variant) and speed (i.e., the difference between the speed of the existing strain and the variant) depend on each other.
@@
 Several studies have tried to predict the future spread of the new variant, including when it will replace the wild type (therefore, its relative speed $\delta$).
-As our baseline scenario we assume $\theta = 1.8$ and ask how assumptions about the generation-interval distribution affect predictions of $\delta$.
+As our baseline scenario we assume $\theta = 1.8$ and ask how assumptions about the generation-interval distribution affect predictions of $\delta$---while this is considerably higher than most previous estimates of around 1.5 \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}, we chose this value as it better reflects the possibility that the new variant may have longer generation intervals (\fref{relstrength}).
 
@@
 \textbf{Relative speed of the new variant given assumed relative strength.}
-(A) True relative speed of the new variant $\delta$ given strength of the wild type $\RR_w$ and the relative strength $\theta$.
-The dashed contour line corresponds to $\delta = 0$.
-True relative speed of the new variant $\delta$ is calculated using $\bar{G}_w = 5\,\textrm{days}$, $\bar{G}_v = 8\,\textrm{days}$, and $\kappa = 1/5$. 
-(B) Estimated relative speed of the new variant $\hat{\delta}$ given strength of the wild type $\RR_w$ and the relative strength $\theta$ assuming identical generation-interval distributions.
-(C) Biases in the estimates of the relative speed of the new variant $\hat{\delta} - \delta$ given strength of the wild type $\RR_w$ and the relative strength $\theta$.
-The dashed contour line corresponds to $\hat{\delta} = \delta$.
+(A) Estimated relative speed of the new variant $\hat{\delta}$ given strength of the wild type $\RR_w$ assuming identical generation-interval distributions.
+(B) True relative speed of the new variant $\delta$ given strength of the wild type $\RR_w$ accounting for longer generation intervals of the new variant.
+(C) Difference between the true and the estimated relative speed of the new variant $\delta - \hat{\delta}$.
+True relative speed of the new variant $\delta$ is calculated using $\theta=1.8$, $\bar{G}_w = 5\,\textrm{days}$, and $\bar{G}_v = 7\,\textrm{days}$.
 }
@@
 We define relative strength as the ratio between reproduction numbers ($\theta=\RR_v/\RR_w$) and relative speed as the difference between growth rates ($\delta=r_v-r_w$) of the variant and the wild type.
-When the generation-interval distributions are asumed to be identical, we find that $\hat{\theta}$ decreases with $$
-
-When relative speed $\delta$ is held constant, faster growth of the existing strain leads to larger relative strength.
-When relative strength $\theta$ is held constant, higher strength of the existing strain leads to smaller relative speed.
-Failing to account for the differences in the generation-interval distributions lead to opposite conclusions, leading to biased estimates of $\delta$ and $\theta$.
+When the generation-interval distributions are asumed to be identical, we find that estimates of $\hat{\theta}$ decreases with $r_w$ whereas $\hat{\delta}$ increase with $\RR_w$.
+When we account for longer generation intervals of the new variant, these relationships can change directions depending on the effective dispersion in generation intervals, which we characterized in terms of the squared coefficient of variation of the gamma distribution $\kappa$.
+Therefore, even a modest differences in the mean generation intervals (2 days, in this case) can lead to considerable biases in the inference---we find that 1.1--1.4 fold under-estimation in the relative strength of the new variant is plausible.
 
@@
 Early in the spread, we typically want to fix the relative speed because we can directly observe it.
-As more information about the transmission and immunity profiles of the new variant becomes available, it is better to fix the relative strength as it better mathces biological assumptions.
+As more information about the transmission and immunity profiles of the new variant becomes available, it is better to fix the relative strength as it better matches biological mechanisms of its spread.
 We suggest researchers to be mindful about which quantity they hold constant and how their assumptions lead to their conclusions.
@@
 However, this is only true if both have the same mean generation intervals (\fref{relspeed}B).
-If the new variant has longer generation intervals than the original strain, imposing strong intervention (therefore reducing both $\RR_w$ and $\RR_v$ by a constant amount) results in a higher relative speed $\delta$, and therefore faster replacement of the original strain (\fref{relspeed}A).
-While it is still necessary to impose stronger intervention to control the epidemic if the new variant has higher $\RR$, it may have unexpected consequences.
+If the new variant has longer generation intervals than the original strain, imposing strong intervention (therefore reducing both $\RR_w$ and $\RR_v$ by a constant amount) can result in a higher relative speed $\delta$, and therefore faster replacement of the original strain (\fref{relspeed}B).
+This is not necesarily a bad news as the absolute speed ($r_v$) of the new variant still decreases as $\RR_v$ decreases.
+These effects should be taken into consideration for accurate projection of the spread of the new variant.
 
-Since the beginning of the SARS-CoV-2 pandemic, the importance of understanding the role of generation-interval distributions in its spread has been constantly emphasized.
+Even though SARS-CoV-2 has been spreading for more than a year, there is still large uncertainty in its generation intervals.
+Observational studies typically focus on serial intervals (i.e., time between symptom onset of the infector and the infectee; \cite{svensson2007note}) because they are easier to measure \citep{griffin2020rapid}, but they are subject to dynamical biases that can be difficult to tease apart \citep{park2021forward}.
+A few studies have tried to estimate the generation-interval distribution from serial intervals with means ranging between 3--6 days and squared coefficient of variations ranging between 0.1--0.6 \citep{ferretti2020quantifying,Ferretti2020timing,ganyani2020estimating,knight2020estimating};
+however, most estimates neglect dynamical biases in serial intervals.
+In addition, serial intervals do not account for asymptomatic transmission, adding further uncertainty to inferences of speed and strength \citep{park2020time}.
+
+Since the beginning of the SARS-CoV-2 pandemic, the importance of understanding the role of generation-interval distributions in its spread has been constantly emphasized \citep{doi:10.1098/rsif.2020.0144}.
 Yet, we still feel that generation intervals are under-appreciated, often viewed as a leeway to estimating $\RR$.

### 2021-02-28 newvariant.tex
spelling

@@
 I think at this level we're OK with saying we approximate by a gamma and kappa is the squared CV.}
+\swp{I tried to talk about it here and there. See what you think.}
 
@@
 Studies of new SARS-CoV-2 variants, however, have mostly focused on characterizing changes in \emph{proportion} of a new variant \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
-This has the advantage of correcting for testing patterns and intensity, and possibly for other transient effects that might affect variants and wildtype viruses similarly.
+This has the advantage of correcting for testing patterns and intensity, and possibly for other transient effects that might affect variants and wild type viruses similarly.
 When incidence is changing exponentially, the proportion of the new variant $p(t)$ follows a logistic equation:
@@
 We define relative strength as the ratio between reproduction numbers ($\theta=\RR_v/\RR_w$) and relative speed as the difference between growth rates ($\delta=r_v-r_w$) of the variant and the wild type.
-When the generation-interval distributions are asumed to be identical, we find that estimates of $\hat{\theta}$ decreases with $r_w$ whereas $\hat{\delta}$ increase with $\RR_w$.
+When the generation-interval distributions are assumed to be identical, we find that estimates of $\hat{\theta}$ decreases with $r_w$ whereas $\hat{\delta}$ increase with $\RR_w$.
 When we account for longer generation intervals of the new variant, these relationships can change directions depending on the effective dispersion in generation intervals, which we characterized in terms of the squared coefficient of variation of the gamma distribution $\kappa$.
@@
 If the new variant has longer generation intervals than the original strain, imposing strong intervention (therefore reducing both $\RR_w$ and $\RR_v$ by a constant amount) can result in a higher relative speed $\delta$, and therefore faster replacement of the original strain (\fref{relspeed}B).
-This is not necesarily a bad news as the absolute speed ($r_v$) of the new variant still decreases as $\RR_v$ decreases.
+This is not necessarily a bad news as the absolute speed ($r_v$) of the new variant still decreases as $\RR_v$ decreases.
 These effects should be taken into consideration for accurate projection of the spread of the new variant.

### 2021-03-01 newvariant.tex
Some quick changes

@@
 Characterizing epidemic dynamics of new SARS-CoV-2 variants is critical to predicting the future course of the pandemic.
-Here, we explore how assumptions about the relative strength (i.e., ratio of reproduction numbers $\RR$) and relative speed (i.e., differences in growth rates $r$) of the new variant affect conclusions about each other.
-We show that failing to account for the possibility that the new variant has longer generation intervals can lead to considerable biases in the inference of the spread of the new variant.
-For example, assuming a mean generation interval of 7 days for the new variant and 5 days for the wild type suggest that 1.1--1.4 fold biases in the degree of increased transmissibility of the new variant is plausible.
-We suggest that estimates of increased transmissibility of new variants should be re-assessed in the light of these findings.
+Here, we explore how assumptions about the relative strength (i.e., ratio of reproduction numbers $\RR$) and relative speed (i.e., differences in growth rates $r$) of the new variant affect conclusions about expected rates of variant takeover and viral spread under various conditions.
+We show that failing to account for the possibility that the new variant has longer generation intervals can lead to considerable biases in the inferences about the relative reproductive number (strength) of the new variant.
+For example, the plausible assumption that the mean generation interval of the new variant is 40\% longer than that of the wild type can lead to biases of up to 40\% in estimates of  the relative transmissibility of the new variant.
+We suggest that better estimates of viral generation intervals could improve forecasts of variant takeover and viral spread.
 
@@
 While both approaches are reasonable, they can lead to different conclusions.
-In many contexts, previous studies have shown that inferences about disease dynamics (e.g., effectiveness of intervention strategy) can changes depending on whether $\RR$ or $r$ is held constant \citep{dushoff2020speed}.
+In many contexts, previous studies have shown that inferences about disease dynamics (e.g., effectiveness of intervention strategy) can changes depending on whether it is assumed that $\RR$ or $r$ is known \citep{dushoff2020speed}.
 
@@
 Most studies have assumed that the generation-interval distributions do not differ between the existing strain and the variant, but recent evidence suggests that the new variant may have longer duration of infection: 13.3 days (90\% CI: 10.1--16.5) for the new variant and 8.2 days (90\% CI: 6.5--9.7) for the wild type \citep{kissler2021densely}.
-Longer duration of infection provides opportunity to transmit for a longer period of time, meaning that the generation intervals of the new variant is likely to be longer than that of the wild type.
-However, there is much uncertainty in how much longer generation intervals might be because generation intervals also depend on other factors that affect transmission such as awareness-driven behavior (e.g., faster self-isolation after symptom onset leading to shorter generation intervals).
+Longer duration of infection is likely to provide opportunity to transmit for a longer period of time, meaning that the mean generation interval of the new variant is likely to be longer than that of the wild type.
+More information is needed, however: generation intervals depend on many other biological and behavioral factors, including for example self-isolation after symptom onset leading .
 
-Here, we explore how different assumptions about the new variant affects estimates of its strength and speed given the possibility that it might have longer generation intervals.
+Here, we explore how different assumptions about the new variant affect estimates of its strength and speed given the possibility that it might have longer generation intervals.
 We ask how the estimates of relative strength (i.e., the ratio between the strength of the existing strain and the variant) and speed (i.e., the difference between the speed of the existing strain and the variant) depend on each other.
@@
 We use the renewal equation framework to characterize the spread of two pathogen strains---in this case, the wild type SARS-CoV-2 virus and the variant of concern.
-Assuming that the effects mutations are negligible, the current incidence of infection $i_x(t)$ caused by the wild type ($x=w$) and the variant ($x=v$) can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary cases are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
+Neglecting the (relatively slow) rate of new mutations, the current incidence of infection $i_x(t)$ caused by the wild type ($x=w$) and the variant ($x=v$) can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary cases are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
 \begin{equation}
@@
 For simplicity we neglect differences in the squared coefficient of variation and assume $\kappa_w = \kappa_v = \kappa$; instead, we focus on the effect of potential differences in the mean generation intervals.
-
-\jd{I like to talk about `effective dispersion' when we're getting into the details of distributions and how good the gamma assumption is.
-I think at this level we're OK with saying we approximate by a gamma and kappa is the squared CV.}
-\swp{I tried to talk about it here and there. See what you think.}
 

### 2021-03-01 newvariant.tex

@@
 Longer duration of infection is likely to provide opportunity to transmit for a longer period of time, meaning that the mean generation interval of the new variant is likely to be longer than that of the wild type.
-More information is needed, however: generation intervals depend on many other biological and behavioral factors, including for example self-isolation after symptom onset leading .
+More information is needed, however: generation intervals depend on many other biological and behavioral factors, including for example self-isolation after symptom onset leading to shorter generation intervals.
 

### 2021-03-02 Rtbias.R
A really cool figure that might be used

@@ -1,0 +1,113 @@
+library(dplyr)
+library(ggplot2); theme_set(theme_bw())
+library(egg)
+source("renewal_det.R")
+
+rr1 <- renewal_det(tmax=150)
+rr2 <- renewal_det(Rfun=function(t) ifelse(t < 50, 2.5, ifelse(t < 80, 0.5, 1.2)) ,
+                   tmax=150)
+
+g1 <- ggplot(rr1) +
+  geom_line(aes(tvec, Ivec1, color="Wild type")) +
+  geom_line(aes(tvec, Ivec2, color="Variant")) +
+  scale_x_continuous("Time (days)", expand=c(0, 0), limits=c(15, 145)) +
+  scale_y_continuous("Instantaneous incidence (1/days)") +
+  scale_color_manual(values=c("red", "black")) +
+  ggtitle("A") +
+  theme(
+    panel.grid = element_blank(),
+    legend.position = c(0.7, 0.8),
+    legend.title = element_blank()
+  )
+
+g2 <- ggplot(rr1) +
+  geom_line(aes(tvec, Ivec2/(Ivec1 + Ivec2))) +
+  scale_x_continuous("Time (days)", expand=c(0, 0), limits=c(15, 145)) +
+  scale_y_continuous("Proportion of the new variant") +
+  scale_color_manual(values=c("red", "black")) +
+  ggtitle("B") +
+  theme(
+    panel.grid = element_blank(),
+    legend.position = "none"
+  )
+
+g3 <- ggplot(rr1) +
+  geom_line(aes(tvec, Rt1, color="Wild type")) +
+  geom_line(aes(tvec, Rt2, color="Variant")) +
+  geom_line(aes(tvec, Rtest2, color="Variant"), lty=2) +
+  scale_x_continuous("Time (days)", expand=c(0, 0), limits=c(15, 145)) +
+  scale_y_continuous("Reproduction number") +
+  scale_color_manual(values=c("red", "black")) +
+  ggtitle("C") +
+  theme(
+    panel.grid = element_blank(),
+    legend.position = "none"
+  )
+
+g4 <- ggplot(rr1) +
+  geom_line(aes(tvec, Rt2/Rt1, color="True ratio")) +
+  geom_line(aes(tvec, Rtest2/Rt1, color="Estimaed ratio")) +
+  scale_x_continuous("Time (days)", expand=c(0, 0), limits=c(15, 145)) +
+  scale_y_continuous("Reproduction number ratio") +
[…]

### 2021-03-02 renewal_det.R

@@ -1,0 +1,46 @@
+renewal_det <- function(Rfun=function(t) 2.5,
+                        theta=1.8,
+                        N=40000,
+                        dt=0.025,
+                        genfun1=function(x) dgamma(x, 5, 1),
+                        genfun2=function(x) dgamma(x, 5, 5/7),
+                        I01=0.1,
+                        I02=0.001,
+                        tmax=100,
+                        genmax=2000) {
+  gen1 <- genfun1(0:genmax*dt+dt)
+  gen1 <- gen1/sum(gen1)
+  gen2 <- genfun2(0:genmax*dt+dt)
+  gen2 <- gen2/sum(gen2)
+  tvec <- seq(0, tmax, by=dt)
+  Ivec1 <- rep(0, tmax/dt)
+  Ivec2 <- rep(0, tmax/dt)
+  
+  Ivec1[1] <- I01
+  Ivec2[1] <- I02
+  Svec <- rep(0, tmax/dt)
+  Svec[1] <- N - I01 - I02
+  
+  for (i in 2:length(tvec)) {
+    Ivec1[i] <- Svec[i-1] * Rfun(tvec[i]) * sum(Ivec1[max(1, i-genmax):(i-1)] * gen1[min(i, genmax+1):2])/N
+    Ivec2[i] <- Svec[i-1] * theta * Rfun(tvec[i]) * sum(Ivec2[max(1, i-genmax):(i-1)] * gen2[min(i, genmax+1):2])/N
+    Svec[i] <- Svec[i-1] - Ivec1[i] - Ivec2[i]
+  }
+  
+  Rt1 <- tail(sapply(tvec, Rfun), -1) * head(Svec, -1)/N
+  Rt2 <- tail(sapply(tvec, Rfun), -1) * head(Svec, -1)/N * theta
+  
+  Rtest1 <- tail(Ivec1, -1)/sapply(2:(length(Ivec1)), function(x) sum(Ivec1[max(1, x-genmax):(x-1)] * gen1[min(x, genmax+1):2]))
+  Rtest2 <- tail(Ivec2, -1)/sapply(2:(length(Ivec2)), function(x) sum(Ivec2[max(1, x-genmax):(x-1)] * gen1[min(x, genmax+1):2]))
+  
+  data.frame(
+    tvec=tvec[-1],
+    Ivec1=Ivec1[-1]/dt,
[…]

### 2021-03-10 Rtbias.R
new figures

@@ -1,113 +1,128 @@
 library(dplyr)
 library(ggplot2); theme_set(theme_bw())
 library(egg)
+library(ggpubr)
+library(gridExtra)
 source("renewal_det.R")
 
-rr1 <- renewal_det(tmax=150)
-rr2 <- renewal_det(Rfun=function(t) ifelse(t < 50, 2.5, ifelse(t < 80, 0.5, 1.2)) ,
-                   tmax=150)
+## fix theta = 1.5
+## fix kappa = 1/5
+## fix assumed GI
+## assume wild type = var but not necessarily the one we're assuming
 
-g1 <- ggplot(rr1) +
-  geom_line(aes(tvec, Ivec1, color="Wild type")) +
-  geom_line(aes(tvec, Ivec2, color="Variant")) +
-  scale_x_continuous("Time (days)", expand=c(0, 0), limits=c(15, 145)) +
-  scale_y_continuous("Instantaneous incidence (1/days)") +
-  scale_color_manual(values=c("red", "black")) +
-  ggtitle("A") +
-  theme(
-    panel.grid = element_blank(),
-    legend.position = c(0.7, 0.8),
-    legend.title = element_blank()
-  )
+theta <- 1.5
+genfun_assumed <- function(x) dgamma(x, 5, 5/5)
+genfun_short <- function(x) dgamma(x, 5, 5/4)
+genfun_long <- function(x) dgamma(x, 5, 5/6)
 
-g2 <- ggplot(rr1) +
-  geom_line(aes(tvec, Ivec2/(Ivec1 + Ivec2))) +
-  scale_x_continuous("Time (days)", expand=c(0, 0), limits=c(15, 145)) +
-  scale_y_continuous("Proportion of the new variant") +
-  scale_color_manual(values=c("red", "black")) +
-  ggtitle("B") +
-  theme(
-    panel.grid = element_blank(),
-    legend.position = "none"
-  )
+svec <- c("Wild type GI = Variant GI = Assumed GI",
+          "Wild type GI = Variant GI < Assumed GI",
+          "Assumed GI < Wild type GI = Variant GI",
+          "Variant GI < Assumed GI < Wild type GI",
+          "Wild type GI < Assumed GI < Variant GI",
+          "Variant GI < Wild type GI = Assumed GI",
+          "Wild type GI = Assumed GI < Variant GI")
 
-g3 <- ggplot(rr1) +
-  geom_line(aes(tvec, Rt1, color="Wild type")) +
[…]

### 2021-03-10 relspeed.R

@@ -4,45 +4,62 @@ library(ggthemes)
 library(egg)
 library(tikzDevice)
 
-theta <- 1.8
+theta <- 1.5
 
-Rwvec <- seq(0.3, 1.7, length.out=21)
+Rwvec <- seq(0.3, 1.7, length.out=31)
 kappavec <- c(0, 0.2, 0.5, 1)
+kappabase <- 0.2
 
 pardata <- expand.grid(kappavec, Rwvec)
 
 Gw <- 5
-Gv <- 7
+
+Gratio <- seq(0.6, 1.4, length.out=31)
+
+pardata2 <- expand.grid(Rwvec, Gratio)
 
 deltadata <- apply(pardata, 1, function(x) {
   kappa <- x[[1]]
   Rw <- x[[2]]
   
   if (kappa == 0) {
-    delta <- log(theta * Rw)/Gv - log(Rw)/Gw
-    
     deltahat <- log(theta * Rw)/Gw - log(Rw)/Gw
   } else {
-    delta <- ((theta * Rw)^kappa - 1)/(kappa*Gv) -  ((Rw)^kappa - 1)/(kappa*Gw)
-    
     deltahat <- ((theta * Rw)^kappa - 1)/(kappa*Gw) -  ((Rw)^kappa - 1)/(kappa*Gw)
-    
   }
   
   data.frame(
     theta=theta,
     Rw=Rw,
+    deltahat=deltahat,
+    kappa=kappa
+  )
+}) %>%
+  bind_rows
+
+deltadata2 <- apply(pardata2, 1, function(x) {
+  Rw <- x[[1]]
+  Gratio <- x[[2]] ## Gv/Gw
+  Gv <- Gratio * Gw
+  
+  delta <- ((theta * Rw)^kappabase - 1)/(kappabase*Gv) -  ((Rw)^kappabase - 1)/(kappabase*Gw)
+  
+  deltahat <- ((theta * Rw)^kappabase - 1)/(kappabase*Gw) -  ((Rw)^kappabase - 1)/(kappabase*Gw)
+  
+  data.frame(
+    theta=theta,
+    Gratio=Gratio,
+    Rw=Rw,
     delta=delta,
     deltahat=deltahat,
-    bias=delta-deltahat,
-    kappa=kappa
+    bias=delta-deltahat
   )
 }) %>%
   bind_rows
 
 g1 <- ggplot(deltadata) +
[…]

### 2021-03-10 relstrength.R

@@ -6,27 +6,25 @@ library(tikzDevice)
 
 delta <- 0.1
 
-rwvec <- seq(-0.13, 0.13, length.out=21)
+rwvec <- seq(-0.13, 0.13, length.out=31)
 kappavec <- c(0, 0.2, 0.5, 1)
+kappabase <- 0.2
 
 pardata <- expand.grid(kappavec, rwvec)
 
 Gw <- 5
-Gv <- 7
+
+Gratio <- seq(0.6, 1.4, length.out=31)
+
+pardata2 <- expand.grid(rwvec, Gratio)
 
 thetadata <- apply(pardata, 1, function(x) {
   kappa <- x[[1]]
   rw <- x[[2]]
   
   if (kappa==0) {
-    
-    theta <- exp(Gv * (rw + delta))/exp(Gw * rw)
-    
     thetahat <- exp(Gw * (rw + delta))/exp(Gw * rw)
   } else {
-    theta <- ((1 + kappa * (rw + delta) * Gv)^(1/kappa))/
-      ((1 + kappa * rw * Gw)^(1/kappa))
-    
     thetahat <- ((1 + kappa * (rw + delta) * Gw)^(1/kappa))/
       ((1 + kappa * rw * Gw)^(1/kappa))
   }
@@ -34,10 +32,30 @@ thetadata <- apply(pardata, 1, function(x) {
   data.frame(
     delta=delta,
     rw=rw,
+    thetahat=thetahat,
+    kappa=kappa
+  )
+}) %>%
+  bind_rows
+
+thetadata2 <- apply(pardata2, 1, function(x) {
+  rw <- x[[1]]
+  Gratio <- x[[2]] ## Gv/Gw
+  Gv <- Gratio * Gw
+  
+  theta <- ((1 + kappabase * (rw + delta) * Gv)^(1/kappabase))/
+    ((1 + kappabase * rw * Gw)^(1/kappabase))
+  
+  thetahat <- ((1 + kappabase * (rw + delta) * Gw)^(1/kappabase))/
+    ((1 + kappabase * rw * Gw)^(1/kappabase))
+  
+  data.frame(
+    delta=delta,
+    Gratio=Gratio,
+    rw=rw,
     theta=theta,
     thetahat=thetahat,
-    bias=theta/thetahat,
-    kappa=kappa
+    bias=theta/thetahat
   )
 }) %>%
   bind_rows
[…]

### 2021-03-10 renewal_det.R

@@ -1,9 +1,10 @@
 renewal_det <- function(Rfun=function(t) 2.5,
-                        theta=1.8,
+                        theta=1.5,
                         N=40000,
                         dt=0.025,
                         genfun1=function(x) dgamma(x, 5, 1),
-                        genfun2=function(x) dgamma(x, 5, 5/7),
+                        genfun2=function(x) dgamma(x, 5, 1),
+                        genfun3=function(x) dgamma(x, 5, 1),
                         I01=0.1,
                         I02=0.001,
                         tmax=100,
@@ -12,6 +13,8 @@ renewal_det <- function(Rfun=function(t) 2.5,
   gen1 <- gen1/sum(gen1)
   gen2 <- genfun2(0:genmax*dt+dt)
   gen2 <- gen2/sum(gen2)
+  gen3 <- genfun3(0:genmax*dt+dt)
+  gen3 <- gen3/sum(gen3)
   tvec <- seq(0, tmax, by=dt)
   Ivec1 <- rep(0, tmax/dt)
   Ivec2 <- rep(0, tmax/dt)
@@ -30,8 +33,8 @@ renewal_det <- function(Rfun=function(t) 2.5,
   Rt1 <- tail(sapply(tvec, Rfun), -1) * head(Svec, -1)/N
   Rt2 <- tail(sapply(tvec, Rfun), -1) * head(Svec, -1)/N * theta
   
-  Rtest1 <- tail(Ivec1, -1)/sapply(2:(length(Ivec1)), function(x) sum(Ivec1[max(1, x-genmax):(x-1)] * gen1[min(x, genmax+1):2]))
-  Rtest2 <- tail(Ivec2, -1)/sapply(2:(length(Ivec2)), function(x) sum(Ivec2[max(1, x-genmax):(x-1)] * gen1[min(x, genmax+1):2]))
+  Rtest1 <- tail(Ivec1, -1)/sapply(2:(length(Ivec1)), function(x) sum(Ivec1[max(1, x-genmax):(x-1)] * gen3[min(x, genmax+1):2]))
+  Rtest2 <- tail(Ivec2, -1)/sapply(2:(length(Ivec2)), function(x) sum(Ivec2[max(1, x-genmax):(x-1)] * gen3[min(x, genmax+1):2]))
   
   data.frame(
     tvec=tvec[-1],

### 2021-03-13 newvariant.tex
updating

@@
 Epidemic strength, characterized by the reproduction number $\RR$, provides information about the final size of the epidemic as well as the amount of intervention required to eliminate the disease \citep{anderson1991infectious}.
-Epidemic speed, characterized by the growth rate $r$ provides information about the population-level time scale of the epidemic and, therefore, when the new variant will replace the existing variant.
-These two quantities are linked by generation intervals---defined as time between infection and transmission---which define individual-level time scale of the epidemic \citep{roberts2007model,svensson2007note,wallinga2007generation}.
+Epidemic speed, characterized by the growth rate $r$ provides information about the population-level time scale of the epidemic and, potentially provides information on when the new variant will replace the existing variant as the dominant strain.
+Strength and speed are linked by generation intervals---defined as time between infection and transmission---which characterizes individual-level time scale of the epidemic \citep{roberts2007model,svensson2007note,wallinga2007generation}.
 
 Many studies have tried to estimate the strength of the novel variant from the observed speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021}.
-On the other hand, some studies have assumed a value of the strength of the novel variant and tried to predict future spread (therefore, the speed) of the variant \citep{davies2021estimated}.
-While both approaches are reasonable, they can lead to different conclusions.
-In many contexts, previous studies have shown that inferences about disease dynamics (e.g., effectiveness of intervention strategy) can changes depending on whether it is assumed that $\RR$ or $r$ is known \citep{dushoff2020speed}.
+In contrast, some studies have assumed a value of the strength of the novel variant and tried to predict future spread (therefore, the speed) of the variant \citep{davies2021estimated}.
+While both approaches are reasonable in principle, holding different quantities constant (i.e., strength or speed) can lead to different conclusions, in practice, about the spread of the disease, and therefore, its control.
+The reason is that inference of speed from strength and strength from speed depends on assumptions about the generation-interval distribution \citep{roberts2007model,svensson2007note,wallinga2007generation,dushoff2020speed}.
+For example, when we assume a fixed value of epidemic strength, longer generation intervals lead to faster epidemic (lower $r$), making the epidemic seem easier to control.
+When we assume a fixed value of epidemic speed, longer generation intervals lead to stronger epidemic (higher $\RR$), making the epidemic seem more difficult to control.
 
-Such inference also depends on assumptions about the underlying generation-interval distribution \citep{roberts2007model,svensson2007note,wallinga2007generation}.
-Most studies have assumed that the generation-interval distributions do not differ between the existing strain and the variant, but recent evidence suggests that the new variant may have longer duration of infection: 13.3 days (90\% CI: 10.1--16.5) for the new variant and 8.2 days (90\% CI: 6.5--9.7) for the wild type \citep{kissler2021densely}.
-Longer duration of infection is likely to provide opportunity to transmit for a longer period of time, meaning that the mean generation interval of the new variant is likely to be longer than that of the wild type.
+In addition, most studies have assumed that the generation-interval distributions do not differ between the existing strain and the variant, but recent evidence suggests that the new variant may have a longer duration of infection: 13.3 days (90\% CI: 10.1--16.5) for the new variant and 8.2 days (90\% CI: 6.5--9.7) for the wild type \citep{kissler2021densely}.
+Longer duration of infection could lead to longer periods of transmission, meaning that the mean generation interval of the new variant is likely to be longer than that of the wild type.
+Other studies have also considered the possibility that the new variant may have shorter generation intervals to account for faster growth rate \citep{davies2021estimated}.
 More information is needed, however: generation intervals depend on many other biological and behavioral factors, including for example self-isolation after symptom onset leading to shorter generation intervals.
+Nonetheless, if the generation-interval distribution of the new variant is different from the generation-interval distribution of the wild type, current estimates of the strength of the new variants may be biased.
 
-Here, we explore how different assumptions about the new variant affect estimates of its strength and speed given the possibility that it might have longer generation intervals.
-We ask how the estimates of relative strength (i.e., the ratio between the strength of the existing strain and the variant) and speed (i.e., the difference between the speed of the existing strain and the variant) depend on each other.
-We then show how neglecting differences in the generation-interval distributions can lead to biased estimates.
+Here, we explore how different assumptions about the new variant affect estimates of its strength and speed given the possibility that it might have different generation intervals.
+We examine how the estimates of relative strength (i.e., the ratio between the strength of the existing strain and the variant) and speed (i.e., the difference between the speed of the existing strain and the variant) depend on each other.
+We then show how neglecting differences in the generation-interval distributions can lead to biased estimates, which, in turn, infleunce mitigation priorities.
 
@@
 
-We use the renewal equation framework to characterize the spread of two pathogen strains---in this case, the wild type SARS-CoV-2 virus and the variant of concern.
+We use the renewal equation framework to characterize the spread of two pathogen strains---in this case, the wild type SARS-CoV-2 virus and a focal variant of concern (e.g., B.1.1.7).
 Neglecting the (relatively slow) rate of new mutations, the current incidence of infection $i_x(t)$ caused by the wild type ($x=w$) and the variant ($x=v$) can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary cases are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
@@
 The normalized kernel $g_x(t, \tau) = K_x(t, \tau)/\RR_x(t)$---which we refer to as the instantaneous generation-interval distribution---describes their relative contribution to current incidence $i_x(t)$ and provides information about the time scale of disease transmission.
-Both the reproduction number and the generation-interval distribution can depend on several factors, including intrinsic infectiousness of an infected individual, non-pharmaceutical interventions, awareness-driven behavior, and population-level susceptibility.
-Many of these factors can affect between both the wild type and the variant in similar ways.
+Both the reproduction number and the generation-interval distribution can depend on several factors, including intrinsic infectiousness of an infected individual, non-pharmaceutical interventions, awareness-driven behavior, and population-level susceptibility \citep{fraser2007estimating}.
 
 Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$.
-In this case, the incidence of both strains exhibit exponential growth or decay at rate $r_x$, satisfying the Euler-Lotka equation \citep{wallinga2007generation}:
+In this case, the incidence of either strains can exhibit exponential growth or decay at rate $r_x$, satisfying the Euler-Lotka equation \citep{wallinga2007generation}:
 \begin{equation}
@@
 This has the advantage of correcting for testing patterns and intensity, and possibly for other transient effects that might affect variants and wild type viruses similarly.
-When incidence is changing exponentially, the proportion of the new variant $p(t)$ follows a logistic equation:
+When incidence is changing exponentially ($i_x(t) = i_x(t_0) \exp(r_x t)$), the proportion of the new variant $p(t)$ follows a logistic equation:
 \begin{align}
-p(t) &= \frac{i_v(0) \exp(r_v t)}{i_w(0) \exp(r_w t) + i_v(0) \exp(r_v t)},
-\\ &= \frac{1}{1 + \left(i_w(0)/i_v(0)\right) \exp(-\delta t)},
+p(t) &= \frac{i_v(t_0) \exp(r_v t)}{i_w(t_0) \exp(r_w t) + i_v(t_0) \exp(r_v t)},
+\\ &= \frac{1}{1 + \left(i_w(t_0)/i_v(t_0)\right) \exp(-\delta t)},
 \end{align}
@@
 As our baseline scenario, we assume $\delta = 0.1/\textrm{days}$ based on the observed relative speed of the new variant in the UK \citep{davies2021estimated}.
-We then ask how the inference of relative strength $\theta = \RR_v/\RR_w$ of the new variant from its relative speed $\delta$ depends on assumptions about the generation-interval distribution.
+This speed can be then used to estimate the relative strength $\theta = \RR_v/\RR_w$ of the new variant with a caveat; such inferece depends on assumptions about the generation-interval distribution.
 The mean generation interval $\bar{G}_w$ of the wild type is assumed to equal 5 days \citep{ferretti2020quantifying}.
-Previous estimates of squared coefficient of variation $\kappa$ in generation intervals are roughly to equal 0.2 \citep{ferretti2020quantifying}.
-However, we expect there to be greater uncertainty in $\kappa$ than in $\bar{G}$, and so we evaluate the estimates of relative strength $\theta$ across a wide range of $\kappa$, ranging from 0 (delta distribution) to 1 (exponential distribution).
+The squared coefficient of variation $\kappa$ in generation intervals os assumed to equal 0.2 \citep{ferretti2020quantifying}---for simplicity, we assume that both the variant and the wild type have equal $\kappa$ and only consider differences in the mean.
+As we expect there to be greater uncertainty in $\kappa$ than in $\bar{G}$, we evaluate the estimates of relative strength $\theta$ across a wide range of $\kappa$, ranging from 0 (delta distribution) to 1 (exponential distribution).
 
 Previous studies have assumed that the generation-interval distributions do not differ between the wild type and the variant: $\bar{G}_v = \bar{G}_w = \bar{G}$.
-In this case, the estimate of the relative strength is given by:
+In this case, the estimate of the relative strength is given by \citep{park2019practical}:
 \begin{equation}
@@
 \end{equation}
-Therefore, if the wild type is spreading faster than we previously thought (higher $r_w$), we expect our estimate of the relative strength $\theta$ to go down (\fref{relstrength}A).
+Therefore, estimates of relative strength $\theta$ depends not only on the relative speed $\delta$ but also on how fast the wild type is spreading in the population $r_w$---a few analyses have implicitly or explicitly neglected this factor by either assuming $r_w = 0$ \citep{switzerland2021variant} or $\kappa = 0$ \citep{davies2021estimated}, in which case we get $\hat{\theta} = \exp(\delta \bar{G})$.
+Instead, if the wild type was spreading faster than previously assumed (higher $r_w$), the estimates of the relative strength $\theta$ will be go down (\fref{relstrength}A).
 The relative strength also depends on the \emph{effective dispersion} in generation intervals (i.e., the amount of variability in generation intervals, as measured by $\kappa$)---using narrower distributions makes $\hat{\theta}$ less sensitive to $r_w$.
 
-However, if the mean generation interval of the new variant is longer than that of the wild type ($\bar{G}_v > \bar{G}_w$), the true relative strength $\theta = \RR_v/\RR_w$ is given by:
+However, if the true mean generation interval of the new variant is different from that of the wild type ($\bar{G}_v \neq \bar{G}_w$), the true relative strength $\theta = \RR_v/\RR_w$ is given by:
 \begin{equation}
@@
 \end{equation}
+In this case,
+
 In this case, the effect of the underlying speed $r_w$ on the relative strength $\theta$ can change directions---\fref{relstrength}B illustrates this effect assuming $\bar{G}_v=7\,\textrm{days}$.

### 2021-03-13 newvariant.tex
editing

@@
 \end{equation}
-In this case,
+In this case, longer (or shorter) mean generation interval of the new variant generally translates to higher (or lower) values of $\theta$---\fref{relstrength}B illustrates this effect assuming $\kappa = 1/5$.
+However, this direction can change when both the new variant and the wild type are decaying exponentially (see area below $r_v = 0$ line in \fref{relstrength}B)---when disease incidence is decaying exponentially, longer generation intervals lead to a \emph{lower} value of strength (and therefore lower relative strength $\theta$).
 
-In this case, the effect of the underlying speed $r_w$ on the relative strength $\theta$ can change directions---\fref{relstrength}B illustrates this effect assuming $\bar{G}_v=7\,\textrm{days}$.
-As $\kappa \to 0$, this approaches
-\begin{equation}
-\theta \to \exp(r_w (\bar{G}_v-\bar{G}_w)+\delta G_v), 
-\end{equation}
-meaning that $\theta$ increases with $r_w$ (\fref{relstrength}B).
-On the other hand, as $\kappa \to 1$, we have
-\begin{equation}
-\theta \to \frac{1 + (r_w + \delta) \bar{G}_v}{1 + r_w \bar{G}_w} = \frac{\bar{G}_v}{\bar{G}_w} + \frac{1+ \delta \bar{G}_v  - \bar{G}_v/\bar{G}_w}{1 + r_w \bar{G}_w},
-\end{equation}
-which decreases with $r_w$ (\fref{relstrength}B).
-
-Therefore, neglecting potential differences in generation intervals generally leads to under-estimation of the relative strength (\fref{relstrength}C).
-In some cases, this can lead to over-estimation (see $r_w < -0.1$ in \fref{relstrength}C):
-Longer generation intervals lead to slower exponential decay when $\RR < 1$, meaning that the new variant can still increase in proportion ($\delta > 0$) even when $\RR_v < \RR_w$.
-However, this is unlikely to be the case as the new variant spread quickly under mixed conditions.
-Instead, it seems more plausible that current estimates have under-estimated the true relative strength by 1.1---1.4 fold.
+Therefore, neglecting potential differences in generation intervals can lead to both under- and over-estimation of the relative strength.
+In \fref{relstrength}C, we quantify the degrees of bias using the ratio between the true value $\theta$, which accounts for differences in generation-interval distributions, and the estimated value $\hat{\theta}$, which assumes equal generation-interval distributions.
+Across the ranges of parameters that we consider, we find that up to 1.4 fold (either upward or downward) adjustments to current estimates may be required.
 
@@
 While relative speed is easier to measure, we often do not expect $\delta$ to be constant---this can depend on epidemic and demographic factors and can vary across time and space.
+Instead, biological mechanisms that cause new variant to have higher strength better translate to assuming a constant value of relative strength.
+For example, if the new variant has higher transmission rate,
+
 Instead, assuming a constant relative strength $\theta$ better matches biological hypotheses about the new variant (e.g., increased transmission and immune escape).

### 2021-03-13 relstrength.R

@@ -77,7 +77,7 @@ g2 <- ggplot(thetadata2) +
   geom_vline(xintercept=1, lty=2, col="white") +
   geom_hline(yintercept=-delta, lty=1, col="white") +
   annotate("text", x=1.04, y=0, label=c("$\\bar{G}_v=\\bar{G}_w$"), col="white", angle=-90) +
-  annotate("text", x=0.8, y=-delta+0.01, label=c("$r_v=1$"), col="white") +
+  annotate("text", x=0.8, y=-delta+0.01, label=c("$r_v=0$"), col="white") +
   scale_x_continuous("Generation interval ratio, $\\bar{G}_v/\\bar{G}_w$", expand=c(0, 0)) +
   scale_y_continuous("Wild type speed, $r_w$ (1/days)", expand=c(0, 0)) +
   scale_fill_viridis_c("$\\theta$") +
@@ -91,7 +91,7 @@ g3 <- ggplot(thetadata2) +
   geom_vline(xintercept=1, lty=2, col="white") +
   geom_hline(yintercept=-delta, lty=1, col="white") +
   annotate("text", x=1.04, y=0, label=c("$\\bar{G}_v=\\bar{G}_w$"), col="white", angle=-90) +
-  annotate("text", x=0.8, y=-delta+0.01, label=c("$r_v=1$"), col="white") +
+  annotate("text", x=0.8, y=-delta+0.01, label=c("$r_v=0$"), col="white") +
   scale_x_continuous("Generation interval ratio, $\\bar{G}_v/\\bar{G}_w$", expand=c(0, 0)) +
   scale_y_continuous("Wild type speed, $r_w$ (1/days)", expand=c(0, 0)) +
   scale_fill_viridis_c("$\\theta/\\hat{\\theta}$", option="A") +

### 2021-03-14 newvariant.tex
done for now?

@@
 
-Characterizing epidemic dynamics of new SARS-CoV-2 variants is critical to predicting the future course of the pandemic.
-Here, we explore how assumptions about the relative strength (i.e., ratio of reproduction numbers $\RR$) and relative speed (i.e., differences in growth rates $r$) of the new variant affect conclusions about expected rates of variant takeover and viral spread under various conditions.
-We show that failing to account for the possibility that the new variant has longer generation intervals can lead to considerable biases in the inferences about the relative reproductive number (strength) of the new variant.
-For example, the plausible assumption that the mean generation interval of the new variant is 40\% longer than that of the wild type can lead to biases of up to 40\% in estimates of  the relative transmissibility of the new variant.
-We suggest that better estimates of viral generation intervals could improve forecasts of variant takeover and viral spread.
-
 \section{Introduction}
@@
 $r_w$ assuming identical generation-interval distributions.
-(B) True relative strength of the new variant $\theta$ given speed of the wild type $r_w$ accounting for longer generation intervals of the new variant.
+(B) True relative strength of the new variant $\theta$ given speed of the wild type $r_w$ and ratio between the mean generation interval of the new variant $\bar{G}_v$ and that of the wild type $\bar{G}_w$.
 (C) Ratio between the true and the estimated relative strength of the new variant $\theta/\hat{\theta}$.
-True relative strength of the new variant $\theta$ is calculated using $\delta=0.1\,\textrm{days}$, $\bar{G}_w = 5\,\textrm{days}$, and $\bar{G}_v = 7\,\textrm{days}$.
+True relative strength of the new variant $\theta$ is calculated using $\delta=0.1\,\textrm{days}$, $\bar{G}_w = 5\,\textrm{days}$, and $\kappa = 1/5$.
 }
@@
 While relative speed is easier to measure, we often do not expect $\delta$ to be constant---this can depend on epidemic and demographic factors and can vary across time and space.
-Instead, biological mechanisms that cause new variant to have higher strength better translate to assuming a constant value of relative strength.
-For example, if the new variant has higher transmission rate,
-
-Instead, assuming a constant relative strength $\theta$ better matches biological hypotheses about the new variant (e.g., increased transmission and immune escape).
-Several studies have tried to predict the future spread of the new variant, including when it will replace the wild type (therefore, its relative speed $\delta$).
-As our baseline scenario we assume $\theta = 1.8$ and ask how assumptions about the generation-interval distribution affect predictions of $\delta$---while this is considerably higher than most previous estimates of around 1.5 \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}, we chose this value as it better reflects the possibility that the new variant may have longer generation intervals (\fref{relstrength}).
+Instead, biological mechanisms that cause new variant to have higher strength better translate to assuming a constant value of relative strength $\theta$.
+For example, if the new variant has a higher transmission rate, the ratio between the strength of the new variant and that of the wild type will remain constant ($=\theta$) under strength-like interventions that reduces the strength of both strains by a constant amount.
+However, this is not necessarily true under speed-like interventions (e.g., contact tracing and self-isolation), which depends on time since infection: if the generation-interval distributions are different, the effectiveness of interventions will also differ.
+For now, we consider consequences of assuming a constant relative strength $\theta$ and predicting the relative speed $\delta$---an assumption often made (either implicitly or explicitly) in studies that have tried to predict the future spread of the new variant, including when it will replace the wild type.
+As our baseline scenario we assume $\theta = 1.5$.
 
@@
 (A) Estimated relative speed of the new variant $\hat{\delta}$ given strength of the wild type $\RR_w$ assuming identical generation-interval distributions.
-(B) True relative speed of the new variant $\delta$ given strength of the wild type $\RR_w$ accounting for longer generation intervals of the new variant.
+(B) True relative strength of the new variant $\delta$ given strength of the wild type $\RR_w$ and ratio between the mean generation interval of the new variant $\bar{G}_v$ and that of the wild type $\bar{G}_w$..
 (C) Difference between the true and the estimated relative speed of the new variant $\delta - \hat{\delta}$.
-True relative speed of the new variant $\delta$ is calculated using $\theta=1.8$, $\bar{G}_w = 5\,\textrm{days}$, and $\bar{G}_v = 7\,\textrm{days}$.
+True relative speed of the new variant $\delta$ is calculated using $\theta=1.5$, $\bar{G}_w = 5\,\textrm{days}$, and $\kappa=1/5$.
 }
@@
 \begin{equation}
-\hat{\delta} = \frac{(\theta \RR_w)^{\kappa_w} - \RR_w^{\kappa_w}}{\kappa_w \bar{G}_w}.
+\hat{\delta} = \frac{(\theta \RR_w)^{\kappa} - \RR_w^{\kappa}}{\kappa \bar{G}_w}.
 \end{equation}
-In this case, the estimated $\hat{\delta}$ increases as the underlying strength of the wild type $\RR_w$ increases (\fref{relspeed}A).
-Assuming a narrower distribution causes $\hat{\delta}$ to be less sensitive to $\RR_w$;
+Here, we see that the estimated $\hat{\delta}$ increases as the underlying strength of the wild type $\RR_w$ increases (\fref{relspeed}A).
+In other words, we expect the relative speed of the new variant to change as epidemiological conditions change, suggesting that a logistic model that assumes constant growth rate may not be appropriate for modeling the proportion of the new variant.
+A narrower distribution causes $\hat{\delta}$ to be less sensitive to $\RR_w$;
 when $\kappa = 0$ (delta-distributed generation intervals), $\hat{\delta}$ remains constant.
 
-When we allow the mean generation interval of the new variant to be long ($\bar{G}_v > \bar{G}_w$), the true $\theta$--$\delta$ relationship is given by:
+When we allow the mean generation interval of the new variant to be different ($\bar{G}_v \neq \bar{G}_w$), the true $\theta$--$\delta$ relationship under gamma approximation is given by \citep{park2019practical}:
 \begin{equation}
@@
 \end{equation}
-In this case, the direction of the effect of $\RR_w$ on $\delta$ can change depending on the value of $\kappa$.
-As $\kappa \to 1$, we have
+In this case, $\delta$ generally decreases as the mean generation interval of the new variant increases, except when both $\RR_v < 1$ and $\RR_w < 1$ (\fref{relspeed}B).
+In some extreme cases (e.g., when $\bar{G}_v$ and $\RR_v$ are very low or when $\bar{G}_v$ and $\RR_v$ are very large), we can obtain $\delta < 0$---an unlikely scenario given that fast replacement of the new variant have been observed under mixed conditions.
+This suggests that differences in generation intervals between the new variant and the wild type are likely to be not too big.
+Finally, assuming equal generation intervals generally leads to both under- and over-estimation of $\delta$ (\fref{relspeed}C).
+
+\section{Inferring relative strength from incidence data, in theory}
+
+In practice, epidemiological conditions change over time due to intervention measures and susceptible depletion effects.
+Such changes can be characterized by ``time-varying'' or ``instantaneous'' reproduction numbers, $\RR(t)$ \citep{fraser2007estimating}.
+Assuming that the generation-interval distribution does not change over time, the instantaneous reproduction numbers of the new variant and of the wild type can be estimated from their corresponding incidence curves---an approach popularized by \cite{cori2013new}:
 \begin{equation}
-\delta \to \frac{(\bar{G}_w \theta - \bar{G}_v) \RR_w + \bar{G}_v - \bar{G}_w}{\bar{G}_v \bar{G}_w},
+\RR_x(t) = \frac{i_x(t)}{\int_0^\infty i_x(t-\tau) g_x(\tau) \dtau}.
 \end{equation}
-which changes linearly with $\RR_w$ with a slope of $(\bar{G}_w \theta - \bar{G}_v)/\bar{G}_v \bar{G}_w$ (which can be either negative or positive).
-For the values we consider, this increases with $\RR_w$ (\fref{relspeed}B).
-On the other hand, as $\kappa \to 0$, we have
+Under strength-like intervention measures that reduce transmission rates by a constant amount, we expect ratios between true reproduction numbers to remain constant and correspond to the true relative strength: $\RR_v(t)/\RR_w(t) = \theta$.
+However, if the assumed generation-interval distribution $\hat{g}(\tau)$ is different from the true distribution, then the ratio between the estimated reproduction numbers $\hat{\theta}(t) = \hat{\RR}_v(t)/\hat{\RR}_w(t)$ need not be constant, where
 \begin{equation}
-\delta \to \frac{\log(\theta \RR_w)}{\bar{G}_v} - \frac{\log(\RR_w)}{\bar{G}_w},
+\hat{\RR}_x(t) = \frac{i_x(t)}{\int_0^\infty i_x(t-\tau) \hat{g}(\tau) \dtau}.
 \end{equation}
-which always decreases with $\RR_w$ provided that $\bar{G}_v > \bar{G}_w$ (\fref{relspeed}B).
-Overall, estimates of $\delta$ become less sensitive to values of $\RR_w$ and range between 0.05/day and 0.15/day, which is consistent with the observed values of $\delta$ from data.
-Therefore, assuming equal generation intervals generally leads to over-estimation of $\delta$ (\fref{relspeed}C).
+
+Here, we simulate three different scenarios using a two strain renewal equation that assumes perfect cross-immunity (\fref{Rtbias}): 
+(1) the wild type and the variant have the same generation-interval distributions, which match the assumed distribution (\fref{Rtbias}A--D);
+(2) the assumed distribution matches the generation-interval distribution of the wild type but has a longer mean than the mean generation interval of the variant (\fref{Rtbias}E--H); and
+(3) the assumed distribution matches the generation-interval distribution of the wild type but has a shorter mean than the mean generation interval of the variant (\fref{Rtbias}I--L).
+Then, we compare the estimated ratio $\hat{\theta}(t) = \hat{\RR}_v(t)/\hat{\RR}_w(t)$ with the true ratio $\theta = \RR_v(t)/\RR_w(t)$.
+For simplicity, we did not model non-pharmaceutical interventions;
+instead, we let epidemic dynamics to be determined by susceptible depletion alone.
+
+\begin{figure}[!th]
+\includegraphics[width=\textwidth]{Rtbias.pdf}
+\caption{
+\textbf{Estimates of relative strength over time under different scenarios.}
+(A, E, I) Daily incidence of infection caused by the new variant and the wild type.
+(B, F, J) True (solid lines) and estimated (dashed lines) reproduction numbers of the new variant and the wild type over time.
+(C, G, K) True (purple) and estimated (orange) ratios between reproduction numbers of the new variant and the wild type over time.
+(D, H, L) Relationship between estimated reproduction numbers on a log-log sacle.
+The dashed lines represent the one-to-one line. 
+The dotted lines represent the $\RR_v=\theta \RR_w$ line. 
+Top row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type and the variant.
+Middle row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type but is longer than that of the varant (3.5 days).
+Bottom row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type but is longer than that of the varant (6.5 days).
+For all simulations, squared coefficient of variation in generation intervals is assumed to equal $\kappa = 1/5$.
+}
+\label{fig:Rtbias}
+\end{figure}
+
+When the assumed distribution matches the true distributions, the estimated ratio remains constant (\fref{Rtbias}C).
+However, if the generation-interval distribution of the variant is different from the assumed distribution, the ratio changes over time (\fref{Rtbias}G,K).
+In particular, if the true generation intervals have a shorter (or longer) mean than the assumed distribution, we over-estimate (or under-estimate) $\RR_v(t)$ during the growth phase and under-estimate (or over-estimate) $\RR_v(t)$ during the decay phase (\fref{Rtbias}F,J), which further translates to biases in the estimated relative strength (\fref{Rtbias}G,K).
+
+We can also assess these biases by plotting the estimated strength of the variant $\hat{\RR}_v(t)$ against the estimated strength of the wild type $\hat{\RR}_w(t)$ on a log-log scale (\fref{Rtbias}D,H,L).
+When generation-interval distributions are correctly specified, we obtain a straight line parallel to the one-to-one line. 
+In this case, the slope corresponds to the logarithm of the relative strength $\log(\theta)$ (\fref{Rtbias}D).
+When the assumed mean generation interval is longer (or shorter) than the that of the variant, we over-estimate (or under-estimate) the slope (\fref{Rtbias}H,L) on a log-log plot.
+
+\section{Inferring relative strength from case data, in practice}
+
+Seb Funk's analysis here.
 

### 2021-03-14 Rtbias.R

@@ -12,32 +12,16 @@ source("renewal_det.R")
 
 theta <- 1.5
 genfun_assumed <- function(x) dgamma(x, 5, 5/5)
-genfun_short <- function(x) dgamma(x, 5, 5/4)
-genfun_long <- function(x) dgamma(x, 5, 5/6)
+genfun_short <- function(x) dgamma(x, 5, 5/3.5)
+genfun_long <- function(x) dgamma(x, 5, 5/6.5)
 
 svec <- c("Wild type GI = Variant GI = Assumed GI",
-          "Wild type GI = Variant GI < Assumed GI",
-          "Assumed GI < Wild type GI = Variant GI",
-          "Variant GI < Assumed GI < Wild type GI",
-          "Wild type GI < Assumed GI < Variant GI",
-          "Variant GI < Wild type GI = Assumed GI",
-          "Wild type GI = Assumed GI < Variant GI")
+          "Variant GI $<$ Wild type GI = Assumed GI",
+          "Wild type GI = Assumed GI $<$ Variant GI")
 
 slist <- list(
   list(genfun1=genfun_assumed,
        genfun2=genfun_assumed,
-       genfun3=genfun_assumed),
-  list(genfun1=genfun_short,
-       genfun2=genfun_short,
-       genfun3=genfun_assumed),
-  list(genfun1=genfun_long,
-       genfun2=genfun_long,
-       genfun3=genfun_assumed),
-  list(genfun1=genfun_long,
-       genfun2=genfun_short,
-       genfun3=genfun_assumed),
-  list(genfun1=genfun_short,
-       genfun2=genfun_long,
        genfun3=genfun_assumed),
   list(genfun1=genfun_assumed,
        genfun2=genfun_short,
@@ -56,18 +40,18 @@ for (i in 1:length(svec)) {
     tmax=150, slist[[i]])
   
   rr <- do.call(renewal_det, tmparg) %>%
-    filter(tvec > 15)
-  
+    filter(tvec > 15, tvec < 95)
   
   g1 <- ggplot(rr) +
     geom_line(aes(tvec, Ivec1, color="Wild type")) +
     geom_line(aes(tvec, Ivec2, color="Variant")) +
-    scale_x_continuous("Time (days)", expand=c(0, 0), limits=c(15, 110)) +
-    scale_y_continuous("Incidence (1/days)") +
[…]

### 2021-03-14 newvariant.tex
done

@@
 \usepackage{natbib}
-\bibliographystyle{chicago}
+\bibliographystyle{unsrt}
 \date{\today}
@@
 When the generation-interval distributions are assumed to be identical, we find that estimates of $\hat{\theta}$ decreases with $r_w$ whereas $\hat{\delta}$ increase with $\RR_w$.
-When we account for longer generation intervals of the new variant, these relationships can change directions depending on the effective dispersion in generation intervals, which we characterized in terms of the squared coefficient of variation of the gamma distribution $\kappa$.
-Therefore, even a modest differences in the mean generation intervals (2 days, in this case) can lead to considerable biases in the inference---we find that 1.1--1.4 fold under-estimation in the relative strength of the new variant is plausible.
+When we account for possible differences in generation-interval distributions, these relationships can change directions and further lead to biases in these estimates.
 
@@
 
-Our simple framework further reveals counterintuitive, but important, insight.
-In general, one would expect that stronger intervention strategies is required in order to slow down the replacement of the original strain by the new variant.
+Our simple framework further reveals counterintuitive, but important, insights on how to control the spread of both wild type and variants.
+In general, one would expect that stronger intervention strategies are required in order to slow down the replacement of the original strain by the new variant.
 However, this is only true if both have the same mean generation intervals (\fref{relspeed}B).
 If the new variant has longer generation intervals than the original strain, imposing strong intervention (therefore reducing both $\RR_w$ and $\RR_v$ by a constant amount) can result in a higher relative speed $\delta$, and therefore faster replacement of the original strain (\fref{relspeed}B).
-This is not necessarily a bad news as the absolute speed ($r_v$) of the new variant still decreases as $\RR_v$ decreases.
+This is not necessarily a bad outcome as the absolute speed ($r_v$) of the new variant still decreases as $\RR_v$ decreases.
 These effects should be taken into consideration for accurate projection of the spread of the new variant.
 
-Even though SARS-CoV-2 has been spreading for more than a year, there is still large uncertainty in its generation intervals.
+Even though SARS-CoV-2 has been spreading for more than a year, there is still considerable uncertainty in its generation intervals.
 Observational studies typically focus on serial intervals (i.e., time between symptom onset of the infector and the infectee; \cite{svensson2007note}) because they are easier to measure \citep{griffin2020rapid}, but they are subject to dynamical biases that can be difficult to tease apart \citep{park2021forward}.
@@
 
-Since the beginning of the SARS-CoV-2 pandemic, the importance of understanding the role of generation-interval distributions in its spread has been constantly emphasized \citep{doi:10.1098/rsif.2020.0144}.
-Yet, we still feel that generation intervals are under-appreciated, often viewed as a leeway to estimating $\RR$.
-Future studies should consider the possibility that the new variant might have longer generation intervals and evaluate how assumptions about generation intervals affect their conclusions.
+Since the beginning of the SARS-CoV-2 pandemic, a few studies have emphasized  the relevance of leveraging generation interval distributions to improve estimates of strength (e.g., \cite{doi:10.1098/rsif.2020.0144}) and the relative importance of modes of transmission (e.g., asymptomatic vs.
+symptomatic, \cite{park2020time}).
+These same lessons have relevance for assessing the spread of variants in a partially susceptible population and improving efforts to control spread.
+Future studies should prioritize detailded assessment of the relative generation intervals of SARS-CoV-2 and widespread variants, as well as evaluate how uncertainty in assumptions about generational intervals might bias conclusions.
 

### 2021-03-19 newvariant.tex
Section 1 thoughts

@@
 
+\newcommand{\rx}[1]{\ensuremath{{r}_{#1}}\xspace} 
+\newcommand{\ly}[1]{\lx{\mathrm{#1}}}
+
 \newcommand{\Rx}[1]{\ensuremath{{\mathcal R}_{#1}}\xspace} 
+\newcommand{\Ry}[1]{\Rx{\mathrm{#1}}}
 \newcommand{\Ro}{\Rx{0}}
@@
 In addition, most studies have assumed that the generation-interval distributions do not differ between the existing strain and the variant, but recent evidence suggests that the new variant may have a longer duration of infection: 13.3 days (90\% CI: 10.1--16.5) for the new variant and 8.2 days (90\% CI: 6.5--9.7) for the wild type \citep{kissler2021densely}.
-Longer duration of infection could lead to longer periods of transmission, meaning that the mean generation interval of the new variant is likely to be longer than that of the wild type.
-Other studies have also considered the possibility that the new variant may have shorter generation intervals to account for faster growth rate \citep{davies2021estimated}.
-More information is needed, however: generation intervals depend on many other biological and behavioral factors, including for example self-isolation after symptom onset leading to shorter generation intervals.
-Nonetheless, if the generation-interval distribution of the new variant is different from the generation-interval distribution of the wild type, current estimates of the strength of the new variants may be biased.
+This implies that the mean generation interval of the new variant is likely to be longer than that of the wild type.
+Other studies have considered the possibility that the faster growth rate of the new variant may be driven by \emph{shorter} generation intervals \citep{davies2021estimated}.
+If the generation-interval distribution of the new variant is different from the generation-interval distribution of the wild type, current estimates of the strength of the new variants may be biased.
+More information is clearly needed, however.
+And the story may be complicated: generation intervals depend on many other biological and behavioral factors, including for example self-isolation after symptom onset leading to shorter generation intervals.
 
-Here, we explore how different assumptions about the new variant affect estimates of its strength and speed given the possibility that it might have different generation intervals.
-We examine how the estimates of relative strength (i.e., the ratio between the strength of the existing strain and the variant) and speed (i.e., the difference between the speed of the existing strain and the variant) depend on each other.
-We then show how neglecting differences in the generation-interval distributions can lead to biased estimates, which, in turn, infleunce mitigation priorities.
+Here, we explore how different assumptions about the relative generation time of the new variant affect estimates of its strength and speed.
+We do this by exploring the relationship between relative strength (the ratio $\Ry{var}/\Ry{wt}$) and relative speed (the difference $\ry{var}-\ry{wt}$).
+We find that neglecting differences in the generation-interval distributions can lead to biased estimates, which, in turn, influence mitigation priorities.
+\jd{Just thoughts.
+I know we haven't introduced this terminology yet, and I know that you're going to use different terminology below.
+The point is that the old version was clunky and I'm trying to sharpen it.
+We can keep working.}
 
@@
 We use the renewal equation framework to characterize the spread of two pathogen strains---in this case, the wild type SARS-CoV-2 virus and a focal variant of concern (e.g., B.1.1.7).
-Neglecting the (relatively slow) rate of new mutations, the current incidence of infection $i_x(t)$ caused by the wild type ($x=w$) and the variant ($x=v$) can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary cases are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
+Neglecting the (relatively slow) rate of new mutations, the current incidence of infection $i_x(t)$ caused by the wild type $w$ and the variant $v$) can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary cases are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
 \begin{equation}

### 2021-03-19 newvariant.tex
control seq typos!
some more random comments

@@
 \newcommand{\rx}[1]{\ensuremath{{r}_{#1}}\xspace} 
-\newcommand{\ly}[1]{\lx{\mathrm{#1}}}
+\newcommand{\ry}[1]{\rx{\mathrm{#1}}}
 
@@
 We use the renewal equation framework to characterize the spread of two pathogen strains---in this case, the wild type SARS-CoV-2 virus and a focal variant of concern (e.g., B.1.1.7).
-Neglecting the (relatively slow) rate of new mutations, the current incidence of infection $i_x(t)$ caused by the wild type $w$ and the variant $v$) can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary cases are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
+Neglecting the (relatively slow) rate of new mutations, the current incidence of infection $i_x(t)$ caused by the wild type $w$ and the variant $v$ can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary cases are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
 \begin{equation}
@@
 The integral of the kernel $\RR_x(t) = \int K_x(t, \tau) \dtau$---referred to as the instantaneous reproduction number---describes the average infectiousness of previously infected individuals at time $t$ \citep{fraser2007estimating}.
+\jd{This is a particular kind of weighted average – in particular, it's not weighted by the actual number of infected individuals present.
+I kind of want to say GI-weighted average, but then we would have to introduce GI as an abbrv.}
 The normalized kernel $g_x(t, \tau) = K_x(t, \tau)/\RR_x(t)$---which we refer to as the instantaneous generation-interval distribution---describes their relative contribution to current incidence $i_x(t)$ and provides information about the time scale of disease transmission.
@@
 Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$.
-In this case, the incidence of either strains can exhibit exponential growth or decay at rate $r_x$, satisfying the Euler-Lotka equation \citep{wallinga2007generation}:
+In this case, the incidence of each strain will exhibit exponential growth (or decay) at rate $r_x$, satisfying the Euler-Lotka equation \citep{wallinga2007generation}:
 \begin{equation}
@@
 Studies of new SARS-CoV-2 variants, however, have mostly focused on characterizing changes in \emph{proportion} of a new variant \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
+\jd{I probably added patterns below, but now I want to drop it.
+Sometimes the testing pattern is: we just noticed the new variant and so let's go and test intensively in the communities where we think it is.}
 This has the advantage of correcting for testing patterns and intensity, and possibly for other transient effects that might affect variants and wild type viruses similarly.
@@
 As our baseline scenario, we assume $\delta = 0.1/\textrm{days}$ based on the observed relative speed of the new variant in the UK \citep{davies2021estimated}.
-This speed can be then used to estimate the relative strength $\theta = \RR_v/\RR_w$ of the new variant with a caveat; such inferece depends on assumptions about the generation-interval distribution.
+This speed can be then used to estimate the relative strength $\theta = \RR_v/\RR_w$ of the new variant with a caveat; such inference depends on assumptions about the generation-interval distribution.
 The mean generation interval $\bar{G}_w$ of the wild type is assumed to equal 5 days \citep{ferretti2020quantifying}.

### 2021-03-25 newvariant.tex
tweaking a bit

--- 2021-03-19 newvariant.tex
@@
 This implies that the mean generation interval of the new variant is likely to be longer than that of the wild type.
-Other studies have considered the possibility that the faster growth rate of the new variant may be driven by \emph{shorter} generation intervals \citep{davies2021estimated}.
+Other studies have considered the possibility that the faster growth rate of the new variant may be driven, in part, by \emph{shorter} generation intervals \citep{davies2021estimated,volz2021transmission}.
 If the generation-interval distribution of the new variant is different from the generation-interval distribution of the wild type, current estimates of the strength of the new variants may be biased.
@@
 
-Here, we explore how different assumptions about the relative generation time of the new variant affect estimates of its strength and speed.
+Here, we explore how different assumptions about the relative generation time of the new variant affect estimates of its strength $\RR$ and speed $r$.
 We do this by exploring the relationship between relative strength (the ratio $\Ry{var}/\Ry{wt}$) and relative speed (the difference $\ry{var}-\ry{wt}$).
@@
 We use the renewal equation framework to characterize the spread of two pathogen strains---in this case, the wild type SARS-CoV-2 virus and a focal variant of concern (e.g., B.1.1.7).
-Neglecting the (relatively slow) rate of new mutations, the current incidence of infection $i_x(t)$ caused by the wild type $w$ and the variant $v$) can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary cases are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
+Neglecting the (relatively slow) rate of new mutations, the current incidence of infection $i_x(t)$ caused by the wild type (``wt'') and the variant (``var'') can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary cases are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
 \begin{equation}

### 2021-03-25 newvariant.tex

--- 2021-03-19 newvariant.tex
@@
 We use the renewal equation framework to characterize the spread of two pathogen strains---in this case, the wild type SARS-CoV-2 virus and a focal variant of concern (e.g., B.1.1.7).
-Neglecting the (relatively slow) rate of new mutations, the current incidence of infection $i_x(t)$ caused by the wild type $w$ and the variant $v$ can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary cases are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
+Neglecting the (relatively slow) rate of new mutations, the current incidence of infection $i_x(t)$ caused by the wild type (``wt'') and the variant (``var'') can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary cases are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
 \begin{equation}

### 2021-03-25 newvariant.tex
more tweaks

@@
 
-The integral of the kernel $\RR_x(t) = \int K_x(t, \tau) \dtau$---referred to as the instantaneous reproduction number---describes the average infectiousness of previously infected individuals at time $t$ \citep{fraser2007estimating}.
-\jd{This is a particular kind of weighted average – in particular, it's not weighted by the actual number of infected individuals present.
-I kind of want to say GI-weighted average, but then we would have to introduce GI as an abbrv.}
+The integral of the kernel $\RR_x(t) = \int K_x(t, \tau) \dtau$ is referred to as the instantaneous reproduction number \citep{fraser2007estimating}.
+This is a particular kind of weighted average which describes the average infectiousness of previously infected individuals at time $t$---in particular, it is not weighted by the actual number of infected individuals present, but rather by their relative infectiousness at time $t$.
 The normalized kernel $g_x(t, \tau) = K_x(t, \tau)/\RR_x(t)$---which we refer to as the instantaneous generation-interval distribution---describes their relative contribution to current incidence $i_x(t)$ and provides information about the time scale of disease transmission.

### 2021-03-25 newvariant.tex
updates

@@
 \newcommand{\dsigma}{\dd{\sigma}}
+
+\newcommand{\ix}[1]{\ensuremath{{i}_{#1}}\xspace} 
+\newcommand{\iy}[1]{\ix{\mathrm{#1}}}
+
+\newcommand{\Gx}[1]{\ensuremath{{\bar G}_{#1}}\xspace} 
+\newcommand{\Gy}[1]{\Gx{\mathrm{#1}}}
 
@@
 We use this framework to investigate how inferences about strength and speed of the variant depend on our assumptions about the underlying generation-interval distributions.
-For simplicity we neglect differences in the squared coefficient of variation and assume $\kappa_w = \kappa_v = \kappa$; instead, we focus on the effect of potential differences in the mean generation intervals.
+For simplicity we neglect differences in the squared coefficient of variation and assume $\kappa_{\mathrm{wt}} = \kappa_{\mathrm{var}} = \kappa$; instead, we focus on the effect of potential differences in the mean generation intervals.
 
@@
 
-Epidemic speed $r_x$ is often estimated directly from case data \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}.
-Studies of new SARS-CoV-2 variants, however, have mostly focused on characterizing changes in \emph{proportion} of a new variant \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
-\jd{I probably added patterns below, but now I want to drop it.
-Sometimes the testing pattern is: we just noticed the new variant and so let's go and test intensively in the communities where we think it is.}
-This has the advantage of correcting for testing patterns and intensity, and possibly for other transient effects that might affect variants and wild type viruses similarly.
+Epidemic speed $r_x$ can be often estimated directly from case data \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}.
+However, estimating epidemic speed from case data can be challenging when case counts are low (e.g., in the beginning of the spread of the new variant) and is sensitive to temporal changes in testing patterns and intensity.
+On the other hand, studies of new SARS-CoV-2 variants have mostly focused on characterizing changes in \emph{proportion} of a new variant \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
+This has the advantage of being less sensitive to such changes in testing, and possibly for other transient effects that might affect variants and wild type viruses similarly.
 When incidence is changing exponentially ($i_x(t) = i_x(t_0) \exp(r_x t)$), the proportion of the new variant $p(t)$ follows a logistic equation:
 \begin{align}
-p(t) &= \frac{i_v(t_0) \exp(r_v t)}{i_w(t_0) \exp(r_w t) + i_v(t_0) \exp(r_v t)},
-\\ &= \frac{1}{1 + \left(i_w(t_0)/i_v(t_0)\right) \exp(-\delta t)},
+p(t) &= \frac{\iy{var}(t_0) \exp(\ry{var} t)}{\iy{wt}(t_0) \exp(\ry{wt} t) + \iy{var}(t_0) \exp(\ry{var} t)},
+\\ &= \frac{1}{1 + \left(\iy{wt}(t_0)/\iy{var}(t_0)\right) \exp(-\delta t)},
 \end{align}
-where the logistic growth rate $\delta = r_v - r_w$ corresponds to the relative speed of the epidemic.
+where the logistic growth rate $\delta = \ry{var} - \ry{wt}$ corresponds to the relative speed of the epidemic.
 
@@
 As our baseline scenario, we assume $\delta = 0.1/\textrm{days}$ based on the observed relative speed of the new variant in the UK \citep{davies2021estimated}.
-This speed can be then used to estimate the relative strength $\theta = \RR_v/\RR_w$ of the new variant with a caveat; such inference depends on assumptions about the generation-interval distribution.
+This speed can be then used to estimate the relative strength $\theta = \Ry{var}/\Ry{wt}$ of the new variant with a caveat; such inference depends on assumptions about the generation-interval distribution.
 The mean generation interval $\bar{G}_w$ of the wild type is assumed to equal 5 days \citep{ferretti2020quantifying}.
@@
 
-Previous studies have assumed that the generation-interval distributions do not differ between the wild type and the variant: $\bar{G}_v = \bar{G}_w = \bar{G}$.
+Previous studies have assumed that the generation-interval distributions do not differ between the wild type and the variant: $\Gy{var} = \Gy{wt} = \bar{G}$.
 In this case, the estimate of the relative strength is given by \citep{park2019practical}:
 \begin{equation}
-\hat{\theta} = \left(1 + \frac{\kappa \delta \bar{G}}{1 + \kappa r_w \bar{G}}\right)^{1/\kappa}.
+\hat{\theta} = \left(1 + \frac{\kappa \delta \bar{G}}{1 + \kappa \ry{wt} \bar{G}}\right)^{1/\kappa}.
 \end{equation}
@@
 Instead, if the wild type was spreading faster than previously assumed (higher $r_w$), the estimates of the relative strength $\theta$ will be go down (\fref{relstrength}A).
-The relative strength also depends on the \emph{effective dispersion} in generation intervals (i.e., the amount of variability in generation intervals, as measured by $\kappa$)---using narrower distributions makes $\hat{\theta}$ less sensitive to $r_w$.
+The relative strength also depends on the \emph{effective dispersion} in generation intervals (i.e., the amount of variability in generation intervals, as measured by $\kappa$)---using narrower distributions makes $\hat{\theta}$ less sensitive to $\ry{wt}$.
 

### 2021-03-25 relstrength.R

@@ -62,7 +62,7 @@ thetadata2 <- apply(pardata2, 1, function(x) {
 
 g1 <- ggplot(thetadata) +
   geom_line(aes(rw, thetahat, lty=as.factor(kappa), col=as.factor(kappa))) +
-  scale_x_continuous("Wild type speed, $r_w$ (1/days)", expand=c(0, 0)) +
+  scale_x_continuous("Wild type speed, $r_{\\mathrm{wt}}$ (1/days)", expand=c(0, 0)) +
   scale_y_continuous("Estimated relative strength, $\\hat{\\theta}$", limits=c(1, 2.7), expand=c(0, 0)) +
   scale_color_colorblind("$\\kappa$") +
   scale_linetype_discrete("$\\kappa$") +
@@ -76,10 +76,10 @@ g2 <- ggplot(thetadata2) +
   geom_tile(aes(Gratio, rw, fill=theta)) +
   geom_vline(xintercept=1, lty=2, col="white") +
   geom_hline(yintercept=-delta, lty=1, col="white") +
-  annotate("text", x=1.04, y=0, label=c("$\\bar{G}_v=\\bar{G}_w$"), col="white", angle=-90) +
-  annotate("text", x=0.8, y=-delta+0.01, label=c("$r_v=0$"), col="white") +
-  scale_x_continuous("Generation interval ratio, $\\bar{G}_v/\\bar{G}_w$", expand=c(0, 0)) +
-  scale_y_continuous("Wild type speed, $r_w$ (1/days)", expand=c(0, 0)) +
+  annotate("text", x=1.04, y=0, label=c("$\\bar{G}_{\\mathrm{var}}=\\bar{G}_{\\mathrm{wt}}$"), col="white", angle=-90) +
+  annotate("text", x=0.8, y=-delta+0.01, label=c("$r_{\\mathrm{var}}=0$"), col="white") +
+  scale_x_continuous("Generation interval ratio, $\\bar{G}_{\\mathrm{var}}/\\bar{G}_{\\mathrm{wt}}$", expand=c(0, 0)) +
+  scale_y_continuous("Wild type speed, $r_{\\mathrm{wt}}$ (1/days)", expand=c(0, 0)) +
   scale_fill_viridis_c("$\\theta$") +
   ggtitle("B. Different generation intervals") +
   theme(
@@ -90,10 +90,10 @@ g3 <- ggplot(thetadata2) +
[…]

### 2021-03-26 newvariant.tex
tweak figure and convert theta to rho

@@
 	\textbf\newline{
-		Characterizing relative epidemic strength and speed of new SARS-CoV-2 variants
+		Roles of generation-interval distributions in characterizing relative epidemic strength and speed of new SARS-CoV-2 variants and their control
 	}
@@
 \textbf{Relative strength of the new variant given observed relative speed.}
-(A) Estimated relative strength of the new variant $\hat{\theta}$ given speed of the wild type
+(A) Estimated relative strength of the new variant $\hat{\rho}$ given speed of the wild type
 $r_w$ assuming identical generation-interval distributions.
-(B) True relative strength of the new variant $\theta$ given speed of the wild type $r_w$ and ratio between the mean generation interval of the new variant $\bar{G}_v$ and that of the wild type $\bar{G}_w$.
-(C) Ratio between the true and the estimated relative strength of the new variant $\theta/\hat{\theta}$.
-True relative strength of the new variant $\theta$ is calculated using $\delta=0.1\,\textrm{days}$, $\bar{G}_w = 5\,\textrm{days}$, and $\kappa = 1/5$.
+(B) True relative strength of the new variant $\rho$ given speed of the wild type $r_w$ and ratio between the mean generation interval of the new variant $\bar{G}_v$ and that of the wild type $\bar{G}_w$.
+(C) Ratio between the true and the estimated relative strength of the new variant $\rho/\hat{\rho}$.
+True relative strength of the new variant $\rho$ is calculated using $\delta=0.1\,\textrm{days}$, $\bar{G}_w = 5\,\textrm{days}$, and $\kappa = 1/5$.
 }
@@
 As our baseline scenario, we assume $\delta = 0.1/\textrm{days}$ based on the observed relative speed of the new variant in the UK \citep{davies2021estimated}.
-This speed can be then used to estimate the relative strength $\theta = \Ry{var}/\Ry{wt}$ of the new variant with a caveat; such inference depends on assumptions about the generation-interval distribution.
+This speed can be then used to estimate the relative strength $\rho = \Ry{var}/\Ry{wt}$ of the new variant with a caveat; such inference depends on assumptions about the generation-interval distribution.
 The mean generation interval $\bar{G}_w$ of the wild type is assumed to equal 5 days \citep{ferretti2020quantifying}.
 The squared coefficient of variation $\kappa$ in generation intervals os assumed to equal 0.2 \citep{ferretti2020quantifying}---for simplicity, we assume that both the variant and the wild type have equal $\kappa$ and only consider differences in the mean.
-As we expect there to be greater uncertainty in $\kappa$ than in $\bar{G}$, we evaluate the estimates of relative strength $\theta$ across a wide range of $\kappa$, ranging from 0 (delta distribution) to 1 (exponential distribution).
+As we expect there to be greater uncertainty in $\kappa$ than in $\bar{G}$, we evaluate the estimates of relative strength $\rho$ across a wide range of $\kappa$, ranging from 0 (delta distribution) to 1 (exponential distribution).
 
@@
 \begin{equation}
-\hat{\theta} = \left(1 + \frac{\kappa \delta \bar{G}}{1 + \kappa \ry{wt} \bar{G}}\right)^{1/\kappa}.
+\hat{\rho} = \left(1 + \frac{\kappa \delta \bar{G}}{1 + \kappa \ry{wt} \bar{G}}\right)^{1/\kappa}.
 \end{equation}
-Therefore, estimates of relative strength $\theta$ depends not only on the relative speed $\delta$ but also on how fast the wild type is spreading in the population $r_w$---a few analyses have implicitly or explicitly neglected this factor by either assuming $r_w = 0$ \citep{switzerland2021variant} or $\kappa = 0$ \citep{davies2021estimated}, in which case we get $\hat{\theta} = \exp(\delta \bar{G})$.
-Instead, if the wild type was spreading faster than previously assumed (higher $r_w$), the estimates of the relative strength $\theta$ will be go down (\fref{relstrength}A).
-The relative strength also depends on the \emph{effective dispersion} in generation intervals (i.e., the amount of variability in generation intervals, as measured by $\kappa$)---using narrower distributions makes $\hat{\theta}$ less sensitive to $\ry{wt}$.
+Therefore, estimates of relative strength $\rho$ depends not only on the relative speed $\delta$ but also on how fast the wild type is spreading in the population $r_w$---a few analyses have implicitly or explicitly neglected this factor by either assuming $r_w = 0$ \citep{switzerland2021variant} or $\kappa = 0$ \citep{davies2021estimated}, in which case we get $\hat{\rho} = \exp(\delta \bar{G})$.
+Instead, if the wild type was spreading faster than previously assumed (higher $r_w$), the estimates of the relative strength $\rho$ will be go down (\fref{relstrength}A).
+The relative strength also depends on the \emph{effective dispersion} in generation intervals (i.e., the amount of variability in generation intervals, as measured by $\kappa$)---using narrower distributions makes $\hat{\rho}$ less sensitive to $\ry{wt}$.
 
-However, if the true mean generation interval of the new variant is different from that of the wild type ($\bar{G}_v \neq \bar{G}_w$), the true relative strength $\theta = \RR_v/\RR_w$ is given by:
+However, if the true mean generation interval of the new variant is different from that of the wild type ($\bar{G}_v \neq \bar{G}_w$), the true relative strength $\rho = \RR_v/\RR_w$ is given by:
 \begin{equation}
-\theta = \left(\frac{1 + \kappa (r_w + \delta) \bar{G}_v}{1 + \kappa r_w \bar{G}_w}\right)^{1/\kappa}.
+\rho = \left(\frac{1 + \kappa (r_w + \delta) \bar{G}_v}{1 + \kappa r_w \bar{G}_w}\right)^{1/\kappa}.
 \end{equation}
-In this case, longer (or shorter) mean generation interval of the new variant generally translates to higher (or lower) values of $\theta$---\fref{relstrength}B illustrates this effect assuming $\kappa = 1/5$.
-However, this direction can change when both the new variant and the wild type are decaying exponentially (see area below $r_v = 0$ line in \fref{relstrength}B)---when disease incidence is decaying exponentially, longer generation intervals lead to a \emph{lower} value of strength (and therefore lower relative strength $\theta$).
+In this case, longer (or shorter) mean generation interval of the new variant generally translates to higher (or lower) values of $\rho$---\fref{relstrength}B illustrates this effect assuming $\kappa = 1/5$.
+However, this direction can change when both the new variant and the wild type are decaying exponentially (see area below $r_v = 0$ line in \fref{relstrength}B)---when disease incidence is decaying exponentially, longer generation intervals lead to a \emph{lower} value of strength (and therefore lower relative strength $\rho$).
 
 Therefore, neglecting potential differences in generation intervals can lead to both under- and over-estimation of the relative strength.
-In \fref{relstrength}C, we quantify the degrees of bias using the ratio between the true value $\theta$, which accounts for differences in generation-interval distributions, and the estimated value $\hat{\theta}$, which assumes equal generation-interval distributions.
+In \fref{relstrength}C, we quantify the degrees of bias using the ratio between the true value $\rho$, which accounts for differences in generation-interval distributions, and the estimated value $\hat{\rho}$, which assumes equal generation-interval distributions.
 Across the ranges of parameters that we consider, we find that up to 1.4 fold (either upward or downward) adjustments to current estimates may be required.
@@
 While relative speed is easier to measure, we often do not expect $\delta$ to be constant---this can depend on epidemic and demographic factors and can vary across time and space.
-Instead, biological mechanisms that cause new variant to have higher strength better translate to assuming a constant value of relative strength $\theta$.
-For example, if the new variant has a higher transmission rate, the ratio between the strength of the new variant and that of the wild type will remain constant ($=\theta$) under strength-like interventions that reduces the strength of both strains by a constant amount.
+Instead, biological mechanisms that cause new variant to have higher strength better translate to assuming a constant value of relative strength $\rho$.
+For example, if the new variant has a higher transmission rate, the ratio between the strength of the new variant and that of the wild type will remain constant ($=\rho$) under strength-like interventions that reduces the strength of both strains by a constant amount.
 However, this is not necessarily true under speed-like interventions (e.g., contact tracing and self-isolation), which depends on time since infection: if the generation-interval distributions are different, the effectiveness of interventions will also differ.
-For now, we consider consequences of assuming a constant relative strength $\theta$ and predicting the relative speed $\delta$---an assumption often made (either implicitly or explicitly) in studies that have tried to predict the future spread of the new variant, including when it will replace the wild type.
-As our baseline scenario we assume $\theta = 1.5$.
+For now, we consider consequences of assuming a constant relative strength $\rho$ and predicting the relative speed $\delta$---an assumption often made (either implicitly or explicitly) in studies that have tried to predict the future spread of the new variant, including when it will replace the wild type.
+As our baseline scenario we assume $\rho = 1.5$.
 
@@
 (C) Difference between the true and the estimated relative speed of the new variant $\delta - \hat{\delta}$.
-True relative speed of the new variant $\delta$ is calculated using $\theta=1.5$, $\bar{G}_w = 5\,\textrm{days}$, and $\kappa=1/5$.
+True relative speed of the new variant $\delta$ is calculated using $\rho=1.5$, $\bar{G}_w = 5\,\textrm{days}$, and $\kappa=1/5$.
 }
@@
 \begin{equation}
-\hat{\delta} = \frac{(\theta \RR_w)^{\kappa} - \RR_w^{\kappa}}{\kappa \bar{G}_w}.
+\hat{\delta} = \frac{(\rho \RR_w)^{\kappa} - \RR_w^{\kappa}}{\kappa \bar{G}_w}.
 \end{equation}
@@
 
-When we allow the mean generation interval of the new variant to be different ($\bar{G}_v \neq \bar{G}_w$), the true $\theta$--$\delta$ relationship under gamma approximation is given by \citep{park2019practical}:
+When we allow the mean generation interval of the new variant to be different ($\bar{G}_v \neq \bar{G}_w$), the true $\rho$--$\delta$ relationship under gamma approximation is given by \citep{park2019practical}:
 \begin{equation}
-\delta = \frac{(\theta \RR_w)^{\kappa} - 1}{\kappa \bar{G}_v} - \frac{\RR_w^{\kappa} - 1}{\kappa \bar{G}_w}.
+\delta = \frac{(\rho \RR_w)^{\kappa} - 1}{\kappa \bar{G}_v} - \frac{\RR_w^{\kappa} - 1}{\kappa \bar{G}_w}.
 \end{equation}
@@
 \end{equation}
-Under strength-like intervention measures that reduce transmission rates by a constant amount, we expect ratios between true reproduction numbers to remain constant and correspond to the true relative strength: $\RR_v(t)/\RR_w(t) = \theta$.
-However, if the assumed generation-interval distribution $\hat{g}(\tau)$ is different from the true distribution, then the ratio between the estimated reproduction numbers $\hat{\theta}(t) = \hat{\RR}_v(t)/\hat{\RR}_w(t)$ need not be constant, where
+Under strength-like intervention measures that reduce transmission rates by a constant amount, we expect ratios between true reproduction numbers to remain constant and correspond to the true relative strength: $\RR_v(t)/\RR_w(t) = \rho$.
+However, if the assumed generation-interval distribution $\hat{g}(\tau)$ is different from the true distribution, then the ratio between the estimated reproduction numbers $\hat{\rho}(t) = \hat{\RR}_v(t)/\hat{\RR}_w(t)$ need not be constant, where
 \begin{equation}
@@
 (3) the assumed distribution matches the generation-interval distribution of the wild type but has a shorter mean than the mean generation interval of the variant (\fref{Rtbias}I--L).
-Then, we compare the estimated ratio $\hat{\theta}(t) = \hat{\RR}_v(t)/\hat{\RR}_w(t)$ with the true ratio $\theta = \RR_v(t)/\RR_w(t)$.
+Then, we compare the estimated ratio $\hat{\rho}(t) = \hat{\RR}_v(t)/\hat{\RR}_w(t)$ with the true ratio $\rho = \RR_v(t)/\RR_w(t)$.
 For simplicity, we did not model non-pharmaceutical interventions;
@@
 The dashed lines represent the one-to-one line. 
-The dotted lines represent the $\RR_v=\theta \RR_w$ line. 
+The dotted lines represent the $\RR_v=\rho \RR_w$ line. 
 Top row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type and the variant.
@@
 When generation-interval distributions are correctly specified, we obtain a straight line parallel to the one-to-one line. 
-In this case, the slope corresponds to the logarithm of the relative strength $\log(\theta)$ (\fref{Rtbias}D).
+In this case, the slope corresponds to the logarithm of the relative strength $\log(\rho)$ (\fref{Rtbias}D).
 When the assumed mean generation interval is longer (or shorter) than the that of the variant, we over-estimate (or under-estimate) the slope (\fref{Rtbias}H,L) on a log-log plot.
@@
 
+\section{Intervention}
+
+If $K(\tau)/L(\tau)$.
+
 \section{Discussion}
@@
 Here, we explored how the generation-interval distribution shapes the link between relative strength and speed of the new variant.
-We define relative strength as the ratio between reproduction numbers ($\theta=\RR_v/\RR_w$) and relative speed as the difference between growth rates ($\delta=r_v-r_w$) of the variant and the wild type.
-When the generation-interval distributions are assumed to be identical, we find that estimates of $\hat{\theta}$ decreases with $r_w$ whereas $\hat{\delta}$ increase with $\RR_w$.
+We define relative strength as the ratio between reproduction numbers ($\rho=\RR_v/\RR_w$) and relative speed as the difference between growth rates ($\delta=r_v-r_w$) of the variant and the wild type.
+When the generation-interval distributions are assumed to be identical, we find that estimates of $\hat{\rho}$ decreases with $r_w$ whereas $\hat{\delta}$ increase with $\RR_w$.
 When we account for possible differences in generation-interval distributions, these relationships can change directions and further lead to biases in these estimates.

### 2021-03-26 relspeed.R

@@ -4,7 +4,7 @@ library(ggthemes)
 library(egg)
 library(tikzDevice)
 
-theta <- 1.5
+rho <- 1.61
 
 Rwvec <- seq(0.3, 1.7, length.out=31)
 kappavec <- c(0, 0.2, 0.5, 1)
@@ -23,13 +23,13 @@ deltadata <- apply(pardata, 1, function(x) {
   Rw <- x[[2]]
   
   if (kappa == 0) {
-    deltahat <- log(theta * Rw)/Gw - log(Rw)/Gw
+    deltahat <- log(rho * Rw)/Gw - log(Rw)/Gw
   } else {
-    deltahat <- ((theta * Rw)^kappa - 1)/(kappa*Gw) -  ((Rw)^kappa - 1)/(kappa*Gw)
+    deltahat <- ((rho * Rw)^kappa - 1)/(kappa*Gw) -  ((Rw)^kappa - 1)/(kappa*Gw)
   }
   
   data.frame(
-    theta=theta,
+    rho=rho,
     Rw=Rw,
     deltahat=deltahat,
     kappa=kappa
@@ -42,12 +42,12 @@ deltadata2 <- apply(pardata2, 1, function(x) {
   Gratio <- x[[2]] ## Gv/Gw
   Gv <- Gratio * Gw
   
-  delta <- ((theta * Rw)^kappabase - 1)/(kappabase*Gv) -  ((Rw)^kappabase - 1)/(kappabase*Gw)
+  delta <- ((rho * Rw)^kappabase - 1)/(kappabase*Gv) -  ((Rw)^kappabase - 1)/(kappabase*Gw)
   
-  deltahat <- ((theta * Rw)^kappabase - 1)/(kappabase*Gw) -  ((Rw)^kappabase - 1)/(kappabase*Gw)
+  deltahat <- ((rho * Rw)^kappabase - 1)/(kappabase*Gw) -  ((Rw)^kappabase - 1)/(kappabase*Gw)
   
   data.frame(
-    theta=theta,
+    rho=rho,
     Gratio=Gratio,
     Rw=Rw,
     delta=delta,
@@ -72,9 +72,9 @@ g1 <- ggplot(deltadata) +
 g2 <- ggplot(deltadata2) +
   geom_tile(aes(Gratio, Rw, fill=delta)) +
   geom_vline(xintercept=1, lty=2, col="white") +
-  geom_hline(yintercept=1/theta, lty=1, col="white") +
[…]

### 2021-03-27 newvariant.tex
updates

@@
 For now, we consider consequences of assuming a constant relative strength $\rho$ and predicting the relative speed $\delta$---an assumption often made (either implicitly or explicitly) in studies that have tried to predict the future spread of the new variant, including when it will replace the wild type.
-As our baseline scenario we assume $\rho = 1.5$.
+As our baseline scenario we assume $\rho = 1.61$, which is the value we obtain for $\delta=0.1\,\textrm{days}$, $\bar{G}_w = \bar{G}_v = 5\,\textrm{days}$, and $\kappa = 1/5$.
 
@@
 (C) Difference between the true and the estimated relative speed of the new variant $\delta - \hat{\delta}$.
-True relative speed of the new variant $\delta$ is calculated using $\rho=1.5$, $\bar{G}_w = 5\,\textrm{days}$, and $\kappa=1/5$.
+True relative speed of the new variant $\delta$ is calculated using $\rho=1.61$, $\bar{G}_w = 5\,\textrm{days}$, and $\kappa=1/5$.
 }
@@
 \end{equation}
-Under strength-like intervention measures that reduce transmission rates by a constant amount, we expect ratios between true reproduction numbers to remain constant and correspond to the true relative strength: $\RR_v(t)/\RR_w(t) = \rho$.
-However, if the assumed generation-interval distribution $\hat{g}(\tau)$ is different from the true distribution, then the ratio between the estimated reproduction numbers $\hat{\rho}(t) = \hat{\RR}_v(t)/\hat{\RR}_w(t)$ need not be constant, where
+Under strength-like intervention measures that reduce transmission rates by a constant amount, we expect ratios between true reproduction numbers to remain constant and correspond to the true relative strength: $\Ry{var}(t)/\Ry{wt}(t) = \rho$.
+However, if the assumed generation-interval distribution $\hat{g}(\tau)$ is different from the true distribution, then the ratio between the estimated reproduction numbers $\hat{\rho}(t) = \hat{\RR}_{\textrm{var}}(t)/\hat{\RR}_{\textrm{wt}}(t)$ need not be constant, where
 \begin{equation}
@@
 Here, we simulate three different scenarios using a two strain renewal equation that assumes perfect cross-immunity (\fref{Rtbias}): 
-(1) the wild type and the variant have the same generation-interval distributions, which match the assumed distribution (\fref{Rtbias}A--D);
-(2) the assumed distribution matches the generation-interval distribution of the wild type but has a longer mean than the mean generation interval of the variant (\fref{Rtbias}E--H); and
-(3) the assumed distribution matches the generation-interval distribution of the wild type but has a shorter mean than the mean generation interval of the variant (\fref{Rtbias}I--L).
-Then, we compare the estimated ratio $\hat{\rho}(t) = \hat{\RR}_v(t)/\hat{\RR}_w(t)$ with the true ratio $\rho = \RR_v(t)/\RR_w(t)$.
-For simplicity, we did not model non-pharmaceutical interventions;
-instead, we let epidemic dynamics to be determined by susceptible depletion alone.
+(1) the wild type and the variant have the same generation-interval distributions, which match the assumed distribution (\fref{Rtbias}A--C);
+(2) the assumed distribution matches the generation-interval distribution of the wild type but has a longer mean than the mean generation interval of the variant (\fref{Rtbias}D--F); and
+(3) the assumed distribution matches the generation-interval distribution of the wild type but has a shorter mean than the mean generation interval of the variant (\fref{Rtbias}G--I).
+Then, we compare the estimated ratio $\hat{\rho}(t) = \hat{\RR}_{\textrm{var}}(t)/\hat{\RR}_{\textrm{wt}}(t)$ with the true ratio $\rho = \Ry{var}(t)/\Ry{wt}(t)$.
+In order to reflect smooth changes in $\RR$ driven by non-pharmaceutical interventions, we model $\Ry{wt}(t)$ phenomenologically and let $\Ry{var}(t) = \rho \Ry{wt}(t)$.
 
@@
 \textbf{Estimates of relative strength over time under different scenarios.}
-(A, E, I) Daily incidence of infection caused by the new variant and the wild type.
-(B, F, J) True (solid lines) and estimated (dashed lines) reproduction numbers of the new variant and the wild type over time.
-(C, G, K) True (purple) and estimated (orange) ratios between reproduction numbers of the new variant and the wild type over time.
-(D, H, L) Relationship between estimated reproduction numbers on a log-log sacle.
-The dashed lines represent the one-to-one line. 
-The dotted lines represent the $\RR_v=\rho \RR_w$ line. 
+(A, D, G) True (solid lines) and estimated (dashed lines) reproduction numbers of the new variant and the wild type over time.
+(B, E, H) True (purple, solid) and estimated (orange, dashed) ratios between reproduction numbers of the new variant and the wild type over time.
+(C, F, I) True (purple, solid) and estimated (orange, dashed) relationships between estimated reproduction numbers on a log-log sacle.
 Top row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type and the variant.
-Middle row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type but is longer than that of the varant (3.5 days).
-Bottom row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type but is longer than that of the varant (6.5 days).
+Middle row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type but is longer than that of the varant (4 days).
+Bottom row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type but is longer than that of the varant (6 days).
 For all simulations, squared coefficient of variation in generation intervals is assumed to equal $\kappa = 1/5$.
+We assume $\Ry{wt}(t)=2.5$ before $t=15\,\textrm{days}$ to reflect unmitigated spread and let $\Ry{wt}(t)=2.5 (1+0.7\cos((t-15)/10))/1.7$ after to model smooth introduction and lifting of non-pharmaceutical interventions.
 }
@@
 
-When the assumed distribution matches the true distributions, the estimated ratio remains constant (\fref{Rtbias}C).
-However, if the generation-interval distribution of the variant is different from the assumed distribution, the ratio changes over time (\fref{Rtbias}G,K).
-In particular, if the true generation intervals have a shorter (or longer) mean than the assumed distribution, we over-estimate (or under-estimate) $\RR_v(t)$ during the growth phase and under-estimate (or over-estimate) $\RR_v(t)$ during the decay phase (\fref{Rtbias}F,J), which further translates to biases in the estimated relative strength (\fref{Rtbias}G,K).
+When the assumed distribution matches the true distributions, the estimated reproduction numbers match the true values (\fref{Rtbias}A), and therefore the ratio remains constant (\fref{Rtbias}B).
+However, if the generation-interval distribution of the variant is different from the assumed distribution, the ratio changes over time (\fref{Rtbias}F,I).
+In particular, if the true generation intervals have a shorter (or longer) mean than the assumed distribution, we over-estimate (or under-estimate) $\Ry{var}(t)$ during the growth phase and under-estimate (or over-estimate) $ $ during the decay phase (\fref{Rtbias}D,G), which further translates to biases in the estimated relative strength (\fref{Rtbias}F,I).
 
-We can also assess these biases by plotting the estimated strength of the variant $\hat{\RR}_v(t)$ against the estimated strength of the wild type $\hat{\RR}_w(t)$ on a log-log scale (\fref{Rtbias}D,H,L).
-When generation-interval distributions are correctly specified, we obtain a straight line parallel to the one-to-one line. 
-In this case, the slope corresponds to the logarithm of the relative strength $\log(\rho)$ (\fref{Rtbias}D).
+We can assess these biases by plotting the estimated strength of the variant $\hat{\RR}_{\textrm{var}}(t)$ against the estimated strength of the wild type $\hat{\RR}_{\textrm{wt}}(t)$ on a log-log scale (\fref{Rtbias}C,F,I).
+When generation-interval distributions are correctly specified, we obtain a straight line with a slope of 1. 
+In this case, the intercept corresponds to the logarithm of the relative strength $\log(\rho)$ (\fref{Rtbias}C).
 When the assumed mean generation interval is longer (or shorter) than the that of the variant, we over-estimate (or under-estimate) the slope (\fref{Rtbias}H,L) on a log-log plot.
+Smooth changes in $\RR$ further generate circular patterns but biases in slopes are clear.
 

### 2021-03-27 relstrength.R

@@ -5,6 +5,7 @@ library(egg)
 library(tikzDevice)
 
 delta <- 0.1
+rho <- 1.61
 
 rwvec <- seq(-0.13, 0.13, length.out=31)
 kappavec <- c(0, 0.2, 0.5, 1)
@@ -18,52 +19,52 @@ Gratio <- seq(0.6, 1.4, length.out=31)
 
 pardata2 <- expand.grid(rwvec, Gratio)
 
-thetadata <- apply(pardata, 1, function(x) {
+rhodata <- apply(pardata, 1, function(x) {
   kappa <- x[[1]]
   rw <- x[[2]]
   
   if (kappa==0) {
-    thetahat <- exp(Gw * (rw + delta))/exp(Gw * rw)
+    rhohat <- exp(Gw * (rw + delta))/exp(Gw * rw)
   } else {
-    thetahat <- ((1 + kappa * (rw + delta) * Gw)^(1/kappa))/
+    rhohat <- ((1 + kappa * (rw + delta) * Gw)^(1/kappa))/
       ((1 + kappa * rw * Gw)^(1/kappa))
   }
   
   data.frame(
     delta=delta,
     rw=rw,
-    thetahat=thetahat,
+    rhohat=rhohat,
     kappa=kappa
   )
 }) %>%
   bind_rows
 
-thetadata2 <- apply(pardata2, 1, function(x) {
+rhodata2 <- apply(pardata2, 1, function(x) {
   rw <- x[[1]]
   Gratio <- x[[2]] ## Gv/Gw
   Gv <- Gratio * Gw
   
-  theta <- ((1 + kappabase * (rw + delta) * Gv)^(1/kappabase))/
+  rho <- ((1 + kappabase * (rw + delta) * Gv)^(1/kappabase))/
     ((1 + kappabase * rw * Gw)^(1/kappabase))
   
-  thetahat <- ((1 + kappabase * (rw + delta) * Gw)^(1/kappabase))/
+  rhohat <- ((1 + kappabase * (rw + delta) * Gw)^(1/kappabase))/
     ((1 + kappabase * rw * Gw)^(1/kappabase))
   
   data.frame(
     delta=delta,
     Gratio=Gratio,
     rw=rw,
-    theta=theta,
-    thetahat=thetahat,
-    bias=theta/thetahat
+    rho=rho,
+    rhohat=rhohat,
+    bias=rho/rhohat
   )
 }) %>%
   bind_rows
 
-g1 <- ggplot(thetadata) +
[…]

### 2021-03-28 control.tex
working on control

% Created by tikzDevice version 0.12.3 on 2021-03-27 20:28:52
% !TEX encoding = UTF-8 Unicode
\documentclass[10pt]{article}
\usepackage{tikz}

\usepackage[active,tightpage,psfixbb]{preview}

\PreviewEnvironment{pgfpicture}

\setlength\PreviewBorder{0pt}
\begin{document}

\begin{tikzpicture}[x=1pt,y=1pt]
\definecolor{fillColor}{RGB}{255,255,255}
\path[use as bounding box,fill=fillColor,fill opacity=0.00] (0,0) rectangle (722.70,361.35);
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (361.35,361.35);
\definecolor{drawColor}{RGB}{255,255,255}
\definecolor{fillColor}{RGB}{255,255,255}

\path[draw=drawColor,line width= 0.7pt,line join=round,line cap=round,fill=fillColor] (  0.00,  0.00) rectangle (361.35,361.35);
\end{scope}
\begin{scope}
\path[clip] ( 49.41, 39.63) rectangle (354.35,332.06);
\definecolor{fillColor}{RGB}{255,255,255}

\path[fill=fillColor] ( 49.41, 39.63) rectangle (354.35,332.06);
\definecolor{drawColor}{RGB}{0,0,0}

\path[draw=drawColor,line width= 1.1pt,line join=round] ( 63.27, 52.92) --
	( 72.51, 59.52) --
	( 81.75, 66.11) --
	( 90.99, 72.69) --
	(100.23, 79.24) --
	(109.48, 85.78) --
	(118.72, 92.31) --
	(127.96, 98.81) --
	(137.20,105.30) --
	(146.44,111.78) --
	(155.68,118.24) --
	(164.92,124.68) --
	(174.16,131.11) --
	(183.40,137.52) --
	(192.64,143.92) --
	(201.88,150.30) --
	(211.12,156.67) --
	(220.36,163.02) --
	(229.60,169.35) --
	(238.84,175.67) --
	(248.08,181.98) --
	(257.32,188.27) --
	(266.56,194.54) --
	(275.81,200.80) --
	(285.05,207.04) --
	(294.29,213.27) --
	(303.53,219.49) --
	(312.77,225.69) --
	(322.01,231.88) --
	(331.25,238.05) --
	(340.49,244.20);
\definecolor{drawColor}{RGB}{230,159,0}

\path[draw=drawColor,line width= 1.1pt,dash pattern=on 4pt off 4pt ,line join=round] ( 63.27,205.70) --
	( 72.51,205.70) --
	( 81.75,205.70) --
	( 90.99,205.70) --
	(100.23,205.70) --
	(109.48,205.70) --
	(118.72,205.70) --
	(127.96,205.70) --
	(137.20,205.70) --
	(146.44,205.70) --
	(155.68,205.70) --
	(164.92,205.70) --
	(174.16,205.70) --
	(183.40,205.70) --
	(192.64,205.70) --
	(201.88,205.70) --
	(211.12,205.70) --
	(220.36,205.70) --
	(229.60,205.70) --
	(238.84,205.70) --
	(248.08,205.70) --
	(257.32,205.70) --
	(266.56,205.70) --
	(275.81,205.70) --
	(285.05,205.70) --
	(294.29,205.70) --
	(303.53,205.70) --
	(312.77,205.70) --
	(322.01,205.70) --
	(331.25,205.70) --
	(340.49,205.70);
\definecolor{drawColor}{RGB}{86,180,233}

\path[draw=drawColor,line width= 1.1pt,dash pattern=on 1pt off 3pt ,line join=round] ( 63.27, 87.55) --
	( 72.51, 95.59) --
	( 81.75,103.61) --
	( 90.99,111.61) --
	(100.23,119.58) --
	(109.48,127.53) --
	(118.72,135.45) --
	(127.96,143.35) --
	(137.20,151.22) --
	(146.44,159.07) --
	(155.68,166.90) --
	(164.92,174.71) --
	(174.16,182.49) --
	(183.40,190.25) --
	(192.64,197.99) --
	(201.88,205.70) --
	(211.12,213.39) --
	(220.36,221.06) --
	(229.60,228.71) --
	(238.84,236.33) --
	(248.08,243.93) --
	(257.32,251.51) --
	(266.56,259.07) --
	(275.81,266.61) --
	(285.05,274.12) --
	(294.29,281.62) --
	(303.53,289.09) --
	(312.77,296.54) --
	(322.01,303.97) --
	(331.25,311.38) --
	(340.49,318.77);
\definecolor{drawColor}{gray}{0.20}

\path[draw=drawColor,line width= 0.7pt,line join=round,line cap=round] ( 49.41, 39.63) rectangle (354.35,332.06);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
\definecolor{drawColor}{gray}{0.30}

\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.12] at ( 43.11,109.07) {1.50};

\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.12] at ( 43.11,190.12) {1.75};

\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.12] at ( 43.11,260.33) {2.00};

\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.12] at ( 43.11,322.26) {2.25};
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
\definecolor{drawColor}{gray}{0.20}

\path[draw=drawColor,line width= 0.7pt,line join=round] ( 45.91,112.92) --
	( 49.41,112.92);

\path[draw=drawColor,line width= 0.7pt,line join=round] ( 45.91,193.98) --
	( 49.41,193.98);

\path[draw=drawColor,line width= 0.7pt,line join=round] ( 45.91,264.19) --
	( 49.41,264.19);

\path[draw=drawColor,line width= 0.7pt,line join=round] ( 45.91,326.12) --
	( 49.41,326.12);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
\definecolor{drawColor}{gray}{0.20}

\path[draw=drawColor,line width= 0.7pt,line join=round] ( 63.27, 36.13) --
	( 63.27, 39.63);

\path[draw=drawColor,line width= 0.7pt,line join=round] (132.58, 36.13) --
	(132.58, 39.63);

\path[draw=drawColor,line width= 0.7pt,line join=round] (201.88, 36.13) --
	(201.88, 39.63);

\path[draw=drawColor,line width= 0.7pt,line join=round] (271.18, 36.13) --
	(271.18, 39.63);

\path[draw=drawColor,line width= 0.7pt,line join=round] (340.49, 36.13) --
	(340.49, 39.63);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
\definecolor{drawColor}{gray}{0.30}

\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.12] at ( 63.27, 25.61) {0.6};

\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.12] at (132.58, 25.61) {0.8};

\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.12] at (201.88, 25.61) {1.0};

\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.12] at (271.18, 25.61) {1.2};

\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.12] at (340.49, 25.61) {1.4};
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.40] at (201.88, 10.06) {Generation interval ratio, $\bar{G}_{\mathrm{var}}/\bar{G}_{\mathrm{wt}}$};
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\node[text=drawColor,rotate= 90.00,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.40] at ( 16.64,185.84) {Strength};
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
\definecolor{fillColor}{RGB}{255,255,255}

\path[fill=fillColor] ( 54.94,251.68) rectangle (226.84,324.71);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
\definecolor{fillColor}{RGB}{255,255,255}

\path[fill=fillColor] ( 61.94,293.37) rectangle ( 79.29,310.71);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
\definecolor{drawColor}{RGB}{86,180,233}

\path[draw=drawColor,line width= 1.1pt,dash pattern=on 1pt off 3pt ,line join=round] ( 63.68,302.04) -- ( 77.55,302.04);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
\definecolor{drawColor}{RGB}{86,180,233}

\path[draw=drawColor,line width= 1.1pt,dash pattern=on 1pt off 3pt ,line join=round] ( 63.68,302.04) -- ( 77.55,302.04);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
\definecolor{drawColor}{RGB}{86,180,233}

\path[draw=drawColor,line width= 1.1pt,dash pattern=on 1pt off 3pt ,line join=round] ( 63.68,302.04) -- ( 77.55,302.04);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
\definecolor{fillColor}{RGB}{255,255,255}

\path[fill=fillColor] ( 61.94,276.02) rectangle ( 79.29,293.37);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
\definecolor{drawColor}{RGB}{230,159,0}

\path[draw=drawColor,line width= 1.1pt,dash pattern=on 4pt off 4pt ,line join=round] ( 63.68,284.69) -- ( 77.55,284.69);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
\definecolor{drawColor}{RGB}{230,159,0}

\path[draw=drawColor,line width= 1.1pt,dash pattern=on 4pt off 4pt ,line join=round] ( 63.68,284.69) -- ( 77.55,284.69);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
\definecolor{drawColor}{RGB}{230,159,0}

\path[draw=drawColor,line width= 1.1pt,dash pattern=on 4pt off 4pt ,line join=round] ( 63.68,284.69) -- ( 77.55,284.69);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
\definecolor{fillColor}{RGB}{255,255,255}

\path[fill=fillColor] ( 61.94,258.68) rectangle ( 79.29,276.02);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\path[draw=drawColor,line width= 1.1pt,line join=round] ( 63.68,267.35) -- ( 77.55,267.35);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\path[draw=drawColor,line width= 1.1pt,line join=round] ( 63.68,267.35) -- ( 77.55,267.35);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\path[draw=drawColor,line width= 1.1pt,line join=round] ( 63.68,267.35) -- ( 77.55,267.35);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\node[text=drawColor,anchor=base west,inner sep=0pt, outer sep=0pt, scale=  1.12] at ( 86.29,298.18) {Constant speed intervention};
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\node[text=drawColor,anchor=base west,inner sep=0pt, outer sep=0pt, scale=  1.12] at ( 86.29,280.84) {Constant strength intervention};
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\node[text=drawColor,anchor=base west,inner sep=0pt, outer sep=0pt, scale=  1.12] at ( 86.29,263.49) {Epidemic};
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\node[text=drawColor,anchor=base west,inner sep=0pt, outer sep=0pt, scale=  1.68] at ( 49.41,342.78) {A};
\end{scope}
\begin{scope}
\path[clip] (361.35,  0.00) rectangle (722.70,361.35);
\definecolor{drawColor}{RGB}{255,255,255}
\definecolor{fillColor}{RGB}{255,255,255}

\path[draw=drawColor,line width= 0.7pt,line join=round,line cap=round,fill=fillColor] (361.35,  0.00) rectangle (722.70,361.35);
\end{scope}
\begin{scope}
\path[clip] (416.36, 39.63) rectangle (715.70,332.06);
\definecolor{fillColor}{RGB}{255,255,255}

\path[fill=fillColor] (416.36, 39.63) rectangle (715.70,332.06);
\definecolor{drawColor}{RGB}{0,0,0}

\path[draw=drawColor,line width= 1.1pt,line join=round] (429.97, 79.69) --
	(439.04, 79.69) --
	(448.11, 79.69) --
	(457.18, 79.69) --
	(466.25, 79.69) --
	(475.32, 79.69) --
	(484.39, 79.69) --
	(493.46, 79.69) --
	(502.53, 79.69) --
	(511.60, 79.69) --
	(520.68, 79.69) --
	(529.75, 79.69) --
	(538.82, 79.69) --
	(547.89, 79.69) --
	(556.96, 79.69) --
	(566.03, 79.69) --
	(575.10, 79.69) --
	(584.17, 79.69) --
	(593.24, 79.69) --
	(602.31, 79.69) --
	(611.38, 79.69) --
	(620.46, 79.69) --
	(629.53, 79.69) --
	(638.60, 79.69) --
	(647.67, 79.69) --
	(656.74, 79.69) --
	(665.81, 79.69) --
	(674.88, 79.69) --
	(683.95, 79.69) --
	(693.02, 79.69) --
	(702.09, 79.69);
\definecolor{drawColor}{RGB}{230,159,0}

\path[draw=drawColor,line width= 1.1pt,dash pattern=on 4pt off 4pt ,line join=round] (429.97,318.77) --
	(439.04,298.97) --
	(448.11,280.79) --
	(457.18,264.03) --
	(466.25,248.54) --
	(475.32,234.18) --
	(484.39,220.82) --
	(493.46,208.37) --
	(502.53,196.74) --
	(511.60,185.84) --
	(520.68,175.62) --
	(529.75,166.00) --
	(538.82,156.95) --
	(547.89,148.40) --
	(556.96,140.32) --
	(566.03,132.67) --
	(575.10,125.42) --
	(584.17,118.54) --
	(593.24,112.00) --
	(602.31,105.77) --
	(611.38, 99.83) --
	(620.46, 94.17) --
	(629.53, 88.76) --
	(638.60, 83.59) --
	(647.67, 78.65) --
	(656.74, 73.91) --
	(665.81, 69.36) --
	(674.88, 65.00) --
	(683.95, 60.82) --
	(693.02, 56.79) --
	(702.09, 52.92);
\definecolor{drawColor}{RGB}{86,180,233}

\path[draw=drawColor,line width= 1.1pt,dash pattern=on 1pt off 3pt ,line join=round] (429.97,139.21) --
	(439.04,139.21) --
	(448.11,139.21) --
	(457.18,139.21) --
	(466.25,139.21) --
	(475.32,139.21) --
	(484.39,139.21) --
	(493.46,139.21) --
	(502.53,139.21) --
	(511.60,139.21) --
	(520.68,139.21) --
	(529.75,139.21) --
	(538.82,139.21) --
	(547.89,139.21) --
	(556.96,139.21) --
	(566.03,139.21) --
	(575.10,139.21) --
	(584.17,139.21) --
	(593.24,139.21) --
	(602.31,139.21) --
	(611.38,139.21) --
	(620.46,139.21) --
	(629.53,139.21) --
	(638.60,139.21) --
	(647.67,139.21) --
	(656.74,139.21) --
	(665.81,139.21) --
	(674.88,139.21) --
	(683.95,139.21) --
	(693.02,139.21) --
	(702.09,139.21);
\definecolor{drawColor}{gray}{0.20}

\path[draw=drawColor,line width= 0.7pt,line join=round,line cap=round] (416.36, 39.63) rectangle (715.70,332.06);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
\definecolor{drawColor}{gray}{0.30}

\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.12] at (410.06, 75.84) {0.100};

\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.12] at (410.06,139.36) {0.125};

\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.12] at (410.06,202.88) {0.150};

\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.12] at (410.06,266.39) {0.175};
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
\definecolor{drawColor}{gray}{0.20}

\path[draw=drawColor,line width= 0.7pt,line join=round] (412.86, 79.69) --
	(416.36, 79.69);

\path[draw=drawColor,line width= 0.7pt,line join=round] (412.86,143.21) --
	(416.36,143.21);

\path[draw=drawColor,line width= 0.7pt,line join=round] (412.86,206.73) --
	(416.36,206.73);

\path[draw=drawColor,line width= 0.7pt,line join=round] (412.86,270.25) --
	(416.36,270.25);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
\definecolor{drawColor}{gray}{0.20}

\path[draw=drawColor,line width= 0.7pt,line join=round] (429.97, 36.13) --
	(429.97, 39.63);

\path[draw=drawColor,line width= 0.7pt,line join=round] (498.00, 36.13) --
	(498.00, 39.63);

\path[draw=drawColor,line width= 0.7pt,line join=round] (566.03, 36.13) --
	(566.03, 39.63);

\path[draw=drawColor,line width= 0.7pt,line join=round] (634.06, 36.13) --
	(634.06, 39.63);

\path[draw=drawColor,line width= 0.7pt,line join=round] (702.09, 36.13) --
	(702.09, 39.63);
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
\definecolor{drawColor}{gray}{0.30}

\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.12] at (429.97, 25.61) {0.6};

\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.12] at (498.00, 25.61) {0.8};

\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.12] at (566.03, 25.61) {1.0};

\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.12] at (634.06, 25.61) {1.2};

\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.12] at (702.09, 25.61) {1.4};
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.40] at (566.03, 10.06) {Generation interval ratio, $\bar{G}_{\mathrm{var}}/\bar{G}_{\mathrm{wt}}$};
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\node[text=drawColor,rotate= 90.00,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.40] at (377.99,185.84) {Speed (1/days)};
\end{scope}
\begin{scope}
\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
\definecolor{drawColor}{RGB}{0,0,0}

\node[text=drawColor,anchor=base west,inner sep=0pt, outer sep=0pt, scale=  1.68] at (416.36,342.78) {B};
\end{scope}
\end{tikzpicture}

\end{document}

### 2021-03-28 newvariant.tex
working on control

@@
 
-\section{Inferring relative strength from case data, in practice}
+\section{Implications for intervention strategies}
 
-Seb Funk's analysis here.
+While the relative speed need not stay constant over the course of an epidemic, at any given point in epidemic, we can measure the speed of the new variant and the wild type and ask how much more intervention is required to control the spread of both strains (relative to measured conditions).
+As a baseline scenario, we assume $\ry{wt}=0\,\textrm{/days}$ and $\delta=0.1\,\textrm{/days}$ (and therfore $\ry{var}=0\,\textrm{/days}$).
+In this case, the wild type is already under control, whereas the variant requires additional intervention measures to stop growing.
 
-\section{Intervention}
+Here, we consider two types of interventions:
+a strength-like intervention which affects the kernel by a constant amount $\theta$ regardless of when they were infected ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau)/\theta$) and a speed-like intervention which affect the kernel at a constant rate $\phi$ ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau) \exp(-\phi \tau)$).
+In this case, it is clear that we can control the spread of the variant when either $\theta > \Ry{var}$ or $\phi > \ry{var}$---these two quantities ($\theta$ and $\phi$) are also referred to as strength and speed of intervention, respectively.
+In general, we can calculate the strength ($\theta = \Ry{pre}/\Ry{post}$) and speed ($\phi = \ry{pre} - \ry{post}$) of intervention, both providing equivalent measures for effectiveness of intervention, for any intervention by comparing epidemiological conditions before and after the intervention.
+We compare how potential differences in generation-inteval distributions affect the control of the spread of the new variant using this strength-and-speed framework.
+\swp{missing citation.}
 
-If $K(\tau)/L(\tau)$.
+\begin{figure}
+\includegraphics[width=\textwidth]{control.pdf}
+\caption{
+\textbf{caption.}
+}
+\end{figure}
 

### 2021-03-28 control.R

@@ -1,0 +1,99 @@
+library(dplyr)
+library(ggplot2); theme_set(theme_bw(base_size=14, base_family = "Times"))
+library(ggthemes)
+library(egg)
+library(tikzDevice)
+
+delta <- 0.1
+
+rw <- 0
+
+kappa <- 0.2
+
+Gw <- 5
+
+Gratio <- seq(0.6, 1.4, length.out=31)
+
+Rv_base <- ((1 + kappa * (rw + delta) * Gw)^(1/kappa))
+target <- 0.9
+
+constant_strength <- Rv_base/target
+
+constant_speed <- optim(0.1,
+                        function(r) {
+                          (integrate(function(y) Rv_base * dgamma(y, shape=1/kappa, rate=1/kappa/Gv) * exp(-r * y),
+                                     0,
+                                     Inf)[[1]] - target)^2
+                        },
+                        lower=0,
+                        upper=0.2,
+                        method="Brent")[[1]]
+
+
+
+intervention <- sapply(Gratio, function(x) {
+  Gr <- x
+  Gv <- Gw * Gr
+  
+  Rv <- ((1 + kappa * (rw + delta) * Gv)^(1/kappa))
+  
+  rv <- rw + delta
+  
+  strength_strength <- constant_strength
+  strength_speed <- (Rv^kappa-1)/kappa/Gv - ((Rv/constant_strength)^kappa-1)/kappa/Gv
+  
+  speed_speed <- constant_speed 
+  
+  speed_strength <- Rv/integrate(function(y) Rv * dgamma(y, shape=1/kappa, rate=1/kappa/Gv) * exp(-constant_speed * y),
+                                 0,
+                                 Inf)[[1]]
+  
+  data.frame(
+    Gratio=Gr,
+    rw=rw,
+    Rv=Rv,
+    rv=rv,
+    strength_strength=strength_strength,
+    strength_speed=strength_speed,
+    speed_strength=speed_strength,
+    speed_speed=speed_speed
+  )
+}, simplify = FALSE) %>%
+  bind_rows
+
+g1 <- ggplot(intervention) +
+  geom_line(aes(Gratio, Rv, col="Epidemic", lty="Epidemic"), lwd=1) +
+  geom_line(aes(Gratio, strength_strength, col="Constant strength intervention", lty="Constant strength intervention"), lwd=1) +
[…]

### 2021-03-28 newvariant.tex
control

@@
 In this case, it is clear that we can control the spread of the variant when either $\theta > \Ry{var}$ or $\phi > \ry{var}$---these two quantities ($\theta$ and $\phi$) are also referred to as strength and speed of intervention, respectively.
-In general, we can calculate the strength ($\theta = \Ry{pre}/\Ry{post}$) and speed ($\phi = \ry{pre} - \ry{post}$) of intervention, both providing equivalent measures for effectiveness of intervention, for any intervention by comparing epidemiological conditions before and after the intervention.
+In general, we can calculate the strength ($\theta = \Ry{pre}/\Ry{post}$) and speed ($\phi = \ry{pre} - \ry{post}$) of intervention, both providing equivalent measures for effectiveness of intervention, for any intervention by comparing epidemiological conditions before and after the intervention \citep{doi:10.1098/rspb.2020.1556}---these quantities, by definition, must be greater than pre-intervention strength and speed of the epidemic to control the spread.
 We compare how potential differences in generation-inteval distributions affect the control of the spread of the new variant using this strength-and-speed framework.
-\swp{missing citation.}
 
-\begin{figure}
+\begin{figure}[!th]
 \includegraphics[width=\textwidth]{control.pdf}
 \caption{
-\textbf{caption.}
+\textbf{Strength and speed of intervention against the new variant.}
+(A) Strength of epidemic (black, solid), constant strength intervention (orange, dashed), and constant speed intervention (blue, dotted).
+(B) Speed of epidemic (black, solid), constant strength intervention (orange, dashed), and constant speed intervention (blue, dotted).
+Strength and speed are calculated assuming $\ry{wt}=0\,\textrm{/days}$, $\delta=0.1\,\textrm{/days}$, $\ry{var}=0\,\textrm{/days}$, $\Gy{wt}=5\,\textrm{days}$, and $\kappa=1/5$.
+Constant strength and speed interventions are modeled such that post intervention strength of the new variant, assuming equal mean generation intervals at 5 days, corresponds to 0.9.
 }
+\label{fig:strengthspeed}
 \end{figure}
+
+\fref{strengthspeed} compares the effectiveness of strength-based and speed-based interventions across a wide range of assumptions about generation-interval distributions.
+We begin by assuming that the variant and the wild type have equal mean generation intervals and model interventions such that they both reduce $\Ry{var}$ to 0.9 after intervention (\fref{strengthspeed}A).
+If the mean generation interval of the variant is longer than that of the wild type, then the strength of the variant will be higher;
+since the strength of the strength-like intervention is insensitive to the underlying generation-interval distribution, we may fail to control the spread of the new variant.
+The strength of the speed-like intervention, however, depends on the new variant:
+since speed-like interventions, such as contact tracing, have disproportionate effects on preventing later transmission, they are more effective (therefore stronger) when the mean generation interval of the new variant is longer.
+
+The speed-based paradigm gives same results about the control but provides important insights (\fref{strengthspeed}B).
+The observed speed of the variant at any given moment does not depend on our estimates of its mean generation interval.
+Therefore, if we can prevent infections at a faster rate than the observed speed of spread (i.e., if the speed of intervention is faster than the epidemic speed), we can control the epidemic---this explains why a constant-speed intervention can prevent the spread of the variant regardless of its mean generation interval.
+On the other hand, differences in mean generation intervals between the variant and the wild type imply different post-intervention epidemic speed (therefore, intervention speed) under a constant-strength intervention, explaining why it might not be effective if the variant has longer generation intervals.
 
@@
 Here, we explored how the generation-interval distribution shapes the link between relative strength and speed of the new variant.
-We define relative strength as the ratio between reproduction numbers ($\rho=\RR_v/\RR_w$) and relative speed as the difference between growth rates ($\delta=r_v-r_w$) of the variant and the wild type.
+We define relative strength as the ratio between reproduction numbers ($\rho=\Ry{var}/\Ry{wt}$) and relative speed as the difference between growth rates ($\delta=\ry{var}-\ry{wt}$) of the variant and the wild type.
 When the generation-interval distributions are assumed to be identical, we find that estimates of $\hat{\rho}$ decreases with $r_w$ whereas $\hat{\delta}$ increase with $\RR_w$.
 When we account for possible differences in generation-interval distributions, these relationships can change directions and further lead to biases in these estimates.
+We also show that differences in generation intervals lead to different conclusions about the effectiveness of interventions:
+if the variant has longer or shorter generation intervals than the wild type, we should prioritize speed- or strength-like interventions, respectively.
 
@@
 Early in the spread, we typically want to fix the relative speed because we can directly observe it.
-As more information about the transmission and immunity profiles of the new variant becomes available, it is better to fix the relative strength as it better matches biological mechanisms of its spread.
+As more information about the transmission and immunity profiles of the new variant becomes available, it is better to fix the relative strength as it better matches biological mechanisms for its higher strength.
 We suggest researchers to be mindful about which quantity they hold constant and how their assumptions lead to their conclusions.

### 2021-03-28 newvariant.tex
typos

@@
 The mean generation interval $\bar{G}_w$ of the wild type is assumed to equal 5 days \citep{ferretti2020quantifying}.
-The squared coefficient of variation $\kappa$ in generation intervals os assumed to equal 0.2 \citep{ferretti2020quantifying}---for simplicity, we assume that both the variant and the wild type have equal $\kappa$ and only consider differences in the mean.
+The squared coefficient of variation $\kappa$ in generation intervals is assumed to equal 0.2 \citep{ferretti2020quantifying}---for simplicity, we assume that both the variant and the wild type have equal $\kappa$ and only consider differences in the mean.
 As we expect there to be greater uncertainty in $\kappa$ than in $\bar{G}$, we evaluate the estimates of relative strength $\rho$ across a wide range of $\kappa$, ranging from 0 (delta distribution) to 1 (exponential distribution).
@@
 (B, E, H) True (purple, solid) and estimated (orange, dashed) ratios between reproduction numbers of the new variant and the wild type over time.
-(C, F, I) True (purple, solid) and estimated (orange, dashed) relationships between estimated reproduction numbers on a log-log sacle.
+(C, F, I) True (purple, solid) and estimated (orange, dashed) relationships between estimated reproduction numbers on a log-log scale.
 Top row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type and the variant.
-Middle row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type but is longer than that of the varant (4 days).
-Bottom row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type but is longer than that of the varant (6 days).
+Middle row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type but is longer than that of the variant (4 days).
+Bottom row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type but is longer than that of the variant (6 days).
 For all simulations, squared coefficient of variation in generation intervals is assumed to equal $\kappa = 1/5$.
@@
 While the relative speed need not stay constant over the course of an epidemic, at any given point in epidemic, we can measure the speed of the new variant and the wild type and ask how much more intervention is required to control the spread of both strains (relative to measured conditions).
-As a baseline scenario, we assume $\ry{wt}=0\,\textrm{/days}$ and $\delta=0.1\,\textrm{/days}$ (and therfore $\ry{var}=0\,\textrm{/days}$).
+As a baseline scenario, we assume $\ry{wt}=0\,\textrm{/days}$ and $\delta=0.1\,\textrm{/days}$ (and therefore $\ry{var}=0\,\textrm{/days}$).
 In this case, the wild type is already under control, whereas the variant requires additional intervention measures to stop growing.
@@
 In general, we can calculate the strength ($\theta = \Ry{pre}/\Ry{post}$) and speed ($\phi = \ry{pre} - \ry{post}$) of intervention, both providing equivalent measures for effectiveness of intervention, for any intervention by comparing epidemiological conditions before and after the intervention \citep{doi:10.1098/rspb.2020.1556}---these quantities, by definition, must be greater than pre-intervention strength and speed of the epidemic to control the spread.
-We compare how potential differences in generation-inteval distributions affect the control of the spread of the new variant using this strength-and-speed framework.
+We compare how potential differences in generation-interval distributions affect the control of the spread of the new variant using this strength-and-speed framework.
 
@@
 These same lessons have relevance for assessing the spread of variants in a partially susceptible population and improving efforts to control spread.
-Future studies should prioritize detailded assessment of the relative generation intervals of SARS-CoV-2 and widespread variants, as well as evaluate how uncertainty in assumptions about generational intervals might bias conclusions.
+Future studies should prioritize detailed assessment of the relative generation intervals of SARS-CoV-2 and widespread variants, as well as evaluate how uncertainty in assumptions about generational intervals might bias conclusions.
 

### 2021-03-28 control.R
more log scales

@@ -12,7 +12,7 @@ kappa <- 0.2
 
 Gw <- 5
 
-Gratio <- seq(0.6, 1.4, length.out=31)
+Gratio <- exp(seq(log(1/1.5), log(1.5), length.out=31))
 
 Rv_base <- ((1 + kappa * (rw + delta) * Gw)^(1/kappa))
 target <- 0.9
@@ -21,7 +21,7 @@ constant_strength <- Rv_base/target
 
 constant_speed <- optim(0.1,
                         function(r) {
-                          (integrate(function(y) Rv_base * dgamma(y, shape=1/kappa, rate=1/kappa/Gv) * exp(-r * y),
+                          (integrate(function(y) Rv_base * dgamma(y, shape=1/kappa, rate=1/kappa/Gw) * exp(-r * y),
                                      0,
                                      Inf)[[1]] - target)^2
                         },
@@ -65,7 +65,8 @@ g1 <- ggplot(intervention) +
   geom_line(aes(Gratio, Rv, col="Epidemic", lty="Epidemic"), lwd=1) +
   geom_line(aes(Gratio, strength_strength, col="Constant strength intervention", lty="Constant strength intervention"), lwd=1) +
   geom_line(aes(Gratio, speed_strength, col="Constant speed intervention", lty="Constant speed intervention"), lwd=1) +
-  scale_x_continuous("Generation interval ratio, $\\bar{G}_{\\mathrm{var}}/\\bar{G}_{\\mathrm{wt}}$") +
+  scale_x_log10("Generation interval ratio, $\\bar{G}_{\\mathrm{var}}/\\bar{G}_{\\mathrm{wt}}$",
+                breaks=c(0.67, 1, 1.5)) +
   scale_y_log10("Strength") +
   scale_linetype_manual("", values=3:1) +
   scale_color_manual("", values=rev(colorblind_pal()(3))) +
@@ -80,7 +81,8 @@ g2 <- ggplot(intervention) +
   geom_line(aes(Gratio, rv, col="Epidemic", lty="Epidemic"), lwd=1) +
   geom_line(aes(Gratio, strength_speed, col="Constant strength intervention", lty="Constant strength intervention"), lwd=1) +
   geom_line(aes(Gratio, speed_speed, col="Constant speed intervention", lty="Constant speed intervention"), lwd=1) +
[…]

### 2021-03-28 relstrength.R

@@ -15,7 +15,7 @@ pardata <- expand.grid(kappavec, rwvec)
 
 Gw <- 5
 
-Gratio <- seq(0.6, 1.4, length.out=31)
+Gratio <- exp(seq(log(1/1.5), log(1.5), length.out=31))
 
 pardata2 <- expand.grid(rwvec, Gratio)
 
@@ -79,7 +79,8 @@ g2 <- ggplot(rhodata2) +
   geom_hline(yintercept=-delta, lty=1, col="white") +
   annotate("text", x=1.04, y=0, label=c("$\\bar{G}_{\\mathrm{var}}=\\bar{G}_{\\mathrm{wt}}$"), col="white", angle=-90) +
   annotate("text", x=0.8, y=-delta+0.01, label=c("$r_{\\mathrm{var}}=0$"), col="white") +
-  scale_x_continuous("Generation interval ratio, $\\bar{G}_{\\mathrm{var}}/\\bar{G}_{\\mathrm{wt}}$", expand=c(0, 0)) +
+  scale_x_log10("Generation interval ratio, $\\bar{G}_{\\mathrm{var}}/\\bar{G}_{\\mathrm{wt}}$", expand=c(0, 0),
+                breaks=c(0.67, 1, 1.5)) +
   scale_y_continuous("Wild type speed, $r_{\\mathrm{wt}}$ (1/days)", expand=c(0, 0)) +
   scale_fill_viridis_c("$\\rho$") +
   ggtitle("B. Different generation intervals") +
@@ -93,7 +94,8 @@ g3 <- ggplot(rhodata2) +
   geom_hline(yintercept=-delta, lty=1, col="white") +
   annotate("text", x=1.04, y=0, label=c("$\\bar{G}_{\\mathrm{var}}=\\bar{G}_{\\mathrm{wt}}$"), col="white", angle=-90) +
   annotate("text", x=0.8, y=-delta+0.01, label=c("$r_{\\mathrm{var}}=0$"), col="white") +
-  scale_x_continuous("Generation interval ratio, $\\bar{G}_{\\mathrm{var}}/\\bar{G}_{\\mathrm{wt}}$", expand=c(0, 0)) +
+  scale_x_log10("Generation interval ratio, $\\bar{G}_{\\mathrm{var}}/\\bar{G}_{\\mathrm{wt}}$", expand=c(0, 0),
+                breaks=c(0.67, 1, 1.5)) +
[…]

### 2021-03-28 newvariant.tex
adding one more line in discussion

@@
 When we account for possible differences in generation-interval distributions, these relationships can change directions and further lead to biases in these estimates.
-We also show that differences in generation intervals lead to different conclusions about the effectiveness of interventions:
+We show how these biases might be assessed in practice.
+We also discuss that differences in generation intervals lead to different conclusions about the effectiveness of interventions:
 if the variant has longer or shorter generation intervals than the wild type, we should prioritize speed- or strength-like interventions, respectively.

### 2021-04-02 newvariant.tex
Trying to clarify early explanations

--- 2021-03-28 newvariant.tex
@@
 \newcommand{\ry}[1]{\rx{\mathrm{#1}}}
+\newcommand{\rw}{\ry{wt}}
 
@@
 \newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}
+\newcommand{\days}{\ensuremath{\, \textrm{day}}}
+\newcommand{\pday}{\ensuremath{/\textrm{day}}}
 \newcommand{\dd}[1]{\ensuremath{\, \mathrm{d}#1}}
@@
 
-Epidemic speed $r_x$ can be often estimated directly from case data \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}.
-However, estimating epidemic speed from case data can be challenging when case counts are low (e.g., in the beginning of the spread of the new variant) and is sensitive to temporal changes in testing patterns and intensity.
-On the other hand, studies of new SARS-CoV-2 variants have mostly focused on characterizing changes in \emph{proportion} of a new variant \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
-This has the advantage of being less sensitive to such changes in testing, and possibly for other transient effects that might affect variants and wild type viruses similarly.
+Epidemic speed $r_x$ can be often estimated directly from case data \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating};
+this can be challenging when case counts are low (as when a new variant begins to spread) and is sensitive to temporal changes in testing patterns and intensity.
+Studies of new SARS-CoV-2 variants have mostly focused on characterizing changes in \emph{proportion} of a new variant \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
+This has the advantage of being less sensitive to changes in testing and to other transient effects that might affect variants and wild type viruses similarly.
 When incidence is changing exponentially ($i_x(t) = i_x(t_0) \exp(r_x t)$), the proportion of the new variant $p(t)$ follows a logistic equation:
@@
 (A) Estimated relative strength of the new variant $\hat{\rho}$ given speed of the wild type
-$r_w$ assuming identical generation-interval distributions.
-(B) True relative strength of the new variant $\rho$ given speed of the wild type $r_w$ and ratio between the mean generation interval of the new variant $\bar{G}_v$ and that of the wild type $\bar{G}_w$.
+$\ry{wt}$ assuming identical generation-interval distributions.
+(B) True relative strength of the new variant $\rho$ given speed of the wild type $\ry{wt}$ and ratio between the mean generation interval of the new variant $\bar{G}_v$ and that of the wild type $\bar{G}_w$.
 (C) Ratio between the true and the estimated relative strength of the new variant $\rho/\hat{\rho}$.
-True relative strength of the new variant $\rho$ is calculated using $\delta=0.1\,\textrm{days}$, $\bar{G}_w = 5\,\textrm{days}$, and $\kappa = 1/5$.
+True relative strength of the new variant $\rho$ is calculated using $\delta=0.1\pday$, $\bar{G}_w = 5\days$, and $\kappa = 1/5$.
 }
@@
 
-As our baseline scenario, we assume $\delta = 0.1/\textrm{days}$ based on the observed relative speed of the new variant in the UK \citep{davies2021estimated}.
-This speed can be then used to estimate the relative strength $\rho = \Ry{var}/\Ry{wt}$ of the new variant with a caveat; such inference depends on assumptions about the generation-interval distribution.
-The mean generation interval $\bar{G}_w$ of the wild type is assumed to equal 5 days \citep{ferretti2020quantifying}.
-The squared coefficient of variation $\kappa$ in generation intervals os assumed to equal 0.2 \citep{ferretti2020quantifying}---for simplicity, we assume that both the variant and the wild type have equal $\kappa$ and only consider differences in the mean.
-As we expect there to be greater uncertainty in $\kappa$ than in $\bar{G}$, we evaluate the estimates of relative strength $\rho$ across a wide range of $\kappa$, ranging from 0 (delta distribution) to 1 (exponential distribution).
+We thus ask: what factors affect the relative strength $\rho = \Ry{var}/\Ry{wt}$ of a new variant, given an observed relative speed $\delta$?
+This inference will depend in general on assumptions about the growth rate of the wild type and on the generation-interval distributions of both variants.
+These have generally been assumed to be the same, but we will also consider the possibility that they are different.
 
-Previous studies have assumed that the generation-interval distributions do not differ between the wild type and the variant: $\Gy{var} = \Gy{wt} = \bar{G}$.
-In this case, the estimate of the relative strength is given by \citep{park2019practical}:
+As our baseline scenario, we take the relative speed of the variant to be $\delta = 0.1/\textrm{day}$ \citep{davies2021estimated}; the mean generation interval of the wild type to be $\bar{G}_w = 5\days$ \citep{ferretti2020quantifying}; and the squared coefficient of variation of generation intervals to be $\kappa=0.2$ \citep{ferretti2020quantifying}.
+For simplicity, we assume that the variant and the wild type have equal $\kappa$ throughout, and only consider differences in the mean.
+\jd{This is awkward; let's skip for now:} As we expect there to be greater uncertainty in $\kappa$ than in $\bar{G}$, we evaluate the estimates of relative strength $\rho$ across a wide range of $\kappa$, ranging from 0 (delta distribution) to 1 (exponential distribution).
+
+When the variant and wild type have the same generation-interval distributions, $\Gy{var} = \Gy{wt} = \bar{G}$, 
+the estimate of the relative strength is given by \citep{park2019practical}:
 \begin{equation}
 \hat{\rho} = \left(1 + \frac{\kappa \delta \bar{G}}{1 + \kappa \ry{wt} \bar{G}}\right)^{1/\kappa}.
+\label{eq:rhoGbar}
 \end{equation}
-Therefore, estimates of relative strength $\rho$ depends not only on the relative speed $\delta$ but also on how fast the wild type is spreading in the population $r_w$---a few analyses have implicitly or explicitly neglected this factor by either assuming $r_w = 0$ \citep{switzerland2021variant} or $\kappa = 0$ \citep{davies2021estimated}, in which case we get $\hat{\rho} = \exp(\delta \bar{G})$.
-Instead, if the wild type was spreading faster than previously assumed (higher $r_w$), the estimates of the relative strength $\rho$ will be go down (\fref{relstrength}A).
+Therefore, estimates of relative strength $\rho$ depend not only on the relative speed $\delta$ but also on how fast the wild type is spreading in the population \rw---a few analyses have implicitly or explicitly neglected this factor by either assuming $\rw = 0$ \citep{switzerland2021variant} or $\kappa = 0$ \citep{davies2021estimated} (in the latter case $\hat{\rho} = \exp(\delta \bar{G})$).
+
+\eref{rhoGbar} shows that, for realistic (non-zero) values of $\kappa$, estimates of the strength $\rho$ go down as estimates of the growth rate $\rw$ go up. 
+\jd{I like effective dispersion, but it does no good to just mention it once. Leave it out for now?}
 The relative strength also depends on the \emph{effective dispersion} in generation intervals (i.e., the amount of variability in generation intervals, as measured by $\kappa$)---using narrower distributions makes $\hat{\rho}$ less sensitive to $\ry{wt}$.
+These effects are illustrated in \fref{relstrength}A.
 
@@
 \begin{equation}
-\rho = \left(\frac{1 + \kappa (r_w + \delta) \bar{G}_v}{1 + \kappa r_w \bar{G}_w}\right)^{1/\kappa}.
+\rho = \left(\frac{1 + \kappa (\rw + \delta) \bar{G}_v}{1 + \kappa \rw \bar{G}_w}\right)^{1/\kappa}.
 \end{equation}
-In this case, longer (or shorter) mean generation interval of the new variant generally translates to higher (or lower) values of $\rho$---\fref{relstrength}B illustrates this effect assuming $\kappa = 1/5$.
-However, this direction can change when both the new variant and the wild type are decaying exponentially (see area below $r_v = 0$ line in \fref{relstrength}B)---when disease incidence is decaying exponentially, longer generation intervals lead to a \emph{lower} value of strength (and therefore lower relative strength $\rho$).
+In this case, longer (or shorter) mean generation interval of the new variant generally translates to higher (or lower) values of $\rho$---\fref{relstrength}B illustrates this effect.
+However, this direction can change when both the new variant and the wild type are decaying exponentially (see area below $r_v = 0$ line in \fref{relstrength}B)---for a given rate of decline, longer generation intervals lead to a \emph{lower} value of strength (and therefore lower relative strength $\rho$).
 
-Therefore, neglecting potential differences in generation intervals can lead to both under- and over-estimation of the relative strength.
-In \fref{relstrength}C, we quantify the degrees of bias using the ratio between the true value $\rho$, which accounts for differences in generation-interval distributions, and the estimated value $\hat{\rho}$, which assumes equal generation-interval distributions.
+Therefore, neglecting potential differences in generation intervals can lead to mis-estimation of the relative strength and the variant, with implications for forecasting and planning.
+In \fref{relstrength}C, we quantify the bias that arises by incorrectly assuming generation intervals are the same using the ratio between the true value $\rho$ and the estimated value $\hat{\rho}$, which assumes equal generation-interval distributions.
 Across the ranges of parameters that we consider, we find that up to 1.4 fold (either upward or downward) adjustments to current estimates may be required.
@@
 
-While relative speed is easier to measure, we often do not expect $\delta$ to be constant---this can depend on epidemic and demographic factors and can vary across time and space.
-Instead, biological mechanisms that cause new variant to have higher strength better translate to assuming a constant value of relative strength $\rho$.
+While relative speed is easier to measure, we do not generally expect $\delta$ to be constant if other factors governing epidemic spread are changing.
+Instead, many biological mechanisms better translate to assuming a constant value of relative strength $\rho$ over changing conditions.
 For example, if the new variant has a higher transmission rate, the ratio between the strength of the new variant and that of the wild type will remain constant ($=\rho$) under strength-like interventions that reduces the strength of both strains by a constant amount.
@@
 For now, we consider consequences of assuming a constant relative strength $\rho$ and predicting the relative speed $\delta$---an assumption often made (either implicitly or explicitly) in studies that have tried to predict the future spread of the new variant, including when it will replace the wild type.
-As our baseline scenario we assume $\rho = 1.61$, which is the value we obtain for $\delta=0.1\,\textrm{days}$, $\bar{G}_w = \bar{G}_v = 5\,\textrm{days}$, and $\kappa = 1/5$.
+As our baseline scenario we assume $\rho = 1.5$.
 
@@
 (C) Difference between the true and the estimated relative speed of the new variant $\delta - \hat{\delta}$.
-True relative speed of the new variant $\delta$ is calculated using $\rho=1.61$, $\bar{G}_w = 5\,\textrm{days}$, and $\kappa=1/5$.
+True relative speed of the new variant $\delta$ is calculated using $\rho=1.5$, $\bar{G}_w = 5\,\textrm{days}$, and $\kappa=1/5$.
 }
@@
 \end{equation}
-Under strength-like intervention measures that reduce transmission rates by a constant amount, we expect ratios between true reproduction numbers to remain constant and correspond to the true relative strength: $\Ry{var}(t)/\Ry{wt}(t) = \rho$.
-However, if the assumed generation-interval distribution $\hat{g}(\tau)$ is different from the true distribution, then the ratio between the estimated reproduction numbers $\hat{\rho}(t) = \hat{\RR}_{\textrm{var}}(t)/\hat{\RR}_{\textrm{wt}}(t)$ need not be constant, where
+Under strength-like intervention measures that reduce transmission rates by a constant amount, we expect ratios between true reproduction numbers to remain constant and correspond to the true relative strength: $\RR_v(t)/\RR_w(t) = \rho$.
+However, if the assumed generation-interval distribution $\hat{g}(\tau)$ is different from the true distribution, then the ratio between the estimated reproduction numbers $\hat{\rho}(t) = \hat{\RR}_v(t)/\hat{\RR}_w(t)$ need not be constant, where
 \begin{equation}
@@
 Here, we simulate three different scenarios using a two strain renewal equation that assumes perfect cross-immunity (\fref{Rtbias}): 
-(1) the wild type and the variant have the same generation-interval distributions, which match the assumed distribution (\fref{Rtbias}A--C);
-(2) the assumed distribution matches the generation-interval distribution of the wild type but has a longer mean than the mean generation interval of the variant (\fref{Rtbias}D--F); and
-(3) the assumed distribution matches the generation-interval distribution of the wild type but has a shorter mean than the mean generation interval of the variant (\fref{Rtbias}G--I).
-Then, we compare the estimated ratio $\hat{\rho}(t) = \hat{\RR}_{\textrm{var}}(t)/\hat{\RR}_{\textrm{wt}}(t)$ with the true ratio $\rho = \Ry{var}(t)/\Ry{wt}(t)$.
-In order to reflect smooth changes in $\RR$ driven by non-pharmaceutical interventions, we model $\Ry{wt}(t)$ phenomenologically and let $\Ry{var}(t) = \rho \Ry{wt}(t)$.
+(1) the wild type and the variant have the same generation-interval distributions, which match the assumed distribution (\fref{Rtbias}A--D);
+(2) the assumed distribution matches the generation-interval distribution of the wild type but has a longer mean than the mean generation interval of the variant (\fref{Rtbias}E--H); and
+(3) the assumed distribution matches the generation-interval distribution of the wild type but has a shorter mean than the mean generation interval of the variant (\fref{Rtbias}I--L).
+Then, we compare the estimated ratio $\hat{\rho}(t) = \hat{\RR}_v(t)/\hat{\RR}_w(t)$ with the true ratio $\rho = \RR_v(t)/\RR_w(t)$.
+For simplicity, we did not model non-pharmaceutical interventions;
+instead, we let epidemic dynamics to be determined by susceptible depletion alone.
 
@@
 \textbf{Estimates of relative strength over time under different scenarios.}
-(A, D, G) True (solid lines) and estimated (dashed lines) reproduction numbers of the new variant and the wild type over time.
-(B, E, H) True (purple, solid) and estimated (orange, dashed) ratios between reproduction numbers of the new variant and the wild type over time.
-(C, F, I) True (purple, solid) and estimated (orange, dashed) relationships between estimated reproduction numbers on a log-log sacle.
+(A, E, I) Daily incidence of infection caused by the new variant and the wild type.
+(B, F, J) True (solid lines) and estimated (dashed lines) reproduction numbers of the new variant and the wild type over time.
+(C, G, K) True (purple) and estimated (orange) ratios between reproduction numbers of the new variant and the wild type over time.
+(D, H, L) Relationship between estimated reproduction numbers on a log-log sacle.
+The dashed lines represent the one-to-one line. 
+The dotted lines represent the $\RR_v=\rho \RR_w$ line. 
 Top row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type and the variant.
-Middle row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type but is longer than that of the varant (4 days).
-Bottom row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type but is longer than that of the varant (6 days).
+Middle row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type but is longer than that of the varant (3.5 days).
+Bottom row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type but is longer than that of the varant (6.5 days).
 For all simulations, squared coefficient of variation in generation intervals is assumed to equal $\kappa = 1/5$.
-We assume $\Ry{wt}(t)=2.5$ before $t=15\,\textrm{days}$ to reflect unmitigated spread and let $\Ry{wt}(t)=2.5 (1+0.7\cos((t-15)/10))/1.7$ after to model smooth introduction and lifting of non-pharmaceutical interventions.
 }
@@
 
-When the assumed distribution matches the true distributions, the estimated reproduction numbers match the true values (\fref{Rtbias}A), and therefore the ratio remains constant (\fref{Rtbias}B).
-However, if the generation-interval distribution of the variant is different from the assumed distribution, the ratio changes over time (\fref{Rtbias}F,I).
-In particular, if the true generation intervals have a shorter (or longer) mean than the assumed distribution, we over-estimate (or under-estimate) $\Ry{var}(t)$ during the growth phase and under-estimate (or over-estimate) $ $ during the decay phase (\fref{Rtbias}D,G), which further translates to biases in the estimated relative strength (\fref{Rtbias}F,I).
+When the assumed distribution matches the true distributions, the estimated ratio remains constant (\fref{Rtbias}C).
+However, if the generation-interval distribution of the variant is different from the assumed distribution, the ratio changes over time (\fref{Rtbias}G,K).
+In particular, if the true generation intervals have a shorter (or longer) mean than the assumed distribution, we over-estimate (or under-estimate) $\RR_v(t)$ during the growth phase and under-estimate (or over-estimate) $\RR_v(t)$ during the decay phase (\fref{Rtbias}F,J), which further translates to biases in the estimated relative strength (\fref{Rtbias}G,K).
 
-We can assess these biases by plotting the estimated strength of the variant $\hat{\RR}_{\textrm{var}}(t)$ against the estimated strength of the wild type $\hat{\RR}_{\textrm{wt}}(t)$ on a log-log scale (\fref{Rtbias}C,F,I).
-When generation-interval distributions are correctly specified, we obtain a straight line with a slope of 1. 
-In this case, the intercept corresponds to the logarithm of the relative strength $\log(\rho)$ (\fref{Rtbias}C).
+We can also assess these biases by plotting the estimated strength of the variant $\hat{\RR}_v(t)$ against the estimated strength of the wild type $\hat{\RR}_w(t)$ on a log-log scale (\fref{Rtbias}D,H,L).
+When generation-interval distributions are correctly specified, we obtain a straight line parallel to the one-to-one line. 
+In this case, the slope corresponds to the logarithm of the relative strength $\log(\rho)$ (\fref{Rtbias}D).
 When the assumed mean generation interval is longer (or shorter) than the that of the variant, we over-estimate (or under-estimate) the slope (\fref{Rtbias}H,L) on a log-log plot.
-Smooth changes in $\RR$ further generate circular patterns but biases in slopes are clear.
 
-\section{Implications for intervention strategies}
+\section{Inferring relative strength from case data, in practice}
 
-While the relative speed need not stay constant over the course of an epidemic, at any given point in epidemic, we can measure the speed of the new variant and the wild type and ask how much more intervention is required to control the spread of both strains (relative to measured conditions).
-As a baseline scenario, we assume $\ry{wt}=0\,\textrm{/days}$ and $\delta=0.1\,\textrm{/days}$ (and therfore $\ry{var}=0\,\textrm{/days}$).
-In this case, the wild type is already under control, whereas the variant requires additional intervention measures to stop growing.
+Seb Funk's analysis here.
 
-Here, we consider two types of interventions:
-a strength-like intervention which affects the kernel by a constant amount $\theta$ regardless of when they were infected ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau)/\theta$) and a speed-like intervention which affect the kernel at a constant rate $\phi$ ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau) \exp(-\phi \tau)$).
-In this case, it is clear that we can control the spread of the variant when either $\theta > \Ry{var}$ or $\phi > \ry{var}$---these two quantities ($\theta$ and $\phi$) are also referred to as strength and speed of intervention, respectively.
-In general, we can calculate the strength ($\theta = \Ry{pre}/\Ry{post}$) and speed ($\phi = \ry{pre} - \ry{post}$) of intervention, both providing equivalent measures for effectiveness of intervention, for any intervention by comparing epidemiological conditions before and after the intervention \citep{doi:10.1098/rspb.2020.1556}---these quantities, by definition, must be greater than pre-intervention strength and speed of the epidemic to control the spread.
-We compare how potential differences in generation-inteval distributions affect the control of the spread of the new variant using this strength-and-speed framework.
+\section{Intervention}
 
-\begin{figure}[!th]
-\includegraphics[width=\textwidth]{control.pdf}
-\caption{
-\textbf{Strength and speed of intervention against the new variant.}
-(A) Strength of epidemic (black, solid), constant strength intervention (orange, dashed), and constant speed intervention (blue, dotted).
-(B) Speed of epidemic (black, solid), constant strength intervention (orange, dashed), and constant speed intervention (blue, dotted).
-Strength and speed are calculated assuming $\ry{wt}=0\,\textrm{/days}$, $\delta=0.1\,\textrm{/days}$, $\ry{var}=0\,\textrm{/days}$, $\Gy{wt}=5\,\textrm{days}$, and $\kappa=1/5$.
-Constant strength and speed interventions are modeled such that post intervention strength of the new variant, assuming equal mean generation intervals at 5 days, corresponds to 0.9.
-}
-\label{fig:strengthspeed}
-\end{figure}
-
-\fref{strengthspeed} compares the effectiveness of strength-based and speed-based interventions across a wide range of assumptions about generation-interval distributions.
-We begin by assuming that the variant and the wild type have equal mean generation intervals and model interventions such that they both reduce $\Ry{var}$ to 0.9 after intervention (\fref{strengthspeed}A).
-If the mean generation interval of the variant is longer than that of the wild type, then the strength of the variant will be higher;
-since the strength of the strength-like intervention is insensitive to the underlying generation-interval distribution, we may fail to control the spread of the new variant.
-The strength of the speed-like intervention, however, depends on the new variant:
-since speed-like interventions, such as contact tracing, have disproportionate effects on preventing later transmission, they are more effective (therefore stronger) when the mean generation interval of the new variant is longer.
-
-The speed-based paradigm gives same results about the control but provides important insights (\fref{strengthspeed}B).
-The observed speed of the variant at any given moment does not depend on our estimates of its mean generation interval.
-Therefore, if we can prevent infections at a faster rate than the observed speed of spread (i.e., if the speed of intervention is faster than the epidemic speed), we can control the epidemic---this explains why a constant-speed intervention can prevent the spread of the variant regardless of its mean generation interval.
-On the other hand, differences in mean generation intervals between the variant and the wild type imply different post-intervention epidemic speed (therefore, intervention speed) under a constant-strength intervention, explaining why it might not be effective if the variant has longer generation intervals.
+If $K(\tau)/L(\tau)$.
 
@@
 Here, we explored how the generation-interval distribution shapes the link between relative strength and speed of the new variant.
-We define relative strength as the ratio between reproduction numbers ($\rho=\Ry{var}/\Ry{wt}$) and relative speed as the difference between growth rates ($\delta=\ry{var}-\ry{wt}$) of the variant and the wild type.
-When the generation-interval distributions are assumed to be identical, we find that estimates of $\hat{\rho}$ decreases with $r_w$ whereas $\hat{\delta}$ increase with $\RR_w$.
+We define relative strength as the ratio between reproduction numbers ($\rho=\RR_v/\RR_w$) and relative speed as the difference between growth rates ($\delta=r_v-\rw$) of the variant and the wild type.
+When the generation-interval distributions are assumed to be identical, we find that estimates of $\hat{\rho}$ decreases with $\rw$ whereas $\hat{\delta}$ increase with $\RR_w$.
 When we account for possible differences in generation-interval distributions, these relationships can change directions and further lead to biases in these estimates.
-We also show that differences in generation intervals lead to different conclusions about the effectiveness of interventions:
-if the variant has longer or shorter generation intervals than the wild type, we should prioritize speed- or strength-like interventions, respectively.
 
@@
 Early in the spread, we typically want to fix the relative speed because we can directly observe it.
-As more information about the transmission and immunity profiles of the new variant becomes available, it is better to fix the relative strength as it better matches biological mechanisms for its higher strength.
+As more information about the transmission and immunity profiles of the new variant becomes available, it is better to fix the relative strength as it better matches biological mechanisms of its spread.
 We suggest researchers to be mindful about which quantity they hold constant and how their assumptions lead to their conclusions.

### 2021-04-02 control.tex

@@
-% Created by tikzDevice version 0.12.3 on 2021-03-27 20:28:52
+% Created by tikzDevice version 0.12.3 on 2021-03-28 18:34:28
 % !TEX encoding = UTF-8 Unicode
@@
 \begin{scope}
-\path[clip] (  0.00,  0.00) rectangle (361.35,361.35);
-\definecolor{drawColor}{RGB}{255,255,255}
-\definecolor{fillColor}{RGB}{255,255,255}
+\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
+\definecolor{drawColor}{gray}{0.30}
 
-\path[draw=drawColor,line width= 0.7pt,line join=round,line cap=round,fill=fillColor] (  0.00,  0.00) rectangle (361.35,361.35);
+\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.12] at ( 37.52, 56.08) {1.4};
+
+\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.12] at ( 37.52,123.48) {1.6};
+
+\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.12] at ( 37.52,182.94) {1.8};
+
+\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.12] at ( 37.52,236.12) {2.0};
+
+\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.12] at ( 37.52,284.23) {2.2};
+
+\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.12] at ( 37.52,328.15) {2.4};
 \end{scope}
 \begin{scope}
-\path[clip] ( 49.41, 39.63) rectangle (354.35,332.06);
-\definecolor{fillColor}{RGB}{255,255,255}
+\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
+\definecolor{drawColor}{gray}{0.20}
 
-\path[fill=fillColor] ( 49.41, 39.63) rectangle (354.35,332.06);
+\path[draw=drawColor,line width= 0.7pt,line join=round] ( 40.32, 59.94) --
+	( 43.82, 59.94);
+
+\path[draw=drawColor,line width= 0.7pt,line join=round] ( 40.32,127.34) --
+	( 43.82,127.34);
+
+\path[draw=drawColor,line width= 0.7pt,line join=round] ( 40.32,186.79) --
+	( 43.82,186.79);
+
+\path[draw=drawColor,line width= 0.7pt,line join=round] ( 40.32,239.98) --
+	( 43.82,239.98);
+
+\path[draw=drawColor,line width= 0.7pt,line join=round] ( 40.32,288.09) --
+	( 43.82,288.09);
+
+\path[draw=drawColor,line width= 0.7pt,line join=round] ( 40.32,332.01) --
+	( 43.82,332.01);
+\end{scope}
+\begin{scope}
+\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
 \definecolor{drawColor}{RGB}{0,0,0}
 
-\path[draw=drawColor,line width= 1.1pt,line join=round] ( 63.27, 52.92) --
-	( 72.51, 59.52) --
-	( 81.75, 66.11) --
-	( 90.99, 72.69) --
-	(100.23, 79.24) --
-	(109.48, 85.78) --
-	(118.72, 92.31) --
-	(127.96, 98.81) --
-	(137.20,105.30) --
-	(146.44,111.78) --
-	(155.68,118.24) --
-	(164.92,124.68) --
-	(174.16,131.11) --
-	(183.40,137.52) --
-	(192.64,143.92) --
-	(201.88,150.30) --
-	(211.12,156.67) --
-	(220.36,163.02) --
-	(229.60,169.35) --
-	(238.84,175.67) --
-	(248.08,181.98) --
-	(257.32,188.27) --
-	(266.56,194.54) --
-	(275.81,200.80) --
-	(285.05,207.04) --
-	(294.29,213.27) --
-	(303.53,219.49) --
-	(312.77,225.69) --
-	(322.01,231.88) --
-	(331.25,238.05) --
-	(340.49,244.20);
-\definecolor{drawColor}{RGB}{230,159,0}
-
-\path[draw=drawColor,line width= 1.1pt,dash pattern=on 4pt off 4pt ,line join=round] ( 63.27,205.70) --
-	( 72.51,205.70) --
-	( 81.75,205.70) --
-	( 90.99,205.70) --
-	(100.23,205.70) --
-	(109.48,205.70) --
-	(118.72,205.70) --
-	(127.96,205.70) --
-	(137.20,205.70) --
-	(146.44,205.70) --
-	(155.68,205.70) --
-	(164.92,205.70) --
-	(174.16,205.70) --
-	(183.40,205.70) --
-	(192.64,205.70) --
-	(201.88,205.70) --
-	(211.12,205.70) --
-	(220.36,205.70) --
-	(229.60,205.70) --
-	(238.84,205.70) --
-	(248.08,205.70) --
-	(257.32,205.70) --
-	(266.56,205.70) --
-	(275.81,205.70) --
-	(285.05,205.70) --
-	(294.29,205.70) --
-	(303.53,205.70) --
-	(312.77,205.70) --
-	(322.01,205.70) --
-	(331.25,205.70) --
-	(340.49,205.70);
-\definecolor{drawColor}{RGB}{86,180,233}
-
-\path[draw=drawColor,line width= 1.1pt,dash pattern=on 1pt off 3pt ,line join=round] ( 63.27, 87.55) --
-	( 72.51, 95.59) --
-	( 81.75,103.61) --
-	( 90.99,111.61) --
-	(100.23,119.58) --
-	(109.48,127.53) --
-	(118.72,135.45) --
-	(127.96,143.35) --
-	(137.20,151.22) --
-	(146.44,159.07) --
-	(155.68,166.90) --
-	(164.92,174.71) --
-	(174.16,182.49) --
-	(183.40,190.25) --
-	(192.64,197.99) --
-	(201.88,205.70) --
-	(211.12,213.39) --
-	(220.36,221.06) --
-	(229.60,228.71) --
-	(238.84,236.33) --
-	(248.08,243.93) --
-	(257.32,251.51) --
-	(266.56,259.07) --
-	(275.81,266.61) --
-	(285.05,274.12) --
-	(294.29,281.62) --
-	(303.53,289.09) --
-	(312.77,296.54) --
-	(322.01,303.97) --
-	(331.25,311.38) --
-	(340.49,318.77);
-\definecolor{drawColor}{gray}{0.20}
-
-\path[draw=drawColor,line width= 0.7pt,line join=round,line cap=round] ( 49.41, 39.63) rectangle (354.35,332.06);
+\node[text=drawColor,rotate= 90.00,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.40] at ( 16.64,185.89) {Strength};
 \end{scope}
@@
 
-\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.12] at ( 43.11,109.07) {1.50};
+\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.12] at (404.47, 94.91) {0.100};
 
-\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.12] at ( 43.11,190.12) {1.75};
+\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.12] at (404.47,167.50) {0.125};
 
-\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.12] at ( 43.11,260.33) {2.00};
+\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.12] at (404.47,240.08) {0.150};
 
-\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.12] at ( 43.11,322.26) {2.25};
+\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.12] at (404.47,312.67) {0.175};
 \end{scope}
@@
 
-\path[draw=drawColor,line width= 0.7pt,line join=round] ( 45.91,112.92) --
-	( 49.41,112.92);
+\path[draw=drawColor,line width= 0.7pt,line join=round] (407.27, 98.77) --
+	(410.77, 98.77);
 
-\path[draw=drawColor,line width= 0.7pt,line join=round] ( 45.91,193.98) --
-	( 49.41,193.98);
+\path[draw=drawColor,line width= 0.7pt,line join=round] (407.27,171.35) --
+	(410.77,171.35);
 
-\path[draw=drawColor,line width= 0.7pt,line join=round] ( 45.91,264.19) --
-	( 49.41,264.19);
+\path[draw=drawColor,line width= 0.7pt,line join=round] (407.27,243.94) --
+	(410.77,243.94);
 
-\path[draw=drawColor,line width= 0.7pt,line join=round] ( 45.91,326.12) --
-	( 49.41,326.12);
+\path[draw=drawColor,line width= 0.7pt,line join=round] (407.27,316.52) --
+	(410.77,316.52);
+\end{scope}
+\begin{scope}
+\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
+\definecolor{drawColor}{RGB}{0,0,0}
+
+\node[text=drawColor,rotate= 90.00,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.40] at (372.39,185.89) {Speed (1/days)};
+\end{scope}
+\begin{scope}
+\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
+\definecolor{drawColor}{RGB}{0,0,0}
+
+\node[text=drawColor,anchor=base west,inner sep=0pt, outer sep=0pt, scale=  1.68] at ( 43.82,342.78) {A};
+\end{scope}
+\begin{scope}
+\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
+\definecolor{drawColor}{RGB}{0,0,0}
+
+\node[text=drawColor,anchor=base west,inner sep=0pt, outer sep=0pt, scale=  1.68] at (410.77,342.78) {B};
+\end{scope}
+\begin{scope}
+\path[clip] ( 43.82, 39.69) rectangle (348.75,332.09);
+\definecolor{fillColor}{RGB}{255,255,255}
+
+\path[fill=fillColor] ( 43.82, 39.69) rectangle (348.75,332.09);
+\definecolor{drawColor}{RGB}{0,0,0}
+
+\path[draw=drawColor,line width= 1.1pt,line join=round] ( 57.68, 52.98) --
+	( 66.92, 57.30) --
+	( 76.16, 61.73) --
+	( 85.40, 66.27) --
+	( 94.64, 70.93) --
+	(103.89, 75.71) --
+	(113.13, 80.60) --
+	(122.37, 85.63) --
+	(131.61, 90.78) --
+	(140.85, 96.06) --
+	(150.09,101.47) --
+	(159.33,107.02) --
+	(168.57,112.71) --
+	(177.81,118.54) --
+	(187.05,124.52) --
+	(196.29,130.64) --
+	(205.53,136.92) --
+	(214.77,143.36) --
+	(224.01,149.95) --
+	(233.25,156.71) --
+	(242.49,163.63) --
+	(251.73,170.73) --
+	(260.97,177.99) --
+	(270.21,185.44) --
+	(279.45,193.06) --
+	(288.69,200.87) --
+	(297.93,208.87) --
+	(307.17,217.06) --
+	(316.41,225.45) --
+	(325.65,234.04) --
+	(334.89,242.84);
+\definecolor{drawColor}{RGB}{230,159,0}
+
+\path[draw=drawColor,line width= 1.1pt,dash pattern=on 4pt off 4pt ,line join=round] ( 57.68,183.83) --
+	( 66.92,183.83) --
+	( 76.16,183.83) --
+	( 85.40,183.83) --
+	( 94.64,183.83) --
+	(103.89,183.83) --
+	(113.13,183.83) --
+	(122.37,183.83) --
+	(131.61,183.83) --
+	(140.85,183.83) --
+	(150.09,183.83) --
+	(159.33,183.83) --
+	(168.57,183.83) --
+	(177.81,183.83) --
+	(187.05,183.83) --
+	(196.29,183.83) --
+	(205.53,183.83) --
+	(214.77,183.83) --
+	(224.01,183.83) --
+	(233.25,183.83) --
+	(242.49,183.83) --
+	(251.73,183.83) --
+	(260.97,183.83) --
+	(270.21,183.83) --
+	(279.45,183.83) --
+	(288.69,183.83) --
+	(297.93,183.83) --
+	(307.17,183.83) --
+	(316.41,183.83) --
+	(325.65,183.83) --
+	(334.89,183.83);
+\definecolor{drawColor}{RGB}{86,180,233}
+
+\path[draw=drawColor,line width= 1.1pt,dash pattern=on 1pt off 3pt ,line join=round] ( 57.68, 89.66) --
+	( 66.92, 94.92) --
+	( 76.16,100.30) --
+	( 85.40,105.82) --
+	( 94.64,111.48) --
+	(103.89,117.28) --
+	(113.13,123.23) --
+	(122.37,129.32) --
+	(131.61,135.57) --
+	(140.85,141.97) --
+	(150.09,148.53) --
+	(159.33,155.25) --
+	(168.57,162.14) --
+	(177.81,169.19) --
+	(187.05,176.42) --
+	(196.29,183.83) --
+	(205.53,191.42) --
+	(214.77,199.19) --
+	(224.01,207.14) --
+	(233.25,215.29) --
+	(242.49,223.64) --
+	(251.73,232.19) --
+	(260.97,240.94) --
+	(270.21,249.90) --
+	(279.45,259.07) --
+	(288.69,268.46) --
+	(297.93,278.07) --
+	(307.17,287.90) --
+	(316.41,297.96) --
+	(325.65,308.26) --
+	(334.89,318.80);
+\definecolor{drawColor}{gray}{0.20}
+
+\path[draw=drawColor,line width= 0.7pt,line join=round,line cap=round] ( 43.82, 39.69) rectangle (348.75,332.09);
+\end{scope}
+\begin{scope}
+\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
+\definecolor{fillColor}{RGB}{255,255,255}
+
+\path[fill=fillColor] ( 48.06,251.71) rectangle (222.54,324.75);
+\end{scope}
+\begin{scope}
+\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
+\definecolor{fillColor}{RGB}{255,255,255}
+
+\path[fill=fillColor] ( 55.06,293.40) rectangle ( 72.41,310.75);
+\end{scope}
+\begin{scope}
+\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
+\definecolor{drawColor}{RGB}{86,180,233}
+
+\path[draw=drawColor,line width= 1.1pt,dash pattern=on 1pt off 3pt ,line join=round] ( 56.80,302.08) -- ( 70.67,302.08);
+\end{scope}
+\begin{scope}
+\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
+\definecolor{drawColor}{RGB}{86,180,233}
+
+\path[draw=drawColor,line width= 1.1pt,dash pattern=on 1pt off 3pt ,line join=round] ( 56.80,302.08) -- ( 70.67,302.08);
+\end{scope}
+\begin{scope}
+\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
+\definecolor{drawColor}{RGB}{86,180,233}
+
+\path[draw=drawColor,line width= 1.1pt,dash pattern=on 1pt off 3pt ,line join=round] ( 56.80,302.08) -- ( 70.67,302.08);
+\end{scope}
+\begin{scope}
+\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
+\definecolor{fillColor}{RGB}{255,255,255}
+
+\path[fill=fillColor] ( 55.06,276.06) rectangle ( 72.41,293.40);
+\end{scope}
+\begin{scope}
+\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
+\definecolor{drawColor}{RGB}{230,159,0}
+
+\path[draw=drawColor,line width= 1.1pt,dash pattern=on 4pt off 4pt ,line join=round] ( 56.80,284.73) -- ( 70.67,284.73);
+\end{scope}
+\begin{scope}
+\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
+\definecolor{drawColor}{RGB}{230,159,0}
+
+\path[draw=drawColor,line width= 1.1pt,dash pattern=on 4pt off 4pt ,line join=round] ( 56.80,284.73) -- ( 70.67,284.73);
+\end{scope}
+\begin{scope}
+\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
+\definecolor{drawColor}{RGB}{230,159,0}
+
+\path[draw=drawColor,line width= 1.1pt,dash pattern=on 4pt off 4pt ,line join=round] ( 56.80,284.73) -- ( 70.67,284.73);
+\end{scope}
+\begin{scope}
+\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
+\definecolor{fillColor}{RGB}{255,255,255}
+
+\path[fill=fillColor] ( 55.06,258.71) rectangle ( 72.41,276.06);
+\end{scope}
+\begin{scope}
+\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
+\definecolor{drawColor}{RGB}{0,0,0}
+
+\path[draw=drawColor,line width= 1.1pt,line join=round] ( 56.80,267.39) -- ( 70.67,267.39);
+\end{scope}
+\begin{scope}
+\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
+\definecolor{drawColor}{RGB}{0,0,0}
+
+\path[draw=drawColor,line width= 1.1pt,line join=round] ( 56.80,267.39) -- ( 70.67,267.39);
+\end{scope}
+\begin{scope}
+\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
+\definecolor{drawColor}{RGB}{0,0,0}
+
+\path[draw=drawColor,line width= 1.1pt,line join=round] ( 56.80,267.39) -- ( 70.67,267.39);
+\end{scope}
+\begin{scope}
+\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
+\definecolor{drawColor}{RGB}{0,0,0}
+
+\node[text=drawColor,anchor=base west,inner sep=0pt, outer sep=0pt, scale=  1.12] at ( 79.41,298.22) {Constant speed intervention};
+\end{scope}
+\begin{scope}
+\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
+\definecolor{drawColor}{RGB}{0,0,0}
+
+\node[text=drawColor,anchor=base west,inner sep=0pt, outer sep=0pt, scale=  1.12] at ( 79.41,280.87) {Constant strength intervention};
+\end{scope}
+\begin{scope}
+\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
+\definecolor{drawColor}{RGB}{0,0,0}
+
+\node[text=drawColor,anchor=base west,inner sep=0pt, outer sep=0pt, scale=  1.12] at ( 79.41,263.53) {Epidemic};
+\end{scope}
+\begin{scope}
+\path[clip] (410.77, 39.69) rectangle (715.70,332.09);
+\definecolor{fillColor}{RGB}{255,255,255}
+
+\path[fill=fillColor] (410.77, 39.69) rectangle (715.70,332.09);
+\definecolor{drawColor}{RGB}{0,0,0}
+
+\path[draw=drawColor,line width= 1.1pt,line join=round] (424.63, 98.77) --
+	(433.87, 98.77) --
+	(443.11, 98.77) --
+	(452.35, 98.77) --
+	(461.59, 98.77) --
+	(470.83, 98.77) --
+	(480.07, 98.77) --
+	(489.31, 98.77) --
+	(498.55, 98.77) --
+	(507.79, 98.77) --
+	(517.03, 98.77) --
+	(526.28, 98.77) --
+	(535.52, 98.77) --
+	(544.76, 98.77) --
+	(554.00, 98.77) --
+	(563.24, 98.77) --
+	(572.48, 98.77) --
+	(581.72, 98.77) --
+	(590.96, 98.77) --
+	(600.20, 98.77) --
+	(609.44, 98.77) --
+	(618.68, 98.77) --
+	(627.92, 98.77) --
+	(637.16, 98.77) --
+	(646.40, 98.77) --
+	(655.64, 98.77) --
+	(664.88, 98.77) --
+	(674.12, 98.77) --
+	(683.36, 98.77) --
+	(692.60, 98.77) --
+	(701.84, 98.77);
+\definecolor{drawColor}{RGB}{230,159,0}
+
+\path[draw=drawColor,line width= 1.1pt,dash pattern=on 4pt off 4pt ,line join=round] (424.63,318.80) --
+	(433.87,306.04) --
+	(443.11,293.62) --
+	(452.35,281.53) --
+	(461.59,269.76) --
+	(470.83,258.31) --
+	(480.07,247.16) --
+	(489.31,236.31) --
+	(498.55,225.75) --
+	(507.79,215.47) --
+	(517.03,205.47) --
+	(526.28,195.73) --
+	(535.52,186.25) --
+	(544.76,177.03) --
+	(554.00,168.05) --
+	(563.24,159.31) --
+	(572.48,150.80) --
+	(581.72,142.52) --
+	(590.96,134.46) --
+	(600.20,126.62) --
+	(609.44,118.98) --
+	(618.68,111.55) --
+	(627.92,104.32) --
+	(637.16, 97.28) --
+	(646.40, 90.42) --
+	(655.64, 83.75) --
+	(664.88, 77.26) --
+	(674.12, 70.94) --
+	(683.36, 64.79) --
+	(692.60, 58.81) --
+	(701.84, 52.98);
+\definecolor{drawColor}{RGB}{86,180,233}
+
+\path[draw=drawColor,line width= 1.1pt,dash pattern=on 1pt off 3pt ,line join=round] (424.63,166.78) --
+	(433.87,166.78) --
+	(443.11,166.78) --
+	(452.35,166.78) --
+	(461.59,166.78) --
+	(470.83,166.78) --
+	(480.07,166.78) --
+	(489.31,166.78) --
+	(498.55,166.78) --
+	(507.79,166.78) --
+	(517.03,166.78) --
+	(526.28,166.78) --
+	(535.52,166.78) --
+	(544.76,166.78) --
+	(554.00,166.78) --
+	(563.24,166.78) --
+	(572.48,166.78) --
+	(581.72,166.78) --
+	(590.96,166.78) --
+	(600.20,166.78) --
+	(609.44,166.78) --
+	(618.68,166.78) --
+	(627.92,166.78) --
+	(637.16,166.78) --
+	(646.40,166.78) --
+	(655.64,166.78) --
+	(664.88,166.78) --
+	(674.12,166.78) --
+	(683.36,166.78) --
+	(692.60,166.78) --
+	(701.84,166.78);
+\definecolor{drawColor}{gray}{0.20}
+
+\path[draw=drawColor,line width= 0.7pt,line join=round,line cap=round] (410.77, 39.69) rectangle (715.70,332.09);
 \end{scope}
@@
 
-\path[draw=drawColor,line width= 0.7pt,line join=round] ( 63.27, 36.13) --
-	( 63.27, 39.63);
+\path[draw=drawColor,line width= 0.7pt,line join=round] ( 59.39, 36.19) --
+	( 59.39, 39.69);
 
-\path[draw=drawColor,line width= 0.7pt,line join=round] (132.58, 36.13) --
-	(132.58, 39.63);
+\path[draw=drawColor,line width= 0.7pt,line join=round] (196.29, 36.19) --
+	(196.29, 39.69);
 
-\path[draw=drawColor,line width= 0.7pt,line join=round] (201.88, 36.13) --
-	(201.88, 39.63);
-
-\path[draw=drawColor,line width= 0.7pt,line join=round] (271.18, 36.13) --
-	(271.18, 39.63);
-
-\path[draw=drawColor,line width= 0.7pt,line join=round] (340.49, 36.13) --
-	(340.49, 39.63);
+\path[draw=drawColor,line width= 0.7pt,line join=round] (334.89, 36.19) --
+	(334.89, 39.69);
 \end{scope}
@@
 
-\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.12] at ( 63.27, 25.61) {0.6};
+\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.12] at ( 59.39, 25.68) {0.67};
 
-\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.12] at (132.58, 25.61) {0.8};
+\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.12] at (196.29, 25.68) {1.00};
 
-\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.12] at (201.88, 25.61) {1.0};
-
-\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.12] at (271.18, 25.61) {1.2};
-
-\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.12] at (340.49, 25.61) {1.4};
+\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.12] at (334.89, 25.68) {1.50};
 \end{scope}
@@
 
-\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.40] at (201.88, 10.06) {Generation interval ratio, $\bar{G}_{\mathrm{var}}/\bar{G}_{\mathrm{wt}}$};
+\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.40] at (196.29, 10.07) {Generation interval ratio, $\bar{G}_{\mathrm{var}}/\bar{G}_{\mathrm{wt}}$};
+\end{scope}
+\begin{scope}
+\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
+\definecolor{drawColor}{gray}{0.20}
+
+\path[draw=drawColor,line width= 0.7pt,line join=round] (426.34, 36.19) --
+	(426.34, 39.69);
+
+\path[draw=drawColor,line width= 0.7pt,line join=round] (563.24, 36.19) --
+	(563.24, 39.69);
+
+\path[draw=drawColor,line width= 0.7pt,line join=round] (701.84, 36.19) --
+	(701.84, 39.69);
+\end{scope}
+\begin{scope}
+\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
+\definecolor{drawColor}{gray}{0.30}
+
+\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.12] at (426.34, 25.68) {0.67};
+
+\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.12] at (563.24, 25.68) {1.00};
+
+\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.12] at (701.84, 25.68) {1.50};
 \end{scope}
@@
 
-\node[text=drawColor,rotate= 90.00,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.40] at ( 16.64,185.84) {Strength};
-\end{scope}
-\begin{scope}
-\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
-\definecolor{fillColor}{RGB}{255,255,255}
-
-\path[fill=fillColor] ( 54.94,251.68) rectangle (226.84,324.71);
-\end{scope}
-\begin{scope}
-\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
-\definecolor{fillColor}{RGB}{255,255,255}
-
-\path[fill=fillColor] ( 61.94,293.37) rectangle ( 79.29,310.71);
-\end{scope}
-\begin{scope}
-\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
-\definecolor{drawColor}{RGB}{86,180,233}
-
-\path[draw=drawColor,line width= 1.1pt,dash pattern=on 1pt off 3pt ,line join=round] ( 63.68,302.04) -- ( 77.55,302.04);
-\end{scope}
-\begin{scope}
-\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
-\definecolor{drawColor}{RGB}{86,180,233}
-
-\path[draw=drawColor,line width= 1.1pt,dash pattern=on 1pt off 3pt ,line join=round] ( 63.68,302.04) -- ( 77.55,302.04);
-\end{scope}
-\begin{scope}
-\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
-\definecolor{drawColor}{RGB}{86,180,233}
-
-\path[draw=drawColor,line width= 1.1pt,dash pattern=on 1pt off 3pt ,line join=round] ( 63.68,302.04) -- ( 77.55,302.04);
-\end{scope}
-\begin{scope}
-\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
-\definecolor{fillColor}{RGB}{255,255,255}
-
-\path[fill=fillColor] ( 61.94,276.02) rectangle ( 79.29,293.37);
-\end{scope}
-\begin{scope}
-\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
-\definecolor{drawColor}{RGB}{230,159,0}
-
-\path[draw=drawColor,line width= 1.1pt,dash pattern=on 4pt off 4pt ,line join=round] ( 63.68,284.69) -- ( 77.55,284.69);
-\end{scope}
-\begin{scope}
-\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
-\definecolor{drawColor}{RGB}{230,159,0}
-
-\path[draw=drawColor,line width= 1.1pt,dash pattern=on 4pt off 4pt ,line join=round] ( 63.68,284.69) -- ( 77.55,284.69);
-\end{scope}
-\begin{scope}
-\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
-\definecolor{drawColor}{RGB}{230,159,0}
-
-\path[draw=drawColor,line width= 1.1pt,dash pattern=on 4pt off 4pt ,line join=round] ( 63.68,284.69) -- ( 77.55,284.69);
-\end{scope}
-\begin{scope}
-\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
-\definecolor{fillColor}{RGB}{255,255,255}
-
-\path[fill=fillColor] ( 61.94,258.68) rectangle ( 79.29,276.02);
-\end{scope}
-\begin{scope}
-\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
-\definecolor{drawColor}{RGB}{0,0,0}
-
-\path[draw=drawColor,line width= 1.1pt,line join=round] ( 63.68,267.35) -- ( 77.55,267.35);
-\end{scope}
-\begin{scope}
-\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
-\definecolor{drawColor}{RGB}{0,0,0}
-
-\path[draw=drawColor,line width= 1.1pt,line join=round] ( 63.68,267.35) -- ( 77.55,267.35);
-\end{scope}
-\begin{scope}
-\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
-\definecolor{drawColor}{RGB}{0,0,0}
-
-\path[draw=drawColor,line width= 1.1pt,line join=round] ( 63.68,267.35) -- ( 77.55,267.35);
-\end{scope}
-\begin{scope}
-\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
-\definecolor{drawColor}{RGB}{0,0,0}
-
-\node[text=drawColor,anchor=base west,inner sep=0pt, outer sep=0pt, scale=  1.12] at ( 86.29,298.18) {Constant speed intervention};
-\end{scope}
-\begin{scope}
-\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
-\definecolor{drawColor}{RGB}{0,0,0}
-
-\node[text=drawColor,anchor=base west,inner sep=0pt, outer sep=0pt, scale=  1.12] at ( 86.29,280.84) {Constant strength intervention};
-\end{scope}
-\begin{scope}
-\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
-\definecolor{drawColor}{RGB}{0,0,0}
-
-\node[text=drawColor,anchor=base west,inner sep=0pt, outer sep=0pt, scale=  1.12] at ( 86.29,263.49) {Epidemic};
-\end{scope}
-\begin{scope}
-\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
-\definecolor{drawColor}{RGB}{0,0,0}
-
-\node[text=drawColor,anchor=base west,inner sep=0pt, outer sep=0pt, scale=  1.68] at ( 49.41,342.78) {A};
-\end{scope}
-\begin{scope}
-\path[clip] (361.35,  0.00) rectangle (722.70,361.35);
-\definecolor{drawColor}{RGB}{255,255,255}
-\definecolor{fillColor}{RGB}{255,255,255}
-
-\path[draw=drawColor,line width= 0.7pt,line join=round,line cap=round,fill=fillColor] (361.35,  0.00) rectangle (722.70,361.35);
-\end{scope}
-\begin{scope}
-\path[clip] (416.36, 39.63) rectangle (715.70,332.06);
-\definecolor{fillColor}{RGB}{255,255,255}
-
-\path[fill=fillColor] (416.36, 39.63) rectangle (715.70,332.06);
-\definecolor{drawColor}{RGB}{0,0,0}
-
-\path[draw=drawColor,line width= 1.1pt,line join=round] (429.97, 79.69) --
-	(439.04, 79.69) --
-	(448.11, 79.69) --
-	(457.18, 79.69) --
-	(466.25, 79.69) --
-	(475.32, 79.69) --
-	(484.39, 79.69) --
-	(493.46, 79.69) --
-	(502.53, 79.69) --
-	(511.60, 79.69) --
-	(520.68, 79.69) --
-	(529.75, 79.69) --
-	(538.82, 79.69) --
-	(547.89, 79.69) --
-	(556.96, 79.69) --
-	(566.03, 79.69) --
-	(575.10, 79.69) --
-	(584.17, 79.69) --
-	(593.24, 79.69) --
-	(602.31, 79.69) --
-	(611.38, 79.69) --
-	(620.46, 79.69) --
-	(629.53, 79.69) --
-	(638.60, 79.69) --
-	(647.67, 79.69) --
-	(656.74, 79.69) --
-	(665.81, 79.69) --
-	(674.88, 79.69) --
-	(683.95, 79.69) --
-	(693.02, 79.69) --
-	(702.09, 79.69);
-\definecolor{drawColor}{RGB}{230,159,0}
-
-\path[draw=drawColor,line width= 1.1pt,dash pattern=on 4pt off 4pt ,line join=round] (429.97,318.77) --
-	(439.04,298.97) --
-	(448.11,280.79) --
-	(457.18,264.03) --
-	(466.25,248.54) --
-	(475.32,234.18) --
-	(484.39,220.82) --
-	(493.46,208.37) --
-	(502.53,196.74) --
-	(511.60,185.84) --
-	(520.68,175.62) --
-	(529.75,166.00) --
-	(538.82,156.95) --
-	(547.89,148.40) --
-	(556.96,140.32) --
-	(566.03,132.67) --
-	(575.10,125.42) --
-	(584.17,118.54) --
-	(593.24,112.00) --
-	(602.31,105.77) --
-	(611.38, 99.83) --
-	(620.46, 94.17) --
-	(629.53, 88.76) --
-	(638.60, 83.59) --
-	(647.67, 78.65) --
-	(656.74, 73.91) --
-	(665.81, 69.36) --
-	(674.88, 65.00) --
-	(683.95, 60.82) --
-	(693.02, 56.79) --
-	(702.09, 52.92);
-\definecolor{drawColor}{RGB}{86,180,233}
-
-\path[draw=drawColor,line width= 1.1pt,dash pattern=on 1pt off 3pt ,line join=round] (429.97,139.21) --
-	(439.04,139.21) --
-	(448.11,139.21) --
-	(457.18,139.21) --
-	(466.25,139.21) --
-	(475.32,139.21) --
-	(484.39,139.21) --
-	(493.46,139.21) --
-	(502.53,139.21) --
-	(511.60,139.21) --
-	(520.68,139.21) --
-	(529.75,139.21) --
-	(538.82,139.21) --
-	(547.89,139.21) --
-	(556.96,139.21) --
-	(566.03,139.21) --
-	(575.10,139.21) --
-	(584.17,139.21) --
-	(593.24,139.21) --
-	(602.31,139.21) --
-	(611.38,139.21) --
-	(620.46,139.21) --
-	(629.53,139.21) --
-	(638.60,139.21) --
-	(647.67,139.21) --
-	(656.74,139.21) --
-	(665.81,139.21) --
-	(674.88,139.21) --
-	(683.95,139.21) --
-	(693.02,139.21) --
-	(702.09,139.21);
-\definecolor{drawColor}{gray}{0.20}
-
-\path[draw=drawColor,line width= 0.7pt,line join=round,line cap=round] (416.36, 39.63) rectangle (715.70,332.06);
-\end{scope}
-\begin{scope}
-\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
-\definecolor{drawColor}{gray}{0.30}
-
-\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.12] at (410.06, 75.84) {0.100};
-
-\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.12] at (410.06,139.36) {0.125};
-
-\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.12] at (410.06,202.88) {0.150};
-
-\node[text=drawColor,anchor=base east,inner sep=0pt, outer sep=0pt, scale=  1.12] at (410.06,266.39) {0.175};
-\end{scope}
-\begin{scope}
-\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
-\definecolor{drawColor}{gray}{0.20}
-
-\path[draw=drawColor,line width= 0.7pt,line join=round] (412.86, 79.69) --
-	(416.36, 79.69);
-
-\path[draw=drawColor,line width= 0.7pt,line join=round] (412.86,143.21) --
-	(416.36,143.21);
-
-\path[draw=drawColor,line width= 0.7pt,line join=round] (412.86,206.73) --
-	(416.36,206.73);
-
-\path[draw=drawColor,line width= 0.7pt,line join=round] (412.86,270.25) --
-	(416.36,270.25);
-\end{scope}
-\begin{scope}
-\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
-\definecolor{drawColor}{gray}{0.20}
-
-\path[draw=drawColor,line width= 0.7pt,line join=round] (429.97, 36.13) --
-	(429.97, 39.63);
-
-\path[draw=drawColor,line width= 0.7pt,line join=round] (498.00, 36.13) --
-	(498.00, 39.63);
-
-\path[draw=drawColor,line width= 0.7pt,line join=round] (566.03, 36.13) --
-	(566.03, 39.63);
-
-\path[draw=drawColor,line width= 0.7pt,line join=round] (634.06, 36.13) --
-	(634.06, 39.63);
-
-\path[draw=drawColor,line width= 0.7pt,line join=round] (702.09, 36.13) --
-	(702.09, 39.63);
-\end{scope}
-\begin{scope}
-\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
-\definecolor{drawColor}{gray}{0.30}
-
-\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.12] at (429.97, 25.61) {0.6};
-
-\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.12] at (498.00, 25.61) {0.8};
-
-\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.12] at (566.03, 25.61) {1.0};
-
-\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.12] at (634.06, 25.61) {1.2};
-
-\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.12] at (702.09, 25.61) {1.4};
-\end{scope}
-\begin{scope}
-\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
-\definecolor{drawColor}{RGB}{0,0,0}
-
-\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.40] at (566.03, 10.06) {Generation interval ratio, $\bar{G}_{\mathrm{var}}/\bar{G}_{\mathrm{wt}}$};
-\end{scope}
-\begin{scope}
-\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
-\definecolor{drawColor}{RGB}{0,0,0}
-
-\node[text=drawColor,rotate= 90.00,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.40] at (377.99,185.84) {Speed (1/days)};
-\end{scope}
-\begin{scope}
-\path[clip] (  0.00,  0.00) rectangle (722.70,361.35);
-\definecolor{drawColor}{RGB}{0,0,0}
-
-\node[text=drawColor,anchor=base west,inner sep=0pt, outer sep=0pt, scale=  1.68] at (416.36,342.78) {B};
+\node[text=drawColor,anchor=base,inner sep=0pt, outer sep=0pt, scale=  1.40] at (563.24, 10.07) {Generation interval ratio, $\bar{G}_{\mathrm{var}}/\bar{G}_{\mathrm{wt}}$};
 \end{scope}

### 2021-04-02 newvariant.tex

--- 2021-03-28 newvariant.tex
@@
 
-As our baseline scenario, we assume $\delta = 0.1/\textrm{days}$ based on the observed relative speed of the new variant in the UK \citep{davies2021estimated}.
-This speed can be then used to estimate the relative strength $\rho = \Ry{var}/\Ry{wt}$ of the new variant with a caveat; such inference depends on assumptions about the generation-interval distribution.
-The mean generation interval $\bar{G}_w$ of the wild type is assumed to equal 5 days \citep{ferretti2020quantifying}.
-The squared coefficient of variation $\kappa$ in generation intervals is assumed to equal 0.2 \citep{ferretti2020quantifying}---for simplicity, we assume that both the variant and the wild type have equal $\kappa$ and only consider differences in the mean.
-As we expect there to be greater uncertainty in $\kappa$ than in $\bar{G}$, we evaluate the estimates of relative strength $\rho$ across a wide range of $\kappa$, ranging from 0 (delta distribution) to 1 (exponential distribution).
+We thus ask: what factors affect the relative strength $\rho = \Ry{var}/\Ry{wt}$ of a new variant, given an observed relative speed $\delta$?
+This inference will depend in general on assumptions about the growth rate of the wild type and on the generation-interval distributions of both variants.
+These have generally been assumed to be the same, but we will also consider the possibility that they are different.
 
-Previous studies have assumed that the generation-interval distributions do not differ between the wild type and the variant: $\Gy{var} = \Gy{wt} = \bar{G}$.
-In this case, the estimate of the relative strength is given by \citep{park2019practical}:
+As our baseline scenario, we take the relative speed of the variant to be $\delta = 0.1/\textrm{day}$ \citep{davies2021estimated}; the mean generation interval of the wild type to be $\bar{G}_w = 5\days$ \citep{ferretti2020quantifying}; and the squared coefficient of variation of generation intervals to be $\kappa=0.2$ \citep{ferretti2020quantifying}.
+For simplicity, we assume that the variant and the wild type have equal $\kappa$ throughout, and only consider differences in the mean.
+\jd{This is awkward; let's skip for now:} As we expect there to be greater uncertainty in $\kappa$ than in $\bar{G}$, we evaluate the estimates of relative strength $\rho$ across a wide range of $\kappa$, ranging from 0 (delta distribution) to 1 (exponential distribution).
+
+When the variant and wild type have the same generation-interval distributions, $\Gy{var} = \Gy{wt} = \bar{G}$, 
+the estimate of the relative strength is given by \citep{park2019practical}:
 \begin{equation}
 \hat{\rho} = \left(1 + \frac{\kappa \delta \bar{G}}{1 + \kappa \ry{wt} \bar{G}}\right)^{1/\kappa}.
+\label{eq:rhoGbar}
 \end{equation}
-Therefore, estimates of relative strength $\rho$ depends not only on the relative speed $\delta$ but also on how fast the wild type is spreading in the population $r_w$---a few analyses have implicitly or explicitly neglected this factor by either assuming $r_w = 0$ \citep{switzerland2021variant} or $\kappa = 0$ \citep{davies2021estimated}, in which case we get $\hat{\rho} = \exp(\delta \bar{G})$.
-Instead, if the wild type was spreading faster than previously assumed (higher $r_w$), the estimates of the relative strength $\rho$ will be go down (\fref{relstrength}A).
+Therefore, estimates of relative strength $\rho$ depend not only on the relative speed $\delta$ but also on how fast the wild type is spreading in the population \rw---a few analyses have implicitly or explicitly neglected this factor by either assuming $\rw = 0$ \citep{switzerland2021variant} or $\kappa = 0$ \citep{davies2021estimated} (in the latter case $\hat{\rho} = \exp(\delta \bar{G})$).
+
+\eref{rhoGbar} shows that, for realistic (non-zero) values of $\kappa$, estimates of the strength $\rho$ go down as estimates of the growth rate $\rw$ go up. 
+\jd{I like effective dispersion, but it does no good to just mention it once. Leave it out for now?}
 The relative strength also depends on the \emph{effective dispersion} in generation intervals (i.e., the amount of variability in generation intervals, as measured by $\kappa$)---using narrower distributions makes $\hat{\rho}$ less sensitive to $\ry{wt}$.
+These effects are illustrated in \fref{relstrength}A.
 
@@
 Here, we explored how the generation-interval distribution shapes the link between relative strength and speed of the new variant.
-We define relative strength as the ratio between reproduction numbers ($\rho=\Ry{var}/\Ry{wt}$) and relative speed as the difference between growth rates ($\delta=\ry{var}-\ry{wt}$) of the variant and the wild type.
-When the generation-interval distributions are assumed to be identical, we find that estimates of $\hat{\rho}$ decreases with $r_w$ whereas $\hat{\delta}$ increase with $\RR_w$.
+We define relative strength as the ratio between reproduction numbers ($\rho=\RR_v/\RR_w$) and relative speed as the difference between growth rates ($\delta=r_v-\rw$) of the variant and the wild type.
+When the generation-interval distributions are assumed to be identical, we find that estimates of $\hat{\rho}$ decreases with $\rw$ whereas $\hat{\delta}$ increase with $\RR_w$.
 When we account for possible differences in generation-interval distributions, these relationships can change directions and further lead to biases in these estimates.

### 2021-04-02 newvariant.tex
Some changes to 4 and 5; also some command organization

@@
 
+%% Consistent, changeable style for subscripts
+\newcommand{\vvvar}{\mathrm{var}}
+\newcommand{\wwwt}{\mathrm{wt}}
+
 \newcommand{\rx}[1]{\ensuremath{{r}_{#1}}\xspace} 
 \newcommand{\ry}[1]{\rx{\mathrm{#1}}} 
-\newcommand{\rw}{\ry{wt}}
+\newcommand{\rw}{\rx{\wwwt}}
+\newcommand{\rv}{\rx{\vvvar}}
 
@@
 \newcommand{\Ro}{\Rx{0}}
-\newcommand{\Rc}{\Rx{\mathrm{c}}}
-\newcommand{\Ri}{\Rx{\mathrm{i}}}
 \newcommand{\RR}{\ensuremath{{\mathcal R}}\xspace}
-\newcommand{\Rhat}{\ensuremath{{\hat\RR}}}
-\newcommand{\Rnaive}{\ensuremath{{\mathcal R}_{\textrm{\tiny naive}}}\xspace}
-\newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}
+\newcommand{\Rw}{\Rx{\wwwt}}
+\newcommand{\Rv}{\Rx{\vvvar}}
+
 \newcommand{\days}{\ensuremath{\, \textrm{day}}}
@@
 
+\newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}
 \newcommand{\tstart}{\ensuremath{\tsub{t}{start}}\xspace}
@@
 Here, we explore how different assumptions about the relative generation time of the new variant affect estimates of its strength $\RR$ and speed $r$.
-We do this by exploring the relationship between relative strength (the ratio $\Ry{var}/\Ry{wt}$) and relative speed (the difference $\ry{var}-\ry{wt}$).
+We do this by exploring the relationship between relative strength (the ratio $\Rv/\Rw$) and relative speed (the difference $\rv-\rw$).
 We find that neglecting differences in the generation-interval distributions can lead to biased estimates, which, in turn, influence mitigation priorities.
@@
 \begin{align}
-p(t) &= \frac{\iy{var}(t_0) \exp(\ry{var} t)}{\iy{wt}(t_0) \exp(\ry{wt} t) + \iy{var}(t_0) \exp(\ry{var} t)},
+p(t) &= \frac{\iy{var}(t_0) \exp(\rv t)}{\iy{wt}(t_0) \exp(\rw t) + \iy{var}(t_0) \exp(\rv t)},
 \\ &= \frac{1}{1 + \left(\iy{wt}(t_0)/\iy{var}(t_0)\right) \exp(-\delta t)},
 \end{align}
-where the logistic growth rate $\delta = \ry{var} - \ry{wt}$ corresponds to the relative speed of the epidemic.
+where the logistic growth rate $\delta = \rv - \rw$ corresponds to the relative speed of the epidemic.
 
@@
 (A) Estimated relative strength of the new variant $\hat{\rho}$ given speed of the wild type
-$\ry{wt}$ assuming identical generation-interval distributions.
-(B) True relative strength of the new variant $\rho$ given speed of the wild type $\ry{wt}$ and ratio between the mean generation interval of the new variant $\bar{G}_v$ and that of the wild type $\bar{G}_w$.
+$\rw$ assuming identical generation-interval distributions.
+(B) True relative strength of the new variant $\rho$ given speed of the wild type $\rw$ and ratio between the mean generation interval of the new variant $\bar{G}_v$ and that of the wild type $\bar{G}_w$.
 (C) Ratio between the true and the estimated relative strength of the new variant $\rho/\hat{\rho}$.
@@
 
-We thus ask: what factors affect the relative strength $\rho = \Ry{var}/\Ry{wt}$ of a new variant, given an observed relative speed $\delta$?
+We thus ask: what factors affect the relative strength $\rho = \Rv/\Rw$ of a new variant, given an observed relative speed $\delta$?
 This inference will depend in general on assumptions about the growth rate of the wild type and on the generation-interval distributions of both variants.
@@
 \begin{equation}
-\hat{\rho} = \left(1 + \frac{\kappa \delta \bar{G}}{1 + \kappa \ry{wt} \bar{G}}\right)^{1/\kappa}.
+\hat{\rho} = \left(1 + \frac{\kappa \delta \bar{G}}{1 + \kappa \rw \bar{G}}\right)^{1/\kappa}.
 \label{eq:rhoGbar}
@@
 \jd{I like effective dispersion, but it does no good to just mention it once. Leave it out for now?}
-The relative strength also depends on the \emph{effective dispersion} in generation intervals (i.e., the amount of variability in generation intervals, as measured by $\kappa$)---using narrower distributions makes $\hat{\rho}$ less sensitive to $\ry{wt}$.
+The relative strength also depends on the \emph{effective dispersion} in generation intervals (i.e., the amount of variability in generation intervals, as measured by $\kappa$)---using narrower distributions makes $\hat{\rho}$ less sensitive to $\rw$.
 These effects are illustrated in \fref{relstrength}A.
 
-However, if the true mean generation interval of the new variant is different from that of the wild type ($\bar{G}_v \neq \bar{G}_w$), the true relative strength $\rho = \RR_v/\RR_w$ is given by:
+However, if the true mean generation interval of the new variant is different from that of the wild type ($\bar{G}_v \neq \bar{G}_w$), the true relative strength $\rho = \Rv/\Rw$ is given by:
 \begin{equation}
@@
 In this case, longer (or shorter) mean generation interval of the new variant generally translates to higher (or lower) values of $\rho$---\fref{relstrength}B illustrates this effect.
-However, this direction can change when both the new variant and the wild type are decaying exponentially (see area below $r_v = 0$ line in \fref{relstrength}B)---for a given rate of decline, longer generation intervals lead to a \emph{lower} value of strength (and therefore lower relative strength $\rho$).
+However, this direction can change when both the new variant and the wild type are decaying exponentially (see area below $\rv = 0$ line in \fref{relstrength}B)---for a given rate of decline, longer generation intervals lead to a \emph{lower} value of strength (and therefore lower relative strength $\rho$).
 
@@
 
-While relative speed is easier to measure, we do not generally expect $\delta$ to be constant if other factors governing epidemic spread are changing.
-Instead, many biological mechanisms better translate to assuming a constant value of relative strength $\rho$ over changing conditions.
-For example, if the new variant has a higher transmission rate, the ratio between the strength of the new variant and that of the wild type will remain constant ($=\rho$) under strength-like interventions that reduces the strength of both strains by a constant amount.
-However, this is not necessarily true under speed-like interventions (e.g., contact tracing and self-isolation), which depends on time since infection: if the generation-interval distributions are different, the effectiveness of interventions will also differ.
-For now, we consider consequences of assuming a constant relative strength $\rho$ and predicting the relative speed $\delta$---an assumption often made (either implicitly or explicitly) in studies that have tried to predict the future spread of the new variant, including when it will replace the wild type.
+We do not generally expect the relative speed $\delta$ to be remain constant if other factors governing epidemic spread are changing.
+Indeed, many biological mechanisms better translate to assuming a constant value of relative strength $\rho$ over changing conditions.
+For example, if the proportion of the population susceptible declines, or the average contact rate changes, while other factors remain constant, the relative strength $\rho$ is expected to remain relatively constant; in general, this will imply a change in relative speed $\delta$.
+
+We thus investigate how $\delta$ is expected to change with \Rw, if $\rho$ remains constant, and how this expectation changes with the ratio of the generation intervals.
 As our baseline scenario we assume $\rho = 1.61$, which is the value we obtain for $\delta=0.1\,\textrm{days}$, $\bar{G}_w = \bar{G}_v = 5\,\textrm{days}$, and $\kappa = 1/5$.
@@
 \textbf{Relative speed of the new variant given assumed relative strength.}
-(A) Estimated relative speed of the new variant $\hat{\delta}$ given strength of the wild type $\RR_w$ assuming identical generation-interval distributions.
-(B) True relative strength of the new variant $\delta$ given strength of the wild type $\RR_w$ and ratio between the mean generation interval of the new variant $\bar{G}_v$ and that of the wild type $\bar{G}_w$..
+(A) Estimated relative speed of the new variant $\hat{\delta}$ given strength of the wild type $\Rw$ assuming identical generation-interval distributions.
+(B) True relative strength of the new variant $\delta$ given strength of the wild type $\Rw$ and ratio between the mean generation interval of the new variant $\bar{G}_v$ and that of the wild type $\bar{G}_w$..
 (C) Difference between the true and the estimated relative speed of the new variant $\delta - \hat{\delta}$.
@@
 
-Like others, we first ignore possible differences between the generation-interval distributions of the wild type and the new variant and try to estimate $\delta$:
+As above, we first ignore possible differences between the generation-interval distributions of the wild type and the new variant and estimate $\delta$:
 \begin{equation}
-\hat{\delta} = \frac{(\rho \RR_w)^{\kappa} - \RR_w^{\kappa}}{\kappa \bar{G}_w}.
+\hat{\delta} = \frac{(\rho \Rw)^{\kappa} - \Rw^{\kappa}}{\kappa \bar{G}_w}.
 \end{equation}
-Here, we see that the estimated $\hat{\delta}$ increases as the underlying strength of the wild type $\RR_w$ increases (\fref{relspeed}A).
+Here, we see that the estimated $\hat{\delta}$ increases as the underlying strength of the wild type $\Rw$ increases (\fref{relspeed}A).
 In other words, we expect the relative speed of the new variant to change as epidemiological conditions change, suggesting that a logistic model that assumes constant growth rate may not be appropriate for modeling the proportion of the new variant.
-A narrower distribution causes $\hat{\delta}$ to be less sensitive to $\RR_w$;
+A narrower distribution causes $\hat{\delta}$ to be less sensitive to $\Rw$;
 when $\kappa = 0$ (delta-distributed generation intervals), $\hat{\delta}$ remains constant.
@@
 \begin{equation}
-\delta = \frac{(\rho \RR_w)^{\kappa} - 1}{\kappa \bar{G}_v} - \frac{\RR_w^{\kappa} - 1}{\kappa \bar{G}_w}.
+\delta = \frac{(\rho \Rw)^{\kappa} - 1}{\kappa \bar{G}_v} - \frac{\Rw^{\kappa} - 1}{\kappa \bar{G}_w}.
 \end{equation}
-In this case, $\delta$ generally decreases as the mean generation interval of the new variant increases, except when both $\RR_v < 1$ and $\RR_w < 1$ (\fref{relspeed}B).
-In some extreme cases (e.g., when $\bar{G}_v$ and $\RR_v$ are very low or when $\bar{G}_v$ and $\RR_v$ are very large), we can obtain $\delta < 0$---an unlikely scenario given that fast replacement of the new variant have been observed under mixed conditions.
-This suggests that differences in generation intervals between the new variant and the wild type are likely to be not too big.
+In this case, $\delta$ generally decreases as the mean generation interval of the new variant increases, except when both $\Rv < 1$ and $\Rw < 1$ (\fref{relspeed}B).
+In some extreme cases---when $\bar{G}_v$ and $\Rv$ are very low or when $\bar{G}_v$ and $\Rv$ are very large---we have $\delta < 0$.
+This is a scenario that does not seem to have occurred for B.1.1.7, where fast replacement has been observed under a wide variety of conditions, perhaps suggesting that any differences in generation interval for this variant are not too big.
 Finally, assuming equal generation intervals generally leads to both under- and over-estimation of $\delta$ (\fref{relspeed}C).
 
-\section{Inferring relative strength from incidence data, in theory}
+\section{Inferring relative strength from incidence data}
 
-In practice, epidemiological conditions change over time due to intervention measures and susceptible depletion effects.
+In practice, epidemiological conditions change over time due to intervention measures and susceptible depletion.
 Such changes can be characterized by ``time-varying'' or ``instantaneous'' reproduction numbers, $\RR(t)$ \citep{fraser2007estimating}.
@@
 \end{equation}
-Under strength-like intervention measures that reduce transmission rates by a constant amount, we expect ratios between true reproduction numbers to remain constant and correspond to the true relative strength: $\Ry{var}(t)/\Ry{wt}(t) = \rho$.
-However, if the assumed generation-interval distribution $\hat{g}(\tau)$ is different from the true distribution, then the ratio between the estimated reproduction numbers $\hat{\rho}(t) = \hat{\RR}_{\textrm{var}}(t)/\hat{\RR}_{\textrm{wt}}(t)$ need not be constant, where
-\begin{equation}
-\hat{\RR}_x(t) = \frac{i_x(t)}{\int_0^\infty i_x(t-\tau) \hat{g}(\tau) \dtau}.
-\end{equation}
+Under strength-like intervention measures that reduce transmission rates by a constant amount, we expect ratios between true reproduction numbers to remain constant and correspond to the true relative strength: $\Rv(t)/\Rw(t) = \rho$.
+However, if the assumed generation-interval distribution $\hat{g}(\tau)$ is different from the true distribution, then the ratio between the estimated reproduction numbers $\hat{\rho}(t) = \hat{\RR}_{\textrm{var}}(t)/\hat{\RR}_{\textrm{wt}}(t)$ may change, even if the true ratio does not.
 
-Here, we simulate three different scenarios using a two strain renewal equation that assumes perfect cross-immunity (\fref{Rtbias}): 
-(1) the wild type and the variant have the same generation-interval distributions, which match the assumed distribution (\fref{Rtbias}A--C);
-(2) the assumed distribution matches the generation-interval distribution of the wild type but has a longer mean than the mean generation interval of the variant (\fref{Rtbias}D--F); and
-(3) the assumed distribution matches the generation-interval distribution of the wild type but has a shorter mean than the mean generation interval of the variant (\fref{Rtbias}G--I).
-Then, we compare the estimated ratio $\hat{\rho}(t) = \hat{\RR}_{\textrm{var}}(t)/\hat{\RR}_{\textrm{wt}}(t)$ with the true ratio $\rho = \Ry{var}(t)/\Ry{wt}(t)$.
-In order to reflect smooth changes in $\RR$ driven by non-pharmaceutical interventions, we model $\Ry{wt}(t)$ phenomenologically and let $\Ry{var}(t) = \rho \Ry{wt}(t)$.
+Here, we investigate under the assumption that the true generation-interval distribution of the wild type is known, and use a two-strain renewal equation which assumes perfect cross-immunity to simulate three different scenarios:
+(1) the wild type and the variant have the same generation-interval distributions (which match the known distribution, shown in \fref{Rtbias}A--C);
+(2) the variant has a shorter mean generation interval (\fref{Rtbias}D--F); and
+(3) the variant has a longter mean generation interval (\fref{Rtbias}G--I).
+Then, we compare the estimated ratio $\hat{\rho}(t) = \hat{\RR}_{\textrm{var}}(t)/\hat{\RR}_{\textrm{wt}}(t)$ with the true ratio $\rho = \Rv(t)/\Rw(t)$.
+In order to reflect smooth changes in $\RR$ driven by non-pharmaceutical interventions and behaviour changes, we model $\Rw(t)$ phenomenologically and let $\Rv(t) = \rho \Rw(t)$.
 
@@
 For all simulations, squared coefficient of variation in generation intervals is assumed to equal $\kappa = 1/5$.
-We assume $\Ry{wt}(t)=2.5$ before $t=15\,\textrm{days}$ to reflect unmitigated spread and let $\Ry{wt}(t)=2.5 (1+0.7\cos((t-15)/10))/1.7$ after to model smooth introduction and lifting of non-pharmaceutical interventions.
+We assume $\Rw(t)=2.5$ before $t=15\,\textrm{days}$ to reflect unmitigated spread and let $\Rw(t)=2.5 (1+0.7\cos((t-15)/10))/1.7$ after to model smooth introduction and lifting of non-pharmaceutical interventions.
 }
@@
 However, if the generation-interval distribution of the variant is different from the assumed distribution, the ratio changes over time (\fref{Rtbias}F,I).
-In particular, if the true generation intervals have a shorter (or longer) mean than the assumed distribution, we over-estimate (or under-estimate) $\Ry{var}(t)$ during the growth phase and under-estimate (or over-estimate) $ $ during the decay phase (\fref{Rtbias}D,G), which further translates to biases in the estimated relative strength (\fref{Rtbias}F,I).
+In particular, if the true generation intervals have a shorter (or longer) mean than the assumed distribution, we over-estimate (or under-estimate) $\Rv(t)$ during the growth phase and under-estimate (or over-estimate) $ $ during the decay phase (\fref{Rtbias}D,G), which further translates to biases in the estimated relative strength (\fref{Rtbias}F,I).
 
@@
 While the relative speed need not stay constant over the course of an epidemic, at any given point in epidemic, we can measure the speed of the new variant and the wild type and ask how much more intervention is required to control the spread of both strains (relative to measured conditions).
-As a baseline scenario, we assume $\ry{wt}=0\,\textrm{/days}$ and $\delta=0.1\,\textrm{/days}$ (and therefore $\ry{var}=0\,\textrm{/days}$).
+As a baseline scenario, we assume $\rw=0\,\textrm{/days}$ and $\delta=0.1\,\textrm{/days}$ (and therefore $\rv=0\,\textrm{/days}$).
 In this case, the wild type is already under control, whereas the variant requires additional intervention measures to stop growing.
@@
 a strength-like intervention which affects the kernel by a constant amount $\theta$ regardless of when they were infected ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau)/\theta$) and a speed-like intervention which affect the kernel at a constant rate $\phi$ ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau) \exp(-\phi \tau)$).
-In this case, it is clear that we can control the spread of the variant when either $\theta > \Ry{var}$ or $\phi > \ry{var}$---these two quantities ($\theta$ and $\phi$) are also referred to as strength and speed of intervention, respectively.
+In this case, it is clear that we can control the spread of the variant when either $\theta > \Rv$ or $\phi > \rv$---these two quantities ($\theta$ and $\phi$) are also referred to as strength and speed of intervention, respectively.
 In general, we can calculate the strength ($\theta = \Ry{pre}/\Ry{post}$) and speed ($\phi = \ry{pre} - \ry{post}$) of intervention, both providing equivalent measures for effectiveness of intervention, for any intervention by comparing epidemiological conditions before and after the intervention \citep{doi:10.1098/rspb.2020.1556}---these quantities, by definition, must be greater than pre-intervention strength and speed of the epidemic to control the spread.
@@
 (B) Speed of epidemic (black, solid), constant strength intervention (orange, dashed), and constant speed intervention (blue, dotted).
-Strength and speed are calculated assuming $\ry{wt}=0\,\textrm{/days}$, $\delta=0.1\,\textrm{/days}$, $\ry{var}=0\,\textrm{/days}$, $\Gy{wt}=5\,\textrm{days}$, and $\kappa=1/5$.
+Strength and speed are calculated assuming $\rw=0\,\textrm{/days}$, $\delta=0.1\,\textrm{/days}$, $\rv=0\,\textrm{/days}$, $\Gy{wt}=5\,\textrm{days}$, and $\kappa=1/5$.
 Constant strength and speed interventions are modeled such that post intervention strength of the new variant, assuming equal mean generation intervals at 5 days, corresponds to 0.9.
@@
 \fref{strengthspeed} compares the effectiveness of strength-based and speed-based interventions across a wide range of assumptions about generation-interval distributions.
-We begin by assuming that the variant and the wild type have equal mean generation intervals and model interventions such that they both reduce $\Ry{var}$ to 0.9 after intervention (\fref{strengthspeed}A).
+We begin by assuming that the variant and the wild type have equal mean generation intervals and model interventions such that they both reduce $\Rv$ to 0.9 after intervention (\fref{strengthspeed}A).
 If the mean generation interval of the variant is longer than that of the wild type, then the strength of the variant will be higher;
@@
 Here, we explored how the generation-interval distribution shapes the link between relative strength and speed of the new variant.
-We define relative strength as the ratio between reproduction numbers ($\rho=\RR_v/\RR_w$) and relative speed as the difference between growth rates ($\delta=r_v-\rw$) of the variant and the wild type.
-When the generation-interval distributions are assumed to be identical, we find that estimates of $\hat{\rho}$ decreases with $\rw$ whereas $\hat{\delta}$ increase with $\RR_w$.
+We define relative strength as the ratio between reproduction numbers ($\rho=\Rv/\Rw$) and relative speed as the difference between growth rates ($\delta=\rv-\rw$) of the variant and the wild type.
+When the generation-interval distributions are assumed to be identical, we find that estimates of $\hat{\rho}$ decreases with $\rw$ whereas $\hat{\delta}$ increase with $\Rw$.
 When we account for possible differences in generation-interval distributions, these relationships can change directions and further lead to biases in these estimates.
@@
 However, this is only true if both have the same mean generation intervals (\fref{relspeed}B).
-If the new variant has longer generation intervals than the original strain, imposing strong intervention (therefore reducing both $\RR_w$ and $\RR_v$ by a constant amount) can result in a higher relative speed $\delta$, and therefore faster replacement of the original strain (\fref{relspeed}B).
-This is not necessarily a bad outcome as the absolute speed ($r_v$) of the new variant still decreases as $\RR_v$ decreases.
+If the new variant has longer generation intervals than the original strain, imposing strong intervention (therefore reducing both $\Rw$ and $\Rv$ by a constant amount) can result in a higher relative speed $\delta$, and therefore faster replacement of the original strain (\fref{relspeed}B).
+This is not necessarily a bad outcome as the absolute speed ($\rv$) of the new variant still decreases as $\Rv$ decreases.
 These effects should be taken into consideration for accurate projection of the spread of the new variant.

### 2021-04-04 newvariant.tex
Finishing pipeline
editing

@@
 \newcommand{\iy}[1]{\ix{\mathrm{#1}}}
+\newcommand{\iw}{\ix{\wwwt}}
+\newcommand{\iv}{\ix{\vvvar}}
 
@@
 \newcommand{\Gy}[1]{\Gx{\mathrm{#1}}}
+\newcommand{\Gw}{\Gx{\wwwt}}
+\newcommand{\Gv}{\Gx{\vvvar}}
 
@@
 Many studies have tried to estimate the strength of the novel variant from the observed speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021}.
-In contrast, some studies have assumed a value of the strength of the novel variant and tried to predict future spread (therefore, the speed) of the variant \citep{davies2021estimated}.
+In contrast, some studies have assumed a value of the strength of the novel variant and tried to predict its future spread (therefore, the speed) of the variant \citep{davies2021estimated}.
 While both approaches are reasonable in principle, holding different quantities constant (i.e., strength or speed) can lead to different conclusions, in practice, about the spread of the disease, and therefore, its control.
@@
 This implies that the mean generation interval of the new variant is likely to be longer than that of the wild type.
-Other studies have considered the possibility that the faster growth rate of the new variant may be driven, in part, by \emph{shorter} generation intervals \citep{davies2021estimated,volz2021transmission}.
+Other studies have considered the possibility that the faster growth rate of the new variant may be driven, in part, by shorter generation intervals \citep{davies2021estimated,volz2021transmission}.
 If the generation-interval distribution of the new variant is different from the generation-interval distribution of the wild type, current estimates of the strength of the new variants may be biased.
@@
 We find that neglecting differences in the generation-interval distributions can lead to biased estimates, which, in turn, influence mitigation priorities.
-\jd{Just thoughts.
-I know we haven't introduced this terminology yet, and I know that you're going to use different terminology below.
-The point is that the old version was clunky and I'm trying to sharpen it.
-We can keep working.}
 
@@
 \begin{align}
-p(t) &= \frac{\iy{var}(t_0) \exp(\rv t)}{\iy{wt}(t_0) \exp(\rw t) + \iy{var}(t_0) \exp(\rv t)},
-\\ &= \frac{1}{1 + \left(\iy{wt}(t_0)/\iy{var}(t_0)\right) \exp(-\delta t)},
+p(t) &= \frac{\iv(t_0) \exp(\rv t)}{\iw(t_0) \exp(\rw t) + \iv(t_0) \exp(\rv t)},
+\\ &= \frac{1}{1 + \left(\iw(t_0)/\iv(t_0)\right) \exp(-\delta t)},
 \end{align}
@@
 \caption{
-\textbf{Relative strength of the new variant given observed relative speed.}
-(A) Estimated relative strength of the new variant $\hat{\rho}$ given speed of the wild type
-$\rw$ assuming identical generation-interval distributions.
-(B) True relative strength of the new variant $\rho$ given speed of the wild type $\rw$ and ratio between the mean generation interval of the new variant $\bar{G}_v$ and that of the wild type $\bar{G}_w$.
-(C) Ratio between the true and the estimated relative strength of the new variant $\rho/\hat{\rho}$.
-True relative strength of the new variant $\rho$ is calculated using $\delta=0.1\pday$, $\bar{G}_w = 5\days$, and $\kappa = 1/5$.
+\textbf{Relative strength of the new variant given observed relative speed under four epidemiological conditions.}
+The relative strength of the new variant $\rho$ given speed of the wild type $\rw$, ratio between the mean generation interval of the new variant $\bar{G}_v$ and that of the wild type $\bar{G}_w$, and squared coefficient of variation in generation intervals $\kappa$.
+Relative strength of the new variant $\rho$ is calculated using $\delta=0.1\pday$, $\bar{G}_w = 5\days$, and $\kappa = 1/5$.
+Four epidemiological conditions assume $\rw=-0.15\pday$ ($\rw < \rv < 0$), $\rw=-0.1\pday$ ($\rw < \rv = 0$), $\rw=0\pday$ ($0 = \rw < \rv$), and $\rw=0.1\pday$ ($0 < \rw < \rv$).
 }
@@
 We thus ask: what factors affect the relative strength $\rho = \Rv/\Rw$ of a new variant, given an observed relative speed $\delta$?
-This inference will depend in general on assumptions about the growth rate of the wild type and on the generation-interval distributions of both variants.
+This inference will depend on assumptions about the generation-interval distributions of both strains.
 These have generally been assumed to be the same, but we will also consider the possibility that they are different.
+Then, given the mean generation interval of the variant $\Gv$ and the wild type $\Gw$, the relative strength $\rho = \Rv/\Rw$ under the gamma approximation is given by \citep{park2019practical}:
+\begin{equation}
+\rho = \left(\frac{1 + \kappa (\rw + \delta) \bar{G}_v}{1 + \kappa \rw \bar{G}_w}\right)^{1/\kappa}.
+\end{equation}
+Therefore, the relative strength $\rho$ depend not only on the relative speed $\delta$ and the generation-interval distributions but also on how fast the wild type is spreading in the population \rw---a few analyses have implicitly or explicitly neglected this factor by either assuming $\rw = 0$ \citep{switzerland2021variant} or $\kappa = 0$ \citep{davies2021estimated} (in the latter case $\hat{\rho} = \exp(\delta \bar{G})$).
 
@@
 For simplicity, we assume that the variant and the wild type have equal $\kappa$ throughout, and only consider differences in the mean.
-\jd{This is awkward; let's skip for now:} As we expect there to be greater uncertainty in $\kappa$ than in $\bar{G}$, we evaluate the estimates of relative strength $\rho$ across a wide range of $\kappa$, ranging from 0 (delta distribution) to 1 (exponential distribution).
+We evaluate the estimates of relative strength $\rho$ across a wide range of $\kappa$, ranging from 0 (delta distribution) to 1 (exponential distribution).
+To further explore how inference depends underlying epidemiological conditions (i.e., whether the variant and the wild type are growing or decaying), we consider four epidemiological conditions: (1) $\rw < \rv < 0$, (2) $\rw < \rv = 0$, (3) $0 = \rw < \rv$, and (4) $0 < \rw < \rv$.
 
-When the variant and wild type have the same generation-interval distributions, $\Gy{var} = \Gy{wt} = \bar{G}$, 
-the estimate of the relative strength is given by \citep{park2019practical}:
-\begin{equation}
-\hat{\rho} = \left(1 + \frac{\kappa \delta \bar{G}}{1 + \kappa \rw \bar{G}}\right)^{1/\kappa}.
-\label{eq:rhoGbar}
-\end{equation}
-Therefore, estimates of relative strength $\rho$ depend not only on the relative speed $\delta$ but also on how fast the wild type is spreading in the population \rw---a few analyses have implicitly or explicitly neglected this factor by either assuming $\rw = 0$ \citep{switzerland2021variant} or $\kappa = 0$ \citep{davies2021estimated} (in the latter case $\hat{\rho} = \exp(\delta \bar{G})$).
+\fref{relstrength} illustrates how the relative strength depends on the shape of the generation-interval distributions as well as underlying epidemiological conditions.
+In general, longer (or shorter) mean generation interval of the new variant translates to higher (or lower) values of $\rho$, except when $\rv \leq 0$.
+When $\rv = 0$ (and also $\rw < 0$), we always have $\Rv = 1$ and so the generation-interval distribution of the new variant does not affect the inference.
+When $\rv < 0$ (and also $\rw < 0$), we see that longer generation intervals decrease $\rho$ because longer generation intervals actually lead to slower decay (higher $r$).
+Assuming a narrower distribution (lower $\kappa$) has qualitatively similar effects as assuming longer generation intervals because they both reduce the amount of early transmission; 
+therefore, narrower distributions lead to higher values of $\rho$ when $\rv > 0$ and lower values of $\rho$ when $\rv \leq 0$.
 
-\eref{rhoGbar} shows that, for realistic (non-zero) values of $\kappa$, estimates of the strength $\rho$ go down as estimates of the growth rate $\rw$ go up. 
-\jd{I like effective dispersion, but it does no good to just mention it once. Leave it out for now?}
-The relative strength also depends on the \emph{effective dispersion} in generation intervals (i.e., the amount of variability in generation intervals, as measured by $\kappa$)---using narrower distributions makes $\hat{\rho}$ less sensitive to $\rw$.
-These effects are illustrated in \fref{relstrength}A.
-
-However, if the true mean generation interval of the new variant is different from that of the wild type ($\bar{G}_v \neq \bar{G}_w$), the true relative strength $\rho = \Rv/\Rw$ is given by:
-\begin{equation}
-\rho = \left(\frac{1 + \kappa (\rw + \delta) \bar{G}_v}{1 + \kappa \rw \bar{G}_w}\right)^{1/\kappa}.
-\end{equation}
-In this case, longer (or shorter) mean generation interval of the new variant generally translates to higher (or lower) values of $\rho$---\fref{relstrength}B illustrates this effect.
-However, this direction can change when both the new variant and the wild type are decaying exponentially (see area below $\rv = 0$ line in \fref{relstrength}B)---for a given rate of decline, longer generation intervals lead to a \emph{lower} value of strength (and therefore lower relative strength $\rho$).
-
-Therefore, neglecting potential differences in generation intervals can lead to mis-estimation of the relative strength and the variant, with implications for forecasting and planning.
-In \fref{relstrength}C, we quantify the bias that arises by incorrectly assuming generation intervals are the same using the ratio between the true value $\rho$ and the estimated value $\hat{\rho}$, which assumes equal generation-interval distributions.
-Across the ranges of parameters that we consider, we find that up to 1.4 fold (either upward or downward) adjustments to current estimates may be required.
+Consistent with earlier studies, we find that the new variant has higer strength than the wild type across a wide range of epidemiological conditions \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
+However, the magnitude of relative strength $\rho$ is sensitive to assumptions about generation intervals.
+For realistic values of $\kappa$ (excluding 0 and 1), we find that the relative strength $\rho$ ranges between 1.1--2.3.
+Even if we restrict the mean generation interval of the variant to differ by 20\% and only consider the scenario under which variant is growing (bottom two panels of \fref{relstrength}), we find that $\rho$ ranges between 1.3--1.8, exhibiting narrower but still large uncertatiny.
 
@@
 Indeed, many biological mechanisms better translate to assuming a constant value of relative strength $\rho$ over changing conditions.
-For example, if the proportion of the population susceptible declines, or the average contact rate changes, while other factors remain constant, the relative strength $\rho$ is expected to remain relatively constant; in general, this will imply a change in relative speed $\delta$.
+For example, if the proportion of the population susceptible declines, or the average contact rate changes, while other factors remain constant, the relative strength $\rho$ is expected to remain relatively constant;
+in general, this will imply a change in relative speed $\delta$.
 
 We thus investigate how $\delta$ is expected to change with \Rw, if $\rho$ remains constant, and how this expectation changes with the ratio of the generation intervals. 
-As our baseline scenario we assume $\rho = 1.61$, which is the value we obtain for $\delta=0.1\,\textrm{days}$, $\bar{G}_w = \bar{G}_v = 5\,\textrm{days}$, and $\kappa = 1/5$.
+Once again, we rely on the gamma approximation to find the relative speed $\delta$ given the mean generation interval of the variant $\Gv$ and the wild type $\Gw$, :
+\begin{equation}
+\delta = \frac{(\rho \Rw)^{\kappa} - 1}{\kappa \Gv} - \frac{\Rw^{\kappa} - 1}{\kappa \Gw}.
+\end{equation}
+As our baseline scenario we assume $\rho = 1.61$, which is the value we obtain for $\delta=0.1\,\textrm{days}$, $\Gw = \Gv = 5\,\textrm{days}$, and $\kappa = 1/5$.
+We also evaluate $\delta$ across four different epidemiological conditions as before: (1) $\Rw < \Rv < 1$, (2) $\Rw < \Rv = 1$, (3) $1 = \Rw < \Rv$, and (4) $1 < \Rw < \Rv$.
 
@@
 \caption{
-\textbf{Relative speed of the new variant given assumed relative strength.}
-(A) Estimated relative speed of the new variant $\hat{\delta}$ given strength of the wild type $\Rw$ assuming identical generation-interval distributions.
-(B) True relative strength of the new variant $\delta$ given strength of the wild type $\Rw$ and ratio between the mean generation interval of the new variant $\bar{G}_v$ and that of the wild type $\bar{G}_w$..
-(C) Difference between the true and the estimated relative speed of the new variant $\delta - \hat{\delta}$.
-True relative speed of the new variant $\delta$ is calculated using $\rho=1.61$, $\bar{G}_w = 5\,\textrm{days}$, and $\kappa=1/5$.
+\textbf{Relative speed of the new variant given assumed relative strength under four epidemiological conditions.}
+The relative speed of the new variant $\delta$ given strength of the wild type $\Rw$, ratio between the mean generation interval of the new variant $\Gv$ and that of the wild type $\Gw$, and squared coefficient of variation in generation intervals $\kappa$.
+Relative speed of the new variant $\delta$ is calculated using $\rho=1.61$, $\bar{G}_w = 5\days$, and $\kappa = 1/5$.
+Four epidemiological conditions assume $\Rw=0.5/\rho$ ($\Rw < \Rv < 1$), $\Rw=1/\rho$ ($\Rw < \Rv = 1$), $\Rw=1$ ($1 = \Rw < \Rv$), and $\Rw=\rho$ ($0 < \Rw < \Rv$).
 }
@@
 
-As above, we first ignore possible differences between the generation-interval distributions of the wild type and the new variant and estimate $\delta$:
-\begin{equation}
-\hat{\delta} = \frac{(\rho \Rw)^{\kappa} - \Rw^{\kappa}}{\kappa \bar{G}_w}.
-\end{equation}
-Here, we see that the estimated $\hat{\delta}$ increases as the underlying strength of the wild type $\Rw$ increases (\fref{relspeed}A).
-In other words, we expect the relative speed of the new variant to change as epidemiological conditions change, suggesting that a logistic model that assumes constant growth rate may not be appropriate for modeling the proportion of the new variant.
-A narrower distribution causes $\hat{\delta}$ to be less sensitive to $\Rw$;
-when $\kappa = 0$ (delta-distributed generation intervals), $\hat{\delta}$ remains constant.
-
-When we allow the mean generation interval of the new variant to be different ($\bar{G}_v \neq \bar{G}_w$), the true $\rho$--$\delta$ relationship under gamma approximation is given by \citep{park2019practical}:
-\begin{equation}
-\delta = \frac{(\rho \Rw)^{\kappa} - 1}{\kappa \bar{G}_v} - \frac{\Rw^{\kappa} - 1}{\kappa \bar{G}_w}.
-\end{equation}
-In this case, $\delta$ generally decreases as the mean generation interval of the new variant increases, except when both $\Rv < 1$ and $\Rw < 1$ (\fref{relspeed}B).
-In some extreme cases---when $\bar{G}_v$ and $\Rv$ are very low or when $\bar{G}_v$ and $\Rv$ are very large---we have $\delta < 0$.
-This is a scenario that does not seem to have occurred for B.1.1.7, where fast replacement has been observed under a wide variety of conditions, perhaps suggesting that any differences in generation interval for this variant are not too big.
-Finally, assuming equal generation intervals generally leads to both under- and over-estimation of $\delta$ (\fref{relspeed}C).
+\fref{relspeed} compares how generation-interval distributions affect predictions about relative speed under different epidemiological conditions.
+In general, longer generation intervals lead to slower relative speed of the variant when the incidence of both strains are increasing (\fref{relspeed}, bottom panels) because slower growth of the variant reduces the differences in absolute speed.
+When $\Rv=1$, the relative speed is insensitive to the generation-interval distribution of the variant because we always have $\rv=0$.
+When $\Rv<1$, longer generation intervals of the variant leads to slower decay ($\rv$ closer to 0), and therefore, greater relative speed.
+Once again, we see that assuming a narrower distribution (lower $\kappa$) has qualitatively similar effects as assuming a longer mean.
 
@@
 
-\bibliography{newvariant}
+\bibliography{newvariant.bib}
 

### 2021-04-04 relspeed.R

@@ -4,105 +4,63 @@ library(ggthemes)
 library(egg)
 library(tikzDevice)
 
-rho <- 1.61
-
-Rwvec <- seq(0.3, 1.7, length.out=31)
-kappavec <- c(0, 0.2, 0.5, 1)
-kappabase <- 0.2
-
-pardata <- expand.grid(kappavec, Rwvec)
-
+rw <- 0
+delta <- 0.1
+kappa_base <- 1/5
 Gw <- 5
+
+rho <- ((1 + kappa_base * (rw+delta) * Gw)^(1/kappa_base))
+
+Rwvec <- c(0.5/rho, 1/rho, 1, rho)
+kappavec <- c(0, 0.2, 0.5, 1)
 
 Gratio <- exp(seq(log(1/1.5), log(1.5), length.out=31))
 
-pardata2 <- expand.grid(Rwvec, Gratio)
+pardata <- expand.grid(kappavec, Rwvec, Gratio)
 
 deltadata <- apply(pardata, 1, function(x) {
   kappa <- x[[1]]
   Rw <- x[[2]]
+  Gratio <- x[[3]]
+  Gv <- Gratio * Gw
   
   if (kappa == 0) {
-    deltahat <- log(rho * Rw)/Gw - log(Rw)/Gw
+    delta <- log(rho * Rw)/Gv - log(Rw)/Gw
   } else {
-    deltahat <- ((rho * Rw)^kappa - 1)/(kappa*Gw) -  ((Rw)^kappa - 1)/(kappa*Gw)
+    delta <- ((rho * Rw)^kappa - 1)/(kappa*Gv) -  ((Rw)^kappa - 1)/(kappa*Gw)
   }
   
   data.frame(
     rho=rho,
     Rw=Rw,
-    deltahat=deltahat,
-    kappa=kappa
-  )
-}) %>%
-  bind_rows
-
-deltadata2 <- apply(pardata2, 1, function(x) {
-  Rw <- x[[1]]
-  Gratio <- x[[2]] ## Gv/Gw
-  Gv <- Gratio * Gw
-  
-  delta <- ((rho * Rw)^kappabase - 1)/(kappabase*Gv) -  ((Rw)^kappabase - 1)/(kappabase*Gw)
-  
-  deltahat <- ((rho * Rw)^kappabase - 1)/(kappabase*Gw) -  ((Rw)^kappabase - 1)/(kappabase*Gw)
-  
-  data.frame(
-    rho=rho,
[…]

### 2021-04-04 relstrength.R

@@ -5,109 +5,62 @@ library(egg)
 library(tikzDevice)
 
 delta <- 0.1
-rho <- 1.61
 
-rwvec <- seq(-0.13, 0.13, length.out=31)
+rwvec <- c(-0.15, -0.1, 0, 0.1)
 kappavec <- c(0, 0.2, 0.5, 1)
-kappabase <- 0.2
-
-pardata <- expand.grid(kappavec, rwvec)
 
 Gw <- 5
 
 Gratio <- exp(seq(log(1/1.5), log(1.5), length.out=31))
 
-pardata2 <- expand.grid(rwvec, Gratio)
+pardata <- expand.grid(kappavec, rwvec, Gratio)
 
 rhodata <- apply(pardata, 1, function(x) {
   kappa <- x[[1]]
   rw <- x[[2]]
+  Gratio <- x[[3]]
+  Gv <- Gratio * Gw
   
   if (kappa==0) {
-    rhohat <- exp(Gw * (rw + delta))/exp(Gw * rw)
+    rho <- exp(Gv * (rw + delta))/exp(Gw * rw)
   } else {
-    rhohat <- ((1 + kappa * (rw + delta) * Gw)^(1/kappa))/
+    rho <- ((1 + kappa * (rw + delta) * Gv)^(1/kappa))/
       ((1 + kappa * rw * Gw)^(1/kappa))
   }
   
   data.frame(
     delta=delta,
     rw=rw,
-    rhohat=rhohat,
+    rho=rho,
+    Gratio=Gratio,
     kappa=kappa
   )
 }) %>%
-  bind_rows
-
-rhodata2 <- apply(pardata2, 1, function(x) {
-  rw <- x[[1]]
-  Gratio <- x[[2]] ## Gv/Gw
-  Gv <- Gratio * Gw
-  
-  rho <- ((1 + kappabase * (rw + delta) * Gv)^(1/kappabase))/
-    ((1 + kappabase * rw * Gw)^(1/kappabase))
-  
-  rhohat <- ((1 + kappabase * (rw + delta) * Gw)^(1/kappabase))/
-    ((1 + kappabase * rw * Gw)^(1/kappabase))
-  
-  data.frame(
-    delta=delta,
-    Gratio=Gratio,
-    rw=rw,
-    rho=rho,
-    rhohat=rhohat,
-    bias=rho/rhohat
+  bind_rows %>%
+  mutate(
[…]

### 2021-04-04 newvariant.tex
one more paragraph

@@
 
+\fref{relspeed} also shows that relative speed depends on underlying epidemiological conditions. 
+For example, even when there are no differences in the generation-interval distributions ($\Gv=\Gw$, in this case), the relative speed $\delta$ can range between 0.08--0.11 when $\kappa=0.2$ and 0.06--0.14 when $\kappa=0.5$.
+Therefore, characterizing the spread of variants assuming constant relative speed (e.g., by fitting a standard logistic equation) without considering how epidemiological conditions change over time should be avoided.
+
 \section{Inferring relative strength from incidence data}

### 2021-04-04 relspeed.R

@@ -48,6 +48,9 @@ deltadata <- apply(pardata, 1, function(x) {
                        "$1 < \\mathcal{R}_{\\textrm{wt}} < \\mathcal{R}_{\\textrm{var}}$"))
   )
 
+deltadata %>%
+  filter(Gratio==1, kappa==0.5)
+
 g1 <- ggplot(deltadata) +
   geom_line(aes(Gratio, delta, lty=kappa, col=kappa)) +
   scale_x_log10("Relative mean generation interval, $\\bar{G}_{\\mathrm{var}}/\\bar{G}_{\\mathrm{wt}}$",

### 2021-04-04 newvariant.tex
bed time

@@
 
-Consistent with earlier studies, we find that the new variant has higer strength than the wild type across a wide range of epidemiological conditions \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
+Consistent with earlier studies, we find that the new variant has higher strength than the wild type across a wide range of epidemiological conditions \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
 However, the magnitude of relative strength $\rho$ is sensitive to assumptions about generation intervals.
 For realistic values of $\kappa$ (excluding 0 and 1), we find that the relative strength $\rho$ ranges between 1.1--2.3.
-Even if we restrict the mean generation interval of the variant to differ by 20\% and only consider the scenario under which variant is growing (bottom two panels of \fref{relstrength}), we find that $\rho$ ranges between 1.3--1.8, exhibiting narrower but still large uncertatiny.
+Even if we restrict the mean generation interval of the variant to differ by 20\% and only consider the scenario under which variant is growing (bottom two panels of \fref{relstrength}), we find that $\rho$ ranges between 1.3--1.8, exhibiting narrower but still large uncertainty.
 
@@
 Such changes can be characterized by ``time-varying'' or ``instantaneous'' reproduction numbers, $\RR(t)$ \citep{fraser2007estimating}.
+Then, we can try to estimate the relative strength of the variant by comparing the ratio between estimates of reproduction numbers over time \citep{volz2021transmission}.
+
 Assuming that the generation-interval distribution does not change over time, the instantaneous reproduction numbers of the new variant and of the wild type can be estimated from their corresponding incidence curves---an approach popularized by \cite{cori2013new}:
@@
 (2) the variant has a shorter mean generation interval (\fref{Rtbias}D--F); and
-(3) the variant has a longter mean generation interval (\fref{Rtbias}G--I).
+(3) the variant has a longer mean generation interval (\fref{Rtbias}G--I).
 Then, we compare the estimated ratio $\hat{\rho}(t) = \hat{\RR}_{\textrm{var}}(t)/\hat{\RR}_{\textrm{wt}}(t)$ with the true ratio $\rho = \Rv(t)/\Rw(t)$.
-In order to reflect smooth changes in $\RR$ driven by non-pharmaceutical interventions and behaviour changes, we model $\Rw(t)$ phenomenologically and let $\Rv(t) = \rho \Rw(t)$.
+In order to reflect smooth changes in $\RR$ driven by non-pharmaceutical interventions and behavior changes, we model $\Rw(t)$ phenomenologically and let $\Rv(t) = \rho \Rw(t)$.
 
@@
 (B, E, H) True (purple, solid) and estimated (orange, dashed) ratios between reproduction numbers of the new variant and the wild type over time.
-(C, F, I) True (purple, solid) and estimated (orange, dashed) relationships between estimated reproduction numbers on a log-log scale.
+(C, F, I) True (purple, solid) and estimated (orange, dashed) relationships between estimated reproduction numbers.
 Top row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type and the variant.
@@
 
-We can assess these biases by plotting the estimated strength of the variant $\hat{\RR}_{\textrm{var}}(t)$ against the estimated strength of the wild type $\hat{\RR}_{\textrm{wt}}(t)$ on a log-log scale (\fref{Rtbias}C,F,I).
-When generation-interval distributions are correctly specified, we obtain a straight line with a slope of 1. 
-In this case, the intercept corresponds to the logarithm of the relative strength $\log(\rho)$ (\fref{Rtbias}C).
-When the assumed mean generation interval is longer (or shorter) than the that of the variant, we over-estimate (or under-estimate) the slope (\fref{Rtbias}H,L) on a log-log plot.
-Smooth changes in $\RR$ further generate circular patterns but biases in slopes are clear.
+We can assess these biases by plotting the estimated strength of the variant $\hat{\RR}_{\textrm{var}}(t)$ against the estimated strength of the wild type $\hat{\RR}_{\textrm{wt}}(t)$ (\fref{Rtbias}C,F,I).
+When generation-interval distributions are correctly specified, we obtain a straight line with a slope of $\rho$ and intercept at zero (\fref{Rtbias}C).
+When the assumed mean generation interval is longer (or shorter) than the that of the variant, we over-estimate (or under-estimate) the slope (\fref{Rtbias}H,L).
+Smooth changes in $\RR$ generate circular patterns but biases in slopes are clear.
 
@@
 
-While the relative speed need not stay constant over the course of an epidemic, at any given point in epidemic, we can measure the speed of the new variant and the wild type and ask how much more intervention is required to control the spread of both strains (relative to measured conditions).
+While relative speed $\delta$ and strength $\rho$ are useful for characterizing the spread of the variant, \emph{absolute} speed $\rv$ and strength $\Rv$ of the variant are the main variables that determine its spread and control.
+In particular, at any given point in epidemic, we can measure the speed of the variant $\rv$ (or infer $\rv$ from $\rw$ and $\delta$) and ask how much more intervention is required to control the spread of both strains (since $\Rv < 1$ implies $\Rw < 1$).
 As a baseline scenario, we assume $\rw=0\,\textrm{/days}$ and $\delta=0.1\,\textrm{/days}$ (and therefore $\rv=0\,\textrm{/days}$).
@@
 In this case, it is clear that we can control the spread of the variant when either $\theta > \Rv$ or $\phi > \rv$---these two quantities ($\theta$ and $\phi$) are also referred to as strength and speed of intervention, respectively.
-In general, we can calculate the strength ($\theta = \Ry{pre}/\Ry{post}$) and speed ($\phi = \ry{pre} - \ry{post}$) of intervention, both providing equivalent measures for effectiveness of intervention, for any intervention by comparing epidemiological conditions before and after the intervention \citep{doi:10.1098/rspb.2020.1556}---these quantities, by definition, must be greater than pre-intervention strength and speed of the epidemic to control the spread.
-We compare how potential differences in generation-interval distributions affect the control of the spread of the new variant using this strength-and-speed framework.
+In general, we can calculate the strength ($\theta = \Ry{pre}/\Ry{post}$) and speed ($\phi = \ry{pre} - \ry{post}$) of intervention, both of which provide equivalent measures for effectiveness of intervention, for any intervention by comparing epidemiological conditions before and after the intervention \citep{doi:10.1098/rspb.2020.1556}---these quantities, by definition, must be greater than their corresponding pre-intervention strength and speed of the epidemic to control the spread.
+Here, we compare how potential differences in generation-interval distributions affect the control of the spread of the new variant using this strength-and-speed framework.
 
@@
 We define relative strength as the ratio between reproduction numbers ($\rho=\Rv/\Rw$) and relative speed as the difference between growth rates ($\delta=\rv-\rw$) of the variant and the wild type.
-When the generation-interval distributions are assumed to be identical, we find that estimates of $\hat{\rho}$ decreases with $\rw$ whereas $\hat{\delta}$ increase with $\Rw$.
-When we account for possible differences in generation-interval distributions, these relationships can change directions and further lead to biases in these estimates.
-We show how these biases might be assessed in practice.
-We also discuss that differences in generation intervals lead to different conclusions about the effectiveness of interventions:
-if the variant has longer or shorter generation intervals than the wild type, we should prioritize speed- or strength-like interventions, respectively.
+We show that longer generation intervals generally lead to higher relative strength and lower relative speed, given that the other quantities are fixed, when the incidence of variant is increasing (i.e., $\Rv > 1$ and $\rv > 0$).
+We also show that propagating uncertainty in possible differences in mean generation intervals alone can result in considerable variation in the estimates of relative strength $\rho$.
+We illustrate how these biases might be assessed in practice and further discuss how differences in generation intervals might lead to different conclusions about the effectiveness of interventions:
+if the variant has longer or shorter generation intervals than the wild type, speed- or strength-like interventions, respectively, will be more effective and therefore should be prioritized.
+
+As the new variants, in particular the B.1.1.7 variant, of SARS-CoV-2 are spreading and becoming the dominant strain in many countries, it is clear that the variants are more transmissible (have higher strength) than the wild type \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
+Our analysis also supports higher strength of the new variant across a wide range of epidemiological conditions and assumptions about generation-interval distributions.
+However, in order to obtain precise and accurate estimates of the relative strength, uncertainty in generation-interval distributions must be taken into account.
+If the variant has longer infectious period \citep{kissler2021densely}, and therefore generation intervals, current estimates may be underestimating the true relative strength.
+
+While previous analyses have typically focused on estimating the relative strength and speed, absolute strength and speed are needed to properly evaluate its spread and control.
+Future studies should prioritize characterizing the spread of variants in absolute terms.
+Epidemiological conditions of the wild type should also be reported with the relative strength and speed in order to provide a proper context to evaluate their dynamics.
+
+We used simulations to show how biases in relative strength from mis-specification of generation-interval distributions can be assessed from data:
+when the assumed generation-interval distribution does not match the true generation-interval distribution of the variant, estimates of absolute strength $\Rv(t)$ will be biased, causing the estimated relative strength to change over time.
+In doing so, we assumed that the incidence of infection caused by the variant and the wild type is known.
+However, estimating $\RR(t)$ from real data, such as case reports, is often associated with several practical challenges, such as delays between infection and case reporting and changes in testing patterns.
+For simplicity, we choose not to present any data analysis in this paper.
+Nonetheless, future studies may consider methods outlined here to assess potential biases in their estimates of relative strength.
 
@@
 As more information about the transmission and immunity profiles of the new variant becomes available, it is better to fix the relative strength as it better matches biological mechanisms for its higher strength.
-We suggest researchers to be mindful about which quantity they hold constant and how their assumptions lead to their conclusions.
-
-Our simple framework further reveals counterintuitive, but important, insights on how to control the spread of both wild type and variants.
-In general, one would expect that stronger intervention strategies are required in order to slow down the replacement of the original strain by the new variant.
-However, this is only true if both have the same mean generation intervals (\fref{relspeed}B).
-If the new variant has longer generation intervals than the original strain, imposing strong intervention (therefore reducing both $\Rw$ and $\Rv$ by a constant amount) can result in a higher relative speed $\delta$, and therefore faster replacement of the original strain (\fref{relspeed}B).
-This is not necessarily a bad outcome as the absolute speed ($\rv$) of the new variant still decreases as $\Rv$ decreases.
-These effects should be taken into consideration for accurate projection of the spread of the new variant.
+We suggest researchers to be mindful about which quantity they hold constant and how their assumptions lead to their conclusions \citep{doi:10.1098/rspb.2020.1556}.
 
@@
 
-Since the beginning of the SARS-CoV-2 pandemic, a few studies have emphasized  the relevance of leveraging generation interval distributions to improve estimates of strength (e.g., \cite{doi:10.1098/rsif.2020.0144}) and the relative importance of modes of transmission (e.g., asymptomatic vs.
+Since the beginning of the SARS-CoV-2 pandemic, a few studies have emphasized the relevance of leveraging generation interval distributions to improve estimates of strength (e.g., \cite{doi:10.1098/rsif.2020.0144,park2021forward}) and the relative importance of modes of transmission (e.g., asymptomatic vs.
 symptomatic, \cite{park2020time}).

### 2021-04-04 Rtbias.R

@@ -56,7 +56,7 @@ for (i in 1:length(svec)) {
     geom_line(aes(tvec, Rtest1, color="Wild type", lty="Estimated"), lwd=1) +
     geom_line(aes(tvec, Rtest2, color="Variant", lty="Estimated"), lwd=1) +
     scale_x_continuous("Time (days)", expand=c(0, 0), limits=c(15, 70)) +
-    scale_y_log10("Reproduction number, $\\mathcal{R}(t)$", limits=c(0.25, 7)) +
+    scale_y_continuous("Reproduction number, $\\mathcal{R}(t)$", limits=c(0.25, 7)) +
     scale_color_manual(values=c("red", "black")) +
     scale_linetype_manual(values=c(2, 1), guide=FALSE) +
     ggtitle(LETTERS[(i-1)*3+1]) +
@@ -77,7 +77,7 @@ for (i in 1:length(svec)) {
       geom_line(aes(tvec, Rtest1, lty="Estimated"), col="black", lwd=1) +
       geom_line(aes(tvec, Rtest2, col="Estimated", lty="Estimated"), lwd=1) +
       scale_x_continuous("Time (days)", expand=c(0, 0), limits=c(15, 70)) +
-      scale_y_log10("Reproduction number, $\\mathcal{R}(t)$", limits=c(0.25, 7)) +
+      scale_y_continuous("Reproduction number, $\\mathcal{R}(t)$", limits=c(0.25, 7)) +
       scale_color_manual("a", values=c("red", "red")) +
       scale_linetype_manual("a", values=c(2, 1)) +
       ggtitle(LETTERS[(i-1)*3+1]) +
@@ -92,13 +92,13 @@ for (i in 1:length(svec)) {
     geom_line(aes(tvec, Rt2/Rt1, col="True", lty="True"), lwd=1) +
     geom_line(aes(tvec, Rtest2/Rtest1, col="Estimated", lty="Estimated"), lwd=1) +
     scale_x_continuous("Time (days)", expand=c(0, 0), limits=c(15, 70)) +
-    scale_y_log10("Relative strength, $\\rho(t)$", limits=c(1, 2.5)) +
+    scale_y_continuous("Relative strength, $\\rho(t)$", limits=c(1, 2.5)) +
     scale_color_manual("a", values=c("orange", "purple")) +
     scale_linetype_manual("a", values=c(2, 1)) +
     ggtitle(LETTERS[(i-1)*3+2]) +
[…]

### 2021-04-04 relspeed.R

@@ -55,7 +55,7 @@ g1 <- ggplot(deltadata) +
   geom_line(aes(Gratio, delta, lty=kappa, col=kappa)) +
   scale_x_log10("Relative mean generation interval, $\\bar{G}_{\\mathrm{var}}/\\bar{G}_{\\mathrm{wt}}$",
                 breaks=c(0.67, 1, 1.5)) +
-  scale_y_continuous("Relative speed, $\\delta$") +
+  scale_y_continuous("Relative speed, $\\delta$ (1/day)") +
   facet_wrap(~Rw) +
   scale_color_colorblind("$\\kappa$") +
   scale_linetype_discrete("$\\kappa$") +

### 2021-04-04 newvariant.tex
bias and sleep

@@
 
-We can assess these biases by plotting the estimated strength of the variant $\hat{\RR}_{\textrm{var}}(t)$ against the estimated strength of the wild type $\hat{\RR}_{\textrm{wt}}(t)$ (\fref{Rtbias}C,F,I).
+We can also assess these biases by plotting the estimated strength of the variant $\hat{\RR}_{\textrm{var}}(t)$ against the estimated strength of the wild type $\hat{\RR}_{\textrm{wt}}(t)$ (\fref{Rtbias}C,F,I).
 When generation-interval distributions are correctly specified, we obtain a straight line with a slope of $\rho$ and intercept at zero (\fref{Rtbias}C).
 When the assumed mean generation interval is longer (or shorter) than the that of the variant, we over-estimate (or under-estimate) the slope (\fref{Rtbias}H,L).
-Smooth changes in $\RR$ generate circular patterns but biases in slopes are clear.
+In addition, the intercept (if we were to perform regression) need not be at zero due to biases in the slope.
+Smooth changes in $\RR$ further generate circular patterns but biases in slopes are clear.
 

### 2021-04-04 newvariant.tex
focuse on S5 and 6

@@
 
-Since the emergence of the new SARS-CoV-2 variant of concern (VoC), estimating its epidemic strength and speed remains one of the key questions in controlling its spread \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
+Since the emergence of SARS-CoV-2 variants of concern (VoCs), estimating variant epidemic strength and speed remains one of the key questions in controlling its spread \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
 Epidemic strength, characterized by the reproduction number $\RR$, provides information about the final size of the epidemic as well as the amount of intervention required to eliminate the disease \citep{anderson1991infectious}.
-Epidemic speed, characterized by the growth rate $r$ provides information about the population-level time scale of the epidemic and, potentially provides information on when the new variant will replace the existing variant as the dominant strain.
+Epidemic speed, characterized by the growth rate $r$ provides information about the population-level time scale of the epidemic and, potentially provides information on when a new variant will replace the existing variant as the dominant strain.
 Strength and speed are linked by generation intervals---defined as time between infection and transmission---which characterizes individual-level time scale of the epidemic \citep{roberts2007model,svensson2007note,wallinga2007generation}.
@@
 
+\jd{I have started editing to make us a little less focused on B.1.1.7; we're not completely clear whether this is mostly a paper about B1, or whether it's just an example.
+I think we should focus less, and think also about new variants that might be coming.
+I stopped here, because this \P\ seems pretty clearly B1-focused and that part seems fine.
+Something to keep in mind.}
 In addition, most studies have assumed that the generation-interval distributions do not differ between the existing strain and the variant, but recent evidence suggests that the new variant may have a longer duration of infection: 13.3 days (90\% CI: 10.1--16.5) for the new variant and 8.2 days (90\% CI: 6.5--9.7) for the wild type \citep{kissler2021densely}.
@@
 Other studies have considered the possibility that the faster growth rate of the new variant may be driven, in part, by shorter generation intervals \citep{davies2021estimated,volz2021transmission}.
-If the generation-interval distribution of the new variant is different from the generation-interval distribution of the wild type, current estimates of the strength of the new variants may be biased.
-More information is clearly needed, however. 
+In general, if the generation-interval distribution of a new variant is different from the generation-interval distribution of the wild type, estimates of variant strength may be biased.
 And the story may be complicated: generation intervals depend on many other biological and behavioral factors, including for example self-isolation after symptom onset leading to shorter generation intervals.
@@
 
+\jd{It would be nice to have the R axes logged.}
+
 When the assumed distribution matches the true distributions, the estimated reproduction numbers match the true values (\fref{Rtbias}A), and therefore the ratio remains constant (\fref{Rtbias}B).
 However, if the generation-interval distribution of the variant is different from the assumed distribution, the ratio changes over time (\fref{Rtbias}F,I).
-In particular, if the true generation intervals have a shorter (or longer) mean than the assumed distribution, we over-estimate (or under-estimate) $\Rv(t)$ during the growth phase and under-estimate (or over-estimate) $ $ during the decay phase (\fref{Rtbias}D,G), which further translates to biases in the estimated relative strength (\fref{Rtbias}F,I).
+In particular, if the true generation intervals have a shorter (or longer) mean than the assumed distribution, we over-estimate (or under-estimate) $\Rv(t)$ during the growth phase and under-estimate (or over-estimate) during the decay phase (\fref{Rtbias}D,G), which further translates to biases in the estimated relative strength (\fref{Rtbias}F,I).
+\jd{Not quite following the logic of how you're linking to the pictures.
+I note that panels C, E and H are not mentioned.}
 
@@
 Smooth changes in $\RR$ further generate circular patterns but biases in slopes are clear.
+\jd{Maybe you didn't finish editing here?
+You've now mentioned F and three times, but still not E or H.
+And I can't find L despite repeatedly updating.
+It also doesn't quite seem to match the caption.}
 
@@
 In particular, at any given point in epidemic, we can measure the speed of the variant $\rv$ (or infer $\rv$ from $\rw$ and $\delta$) and ask how much more intervention is required to control the spread of both strains (since $\Rv < 1$ implies $\Rw < 1$).
-As a baseline scenario, we assume $\rw=0\,\textrm{/days}$ and $\delta=0.1\,\textrm{/days}$ (and therefore $\rv=0\,\textrm{/days}$).
-In this case, the wild type is already under control, whereas the variant requires additional intervention measures to stop growing.
+As a baseline scenario, we assume $\rw=0$ and $\delta=0.1\pday$ (and therefore $\rv=0.1\pday$).
+\jd{This sentence invites arguments, and we don't need it anyway:} In this case, the wild type is already under control, whereas the variant requires additional intervention measures to stop growing.
 
-Here, we consider two types of interventions:
-a strength-like intervention which affects the kernel by a constant amount $\theta$ regardless of when they were infected ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau)/\theta$) and a speed-like intervention which affect the kernel at a constant rate $\phi$ ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau) \exp(-\phi \tau)$).
-In this case, it is clear that we can control the spread of the variant when either $\theta > \Rv$ or $\phi > \rv$---these two quantities ($\theta$ and $\phi$) are also referred to as strength and speed of intervention, respectively.
-In general, we can calculate the strength ($\theta = \Ry{pre}/\Ry{post}$) and speed ($\phi = \ry{pre} - \ry{post}$) of intervention, both of which provide equivalent measures for effectiveness of intervention, for any intervention by comparing epidemiological conditions before and after the intervention \citep{doi:10.1098/rspb.2020.1556}---these quantities, by definition, must be greater than their corresponding pre-intervention strength and speed of the epidemic to control the spread.
-Here, we compare how potential differences in generation-interval distributions affect the control of the spread of the new variant using this strength-and-speed framework.
+We consider two types of intervention:
+an intervention of constant ``strength'', reduces transmission by a constant factor $\theta$ regardless of age of infection ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau)/\theta$); and an intervention of constant ``speed'' which reduces transmission after infection by a constant rate $\phi$ ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau) \exp(-\phi \tau)$).
+In this case, we can control the spread of the variant when $\theta > \Rv$ or $\phi > \rv$, respectively.
+In general, we can calculate the strength ($\theta = \Ry{pre}/\Ry{post}$) and speed ($\phi = \ry{pre} - \ry{post}$) of intervention for any intervention by comparing epidemiological conditions before and after the intervention \citep{doi:10.1098/rspb.2020.1556}---these quantities, by definition, must be greater than their corresponding pre-intervention strength and speed of the epidemic to control the spread.
+Here, we use the strength-and-speed framework to compare how differences in generation-interval distributions would affect the control of the spread of a new variant.
 
@@
 since the strength of the strength-like intervention is insensitive to the underlying generation-interval distribution, we may fail to control the spread of the new variant.
-The strength of the speed-like intervention, however, depends on the new variant:
-since speed-like interventions, such as contact tracing, have disproportionate effects on preventing later transmission, they are more effective (therefore stronger) when the mean generation interval of the new variant is longer.
+With speed-like interventions, however, there is a countervailing effect:
+since speed-like interventions like contact tracing have disproportionate effects on preventing later transmission, they are more effective (therefore stronger) when the mean generation interval of the new variant is longer.
 
-The speed-based paradigm gives same results about the control but provides important insights (\fref{strengthspeed}B).
+The speed-based paradigm gives the same results about the control but provides important insights (\fref{strengthspeed}B).
 The observed speed of the variant at any given moment does not depend on our estimates of its mean generation interval.
-Therefore, if we can prevent infections at a faster rate than the observed speed of spread (i.e., if the speed of intervention is faster than the epidemic speed), we can control the epidemic---this explains why a constant-speed intervention can prevent the spread of the variant regardless of its mean generation interval.
+Therefore, if we can reduce infectiousness at a faster rate than the observed speed of spread (i.e., if the speed of intervention is faster than the epidemic speed), we can control the epidemic---this explains why a constant-speed intervention can prevent the spread of the variant regardless of its mean generation interval.
+\jd{This s.~feels repetitive, do we need it?:}
 On the other hand, differences in mean generation intervals between the variant and the wild type imply different post-intervention epidemic speed (therefore, intervention speed) under a constant-strength intervention, explaining why it might not be effective if the variant has longer generation intervals.
@@
 We define relative strength as the ratio between reproduction numbers ($\rho=\Rv/\Rw$) and relative speed as the difference between growth rates ($\delta=\rv-\rw$) of the variant and the wild type.
-We show that longer generation intervals generally lead to higher relative strength and lower relative speed, given that the other quantities are fixed, when the incidence of variant is increasing (i.e., $\Rv > 1$ and $\rv > 0$).
-We also show that propagating uncertainty in possible differences in mean generation intervals alone can result in considerable variation in the estimates of relative strength $\rho$.
-We illustrate how these biases might be assessed in practice and further discuss how differences in generation intervals might lead to different conclusions about the effectiveness of interventions:
-if the variant has longer or shorter generation intervals than the wild type, speed- or strength-like interventions, respectively, will be more effective and therefore should be prioritized.
+We show that longer generation intervals generally lead to higher relative strength for a given relative speed (and conversely, lower relative speed for a given relative strength); these can be reversed when incidence is decreasing.e., $\Rv > 1$ and $\rv > 0$).
+We also that uncertainty about the relative mean generation interval of the variant has important implications for estimates of relative strength of the variantj.
+We illustrate how these biases might be assessed and discuss how differences in generation intervals might lead to different conclusions about the effectiveness of interventions:
+if the variant has longer (respectively, shorter) generation intervals than the wild type, speed-like (strength-like) interventions, will be more relatively more effective than naive estimates would suggest.
 

### 2021-04-05 newvariant.tex
done?

@@
 Epidemic strength, characterized by the reproduction number $\RR$, provides information about the final size of the epidemic as well as the amount of intervention required to eliminate the disease \citep{anderson1991infectious}.
-Epidemic speed, characterized by the growth rate $r$ provides information about the population-level time scale of the epidemic and, potentially provides information on when a new variant will replace the existing variant as the dominant strain.
+Epidemic speed, characterized by the growth rate $r$, provides information about the population-level time scale of the epidemic and, potentially when a new variant will replace the existing variant as the dominant strain.
 Strength and speed are linked by generation intervals---defined as time between infection and transmission---which characterizes individual-level time scale of the epidemic \citep{roberts2007model,svensson2007note,wallinga2007generation}.
@@
 In contrast, some studies have assumed a value of the strength of the novel variant and tried to predict its future spread (therefore, the speed) of the variant \citep{davies2021estimated}.
-While both approaches are reasonable in principle, holding different quantities constant (i.e., strength or speed) can lead to different conclusions, in practice, about the spread of the disease, and therefore, its control.
-The reason is that inference of speed from strength and strength from speed depends on assumptions about the generation-interval distribution \citep{roberts2007model,svensson2007note,wallinga2007generation,dushoff2020speed}.
+While both approaches are reasonable in principle, holding different quantities constant (i.e., strength or speed) can lead to different conclusions, in practice, about the spread of the disease, and therefore, its control \citep{doi:10.1098/rspb.2020.1556}.
+The reason is that inference of speed from strength and strength from speed depends on assumptions about the generation-interval distribution \citep{roberts2007model,svensson2007note,wallinga2007generation}.
 For example, when we assume a fixed value of epidemic strength, longer generation intervals lead to faster epidemic (lower $r$), making the epidemic seem easier to control.
@@
 
-\jd{I have started editing to make us a little less focused on B.1.1.7; we're not completely clear whether this is mostly a paper about B1, or whether it's just an example.
-I think we should focus less, and think also about new variants that might be coming.
-I stopped here, because this \P\ seems pretty clearly B1-focused and that part seems fine.
-Something to keep in mind.}
-In addition, most studies have assumed that the generation-interval distributions do not differ between the existing strain and the variant, but recent evidence suggests that the new variant may have a longer duration of infection: 13.3 days (90\% CI: 10.1--16.5) for the new variant and 8.2 days (90\% CI: 6.5--9.7) for the wild type \citep{kissler2021densely}.
-This implies that the mean generation interval of the new variant is likely to be longer than that of the wild type.
-Other studies have considered the possibility that the faster growth rate of the new variant may be driven, in part, by shorter generation intervals \citep{davies2021estimated,volz2021transmission}.
+In addition, most studies have assumed that the generation-interval distributions do not differ between the existing strain and the variant, but recent evidence suggests that the B.1.1.7 variant may have a longer duration of infection: 13.3 days (90\% CI: 10.1--16.5) for the new variant and 8.2 days (90\% CI: 6.5--9.7) for the wild type \citep{kissler2021densely}.
+This implies that the mean generation interval of B.1.1.7 is likely to be longer than that of the wild type.
+Other studies have considered the possibility that the faster growth rate of new variants may be driven, in part, by shorter generation intervals \citep{davies2021estimated,volz2021transmission}.
 In general, if the generation-interval distribution of a new variant is different from the generation-interval distribution of the wild type, estimates of variant strength may be biased.
 And the story may be complicated: generation intervals depend on many other biological and behavioral factors, including for example self-isolation after symptom onset leading to shorter generation intervals.
+We also do not know whether duration of infection of other variants will be same or different among each other.
 
 Here, we explore how different assumptions about the relative generation time of the new variant affect estimates of its strength $\RR$ and speed $r$.
-We do this by exploring the relationship between relative strength (the ratio $\Rv/\Rw$) and relative speed (the difference $\rv-\rw$).
+We do this by exploring the relationship between relative strength (the ratio $\Rv/\Rw$) and relative speed (the difference $\rv-\rw$) under a wide range of assumptions about generation-interval distributions.
 We find that neglecting differences in the generation-interval distributions can lead to biased estimates, which, in turn, influence mitigation priorities.
+We also discuss how such biases might be assessed in practice.
 
@@
 
-We use the renewal equation framework to characterize the spread of two pathogen strains---in this case, the wild type SARS-CoV-2 virus and a focal variant of concern (e.g., B.1.1.7).
+We use the renewal equation framework to characterize the spread of two pathogen strains---in this case, the wild type SARS-CoV-2 virus and a focal variant of concern, such as B.1.1.7.
 Neglecting the (relatively slow) rate of new mutations, the current incidence of infection $i_x(t)$ caused by the wild type (``wt'') and the variant (``var'') can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary cases are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
@@
 
-As our baseline scenario, we take the relative speed of the variant to be $\delta = 0.1/\textrm{day}$ \citep{davies2021estimated}; the mean generation interval of the wild type to be $\bar{G}_w = 5\days$ \citep{ferretti2020quantifying}; and the squared coefficient of variation of generation intervals to be $\kappa=0.2$ \citep{ferretti2020quantifying}.
+As our baseline scenario, we take the relative speed of the variant to be $\delta = 0.1/\textrm{day}$ based on the observed relaitve growth rate of B.1.1.7 \citep{davies2021estimated}; the mean generation interval of the wild type to be $\bar{G}_w = 5\days$ \citep{ferretti2020quantifying}; and the squared coefficient of variation of generation intervals to be $\kappa=0.2$ \citep{ferretti2020quantifying}.
+We use relative speed of B.1.1.7 as an example but note that the qualitative conclusions should hold for other variants as well.
 For simplicity, we assume that the variant and the wild type have equal $\kappa$ throughout, and only consider differences in the mean.
@@
 
-\jd{It would be nice to have the R axes logged.}
-
 When the assumed distribution matches the true distributions, the estimated reproduction numbers match the true values (\fref{Rtbias}A), and therefore the ratio remains constant (\fref{Rtbias}B).
-However, if the generation-interval distribution of the variant is different from the assumed distribution, the ratio changes over time (\fref{Rtbias}F,I).
+However, if the generation-interval distribution of the variant is different from the assumed distribution, the ratio changes over time (\fref{Rtbias}E,H).
 In particular, if the true generation intervals have a shorter (or longer) mean than the assumed distribution, we over-estimate (or under-estimate) $\Rv(t)$ during the growth phase and under-estimate (or over-estimate) during the decay phase (\fref{Rtbias}D,G), which further translates to biases in the estimated relative strength (\fref{Rtbias}F,I).
-\jd{Not quite following the logic of how you're linking to the pictures.
-I note that panels C, E and H are not mentioned.}
 
@@
 When generation-interval distributions are correctly specified, we obtain a straight line with a slope of $\rho$ and intercept at zero (\fref{Rtbias}C).
-When the assumed mean generation interval is longer (or shorter) than the that of the variant, we over-estimate (or under-estimate) the slope (\fref{Rtbias}H,L).
+When the assumed mean generation interval is longer (or shorter) than the that of the variant, we over-estimate (or under-estimate) the slope (\fref{Rtbias}F,I).
 In addition, the intercept (if we were to perform regression) need not be at zero due to biases in the slope.
 Smooth changes in $\RR$ further generate circular patterns but biases in slopes are clear.
-\jd{Maybe you didn't finish editing here?
-You've now mentioned F and three times, but still not E or H.
-And I can't find L despite repeatedly updating.
-It also doesn't quite seem to match the caption.}
 
@@
 In particular, at any given point in epidemic, we can measure the speed of the variant $\rv$ (or infer $\rv$ from $\rw$ and $\delta$) and ask how much more intervention is required to control the spread of both strains (since $\Rv < 1$ implies $\Rw < 1$).
-As a baseline scenario, we assume $\rw=0$ and $\delta=0.1\pday$ (and therefore $\rv=0.1\pday$).
-\jd{This sentence invites arguments, and we don't need it anyway:} In this case, the wild type is already under control, whereas the variant requires additional intervention measures to stop growing.
+As a baseline scenario, we assume $\rw=0$ and $\delta=0.1\pday$ (and therefore $\rv=0.1\pday$), in which case additional intervention is required to bring $\rv$ below 0 (or, equivalently, $\Rv$ below 1)
 
 We consider two types of intervention:
-an intervention of constant ``strength'', reduces transmission by a constant factor $\theta$ regardless of age of infection ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau)/\theta$); and an intervention of constant ``speed'' which reduces transmission after infection by a constant rate $\phi$ ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau) \exp(-\phi \tau)$).
+an intervention of constant ``strength'', which reduces transmission by a constant factor $\theta$ regardless of age of infection ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau)/\theta$); and an intervention of constant ``speed'', which reduces transmission after infection by a constant rate $\phi$ ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau) \exp(-\phi \tau)$).
 In this case, we can control the spread of the variant when $\theta > \Rv$ or $\phi > \rv$, respectively.
@@
 Therefore, if we can reduce infectiousness at a faster rate than the observed speed of spread (i.e., if the speed of intervention is faster than the epidemic speed), we can control the epidemic---this explains why a constant-speed intervention can prevent the spread of the variant regardless of its mean generation interval.
-\jd{This s.~feels repetitive, do we need it?:}
-On the other hand, differences in mean generation intervals between the variant and the wild type imply different post-intervention epidemic speed (therefore, intervention speed) under a constant-strength intervention, explaining why it might not be effective if the variant has longer generation intervals.
 

### 2021-04-05 control.R

@@ -67,7 +67,7 @@ g1 <- ggplot(intervention) +
   geom_line(aes(Gratio, speed_strength, col="Constant speed intervention", lty="Constant speed intervention"), lwd=1) +
   scale_x_log10("Generation interval ratio, $\\bar{G}_{\\mathrm{var}}/\\bar{G}_{\\mathrm{wt}}$",
                 breaks=c(0.67, 1, 1.5)) +
-  scale_y_log10("Strength") +
+  scale_y_log10("Strength", breaks=c(1, 2, 4), limits=c(1, 4)) +
   scale_linetype_manual("", values=3:1) +
   scale_color_manual("", values=rev(colorblind_pal()(3))) +
   ggtitle("A") +

### 2021-04-05 relspeed.R

@@ -52,6 +52,7 @@ deltadata %>%
   filter(Gratio==1, kappa==0.5)
 
 g1 <- ggplot(deltadata) +
+  geom_hline(data=deltadata2, aes(yintercept=delta)) +
   geom_line(aes(Gratio, delta, lty=kappa, col=kappa)) +
   scale_x_log10("Relative mean generation interval, $\\bar{G}_{\\mathrm{var}}/\\bar{G}_{\\mathrm{wt}}$",
                 breaks=c(0.67, 1, 1.5)) +
@@ -62,6 +63,8 @@ g1 <- ggplot(deltadata) +
   theme(
     panel.grid = element_blank()
   )
+
+
 
 tikz(file = "relspeed.tex", width = 6*1.2, height = 4*1.2, standAlone = T)
 g1

### 2021-04-05 relstrength.R

@@ -52,7 +52,7 @@ g1 <- ggplot(rhodata) +
   geom_line(aes(Gratio, rho, lty=kappa, col=kappa)) +
   scale_x_log10("Relative mean generation interval, $\\bar{G}_{\\mathrm{var}}/\\bar{G}_{\\mathrm{wt}}$",
                 breaks=c(0.67, 1, 1.5)) +
-  scale_y_continuous("Relative strength, $\\rho$") +
+  scale_y_log10("Relative strength, $\\rho$", breaks=c(1, 2, 4), limits=c(1, 4)) +
   facet_wrap(~rw) +
   scale_color_colorblind("$\\kappa$") +
   scale_linetype_discrete("$\\kappa$") +

### 2021-04-05 Rtbias.R
forgot to commit

@@ -56,7 +56,8 @@ for (i in 1:length(svec)) {
     geom_line(aes(tvec, Rtest1, color="Wild type", lty="Estimated"), lwd=1) +
     geom_line(aes(tvec, Rtest2, color="Variant", lty="Estimated"), lwd=1) +
     scale_x_continuous("Time (days)", expand=c(0, 0), limits=c(15, 70)) +
-    scale_y_continuous("Reproduction number, $\\mathcal{R}(t)$", limits=c(0.25, 7)) +
+    scale_y_log10("Reproduction number, $\\mathcal{R}(t)$", limits=c(0.25, 8),
+                  breaks=c(0.25, 0.5, 1, 2, 4, 8)) +
     scale_color_manual(values=c("red", "black")) +
     scale_linetype_manual(values=c(2, 1), guide=FALSE) +
     ggtitle(LETTERS[(i-1)*3+1]) +
@@ -77,7 +78,8 @@ for (i in 1:length(svec)) {
       geom_line(aes(tvec, Rtest1, lty="Estimated"), col="black", lwd=1) +
       geom_line(aes(tvec, Rtest2, col="Estimated", lty="Estimated"), lwd=1) +
       scale_x_continuous("Time (days)", expand=c(0, 0), limits=c(15, 70)) +
-      scale_y_continuous("Reproduction number, $\\mathcal{R}(t)$", limits=c(0.25, 7)) +
+      scale_y_log10("Reproduction number, $\\mathcal{R}(t)$", limits=c(0.25, 8),
+                    breaks=c(0.25, 0.5, 1, 2, 4, 8)) +
       scale_color_manual("a", values=c("red", "red")) +
       scale_linetype_manual("a", values=c(2, 1)) +
       ggtitle(LETTERS[(i-1)*3+1]) +
@@ -92,7 +94,7 @@ for (i in 1:length(svec)) {
     geom_line(aes(tvec, Rt2/Rt1, col="True", lty="True"), lwd=1) +
     geom_line(aes(tvec, Rtest2/Rtest1, col="Estimated", lty="Estimated"), lwd=1) +
     scale_x_continuous("Time (days)", expand=c(0, 0), limits=c(15, 70)) +
-    scale_y_continuous("Relative strength, $\\rho(t)$", limits=c(1, 2.5)) +
+    scale_y_log10("Relative strength, $\\rho(t)$", limits=c(1, 2.5)) +
[…]

### 2021-04-05 relspeed.R

@@ -52,7 +52,6 @@ deltadata %>%
   filter(Gratio==1, kappa==0.5)
 
 g1 <- ggplot(deltadata) +
-  geom_hline(data=deltadata2, aes(yintercept=delta)) +
   geom_line(aes(Gratio, delta, lty=kappa, col=kappa)) +
   scale_x_log10("Relative mean generation interval, $\\bar{G}_{\\mathrm{var}}/\\bar{G}_{\\mathrm{wt}}$",
                 breaks=c(0.67, 1, 1.5)) +
@@ -63,8 +62,6 @@ g1 <- ggplot(deltadata) +
   theme(
     panel.grid = element_blank()
   )
-
-
 
 tikz(file = "relspeed.tex", width = 6*1.2, height = 4*1.2, standAlone = T)
 g1

### 2021-04-08 newvariant.tex
edits

@@
 
+Inferring relative strength (i.e., ratio between reproduction numbers, $\Rv/\Rw$) and relative speed (i.e., difference between growth rate, $\rv-\rw$) of new SARS-CoV-2 variants compared to their wild type is critical to predicting the course of the current pandemic and controlling it.
+Many studies have already estimated the relative strength of new variants from their relative speed, but they typically neglect the possibility that the new variants may have different generation intervals (i.e., time between infection and transmission), which determines the relationship between relative strength and speed.
+However, a recent study suggests that B.1.1.7 strain may have longer infectious period, meaning that some variants, including B.1.1.7, may be able to transmit for a longer period of time and have different generation intervals from the wild type.
+Here, we explore how differences in generation intervals between a new variant and the wild type affects the relationship between relative strength and speed.
+We use simulations to show how neglecting their differences can lead to biases and illustrate how such biases can be assessed in practice.
+Finally, we discuss implications for control: if new variants have longer generation intervals, speed-like interventions (e.g., contact tracing) will be more relatively effective than strength-like interventions (e.g., social distancing), and therefore should be prioritized (and vice versa).
+
 \section{Introduction}
@@
 
-As our baseline scenario, we take the relative speed of the variant to be $\delta = 0.1/\textrm{day}$ based on the observed relaitve growth rate of B.1.1.7 \citep{davies2021estimated}; the mean generation interval of the wild type to be $\bar{G}_w = 5\days$ \citep{ferretti2020quantifying}; and the squared coefficient of variation of generation intervals to be $\kappa=0.2$ \citep{ferretti2020quantifying}.
+As our baseline scenario, we take the relative speed of the variant to be $\delta = 0.1/\textrm{day}$ based on the observed relative growth rate of B.1.1.7 \citep{davies2021estimated}; the mean generation interval of the wild type to be $\bar{G}_w = 5\days$ \citep{ferretti2020quantifying}; and the squared coefficient of variation of generation intervals to be $\kappa=0.2$ \citep{ferretti2020quantifying}.
 We use relative speed of B.1.1.7 as an example but note that the qualitative conclusions should hold for other variants as well.
@@
 We show that longer generation intervals generally lead to higher relative strength for a given relative speed (and conversely, lower relative speed for a given relative strength); these can be reversed when incidence is decreasing.e., $\Rv > 1$ and $\rv > 0$).
-We also that uncertainty about the relative mean generation interval of the variant has important implications for estimates of relative strength of the variantj.
+We also that uncertainty about the relative mean generation interval of the variant has important implications for estimates of relative strength of the variant.
 We illustrate how these biases might be assessed and discuss how differences in generation intervals might lead to different conclusions about the effectiveness of interventions:

### 2021-04-11 newvariant.tex
edits

--- 2021-04-05 newvariant.tex
@@
 
+Inferring relative strength (i.e., ratio between reproduction numbers, $\Rv/\Rw$) and relative speed (i.e., difference between growth rate, $\rv-\rw$) of new SARS-CoV-2 variants compared to their wild type is critical to predicting and controlling the course of the current pandemic.
+Multiple studies have estimated the relative strength of new variants from their relative speed, but they typically neglect the possibility that the new variants have different generation intervals (i.e., time between infection and transmission), which determines the relationship between relative strength and speed.
+However, the B.1.1.7 strain may have a longer infectious period than prior dominant lineages.
+Here, we explore how differences in generation intervals between a new variant and the wild type affects the relationship between relative strength and speed.
+We use simulations to show how neglecting their differences can lead to biases and illustrate how such biases can be assessed in practice.
+Finally, we discuss implications for control: if new variants have longer generation intervals then speed-like interventions (e.g., contact tracing) will be more relatively effective than strength-like interventions (e.g., social distancing), and therefore should be prioritized.
+
 \section{Introduction}
 
-Since the emergence of SARS-CoV-2 variants of concern (VoCs), estimating variant epidemic strength and speed remains one of the key questions in controlling its spread \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
-Epidemic strength, characterized by the reproduction number $\RR$, provides information about the final size of the epidemic as well as the amount of intervention required to eliminate the disease \citep{anderson1991infectious}.
-Epidemic speed, characterized by the growth rate $r$, provides information about the population-level time scale of the epidemic and, potentially when a new variant will replace the existing variant as the dominant strain.
-Strength and speed are linked by generation intervals---defined as time between infection and transmission---which characterizes individual-level time scale of the epidemic \citep{roberts2007model,svensson2007note,wallinga2007generation}.
+Estimating variant epidemic strength and speed remains one of the key questions in controlling the spread of SARS-CoV-2 variants of concerns (VoCs) \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
+Epidemic strength is characterized by the reproduction number $\RR$, which determines the number of new infections per average infection as well as the ``strength'' of intervention required to eliminate the disease \citep{anderson1991infectious}.
+It also provides information about the final size of an epidemic under the homogeneous assumption \citep{anderson1991infectious}.
+Epidemic speed is characterized by the growth rate $r$, which provides information about the population-level time scale of the epidemic.
+Like epidemic strength, epidemic speed also determines conditions for disease elimination, where $r=0$ is a threshold like $\RR=1$ \citep{doi:10.1098/rspb.2020.1556}.
+Strength and speed are linked by generation intervals---defined as time between infection and transmission---which characterizes the individual-level time scale of the epidemic \citep{roberts2007model,svensson2007note,wallinga2007generation}.
 
-Many studies have tried to estimate the strength of the novel variant from the observed speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021}.
-In contrast, some studies have assumed a value of the strength of the novel variant and tried to predict its future spread (therefore, the speed) of the variant \citep{davies2021estimated}.
+Analyses of new variants have prioritized characterizing \emph{relative} strength (ratios between reproduction numbers $\Rv/\Rw$) and speed (differences between growth rates $\rv-\rw$) of the variants.
+Many studies have tried to estimate the relative strength of the novel variant from the observed relative speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021}.
+In contrast, some studies have assumed a value of the relative strength of the novel variant and tried to predict its relative speed to determine if and when a new variant will become the dominant strain.
+\citep{davies2021estimated}.
 While both approaches are reasonable in principle, holding different quantities constant (i.e., strength or speed) can lead to different conclusions, in practice, about the spread of the disease, and therefore, its control \citep{doi:10.1098/rspb.2020.1556}.
 The reason is that inference of speed from strength and strength from speed depends on assumptions about the generation-interval distribution \citep{roberts2007model,svensson2007note,wallinga2007generation}.
-For example, when we assume a fixed value of epidemic strength, longer generation intervals lead to faster epidemic (lower $r$), making the epidemic seem easier to control.
+For example, when we assume a fixed value of epidemic strength, longer generation intervals are associated with a slower epidemic (lower $r$), making the epidemic seem easier to control.
 When we assume a fixed value of epidemic speed, longer generation intervals lead to stronger epidemic (higher $\RR$), making the epidemic seem more difficult to control.
 
-In addition, most studies have assumed that the generation-interval distributions do not differ between the existing strain and the variant, but recent evidence suggests that the B.1.1.7 variant may have a longer duration of infection: 13.3 days (90\% CI: 10.1--16.5) for the new variant and 8.2 days (90\% CI: 6.5--9.7) for the wild type \citep{kissler2021densely}.
+In practice, most studies have assumed that the generation-interval distributions do not differ between the existing strain and the variant, but recent evidence suggests that the B.1.1.7 variant may have a longer duration of infection: 13.3 days (90\% CI: 10.1--16.5) for the new variant and 8.2 days (90\% CI: 6.5--9.7) for the previously dominant variant, which we refer to as wild type \citep{kissler2021densely}.
 This implies that the mean generation interval of B.1.1.7 is likely to be longer than that of the wild type.
@@
 In general, if the generation-interval distribution of a new variant is different from the generation-interval distribution of the wild type, estimates of variant strength may be biased.
-And the story may be complicated: generation intervals depend on many other biological and behavioral factors, including for example self-isolation after symptom onset leading to shorter generation intervals.
+However, linking strength and speed is complicated given that generation intervals depend on many other biological and behavioral factors, including for example self-isolation after symptom onset leading to shorter generation intervals.
 We also do not know whether duration of infection of other variants will be same or different among each other.
+\swp{JSW wanted to remove the last sentence but I feel like need a sentence that alludes to other variants.}
 
-Here, we explore how different assumptions about the relative generation time of the new variant affect estimates of its strength $\RR$ and speed $r$.
+Here, we explore how different assumptions affect estimates of the relative affect estimates of its strength $\RR$ and speed $r$ for a new variant (e.g., B.1.1.7).
 We do this by exploring the relationship between relative strength (the ratio $\Rv/\Rw$) and relative speed (the difference $\rv-\rw$) under a wide range of assumptions about generation-interval distributions.
-We find that neglecting differences in the generation-interval distributions can lead to biased estimates, which, in turn, influence mitigation priorities.
-We also discuss how such biases might be assessed in practice.
+We find that neglecting differences in the generation-interval distributions can lead to biased estimates.
+We also discuss how such biases might be assessed in practice and how information on differences in generation interval distributions might also influence priorities for controlling the spread of VoCs.
 
@@
 this can be challenging when case counts are low (as when a new variant begins to spread) and is sensitive to temporal changes in testing patterns and intensity.
-Studies of new SARS-CoV-2 variants have mostly focused on characterizing changes in \emph{proportion} of a new variant \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
-This has the advantage of being less sensitive to changes in testing and to other transient effects that might affect variants and wild type viruses similarly.
+Studies of new SARS-CoV-2 variants have mostly focused on characterizing changes in the \emph{proportion} of a new variant \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
+Assessing frequency dynamics has the advantage of being less sensitive to changes in testing and to other transient effects that might affect variants and wild type viruses similarly.
 When incidence is changing exponentially ($i_x(t) = i_x(t_0) \exp(r_x t)$), the proportion of the new variant $p(t)$ follows a logistic equation:
@@
 \textbf{Relative strength of the new variant given observed relative speed under four epidemiological conditions.}
-The relative strength of the new variant $\rho$ given speed of the wild type $\rw$, ratio between the mean generation interval of the new variant $\bar{G}_v$ and that of the wild type $\bar{G}_w$, and squared coefficient of variation in generation intervals $\kappa$.
-Relative strength of the new variant $\rho$ is calculated using $\delta=0.1\pday$, $\bar{G}_w = 5\days$, and $\kappa = 1/5$.
+The relative strength of the new variant $\rho$ given speed of the wild type $\rw$, ratio between the mean generation interval of the new variant $\Gv$ and that of the wild type $\Gw$, and squared coefficient of variation in generation intervals $\kappa$.
+Relative strength of the new variant $\rho$ is calculated using $\delta=0.1\pday$, $\Gw = 5\days$, and $\kappa = 1/5$.
 Four epidemiological conditions assume $\rw=-0.15\pday$ ($\rw < \rv < 0$), $\rw=-0.1\pday$ ($\rw < \rv = 0$), $\rw=0\pday$ ($0 = \rw < \rv$), and $\rw=0.1\pday$ ($0 < \rw < \rv$).
@@
 This inference will depend on assumptions about the generation-interval distributions of both strains.
-These have generally been assumed to be the same, but we will also consider the possibility that they are different.
-Then, given the mean generation interval of the variant $\Gv$ and the wild type $\Gw$, the relative strength $\rho = \Rv/\Rw$ under the gamma approximation is given by \citep{park2019practical}:
+Given the mean generation interval of the variant $\Gv$ and the wild type $\Gw$, the relative strength $\rho = \Rv/\Rw$ under the gamma approximation is given by \citep{park2019practical}:
 \begin{equation}
-\rho = \left(\frac{1 + \kappa (\rw + \delta) \bar{G}_v}{1 + \kappa \rw \bar{G}_w}\right)^{1/\kappa}.
+\rho = \left(\frac{1 + \kappa (\rw + \delta) \Gv}{1 + \kappa \rw \Gw}\right)^{1/\kappa}.
 \end{equation}
@@
 
-As our baseline scenario, we take the relative speed of the variant to be $\delta = 0.1/\textrm{day}$ based on the observed relaitve growth rate of B.1.1.7 \citep{davies2021estimated}; the mean generation interval of the wild type to be $\bar{G}_w = 5\days$ \citep{ferretti2020quantifying}; and the squared coefficient of variation of generation intervals to be $\kappa=0.2$ \citep{ferretti2020quantifying}.
+As our baseline scenario, we take the relative speed of the variant to be $\delta = 0.1/\textrm{day}$ based on the observed relative growth rate of B.1.1.7 \citep{davies2021estimated}; the mean generation interval of the wild type to be $\bar{G}_w = 5\days$ \citep{ferretti2020quantifying}; and the squared coefficient of variation of generation intervals to be $\kappa=0.2$ \citep{ferretti2020quantifying}.
 We use relative speed of B.1.1.7 as an example but note that the qualitative conclusions should hold for other variants as well.
-For simplicity, we assume that the variant and the wild type have equal $\kappa$ throughout, and only consider differences in the mean.
-We evaluate the estimates of relative strength $\rho$ across a wide range of $\kappa$, ranging from 0 (delta distribution) to 1 (exponential distribution).
-To further explore how inference depends underlying epidemiological conditions (i.e., whether the variant and the wild type are growing or decaying), we consider four epidemiological conditions: (1) $\rw < \rv < 0$, (2) $\rw < \rv = 0$, (3) $0 = \rw < \rv$, and (4) $0 < \rw < \rv$.
+As noted above, we assume that the variant and the wild type have equal $\kappa$ throughout, and only consider differences in the mean.
+We evaluate the estimates of relative strength $\rho$ across a wide range of $\kappa$ from 0 (delta distribution) to 1 (exponential distribution).
+To further explore how inference depends on underlying epidemiological conditions (i.e., whether the variant and the wild type are growing or decaying), we consider four epidemiological conditions: (1) $\rw < \rv < 0$, (2) $\rw < \rv = 0$, (3) $0 = \rw < \rv$, and (4) $0 < \rw < \rv$.
 
@@
 Indeed, many biological mechanisms better translate to assuming a constant value of relative strength $\rho$ over changing conditions.
-For example, if the proportion of the population susceptible declines, or the average contact rate changes, while other factors remain constant, the relative strength $\rho$ is expected to remain relatively constant;
+For example, if the proportion of the population susceptible declines, or the average contact rate changes, while other factors remain constant, the relative strength $\rho$ is expected to remain constant;
 in general, this will imply a change in relative speed $\delta$.
@@
 
-In practice, epidemiological conditions change over time due to intervention measures and susceptible depletion.
-Such changes can be characterized by ``time-varying'' or ``instantaneous'' reproduction numbers, $\RR(t)$ \citep{fraser2007estimating}.
-Then, we can try to estimate the relative strength of the variant by comparing the ratio between estimates of reproduction numbers over time \citep{volz2021transmission}.
-
+Instead of estimating relative strength from speed, it is possible to directly estimate ``time-varying'' or ``instantaneous'' reproduction numbers of the variant and the wild type from incidence data, and compare their ratios \citep{fraser2007estimating};
+such methods have been used in previous analyses of the B.1.1.7 strain by \cite{volz2021transmission}.
 Assuming that the generation-interval distribution does not change over time, the instantaneous reproduction numbers of the new variant and of the wild type can be estimated from their corresponding incidence curves---an approach popularized by \cite{cori2013new}:
@@
 \end{equation}
-Under strength-like intervention measures that reduce transmission rates by a constant amount, we expect ratios between true reproduction numbers to remain constant and correspond to the true relative strength: $\Rv(t)/\Rw(t) = \rho$.
+Under strength-like intervention measures that reduce transmission rates by a constant amount, we expect ratios between reproduction numbers to remain constant and correspond to the true relative strength: $\Rv(t)/\Rw(t) = \rho$.
 However, if the assumed generation-interval distribution $\hat{g}(\tau)$ is different from the true distribution, then the ratio between the estimated reproduction numbers $\hat{\rho}(t) = \hat{\RR}_{\textrm{var}}(t)/\hat{\RR}_{\textrm{wt}}(t)$ may change, even if the true ratio does not.
@@
 
-While relative speed $\delta$ and strength $\rho$ are useful for characterizing the spread of the variant, \emph{absolute} speed $\rv$ and strength $\Rv$ of the variant are the main variables that determine its spread and control.
+While relative speed $\delta$ and strength $\rho$ are useful for characterizing the spread of the variant in an epidemiological context with a previously dominant wild type, \emph{absolute} speed $\rv$ and strength $\Rv$ of the variant are the main variables that determine the spread and control of the variant over the long term.
 In particular, at any given point in epidemic, we can measure the speed of the variant $\rv$ (or infer $\rv$ from $\rw$ and $\delta$) and ask how much more intervention is required to control the spread of both strains (since $\Rv < 1$ implies $\Rw < 1$).
@@
 
-Here, we explored how the generation-interval distribution shapes the link between relative strength and speed of the new variant.
-We define relative strength as the ratio between reproduction numbers ($\rho=\Rv/\Rw$) and relative speed as the difference between growth rates ($\delta=\rv-\rw$) of the variant and the wild type.
-We show that longer generation intervals generally lead to higher relative strength for a given relative speed (and conversely, lower relative speed for a given relative strength); these can be reversed when incidence is decreasing.e., $\Rv > 1$ and $\rv > 0$).
-We also that uncertainty about the relative mean generation interval of the variant has important implications for estimates of relative strength of the variantj.
-We illustrate how these biases might be assessed and discuss how differences in generation intervals might lead to different conclusions about the effectiveness of interventions:
+We explored how the generation-interval distribution shapes the link between relative strength and speed of the new variant.
+We showed that longer generation intervals generally lead to higher relative strength for a given relative speed (and conversely, lower relative speed for a given relative strength); these can be reversed when incidence is decreasing.
+We also found that uncertainty about the relative mean generation interval of the variant has important implications for estimates of relative strength of the variant.
+We illustrated how these biases might be assessed and discussed how differences in generation intervals might lead to different conclusions about the effectiveness of interventions:
 if the variant has longer (respectively, shorter) generation intervals than the wild type, speed-like (strength-like) interventions, will be more relatively more effective than naive estimates would suggest.
@@
 As the new variants, in particular the B.1.1.7 variant, of SARS-CoV-2 are spreading and becoming the dominant strain in many countries, it is clear that the variants are more transmissible (have higher strength) than the wild type \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
-Our analysis also supports higher strength of the new variant across a wide range of epidemiological conditions and assumptions about generation-interval distributions.
-However, in order to obtain precise and accurate estimates of the relative strength, uncertainty in generation-interval distributions must be taken into account.
-If the variant has longer infectious period \citep{kissler2021densely}, and therefore generation intervals, current estimates may be underestimating the true relative strength.
-
+Our analysis supports estimates of a higher strength of the new variant across a wide range of epidemiological conditions and assumptions about generation-interval distributions.
+However, our analysis also shows how uncertainty in generation-interval distributions must be taken into account to obtain accurate estimates of the relative strength of variants.
+If the variant has longer infectious period \citep{kissler2021densely}, and therefore extended generation-interval distributions, current estimates may be underestimating the true relative strength.
 While previous analyses have typically focused on estimating the relative strength and speed, absolute strength and speed are needed to properly evaluate its spread and control.
@@
 In doing so, we assumed that the incidence of infection caused by the variant and the wild type is known.
-However, estimating $\RR(t)$ from real data, such as case reports, is often associated with several practical challenges, such as delays between infection and case reporting and changes in testing patterns.
-For simplicity, we choose not to present any data analysis in this paper.
-Nonetheless, future studies may consider methods outlined here to assess potential biases in their estimates of relative strength.
+However, estimating $\RR(t)$ from real data, such as case incidence, is often associated with several practical challenges, such as delays between infection and reporting and changes in testing patterns.
+We have chosen to focus here on the underlying dynamical mechanisms that may impact inference.
+Future studies may consider the development of methods based on mechanism outlined here to assess potential biases in their estimates of relative strength.
 
-We argue that both perspectives (i.e., whether we hold relative strength or speed constant) are useful in understanding the dynamics of the new variant.
-Early in the spread, we typically want to fix the relative speed because we can directly observe it.
-As more information about the transmission and immunity profiles of the new variant becomes available, it is better to fix the relative strength as it better matches biological mechanisms for its higher strength.
+In moving from mechanism to improved estimates, we argue that both perspectives (i.e., whether we hold relative strength or speed constant) are useful in understanding the dynamics of the new variant.
+Early in the spread of an emerging variant, it is likely more convenient to fix the relative speed, given that speed can be directly observed.
+As more information about the transmission and immunity profiles of the new variant becomes available, then we advise fixing the relative strength and inferring the speed as it better matches biological mechanisms for its higher strength.
 We suggest researchers to be mindful about which quantity they hold constant and how their assumptions lead to their conclusions \citep{doi:10.1098/rspb.2020.1556}.
@@
 
+The spread of new SARS-CoV-2 variants and the replacement of previously dominant lineages represent ongoing challenges for controlling the SARS-CoV-2 pandemic.  
+By explicitly considering epidemiological context and generation interval differences together, we have also shown that improving estimates of the the relative duration of infectiousness at the individual scale may represent a pathway towards more effective interventions.
+Specifically, we have shown that
+
 \bibliography{newvariant.bib}

### 2021-04-11 newvariant.tex
a bit more fixes

@@
 
-As our baseline scenario, we take the relative speed of the variant to be $\delta = 0.1/\textrm{day}$ based on the observed relative growth rate of B.1.1.7 \citep{davies2021estimated}; the mean generation interval of the wild type to be $\bar{G}_w = 5\days$ \citep{ferretti2020quantifying}; and the squared coefficient of variation of generation intervals to be $\kappa=0.2$ \citep{ferretti2020quantifying}.
+As our baseline scenario, we take the relative speed of the variant to be $\delta = 0.1/\textrm{day}$ based on the observed relative growth rate of B.1.1.7 \citep{davies2021estimated}; the mean generation interval of the wild type to be $\Gw = 5\days$ \citep{ferretti2020quantifying}; and the squared coefficient of variation of generation intervals to be $\kappa=0.2$ \citep{ferretti2020quantifying}.
 We use relative speed of B.1.1.7 as an example but note that the qualitative conclusions should hold for other variants as well.
@@
 The relative speed of the new variant $\delta$ given strength of the wild type $\Rw$, ratio between the mean generation interval of the new variant $\Gv$ and that of the wild type $\Gw$, and squared coefficient of variation in generation intervals $\kappa$.
-Relative speed of the new variant $\delta$ is calculated using $\rho=1.61$, $\bar{G}_w = 5\days$, and $\kappa = 1/5$.
+Relative speed of the new variant $\delta$ is calculated using $\rho=1.61$, $\Gw = 5\days$, and $\kappa = 1/5$.
 Four epidemiological conditions assume $\Rw=0.5/\rho$ ($\Rw < \Rv < 1$), $\Rw=1/\rho$ ($\Rw < \Rv = 1$), $\Rw=1$ ($1 = \Rw < \Rv$), and $\Rw=\rho$ ($0 < \Rw < \Rv$).

### 2021-04-14 newvariant.tex
various edits from Ben

@@
 \usepackage{natbib}
-\bibliographystyle{unsrt}
+\bibliographystyle{chicago}
 \date{\today}
@@
 
-Inferring relative strength (i.e., ratio between reproduction numbers, $\Rv/\Rw$) and relative speed (i.e., difference between growth rate, $\rv-\rw$) of new SARS-CoV-2 variants compared to their wild type is critical to predicting and controlling the course of the current pandemic.
-Multiple studies have estimated the relative strength of new variants from their relative speed, but they typically neglect the possibility that the new variants have different generation intervals (i.e., time between infection and transmission), which determines the relationship between relative strength and speed.
-However, the B.1.1.7 strain may have a longer infectious period than prior dominant lineages.
-Here, we explore how differences in generation intervals between a new variant and the wild type affects the relationship between relative strength and speed.
+Inferring the relative strength (i.e., the ratio of reproduction numbers, $\Rv/\Rw$) and relative speed (i.e., the difference between growth rate, $\rv-\rw$) of new SARS-CoV-2 variants compared to their wild types is critical to predicting and controlling the course of the current pandemic.
+Multiple studies have estimated the relative strength of new variants from the observed relative speed, but they typically neglect the possibility that the new variants have different generation intervals (i.e., time between infection and transmission), which determines the relationship between relative strength and speed.
+However, the B.1.1.7 strain may have a longer infectious period (and therefore, generation interval) than prior dominant lineages.
+Here, we explore how differences in generation intervals between a new variant and the wild type affect the relationship between relative strength and speed.
 We use simulations to show how neglecting their differences can lead to biases and illustrate how such biases can be assessed in practice.
-Finally, we discuss implications for control: if new variants have longer generation intervals then speed-like interventions (e.g., contact tracing) will be more relatively effective than strength-like interventions (e.g., social distancing), and therefore should be prioritized.
+Finally, we discuss implications for control: if new variants have longer generation intervals then speed-like interventions (e.g., contact tracing) will be more relatively effective than strength-like interventions (e.g., social distancing).
 
@@
 
-Estimating variant epidemic strength and speed remains one of the key questions in controlling the spread of SARS-CoV-2 variants of concerns (VoCs) \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
-Epidemic strength is characterized by the reproduction number $\RR$, which determines the number of new infections per average infection as well as the ``strength'' of intervention required to eliminate the disease \citep{anderson1991infectious}.
-It also provides information about the final size of an epidemic under the homogeneous assumption \citep{anderson1991infectious}.
-Epidemic speed is characterized by the growth rate $r$, which provides information about the population-level time scale of the epidemic.
-Like epidemic strength, epidemic speed also determines conditions for disease elimination, where $r=0$ is a threshold like $\RR=1$ \citep{doi:10.1098/rspb.2020.1556}.
+Estimating variant epidemic strength and speed remains one of the key questions in controlling the spread of SARS-CoV-2 variants of concern (VoCs) \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
+Epidemic strength is characterized by the reproduction number $\RR$, which determines the average number of new infections per infection as well as the strength of intervention required to eliminate the disease \citep{anderson1991infectious}.
+It also provides information about the final size of an epidemic in a homogeneous population \citep{anderson1991infectious}.
+Epidemic speed is characterized by the growth rate $r$, which describes how fast a disease spreads at the population-level.
+Like epidemic strength, epidemic speed also determines conditions for disease eliminationl $r=0$ is a threshold equivalent to $\RR=1$ \citep{doi:10.1098/rspb.2020.1556}.
 Strength and speed are linked by generation intervals---defined as time between infection and transmission---which characterizes the individual-level time scale of the epidemic \citep{roberts2007model,svensson2007note,wallinga2007generation}.
 
-Analyses of new variants have prioritized characterizing \emph{relative} strength (ratios between reproduction numbers $\Rv/\Rw$) and speed (differences between growth rates $\rv-\rw$) of the variants.
+Analyses of new variants have typically characterized \emph{relative} strength (ratios between reproduction numbers $\Rv/\Rw$) and speed (differences between growth rates $\rv-\rw$) of the variants.
 Many studies have tried to estimate the relative strength of the novel variant from the observed relative speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021}.
-In contrast, some studies have assumed a value of the relative strength of the novel variant and tried to predict its relative speed to determine if and when a new variant will become the dominant strain.
+In contrast, some studies have assumed a value of the relative strength of the novel variant and tried to predict its relative speed to determine when a new variant will become the dominant strain.
 \citep{davies2021estimated}.
-While both approaches are reasonable in principle, holding different quantities constant (i.e., strength or speed) can lead to different conclusions, in practice, about the spread of the disease, and therefore, its control \citep{doi:10.1098/rspb.2020.1556}.
-The reason is that inference of speed from strength and strength from speed depends on assumptions about the generation-interval distribution \citep{roberts2007model,svensson2007note,wallinga2007generation}.
-For example, when we assume a fixed value of epidemic strength, longer generation intervals are associated with a slower epidemic (lower $r$), making the epidemic seem easier to control.
+While both approaches are reasonable in principle, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the disease and its control \citep{doi:10.1098/rspb.2020.1556}:
+estimation of speed from strength or strength from speed depends on assumptions about the generation-interval distribution \citep{roberts2007model,svensson2007note,wallinga2007generation}.
+For example, when we assume a fixed value of epidemic strength, longer generation intervals lead to a slower epidemic (lower $r$), making the epidemic seem easier to control.
 When we assume a fixed value of epidemic speed, longer generation intervals lead to stronger epidemic (higher $\RR$), making the epidemic seem more difficult to control.
 
-In practice, most studies have assumed that the generation-interval distributions do not differ between the existing strain and the variant, but recent evidence suggests that the B.1.1.7 variant may have a longer duration of infection: 13.3 days (90\% CI: 10.1--16.5) for the new variant and 8.2 days (90\% CI: 6.5--9.7) for the previously dominant variant, which we refer to as wild type \citep{kissler2021densely}.
-This implies that the mean generation interval of B.1.1.7 is likely to be longer than that of the wild type.
+In practice, most studies have assumed that the previous dominant strain (wild type) and variants have identical generation intervals, but recent evidence suggests that the B.1.1.7 variant may have a longer duration of infection: 13.3 days (90\% CI: 10.1--16.5) for the new variant and 8.2 days (90\% CI: 6.5--9.7) for the wild type \citep{kissler2021densely}.
+Longer duration of infection implies that the mean generation interval of B.1.1.7 is likely to be longer than that of the wild type.
 Other studies have considered the possibility that the faster growth rate of new variants may be driven, in part, by shorter generation intervals \citep{davies2021estimated,volz2021transmission}.
 In general, if the generation-interval distribution of a new variant is different from the generation-interval distribution of the wild type, estimates of variant strength may be biased.
-However, linking strength and speed is complicated given that generation intervals depend on many other biological and behavioral factors, including for example self-isolation after symptom onset leading to shorter generation intervals.
-We also do not know whether duration of infection of other variants will be same or different among each other.
-\swp{JSW wanted to remove the last sentence but I feel like need a sentence that alludes to other variants.}
+However, linking strength and speed is complicated given that generation intervals depend on many other biological and behavioral factors;
+for example, self-isolation after symptom onset will lead to shorter generation intervals.
 
-Here, we explore how different assumptions affect estimates of the relative affect estimates of its strength $\RR$ and speed $r$ for a new variant (e.g., B.1.1.7).
-We do this by exploring the relationship between relative strength (the ratio $\Rv/\Rw$) and relative speed (the difference $\rv-\rw$) under a wide range of assumptions about generation-interval distributions.
+Here, we explore how different assumptions affect estimates of the relative strength $\RR$ and speed $r$ for a new variant (e.g., B.1.1.7).
+We compare the relationship between relative strength (the ratio $\Rv/\Rw$) and relative speed (the difference $\rv-\rw$) under a wide range of assumptions about generation-interval distributions.
 We find that neglecting differences in the generation-interval distributions can lead to biased estimates.
@@
 The integral of the kernel $\RR_x(t) = \int K_x(t, \tau) \dtau$ is referred to as the instantaneous reproduction number \citep{fraser2007estimating}.
-This is a particular kind of weighted average which describes the average infectiousness of previously infected individuals at time $t$---in particular, it is not weighted by the actual number of infected individuals present, but rather by their relative infectiousness at time $t$.
-The normalized kernel $g_x(t, \tau) = K_x(t, \tau)/\RR_x(t)$---which we refer to as the instantaneous generation-interval distribution---describes their relative contribution to current incidence $i_x(t)$ and provides information about the time scale of disease transmission.
-Both the reproduction number and the generation-interval distribution can depend on several factors, including intrinsic infectiousness of an infected individual, non-pharmaceutical interventions, awareness-driven behavior, and population-level susceptibility \citep{fraser2007estimating}.
+The instantaneous reproduction number is a particular kind of weighted average which describes the average infectiousness of previously infected individuals at time $t$---in particular, it is weighted by their relative infectiousness at time $t$, rather than by the actual number of infected individuals present.
+The normalized kernel $g_x(t, \tau) = K_x(t, \tau)/\RR_x(t)$---which we refer to as the instantaneous generation-interval distribution---describes the relative contribution of previously infected individuals to current incidence $i_x(t)$ and provides information about the time scale of disease transmission.
+Both the reproduction number and the generation-interval distribution depend on many factors, including intrinsic infectiousness of an infected individual, non-pharmaceutical interventions, awareness-driven behavior, and population-level susceptibility \citep{fraser2007estimating}.
 
@@
 \end{equation}
-We can approximate the $r$--$\RR$ relationship by assuming that the generation-interval distribution is approximately gamma-distributed, and summarizing it using the mean generation interval $\bar{G}_x$ and the squared coefficient of variation $\kappa_x$ \citep{park2019practical}:
+We can approximate the $r$--$\RR$ relationship by assuming that the generation-interval distribution is Gamma-distributed, and summarizing it using the mean generation interval $\bar{G}_x$ and the squared coefficient of variation $\kappa_x$ \citep{park2019practical}:
 \begin{equation}
@@
 \end{equation}
-The gamma assumption covers models that assume exponentially distributed generation intervals (when $\kappa=1$), corresponding to the SIR model \citep{anderson1991infectious}; various gamma assumptions and approximations are widely used in epidemic modeling, including for models of SARS-CoV-2 \citep{doi:10.1098/rsif.2020.0144}.
+The Gamma assumption includes as a special case models that assume exponentially distributed generation intervals (when $\kappa=1$), corresponding to the SIR model \citep{anderson1991infectious}; various gamma assumptions and approximations are widely used in epidemic modeling, including for models of SARS-CoV-2 \citep{doi:10.1098/rsif.2020.0144}.
 We use this framework to investigate how inferences about strength and speed of the variant depend on our assumptions about the underlying generation-interval distributions.
@@
 
-Epidemic speed $r_x$ can be often estimated directly from case data \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating};
-this can be challenging when case counts are low (as when a new variant begins to spread) and is sensitive to temporal changes in testing patterns and intensity.
+Epidemic speed $r_x$ can often be estimated directly from case data \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating};
+however, the estimating $r_x$ can be challenging when case counts are low (as when a new variant begins to spread) and is sensitive to temporal changes in testing patterns and intensity.
 Studies of new SARS-CoV-2 variants have mostly focused on characterizing changes in the \emph{proportion} of a new variant \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
-Assessing frequency dynamics has the advantage of being less sensitive to changes in testing and to other transient effects that might affect variants and wild type viruses similarly.
-When incidence is changing exponentially ($i_x(t) = i_x(t_0) \exp(r_x t)$), the proportion of the new variant $p(t)$ follows a logistic equation:
+Assessing frequency dynamics has the advantage of being less sensitive to changes in testing and to other transient effects that would affect variants and wild type viruses similarly.
+When incidence is changing exponentially ($i_x(t) = i_x(t_0) \exp(r_x t)$), the proportion of the new variant $p(t)$ follows a logistic growth curve:
 \begin{align}
@@
 \caption{
-\textbf{Relative strength of the new variant given observed relative speed under four epidemiological conditions.}
-The relative strength of the new variant $\rho$ given speed of the wild type $\rw$, ratio between the mean generation interval of the new variant $\Gv$ and that of the wild type $\Gw$, and squared coefficient of variation in generation intervals $\kappa$.
-Relative strength of the new variant $\rho$ is calculated using $\delta=0.1\pday$, $\Gw = 5\days$, and $\kappa = 1/5$.
+\textbf{Relative strength of the new variant conditional on observed relative speed under four epidemiological conditions.}
+The relative strength of the new variant $\rho$ conditional on the speed of the wild type $\rw$, the ratio between the mean generation interval of the new variant $\Gv$ and that of the wild type $\Gw$, and the squared coefficient of variation in generation intervals $\kappa$.
+The relative strength of the new variant $\rho$ is calculated using $\delta=0.1\pday$, $\Gw = 5\days$, and $\kappa = 1/5$.
 Four epidemiological conditions assume $\rw=-0.15\pday$ ($\rw < \rv < 0$), $\rw=-0.1\pday$ ($\rw < \rv = 0$), $\rw=0\pday$ ($0 = \rw < \rv$), and $\rw=0.1\pday$ ($0 < \rw < \rv$).
@@
 
-We thus ask: what factors affect the relative strength $\rho = \Rv/\Rw$ of a new variant, given an observed relative speed $\delta$?
-This inference will depend on assumptions about the generation-interval distributions of both strains.
+We thus ask: what factors affect the relative strength $\rho = \Rv/\Rw$ of a new variant, conditional on an observed relative speed $\delta$?
+Inference of $\rho$ from $\delta$ will depend on assumptions about the generation-interval distributions of both strains.
 Given the mean generation interval of the variant $\Gv$ and the wild type $\Gw$, the relative strength $\rho = \Rv/\Rw$ under the gamma approximation is given by \citep{park2019practical}:
@@
 \end{equation}
-Therefore, the relative strength $\rho$ depend not only on the relative speed $\delta$ and the generation-interval distributions but also on how fast the wild type is spreading in the population \rw---a few analyses have implicitly or explicitly neglected this factor by either assuming $\rw = 0$ \citep{switzerland2021variant} or $\kappa = 0$ \citep{davies2021estimated} (in the latter case $\hat{\rho} = \exp(\delta \bar{G})$).
+Therefore, the relative strength $\rho$ depends not only on the relative speed $\delta$ and the generation-interval distributions but also on how fast the wild type is spreading in the population (\rw)---
+some analyses have implicitly or explicitly neglected this factor by either assuming $\rw = 0$ \citep{switzerland2021variant} or $\kappa = 0$ \citep{davies2021estimated} (in the latter case $\hat{\rho} = \exp(\delta \bar{G})$).
 
-As our baseline scenario, we take the relative speed of the variant to be $\delta = 0.1/\textrm{day}$ based on the observed relative growth rate of B.1.1.7 \citep{davies2021estimated}; the mean generation interval of the wild type to be $\Gw = 5\days$ \citep{ferretti2020quantifying}; and the squared coefficient of variation of generation intervals to be $\kappa=0.2$ \citep{ferretti2020quantifying}.
-We use relative speed of B.1.1.7 as an example but note that the qualitative conclusions should hold for other variants as well.
+Based on the observed relative growth rate of B.1.1.7 in the UK, we start by  taking the relative speed of the variant to be $\delta = 0.1/\textrm{day}$  \citep{davies2021estimated}; the mean generation interval of the wild type to be $\Gw = 5\days$ \citep{ferretti2020quantifying}; and the squared coefficient of variation of generation intervals to be $\kappa=0.2$ \citep{ferretti2020quantifying}.
+We consider the spread of B.1.1.7 as an example but out qualitative conclusions should hold for other VoCs.
 As noted above, we assume that the variant and the wild type have equal $\kappa$ throughout, and only consider differences in the mean.
 We evaluate the estimates of relative strength $\rho$ across a wide range of $\kappa$ from 0 (delta distribution) to 1 (exponential distribution).
-To further explore how inference depends on underlying epidemiological conditions (i.e., whether the variant and the wild type are growing or decaying), we consider four epidemiological conditions: (1) $\rw < \rv < 0$, (2) $\rw < \rv = 0$, (3) $0 = \rw < \rv$, and (4) $0 < \rw < \rv$.
+To further explore how inference depends on underlying epidemiological conditions (i.e., whether the incidence of the variant and the wild type are growing or shrinking), we consider four epidemiological conditions: (1) $\rw < \rv < 0$, (2) $\rw < \rv = 0$, (3) $0 = \rw < \rv$, and (4) $0 < \rw < \rv$.
 
-\fref{relstrength} illustrates how the relative strength depends on the shape of the generation-interval distributions as well as underlying epidemiological conditions.
-In general, longer (or shorter) mean generation interval of the new variant translates to higher (or lower) values of $\rho$, except when $\rv \leq 0$.
-When $\rv = 0$ (and also $\rw < 0$), we always have $\Rv = 1$ and so the generation-interval distribution of the new variant does not affect the inference.
-When $\rv < 0$ (and also $\rw < 0$), we see that longer generation intervals decrease $\rho$ because longer generation intervals actually lead to slower decay (higher $r$).
-Assuming a narrower distribution (lower $\kappa$) has qualitatively similar effects as assuming longer generation intervals because they both reduce the amount of early transmission; 
+\fref{relstrength} illustrates how the relative strength depends on the shape of the generation-interval distributions and on underlying epidemiological conditions.
+In general, longer mean generation interval of the new variant translates to higher values of $\rho$ (and vice versa), except when $\rv \leq 0$.
+When $\rv = 0$ (and therefore $\rw < 0$), we always have $\Rv = 1$ and so the inferred value of $\rho$ is independent of the generation-interval distribution of the new variant.
+When $\rv < 0$ (and therefore $\rw < 0$), we see that longer generation intervals decrease $\rho$ because longer generation intervals actually lead to slower decay (higher $r$).
+Assuming a narrower distribution (lower $\kappa$) has qualitatively similar effects as assuming longer generation intervals because both reduce the amount of early transmission; 
 therefore, narrower distributions lead to higher values of $\rho$ when $\rv > 0$ and lower values of $\rho$ when $\rv \leq 0$.
@@
 However, the magnitude of relative strength $\rho$ is sensitive to assumptions about generation intervals.
-For realistic values of $\kappa$ (excluding 0 and 1), we find that the relative strength $\rho$ ranges between 1.1--2.3.
-Even if we restrict the mean generation interval of the variant to differ by 20\% and only consider the scenario under which variant is growing (bottom two panels of \fref{relstrength}), we find that $\rho$ ranges between 1.3--1.8, exhibiting narrower but still large uncertainty.
+For realistic values of $\kappa$ (excluding 0 and 1), we find that the inferred relative strength $\rho$ ranges between 1.1--2.3.
+Even if we restrict the mean generation interval of the variant to differ by 20\% and only consider the scenario under which variant is growing (bottom two panels of \fref{relstrength}), we find that $\rho$ ranges between 1.3--1.8, providing narrower but still large uncertainty.
 
@@
 
-We do not generally expect the relative speed $\delta$ to be remain constant if other factors governing epidemic spread are changing.
+We do not generally expect the relative speed $\delta$ to remain constant if other factors governing epidemic spread are changing.
 Indeed, many biological mechanisms better translate to assuming a constant value of relative strength $\rho$ over changing conditions.
@@
 
-We thus investigate how $\delta$ is expected to change with \Rw, if $\rho$ remains constant, and how this expectation changes with the ratio of the generation intervals. 
+We thus investigate how $\delta$ is expected to change with \Rw if $\rho$ remains constant, and how this expectation changes with the ratio of the generation intervals. 
 Once again, we rely on the gamma approximation to find the relative speed $\delta$ given the mean generation interval of the variant $\Gv$ and the wild type $\Gw$, :
@@
 As our baseline scenario we assume $\rho = 1.61$, which is the value we obtain for $\delta=0.1\,\textrm{days}$, $\Gw = \Gv = 5\,\textrm{days}$, and $\kappa = 1/5$.
-We also evaluate $\delta$ across four different epidemiological conditions as before: (1) $\Rw < \Rv < 1$, (2) $\Rw < \Rv = 1$, (3) $1 = \Rw < \Rv$, and (4) $1 < \Rw < \Rv$.
+We evaluate $\delta$ across four different epidemiological conditions as before: (1) $\Rw < \Rv < 1$, (2) $\Rw < \Rv = 1$, (3) $1 = \Rw < \Rv$, and (4) $1 < \Rw < \Rv$.
 
@@
 
-\fref{relspeed} compares how generation-interval distributions affect predictions about relative speed under different epidemiological conditions.
-In general, longer generation intervals lead to slower relative speed of the variant when the incidence of both strains are increasing (\fref{relspeed}, bottom panels) because slower growth of the variant reduces the differences in absolute speed.
+In general, longer generation intervals lead to slower relative speed of the variant when the incidence of both strains is increasing (\fref{relspeed}, bottom panels) because slower growth of the variant reduces the differences in absolute speed.
 When $\Rv=1$, the relative speed is insensitive to the generation-interval distribution of the variant because we always have $\rv=0$.
-When $\Rv<1$, longer generation intervals of the variant leads to slower decay ($\rv$ closer to 0), and therefore, greater relative speed.
+When $\Rv<1$, longer generation intervals of the variant lead to slower decay ($\rv$ closer to 0), and therefore, greater relative speed.
 Once again, we see that assuming a narrower distribution (lower $\kappa$) has qualitatively similar effects as assuming a longer mean.
 
-\fref{relspeed} also shows that relative speed depends on underlying epidemiological conditions. 
+\fref{relspeed} also shows that when $\rho$ is fixed relative speed depends on underlying epidemiological conditions. 
 For example, even when there are no differences in the generation-interval distributions ($\Gv=\Gw$, in this case), the relative speed $\delta$ can range between 0.08--0.11 when $\kappa=0.2$ and 0.06--0.14 when $\kappa=0.5$.
@@
 
-Instead of estimating relative strength from speed, it is possible to directly estimate ``time-varying'' or ``instantaneous'' reproduction numbers of the variant and the wild type from incidence data, and compare their ratios \citep{fraser2007estimating};
+Instead of estimating relative strength from speed, one can directly estimate time-varying or instantaneous reproduction numbers of the variant and the wild type from incidence data, and compare their ratios \citep{fraser2007estimating};
 such methods have been used in previous analyses of the B.1.1.7 strain by \cite{volz2021transmission}.
-Assuming that the generation-interval distribution does not change over time, the instantaneous reproduction numbers of the new variant and of the wild type can be estimated from their corresponding incidence curves---an approach popularized by \cite{cori2013new}:
+Assuming that the generation-interval distribution remains constant, the instantaneous reproduction numbers of the new variant and of the wild type can be estimated from their corresponding incidence curves---an approach popularized by \cite{cori2013new}:
 \begin{equation}
@@
 
-Here, we investigate under the assumption that the true generation-interval distribution of the wild type is known, and use a two-strain renewal equation which assumes perfect cross-immunity to simulate three different scenarios:
+Here, we investigate how misspecification of the generation-interval distribution of the variant affects our inference of relative strength from inference data under the assumption that the true generation-interval distribution of the wild type is known.
+We use a two-strain renewal equation that assumes perfect cross-immunity to simulate three different scenarios (see Methods):  
 (1) the wild type and the variant have the same generation-interval distributions (which match the known distribution, shown in \fref{Rtbias}A--C);
@@
 Then, we compare the estimated ratio $\hat{\rho}(t) = \hat{\RR}_{\textrm{var}}(t)/\hat{\RR}_{\textrm{wt}}(t)$ with the true ratio $\rho = \Rv(t)/\Rw(t)$.
-In order to reflect smooth changes in $\RR$ driven by non-pharmaceutical interventions and behavior changes, we model $\Rw(t)$ phenomenologically and let $\Rv(t) = \rho \Rw(t)$.
+In order to allow for smooth changes in $\RR$ driven by non-pharmaceutical interventions and behavior changes, we model $\Rw(t)$ phenomenologically (\fref{Rtbias}) and let $\Rv(t) = \rho \Rw(t)$.
 
@@
 
-When the assumed distribution matches the true distributions, the estimated reproduction numbers match the true values (\fref{Rtbias}A), and therefore the ratio remains constant (\fref{Rtbias}B).
-However, if the generation-interval distribution of the variant is different from the assumed distribution, the ratio changes over time (\fref{Rtbias}E,H).
-In particular, if the true generation intervals have a shorter (or longer) mean than the assumed distribution, we over-estimate (or under-estimate) $\Rv(t)$ during the growth phase and under-estimate (or over-estimate) during the decay phase (\fref{Rtbias}D,G), which further translates to biases in the estimated relative strength (\fref{Rtbias}F,I).
+When the assumed distribution matches the true distribution, the estimated reproduction numbers match the true values (\fref{Rtbias}A), and therefore the ratio remains constant (\fref{Rtbias}B).
+However, if the generation-interval distribution of the variant differs from the assumed distribution, the ratio changes over time (\fref{Rtbias}E,H).
+In particular, if the true generation intervals have a shorter mean than the assumed distribution, we over-estimate $\Rv(t)$ during the growth phase and under-estimate during the decay phase (and vice versa, \fref{Rtbias}D,G), which further translates to biases in the estimated relative strength (\fref{Rtbias}F,I).
 
@@
 When generation-interval distributions are correctly specified, we obtain a straight line with a slope of $\rho$ and intercept at zero (\fref{Rtbias}C).
-When the assumed mean generation interval is longer (or shorter) than the that of the variant, we over-estimate (or under-estimate) the slope (\fref{Rtbias}F,I).
+When the assumed mean generation interval is longer than the that of the variant, we over-estimate the slope (and vice versa, \fref{Rtbias}F,I).
 In addition, the intercept (if we were to perform regression) need not be at zero due to biases in the slope.
@@
 In particular, at any given point in epidemic, we can measure the speed of the variant $\rv$ (or infer $\rv$ from $\rw$ and $\delta$) and ask how much more intervention is required to control the spread of both strains (since $\Rv < 1$ implies $\Rw < 1$).
-As a baseline scenario, we assume $\rw=0$ and $\delta=0.1\pday$ (and therefore $\rv=0.1\pday$), in which case additional intervention is required to bring $\rv$ below 0 (or, equivalently, $\Rv$ below 1)
+As a baseline scenario, we assume $\rw=0$ and $\delta=0.1\pday$ (and therefore $\rv=0.1\pday$), in which case additional intervention is required to reduce $\rv$ below 0 (or, equivalently, $\Rv$ below 1).
 
 We consider two types of intervention:
-an intervention of constant ``strength'', which reduces transmission by a constant factor $\theta$ regardless of age of infection ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau)/\theta$); and an intervention of constant ``speed'', which reduces transmission after infection by a constant rate $\phi$ ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau) \exp(-\phi \tau)$).
+an intervention of constant strength, which reduces transmission by a constant factor $\theta$ regardless of age of infection ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau)/\theta$); and an intervention of constant speed, which reduces transmission after infection by a constant rate $\phi$ ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau) \exp(-\phi \tau)$).
 In this case, we can control the spread of the variant when $\theta > \Rv$ or $\phi > \rv$, respectively.
-In general, we can calculate the strength ($\theta = \Ry{pre}/\Ry{post}$) and speed ($\phi = \ry{pre} - \ry{post}$) of intervention for any intervention by comparing epidemiological conditions before and after the intervention \citep{doi:10.1098/rspb.2020.1556}---these quantities, by definition, must be greater than their corresponding pre-intervention strength and speed of the epidemic to control the spread.
+In general, we can calculate the strength ($\theta = \Ry{pre}/\Ry{post}$) and speed ($\phi = \ry{pre} - \ry{post}$) for any intervention by comparing epidemiological conditions before and after the intervention \citep{doi:10.1098/rspb.2020.1556}---these quantities, by definition, must be greater than the pre-intervention strength or speed of the epidemic in order to control the epidemic.
 Here, we use the strength-and-speed framework to compare how differences in generation-interval distributions would affect the control of the spread of a new variant.
@@
 The speed-based paradigm gives the same results about the control but provides important insights (\fref{strengthspeed}B).
-The observed speed of the variant at any given moment does not depend on our estimates of its mean generation interval.
+The observed speed of the variant at any given moment is independent of our estimates of its mean generation interval.
 Therefore, if we can reduce infectiousness at a faster rate than the observed speed of spread (i.e., if the speed of intervention is faster than the epidemic speed), we can control the epidemic---this explains why a constant-speed intervention can prevent the spread of the variant regardless of its mean generation interval.
@@
 
+\swp{Stylistic edits from Ben: remove ``we did this we did that''}
 We explored how the generation-interval distribution shapes the link between relative strength and speed of the new variant.
@@
 
-As the new variants, in particular the B.1.1.7 variant, of SARS-CoV-2 are spreading and becoming the dominant strain in many countries, it is clear that the variants are more transmissible (have higher strength) than the wild type \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
+As the new variants, in particular the B.1.1.7 variant, of SARS-CoV-2 are spreading and becoming dominant in many countries, it is clear that the variants are more transmissible (have higher strength) than the wild type \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
 Our analysis supports estimates of a higher strength of the new variant across a wide range of epidemiological conditions and assumptions about generation-interval distributions.
-However, our analysis also shows how uncertainty in generation-interval distributions must be taken into account to obtain accurate estimates of the relative strength of variants.
-If the variant has longer infectious period \citep{kissler2021densely}, and therefore extended generation-interval distributions, current estimates may be underestimating the true relative strength.
+However, our analysis also shows thatuncertainty in generation-interval distributions must be taken into account to obtain accurate estimates of the relative strength of variants.
+If the variant has a longer infectious period \citep{kissler2021densely}, and therefore extended generation-interval distributions, current estimates may be underestimating the true relative strength.
 While previous analyses have typically focused on estimating the relative strength and speed, absolute strength and speed are needed to properly evaluate its spread and control.
@@
 
-We used simulations to show how biases in relative strength from mis-specification of generation-interval distributions can be assessed from data:
+We used simulations to show how mis-specification of generation-interval distributions can bias the inference of relative strenth from incidence data:
 when the assumed generation-interval distribution does not match the true generation-interval distribution of the variant, estimates of absolute strength $\Rv(t)$ will be biased, causing the estimated relative strength to change over time.
@@
 Early in the spread of an emerging variant, it is likely more convenient to fix the relative speed, given that speed can be directly observed.
-As more information about the transmission and immunity profiles of the new variant becomes available, then we advise fixing the relative strength and inferring the speed as it better matches biological mechanisms for its higher strength.
-We suggest researchers to be mindful about which quantity they hold constant and how their assumptions lead to their conclusions \citep{doi:10.1098/rspb.2020.1556}.
+As more information about the transmission and immunity profiles of the new variant becomes available, we advise fixing the relative strength and inferring the speed as this assumption better matches biological mechanisms for its higher strength.
+Researchers should be mindful about which quantity they hold constant and how their assumptions lead to their conclusions \citep{doi:10.1098/rspb.2020.1556}.
 
 Even though SARS-CoV-2 has been spreading for more than a year, there is still considerable uncertainty in its generation intervals.
-Observational studies typically focus on serial intervals (i.e., time between symptom onset of the infector and the infectee; \cite{svensson2007note}) because they are easier to measure \citep{griffin2020rapid}, but they are subject to dynamical biases that can be difficult to tease apart \citep{park2021forward}.
-A few studies have tried to estimate the generation-interval distribution from serial intervals with means ranging between 3--6 days and squared coefficient of variations ranging between 0.1--0.6 \citep{ferretti2020quantifying,Ferretti2020timing,ganyani2020estimating,knight2020estimating};
-however, most estimates neglect dynamical biases in serial intervals.
+Observational studies typically focus on serial intervals (i.e., time between symptom onset of the infector and the infectee; \cite{svensson2007note}) because they are easier to measure \citep{griffin2020rapid}.
+Furthermore, realized serial intervals are subject to dynamical biases that can be difficult to tease apart \citep{park2021forward}.
+A few studies have tried to estimate the generation-interval distribution from serial intervals, with means ranging between 3--6 days and squared coefficients of variation ranging between 0.1--0.6 \citep{ferretti2020quantifying,Ferretti2020timing,ganyani2020estimating,knight2020estimating};
+however, most estimates neglect dynamical biases in serial intervals caused by underlying epidemiological dynamics in their estimates of generation intervals.
 In addition, serial intervals do not account for asymptomatic transmission, adding further uncertainty to inferences of speed and strength \citep{park2020time}.
@@
 symptomatic, \cite{park2020time}).
-These same lessons have relevance for assessing the spread of variants in a partially susceptible population and improving efforts to control spread.
-Future studies should prioritize detailed assessment of the relative generation intervals of SARS-CoV-2 and widespread variants, as well as evaluate how uncertainty in assumptions about generational intervals might bias conclusions.
+These same lessons should be considered when assessing the spread of variants in a partially susceptible population and improving efforts to control spread.
+Future studies should prioritize detailed assessment of the generation intervals of SARS-CoV-2 and widespread variants, as well as consider how uncertainty in generational intervals might bias conclusions.
 

### 2021-04-14 newvariant.tex
more edits

@@
 
-Estimating variant epidemic strength and speed remains one of the key questions in controlling the spread of SARS-CoV-2 variants of concern (VoCs) \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
+Estimating variant epidemic strength and speed remains one of the key questions in controlling the spread and understanding the threat of SARS-CoV-2 variants of concern (VoCs) \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
 Epidemic strength is characterized by the reproduction number $\RR$, which determines the average number of new infections per infection as well as the strength of intervention required to eliminate the disease \citep{anderson1991infectious}.
@@
 Many studies have tried to estimate the relative strength of the novel variant from the observed relative speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021}.
-In contrast, some studies have assumed a value of the relative strength of the novel variant and tried to predict its relative speed to determine when a new variant will become the dominant strain.
-\citep{davies2021estimated}.
+Some studies have instead assumed a value of the relative strength of the novel variant and tried to predict its relative speed to determine when a new variant will become the dominant strain.
+\citep{davies2021estimated,di2021impact}.
 While both approaches are reasonable in principle, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the disease and its control \citep{doi:10.1098/rspb.2020.1556}:
@@
 We use the renewal equation framework to characterize the spread of two pathogen strains---in this case, the wild type SARS-CoV-2 virus and a focal variant of concern, such as B.1.1.7.
-Neglecting the (relatively slow) rate of new mutations, the current incidence of infection $i_x(t)$ caused by the wild type (``wt'') and the variant (``var'') can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary cases are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
+Neglecting the (relatively slow) rate of new mutations, the current incidence of infection $i_x(t)$ caused by strain $x$---wild type (``wt'') and the variant (``var'')---can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary cases are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
 \begin{equation}
@@
 
-Epidemic speed $r_x$ can often be estimated directly from case data \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating};
-however, the estimating $r_x$ can be challenging when case counts are low (as when a new variant begins to spread) and is sensitive to temporal changes in testing patterns and intensity.
+Epidemic speed $r_x$ can often be estimated directly from case data during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}---in theory both $r$ and $\RR$ relate to latent infections and therefore using case data can lead to biases due to delays in reporting \citep{goldstein2009reconstructing,gostic2020practical}.
+However, estimating $r_x$ can be challenging when case counts are low (as when a new variant begins to spread) and is sensitive to temporal changes in testing patterns and intensity.
 Studies of new SARS-CoV-2 variants have mostly focused on characterizing changes in the \emph{proportion} of a new variant \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
@@
 As noted above, we assume that the variant and the wild type have equal $\kappa$ throughout, and only consider differences in the mean.
-We evaluate the estimates of relative strength $\rho$ across a wide range of $\kappa$ from 0 (delta distribution) to 1 (exponential distribution).
+We evaluate the estimates of relative strength $\rho$ across a wide range of $\kappa$ from 0 (fixed generation interval; delta distribution) to 1 (exponential distribution).
 To further explore how inference depends on underlying epidemiological conditions (i.e., whether the incidence of the variant and the wild type are growing or shrinking), we consider four epidemiological conditions: (1) $\rw < \rv < 0$, (2) $\rw < \rv = 0$, (3) $0 = \rw < \rv$, and (4) $0 < \rw < \rv$.
@@
 
-Consistent with earlier studies, we find that the new variant has higher strength than the wild type across a wide range of epidemiological conditions \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
+We find that an increased speed of $\delta=0.1\pday$ for the variant is consistent with higher strength than the wild type across a wide range of epidemiological conditions \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
 However, the magnitude of relative strength $\rho$ is sensitive to assumptions about generation intervals.
@@
 
-While relative speed $\delta$ and strength $\rho$ are useful for characterizing the spread of the variant in an epidemiological context with a previously dominant wild type, \emph{absolute} speed $\rv$ and strength $\Rv$ of the variant are the main variables that determine the spread and control of the variant over the long term.
+While relative speed $\delta$ and strength $\rho$ are useful for characterizing the spread of the variant in an epidemiological context with a previously dominant wild type, \emph{absolute} speed $\rv$ and strength $\Rv$ of the variant are the main variables that determine the spread and conditions for control of the variant over the long term.
 In particular, at any given point in epidemic, we can measure the speed of the variant $\rv$ (or infer $\rv$ from $\rw$ and $\delta$) and ask how much more intervention is required to control the spread of both strains (since $\Rv < 1$ implies $\Rw < 1$).
@@
 As the new variants, in particular the B.1.1.7 variant, of SARS-CoV-2 are spreading and becoming dominant in many countries, it is clear that the variants are more transmissible (have higher strength) than the wild type \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
-Our analysis supports estimates of a higher strength of the new variant across a wide range of epidemiological conditions and assumptions about generation-interval distributions.
+Our analysis supports estimates of a higher strength of the new variant bounding on a wide range of parameter ranges about epidemiological conditions and generation-interval distributions.
 However, our analysis also shows thatuncertainty in generation-interval distributions must be taken into account to obtain accurate estimates of the relative strength of variants.

### 2021-04-14 newvariant.tex
remaking a lot of figures

@@
 We use simulations to show how neglecting their differences can lead to biases and illustrate how such biases can be assessed in practice.
-Finally, we discuss implications for control: if new variants have longer generation intervals then speed-like interventions (e.g., contact tracing) will be more relatively effective than strength-like interventions (e.g., social distancing).
+Finally, we discuss implications for control: if new variants have longer generation intervals then speed-like interventions (e.g., contact tracing) become more effective, whereas strength-like interventions become less effective (e.g., social distancing).
 
@@
 
-\begin{figure}[!th]
+\begin{figure}[!t]
 \includegraphics[width=\textwidth]{relstrength.pdf}
@@
 The relative strength of the new variant $\rho$ is calculated using $\delta=0.1\pday$, $\Gw = 5\days$, and $\kappa = 1/5$.
-Four epidemiological conditions assume $\rw=-0.15\pday$ ($\rw < \rv < 0$), $\rw=-0.1\pday$ ($\rw < \rv = 0$), $\rw=0\pday$ ($0 = \rw < \rv$), and $\rw=0.1\pday$ ($0 < \rw < \rv$).
+Five epidemiological conditions assume $\rw=-0.15\pday$ ($\rw < \rv < 0$), $\rw=-0.1\pday$ ($\rw < \rv = 0$), $\rw=-0.05\pday$ ($\rw < 0 < \rv$), $\rw=0\pday$ ($0 = \rw < \rv$), and $\rw=0.05\pday$ ($0 < \rw < \rv$).
 }
@@
 Inference of $\rho$ from $\delta$ will depend on assumptions about the generation-interval distributions of both strains.
-Given the mean generation interval of the variant $\Gv$ and the wild type $\Gw$, the relative strength $\rho = \Rv/\Rw$ under the gamma approximation is given by \citep{park2019practical}:
+Given the mean generation interval of the variant $\Gv$ and the wild type $\Gw$, the relative strength $\rho = \Rv/\Rw$ under the gamma approximation \citep{park2019practical} is given by:
 \begin{equation}
@@
 We evaluate the estimates of relative strength $\rho$ across a wide range of $\kappa$ from 0 (fixed generation interval; delta distribution) to 1 (exponential distribution).
-To further explore how inference depends on underlying epidemiological conditions (i.e., whether the incidence of the variant and the wild type are growing or shrinking), we consider four epidemiological conditions: (1) $\rw < \rv < 0$, (2) $\rw < \rv = 0$, (3) $0 = \rw < \rv$, and (4) $0 < \rw < \rv$.
+To further explore how inference depends on underlying epidemiological conditions (i.e., whether the incidence of the variant and the wild type are growing or shrinking), we consider five epidemiological conditions: (1) $\rw < \rv < 0$, (2) $\rw < \rv = 0$, (3) $\rw < 0 < \rv$, (4) $0 = \rw < \rv$, and (5) $0 < \rw < \rv$.
 
@@
 therefore, narrower distributions lead to higher values of $\rho$ when $\rv > 0$ and lower values of $\rho$ when $\rv \leq 0$.
+When $\rw < 0 < \rv$, inference of $\rho$ is relatively insensitive to values of $\kappa$.
 
@@
 As our baseline scenario we assume $\rho = 1.61$, which is the value we obtain for $\delta=0.1\,\textrm{days}$, $\Gw = \Gv = 5\,\textrm{days}$, and $\kappa = 1/5$.
-We evaluate $\delta$ across four different epidemiological conditions as before: (1) $\Rw < \Rv < 1$, (2) $\Rw < \Rv = 1$, (3) $1 = \Rw < \Rv$, and (4) $1 < \Rw < \Rv$.
+We evaluate $\delta$ across five different epidemiological conditions as before: (1) $\Rw < \Rv < 1$, (2) $\Rw < \Rv = 1$, (3) $\Rw < 1 < \Rv$, (4) $1 = \Rw < \Rv$, and (5) $1 < \Rw < \Rv$.
 
@@
 Relative speed of the new variant $\delta$ is calculated using $\rho=1.61$, $\Gw = 5\days$, and $\kappa = 1/5$.
-Four epidemiological conditions assume $\Rw=0.5/\rho$ ($\Rw < \Rv < 1$), $\Rw=1/\rho$ ($\Rw < \Rv = 1$), $\Rw=1$ ($1 = \Rw < \Rv$), and $\Rw=\rho$ ($0 < \Rw < \Rv$).
+Five epidemiological conditions assume $\Rw=1/\rho^{3/2}$ ($\Rw < \Rv < 1$), $\Rw=1/\rho$ ($\Rw < \Rv = 1$), $\Rw=1/\rho^{1/2}$ ($\Rw < 1 < \Rv$), $\Rw=1$ ($1 = \Rw < \Rv$), and $\Rw=\rho^{1/2}$ ($0 < \Rw < \Rv$).
 }
@@
 For example, even when there are no differences in the generation-interval distributions ($\Gv=\Gw$, in this case), the relative speed $\delta$ can range between 0.08--0.11 when $\kappa=0.2$ and 0.06--0.14 when $\kappa=0.5$.
-Therefore, characterizing the spread of variants assuming constant relative speed (e.g., by fitting a standard logistic equation) without considering how epidemiological conditions change over time should be avoided.
+Therefore, characterizing the spread of variants assuming constant relative speed (e.g., by fitting a standard logistic growth curve) without considering how epidemiological conditions change over time should be avoided.
 
@@
 
-Instead of estimating relative strength from speed, one can directly estimate time-varying or instantaneous reproduction numbers of the variant and the wild type from incidence data, and compare their ratios \citep{fraser2007estimating};
+Instead of estimating relative strength from speed, one can estimate time-varying or instantaneous reproduction numbers $\RR(t)$ of the variant and the wild type from incidence data \citep{fraser2007estimating}, and directly compare their ratios;
 such methods have been used in previous analyses of the B.1.1.7 strain by \cite{volz2021transmission}.
@@
 Then, we compare the estimated ratio $\hat{\rho}(t) = \hat{\RR}_{\textrm{var}}(t)/\hat{\RR}_{\textrm{wt}}(t)$ with the true ratio $\rho = \Rv(t)/\Rw(t)$.
-In order to allow for smooth changes in $\RR$ driven by non-pharmaceutical interventions and behavior changes, we model $\Rw(t)$ phenomenologically (\fref{Rtbias}) and let $\Rv(t) = \rho \Rw(t)$.
+In order to allow for changes in $\RR$ driven by the introduction and lifting of non-pharmaceutical interventions \citep{flaxman2020Rt}, we model $\Rw(t)$ using a step function (\fref{Rtbias}) and let $\Rv(t) = \rho \Rw(t)$.
 
@@
 For all simulations, squared coefficient of variation in generation intervals is assumed to equal $\kappa = 1/5$.
-We assume $\Rw(t)=2.5$ before $t=15\,\textrm{days}$ to reflect unmitigated spread and let $\Rw(t)=2.5 (1+0.7\cos((t-15)/10))/1.7$ after to model smooth introduction and lifting of non-pharmaceutical interventions.
 }

### 2021-04-14 Rtbias.R

@@ -1,5 +1,5 @@
 library(dplyr)
-library(ggplot2); theme_set(theme_bw(base_size = 12))
+library(ggplot2); theme_set(theme_bw(base_size = 14))
 library(egg)
 library(ggpubr)
 library(gridExtra)
@@ -40,10 +40,12 @@ for (i in 1:length(svec)) {
   tmparg <- c(theta=theta,
     tmax=150, slist[[i]],
     Rfun=function(t) {
-      if (t < 15) {
-        return(2.5)
+      if (t < 30) {
+        return(2)
+      } else if (t < 60) {
+        return(0.5)
       } else {
-        return(2.5 * (1 + 0.7 * cos((t-15)/10))/1.7)
+        return(1.2)
       }
     })
   
@@ -51,10 +53,10 @@ for (i in 1:length(svec)) {
     filter(tvec > 15, tvec < 70)
   
   g1 <- ggplot(rr) +
-    geom_line(aes(tvec, Rt1, color="Wild type", lty="True"), lwd=1) +
-    geom_line(aes(tvec, Rt2, color="Variant", lty="True"), lwd=1) +
-    geom_line(aes(tvec, Rtest1, color="Wild type", lty="Estimated"), lwd=1) +
-    geom_line(aes(tvec, Rtest2, color="Variant", lty="Estimated"), lwd=1) +
+    geom_line(aes(tvec, Rt1, color="Wild type", lty="True"), lwd=2) +
+    geom_line(aes(tvec, Rt2, color="Variant", lty="True"), lwd=2) +
+    geom_line(aes(tvec, Rtest1, color="Wild type", lty="Estimated"), lwd=2) +
+    geom_line(aes(tvec, Rtest2, color="Variant", lty="Estimated"), lwd=2) +
     scale_x_continuous("Time (days)", expand=c(0, 0), limits=c(15, 70)) +
     scale_y_log10("Reproduction number, $\\mathcal{R}(t)$", limits=c(0.25, 8),
                   breaks=c(0.25, 0.5, 1, 2, 4, 8)) +
@@ -73,10 +75,10 @@ for (i in 1:length(svec)) {
   
   if (i==2) {
     g1 <- ggplot(rr) +
-      geom_line(aes(tvec, Rt1, lty="True"), color="black", lwd=1) +
-      geom_line(aes(tvec, Rt2, col="True", lty="True"), lwd=1) +
[…]

### 2021-04-14 relspeed.R

@@ -1,5 +1,5 @@
 library(dplyr)
-library(ggplot2); theme_set(theme_bw(base_size=14, base_family = "Times"))
+library(ggplot2); theme_set(theme_bw(base_size=18, base_family = "Times"))
 library(ggthemes)
 library(egg)
 library(tikzDevice)
@@ -11,7 +11,7 @@ Gw <- 5
 
 rho <- ((1 + kappa_base * (rw+delta) * Gw)^(1/kappa_base))
 
-Rwvec <- c(0.5/rho, 1/rho, 1, rho)
+Rwvec <- c(1/rho^{3/2}, 1/rho, 1/sqrt(rho), 1, sqrt(rho))
 kappavec <- c(0, 0.2, 0.5, 1)
 
 Gratio <- exp(seq(log(1/1.5), log(1.5), length.out=31))
@@ -41,29 +41,52 @@ deltadata <- apply(pardata, 1, function(x) {
   bind_rows %>%
   mutate(
     kappa=factor(kappa, levels=c(0, 0.2, 0.5, 1)),
-    Rw=factor(Rw, levels=Rwvec,
-              labels=c("$\\mathcal{R}_{\\textrm{wt}} < \\mathcal{R}_{\\textrm{var}} < 1$",
-                       "$\\mathcal{R}_{\\textrm{wt}} < \\mathcal{R}_{\\textrm{var}} = 1$",
+    Rw=factor(Rw, levels=c(Rwvec[3:5], Rwvec[1:2]),
+              labels=c("$\\mathcal{R}_{\\textrm{wt}} < 1 < \\mathcal{R}_{\\textrm{var}}$",
                        "$1 = \\mathcal{R}_{\\textrm{wt}} < \\mathcal{R}_{\\textrm{var}}$",
-                       "$1 < \\mathcal{R}_{\\textrm{wt}} < \\mathcal{R}_{\\textrm{var}}$"))
+                       "$1 < \\mathcal{R}_{\\textrm{wt}} < \\mathcal{R}_{\\textrm{var}}$",
+                       "$\\mathcal{R}_{\\textrm{wt}} < \\mathcal{R}_{\\textrm{var}} < 1$",
+                       "$\\mathcal{R}_{\\textrm{wt}} < \\mathcal{R}_{\\textrm{var}} = 1$"))
   )
 
 deltadata %>%
   filter(Gratio==1, kappa==0.5)
 
+kappa_title1 <- data.frame(
+  x=c(1.3, 1.3),
+  y=c(0.06, 0.12),
[…]

### 2021-04-14 relstrength.R

@@ -1,12 +1,13 @@
 library(dplyr)
-library(ggplot2); theme_set(theme_bw(base_size=14, base_family = "Times"))
+library(ggplot2); theme_set(theme_bw(base_size=18, base_family = "Times"))
 library(ggthemes)
 library(egg)
+library(directlabels)
 library(tikzDevice)
 
 delta <- 0.1
 
-rwvec <- c(-0.15, -0.1, 0, 0.1)
+rwvec <- c(-0.15, -0.1, -0.05, 0, 0.05)
 kappavec <- c(0, 0.2, 0.5, 1)
 
 Gw <- 5
@@ -39,28 +40,49 @@ rhodata <- apply(pardata, 1, function(x) {
   bind_rows %>%
   mutate(
     kappa=factor(kappa, levels=c(0, 0.2, 0.5, 1)),
-    rw=factor(rw, levels=rwvec,
-              labels=c("$r_{\\textrm{wt}} < r_{\\textrm{var}} < 0$",
-                       "$r_{\\textrm{wt}} < r_{\\textrm{var}} = 0$",
+    rw=factor(rw, levels=c(rwvec[3:5], rwvec[1:2]),
+              labels=c("$r_{\\textrm{wt}} < 0 < r_{\\textrm{var}}$",
                        "$0 = r_{\\textrm{wt}} < r_{\\textrm{var}}$",
-                       "$0 < r_{\\textrm{wt}} < r_{\\textrm{var}}$"))
+                       "$0 < r_{\\textrm{wt}} < r_{\\textrm{var}}$",
+                       "$r_{\\textrm{wt}} < r_{\\textrm{var}} < 0$",
+                       "$r_{\\textrm{wt}} < r_{\\textrm{var}} = 0$"))
   )
 
 range(filter(rhodata, !kappa %in% c(0, 1))$rho)
 
+kappa_title1 <- data.frame(
+  x=c(1.3, 1.3),
+  y=c(1.5, 2.2),
+  labels=c("$\\kappa=0$", "$\\kappa=1$"),
+  rw="$r_{\\textrm{wt}} < r_{\\textrm{var}} = 0$"
+)
+
+kappa_title2 <- data.frame(
+  x=c(1.3, 1.3),
+  y=c(1.4, 2.4),
+  labels=c("$\\kappa=1$", "$\\kappa=0$"),
+  rw="$0 < r_{\\textrm{wt}} < r_{\\textrm{var}}$"
+)
+
 g1 <- ggplot(rhodata) +
[…]

### 2021-04-14 newvariant.tex
rewriting the entire section

@@
 However, if the generation-interval distribution of the variant differs from the assumed distribution, the ratio changes over time (\fref{Rtbias}E,H).
-In particular, if the true generation intervals have a shorter mean than the assumed distribution, we over-estimate $\Rv(t)$ during the growth phase and under-estimate during the decay phase (and vice versa, \fref{Rtbias}D,G), which further translates to biases in the estimated relative strength (\fref{Rtbias}F,I).
+In particular, if the true generation intervals have a shorter mean than the assumed distribution, we over-estimate $\Rv(t)$ during the growth phase and under-estimate during the decay phase (and vice versa, \fref{Rtbias}D,G), which further translates to biases in the estimated relative strength (\fref{Rtbias}E,H).
+Therefore, biases in the estimated relative strength will correlate with the strength of the variant, (i.e., whether $\Rv(t)$ is below or above 1).
 
-We can also assess these biases by plotting the estimated strength of the variant $\hat{\RR}_{\textrm{var}}(t)$ against the estimated strength of the wild type $\hat{\RR}_{\textrm{wt}}(t)$ (\fref{Rtbias}C,F,I).
+In practice, estimates of instantaneous reproduction numbers $\RR(t)$ (and therefore, their ratios) can be noisy due to limited data availability or model assumptions;
+instead, we might want to estimate a single value of relative strength $\rho$.
+For example, we can estimate $\rho$ by plotting the estimated strength of the variant $\hat{\RR}_{\textrm{var}}(t)$ against the estimated strength of the wild type $\hat{\RR}_{\textrm{wt}}(t)$---as presented in Figure 2 of \cite{volz2021transmission}---and performing a linear regression (\fref{Rtbias}C,F,I).
 When generation-interval distributions are correctly specified, we obtain a straight line with a slope of $\rho$ and intercept at zero (\fref{Rtbias}C).
-When the assumed mean generation interval is longer than the that of the variant, we over-estimate the slope (and vice versa, \fref{Rtbias}F,I).
-In addition, the intercept (if we were to perform regression) need not be at zero due to biases in the slope.
-Smooth changes in $\RR$ further generate circular patterns but biases in slopes are clear.
+However, when the assumed mean generation interval is longer than the that of the variant, we over-estimate the slope (and vice versa, \fref{Rtbias}F,I).
+In theory, we expect the regression line to go through the origin, but when the assumed generation-interval distribution differs from the true distribution, this is not necessarily the case: biases in slopes due to mis-specified generation-interval distributions also lead to biases in intercepts.
 
@@
 In this case, we can control the spread of the variant when $\theta > \Rv$ or $\phi > \rv$, respectively.
-In general, we can calculate the strength ($\theta = \Ry{pre}/\Ry{post}$) and speed ($\phi = \ry{pre} - \ry{post}$) for any intervention by comparing epidemiological conditions before and after the intervention \citep{doi:10.1098/rspb.2020.1556}---these quantities, by definition, must be greater than the pre-intervention strength or speed of the epidemic in order to control the epidemic.
-Here, we use the strength-and-speed framework to compare how differences in generation-interval distributions would affect the control of the spread of a new variant.
+As a baseline scenario, we consider constant-strength and speed interventions that reduce $\Rv$ to 0.9 when both the variant and the wild type have equal mean generation intervals (\fref{strengthspeed}A--B).
+While both interventions are equally effective on strength-scale (that is, $\Ry{post}=\int  K_{\mathrm{post}}(\tau) \dtau = 0.9$), they have different dynamical implications.
+The constant-strength intervention affects transmission equally throughout the course of infection, whereas the constant-speed intervention has greater impact of transmission that occur later;
+as a result, the constant-speed intervention reduces the post-intervention mean generation interval (\fref{strengthspeed}B) and leads to (slightly) faster exponential decay (therefore, lower $\ry{post}$).
 
@@
 
-\fref{strengthspeed} compares the effectiveness of strength-based and speed-based interventions across a wide range of assumptions about generation-interval distributions.
-We begin by assuming that the variant and the wild type have equal mean generation intervals and model interventions such that they both reduce $\Rv$ to 0.9 after intervention (\fref{strengthspeed}A).
-If the mean generation interval of the variant is longer than that of the wild type, then the strength of the variant will be higher;
-since the strength of the strength-like intervention is insensitive to the underlying generation-interval distribution, we may fail to control the spread of the new variant.
-With speed-like interventions, however, there is a countervailing effect:
-since speed-like interventions like contact tracing have disproportionate effects on preventing later transmission, they are more effective (therefore stronger) when the mean generation interval of the new variant is longer.
+However, if the variant has longer generation intervals than the wild type (\fref{strengthspeed}C--D),  then the strength of the variant will be higher conditional on the observed speed (\fref{strengthspeed}E).
+In this case, the same constant-strength intervention can fail to control the epidemic (i.e., $\Ry{post} > 1$; \fref{strengthspeed}E) because this intervention reduces the transmission by a constant amount regardless of age of infection (\fref{strengthspeed}C).
+On the other hand, the same constant-speed intervention will prevent a larger proportion of transmission, leading to lower $\Ry{post}$ (\fref{strengthspeed}E), because it is more effective against late-stage transmission (\fref{strengthspeed}D).
+The constant-speed intervention also has a greater impact on reducing the mean generation interval (\fref{strengthspeed}D).
 
-The speed-based paradigm gives the same results about the control but provides important insights (\fref{strengthspeed}B).
-The observed speed of the variant at any given moment is independent of our estimates of its mean generation interval.
-Therefore, if we can reduce infectiousness at a faster rate than the observed speed of spread (i.e., if the speed of intervention is faster than the epidemic speed), we can control the epidemic---this explains why a constant-speed intervention can prevent the spread of the variant regardless of its mean generation interval.
+The speed-based paradigm gives the same results about the control but provides important insights (\fref{strengthspeed}F).
+The observed speed of the variant $\rv$ at a given moment is independent of our estimates of its mean generation interval.
+Likewise, the post-intervention speed of the variant under the constant-speed intevention is also independent of the mean generation interval;
+therefore, the constant-speed intevention will be always equally effective (on the speed scale) in preventing the spread.
+
+The constant post-intervention speed can be understood by considering a speed-based decomposition of the kernel: $K_{\mathrm{pre}}(\tau) = \exp(\rv) b(\tau)$, where $b(\tau)$ is the initial backward generation-interval distribution and therefore integrates to 1 \citep{doi:10.1098/rspb.2020.1556}.
+This decomposition allows us to re-write the post-intervention kernel (under the constant-speed intervention) as follows:
+\begin{equation}
+K_{\mathrm{post}}(\tau) = \exp(\rv-\phi) b(\tau).
+\end{equation}
+Since the post-intervention speed $\ry{post}$ is determined by the following implicit equation,
+\begin{equation}
+1 = \int_0^\infty \exp(-\ry{post} \tau) K_{\mathrm{post}}(\tau) \dtau = \int_0^\infty \exp(\rv-\phi-\ry{post}) b(\tau)  \dtau,
+\end{equation}
+we have $\ry{post} = \phi - \rv$.
+Therefore, if we can reduce infectiousness at a faster rate than the observed speed of spread (i.e., if the speed of intervention is faster than the epidemic speed, $\phi > \rv$), we can control the epidemic regardless of the underlying generation-interval distribution.
 

### 2021-04-14 control.R

@@ -1,5 +1,5 @@
 library(dplyr)
-library(ggplot2); theme_set(theme_bw(base_size=14, base_family = "Times"))
+library(ggplot2); theme_set(theme_bw(base_size=18, base_family = "Times"))
 library(ggthemes)
 library(egg)
 library(tikzDevice)
@@ -29,8 +29,6 @@ constant_speed <- optim(0.1,
                         upper=0.2,
                         method="Brent")[[1]]
 
-
-
 intervention <- sapply(Gratio, function(x) {
   Gr <- x
   Gv <- Gw * Gr
@@ -39,63 +37,156 @@ intervention <- sapply(Gratio, function(x) {
   
   rv <- rw + delta
   
-  strength_strength <- constant_strength
-  strength_speed <- (Rv^kappa-1)/kappa/Gv - ((Rv/constant_strength)^kappa-1)/kappa/Gv
+  strength_post_strength <- Rv/constant_strength
+  strength_post_speed <- ((strength_post_strength)^kappa-1)/kappa/Gv
   
-  speed_speed <- constant_speed 
-  
-  speed_strength <- Rv/integrate(function(y) Rv * dgamma(y, shape=1/kappa, rate=1/kappa/Gv) * exp(-constant_speed * y),
+  speed_post_strength <- integrate(function(y) Rv * dgamma(y, shape=1/kappa, rate=1/kappa/Gv) * exp(-constant_speed * y),
                                  0,
                                  Inf)[[1]]
+  
+  speed_post_speed <- optim(-0.05,
+        function(r) {
+          (integrate(function(y) Rv * dgamma(y, shape=1/kappa, rate=1/kappa/Gv) * exp(-constant_speed * y) * exp(-r * y),
+                     0,
+                     Inf)[[1]] - 1)^2
+        },
+        lower=-0.1,
+        upper=0.1,
+        method="Brent")[[1]]
   
   data.frame(
     Gratio=Gr,
     rw=rw,
     Rv=Rv,
     rv=rv,
-    strength_strength=strength_strength,
-    strength_speed=strength_speed,
-    speed_strength=speed_strength,
-    speed_speed=speed_speed
+    strength_post_strength=strength_post_strength,
+    strength_post_speed=strength_post_speed,
+    speed_post_strength=speed_post_strength,
[…]

### 2021-04-15 newvariant.tex
done..?

@@
 \usepackage{natbib}
-\bibliographystyle{chicago}
+\bibliographystyle{apalike}
 \date{\today}
@@
 
-\newcommand{\days}{\ensuremath{\, \textrm{day}}}
+\newcommand{\days}{\ensuremath{\, \textrm{days}}}
 \newcommand{\pday}{\ensuremath{/\textrm{day}}}
@@
 Epidemic speed is characterized by the growth rate $r$, which describes how fast a disease spreads at the population-level.
-Like epidemic strength, epidemic speed also determines conditions for disease eliminationl $r=0$ is a threshold equivalent to $\RR=1$ \citep{doi:10.1098/rspb.2020.1556}.
+Like epidemic strength, epidemic speed also determines conditions for disease elimination $r=0$ is a threshold equivalent to $\RR=1$ \citep{doi:10.1098/rspb.2020.1556}.
 Strength and speed are linked by generation intervals---defined as time between infection and transmission---which characterizes the individual-level time scale of the epidemic \citep{roberts2007model,svensson2007note,wallinga2007generation}.
@@
 (C, F, I) True (purple, solid) and estimated (orange, dashed) relationships between estimated reproduction numbers.
+Blue dotted lines represent the regression lines of the estimated variant reproduction numbers against the estimated wild type reproduction numbers.
 Top row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type and the variant.
@@
 In theory, we expect the regression line to go through the origin, but when the assumed generation-interval distribution differs from the true distribution, this is not necessarily the case: biases in slopes due to mis-specified generation-interval distributions also lead to biases in intercepts.
+This analysis suggests another method for assessing potential biases in the estimates of relative strength.
 
@@
 \caption{
-\textbf{Strength and speed of intervention against the new variant.}
-(A) Strength of epidemic (black, solid), constant strength intervention (orange, dashed), and constant speed intervention (blue, dotted).
-(B) Speed of epidemic (black, solid), constant strength intervention (orange, dashed), and constant speed intervention (blue, dotted).
-Strength and speed are calculated assuming $\rw=0\,\textrm{/days}$, $\delta=0.1\,\textrm{/days}$, $\rv=0\,\textrm{/days}$, $\Gy{wt}=5\,\textrm{days}$, and $\kappa=1/5$.
+\textbf{Effects of constant-strength and speed of interventions against the spread new variant.}
+(A--B) Pre-intervention (black) and post-intervention kernel (colored) of the new variant under constant-strength (A) and speed (B) interventions assuming that the variant has equal generation intervals as the wild type.
+(C--D) Pre-intervention (black) and post-intervention kernel (colored) of the new variant under constant-strength (C) and speed (D) interventions assuming that the variant has longer generation intervals as the wild type.
+(E) Pre-intervention (black) and post-intervention strength conditional on the mean generation interval of the variant (colored).
+(F) Pre-intervention (black) and post-intervention speed conditional on the mean generation interval of the variant (colored).
+Strength and speed are calculated assuming$\rw=0\,\textrm{/days}$, $\delta=0.1\,\textrm{/days}$, $\rv=0\,\textrm{/days}$, $\Gy{wt}=5\,\textrm{days}$, and $\kappa=1/5$ for pre-intervention conditions.
 Constant strength and speed interventions are modeled such that post intervention strength of the new variant, assuming equal mean generation intervals at 5 days, corresponds to 0.9.
@@
 The observed speed of the variant $\rv$ at a given moment is independent of our estimates of its mean generation interval.
-Likewise, the post-intervention speed of the variant under the constant-speed intevention is also independent of the mean generation interval;
-therefore, the constant-speed intevention will be always equally effective (on the speed scale) in preventing the spread.
+Likewise, the post-intervention speed of the variant under the constant-speed intervention is also independent of the mean generation interval;
+therefore, the constant-speed intervention will be always equally effective (on the speed scale) in preventing the spread.
 
@@
 
-\swp{Stylistic edits from Ben: remove ``we did this we did that''}
 We explored how the generation-interval distribution shapes the link between relative strength and speed of the new variant.
-We showed that longer generation intervals generally lead to higher relative strength for a given relative speed (and conversely, lower relative speed for a given relative strength); these can be reversed when incidence is decreasing.
-We also found that uncertainty about the relative mean generation interval of the variant has important implications for estimates of relative strength of the variant.
-We illustrated how these biases might be assessed and discussed how differences in generation intervals might lead to different conclusions about the effectiveness of interventions:
+Longer generation intervals generally lead to higher relative strength for a given relative speed (and conversely, lower relative speed for a given relative strength); these can be reversed when incidence is decreasing.
+Neglecting potential differences in the mean generation intervals between the variant and the wild type can bias estimates of the relative strength from incidence data;
+these biases may be systematically assessed by considering whether apparent changes in relative strength over time are associated with underlying epidemiological dynamics of the variant and the wild type.
+Finally, differences in generation intervals can further lead to different conclusions about the effectiveness of interventions:
 if the variant has longer (respectively, shorter) generation intervals than the wild type, speed-like (strength-like) interventions, will be more relatively more effective than naive estimates would suggest.
@@
 Our analysis supports estimates of a higher strength of the new variant bounding on a wide range of parameter ranges about epidemiological conditions and generation-interval distributions.
-However, our analysis also shows thatuncertainty in generation-interval distributions must be taken into account to obtain accurate estimates of the relative strength of variants.
+However, our analysis also shows that uncertainty in generation-interval distributions must be taken into account to obtain accurate estimates of the relative strength of variants.
 If the variant has a longer infectious period \citep{kissler2021densely}, and therefore extended generation-interval distributions, current estimates may be underestimating the true relative strength.
-While previous analyses have typically focused on estimating the relative strength and speed, absolute strength and speed are needed to properly evaluate its spread and control.
-Future studies should prioritize characterizing the spread of variants in absolute terms.
-Epidemiological conditions of the wild type should also be reported with the relative strength and speed in order to provide a proper context to evaluate their dynamics.
 
-We used simulations to show how mis-specification of generation-interval distributions can bias the inference of relative strenth from incidence data:
+There is currently a (minor) discrepancy in the estimates of relative strength of new variants, particularly for B.1.1.7.
+Mathematical analyses have typically reported greater than 1.4 fold increase in reproduction number for the B.1.1.7 variant ($\rho > 1.4$) whereas an independent analysis of secondary attack rates from contact tracing data suggests 1.25--1.4 fold increase \citep{ukinvest}.
+We can ask whether different estimates can be reconciled by generation intervals alone---for simplicity, we assume that the variant and the wild type have identical generation-interval distributions as other studies have.
+For example, \cite{davies2021estimated} estimated $\rho=1.77$ assuming a fixed generation generation interval at 5.5 days from $\delta=0.104\pday$ in their regression analysis;
+similarly, \cite{volz2021transmission} estimated $\rho=1.79$ by directly comparing the ratio between $\Rv$ and $\Rw$ assuming a mean generation interval of 6.5 days and a squared coefficient of variation of 0.38.
+While \cite{volz2021transmission} assumed a longer mean generation interval (which typically increases $\rho$), their estimate is similar to \cite{davies2021estimated}, likely because the effect of a wide distribution (which typically reduces $\rho$) can cancel out with the effect of longer intervals---in fact, we obtain $\rho=1.81$ from $\delta=0.104\pday$, $\rw=0\pday$, $\Gv=\Gw=6.4\days$, and $\kappa=0.38$ under the gamma assumption, explaining the lack of differences in the estimates of $\rho$ despite the differences in the assumed generation-interval distributions.
+
+However, a fixed generation generation interval assumed by \cite{davies2021estimated} is too narrow.
+The mean generation interval of 6.5 days assumed by \cite{volz2021transmission} is also likely too long:
+first, this estimate actually corresponds to the observed mean serial intervals from China in between January and February, 2020 \citep{bi2020epidemiology}, during which they were more likely to observe longer serial intervals due to initial epidemic growth \citep{park2021forward, ali2020serial}; and second, recently infected individuals may be less likely to transmit after symptom onset due to increased awareness of the pandemic compared to earlier on, leading to shorter generation intervals.
+Instead, if we replace the assumed generation-interval distributions for \cite{volz2021transmission} with shorter generation interval estimates from Tianjin, China ($\Gw=2.57\days$ and $\kappa=1$; \cite{ganyani2020estimating}), we obtain $\rho=1.27$, which is consistent with the attack rate analysis \citep{ukinvest}.
+We do not claim that these estimates of $\rho$ or $\Gw$ are more accurate;
+instead, this back-of-the-envelope calculation highlights the importance of propagating uncertainty in the generation-interval distribution in assessing the relative strength and speed of SARS-CoV-2 variants.
+
+We further used simulations to show how mis-specification of generation-interval distributions can bias the inference of relative strength from incidence data:
 when the assumed generation-interval distribution does not match the true generation-interval distribution of the variant, estimates of absolute strength $\Rv(t)$ will be biased, causing the estimated relative strength to change over time.
-In doing so, we assumed that the incidence of infection caused by the variant and the wild type is known.
-However, estimating $\RR(t)$ from real data, such as case incidence, is often associated with several practical challenges, such as delays between infection and reporting and changes in testing patterns.
+In doing so, we assumed that the intervention will reduce transmission caused by the variant and the wild type by equal amounts, thereby preserving the relative strength over time;
+however, this is only true under strength-like interventions, which is insensitive to age of infection, but not under speed-like interventions. 
+As we demonstrated earlier, the effectiveness of speed-like interventions depend on the shape of the generation-interval distribution;
+therefore, if the variant and the wild type have different generation intervals, speed-like interventions, such as contact tracing, can affect them differently.
+
+We also assumed that the incidence of infection caused by the variant and the wild type is known.
+However, estimating $\RR(t)$ from real data, such as case incidence, is often associated with several practical challenges, such as delays between infection and reporting and changes in testing patterns \citep{gostic2020practical}.
 We have chosen to focus here on the underlying dynamical mechanisms that may impact inference.
@@
 
+Here, we lay out a few practical advices in analyzing the epidemiological dynamics of new variants for future studies.
+First, models that assume a constant relative speed, such as standard logistic growth model, should be avoided for characterizing the growth of new variants---the relative speed changes over time depending on epidemiological conditions.
+Second, the absolute strength and speed should not be neglected over their relative values, and both relative and absolute values should be reported together.
+While the relative strength and speed are useful for describing the spread of new variants, the absolute values determine their spread and control.
+In addition, as we showed in our analysis, it is critical to understand the underlying epidemiological conditions of the variant and the wild type in interpreting the relative strength and speed.
+Finally, uncertainty in generation intervals should be carefully considered.
+
 Even though SARS-CoV-2 has been spreading for more than a year, there is still considerable uncertainty in its generation intervals.
@@
 Furthermore, realized serial intervals are subject to dynamical biases that can be difficult to tease apart \citep{park2021forward}.
-A few studies have tried to estimate the generation-interval distribution from serial intervals, with means ranging between 3--6 days and squared coefficients of variation ranging between 0.1--0.6 \citep{ferretti2020quantifying,Ferretti2020timing,ganyani2020estimating,knight2020estimating};
-however, most estimates neglect dynamical biases in serial intervals caused by underlying epidemiological dynamics in their estimates of generation intervals.
-In addition, serial intervals do not account for asymptomatic transmission, adding further uncertainty to inferences of speed and strength \citep{park2020time}.
+A few studies have tried to estimate the generation-interval distribution from serial intervals, with means ranging between 3--6 days and squared coefficients of variation ranging between 0.1--1 \citep{ferretti2020quantifying,Ferretti2020timing,ganyani2020estimating,knight2020estimating};
+as we have shown, this uncertainty alone can generate large uncertainty in the inference of relative strength.
+Furthermore, most estimates neglect dynamical biases in serial intervals caused by underlying epidemiological dynamics in their estimates of generation intervals.
+Serial intervals also do not account for asymptomatic transmission, adding further uncertainty to inferences of speed and strength \citep{park2020time}.
 
@@
 The spread of new SARS-CoV-2 variants and the replacement of previously dominant lineages represent ongoing challenges for controlling the SARS-CoV-2 pandemic.  
-By explicitly considering epidemiological context and generation interval differences together, we have also shown that improving estimates of the the relative duration of infectiousness at the individual scale may represent a pathway towards more effective interventions.
-Specifically, we have shown that
+By explicitly considering epidemiological context and generation interval differences together, we have shown that improving estimates of the the relative duration of infectiousness at the individual scale may represent a pathway towards more effective interventions.
+Specifically, we have shown that speed-like interventions, such as contact tracing, will be more effective if variants have longer genenration intervals.
+Most intervention strategies throughout the current pandemic have focused on strength-like interventions \citep{flaxman2020Rt}, such as lockdowns, partly because pre-symptomatic transmission of SARS-CoV2 have limited the effectiveness of contact tracing efforts \citep{hellewell2020feasibility}.
+However, given the possibility that new variants can have different infection characteristics, future studies should consider whether their transmission dynamics also differ (e.g., the amount of pre-symptomatic transmission) and evaluate intervention strategies accordingly.
+
+\section{Methods}
+
+\subsection{Two-strain renewal equation}
+
+We use a two-strain renewal equation to simultaneously model the spread of the variant and the wild type.
+Ignoring birth and death, the incidence of infection caused by the variant $\iv$ and the wild type $\iw$ assuming perfect cross immunity is given by:
+\begin{align}
+\frac{dS}{dt} &= - \iv(t) - \iw(t)\\
+\iv(t) &= \Rv(t) \int_{0}^\infty \iv(t-\tau) g_{\mathrm{var}}(\tau) \dtau\\
+\iw(t) &= \Rw(t) \int_{0}^\infty \iw(t-\tau) g_{\mathrm{wt}}(\tau) \dtau
+\end{align}
+where $S$ represents the proportion of susceptible individuals.
+We discretize the model at the time scale of 0.025 days as described in \citep{park2021forward} and simulated the model with $\iv(0) = 0.001$ and $\iw(0) = 0.1$.
+We assume that $\Rw(t)$ remains at 2 for 30 days, immediately decreases and stays at 0.5 until day 60, and immediately increases back to 1.2.
+Finally, we assume $\Rv(t) = \rho \Rw(t)$.
 

### 2021-04-15 control.R

@@ -90,14 +90,14 @@ Gpost_long <- integrate(function(x) x*Rv_base*dgamma(x, shape=1/kappa, rate=1/ka
                         Inf)[[1]]/filter(intervention, Gratio==tail(Gratio,1))$speed_post_strength
 
 g1 <- ggplot(gendata_base) +
-  geom_line(aes(tvec, density, col="Epidemic", lty="Epidemic"), lwd=2) +
+  geom_line(aes(tvec, density, col="Pre-intervention", lty="Pre-intervention"), lwd=2) +
   geom_line(aes(tvec, strength, col="Post-intervention (constant-strength)", lty="Post-intervention (constant-strength)"), lwd=2) +
   geom_vline(xintercept = Gw, lwd=2) +
   geom_vline(xintercept = Gw, lwd=2, col=colorblind_pal()(3)[3], lty=3) +
   scale_x_continuous("Generation intervals (days)") +
   scale_y_continuous("Kernel density", limits=c(0, 0.4)) +
-  scale_color_manual("", values=(colorblind_pal()(3)[c(1, 3)])) +
-  scale_linetype_manual("", values=c(1, 3)) +
+  scale_color_manual("", values=(colorblind_pal()(3)[c(3, 1)])) +
+  scale_linetype_manual("", values=c(3, 1)) +
   ggtitle("A. Equal generation intervals") +
   theme(
     panel.grid = element_blank(),
@@ -106,14 +106,14 @@ g1 <- ggplot(gendata_base) +
   )
 
 g2 <- ggplot(gendata_base) +
-  geom_line(aes(tvec, density, col="Epidemic", lty="Epidemic"), lwd=2) +
+  geom_line(aes(tvec, density, col="Pre-intervention", lty="Pre-intervention"), lwd=2) +
   geom_line(aes(tvec, speed, col="Post-intervention (constant-strength)", lty="Post-intervention (constant-strength)"), lwd=2) +
   geom_vline(xintercept = Gw, lwd=2) +
   geom_vline(xintercept = Gpost_base, col=colorblind_pal()(3)[2], lwd=2, lty=2) +
   scale_x_continuous("Generation intervals (days)") +
   scale_y_continuous("Kernel density", limits=c(0, 0.4)) +
-  scale_color_manual("", values=(colorblind_pal()(3)[c(1, 2)])) +
-  scale_linetype_manual("", values=c(1, 2)) +
[…]

### 2021-04-15 newvariant.tex
done

@@
 
-\begin{figure}[!th]
+\begin{figure}[!pht]
 \includegraphics[width=\textwidth]{Rtbias.pdf}
@@
 Blue dotted lines represent the regression lines of the estimated variant reproduction numbers against the estimated wild type reproduction numbers.
+Gray lines represent the one-to-one line.
 Top row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type and the variant.
@@
 
-Here, we lay out a few practical advices in analyzing the epidemiological dynamics of new variants for future studies.
+Here, we lay out a few practical advice in analyzing the epidemiological dynamics of new variants for future studies.
 First, models that assume a constant relative speed, such as standard logistic growth model, should be avoided for characterizing the growth of new variants---the relative speed changes over time depending on epidemiological conditions.
@@
 By explicitly considering epidemiological context and generation interval differences together, we have shown that improving estimates of the the relative duration of infectiousness at the individual scale may represent a pathway towards more effective interventions.
-Specifically, we have shown that speed-like interventions, such as contact tracing, will be more effective if variants have longer genenration intervals.
-Most intervention strategies throughout the current pandemic have focused on strength-like interventions \citep{flaxman2020Rt}, such as lockdowns, partly because pre-symptomatic transmission of SARS-CoV2 have limited the effectiveness of contact tracing efforts \citep{hellewell2020feasibility}.
+Specifically, we have shown that speed-like interventions, such as contact tracing, will be more effective if variants have longer generation intervals.
+Most intervention strategies throughout the current pandemic have focused on strength-like interventions \citep{flaxman2020Rt}, such as lock-downs, partly because pre-symptomatic transmission of SARS-CoV2 have limited the effectiveness of contact tracing efforts \citep{hellewell2020feasibility}.
 However, given the possibility that new variants can have different infection characteristics, future studies should consider whether their transmission dynamics also differ (e.g., the amount of pre-symptomatic transmission) and evaluate intervention strategies accordingly.
@@
 
-We use a two-strain renewal equation to simultaneously model the spread of the variant and the wild type.
+We use a two-strain renewal equation to simulate the spread of the variant and the wild type.
 Ignoring birth and death, the incidence of infection caused by the variant $\iv$ and the wild type $\iw$ assuming perfect cross immunity is given by:

### 2021-04-15 newvariant.tex
minor changes

@@
 \RR_x(t) = \frac{i_x(t)}{\int_0^\infty i_x(t-\tau) g_x(\tau) \dtau}.
+\label{eq:rt}
 \end{equation}
@@
 instead, this back-of-the-envelope calculation highlights the importance of propagating uncertainty in the generation-interval distribution in assessing the relative strength and speed of SARS-CoV-2 variants.
+Obtaining an accurate and precise estimate of $\rho$ will need to take into account changes in $\rho$ as well as underlying $\rw$.
 
@@
 Finally, we assume $\Rv(t) = \rho \Rw(t)$.
+From simulated incidence curves between day 15 and 70, we estimate the instaneous reproduction number using \eref{rt}.
+We ignore incidence before day 15 to remove any potential transient effects.
 

### 2021-04-16 newvariant.tex
typo
Going to a talk

@@
 Estimating variant epidemic strength and speed remains one of the key questions in controlling the spread and understanding the threat of SARS-CoV-2 variants of concern (VoCs) \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
-Epidemic strength is characterized by the reproduction number $\RR$, which determines the average number of new infections per infection as well as the strength of intervention required to eliminate the disease \citep{anderson1991infectious}.
+Epidemic ``strength'' is measured by the reproduction number $\RR$ -- the average number of new infections caused by a ``typical'' infection \jd{Eliminate from here, stay sharp: as well as the strength of intervention required to eliminate the disease} \jd{Maybe add: disease can spread in a population only if $\RR>1$} \citep{anderson1991infectious} \jd{Cite Heesterbeek (for typical, also because they did the best job of characterizing classical Ro.
+Could drop anderson here, since it's already in the next s.}.
+
 It also provides information about the final size of an epidemic in a homogeneous population \citep{anderson1991infectious}.
@@
 
+\jd{The middle of this s. is a bad place for this caveat. You should talk about it in terms of ``incidence of infection'', I think, and deal with it above here.}
 Epidemic speed $r_x$ can often be estimated directly from case data during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}---in theory both $r$ and $\RR$ relate to latent infections and therefore using case data can lead to biases due to delays in reporting \citep{goldstein2009reconstructing,gostic2020practical}.
@@
 Studies of new SARS-CoV-2 variants have mostly focused on characterizing changes in the \emph{proportion} of a new variant \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
-Assessing frequency dynamics has the advantage of being less sensitive to changes in testing and to other transient effects that would affect variants and wild type viruses similarly.
+Focusing on proportions has the advantage of being less sensitive to changes in testing and to other transient effects that would affect variants and wild type viruses similarly.
 When incidence is changing exponentially ($i_x(t) = i_x(t_0) \exp(r_x t)$), the proportion of the new variant $p(t)$ follows a logistic growth curve:
@@
 \caption{
-\textbf{Relative strength of the new variant conditional on observed relative speed under four epidemiological conditions.}
+\textbf{Relative strength of the new variant conditional on observed relative speed under five epidemiological conditions.}
 The relative strength of the new variant $\rho$ conditional on the speed of the wild type $\rw$, the ratio between the mean generation interval of the new variant $\Gv$ and that of the wild type $\Gw$, and the squared coefficient of variation in generation intervals $\kappa$.
 The relative strength of the new variant $\rho$ is calculated using $\delta=0.1\pday$, $\Gw = 5\days$, and $\kappa = 1/5$.
-Five epidemiological conditions assume $\rw=-0.15\pday$ ($\rw < \rv < 0$), $\rw=-0.1\pday$ ($\rw < \rv = 0$), $\rw=-0.05\pday$ ($\rw < 0 < \rv$), $\rw=0\pday$ ($0 = \rw < \rv$), and $\rw=0.05\pday$ ($0 < \rw < \rv$).
+\jd{If you add $\rw$ to the figure titles, you could skip this sentence and make life easier for readers.} Five epidemiological conditions assume $\rw=-0.15\pday$ ($\rw < \rv < 0$), $\rw=-0.1\pday$ ($\rw < \rv = 0$), $\rw=-0.05\pday$ ($\rw < 0 < \rv$), $\rw=0\pday$ ($0 = \rw < \rv$), and $\rw=0.05\pday$ ($0 < \rw < \rv$).
 }
@@
 Based on the observed relative growth rate of B.1.1.7 in the UK, we start by  taking the relative speed of the variant to be $\delta = 0.1/\textrm{day}$  \citep{davies2021estimated}; the mean generation interval of the wild type to be $\Gw = 5\days$ \citep{ferretti2020quantifying}; and the squared coefficient of variation of generation intervals to be $\kappa=0.2$ \citep{ferretti2020quantifying}.
-We consider the spread of B.1.1.7 as an example but out qualitative conclusions should hold for other VoCs.
+We consider the spread of B.1.1.7 as an example but our qualitative conclusions should hold for other VoCs.
 As noted above, we assume that the variant and the wild type have equal $\kappa$ throughout, and only consider differences in the mean.
-We evaluate the estimates of relative strength $\rho$ across a wide range of $\kappa$ from 0 (fixed generation interval; delta distribution) to 1 (exponential distribution).
+We evaluate the estimates of relative strength $\rho$ across a wide range of $\kappa$ from 0 (fixed-length generation intervals) to 1 (exponential distribution).
 To further explore how inference depends on underlying epidemiological conditions (i.e., whether the incidence of the variant and the wild type are growing or shrinking), we consider five epidemiological conditions: (1) $\rw < \rv < 0$, (2) $\rw < \rv = 0$, (3) $\rw < 0 < \rv$, (4) $0 = \rw < \rv$, and (5) $0 < \rw < \rv$.
+\jd{This looks like the right order to me; it does not seem to match F1.}
 
 \fref{relstrength} illustrates how the relative strength depends on the shape of the generation-interval distributions and on underlying epidemiological conditions.
-In general, longer mean generation interval of the new variant translates to higher values of $\rho$ (and vice versa), except when $\rv \leq 0$.
-When $\rv = 0$ (and therefore $\rw < 0$), we always have $\Rv = 1$ and so the inferred value of $\rho$ is independent of the generation-interval distribution of the new variant.
-When $\rv < 0$ (and therefore $\rw < 0$), we see that longer generation intervals decrease $\rho$ because longer generation intervals actually lead to slower decay (higher $r$).
+In general, longer mean generation interval of the new variant translates to higher values of $\rho$ (and vice versa), except when $\rv \leq 0$ (recall, we always assume $\rw<\rv$).
+When $\rv = 0$, we always have $\Rv = 1$ and so the inferred value of $\rho$ is independent of the generation-interval distribution of the new variant.
+When $\rv < 0$, we see that longer generation intervals decrease $\rho$ because longer generation intervals actually lead to slower decay (higher $r$).
 Assuming a narrower distribution (lower $\kappa$) has qualitatively similar effects as assuming longer generation intervals because both reduce the amount of early transmission; 
@@
 
-We find that an increased speed of $\delta=0.1\pday$ for the variant is consistent with higher strength than the wild type across a wide range of epidemiological conditions \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
+Unsurprisingly, we find that an increased speed of $\delta=0.1\pday$ for the variant is consistent with higher strength than the wild type across the range of epidemiological conditions considered \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
 However, the magnitude of relative strength $\rho$ is sensitive to assumptions about generation intervals.
 For realistic values of $\kappa$ (excluding 0 and 1), we find that the inferred relative strength $\rho$ ranges between 1.1--2.3.
-Even if we restrict the mean generation interval of the variant to differ by 20\% and only consider the scenario under which variant is growing (bottom two panels of \fref{relstrength}), we find that $\rho$ ranges between 1.3--1.8, providing narrower but still large uncertainty.
+Even if we restrict the difference in generation interval to $<20\%$ and only consider scenarios with $\rv>0$ (bottom panels of \fref{relstrength}), we find that $\rho$ ranges between 1.3--1.8.
+\jd{Is it better to consider just one scenario here?
+$\rw=0$?}
 
@@
 \caption{
-\textbf{Relative speed of the new variant given assumed relative strength under four epidemiological conditions.}
+\textbf{Relative speed of the new variant given assumed relative strength under five epidemiological conditions.}
 The relative speed of the new variant $\delta$ given strength of the wild type $\Rw$, ratio between the mean generation interval of the new variant $\Gv$ and that of the wild type $\Gw$, and squared coefficient of variation in generation intervals $\kappa$.
@@
 Finally, we assume $\Rv(t) = \rho \Rw(t)$.
-From simulated incidence curves between day 15 and 70, we estimate the instaneous reproduction number using \eref{rt}.
+From simulated incidence curves between day 15 and 70, we estimate the instantaneous reproduction number using \eref{rt}.
 We ignore incidence before day 15 to remove any potential transient effects.

### 2021-04-22 newvariant.tex
Handing back

@@
 The relative strength of the new variant $\rho$ is calculated using $\delta=0.1\pday$, $\Gw = 5\days$, and $\kappa = 1/5$.
-\jd{If you add $\rw$ to the figure titles, you could skip this sentence and make life easier for readers.} Five epidemiological conditions assume $\rw=-0.15\pday$ ($\rw < \rv < 0$), $\rw=-0.1\pday$ ($\rw < \rv = 0$), $\rw=-0.05\pday$ ($\rw < 0 < \rv$), $\rw=0\pday$ ($0 = \rw < \rv$), and $\rw=0.05\pday$ ($0 < \rw < \rv$).
+\jd{If you add $\rw$ to the figure titles, you could skip this sentence and make life easier for readers.} Five epidemiological conditions assume $\rw=-0.15\pday$ ($\rw < \rv < 0$), $\rw=-0.1\pday$ ($\rw < \rv = 0$), $\rw=-0.05\pday$ ($\rw < 0 < \rv$), $\rw=0$ ($0 = \rw < \rv$), and $\rw=0.05\pday$ ($0 < \rw < \rv$).
 }
@@
 
-Based on the observed relative growth rate of B.1.1.7 in the UK, we start by  taking the relative speed of the variant to be $\delta = 0.1/\textrm{day}$  \citep{davies2021estimated}; the mean generation interval of the wild type to be $\Gw = 5\days$ \citep{ferretti2020quantifying}; and the squared coefficient of variation of generation intervals to be $\kappa=0.2$ \citep{ferretti2020quantifying}.
+Based on the observed relative growth rate of B.1.1.7 in the UK, we start by  taking the relative speed of the variant to be $\delta = 0.1/\pday$  \citep{davies2021estimated}; the mean generation interval of the wild type to be $\Gw = 5\days$ \citep{ferretti2020quantifying}; and the squared coefficient of variation of generation intervals to be $\kappa=0.2$ \citep{ferretti2020quantifying}.
 We consider the spread of B.1.1.7 as an example but our qualitative conclusions should hold for other VoCs.
@@
 \textbf{Relative speed of the new variant given assumed relative strength under five epidemiological conditions.}
-The relative speed of the new variant $\delta$ given strength of the wild type $\Rw$, ratio between the mean generation interval of the new variant $\Gv$ and that of the wild type $\Gw$, and squared coefficient of variation in generation intervals $\kappa$.
+\jd{Again, I'd like to see these ordered better.
+I also still like the idea of putting the value of Rwt in the figure titles (the geometric progression is cute, and we should keep those values, but maybe just mention it briefly and show the values to 2dp.} The relative speed of the new variant $\delta$ given strength of the wild type $\Rw$, ratio between the mean generation interval of the new variant $\Gv$ and that of the wild type $\Gw$, and squared coefficient of variation in generation intervals $\kappa$.
 Relative speed of the new variant $\delta$ is calculated using $\rho=1.61$, $\Gw = 5\days$, and $\kappa = 1/5$.
@@
 Then, we compare the estimated ratio $\hat{\rho}(t) = \hat{\RR}_{\textrm{var}}(t)/\hat{\RR}_{\textrm{wt}}(t)$ with the true ratio $\rho = \Rv(t)/\Rw(t)$.
+%% JD: How does this convoluted LaTeX not make you crazy?
 In order to allow for changes in $\RR$ driven by the introduction and lifting of non-pharmaceutical interventions \citep{flaxman2020Rt}, we model $\Rw(t)$ using a step function (\fref{Rtbias}) and let $\Rv(t) = \rho \Rw(t)$.
@@
 
-When the assumed distribution matches the true distribution, the estimated reproduction numbers match the true values (\fref{Rtbias}A), and therefore the ratio remains constant (\fref{Rtbias}B).
+When the assumed distribution matches the true distribution, the estimated reproduction numbers match the true values (\fref{Rtbias}A); thus, their ratio remains constant (\fref{Rtbias}B), and Rhatvar=Rvar (\fref{Rtbias}B).
 However, if the generation-interval distribution of the variant differs from the assumed distribution, the ratio changes over time (\fref{Rtbias}E,H).
-In particular, if the true generation intervals have a shorter mean than the assumed distribution, we over-estimate $\Rv(t)$ during the growth phase and under-estimate during the decay phase (and vice versa, \fref{Rtbias}D,G), which further translates to biases in the estimated relative strength (\fref{Rtbias}E,H).
-Therefore, biases in the estimated relative strength will correlate with the strength of the variant, (i.e., whether $\Rv(t)$ is below or above 1).
+In particular, if the true generation intervals have a shorter mean than the assumed distribution, we over-estimate $\Rv(t)$ during the growth phase and under-estimate during the decay phase (and conversely, \fref{Rtbias}D,G), which further translates to biases in the estimated relative strength (\fref{Rtbias}E,H).
+\jd{Skip this:} Therefore, biases in the estimated relative strength will correlate with the strength of the variant, (i.e., whether $\Rv(t)$ is below or above 1).
 
@@
 For example, we can estimate $\rho$ by plotting the estimated strength of the variant $\hat{\RR}_{\textrm{var}}(t)$ against the estimated strength of the wild type $\hat{\RR}_{\textrm{wt}}(t)$---as presented in Figure 2 of \cite{volz2021transmission}---and performing a linear regression (\fref{Rtbias}C,F,I).
-When generation-interval distributions are correctly specified, we obtain a straight line with a slope of $\rho$ and intercept at zero (\fref{Rtbias}C).
+If $\rho$ is constant, and generation-interval distributions are correctly specified, we obtain a straight line with a slope of $\rho$ and intercept at zero (\fref{Rtbias}C).
 However, when the assumed mean generation interval is longer than the that of the variant, we over-estimate the slope (and vice versa, \fref{Rtbias}F,I).
+\jd{intercept stuff is long and vague; is that because we're not confident of the pattern?}
 In theory, we expect the regression line to go through the origin, but when the assumed generation-interval distribution differs from the true distribution, this is not necessarily the case: biases in slopes due to mis-specified generation-interval distributions also lead to biases in intercepts.
@@
 an intervention of constant strength, which reduces transmission by a constant factor $\theta$ regardless of age of infection ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau)/\theta$); and an intervention of constant speed, which reduces transmission after infection by a constant rate $\phi$ ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau) \exp(-\phi \tau)$).
-In this case, we can control the spread of the variant when $\theta > \Rv$ or $\phi > \rv$, respectively.
-As a baseline scenario, we consider constant-strength and speed interventions that reduce $\Rv$ to 0.9 when both the variant and the wild type have equal mean generation intervals (\fref{strengthspeed}A--B).
-While both interventions are equally effective on strength-scale (that is, $\Ry{post}=\int  K_{\mathrm{post}}(\tau) \dtau = 0.9$), they have different dynamical implications.
-The constant-strength intervention affects transmission equally throughout the course of infection, whereas the constant-speed intervention has greater impact of transmission that occur later;
+In this case, we can control the spread of the variant when $\theta > \Rv$ or $\phi > \rv$, respectively. [[Cite speedstrength]]
+We consider constant-strength and speed interventions that reduce $\Rv$ to 0.9 when both the variant and the wild type have equal mean generation intervals (\fref{strengthspeed}A--B).
+While both interventions are equally effective on the strength scale (that is, $\Ry{post}=\int  K_{\mathrm{post}}(\tau) \dtau = 0.9$), they have different dynamical implications.
+The constant-strength intervention affects transmission equally throughout the course of infection, whereas the constant-speed intervention has greater impact on transmission that occurs later in infection;
 as a result, the constant-speed intervention reduces the post-intervention mean generation interval (\fref{strengthspeed}B) and leads to (slightly) faster exponential decay (therefore, lower $\ry{post}$).
@@
 \caption{
-\textbf{Effects of constant-strength and speed of interventions against the spread new variant.}
-(A--B) Pre-intervention (black) and post-intervention kernel (colored) of the new variant under constant-strength (A) and speed (B) interventions assuming that the variant has equal generation intervals as the wild type.
-(C--D) Pre-intervention (black) and post-intervention kernel (colored) of the new variant under constant-strength (C) and speed (D) interventions assuming that the variant has longer generation intervals as the wild type.
+\textbf{Effects of constant-strength and constant-speed interventions on the spread of the new variant.}
+\jd{B legend is wrong (-strength)}
+(A--B) Pre-intervention (black) and post-intervention (colored) kernel of the new variant under constant-strength (A) and -speed (B) interventions assuming that the variant has equal generation intervals as the wild type.
+(C--D) Pre-intervention (black) and post-intervention kernel of the new variant assuming that the variant has longer generation intervals than the wild type.
 (E) Pre-intervention (black) and post-intervention strength conditional on the mean generation interval of the variant (colored).
 (F) Pre-intervention (black) and post-intervention speed conditional on the mean generation interval of the variant (colored).
-Strength and speed are calculated assuming$\rw=0\,\textrm{/days}$, $\delta=0.1\,\textrm{/days}$, $\rv=0\,\textrm{/days}$, $\Gy{wt}=5\,\textrm{days}$, and $\kappa=1/5$ for pre-intervention conditions.
-Constant strength and speed interventions are modeled such that post intervention strength of the new variant, assuming equal mean generation intervals at 5 days, corresponds to 0.9.
+\jd{Something is off here; we can't have both $r=0$.} Strength and speed are calculated assuming $\rw=0$, $\delta=0.1\pday$, $\rv=0$, $\Gy{wt}=5\,\textrm{days}$, and $\kappa=1/5$ for pre-intervention conditions.
+Intervention strength and speed are chosen so that post intervention strength of the new variant is 0.9 when its mean generation interval is 5 days.9.
 }
@@
 we have $\ry{post} = \phi - \rv$.
-Therefore, if we can reduce infectiousness at a faster rate than the observed speed of spread (i.e., if the speed of intervention is faster than the epidemic speed, $\phi > \rv$), we can control the epidemic regardless of the underlying generation-interval distribution.
+In particular, if speed of intervention is faster than the observed speed of spread (i.e., if $\phi > \rv$), we can control the epidemic regardless of the underlying generation-interval distribution.
 
@@
 
-We explored how the generation-interval distribution shapes the link between relative strength and speed of the new variant.
+We explored how the generation-interval distribution shapes the link between relative strength and speed of an invadiang disease variant.
 Longer generation intervals generally lead to higher relative strength for a given relative speed (and conversely, lower relative speed for a given relative strength); these can be reversed when incidence is decreasing.
 Neglecting potential differences in the mean generation intervals between the variant and the wild type can bias estimates of the relative strength from incidence data;
-these biases may be systematically assessed by considering whether apparent changes in relative strength over time are associated with underlying epidemiological dynamics of the variant and the wild type.
-Finally, differences in generation intervals can further lead to different conclusions about the effectiveness of interventions:
+these biases may be assessed by considering whether relative strength appears to vary systematically with underlying epidemiological dynamics.
+Finally, differences in generation intervals can also lead to different conclusions about the effectiveness of interventions:
 if the variant has longer (respectively, shorter) generation intervals than the wild type, speed-like (strength-like) interventions, will be more relatively more effective than naive estimates would suggest.
@@
 As the new variants, in particular the B.1.1.7 variant, of SARS-CoV-2 are spreading and becoming dominant in many countries, it is clear that the variants are more transmissible (have higher strength) than the wild type \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
-Our analysis supports estimates of a higher strength of the new variant bounding on a wide range of parameter ranges about epidemiological conditions and generation-interval distributions.
+Our analysis supports estimates of a higher strength of the new variant across a wide range of assumptions about epidemiological conditions and generation-interval distributions.
 However, our analysis also shows that uncertainty in generation-interval distributions must be taken into account to obtain accurate estimates of the relative strength of variants.
@@
 Mathematical analyses have typically reported greater than 1.4 fold increase in reproduction number for the B.1.1.7 variant ($\rho > 1.4$) whereas an independent analysis of secondary attack rates from contact tracing data suggests 1.25--1.4 fold increase \citep{ukinvest}.
-We can ask whether different estimates can be reconciled by generation intervals alone---for simplicity, we assume that the variant and the wild type have identical generation-interval distributions as other studies have.
+We can ask whether different estimates can be reconciled by generation intervals alone---for simplicity, we will assume that the variant and the wild type have identical generation-interval distributions as other studies have.
+
 For example, \cite{davies2021estimated} estimated $\rho=1.77$ assuming a fixed generation generation interval at 5.5 days from $\delta=0.104\pday$ in their regression analysis;
 similarly, \cite{volz2021transmission} estimated $\rho=1.79$ by directly comparing the ratio between $\Rv$ and $\Rw$ assuming a mean generation interval of 6.5 days and a squared coefficient of variation of 0.38.
-While \cite{volz2021transmission} assumed a longer mean generation interval (which typically increases $\rho$), their estimate is similar to \cite{davies2021estimated}, likely because the effect of a wide distribution (which typically reduces $\rho$) can cancel out with the effect of longer intervals---in fact, we obtain $\rho=1.81$ from $\delta=0.104\pday$, $\rw=0\pday$, $\Gv=\Gw=6.4\days$, and $\kappa=0.38$ under the gamma assumption, explaining the lack of differences in the estimates of $\rho$ despite the differences in the assumed generation-interval distributions.
+In this case, the former study's longer mean generation interval (which typically increases $\rho$) is counter-balanced by the effects of their wider distribution.
+\jd{Not clear how to interpret this. Can we match both results under a gamma assumption changing only kappa and Gbar?}
+In fact, we obtain $\rho=1.81$ from $\delta=0.104\pday$, $\rw=0$, $\Gv=\Gw=6.4\days$, and $\kappa=0.38$ under the gamma assumption, explaining the lack of differences in the estimates of $\rho$ despite the differences in the assumed generation-interval distributions.
 
-However, a fixed generation generation interval assumed by \cite{davies2021estimated} is too narrow.
+However, the fixed generation generation interval assumed by \cite{davies2021estimated} is too narrow.
 The mean generation interval of 6.5 days assumed by \cite{volz2021transmission} is also likely too long:
 first, this estimate actually corresponds to the observed mean serial intervals from China in between January and February, 2020 \citep{bi2020epidemiology}, during which they were more likely to observe longer serial intervals due to initial epidemic growth \citep{park2021forward, ali2020serial}; and second, recently infected individuals may be less likely to transmit after symptom onset due to increased awareness of the pandemic compared to earlier on, leading to shorter generation intervals.
-Instead, if we replace the assumed generation-interval distributions for \cite{volz2021transmission} with shorter generation interval estimates from Tianjin, China ($\Gw=2.57\days$ and $\kappa=1$; \cite{ganyani2020estimating}), we obtain $\rho=1.27$, which is consistent with the attack rate analysis \citep{ukinvest}.
+If we replace the assumed generation-interval distributions for \cite{volz2021transmission} with shorter generation interval estimates from Tianjin, China ($\Gw=2.57\days$ and $\kappa=1$; \cite{ganyani2020estimating}), we obtain $\rho=1.27$, which is consistent with the attack rate analysis \citep{ukinvest}.
 We do not claim that these estimates of $\rho$ or $\Gw$ are more accurate;
-instead, this back-of-the-envelope calculation highlights the importance of propagating uncertainty in the generation-interval distribution in assessing the relative strength and speed of SARS-CoV-2 variants.
-Obtaining an accurate and precise estimate of $\rho$ will need to take into account changes in $\rho$ as well as underlying $\rw$.
+instead, this back-of-the-envelope calculation highlights the importance of propagating uncertainty in generation-interval distributions in assessing the relative strength and speed of SARS-CoV-2 variants.
+\jd{I want to drop this here.
+Things are getting tangled, and this \P\ is about the equal-GI case.} Obtaining accurate estimates of $\rho$ will need to take into account changes in $\rho$ as well as underlying $\rw$.
 
@@
 We also assumed that the incidence of infection caused by the variant and the wild type is known.
-However, estimating $\RR(t)$ from real data, such as case incidence, is often associated with several practical challenges, such as delays between infection and reporting and changes in testing patterns \citep{gostic2020practical}.
+However, estimating $\RR(t)$ from real data, such as case incidence, is often associated with a variety of practical challenges, including delays between infection and reporting and changes in testing patterns \citep{gostic2020practical}.
 We have chosen to focus here on the underlying dynamical mechanisms that may impact inference.
@@
 As more information about the transmission and immunity profiles of the new variant becomes available, we advise fixing the relative strength and inferring the speed as this assumption better matches biological mechanisms for its higher strength.
-Researchers should be mindful about which quantity they hold constant and how their assumptions lead to their conclusions \citep{doi:10.1098/rspb.2020.1556}.
+Researchers should be mindful about which quantity they hold constant and how their conclusions follow from their assumptions \citep{doi:10.1098/rspb.2020.1556}.
 
-Here, we lay out a few practical advice in analyzing the epidemiological dynamics of new variants for future studies.
+This study has some practical implications for analyzing the epidemiological dynamics of new variants.
 First, models that assume a constant relative speed, such as standard logistic growth model, should be avoided for characterizing the growth of new variants---the relative speed changes over time depending on epidemiological conditions.
-Second, the absolute strength and speed should not be neglected over their relative values, and both relative and absolute values should be reported together.
+\jd{need to reconcile this with advice above about Early.}
+Second, the absolute strength and speed should not be neglected in favor or relative values, and both relative and absolute values should be discussed together.
 While the relative strength and speed are useful for describing the spread of new variants, the absolute values determine their spread and control.
-In addition, as we showed in our analysis, it is critical to understand the underlying epidemiological conditions of the variant and the wild type in interpreting the relative strength and speed.
+\jd{I changed this s., but maybe we can delete it; it's covered better in the next \P.} In addition, as we showed in our analysis, looking at absolute and relative values together can be critical to understanding dynamics.
 Finally, uncertainty in generation intervals should be carefully considered.
@@
 Observational studies typically focus on serial intervals (i.e., time between symptom onset of the infector and the infectee; \cite{svensson2007note}) because they are easier to measure \citep{griffin2020rapid}.
-Furthermore, realized serial intervals are subject to dynamical biases that can be difficult to tease apart \citep{park2021forward}.
+However, realized serial intervals are subject to dynamical biases that can be difficult to tease apart \citep{park2021forward}.
 A few studies have tried to estimate the generation-interval distribution from serial intervals, with means ranging between 3--6 days and squared coefficients of variation ranging between 0.1--1 \citep{ferretti2020quantifying,Ferretti2020timing,ganyani2020estimating,knight2020estimating};
@@
 
-Since the beginning of the SARS-CoV-2 pandemic, a few studies have emphasized the relevance of leveraging generation interval distributions to improve estimates of strength (e.g., \cite{doi:10.1098/rsif.2020.0144,park2021forward}) and the relative importance of modes of transmission (e.g., asymptomatic vs.
+\jd{If this \P\ is going to be mostly about us, we should say so more clearly.} Since the beginning of the SARS-CoV-2 pandemic, a few studies have emphasized the relevance of leveraging generation interval distributions to improve estimates of strength (e.g., \cite{doi:10.1098/rsif.2020.0144,park2021forward}) and the relative importance of modes of transmission (e.g., asymptomatic vs.
 symptomatic, \cite{park2020time}).
@@
 By explicitly considering epidemiological context and generation interval differences together, we have shown that improving estimates of the the relative duration of infectiousness at the individual scale may represent a pathway towards more effective interventions.
-Specifically, we have shown that speed-like interventions, such as contact tracing, will be more effective if variants have longer generation intervals.
+Specifically, we have shown that speed-like interventions, such as contact tracing, will be relatively more effective if variants have longer generation intervals.
 Most intervention strategies throughout the current pandemic have focused on strength-like interventions \citep{flaxman2020Rt}, such as lock-downs, partly because pre-symptomatic transmission of SARS-CoV2 have limited the effectiveness of contact tracing efforts \citep{hellewell2020feasibility}.

### 2021-04-23 newvariant.tex
done for tonight

@@
 Estimating variant epidemic strength and speed remains one of the key questions in controlling the spread and understanding the threat of SARS-CoV-2 variants of concern (VoCs) \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
-Epidemic ``strength'' is measured by the reproduction number $\RR$ -- the average number of new infections caused by a ``typical'' infection \jd{Eliminate from here, stay sharp: as well as the strength of intervention required to eliminate the disease} \jd{Maybe add: disease can spread in a population only if $\RR>1$} \citep{anderson1991infectious} \jd{Cite Heesterbeek (for typical, also because they did the best job of characterizing classical Ro.
-Could drop anderson here, since it's already in the next s.}.
-
+Epidemic ``strength'' is measured by the reproduction number $\RR$ -- the average number of new infections caused by a ``typical'' infection. 
+Disease can spread in a population only if $\RR>1$ \citep{diekmann1990definition}.
 It also provides information about the final size of an epidemic in a homogeneous population \citep{anderson1991infectious}.
@@
 We use the renewal equation framework to characterize the spread of two pathogen strains---in this case, the wild type SARS-CoV-2 virus and a focal variant of concern, such as B.1.1.7.
-Neglecting the (relatively slow) rate of new mutations, the current incidence of infection $i_x(t)$ caused by strain $x$---wild type (``wt'') and the variant (``var'')---can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary cases are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
+In particular, we focus on characterizing the incidence of \emph{infection} (i.e., the rate at which new infections are generated), which is directly related to $r$ and $\RR$.
+In contrast, observed case reports are subject to reporting delays, which must be taken into account in order to correctly infer $r$ or $\RR$ in practice \citep{goldstein2009reconstructing,gostic2020practical}.
+
+Neglecting the (relatively slow) rate of new mutations, the current incidence of infection $i_x(t)$ caused by strain $x$---wild type (``wt'') and the variant (``var'')---can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary infections are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
 \begin{equation}
@@
 
-\jd{The middle of this s. is a bad place for this caveat. You should talk about it in terms of ``incidence of infection'', I think, and deal with it above here.}
-Epidemic speed $r_x$ can often be estimated directly from case data during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}---in theory both $r$ and $\RR$ relate to latent infections and therefore using case data can lead to biases due to delays in reporting \citep{goldstein2009reconstructing,gostic2020practical}.
+Epidemic speed $r_x$ can often be estimated directly from incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}.
 However, estimating $r_x$ can be challenging when case counts are low (as when a new variant begins to spread) and is sensitive to temporal changes in testing patterns and intensity.
@@
 The relative strength of the new variant $\rho$ is calculated using $\delta=0.1\pday$, $\Gw = 5\days$, and $\kappa = 1/5$.
-\jd{If you add $\rw$ to the figure titles, you could skip this sentence and make life easier for readers.} Five epidemiological conditions assume $\rw=-0.15\pday$ ($\rw < \rv < 0$), $\rw=-0.1\pday$ ($\rw < \rv = 0$), $\rw=-0.05\pday$ ($\rw < 0 < \rv$), $\rw=0$ ($0 = \rw < \rv$), and $\rw=0.05\pday$ ($0 < \rw < \rv$).
+Assumed values of $\rw$ (and therefore $\rv$) are written on top right corners of each panel.
 }
@@
 To further explore how inference depends on underlying epidemiological conditions (i.e., whether the incidence of the variant and the wild type are growing or shrinking), we consider five epidemiological conditions: (1) $\rw < \rv < 0$, (2) $\rw < \rv = 0$, (3) $\rw < 0 < \rv$, (4) $0 = \rw < \rv$, and (5) $0 < \rw < \rv$.
-\jd{This looks like the right order to me; it does not seem to match F1.}
 
@@
 For realistic values of $\kappa$ (excluding 0 and 1), we find that the inferred relative strength $\rho$ ranges between 1.1--2.3.
-Even if we restrict the difference in generation interval to $<20\%$ and only consider scenarios with $\rv>0$ (bottom panels of \fref{relstrength}), we find that $\rho$ ranges between 1.3--1.8.
-\jd{Is it better to consider just one scenario here?
-$\rw=0$?}
+Even if we restrict the difference in generation interval to $<20\%$ and only consider scenarios with $\rv>0$ (bottom panels of \fref{relstrength}), we find that $\rho$ ranges between 1.4--1.8.
 
@@
 \textbf{Relative speed of the new variant given assumed relative strength under five epidemiological conditions.}
-\jd{Again, I'd like to see these ordered better.
-I also still like the idea of putting the value of Rwt in the figure titles (the geometric progression is cute, and we should keep those values, but maybe just mention it briefly and show the values to 2dp.} The relative speed of the new variant $\delta$ given strength of the wild type $\Rw$, ratio between the mean generation interval of the new variant $\Gv$ and that of the wild type $\Gw$, and squared coefficient of variation in generation intervals $\kappa$.
+The relative speed of the new variant $\delta$ given strength of the wild type $\Rw$, ratio between the mean generation interval of the new variant $\Gv$ and that of the wild type $\Gw$, and squared coefficient of variation in generation intervals $\kappa$.
 Relative speed of the new variant $\delta$ is calculated using $\rho=1.61$, $\Gw = 5\days$, and $\kappa = 1/5$.
-Five epidemiological conditions assume $\Rw=1/\rho^{3/2}$ ($\Rw < \Rv < 1$), $\Rw=1/\rho$ ($\Rw < \Rv = 1$), $\Rw=1/\rho^{1/2}$ ($\Rw < 1 < \Rv$), $\Rw=1$ ($1 = \Rw < \Rv$), and $\Rw=\rho^{1/2}$ ($0 < \Rw < \Rv$).
+Assumed values of $\Rw$ (and therefore $\Rv$) are written on top right corners of each panel.
 }
@@
 In particular, if the true generation intervals have a shorter mean than the assumed distribution, we over-estimate $\Rv(t)$ during the growth phase and under-estimate during the decay phase (and conversely, \fref{Rtbias}D,G), which further translates to biases in the estimated relative strength (\fref{Rtbias}E,H).
-\jd{Skip this:} Therefore, biases in the estimated relative strength will correlate with the strength of the variant, (i.e., whether $\Rv(t)$ is below or above 1).
 
@@
 However, when the assumed mean generation interval is longer than the that of the variant, we over-estimate the slope (and vice versa, \fref{Rtbias}F,I).
-\jd{intercept stuff is long and vague; is that because we're not confident of the pattern?}
-In theory, we expect the regression line to go through the origin, but when the assumed generation-interval distribution differs from the true distribution, this is not necessarily the case: biases in slopes due to mis-specified generation-interval distributions also lead to biases in intercepts.
-This analysis suggests another method for assessing potential biases in the estimates of relative strength.
+Biases in slopes further translate to biases in intercepts; in theory, we would expect the regression line to go through the origin (because $\Rv = 0$ when $\Rw = 0$).
 
@@
 \textbf{Effects of constant-strength and constant-speed interventions on the spread of the new variant.}
-\jd{B legend is wrong (-strength)}
 (A--B) Pre-intervention (black) and post-intervention (colored) kernel of the new variant under constant-strength (A) and -speed (B) interventions assuming that the variant has equal generation intervals as the wild type.
@@
 (F) Pre-intervention (black) and post-intervention speed conditional on the mean generation interval of the variant (colored).
-\jd{Something is off here; we can't have both $r=0$.} Strength and speed are calculated assuming $\rw=0$, $\delta=0.1\pday$, $\rv=0$, $\Gy{wt}=5\,\textrm{days}$, and $\kappa=1/5$ for pre-intervention conditions.
+Strength and speed are calculated assuming $\rw=0\pday$, $\delta=0.1\pday$, $\rv=0.1\pday$, $\Gy{wt}=5\,\textrm{days}$, and $\kappa=1/5$ for pre-intervention conditions.
 Intervention strength and speed are chosen so that post intervention strength of the new variant is 0.9 when its mean generation interval is 5 days.9.
@@
 instead, this back-of-the-envelope calculation highlights the importance of propagating uncertainty in generation-interval distributions in assessing the relative strength and speed of SARS-CoV-2 variants.
-\jd{I want to drop this here.
-Things are getting tangled, and this \P\ is about the equal-GI case.} Obtaining accurate estimates of $\rho$ will need to take into account changes in $\rho$ as well as underlying $\rw$.
 
@@
 This study has some practical implications for analyzing the epidemiological dynamics of new variants.
-First, models that assume a constant relative speed, such as standard logistic growth model, should be avoided for characterizing the growth of new variants---the relative speed changes over time depending on epidemiological conditions.
+First, models that assume a constant relative speed, such as standard logistic growth model, should be avoided, if possible, for characterizing the growth of new variants---the relative speed changes over time depending on epidemiological conditions.
+\swp{In some cases, such as early in the spread of new variants, assuming a constant relative speed may not be a bad idea if epidemiological conditions do not change much.
+Such simplifying assumption can also help obtain crude estimates of relative speed.
+However, epidemiological conditions will change over time in response to spreading new variants, causing relative speed to change and invalidating this assumption.}
 \jd{need to reconcile this with advice above about Early.}
@@
 While the relative strength and speed are useful for describing the spread of new variants, the absolute values determine their spread and control.
-\jd{I changed this s., but maybe we can delete it; it's covered better in the next \P.} In addition, as we showed in our analysis, looking at absolute and relative values together can be critical to understanding dynamics.
 Finally, uncertainty in generation intervals should be carefully considered.

### 2021-04-23 control.R

@@ -90,8 +90,8 @@ Gpost_long <- integrate(function(x) x*Rv_base*dgamma(x, shape=1/kappa, rate=1/ka
                         Inf)[[1]]/filter(intervention, Gratio==tail(Gratio,1))$speed_post_strength
 
 g1 <- ggplot(gendata_base) +
-  geom_line(aes(tvec, density, col="Pre-intervention", lty="Pre-intervention"), lwd=2) +
-  geom_line(aes(tvec, strength, col="Post-intervention (constant-strength)", lty="Post-intervention (constant-strength)"), lwd=2) +
+  geom_line(aes(tvec, density, col="Pre-intervention", lty="Pre-intervention"), lwd=3) +
+  geom_line(aes(tvec, strength, col="Post-intervention (constant-strength)", lty="Post-intervention (constant-strength)"), lwd=3) +
   geom_vline(xintercept = Gw, lwd=2) +
   geom_vline(xintercept = Gw, lwd=2, col=colorblind_pal()(3)[3], lty=3) +
   scale_x_continuous("Generation intervals (days)") +
@@ -106,8 +106,8 @@ g1 <- ggplot(gendata_base) +
   )
 
 g2 <- ggplot(gendata_base) +
-  geom_line(aes(tvec, density, col="Pre-intervention", lty="Pre-intervention"), lwd=2) +
-  geom_line(aes(tvec, speed, col="Post-intervention (constant-strength)", lty="Post-intervention (constant-strength)"), lwd=2) +
+  geom_line(aes(tvec, density, col="Pre-intervention", lty="Pre-intervention"), lwd=3) +
+  geom_line(aes(tvec, speed, col="Post-intervention (constant-speed)", lty="Post-intervention (constant-speed)"), lwd=3) +
   geom_vline(xintercept = Gw, lwd=2) +
   geom_vline(xintercept = Gpost_base, col=colorblind_pal()(3)[2], lwd=2, lty=2) +
   scale_x_continuous("Generation intervals (days)") +
@@ -122,8 +122,8 @@ g2 <- ggplot(gendata_base) +
   )
 
 g3 <- ggplot(gendata_long) +
-  geom_line(aes(tvec, density), lwd=2, col=colorblind_pal()(3)[1]) +
-  geom_line(aes(tvec, strength), lwd=2, col=colorblind_pal()(3)[3], lty=3) +
[…]

### 2021-04-23 relspeed.R

@@ -49,6 +49,30 @@ deltadata <- apply(pardata, 1, function(x) {
                        "$\\mathcal{R}_{\\textrm{wt}} < \\mathcal{R}_{\\textrm{var}} = 1$"))
   )
 
+
+textdata <- data.frame(
+  Rw=c("$\\mathcal{R}_{\\textrm{wt}} < \\mathcal{R}_{\\textrm{var}} < 1$",
+       "$\\mathcal{R}_{\\textrm{wt}} < \\mathcal{R}_{\\textrm{var}} = 1$",
+       "$\\mathcal{R}_{\\textrm{wt}} < 1 < \\mathcal{R}_{\\textrm{var}}$",
+       "$1 = \\mathcal{R}_{\\textrm{wt}} < \\mathcal{R}_{\\textrm{var}}$",
+       "$1 < \\mathcal{R}_{\\textrm{wt}} < \\mathcal{R}_{\\textrm{var}}$"),
+  label_Rw=c("$\\mathcal{R}_{\\textrm{wt}}=\\rho^{-3/2}$",
+             "$\\mathcal{R}_{\\textrm{wt}}=\\rho^{-1}$",
+             "$\\mathcal{R}_{\\textrm{wt}}=\\rho^{-1/2}$",
+             "$\\mathcal{R}_{\\textrm{wt}}=1$",
+             "$\\mathcal{R}_{\\textrm{wt}}=\\rho^{1/2}$"),
+  label_Rv=c("$\\mathcal{R}_{\\textrm{var}}=\\rho^{-1/2}$",
+             "$\\mathcal{R}_{\\textrm{var}}=1$",
+             "$\\mathcal{R}_{\\textrm{var}}=\\rho^{1/2}$",
+             "$\\mathcal{R}_{\\textrm{var}}=\\rho$",
+             "$\\mathcal{R}_{\\textrm{var}}=\\rho^{3/2}$")
+)
+
+textdata2 <- data.frame(
+  Rw=c("$\\mathcal{R}_{\\textrm{wt}} < \\mathcal{R}_{\\textrm{var}} < 1$"),
+  label=c("$\\rho=1.61$")
+)
+
 deltadata %>%
   filter(Gratio==1, kappa==0.5)
 
@@ -70,6 +94,9 @@ g1 <- ggplot(deltadata) +
   geom_line(aes(Gratio, delta, lty=kappa, col=kappa), lwd=2) +
[…]

### 2021-04-23 relstrength.R

@@ -50,6 +50,24 @@ rhodata <- apply(pardata, 1, function(x) {
 
 range(filter(rhodata, !kappa %in% c(0, 1))$rho)
 
+textdata <- data.frame(
+  rw=c("$r_{\\textrm{wt}} < 0 < r_{\\textrm{var}}$",
+                     "$0 = r_{\\textrm{wt}} < r_{\\textrm{var}}$",
+                     "$0 < r_{\\textrm{wt}} < r_{\\textrm{var}}$",
+                     "$r_{\\textrm{wt}} < r_{\\textrm{var}} < 0$",
+                     "$r_{\\textrm{wt}} < r_{\\textrm{var}} = 0$"),
+  label_rw=c("$r_{\\textrm{wt}}=-0.05/\\textrm{day}$",
+             "$r_{\\textrm{wt}}=0/\\textrm{day}$",
+             "$r_{\\textrm{wt}}=0.05/\\textrm{day}$",
+             "$r_{\\textrm{wt}}=-0.15/\\textrm{day}$",
+             "$r_{\\textrm{wt}}=-0.1/\\textrm{day}$"),
+  label_rv=c("$r_{\\textrm{var}}=0.05/\\textrm{day}$",
+             "$r_{\\textrm{var}}=0.1/\\textrm{day}$",
+             "$r_{\\textrm{var}}=0.15/\\textrm{day}$",
+             "$r_{\\textrm{var}}=-0.05/\\textrm{day}$",
+             "$r_{\\textrm{var}}=0/\\textrm{day}$")
+)
+
 kappa_title1 <- data.frame(
   x=c(1.3, 1.3),
   y=c(1.5, 2.2),
@@ -68,6 +86,8 @@ g1 <- ggplot(rhodata) +
   geom_line(aes(Gratio, rho, lty=kappa, col=kappa), lwd=2) +
   geom_text(data=kappa_title1, aes(x, y, label=labels), col=(colorblind_pal()(4))[c(1, 4)], size=5) +
   geom_text(data=kappa_title2, aes(x, y, label=labels), col=(colorblind_pal()(4))[c(4, 1)], size=5) +
+  geom_text(data=textdata, aes(x=3/2, y=3.9, label=label_rw), hjust=1, vjust=1) +
+  geom_text(data=textdata, aes(x=3/2, y=3.5, label=label_rv), hjust=1, vjust=1) +
[…]

### 2021-04-23 newvariant.tex
little change with Daniel

@@
 This study has some practical implications for analyzing the epidemiological dynamics of new variants.
-First, models that assume a constant relative speed, such as standard logistic growth model, should be avoided, if possible, for characterizing the growth of new variants---the relative speed changes over time depending on epidemiological conditions.
-\swp{In some cases, such as early in the spread of new variants, assuming a constant relative speed may not be a bad idea if epidemiological conditions do not change much.
+First, models that assume a constant relative speed, such as standard logistic growth model, should be used with care---it is important to remember that the relative speed is expected to change with epidemiological conditions.
+In some cases, such as early in the spread of new variants, assuming a constant relative speed may not be a bad idea if epidemiological conditions do not change much.
 Such simplifying assumption can also help obtain crude estimates of relative speed.
-However, epidemiological conditions will change over time in response to spreading new variants, causing relative speed to change and invalidating this assumption.}
-\jd{need to reconcile this with advice above about Early.}
+However, epidemiological conditions will change over time in response to spreading new variants, causing relative speed to change and invalidating this assumption.
 Second, the absolute strength and speed should not be neglected in favor or relative values, and both relative and absolute values should be discussed together.

### 2021-04-23 newvariant.tex
commit

--- 2021-04-23 newvariant.tex
@@
 an intervention of constant strength, which reduces transmission by a constant factor $\theta$ regardless of age of infection ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau)/\theta$); and an intervention of constant speed, which reduces transmission after infection by a constant rate $\phi$ ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau) \exp(-\phi \tau)$).
-In this case, we can control the spread of the variant when $\theta > \Rv$ or $\phi > \rv$, respectively. [[Cite speedstrength]]
+In this case, we can control the spread of the variant when $\theta > \Rv$ or $\phi > \rv$, respectively \citep{doi:10.1098/rspb.2020.1556}.
 We consider constant-strength and speed interventions that reduce $\Rv$ to 0.9 when both the variant and the wild type have equal mean generation intervals (\fref{strengthspeed}A--B).
@@
 
-We explored how the generation-interval distribution shapes the link between relative strength and speed of an invadiang disease variant.
+We explored how the generation-interval distribution shapes the link between relative strength and speed of an invading disease variant.
 Longer generation intervals generally lead to higher relative strength for a given relative speed (and conversely, lower relative speed for a given relative strength); these can be reversed when incidence is decreasing.
@@
 similarly, \cite{volz2021transmission} estimated $\rho=1.79$ by directly comparing the ratio between $\Rv$ and $\Rw$ assuming a mean generation interval of 6.5 days and a squared coefficient of variation of 0.38.
-In this case, the former study's longer mean generation interval (which typically increases $\rho$) is counter-balanced by the effects of their wider distribution.
-\jd{Not clear how to interpret this. Can we match both results under a gamma assumption changing only kappa and Gbar?}
-In fact, we obtain $\rho=1.81$ from $\delta=0.104\pday$, $\rw=0$, $\Gv=\Gw=6.4\days$, and $\kappa=0.38$ under the gamma assumption, explaining the lack of differences in the estimates of $\rho$ despite the differences in the assumed generation-interval distributions.
+In this case, the latter study's longer mean generation interval (which typically increases $\rho$) is counter-balanced by the effects of their wider distribution, explaining the lack of differences in the estimates of $\rho$ despite the differences in the assumed generation-interval distributions.
 

### 2021-04-23 newvariant.tex
done?

@@
 
-\jd{If this \P\ is going to be mostly about us, we should say so more clearly.} Since the beginning of the SARS-CoV-2 pandemic, a few studies have emphasized the relevance of leveraging generation interval distributions to improve estimates of strength (e.g., \cite{doi:10.1098/rsif.2020.0144,park2021forward}) and the relative importance of modes of transmission (e.g., asymptomatic vs.
+\jd{If this \P\ is going to be mostly about us, we should say so more clearly.} Since the beginning of the SARS-CoV-2 pandemic, a few studies have emphasized the relevance of leveraging generation interval distributions to improve estimates of strength (e.g., \cite{doi:10.1098/rsif.2020.0144,ali2020serial,gostic2020practical,park2021forward}) and the relative importance of modes of transmission (e.g., asymptomatic vs.
 symptomatic, \cite{park2020time}).

### 2021-04-23 newvariant.tex
done?

--- 2021-04-23 newvariant.tex
@@
 }
+\\
+Sang Woo Park\textsuperscript{1,*}
+Benjamin M.\ Bolker\textsuperscript{2,3,4}
+Sebastian Funk\textsuperscript{5,6}
+C.\ Jessica E.\ Metcalf\textsuperscript{1,7}
+Joshua S.\ Weitz\textsuperscript{8,9}
+Bryan T.\ Grenfell\textsuperscript{1,7,10}
+Jonathan Dushoff\textsuperscript{2,3,4}
+\\
+\bigskip
+\textbf{1} Department of Ecology and Evolutionary Biology, Princeton University, Princeton, NJ, USA
+\\
+\textbf{2} Department of Biology, McMaster University, Hamilton, ON, Canada
+\\
+\textbf{3} Department of Mathematics and Statistics, McMaster University, Hamilton, ON, Canada
+\\
+\textbf{4} M.\,G.\,DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, ON, Canada
+\\
+\textbf{5} Department for Infectious Disease Epidemiology, London School of Hygiene and Tropical Medicine, London, UK
+\\
+\textbf{6} Centre for Mathematical Modelling of Infectious Diseases, London School of Hygiene and Tropical Medicine, London, UK
+\\
+\textbf{7} Princeton School of Public and International Affairs, Princeton University, Princeton, NJ, USA
+\\
+\textbf{8} School of Biological Sciences, Georgia Institute of Technology, Atlanta, GA, USA
+\\
+\textbf{9} School of Physics, Georgia Institute of Technology, Atlanta, GA, USA
+\\
+\textbf{10} Fogarty International Center, National Institutes of Health, Bethesda, MD, USA
+\\
+\bigskip
+
+*Corresponding author: swp2@princeton.edu
+\bigskip
+
 \end{flushleft}
@@
 Strength and speed are calculated assuming $\rw=0\pday$, $\delta=0.1\pday$, $\rv=0.1\pday$, $\Gy{wt}=5\,\textrm{days}$, and $\kappa=1/5$ for pre-intervention conditions.
-Intervention strength and speed are chosen so that post intervention strength of the new variant is 0.9 when its mean generation interval is 5 days.9.
+Intervention strength and speed are chosen so that post intervention strength of the new variant is 0.9 when its mean generation interval is 5 days.
 }
@@
 we have $\ry{post} = \phi - \rv$.
-In particular, if speed of intervention is faster than the observed speed of spread (i.e., if $\phi > \rv$), we can control the epidemic regardless of the underlying generation-interval distribution.
+In particular, if speed of intervention is faster than the observed speed of spread (i.e., if $\phi > \rv$), we can control the epidemic (i.e., $\ry{post} < 0$) regardless of the underlying generation-interval distribution.
 
@@
 
-\jd{If this \P\ is going to be mostly about us, we should say so more clearly.} Since the beginning of the SARS-CoV-2 pandemic, a few studies have emphasized the relevance of leveraging generation interval distributions to improve estimates of strength (e.g., \cite{doi:10.1098/rsif.2020.0144,park2021forward}) and the relative importance of modes of transmission (e.g., asymptomatic vs.
+Since the beginning of the SARS-CoV-2 pandemic, a few studies have emphasized the relevance of leveraging generation interval distributions to improve estimates of strength \citep{doi:10.1098/rsif.2020.0144,ali2020serial,gostic2020practical,park2021forward} and the relative importance of modes of transmission (e.g., asymptomatic vs.
 symptomatic, \cite{park2020time}).

### 2021-04-26 newvariant.tex
fix

@@
 Mathematical analyses have typically reported greater than 1.4 fold increase in reproduction number for the B.1.1.7 variant ($\rho > 1.4$) whereas an independent analysis of secondary attack rates from contact tracing data suggests 1.25--1.4 fold increase \citep{ukinvest}.
-We can ask whether different estimates can be reconciled by generation intervals alone---for simplicity, we will assume that the variant and the wild type have identical generation-interval distributions as other studies have.
+We can ask whether different estimates can be reconciled by generation intervals alone.
 
-For example, \cite{davies2021estimated} estimated $\rho=1.77$ assuming a fixed generation generation interval at 5.5 days from $\delta=0.104\pday$ in their regression analysis;
-similarly, \cite{volz2021transmission} estimated $\rho=1.79$ by directly comparing the ratio between $\Rv$ and $\Rw$ assuming a mean generation interval of 6.5 days and a squared coefficient of variation of 0.38.
+\cite{davies2021estimated} estimated $\rho=1.77$ assuming a fixed generation generation interval at 5.5 days from $\delta=0.104\pday$ in their regression analysis.
+Similarly, \cite{volz2021transmission} estimated $\rho=1.79$ by directly comparing the ratio between $\Rv$ and $\Rw$ assuming a mean generation interval of 6.5 days and a squared coefficient of variation of 0.38.
 In this case, the latter study's longer mean generation interval (which typically increases $\rho$) is counter-balanced by the effects of their wider distribution, explaining the lack of differences in the estimates of $\rho$ despite the differences in the assumed generation-interval distributions.
+We note that both studies performed several analyses to estimate $\rho$ and propagated uncertainty in the assumed generation-interval distributions; these two were chosen for illustrative purposes rather than to question the validity of their estimates.
 
-However, the fixed generation generation interval assumed by \cite{davies2021estimated} is too narrow.
+The fixed generation generation interval assumed by \cite{davies2021estimated} is, however, too narrow.
 The mean generation interval of 6.5 days assumed by \cite{volz2021transmission} is also likely too long: 
-first, this estimate actually corresponds to the observed mean serial intervals from China in between January and February, 2020 \citep{bi2020epidemiology}, during which they were more likely to observe longer serial intervals due to initial epidemic growth \citep{park2021forward, ali2020serial}; and second, recently infected individuals may be less likely to transmit after symptom onset due to increased awareness of the pandemic compared to earlier on, leading to shorter generation intervals.
-If we replace the assumed generation-interval distributions for \cite{volz2021transmission} with shorter generation interval estimates from Tianjin, China ($\Gw=2.57\days$ and $\kappa=1$; \cite{ganyani2020estimating}), we obtain $\rho=1.27$, which is consistent with the attack rate analysis \citep{ukinvest}.
-We do not claim that these estimates of $\rho$ or $\Gw$ are more accurate;
-instead, this back-of-the-envelope calculation highlights the importance of propagating uncertainty in generation-interval distributions in assessing the relative strength and speed of SARS-CoV-2 variants.
+first, this estimate actually corresponds to the observed mean serial intervals from China in between January and February, 2020 \citep{bi2020epidemiology}, during which they were more likely to observe longer serial intervals due to initial epidemic growth \citep{park2021forward, ali2020serial}; and second, recently infected individuals may be less likely to transmit after symptom onset due to increased awareness of the pandemic, leading to shorter generation intervals.
+For example, \cite{davies2021estimated} showed that using shorter generation interval estimates from \cite{ganyani2020estimating} ($\Gw=\Gv=3.6\days$ and $\kappa=0\ \textrm{or}\ 0.7$) can lead to considerable reduction in the estimates of $\rho$ (see Supplementary Table S1 in the original text)---in the case of their regression analysis, $\rho$ decreased from 1.77 to 1.46 (assuming $\kappa = 0$).
+Similarly, \cite{volz2021transmission} considered the possibility that the variant may have shorter generation intervals ($\Gv=0.75\Gw$), decreasing $\rho$ from 1.79 to 1.60.
+If we consider even shorter generation interval estimates from Tianjin, China ($\Gw=\Gv=2.57\days$ and $\kappa=1$; \cite{ganyani2020estimating}), we obtain $\rho=1.27$ (from $\delta=0.104\pday$ and $\rw = 0\pday$; \cite{davies2021estimated}).
+Although these estimates are more consistent with the attack rate analysis \citep{ukinvest},
+we do not claim that these estimates of $\rho$ are more accurate.
+Instead, this back-of-the-envelope calculation highlights the importance of propagating uncertainty in generation-interval distributions in assessing the relative strength and speed of SARS-CoV-2 variants.
 

### 2021-04-26 newvariant.tex
another minor edit

@@
 Specifically, we have shown that speed-like interventions, such as contact tracing, will be relatively more effective if variants have longer generation intervals.
-Most intervention strategies throughout the current pandemic have focused on strength-like interventions \citep{flaxman2020Rt}, such as lock-downs, partly because pre-symptomatic transmission of SARS-CoV2 have limited the effectiveness of contact tracing efforts \citep{hellewell2020feasibility}.
+Most intervention strategies throughout the current pandemic have focused on strength-like interventions \citep{flaxman2020Rt}, such as lock-downs, partly because pre-symptomatic transmission of SARS-CoV-2 has limited the effectiveness of contact tracing efforts \citep{hellewell2020feasibility}.
 However, given the possibility that new variants can have different infection characteristics, future studies should consider whether their transmission dynamics also differ (e.g., the amount of pre-symptomatic transmission) and evaluate intervention strategies accordingly.

### 2021-04-30 newvariant.tex
tiny edits

@@
 While relative speed $\delta$ and strength $\rho$ are useful for characterizing the spread of the variant in an epidemiological context with a previously dominant wild type, \emph{absolute} speed $\rv$ and strength $\Rv$ of the variant are the main variables that determine the spread and conditions for control of the variant over the long term.
-In particular, at any given point in epidemic, we can measure the speed of the variant $\rv$ (or infer $\rv$ from $\rw$ and $\delta$) and ask how much more intervention is required to control the spread of both strains (since $\Rv < 1$ implies $\Rw < 1$).
+In particular, at any given point in the epidemic, we can measure the speed of the variant $\rv$ (or infer $\rv$ from $\rw$ and $\delta$) and ask how much more intervention is required to control the spread of both strains (since $\Rv < 1$ implies $\Rw < 1$).
 As a baseline scenario, we assume $\rw=0$ and $\delta=0.1\pday$ (and therefore $\rv=0.1\pday$), in which case additional intervention is required to reduce $\rv$ below 0 (or, equivalently, $\Rv$ below 1).
@@
 There is currently a (minor) discrepancy in the estimates of relative strength of new variants, particularly for B.1.1.7.
-Mathematical analyses have typically reported greater than 1.4 fold increase in reproduction number for the B.1.1.7 variant ($\rho > 1.4$) whereas an independent analysis of secondary attack rates from contact tracing data suggests 1.25--1.4 fold increase \citep{ukinvest}.
+Mathematical analyses have typically reported greater than a 1.4 fold increase in reproduction number for the B.1.1.7 variant ($\rho > 1.4$) whereas an independent analysis of secondary attack rates from contact tracing data suggests 1.25--1.4 fold increase \citep{ukinvest}.
 We can ask whether different estimates can be reconciled by generation intervals alone.
@@
 This study has some practical implications for analyzing the epidemiological dynamics of new variants.
-First, models that assume a constant relative speed, such as standard logistic growth model, should be used with care---it is important to remember that the relative speed is expected to change with epidemiological conditions.
+First, models that assume a constant relative speed, such as the standard logistic growth model, should be used with care---it is important to remember that the relative speed is expected to change with epidemiological conditions.
 In some cases, such as early in the spread of new variants, assuming a constant relative speed may not be a bad idea if epidemiological conditions do not change much.

### 2021-04-30 newvariant.tex
various edits

@@
 	\textbf\newline{
-		Roles of generation-interval distributions in characterizing relative epidemic strength and speed of new SARS-CoV-2 variants and their control
+		Roles of generation-interval distributions in shaping relative epidemic strength, speed, and control of new SARS-CoV-2 variants
 	}
@@
 Multiple studies have estimated the relative strength of new variants from the observed relative speed, but they typically neglect the possibility that the new variants have different generation intervals (i.e., time between infection and transmission), which determines the relationship between relative strength and speed.
-However, the B.1.1.7 strain may have a longer infectious period (and therefore, generation interval) than prior dominant lineages.
+Notably, the increasingly predominant B.1.1.7 variant may have a longer infectious period (and therefore, generation interval) than prior dominant lineages.
 Here, we explore how differences in generation intervals between a new variant and the wild type affect the relationship between relative strength and speed.
-We use simulations to show how neglecting their differences can lead to biases and illustrate how such biases can be assessed in practice.
+We use simulations to show how neglecting these differences can lead to biases in estimates of relative strength  in practice and illustrate how such biases can be assessed.
 Finally, we discuss implications for control: if new variants have longer generation intervals then speed-like interventions (e.g., contact tracing) become more effective, whereas strength-like interventions become less effective (e.g., social distancing).
@@
 
+\swp{1/day or 1/days?}
+
 Estimating variant epidemic strength and speed remains one of the key questions in controlling the spread and understanding the threat of SARS-CoV-2 variants of concern (VoCs) \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
 Epidemic ``strength'' is measured by the reproduction number $\RR$ -- the average number of new infections caused by a ``typical'' infection. 
-Disease can spread in a population only if $\RR>1$ \citep{diekmann1990definition}.
-It also provides information about the final size of an epidemic in a homogeneous population \citep{anderson1991infectious}.
+Disease can spread in a population if $\RR>1$ \citep{diekmann1990definition}.
+The strength also provides information about the final size of an epidemic in a homogeneous population \citep{anderson1991infectious}.
 Epidemic speed is characterized by the growth rate $r$, which describes how fast a disease spreads at the population-level.
-Like epidemic strength, epidemic speed also determines conditions for disease elimination $r=0$ is a threshold equivalent to $\RR=1$ \citep{doi:10.1098/rspb.2020.1556}.
+Like epidemic strength, epidemic speed also determines conditions for disease elimination: $r=0$ is a threshold equivalent to $\RR=1$.
 Strength and speed are linked by generation intervals---defined as time between infection and transmission---which characterizes the individual-level time scale of the epidemic \citep{roberts2007model,svensson2007note,wallinga2007generation}.
@@
 \end{equation}
-The Gamma assumption includes as a special case models that assume exponentially distributed generation intervals (when $\kappa=1$), corresponding to the SIR model \citep{anderson1991infectious}; various gamma assumptions and approximations are widely used in epidemic modeling, including for models of SARS-CoV-2 \citep{doi:10.1098/rsif.2020.0144}.
+The Gamma assumption includes as a special case models that assume exponentially distributed generation intervals (when $\kappa=1$), corresponding to the SIR model \citep{anderson1991infectious}; various Gamma assumptions are widely used in epidemic modeling, including for models of SARS-CoV-2 \citep{doi:10.1098/rsif.2020.0144}.
 We use this framework to investigate how inferences about strength and speed of the variant depend on our assumptions about the underlying generation-interval distributions.
@@
 Inference of $\rho$ from $\delta$ will depend on assumptions about the generation-interval distributions of both strains.
-Given the mean generation interval of the variant $\Gv$ and the wild type $\Gw$, the relative strength $\rho = \Rv/\Rw$ under the gamma approximation \citep{park2019practical} is given by:
+Given the mean generation interval of the variant $\Gv$ and the wild type $\Gw$, the relative strength $\rho = \Rv/\Rw$ under the Gamma assumption \citep{park2019practical} is given by:
 \begin{equation}
@@
 
-Based on the observed relative growth rate of B.1.1.7 in the UK, we start by  taking the relative speed of the variant to be $\delta = 0.1/\pday$  \citep{davies2021estimated}; the mean generation interval of the wild type to be $\Gw = 5\days$ \citep{ferretti2020quantifying}; and the squared coefficient of variation of generation intervals to be $\kappa=0.2$ \citep{ferretti2020quantifying}.
+Based on the observed relative growth rate of B.1.1.7 in the UK, we start by  taking the relative speed of the variant to be $\delta = 0.1\pday$  \citep{davies2021estimated}; the mean generation interval of the wild type to be $\Gw = 5\days$ \citep{ferretti2020quantifying}; and the squared coefficient of variation of generation intervals to be $\kappa=0.2$ \citep{ferretti2020quantifying}.
 We consider the spread of B.1.1.7 as an example but our qualitative conclusions should hold for other VoCs.
@@
 \fref{relstrength} illustrates how the relative strength depends on the shape of the generation-interval distributions and on underlying epidemiological conditions.
-In general, longer mean generation interval of the new variant translates to higher values of $\rho$ (and vice versa), except when $\rv \leq 0$ (recall, we always assume $\rw<\rv$).
+In general, longer mean generation intervals of the new variant translates to higher values of $\rho$ (and vice versa), except when $\rv \leq 0$ (recall, we always assume $\rw<\rv$).
 When $\rv = 0$, we always have $\Rv = 1$ and so the inferred value of $\rho$ is independent of the generation-interval distribution of the new variant.
@@
 We do not generally expect the relative speed $\delta$ to remain constant if other factors governing epidemic spread are changing.
-Indeed, many biological mechanisms better translate to assuming a constant value of relative strength $\rho$ over changing conditions.
+Instead, many biological mechanisms appear compatible to assuming a constant value of relative strength $\rho$ over changing conditions.
 For example, if the proportion of the population susceptible declines, or the average contact rate changes, while other factors remain constant, the relative strength $\rho$ is expected to remain constant;
@@
 We thus investigate how $\delta$ is expected to change with \Rw if $\rho$ remains constant, and how this expectation changes with the ratio of the generation intervals. 
-Once again, we rely on the gamma approximation to find the relative speed $\delta$ given the mean generation interval of the variant $\Gv$ and the wild type $\Gw$, :
+Once again, we rely on the Gamma assumption to find the relative speed $\delta$ given the mean generation interval of the variant $\Gv$ and the wild type $\Gw$, :
 \begin{equation}
@@
 \end{equation}
-As our baseline scenario we assume $\rho = 1.61$, which is the value we obtain for $\delta=0.1\,\textrm{days}$, $\Gw = \Gv = 5\,\textrm{days}$, and $\kappa = 1/5$.
+As our baseline scenario we assume $\rho = 1.61$, which is the value we obtain for $\delta=0.1\pday$, $\Gw = \Gv = 5\,\textrm{days}$, and $\kappa = 1/5$.
 We evaluate $\delta$ across five different epidemiological conditions as before: (1) $\Rw < \Rv < 1$, (2) $\Rw < \Rv = 1$, (3) $\Rw < 1 < \Rv$, (4) $1 = \Rw < \Rv$, and (5) $1 < \Rw < \Rv$.
@@
 \end{figure}
+\swp{This figure also needs to be fixed}
 
-When the assumed distribution matches the true distribution, the estimated reproduction numbers match the true values (\fref{Rtbias}A); thus, their ratio remains constant (\fref{Rtbias}B), and Rhatvar=Rvar (\fref{Rtbias}B).
+When the assumed distribution matches the true distribution, the estimated reproduction numbers match the true values (\fref{Rtbias}A); thus, their ratio remains constant and $\hat{\RR}_{\textrm{var}}(t)=\Rv(t)$ (\fref{Rtbias}B).
 However, if the generation-interval distribution of the variant differs from the assumed distribution, the ratio changes over time (\fref{Rtbias}E,H).
-In particular, if the true generation intervals have a shorter mean than the assumed distribution, we over-estimate $\Rv(t)$ during the growth phase and under-estimate during the decay phase (and conversely, \fref{Rtbias}D,G), which further translates to biases in the estimated relative strength (\fref{Rtbias}E,H).
+In particular, if the true generation intervals of the variant have a shorter mean than the assumed distribution, we over-estimate $\Rv(t)$ during the growth phase and under-estimate during the decay phase (and conversely, \fref{Rtbias}D,G), which further translates to biases in the estimated relative strength (\fref{Rtbias}E,H).
 
@@
 \end{figure}
+\swp{JSW: I guess we can't say these are apples-to-apples comparisons...
+because it seems that constant speed is always better; always control, whereas constant strength does not...
+I see that when ratio of means <1 then better to do strength but still notable that speed ALWAYS works here.}
+\swp{He also said something about constant speed not being constant effort}
 
@@
 
-The speed-based paradigm gives the same results about the control but provides important insights (\fref{strengthspeed}F).
+The speed-based paradigm gives the same results regarding control but provides important insights (\fref{strengthspeed}F).
 The observed speed of the variant $\rv$ at a given moment is independent of our estimates of its mean generation interval.
@@
 
-\cite{davies2021estimated} estimated $\rho=1.77$ assuming a fixed generation generation interval at 5.5 days from $\delta=0.104\pday$ in their regression analysis.
+\cite{davies2021estimated} estimated a relative strength of $\rho=1.77$ assuming a fixed generation generation interval at 5.5 days from $\delta=0.104\pday$ in their regression analysis.
 Similarly, \cite{volz2021transmission} estimated $\rho=1.79$ by directly comparing the ratio between $\Rv$ and $\Rw$ assuming a mean generation interval of 6.5 days and a squared coefficient of variation of 0.38.
@@
 we do not claim that these estimates of $\rho$ are more accurate.
-Instead, this back-of-the-envelope calculation highlights the importance of propagating uncertainty in generation-interval distributions in assessing the relative strength and speed of SARS-CoV-2 variants.
+Instead, this illustrative calculation highlights the importance of propagating uncertainty in generation-interval distributions in assessing the relative strength and speed of SARS-CoV-2 variants.
+\swp{Let's talk.}
 
@@
 however, this is only true under strength-like interventions, which is insensitive to age of infection, but not under speed-like interventions. 
-As we demonstrated earlier, the effectiveness of speed-like interventions depend on the shape of the generation-interval distribution;
+As we demonstrated, the effectiveness of speed-like interventions depend on the shape of the generation-interval distribution;
 therefore, if the variant and the wild type have different generation intervals, speed-like interventions, such as contact tracing, can affect them differently.
@@
 However, epidemiological conditions will change over time in response to spreading new variants, causing relative speed to change and invalidating this assumption.
-Second, the absolute strength and speed should not be neglected in favor or relative values, and both relative and absolute values should be discussed together.
+Second, the absolute strength and speed should not be neglected in favor of relative values, and both relative and absolute values should be discussed together.
 While the relative strength and speed are useful for describing the spread of new variants, the absolute values determine their spread and control.
@@
 Future studies should prioritize detailed assessment of the generation intervals of SARS-CoV-2 and widespread variants, as well as consider how uncertainty in generational intervals might bias conclusions.
+
+\swp{perhaps comment that, if new variants escape vaccinal or natural immunity to some extent (ref), this will alter the calculation: eg increase strength?}
 

### 2021-04-30 relspeed.R

@@ -101,7 +101,7 @@ g1 <- ggplot(deltadata) +
                 breaks=c(2/3, 1, 3/2),
                 labels=c("2/3", 1, "3/2"),
                 expand=c(0, 0.03)) +
-  scale_y_continuous("Relative speed, $\\delta$ (1/day)", limits=c(0, 0.3), expand=c(0, 0),
+  scale_y_continuous("Relative speed, $\\delta$ (1/days)", limits=c(0, 0.3), expand=c(0, 0),
                      breaks=c(0, 0.1, 0.2, 0.3),
                      labels=c(0, 0.1, 0.2, 0.3)) +
   facet_wrap(~Rw, as.table=F) +

### 2021-04-30 newvariant.tex
Meeting with 朴

@@
 \section{Introduction}
-
-\swp{1/day or 1/days?}
 
@@
 (B, E, H) True (purple, solid) and estimated (orange, dashed) ratios between reproduction numbers of the new variant and the wild type over time.
-(C, F, I) True (purple, solid) and estimated (orange, dashed) relationships between estimated reproduction numbers.
+(C, F, I) Phase planes (time is implicit) showing true (purple, solid) and estimated (orange, dashed) relationships between estimated reproduction numbers.
 Blue dotted lines represent the regression lines of the estimated variant reproduction numbers against the estimated wild type reproduction numbers.
@@
 \end{figure}
-\swp{This figure also needs to be fixed}
 
@@
 \caption{
-\textbf{Effects of constant-strength and constant-speed interventions on the spread of the new variant.}
+\textbf{Effects of constant-strength and constant-speed interventions on the spread of a new variant with known speed \rv.}
 (A--B) Pre-intervention (black) and post-intervention (colored) kernel of the new variant under constant-strength (A) and -speed (B) interventions assuming that the variant has equal generation intervals as the wild type.
@@
 \end{figure}
-\swp{JSW: I guess we can't say these are apples-to-apples comparisons...
-because it seems that constant speed is always better; always control, whereas constant strength does not...
-I see that when ratio of means <1 then better to do strength but still notable that speed ALWAYS works here.}
-\swp{He also said something about constant speed not being constant effort}
 
@@
 Similarly, \cite{volz2021transmission} estimated $\rho=1.79$ by directly comparing the ratio between $\Rv$ and $\Rw$ assuming a mean generation interval of 6.5 days and a squared coefficient of variation of 0.38.
-In this case, the latter study's longer mean generation interval (which typically increases $\rho$) is counter-balanced by the effects of their wider distribution, explaining the lack of differences in the estimates of $\rho$ despite the differences in the assumed generation-interval distributions.
-We note that both studies performed several analyses to estimate $\rho$ and propagated uncertainty in the assumed generation-interval distributions; these two were chosen for illustrative purposes rather than to question the validity of their estimates.
+In this case, the latter study's longer mean generation interval (which increases $\rho$) is counter-balanced by the effects of their wider distribution (which decreases $\rho$).
+We note that both studies performed several analyses to estimate $\rho$ and propagated uncertainty in the assumed generation-interval distributions; these examples were chosen for illustrative purposes rather than to question the validity of their estimates.
 

### 2021-04-30 newvariant.tex
how about this

@@
 Mathematical analyses have typically reported greater than a 1.4 fold increase in reproduction number for the B.1.1.7 variant ($\rho > 1.4$) whereas an independent analysis of secondary attack rates from contact tracing data suggests 1.25--1.4 fold increase \citep{ukinvest}.
-We can ask whether different estimates can be reconciled by generation intervals alone.
+As shown in \cite{davies2021estimated} and \cite{volz2021transmission}, propagating uncertainty in generation interval estimates (and, in particular, assuming shorter generation intervals) can help explain these differences to some extent.
 
-\cite{davies2021estimated} estimated a relative strength of $\rho=1.77$ assuming a fixed generation generation interval at 5.5 days from $\delta=0.104\pday$ in their regression analysis.
+Using the Gamma framework \citep{doi:10.1098/rsif.2020.0144} can help us evaluate these differences more clearly.
+For example, \cite{davies2021estimated} estimated a relative strength of $\rho=1.77$ assuming a fixed generation generation interval at 5.5 days from $\delta=0.104\pday$ in one of their regression analysis.
 Similarly, \cite{volz2021transmission} estimated $\rho=1.79$ by directly comparing the ratio between $\Rv$ and $\Rw$ assuming a mean generation interval of 6.5 days and a squared coefficient of variation of 0.38.
 In this case, the latter study's longer mean generation interval (which increases $\rho$) is counter-balanced by the effects of their wider distribution (which decreases $\rho$).
-We note that both studies performed several analyses to estimate $\rho$ and propagated uncertainty in the assumed generation-interval distributions; these examples were chosen for illustrative purposes rather than to question the validity of their estimates.
-
-The fixed generation generation interval assumed by \cite{davies2021estimated} is, however, too narrow.
-The mean generation interval of 6.5 days assumed by \cite{volz2021transmission} is also likely too long: 
-first, this estimate actually corresponds to the observed mean serial intervals from China in between January and February, 2020 \citep{bi2020epidemiology}, during which they were more likely to observe longer serial intervals due to initial epidemic growth \citep{park2021forward, ali2020serial}; and second, recently infected individuals may be less likely to transmit after symptom onset due to increased awareness of the pandemic, leading to shorter generation intervals.
-For example, \cite{davies2021estimated} showed that using shorter generation interval estimates from \cite{ganyani2020estimating} ($\Gw=\Gv=3.6\days$ and $\kappa=0\ \textrm{or}\ 0.7$) can lead to considerable reduction in the estimates of $\rho$ (see Supplementary Table S1 in the original text)---in the case of their regression analysis, $\rho$ decreased from 1.77 to 1.46 (assuming $\kappa = 0$).
-Similarly, \cite{volz2021transmission} considered the possibility that the variant may have shorter generation intervals ($\Gv=0.75\Gw$), decreasing $\rho$ from 1.79 to 1.60.
-If we consider even shorter generation interval estimates from Tianjin, China ($\Gw=\Gv=2.57\days$ and $\kappa=1$; \cite{ganyani2020estimating}), we obtain $\rho=1.27$ (from $\delta=0.104\pday$ and $\rw = 0\pday$; \cite{davies2021estimated}).
+Instead, if we consider short and wide generation interval estimates from Tianjin, China ($\Gw=\Gv=2.57\days$ and $\kappa=1$; \cite{ganyani2020estimating}), we obtain $\rho=1.27$ (from $\delta=0.104\pday$ and $\rw = 0\pday$; \cite{davies2021estimated}).
 Although these estimates are more consistent with the attack rate analysis \citep{ukinvest},
@@
 Instead, this illustrative calculation highlights the importance of propagating uncertainty in generation-interval distributions in assessing the relative strength and speed of SARS-CoV-2 variants.
-\swp{Let's talk.}
 

### 2021-04-30 newvariant.tex
done?

@@
 
-Using the Gamma framework \citep{doi:10.1098/rsif.2020.0144} can help us evaluate these differences more clearly.
+Using the Gamma framework \citep{doi:10.1098/rsif.2020.0144} can help us reconcile different estimates more clearly.
 For example, \cite{davies2021estimated} estimated a relative strength of $\rho=1.77$ assuming a fixed generation generation interval at 5.5 days from $\delta=0.104\pday$ in one of their regression analysis.
@@
 we do not claim that these estimates of $\rho$ are more accurate.
-Instead, this illustrative calculation highlights the importance of propagating uncertainty in generation-interval distributions in assessing the relative strength and speed of SARS-CoV-2 variants.
+This illustrative calculation highlights the importance of propagating uncertainty in generation-interval distributions in assessing the relative strength and speed of SARS-CoV-2 variants.
 

### 2021-04-30 newvariant.tex
Fiddling

@@
 
-Using the Gamma framework \citep{doi:10.1098/rsif.2020.0144} can help us reconcile different estimates more clearly.
+Using the gamma-approximation framework \citep{doi:10.1098/rsif.2020.0144} can help us reconcile different estimates more clearly.
 For example, \cite{davies2021estimated} estimated a relative strength of $\rho=1.77$ assuming a fixed generation generation interval at 5.5 days from $\delta=0.104\pday$ in one of their regression analysis.
 Similarly, \cite{volz2021transmission} estimated $\rho=1.79$ by directly comparing the ratio between $\Rv$ and $\Rw$ assuming a mean generation interval of 6.5 days and a squared coefficient of variation of 0.38.
-In this case, the latter study's longer mean generation interval (which increases $\rho$) is counter-balanced by the effects of their wider distribution (which decreases $\rho$).
+In this case, the latter study's longer mean generation interval (which increases $\rho$) is counter-balanced by the effects of their wider distribution (which decreases $\rho$); calculations using our approximation framework produce qualitatively similar results.
 Instead, if we consider short and wide generation interval estimates from Tianjin, China ($\Gw=\Gv=2.57\days$ and $\kappa=1$; \cite{ganyani2020estimating}), we obtain $\rho=1.27$ (from $\delta=0.104\pday$ and $\rw = 0\pday$; \cite{davies2021estimated}).

### 2021-04-30 newvariant.tex
back to writing generals proposal

--- 2021-04-30 newvariant.tex
@@
 If the variant has a longer infectious period \citep{kissler2021densely}, and therefore extended generation-interval distributions, current estimates may be underestimating the true relative strength.
+If new variants escape vaccinal or natural immunity to some extent \citep{shaman2020will}, their relative strength may increase further.
 
@@
 
-Using the Gamma framework \citep{doi:10.1098/rsif.2020.0144} can help us reconcile different estimates more clearly.
+Using the Gamma-approximation framework \citep{doi:10.1098/rsif.2020.0144} can help us reconcile different estimates more clearly.
 For example, \cite{davies2021estimated} estimated a relative strength of $\rho=1.77$ assuming a fixed generation generation interval at 5.5 days from $\delta=0.104\pday$ in one of their regression analysis.
 Similarly, \cite{volz2021transmission} estimated $\rho=1.79$ by directly comparing the ratio between $\Rv$ and $\Rw$ assuming a mean generation interval of 6.5 days and a squared coefficient of variation of 0.38.
-In this case, the latter study's longer mean generation interval (which increases $\rho$) is counter-balanced by the effects of their wider distribution (which decreases $\rho$).
-Instead, if we consider short and wide generation interval estimates from Tianjin, China ($\Gw=\Gv=2.57\days$ and $\kappa=1$; \cite{ganyani2020estimating}), we obtain $\rho=1.27$ (from $\delta=0.104\pday$ and $\rw = 0\pday$; \cite{davies2021estimated}).
+In this case, the latter study's longer mean generation interval (which increases $\rho$) is counter-balanced by the effects of their wider distribution (which decreases $\rho$); calculations using our approximation framework produce qualitatively similar results ($\rho=1.81$ from $\Gw=\Gv=6.4\days$, $\kappa=0.38$, $\delta=0.104\pday$, and $\rw=0\pday$).
+If we consider short and wide generation interval estimates from Tianjin, China ($\Gw=\Gv=2.57\days$ and $\kappa=1$; \cite{ganyani2020estimating}), we obtain $\rho=1.27$ (from $\delta=0.104\pday$ and $\rw = 0\pday$).
 Although these estimates are more consistent with the attack rate analysis \citep{ukinvest},
@@
 
-\swp{perhaps comment that, if new variants escape vaccinal or natural immunity to some extent (ref), this will alter the calculation: eg increase strength?}
-
 The spread of new SARS-CoV-2 variants and the replacement of previously dominant lineages represent ongoing challenges for controlling the SARS-CoV-2 pandemic.  

### 2021-05-01 newvariant.tex
done?

@@
 \caption{
-\textbf{Relative strength of the new variant conditional on observed relative speed under five epidemiological conditions.}
+\textbf{Relative strength of the new variant assuming a fixed speed advantage $\delta$ under five epidemiological conditions.}
 The relative strength of the new variant $\rho$ conditional on the speed of the wild type $\rw$, the ratio between the mean generation interval of the new variant $\Gv$ and that of the wild type $\Gw$, and the squared coefficient of variation in generation intervals $\kappa$.
@@
 \caption{
-\textbf{Relative speed of the new variant given assumed relative strength under five epidemiological conditions.}
+\textbf{Relative speed of the new variant assuming a fixed strength advantage $\rho$ u under five epidemiological conditions.}
 The relative speed of the new variant $\delta$ given strength of the wild type $\Rw$, ratio between the mean generation interval of the new variant $\Gv$ and that of the wild type $\Gw$, and squared coefficient of variation in generation intervals $\kappa$.
@@
 Then, we compare the estimated ratio $\hat{\rho}(t) = \hat{\RR}_{\textrm{var}}(t)/\hat{\RR}_{\textrm{wt}}(t)$ with the true ratio $\rho = \Rv(t)/\Rw(t)$.
-%% JD: How does this convoluted LaTeX not make you crazy?
-In order to allow for changes in $\RR$ driven by the introduction and lifting of non-pharmaceutical interventions \citep{flaxman2020Rt}, we model $\Rw(t)$ using a step function (\fref{Rtbias}) and let $\Rv(t) = \rho \Rw(t)$.
+In order to allow for changes in $\RR$ driven by the introduction and lifting of non-pharmaceutical interventions, we let $\Rw(t)$ to decrease from 2 to 0.4 around day 30 and increase back up to 1 around day 60 and assume $\Rv(t) = \rho \Rw(t)$.
+Although previous studies have modeled the impact of non-pharmaceutical interventions as a step function \citep{flaxman2020Rt}, we choose to use a smooth function to model $\Rw(t)$ (\fref{Rtbias}; see Methods) given the possibility that behavioral changes may affect transmission before and after interventions take place.
+In Supplementary Figure S1, we show that using a step function does not affect our conclusions.
 
 \begin{figure}[!pht]
-\includegraphics[width=\textwidth]{Rtbias.pdf}
+\begin{center}
+\includegraphics[width=0.9\textwidth]{Rtbias_smooth.pdf}
 \caption{
@@
 }
+\end{center}
 \label{fig:Rtbias}
@@
 We discretize the model at the time scale of 0.025 days as described in \citep{park2021forward} and simulated the model with $\iv(0) = 0.001$ and $\iw(0) = 0.1$.
-We assume that $\Rw(t)$ remains at 2 for 30 days, immediately decreases and stays at 0.5 until day 60, and immediately increases back to 1.2.
+To allow for smooth changes in $\Rw(t)$ around day 30 and 60, we let:
+\begin{equation}
+\Rw(t) = 2 \times \left[1 - 0.8 \left(\frac{1}{2} + \frac{\arctan(t-30)}{\pi} \right) \right] \times \left[1 + 1.5 \left(\frac{1}{2} + \frac{\arctan(t-60)}{\pi} \right) \right].
+\end{equation}
+In this case, $\Rw(t)$ remains around 2 and slowly decreases to $0.4 = 2 \times (1-0.8)$ around day 30, and slowly increases back up to $1 = 0.4 \times (1 + 1.5)$ around day 60.
 Finally, we assume $\Rv(t) = \rho \Rw(t)$.
@@
 
+\pagebreak
+
+\section*{Supplementary Figure}
+
+\setcounter{figure}{0}    
+\renewcommand\thefigure{S\arabic{figure}}
+
+\begin{figure}[!pht]
+\begin{center}
+\includegraphics[width=0.9\textwidth]{Rtbias.pdf}
+\caption{
+\textbf{Estimates of relative strength over time under different scenarios assuming step changes in reproduction number.}
+See Figure 3 of the original text for figure caption.
+}
+\end{center}
+\label{fig:Rtbias}
+\end{figure}
+
+\pagebreak
+
 \bibliography{newvariant.bib}

### 2021-05-01 Rtbias_smooth.R

@@ -40,13 +40,8 @@ for (i in 1:length(svec)) {
   tmparg <- c(theta=theta,
     tmax=150, slist[[i]],
     Rfun=function(t) {
-      if (t < 30) {
-        return(2)
-      } else if (t < 60) {
-        return(0.4)
-      } else {
-        return(1)
-      }
+      2 * (1 - 0.8 * (1/2 + 1/pi*atan((t-30)))) *
+        (1 + 1.5 * (1/2 + 1/pi*atan((t-60))))
     })
   
   rr <- do.call(renewal_det, tmparg) %>%
@@ -58,7 +53,7 @@ for (i in 1:length(svec)) {
     geom_line(aes(tvec, Rtest1, color="Wild type", lty="Estimated"), lwd=2) +
     geom_line(aes(tvec, Rtest2, color="Variant", lty="Estimated"), lwd=2) +
     scale_x_continuous("Time (days)", expand=c(0, 0), limits=c(15, 70)) +
-    scale_y_log10("Reproduction number, $\\mathcal{R}(t)$", limits=c(0.39, 8),
+    scale_y_log10("Reproduction number, $\\mathcal{R}(t)$", limits=c(0.4, 8),
                   breaks=c(0.25, 0.5, 1, 2, 4, 8)) +
     scale_color_manual(values=c("red", "black")) +
     scale_linetype_manual(values=c(2, 1), guide=FALSE) +
@@ -81,7 +76,7 @@ for (i in 1:length(svec)) {
       geom_line(aes(tvec, Rtest1, lty="Estimated"), col="black", lwd=2) +
       geom_line(aes(tvec, Rtest2, col="Estimated", lty="Estimated"), lwd=2) +
       scale_x_continuous("Time (days)", expand=c(0, 0), limits=c(15, 70)) +
-      scale_y_log10("Reproduction number, $\\mathcal{R}(t)$", limits=c(0.39, 8),
+      scale_y_log10("Reproduction number, $\\mathcal{R}(t)$", limits=c(0.4, 8),
                     breaks=c(0.25, 0.5, 1, 2, 4, 8)) +
       scale_color_manual("a", values=c("red", "red", "black")) +
       scale_linetype_manual("a", values=c(2, 1, 1)) +
@@ -148,7 +143,7 @@ for (i in 1:length(svec)) {
   if (i != 1) {
     g3 <- g3 +
[…]

### 2021-05-03 newvariant.tex
almost done

@@
 
-Inferring the relative strength (i.e., the ratio of reproduction numbers, $\Rv/\Rw$) and relative speed (i.e., the difference between growth rate, $\rv-\rw$) of new SARS-CoV-2 variants compared to their wild types is critical to predicting and controlling the course of the current pandemic.
+Inferring the relative strength (i.e., the ratio of reproduction numbers, $\Rv/\Rw$) and relative speed (i.e., the difference between growth rates, $\rv-\rw$) of new SARS-CoV-2 variants compared to their wild types is critical to predicting and controlling the course of the current pandemic.
 Multiple studies have estimated the relative strength of new variants from the observed relative speed, but they typically neglect the possibility that the new variants have different generation intervals (i.e., time between infection and transmission), which determines the relationship between relative strength and speed.
-Notably, the increasingly predominant B.1.1.7 variant may have a longer infectious period (and therefore, generation interval) than prior dominant lineages.
+Notably, the increasingly predominant B.1.1.7 variant may have a longer infectious period (and therefore, a longer generation interval) than prior dominant lineages.
 Here, we explore how differences in generation intervals between a new variant and the wild type affect the relationship between relative strength and speed.
-We use simulations to show how neglecting these differences can lead to biases in estimates of relative strength  in practice and illustrate how such biases can be assessed.
-Finally, we discuss implications for control: if new variants have longer generation intervals then speed-like interventions (e.g., contact tracing) become more effective, whereas strength-like interventions become less effective (e.g., social distancing).
+We use simulations to show how neglecting these differences can lead to biases in estimates of relative strength in practice and to illustrate how such biases can be assessed.
+Finally, we discuss implications for control: if new variants have longer generation intervals then speed-like interventions such as contact tracing become more effective, whereas strength-like interventions such as social distancing become less effective.
 
@@
 Estimating variant epidemic strength and speed remains one of the key questions in controlling the spread and understanding the threat of SARS-CoV-2 variants of concern (VoCs) \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
-Epidemic ``strength'' is measured by the reproduction number $\RR$ -- the average number of new infections caused by a ``typical'' infection. 
+Epidemic ``strength'' is measured by the reproduction number $\RR$ -- the average number of new infections caused by a typical infection. 
 Disease can spread in a population if $\RR>1$ \citep{diekmann1990definition}.
-The strength also provides information about the final size of an epidemic in a homogeneous population \citep{anderson1991infectious}.
-Epidemic speed is characterized by the growth rate $r$, which describes how fast a disease spreads at the population-level.
+The epidemic strength also determines the final size of an epidemic in a homogeneous population \citep{anderson1991infectious}.
+Epidemic ``speed'' is characterized by the growth rate $r$, which describes how fast a disease spreads at the population level.
 Like epidemic strength, epidemic speed also determines conditions for disease elimination: $r=0$ is a threshold equivalent to $\RR=1$.
-Strength and speed are linked by generation intervals---defined as time between infection and transmission---which characterizes the individual-level time scale of the epidemic \citep{roberts2007model,svensson2007note,wallinga2007generation}.
+Strength and speed are linked by generation intervals---defined as the time between infection and transmission---which characterizes the individual-level time scale of the epidemic \citep{roberts2007model,svensson2007note,wallinga2007generation}.
 
@@
 Many studies have tried to estimate the relative strength of the novel variant from the observed relative speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021}.
-Some studies have instead assumed a value of the relative strength of the novel variant and tried to predict its relative speed to determine when a new variant will become the dominant strain.
-\citep{davies2021estimated,di2021impact}.
-While both approaches are reasonable in principle, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the disease and its control \citep{doi:10.1098/rspb.2020.1556}:
+Some studies have instead assumed a value of the relative strength of the novel variant and tried to predict its relative speed to determine when a new variant will become the dominant strain \citep{davies2021estimated,di2021impact}.
+While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the disease and its control \citep{doi:10.1098/rspb.2020.1556}:
 estimation of speed from strength or strength from speed depends on assumptions about the generation-interval distribution \citep{roberts2007model,svensson2007note,wallinga2007generation}.
-For example, when we assume a fixed value of epidemic strength, longer generation intervals lead to a slower epidemic (lower $r$), making the epidemic seem easier to control.
-When we assume a fixed value of epidemic speed, longer generation intervals lead to stronger epidemic (higher $\RR$), making the epidemic seem more difficult to control.
+For example, when we assume a fixed value of epidemic strength, longer generation intervals lead to a slower epidemic (lower $r$), making the epidemic easier to control.
+When we assume a fixed value of epidemic speed, longer generation intervals lead to a stronger epidemic (higher $\RR$), making the epidemic more difficult to control.
 
 In practice, most studies have assumed that the previous dominant strain (wild type) and variants have identical generation intervals, but recent evidence suggests that the B.1.1.7 variant may have a longer duration of infection: 13.3 days (90\% CI: 10.1--16.5) for the new variant and 8.2 days (90\% CI: 6.5--9.7) for the wild type \citep{kissler2021densely}.
-Longer duration of infection implies that the mean generation interval of B.1.1.7 is likely to be longer than that of the wild type.
+Longer duration of infection suggests that the mean generation interval of B.1.1.7 is likely to be longer than that of the wild type.
 Other studies have considered the possibility that the faster growth rate of new variants may be driven, in part, by shorter generation intervals \citep{davies2021estimated,volz2021transmission}.
 In general, if the generation-interval distribution of a new variant is different from the generation-interval distribution of the wild type, estimates of variant strength may be biased.
-However, linking strength and speed is complicated given that generation intervals depend on many other biological and behavioral factors;
+However, linking strength and speed is complicated given that generation intervals depend on many factors including behavior:
 for example, self-isolation after symptom onset will lead to shorter generation intervals.
 
-Here, we explore how different assumptions affect estimates of the relative strength $\RR$ and speed $r$ for a new variant (e.g., B.1.1.7).
+Here, we explore how different assumptions affect estimates of the relative strength $\RR$ and speed $r$ for a new variant, such as B.1.1.7.
 We compare the relationship between relative strength (the ratio $\Rv/\Rw$) and relative speed (the difference $\rv-\rw$) under a wide range of assumptions about generation-interval distributions.
 We find that neglecting differences in the generation-interval distributions can lead to biased estimates.
-We also discuss how such biases might be assessed in practice and how information on differences in generation interval distributions might also influence priorities for controlling the spread of VoCs.
+We also discuss how such biases might be assessed in practice and how information on differences in generation interval distributions might influence priorities for controlling the spread of VoCs.
 
@@
 
-We use the renewal equation framework to characterize the spread of two pathogen strains---in this case, the wild type SARS-CoV-2 virus and a focal variant of concern, such as B.1.1.7.
-In particular, we focus on characterizing the incidence of \emph{infection} (i.e., the rate at which new infections are generated), which is directly related to $r$ and $\RR$.
+We use the renewal equation framework to characterize the spread of two pathogen strains---in this case, the wild type SARS-CoV-2 virus and a focal variant of concern.
+In particular, we focus on characterizing the incidence of infection (i.e., the rate at which new infections are generated), which is directly related to $r$ and $\RR$.
 In contrast, observed case reports are subject to reporting delays, which must be taken into account in order to correctly infer $r$ or $\RR$ in practice \citep{goldstein2009reconstructing,gostic2020practical}.
 
-Neglecting the (relatively slow) rate of new mutations, the current incidence of infection $i_x(t)$ caused by strain $x$---wild type (``wt'') and the variant (``var'')---can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary infections are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
+Neglecting the (relatively slow) rate of new mutations, the current incidence of infection $i_x(t)$ caused by each strain $x$---either the wild type (``wt'') or the variant (``var'')---can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary infections are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
 \begin{equation}
@@
 \end{equation}
-This framework provides a flexible way of modeling disease dynamics and generalizes compartmental models, such as the SEIR model \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, roberts2007model, champredon2018equivalence}.
+This framework provides a flexible way of modeling disease dynamics and generalizes compartmental model, such as the SEIR model \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, roberts2007model, champredon2018equivalence}.
 
 The integral of the kernel $\RR_x(t) = \int K_x(t, \tau) \dtau$ is referred to as the instantaneous reproduction number \citep{fraser2007estimating}.
-The instantaneous reproduction number is a particular kind of weighted average which describes the average infectiousness of previously infected individuals at time $t$---in particular, it is weighted by their relative infectiousness at time $t$, rather than by the actual number of infected individuals present.
+The instantaneous reproduction number is a particular kind of weighted average of infectiousness of previously infected individuals at time $t$---in particular, it is weighted by the total relative infectiousness at time $t$, rather than by the actual number of infected individuals present.
 The normalized kernel $g_x(t, \tau) = K_x(t, \tau)/\RR_x(t)$---which we refer to as the instantaneous generation-interval distribution---describes the relative contribution of previously infected individuals to current incidence $i_x(t)$ and provides information about the time scale of disease transmission.
@@
 \end{equation}
-We can approximate the $r$--$\RR$ relationship by assuming that the generation-interval distribution is Gamma-distributed, and summarizing it using the mean generation interval $\bar{G}_x$ and the squared coefficient of variation $\kappa_x$ \citep{park2019practical}:
+We can approximate this $r$--$\RR$ relationship by assuming that the generation-interval distribution is Gamma-distributed, and summarizing it using the mean generation interval $\bar{G}_x$ and the squared coefficient of variation $\kappa_x$ \citep{park2019practical}:
 \begin{equation}
@@
 Epidemic speed $r_x$ can often be estimated directly from incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}.
-However, estimating $r_x$ can be challenging when case counts are low (as when a new variant begins to spread) and is sensitive to temporal changes in testing patterns and intensity.
+However, estimating $r_x$ can be challenging when case counts are low (as when a new variant begins to spread) and is sensitive to temporal changes in the distribution of testing effort in the population and the overall testing intensity.
 Studies of new SARS-CoV-2 variants have mostly focused on characterizing changes in the \emph{proportion} of a new variant \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
-Focusing on proportions has the advantage of being less sensitive to changes in testing and to other transient effects that would affect variants and wild type viruses similarly.
+Focusing on proportions has the advantage, because changes in proportions are less sensitive to changes in testing and to other transient effects that would affect variants and wild type viruses similarly.
 When incidence is changing exponentially ($i_x(t) = i_x(t_0) \exp(r_x t)$), the proportion of the new variant $p(t)$ follows a logistic growth curve:
@@
 \textbf{Relative strength of the new variant assuming a fixed speed advantage $\delta$ under five epidemiological conditions.}
-The relative strength of the new variant $\rho$ conditional on the speed of the wild type $\rw$, the ratio between the mean generation interval of the new variant $\Gv$ and that of the wild type $\Gw$, and the squared coefficient of variation in generation intervals $\kappa$.
+The relative strength of the new variant $\rho$ conditional on the speed of the wild type $\rw$; the ratio between the mean generation interval of the new variant $\Gv$ and that of the wild type $\Gw$; and the squared coefficient of variation in generation intervals $\kappa$.
 The relative strength of the new variant $\rho$ is calculated using $\delta=0.1\pday$, $\Gw = 5\days$, and $\kappa = 1/5$.
-Assumed values of $\rw$ (and therefore $\rv$) are written on top right corners of each panel.
+Assumed values of $\rw$ (and therefore $\rv$) are shown in the top right corners of each panel.
 }
@@
 We evaluate the estimates of relative strength $\rho$ across a wide range of $\kappa$ from 0 (fixed-length generation intervals) to 1 (exponential distribution).
-To further explore how inference depends on underlying epidemiological conditions (i.e., whether the incidence of the variant and the wild type are growing or shrinking), we consider five epidemiological conditions: (1) $\rw < \rv < 0$, (2) $\rw < \rv = 0$, (3) $\rw < 0 < \rv$, (4) $0 = \rw < \rv$, and (5) $0 < \rw < \rv$.
+To further explore how inference depends on underlying epidemiological conditions (i.e., whether the incidences of the variant and the wild type are growing or shrinking), we consider five epidemiological conditions: (1) $\rw < \rv < 0$, (2) $\rw < \rv = 0$, (3) $\rw < 0 < \rv$, (4) $0 = \rw < \rv$, and (5) $0 < \rw < \rv$.
 
 \fref{relstrength} illustrates how the relative strength depends on the shape of the generation-interval distributions and on underlying epidemiological conditions.
-In general, longer mean generation intervals of the new variant translates to higher values of $\rho$ (and vice versa), except when $\rv \leq 0$ (recall, we always assume $\rw<\rv$).
+In general, longer mean generation intervals of the new variant translate to higher values of $\rho$ (and vice versa), except when $\rv \leq 0$ (recall, we always assume $\rw<\rv$).
 When $\rv = 0$, we always have $\Rv = 1$ and so the inferred value of $\rho$ is independent of the generation-interval distribution of the new variant.
@@
 For realistic values of $\kappa$ (excluding 0 and 1), we find that the inferred relative strength $\rho$ ranges between 1.1--2.3.
-Even if we restrict the difference in generation interval to $<20\%$ and only consider scenarios with $\rv>0$ (bottom panels of \fref{relstrength}), we find that $\rho$ ranges between 1.4--1.8.
+Even if we restrict the difference in generation interval to less than 20\% and only consider scenarios with $\rv>0$ (bottom panels of \fref{relstrength}), we find that $\rho$ ranges between 1.4--1.8.
 
@@
 We do not generally expect the relative speed $\delta$ to remain constant if other factors governing epidemic spread are changing.
-Instead, many biological mechanisms appear compatible to assuming a constant value of relative strength $\rho$ over changing conditions.
+Instead, many biological mechanisms appear compatible with assuming a constant value of relative strength $\rho$ over changing conditions.
 For example, if the proportion of the population susceptible declines, or the average contact rate changes, while other factors remain constant, the relative strength $\rho$ is expected to remain constant;
@@
 We thus investigate how $\delta$ is expected to change with \Rw if $\rho$ remains constant, and how this expectation changes with the ratio of the generation intervals. 
-Once again, we rely on the Gamma assumption to find the relative speed $\delta$ given the mean generation interval of the variant $\Gv$ and the wild type $\Gw$, :
+Once again, we rely on the Gamma assumption to find the relative speed $\delta$ given the mean generation interval of the variant $\Gv$ and the wild type $\Gw$,:
 \begin{equation}
@@
 \caption{
-\textbf{Relative speed of the new variant assuming a fixed strength advantage $\rho$ u under five epidemiological conditions.}
-The relative speed of the new variant $\delta$ given strength of the wild type $\Rw$, ratio between the mean generation interval of the new variant $\Gv$ and that of the wild type $\Gw$, and squared coefficient of variation in generation intervals $\kappa$.
+\textbf{Relative speed of the new variant assuming a fixed strength advantage $\rho$ under five epidemiological conditions.}
+The relative speed of the new variant $\delta$ conditional on the strength of the wild type $\Rw$; ratio between the mean generation interval of the new variant $\Gv$ and that of the wild type $\Gw$; and squared coefficient of variation in generation intervals $\kappa$.
 Relative speed of the new variant $\delta$ is calculated using $\rho=1.61$, $\Gw = 5\days$, and $\kappa = 1/5$.
@@
 Then, we compare the estimated ratio $\hat{\rho}(t) = \hat{\RR}_{\textrm{var}}(t)/\hat{\RR}_{\textrm{wt}}(t)$ with the true ratio $\rho = \Rv(t)/\Rw(t)$.
-In order to allow for changes in $\RR$ driven by the introduction and lifting of non-pharmaceutical interventions, we let $\Rw(t)$ to decrease from 2 to 0.4 around day 30 and increase back up to 1 around day 60 and assume $\Rv(t) = \rho \Rw(t)$.
+In order to allow for changes in $\RR$ driven by the introduction and lifting of non-pharmaceutical interventions, we let $\Rw(t)$ decrease from 2 to 0.4 around day 30 and increase back up to 1 around day 60 and assume $\Rv(t) = \rho \Rw(t)$.
 Although previous studies have modeled the impact of non-pharmaceutical interventions as a step function \citep{flaxman2020Rt}, we choose to use a smooth function to model $\Rw(t)$ (\fref{Rtbias}; see Methods) given the possibility that behavioral changes may affect transmission before and after interventions take place.
-In Supplementary Figure S1, we show that using a step function does not affect our conclusions.
+We reach similar conclusiosn if we use a step function instead (Supplementary Figure S1).
 
@@
 If $\rho$ is constant, and generation-interval distributions are correctly specified, we obtain a straight line with a slope of $\rho$ and intercept at zero (\fref{Rtbias}C).
-However, when the assumed mean generation interval is longer than the that of the variant, we over-estimate the slope (and vice versa, \fref{Rtbias}F,I).
+However, when the assumed mean generation interval is longer than the that of the variant, we over-estimate the slope (and conversely, \fref{Rtbias}F,I).
 Biases in slopes further translate to biases in intercepts; in theory, we would expect the regression line to go through the origin (because $\Rv = 0$ when $\Rw = 0$).
@@
 
-While relative speed $\delta$ and strength $\rho$ are useful for characterizing the spread of the variant in an epidemiological context with a previously dominant wild type, \emph{absolute} speed $\rv$ and strength $\Rv$ of the variant are the main variables that determine the spread and conditions for control of the variant over the long term.
+While relative speed $\delta$ and strength $\rho$ are useful for characterizing the spread of the variant in an epidemiological context with a previously dominant wild type, the \emph{absolute} speed $\rv$ and strength $\Rv$ of the variant are the main variables that determine the spread and conditions for control of the variant over the long term.
 In particular, at any given point in the epidemic, we can measure the speed of the variant $\rv$ (or infer $\rv$ from $\rw$ and $\delta$) and ask how much more intervention is required to control the spread of both strains (since $\Rv < 1$ implies $\Rw < 1$).
@@
 On the other hand, the same constant-speed intervention will prevent a larger proportion of transmission, leading to lower $\Ry{post}$ (\fref{strengthspeed}E), because it is more effective against late-stage transmission (\fref{strengthspeed}D).
-The constant-speed intervention also has a greater impact on reducing the mean generation interval (\fref{strengthspeed}D).
+The constant-speed intervention also reduces the mean generation interval by a larger factor (\fref{strengthspeed}D).
 
@@
 Likewise, the post-intervention speed of the variant under the constant-speed intervention is also independent of the mean generation interval;
-therefore, the constant-speed intervention will be always equally effective (on the speed scale) in preventing the spread.
+therefore, the constant-speed intervention will always be equally effective (on the speed scale) in preventing the spread.
 
@@
 We explored how the generation-interval distribution shapes the link between relative strength and speed of an invading disease variant.
-Longer generation intervals generally lead to higher relative strength for a given relative speed (and conversely, lower relative speed for a given relative strength); these can be reversed when incidence is decreasing.
+Longer generation intervals generally lead to higher relative strength for a given relative speed (and conversely, lower relative speed for a given relative strength); these relationships are reversed when incidence is decreasing.
 Neglecting potential differences in the mean generation intervals between the variant and the wild type can bias estimates of the relative strength from incidence data;
@@
 Finally, differences in generation intervals can also lead to different conclusions about the effectiveness of interventions:
-if the variant has longer (respectively, shorter) generation intervals than the wild type, speed-like (strength-like) interventions, will be more relatively more effective than naive estimates would suggest.
+if the variant has longer generation intervals than the wild type, speed-like interventions, will be relatively more effective than naive estimates would suggest (and conversely, strength-like intervention will be relatively more effective if the variant has shorter generation intervals).
 
@@
 However, our analysis also shows that uncertainty in generation-interval distributions must be taken into account to obtain accurate estimates of the relative strength of variants.
-If the variant has a longer infectious period \citep{kissler2021densely}, and therefore extended generation-interval distributions, current estimates may be underestimating the true relative strength.
+If the variant has a longer infectious period \citep{kissler2021densely}, and therefore extended generation-interval distributions, current estimates may underestimate the true relative strength.
 If new variants escape vaccinal or natural immunity to some extent \citep{shaman2020will}, their relative strength may increase further.
@@
 There is currently a (minor) discrepancy in the estimates of relative strength of new variants, particularly for B.1.1.7.
-Mathematical analyses have typically reported greater than a 1.4 fold increase in reproduction number for the B.1.1.7 variant ($\rho > 1.4$) whereas an independent analysis of secondary attack rates from contact tracing data suggests 1.25--1.4 fold increase \citep{ukinvest}.
-As shown in \cite{davies2021estimated} and \cite{volz2021transmission}, propagating uncertainty in generation interval estimates (and, in particular, assuming shorter generation intervals) can help explain these differences to some extent.
+Mathematical analyses have typically reported greater than a 1.4 fold increase in reproduction number for the B.1.1.7 variant ($\rho > 1.4$) whereas an independent analysis of secondary attack rates from contact tracing data suggests a 1.25--1.4 fold increase \citep{ukinvest}.
+As shown in \cite{davies2021estimated} and \cite{volz2021transmission}, propagating uncertainty in generation interval estimates (and, in particular, assuming shorter generation intervals) can partially explain these differences.
 
@@
 Although these estimates are more consistent with the attack rate analysis \citep{ukinvest},
-we do not claim that these estimates of $\rho$ are more accurate.
-This illustrative calculation highlights the importance of propagating uncertainty in generation-interval distributions in assessing the relative strength and speed of SARS-CoV-2 variants.
+we do not claim that our estimates of $\rho$ are more accurate.
+This calculation simply highlights the importance of propagating uncertainty in generation-interval distributions in assessing the relative strength and speed of SARS-CoV-2 variants.
 
@@
 when the assumed generation-interval distribution does not match the true generation-interval distribution of the variant, estimates of absolute strength $\Rv(t)$ will be biased, causing the estimated relative strength to change over time.
-In doing so, we assumed that the intervention will reduce transmission caused by the variant and the wild type by equal amounts, thereby preserving the relative strength over time;
-however, this is only true under strength-like interventions, which is insensitive to age of infection, but not under speed-like interventions. 
-As we demonstrated, the effectiveness of speed-like interventions depend on the shape of the generation-interval distribution;
-therefore, if the variant and the wild type have different generation intervals, speed-like interventions, such as contact tracing, can affect them differently.
+In our simulations, we assumed that the intervention would reduce transmission caused by the variant and the wild type by equal amounts, thereby preserving the relative strength over time;
+however, this assumption only holds under strength-like interventions, which are insensitive to age of infection, but not under speed-like interventions. 
+As we demonstrated, the effectiveness of speed-like interventions depends on the shape of the generation-interval distribution;
+therefore, if the variant and the wild type have different generation intervals, speed-like interventions such as contact tracing can affect them differently.
 
@@
 However, estimating $\RR(t)$ from real data, such as case incidence, is often associated with a variety of practical challenges, including delays between infection and reporting and changes in testing patterns \citep{gostic2020practical}.
-We have chosen to focus here on the underlying dynamical mechanisms that may impact inference.
+We have chosen to focus here on the underlying dynamical mechanisms that may affect inference.
 Future studies may consider the development of methods based on mechanism outlined here to assess potential biases in their estimates of relative strength.
@@
 Early in the spread of an emerging variant, it is likely more convenient to fix the relative speed, given that speed can be directly observed.
-As more information about the transmission and immunity profiles of the new variant becomes available, we advise fixing the relative strength and inferring the speed as this assumption better matches biological mechanisms for its higher strength.
+As more information about the transmission and immunity profiles of the new variant becomes available, we advise instead fixing the relative strength and inferring the speed, as this assumption better matches biological mechanisms for the variants' higher strength (e.g., higher rates of transmission and immune evasion).
 Researchers should be mindful about which quantity they hold constant and how their conclusions follow from their assumptions \citep{doi:10.1098/rspb.2020.1556}.
 
-This study has some practical implications for analyzing the epidemiological dynamics of new variants.
+This study has practical implications for analyzing the epidemiological dynamics of new variants.
 First, models that assume a constant relative speed, such as the standard logistic growth model, should be used with care---it is important to remember that the relative speed is expected to change with epidemiological conditions.
-In some cases, such as early in the spread of new variants, assuming a constant relative speed may not be a bad idea if epidemiological conditions do not change much.
+In some cases, such as early in the spread of new variants, assuming a constant relative speed may be justified if epidemiological conditions remain relatively constant. 
 Such simplifying assumption can also help obtain crude estimates of relative speed. 
-However, epidemiological conditions will change over time in response to spreading new variants, causing relative speed to change and invalidating this assumption.
-Second, the absolute strength and speed should not be neglected in favor of relative values, and both relative and absolute values should be discussed together.
+However, epidemiological conditions \textit{will} change over time in response to spreading new variants, causing relative speed to change and invalidating this assumption.
+Second, the absolute strength and speed should not be neglected in favor of relative values; relative and absolute values should be given roughly equal weight when presenting conclusions.
 While the relative strength and speed are useful for describing the spread of new variants, the absolute values determine their spread and control.
@@
 
-Even though SARS-CoV-2 has been spreading for more than a year, there is still considerable uncertainty in its generation intervals.
+Even though SARS-CoV-2 has been spreading for more than a year, there is still considerable uncertainty about its generation intervals.
 Observational studies typically focus on serial intervals (i.e., time between symptom onset of the infector and the infectee; \cite{svensson2007note}) because they are easier to measure \citep{griffin2020rapid}.
@@
 Furthermore, most estimates neglect dynamical biases in serial intervals caused by underlying epidemiological dynamics in their estimates of generation intervals.
-Serial intervals also do not account for asymptomatic transmission, adding further uncertainty to inferences of speed and strength \citep{park2020time}.
+Serial intervals also fail to account for asymptomatic transmission, adding further uncertainty to inferences of speed and strength \citep{park2020time}.
 
@@
 where $S$ represents the proportion of susceptible individuals.
-We discretize the model at the time scale of 0.025 days as described in \citep{park2021forward} and simulated the model with $\iv(0) = 0.001$ and $\iw(0) = 0.1$.
+We discretize the model at the time scale of 0.025 days as described in \cite{park2021forward} and simulated the model with $\iv(0) = 0.001$ and $\iw(0) = 0.1$.
 To allow for smooth changes in $\Rw(t)$ around day 30 and 60, we let:

### 2021-05-03 newvariant.tex
typo
minor edits and funding statement

@@
 estimation of speed from strength or strength from speed depends on assumptions about the generation-interval distribution \citep{roberts2007model,svensson2007note,wallinga2007generation}.
-For example, when we assume a fixed value of epidemic strength, longer generation intervals lead to a slower epidemic (lower $r$), making the epidemic easier to control.
-When we assume a fixed value of epidemic speed, longer generation intervals lead to a stronger epidemic (higher $\RR$), making the epidemic more difficult to control.
+For example, when epidemic strength is fixed, assuming longer generation intervals lead to a slower epidemic (lower $r$), making the epidemic look easier to control with speed-like interventions such as contact tracing.
+When epidemic speed is fixed, assuming longer generation intervals lead to a stronger epidemic (higher $\RR$), making the epidemic look harder to control with strength-like interventions such as social distancing.
 
@@
 Although previous studies have modeled the impact of non-pharmaceutical interventions as a step function \citep{flaxman2020Rt}, we choose to use a smooth function to model $\Rw(t)$ (\fref{Rtbias}; see Methods) given the possibility that behavioral changes may affect transmission before and after interventions take place.
-We reach similar conclusiosn if we use a step function instead (Supplementary Figure S1).
+We reach similar conclusions if we use a step function instead (Supplementary Figure S1).
 
@@
 
+\section*{Data availability}
+
+All data and code are stored in a publicly available GitHub repository (\url{https://github.com/parksw3/newvariant}).
+
+\section*{Competing interests}
+
+We declare no competing interests.
+
+\section*{Funding}
+
+BMB was supported by Natural Sciences and Engineering Research Council (NSERC). 
+% J.D. was supported by Canadian Institutes of Health Research (CIHR). 
+% J.S.W. was supported by Army Research Office (W911NF1910384).
+The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.
+
 \pagebreak

### 2021-05-03 relstrength.R

@@ -39,7 +39,7 @@ rhodata <- apply(pardata, 1, function(x) {
 }) %>%
   bind_rows %>%
   mutate(
-    kappa=factor(kappa, levels=c(0, 0.2, 0.5, 1)),
+    kappa=factor(kappa, levels=rev(c(0, 0.2, 0.5, 1))),
     rw=factor(rw, levels=c(rwvec[3:5], rwvec[1:2]),
               labels=c("$r_{\\textrm{wt}} < 0 < r_{\\textrm{var}}$",
                        "$0 = r_{\\textrm{wt}} < r_{\\textrm{var}}$",
@@ -94,8 +94,8 @@ g1 <- ggplot(rhodata) +
                 expand=c(0, 0.03)) +
   scale_y_log10("Relative strength, $\\rho$", breaks=c(1, 2, 4), limits=c(1, 4), expand=c(0, 0)) +
   facet_wrap(~rw, as.table=F) +
-  scale_color_colorblind("$\\kappa$") +
-  scale_linetype_discrete("$\\kappa$") +
+  scale_color_manual("$\\kappa$", values=rev(colorblind_pal()(4))) +
+  scale_linetype_manual("$\\kappa$", values=rev(c("solid", "22", "42", "44"))) +
   theme(
     panel.grid = element_blank(),
     panel.spacing = unit(0, "cm"),

### 2021-05-03 newvariant.tex
tweaks

@@
 Relative speed of the new variant $\delta$ is calculated using $\rho=1.61$, $\Gw = 5\days$, and $\kappa = 1/5$.
-Assumed values of $\Rw$ (and therefore $\Rv$) are written on top right corners of each panel.
+Assumed values of $\Rw$ (and therefore $\Rv$) are shown in the top right corners of each panel.
 }
@@
 A few studies have tried to estimate the generation-interval distribution from serial intervals, with means ranging between 3--6 days and squared coefficients of variation ranging between 0.1--1 \citep{ferretti2020quantifying,Ferretti2020timing,ganyani2020estimating,knight2020estimating};
-as we have shown, this uncertainty alone can generate large uncertainty in the inference of relative strength.
-Furthermore, most estimates neglect dynamical biases in serial intervals caused by underlying epidemiological dynamics in their estimates of generation intervals.
+as we have shown, this uncertainty in the generation-interval distribution alone can generate large uncertainty in the inference of relative strength.
+Furthermore, most estimates neglect dynamical biases in serial intervals caused by underlying epidemiological dynamics in their estimates of generation intervals \citep{park2021forward}.
 Serial intervals also fail to account for asymptomatic transmission, adding further uncertainty to inferences of speed and strength \citep{park2020time}.
@@
 
-The spread of new SARS-CoV-2 variants and the replacement of previously dominant lineages represent ongoing challenges for controlling the SARS-CoV-2 pandemic.  
+The spread of new SARS-CoV-2 variants and the replacement of previously dominant lineages represent ongoing challenges for controlling the SARS-CoV-2 pandemic \citep{abdool2021new,fontanet2021sars,walensky2021sars}.
 By explicitly considering epidemiological context and generation interval differences together, we have shown that improving estimates of the the relative duration of infectiousness at the individual scale may represent a pathway towards more effective interventions.
@@
 \end{center}
-\label{fig:Rtbias}
 \end{figure}

### 2021-05-03 control.R
really done

@@ -96,8 +96,10 @@ g1 <- ggplot(gendata_base) +
   geom_vline(xintercept = Gw, lwd=2, col=colorblind_pal()(3)[3], lty=3) +
   scale_x_continuous("Generation intervals (days)") +
   scale_y_continuous("Kernel density", limits=c(0, 0.4)) +
-  scale_color_manual("", values=(colorblind_pal()(3)[c(3, 1)])) +
-  scale_linetype_manual("", values=c(3, 1)) +
+  scale_color_manual("", values=(colorblind_pal()(3)[c(1, 3)]),
+                     breaks=c("Pre-intervention", "Post-intervention (constant-strength)")) +
+  scale_linetype_manual("", values=c(1, 3),
+                        breaks=c("Pre-intervention", "Post-intervention (constant-strength)")) +
   ggtitle("A. Equal generation intervals") +
   theme(
     panel.grid = element_blank(),
@@ -112,8 +114,10 @@ g2 <- ggplot(gendata_base) +
   geom_vline(xintercept = Gpost_base, col=colorblind_pal()(3)[2], lwd=2, lty=2) +
   scale_x_continuous("Generation intervals (days)") +
   scale_y_continuous("Kernel density", limits=c(0, 0.4)) +
-  scale_color_manual("", values=(colorblind_pal()(3)[c(2, 1)])) +
-  scale_linetype_manual("", values=c(2, 1)) +
+  scale_color_manual("", values=(colorblind_pal()(3)[c(1, 2)]),
+                     breaks=c("Pre-intervention", "Post-intervention (constant-speed)")) +
+  scale_linetype_manual("", values=c(1, 2),
+                        breaks=c("Pre-intervention", "Post-intervention (constant-speed)")) +
   ggtitle("B. Equal generation intervals") +
   theme(
     panel.grid = element_blank(),
@@ -155,8 +159,10 @@ g5 <- ggplot(intervention) +
                 breaks=c(2/3, 1, 3/2),
                 labels=c("2/3", 1, "3/2")) +
   scale_y_log10("Strength", breaks=c(0.5, 1, 2, 4), limits=c(0.5, 4)) +
-  scale_linetype_manual("", values=c(2, 3, 1)) +
-  scale_color_manual("", values=(colorblind_pal()(3)[c(2, 3, 1)])) +
+  scale_linetype_manual("", values=c(1, 2, 3),
[…]

### 2021-05-03 newvariant.tex
new line
added weitz funding
minor

@@
 }
+\newline
 \\
@@
 \textbf{9} School of Physics, Georgia Institute of Technology, Atlanta, GA, USA
-\\
-\textbf{10} Fogarty International Center, National Institutes of Health, Bethesda, MD, USA
 \\
@@
 
-BMB was supported by Natural Sciences and Engineering Research Council (NSERC). 
+B.M.B. was supported by Natural Sciences and Engineering Research Council of Canada (NSERC). 
+J.S.W. was supported by National Science Foundation (2032082).
 % J.D. was supported by Canadian Institutes of Health Research (CIHR).
-% J.S.W. was supported by Army Research Office (W911NF1910384).
 The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.

### 2021-05-03 newvariant.tex
funding done

@@
 
-B.M.B. was supported by Natural Sciences and Engineering Research Council of Canada (NSERC). 
-J.S.W. was supported by National Science Foundation (2032082).
-% J.D. was supported by Canadian Institutes of Health Research (CIHR).
+B.M.B. was supported by the Natural Sciences and Engineering Research Council of Canada 
+J.D. was supported by the Canadian Institutes of Health Research, 
+the The Natural Sciences and Engineering Research Council of Canada, 
+and the Michael G. DeGroote Institute for Infectious Disease Research.
+J.S.W. was supported by the National Science Foundation (2032082).
 The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.

### 2021-05-03 newvariant.tex
minor edits before the final submission

@@
 
-Estimating variant epidemic strength and speed remains one of the key questions in controlling the spread and understanding the threat of SARS-CoV-2 variants of concern (VoCs) \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
+Estimating variant epidemic strength and speed remains one of the key questions in controlling the spread and understanding the threat of SARS-CoV-2 variants of concern (VoCs) \citep{switzerland2021variant, davies2021estimated, di2021impact,graham2021changes, leung2021early, volz2021transmission,zhao2021}.
 Epidemic ``strength'' is measured by the reproduction number $\RR$ -- the average number of new infections caused by a typical infection. 
@@
 
-Analyses of new variants have typically characterized \emph{relative} strength (ratios between reproduction numbers $\Rv/\Rw$) and speed (differences between growth rates $\rv-\rw$) of the variants.
+Analyses of new variants have typically characterized \emph{relative} strength (i.e., the ratio of reproduction numbers $\Rv/\Rw$) and speed (i.e., the difference between growth rates $\rv-\rw$) of the variants.
 Many studies have tried to estimate the relative strength of the novel variant from the observed relative speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021}.
@@
 However, estimating $r_x$ can be challenging when case counts are low (as when a new variant begins to spread) and is sensitive to temporal changes in the distribution of testing effort in the population and the overall testing intensity.
-Studies of new SARS-CoV-2 variants have mostly focused on characterizing changes in the \emph{proportion} of a new variant \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
+Studies of new SARS-CoV-2 variants have mostly focused on characterizing changes in the \emph{proportion} of a new variant \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
 Focusing on proportions has the advantage, because changes in proportions are less sensitive to changes in testing and to other transient effects that would affect variants and wild type viruses similarly.
@@
 
-Unsurprisingly, we find that an increased speed of $\delta=0.1\pday$ for the variant is consistent with higher strength than the wild type across the range of epidemiological conditions considered \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
+Unsurprisingly, we find that an increased speed of $\delta=0.1\pday$ for the variant is consistent with higher strength than the wild type across the range of epidemiological conditions considered \citep{switzerland2021variant, davies2021estimated, di2021impact, ,graham2021changes, leung2021early, volz2021transmission,zhao2021}.
 However, the magnitude of relative strength $\rho$ is sensitive to assumptions about generation intervals.
@@
 \end{equation}
-As our baseline scenario we assume $\rho = 1.61$, which is the value we obtain for $\delta=0.1\pday$, $\Gw = \Gv = 5\,\textrm{days}$, and $\kappa = 1/5$.
+As our baseline scenario we assume $\rho = 1.61$, which is the value we obtain for $\delta=0.1\pday$, $\rw=0\pday$, $\Gw = \Gv = 5\,\textrm{days}$, and $\kappa = 1/5$.
 We evaluate $\delta$ across five different epidemiological conditions as before: (1) $\Rw < \Rv < 1$, (2) $\Rw < \Rv = 1$, (3) $\Rw < 1 < \Rv$, (4) $1 = \Rw < \Rv$, and (5) $1 < \Rw < \Rv$.
@@
 For example, even when there are no differences in the generation-interval distributions ($\Gv=\Gw$, in this case), the relative speed $\delta$ can range between 0.08--0.11 when $\kappa=0.2$ and 0.06--0.14 when $\kappa=0.5$.
-Therefore, characterizing the spread of variants assuming constant relative speed (e.g., by fitting a standard logistic growth curve) without considering how epidemiological conditions change over time should be avoided.
+Therefore, characterizing the spread of variants assuming constant relative speed (e.g., by fitting a standard logistic growth curve) without considering how epidemiological conditions change over time should be used with care.
 
@@
 
-As the new variants, in particular the B.1.1.7 variant, of SARS-CoV-2 are spreading and becoming dominant in many countries, it is clear that the variants are more transmissible (have higher strength) than the wild type \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
+As new variants, in particular the B.1.1.7 variant, of SARS-CoV-2 are spreading and becoming dominant in many countries, it is clear that the variants are more transmissible (have higher strength) than the wild type \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
 Our analysis supports estimates of a higher strength of the new variant across a wide range of assumptions about epidemiological conditions and generation-interval distributions.
 However, our analysis also shows that uncertainty in generation-interval distributions must be taken into account to obtain accurate estimates of the relative strength of variants.
-If the variant has a longer infectious period \citep{kissler2021densely}, and therefore extended generation-interval distributions, current estimates may underestimate the true relative strength.
-If new variants escape vaccinal or natural immunity to some extent \citep{shaman2020will}, their relative strength may increase further.
+If new variants has a longer infectious period \citep{kissler2021densely}, and therefore extended generation-interval distributions, current estimates may underestimate the true relative strength.
+If new variants escape vaccinal or natural immunity to some extent \citep{Kupferschmidt329}, their relative strength may increase further.
 
@@
 By explicitly considering epidemiological context and generation interval differences together, we have shown that improving estimates of the the relative duration of infectiousness at the individual scale may represent a pathway towards more effective interventions.
-Specifically, we have shown that speed-like interventions, such as contact tracing, will be relatively more effective if variants have longer generation intervals.
+Specifically, speed-like interventions, such as contact tracing, will be relatively more effective if variants have longer generation intervals.
 Most intervention strategies throughout the current pandemic have focused on strength-like interventions \citep{flaxman2020Rt}, such as lock-downs, partly because pre-symptomatic transmission of SARS-CoV-2 has limited the effectiveness of contact tracing efforts \citep{hellewell2020feasibility}.
-However, given the possibility that new variants can have different infection characteristics, future studies should consider whether their transmission dynamics also differ (e.g., the amount of pre-symptomatic transmission) and evaluate intervention strategies accordingly.
+However, given the possibility that new variants can have different infection characteristics \citep{kissler2021densely}, future studies should consider whether their transmission dynamics also differ (e.g., the amount of pre-symptomatic transmission) and evaluate intervention strategies accordingly.
 

### 2021-05-03 newvariant.tex
added a period
removed the

@@
 
-B.M.B. was supported by the Natural Sciences and Engineering Research Council of Canada 
+B.M.B. was supported by the Natural Sciences and Engineering Research Council of Canada. 
 J.D. was supported by the Canadian Institutes of Health Research, 
-the The Natural Sciences and Engineering Research Council of Canada, 
+the Natural Sciences and Engineering Research Council of Canada, 
 and the Michael G. DeGroote Institute for Infectious Disease Research.

### 2021-05-05 newvariant.tex
adding one more sentence and funding

@@
 Although these estimates are more consistent with the attack rate analysis \citep{ukinvest},
-we do not claim that our estimates of $\rho$ are more accurate.
-This calculation simply highlights the importance of propagating uncertainty in generation-interval distributions in assessing the relative strength and speed of SARS-CoV-2 variants.
+we do not claim that they are more accurate---
+in particular, contact tracing data can be biased towards particular types of contacts that are easily identified such as household contacts, which could also affect the estimate of $\rho$.
+Instead, this calculation simply highlights the importance of propagating uncertainty in generation-interval distributions in assessing the relative strength and speed of SARS-CoV-2 variants.
 
@@
 J.S.W. was supported by the National Science Foundation (2032082).
+S.F. was supported by the Wellcome Trust (210758/Z/18/Z).
 The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.

### 2021-05-07 relspeed_new.R
trying out new figures

@@ -1,0 +1,75 @@
+library(dplyr)
+library(ggplot2); theme_set(theme_bw(base_size=18, base_family = "Times"))
+library(ggthemes)
+library(egg)
+library(tikzDevice)
+
+rw <- 0
+delta <- 0.1
+kappa_base <- 1/5
+Gw <- 5
+
+rho <- ((1 + kappa_base * (rw+delta) * Gw)^(1/kappa_base))
+
+Rwvec <- exp(seq(log(1/rho^{3/2}), log(sqrt(rho)), length.out=31))
+kappavec <- c(0, 1)
+
+Gratiovec <- c(1/1.5, 1, 1.5)
+
+pardata <- expand.grid(kappavec, Rwvec, Gratiovec)
+
+deltadata <- apply(pardata, 1, function(x) {
+  kappa <- x[[1]]
+  Rw <- x[[2]]
+  Gratio <- x[[3]]
+  Gv <- Gratio * Gw
+  
+  if (kappa == 0) {
+    delta <- log(rho * Rw)/Gv - log(Rw)/Gw
+  } else {
+    delta <- ((rho * Rw)^kappa - 1)/(kappa*Gv) -  ((Rw)^kappa - 1)/(kappa*Gw)
+  }
+  
+  data.frame(
+    rho=rho,
+    Rw=Rw,
+    delta=delta,
+    kappa=kappa,
+    Gratio=Gratio
+  )
+}) %>%
+  bind_rows %>%
+  mutate(
+    kappa=factor(kappa, levels=c(kappavec),
+                 labels=paste0("$\\kappa=", kappavec, "$")),
+    Gratio2=factor(Gratio, levels=c(Gratiovec),
+                   labels=c("$\\bar{G}_{\\textrm{var}} = 2/3 \\bar{G}_{\\textrm{wt}}$",
+                            "$\\bar{G}_{\\textrm{var}} = \\bar{G}_{\\textrm{wt}}$",
+                            "$\\bar{G}_{\\textrm{var}} = 3/2 \\bar{G}_{\\textrm{wt}}$"))
+  )
+
+g1 <- ggplot(deltadata) +
+  geom_line(aes(Rw, delta, lty=Gratio2, col=Gratio2), lwd=2)  +
+  scale_x_log10("Wild type strength, $\\mathcal{R}_{\\textrm{wt}}$",
+                breaks=c(2/3, 1, 3/2),
+                labels=c("2/3", 1, "3/2"),
+                expand=c(0, 0.03)) +
[…]

### 2021-05-07 relstrength_new.R

@@ -1,0 +1,73 @@
+library(dplyr)
+library(ggplot2); theme_set(theme_bw(base_size=18, base_family = "Times"))
+library(ggthemes)
+library(egg)
+library(directlabels)
+library(tikzDevice)
+
+deltavec <- c(0.1)
+
+rwvec <- seq(-0.15, 0.05, length.out=31)
+kappavec <- c(0, 1)
+
+Gw <- 5
+
+Gratiovec <- c(1/1.5, 1, 1.5)
+
+pardata <- expand.grid(kappavec, rwvec, Gratiovec, deltavec)
+
+rhodata <- apply(pardata, 1, function(x) {
+  kappa <- x[[1]]
+  rw <- x[[2]]
+  Gratio <- x[[3]]
+  delta <- x[[4]]
+  
+  Gv <- Gratio * Gw
+  
+  if (kappa==0) {
+    rho <- exp(Gv * (rw + delta))/exp(Gw * rw)
+  } else {
+    rho <- ((1 + kappa * (rw + delta) * Gv)^(1/kappa))/
+      ((1 + kappa * rw * Gw)^(1/kappa))
+  }
+  
+  data.frame(
+    delta=delta,
+    rw=rw,
+    rho=rho,
+    Gratio=Gratio,
+    kappa=kappa
+  )
+}) %>%
+  bind_rows %>%
+  mutate(
+    kappa=factor(kappa, levels=c(kappavec),
+                 labels=paste0("$\\kappa=", kappavec, "$")),
+    Gratio2=factor(Gratio, levels=c(Gratiovec),
+                   labels=c("$\\bar{G}_{\\textrm{var}} = 2/3 \\bar{G}_{\\textrm{wt}}$",
+                            "$\\bar{G}_{\\textrm{var}} = \\bar{G}_{\\textrm{wt}}$",
+                            "$\\bar{G}_{\\textrm{var}} = 3/2 \\bar{G}_{\\textrm{wt}}$")),
+    delta2=factor(delta, levels=c(deltavec),
+                  labels=paste0("$\\delta=", deltavec, "/\\textrm{day}$"))
+  )
+
+g1 <- ggplot(rhodata) +
+  geom_line(aes(rw, rho, col=Gratio2, lty=Gratio2), lwd=2) +
+  facet_grid(~kappa) +
+  scale_x_continuous("Wild type speed, $r_{\\textrm{wt}}$ (1/day)", breaks=c(-0.1, 0)) +
[…]

### 2021-05-07 relspeed_new2.R
probably better

@@ -1,0 +1,129 @@
+library(dplyr)
+library(ggplot2); theme_set(theme_bw(base_size=18, base_family = "Times"))
+library(ggthemes)
+library(egg)
+library(tikzDevice)
+
+rw <- 0
+delta <- 0.1
+kappa_base <- 1/5
+Gw <- 5
+
+rho <- ((1 + kappa_base * (rw+delta) * Gw)^(1/kappa_base))
+
+Rwvec <- exp(seq(log(1/rho^{3/2}), log(sqrt(rho)), length.out=31))
+kappavec <- c(0, 0.2, 1)
+
+Gratiovec <- c(1/1.5, 1, 1.5)
+
+pardata <- expand.grid(kappavec, Rwvec, Gratiovec)
+
+deltadata <- apply(pardata, 1, function(x) {
+  kappa <- x[[1]]
+  Rw <- x[[2]]
+  Gratio <- x[[3]]
+  Gv <- Gratio * Gw
+  
+  if (kappa == 0) {
+    delta <- log(rho * Rw)/Gv - log(Rw)/Gw
+  } else {
+    delta <- ((rho * Rw)^kappa - 1)/(kappa*Gv) -  ((Rw)^kappa - 1)/(kappa*Gw)
+  }
+  
+  data.frame(
+    rho=rho,
+    Rw=Rw,
+    delta=delta,
+    kappa=kappa,
+    Gratio=Gratio
+  )
+}) %>%
+  bind_rows %>%
+  mutate(
+    kappa=factor(kappa, levels=c(kappavec),
+                 labels=paste0("$\\kappa=", kappavec, "$")),
+    Gratio2=factor(Gratio, levels=c(Gratiovec),
+                   labels=c("$\\bar{G}_{\\textrm{var}} = 2/3 \\bar{G}_{\\textrm{wt}}$",
+                            "$\\bar{G}_{\\textrm{var}} = \\bar{G}_{\\textrm{wt}}$",
+                            "$\\bar{G}_{\\textrm{var}} = 3/2 \\bar{G}_{\\textrm{wt}}$"))
+  )
+
+textdata1 <- data.frame(
+  x=sqrt(rho),
+  y=0.105,
+  Gratio2="$\\bar{G}_{\\textrm{var}} = \\bar{G}_{\\textrm{wt}}$",
+  kappa=c("$\\kappa=0$")
+)
+
+textdata2 <- data.frame(
+  x=sqrt(rho),
+  y=0.056,
[…]

### 2021-05-07 relstrength_new2.R

@@ -1,0 +1,128 @@
+library(dplyr)
+library(ggplot2); theme_set(theme_bw(base_size=18, base_family = "Times"))
+library(ggthemes)
+library(egg)
+library(directlabels)
+library(tikzDevice)
+
+deltavec <- c(0.1)
+
+rwvec <- seq(-0.15, 0.05, length.out=31)
+kappavec <- c(0, 0.2, 1)
+
+Gw <- 5
+
+Gratiovec <- c(1/1.5, 1, 1.5)
+
+pardata <- expand.grid(kappavec, rwvec, Gratiovec, deltavec)
+
+rhodata <- apply(pardata, 1, function(x) {
+  kappa <- x[[1]]
+  rw <- x[[2]]
+  Gratio <- x[[3]]
+  delta <- x[[4]]
+  
+  Gv <- Gratio * Gw
+  
+  if (kappa==0) {
+    rho <- exp(Gv * (rw + delta))/exp(Gw * rw)
+  } else {
+    rho <- ((1 + kappa * (rw + delta) * Gv)^(1/kappa))/
+      ((1 + kappa * rw * Gw)^(1/kappa))
+  }
+  
+  data.frame(
+    delta=delta,
+    rw=rw,
+    rho=rho,
+    Gratio=Gratio,
+    kappa=kappa
+  )
+}) %>%
+  bind_rows %>%
+  mutate(
+    kappa=factor(kappa, levels=c(kappavec),
+                 labels=paste0("$\\kappa=", kappavec, "$")),
+    Gratio2=factor(Gratio, levels=c(Gratiovec),
+                   labels=c("$\\bar{G}_{\\textrm{var}} = 2/3 \\bar{G}_{\\textrm{wt}}$",
+                            "$\\bar{G}_{\\textrm{var}} = \\bar{G}_{\\textrm{wt}}$",
+                            "$\\bar{G}_{\\textrm{var}} = 3/2 \\bar{G}_{\\textrm{wt}}$")),
+    delta2=factor(delta, levels=c(deltavec),
+                  labels=paste0("$\\delta=", deltavec, "/\\textrm{day}$"))
+  )
+
+textdata1 <- data.frame(
+  x=0.05,
+  y=1.71,
+  Gratio2="$\\bar{G}_{\\textrm{var}} = \\bar{G}_{\\textrm{wt}}$",
+  kappa=c("$\\kappa=0$")
+)
+
+textdata2 <- data.frame(
+  x=0.05,
+  y=2.5,
[…]

### 2021-05-14 newvariant.tex
cut down 600 words

@@
 \usepackage{natbib}
-\bibliographystyle{apalike}
+\setcitestyle{numbers} 
+\setcitestyle{square}
+\bibliographystyle{prsb}
 \date{\today}
@@
 \end{flushleft}
+
+%% 6635 words
+%% 6040 words
 
@@
 
-Unsurprisingly, we find that an increased speed of $\delta=0.1\pday$ for the variant is consistent with higher strength than the wild type across the range of epidemiological conditions considered \citep{switzerland2021variant, davies2021estimated, di2021impact, ,graham2021changes, leung2021early, volz2021transmission,zhao2021}.
+Unsurprisingly, we find that an increased speed of $\delta=0.1\pday$ for the variant is consistent with higher strength than the wild type across the range of epidemiological conditions considered \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
 However, the magnitude of relative strength $\rho$ is sensitive to assumptions about generation intervals.
@@
 Here, we investigate how misspecification of the generation-interval distribution of the variant affects our inference of relative strength from inference data under the assumption that the true generation-interval distribution of the wild type is known.
-We use a two-strain renewal equation that assumes perfect cross-immunity to simulate three different scenarios (see Methods):  
+We use a two-strain renewal equation that assumes perfect cross-immunity to simulate three different scenarios (see Supplementary Text):  
 (1) the wild type and the variant have the same generation-interval distributions (which match the known distribution, shown in \fref{Rtbias}A--C);
@@
 In order to allow for changes in $\RR$ driven by the introduction and lifting of non-pharmaceutical interventions, we let $\Rw(t)$ decrease from 2 to 0.4 around day 30 and increase back up to 1 around day 60 and assume $\Rv(t) = \rho \Rw(t)$.
-Although previous studies have modeled the impact of non-pharmaceutical interventions as a step function \citep{flaxman2020Rt}, we choose to use a smooth function to model $\Rw(t)$ (\fref{Rtbias}; see Methods) given the possibility that behavioral changes may affect transmission before and after interventions take place.
+Although previous studies have modeled the impact of non-pharmaceutical interventions as a step function \citep{flaxman2020Rt}, we choose to use a smooth function to model $\Rw(t)$ (\fref{Rtbias}; see Supplementary Text) given the possibility that behavioral changes may affect transmission before and after interventions take place.
 We reach similar conclusions if we use a step function instead (Supplementary Figure S1).
@@
 
-\section{Methods}
-
-\subsection{Two-strain renewal equation}
-
-We use a two-strain renewal equation to simulate the spread of the variant and the wild type.
-Ignoring birth and death, the incidence of infection caused by the variant $\iv$ and the wild type $\iw$ assuming perfect cross immunity is given by:
-\begin{align}
-\frac{dS}{dt} &= - \iv(t) - \iw(t)\\
-\iv(t) &= \Rv(t) \int_{0}^\infty \iv(t-\tau) g_{\mathrm{var}}(\tau) \dtau\\
-\iw(t) &= \Rw(t) \int_{0}^\infty \iw(t-\tau) g_{\mathrm{wt}}(\tau) \dtau
-\end{align}
-where $S$ represents the proportion of susceptible individuals.
-We discretize the model at the time scale of 0.025 days as described in \cite{park2021forward} and simulated the model with $\iv(0) = 0.001$ and $\iw(0) = 0.1$.
-To allow for smooth changes in $\Rw(t)$ around day 30 and 60, we let:
-\begin{equation}
-\Rw(t) = 2 \times \left[1 - 0.8 \left(\frac{1}{2} + \frac{\arctan(t-30)}{\pi} \right) \right] \times \left[1 + 1.5 \left(\frac{1}{2} + \frac{\arctan(t-60)}{\pi} \right) \right].
-\end{equation}
-In this case, $\Rw(t)$ remains around 2 and slowly decreases to $0.4 = 2 \times (1-0.8)$ around day 30, and slowly increases back up to $1 = 0.4 \times (1 + 1.5)$ around day 60.
-Finally, we assume $\Rv(t) = \rho \Rw(t)$.
-From simulated incidence curves between day 15 and 70, we estimate the instantaneous reproduction number using \eref{rt}.
-We ignore incidence before day 15 to remove any potential transient effects.
-
 \section*{Data availability}
@@
 
-\section*{Supplementary Figure}
-
-\setcounter{figure}{0}    
-\renewcommand\thefigure{S\arabic{figure}}
-
-\begin{figure}[!pht]
-\begin{center}
-\includegraphics[width=0.9\textwidth]{Rtbias.pdf}
-\caption{
-\textbf{Estimates of relative strength over time under different scenarios assuming step changes in reproduction number.}
-See Figure 3 of the original text for figure caption.
-}
-\end{center}
-\end{figure}
-
-\pagebreak
-
-\bibliography{newvariant.bib}
+\bibliography{newvariant_abbv.bib}
 

### 2021-05-14 supp.tex
cut down 600 words

--- 2021-05-03 newvariant.tex
@@
 	\textbf\newline{
-		Roles of generation-interval distributions in shaping relative epidemic strength, speed, and control of new SARS-CoV-2 variants
+		Roles of generation-interval distributions in shaping relative epidemic strength, speed, and control of new SARS-CoV-2 variants---Supplementary Materials
 	}
@@
 
-\section*{Abstract}
+%% 6635 words
 
-Inferring the relative strength (i.e., the ratio of reproduction numbers, $\Rv/\Rw$) and relative speed (i.e., the difference between growth rates, $\rv-\rw$) of new SARS-CoV-2 variants compared to their wild types is critical to predicting and controlling the course of the current pandemic.
-Multiple studies have estimated the relative strength of new variants from the observed relative speed, but they typically neglect the possibility that the new variants have different generation intervals (i.e., time between infection and transmission), which determines the relationship between relative strength and speed.
-Notably, the increasingly predominant B.1.1.7 variant may have a longer infectious period (and therefore, a longer generation interval) than prior dominant lineages.
-Here, we explore how differences in generation intervals between a new variant and the wild type affect the relationship between relative strength and speed.
-We use simulations to show how neglecting these differences can lead to biases in estimates of relative strength in practice and to illustrate how such biases can be assessed.
-Finally, we discuss implications for control: if new variants have longer generation intervals then speed-like interventions such as contact tracing become more effective, whereas strength-like interventions such as social distancing become less effective.
-
-\section{Introduction}
-
-Estimating variant epidemic strength and speed remains one of the key questions in controlling the spread and understanding the threat of SARS-CoV-2 variants of concern (VoCs) \citep{switzerland2021variant, davies2021estimated, di2021impact,graham2021changes, leung2021early, volz2021transmission,zhao2021}.
-Epidemic ``strength'' is measured by the reproduction number $\RR$ -- the average number of new infections caused by a typical infection. 
-Disease can spread in a population if $\RR>1$ \citep{diekmann1990definition}.
-The epidemic strength also determines the final size of an epidemic in a homogeneous population \citep{anderson1991infectious}.
-Epidemic ``speed'' is characterized by the growth rate $r$, which describes how fast a disease spreads at the population level.
-Like epidemic strength, epidemic speed also determines conditions for disease elimination: $r=0$ is a threshold equivalent to $\RR=1$.
-Strength and speed are linked by generation intervals---defined as the time between infection and transmission---which characterizes the individual-level time scale of the epidemic \citep{roberts2007model,svensson2007note,wallinga2007generation}.
-
-Analyses of new variants have typically characterized \emph{relative} strength (i.e., the ratio of reproduction numbers $\Rv/\Rw$) and speed (i.e., the difference between growth rates $\rv-\rw$) of the variants.
-Many studies have tried to estimate the relative strength of the novel variant from the observed relative speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021}.
-Some studies have instead assumed a value of the relative strength of the novel variant and tried to predict its relative speed to determine when a new variant will become the dominant strain \citep{davies2021estimated,di2021impact}.
-While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the disease and its control \citep{doi:10.1098/rspb.2020.1556}:
-estimation of speed from strength or strength from speed depends on assumptions about the generation-interval distribution \citep{roberts2007model,svensson2007note,wallinga2007generation}.
-For example, when epidemic strength is fixed, assuming longer generation intervals lead to a slower epidemic (lower $r$), making the epidemic look easier to control with speed-like interventions such as contact tracing.
-When epidemic speed is fixed, assuming longer generation intervals lead to a stronger epidemic (higher $\RR$), making the epidemic look harder to control with strength-like interventions such as social distancing.
-
-In practice, most studies have assumed that the previous dominant strain (wild type) and variants have identical generation intervals, but recent evidence suggests that the B.1.1.7 variant may have a longer duration of infection: 13.3 days (90\% CI: 10.1--16.5) for the new variant and 8.2 days (90\% CI: 6.5--9.7) for the wild type \citep{kissler2021densely}.
-Longer duration of infection suggests that the mean generation interval of B.1.1.7 is likely to be longer than that of the wild type.
-Other studies have considered the possibility that the faster growth rate of new variants may be driven, in part, by shorter generation intervals \citep{davies2021estimated,volz2021transmission}.
-In general, if the generation-interval distribution of a new variant is different from the generation-interval distribution of the wild type, estimates of variant strength may be biased.
-However, linking strength and speed is complicated given that generation intervals depend on many factors including behavior:
-for example, self-isolation after symptom onset will lead to shorter generation intervals.
-
-Here, we explore how different assumptions affect estimates of the relative strength $\RR$ and speed $r$ for a new variant, such as B.1.1.7.
-We compare the relationship between relative strength (the ratio $\Rv/\Rw$) and relative speed (the difference $\rv-\rw$) under a wide range of assumptions about generation-interval distributions.
-We find that neglecting differences in the generation-interval distributions can lead to biased estimates.
-We also discuss how such biases might be assessed in practice and how information on differences in generation interval distributions might influence priorities for controlling the spread of VoCs.
-
-\section{Renewal equation framework}
-
-We use the renewal equation framework to characterize the spread of two pathogen strains---in this case, the wild type SARS-CoV-2 virus and a focal variant of concern.
-In particular, we focus on characterizing the incidence of infection (i.e., the rate at which new infections are generated), which is directly related to $r$ and $\RR$.
-In contrast, observed case reports are subject to reporting delays, which must be taken into account in order to correctly infer $r$ or $\RR$ in practice \citep{goldstein2009reconstructing,gostic2020practical}.
-
-Neglecting the (relatively slow) rate of new mutations, the current incidence of infection $i_x(t)$ caused by each strain $x$---either the wild type (``wt'') or the variant (``var'')---can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary infections are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
-\begin{equation}
-i_x(t) = \int_0^\infty i_x(t-\tau) K_x(t, \tau) \dtau.
-\end{equation}
-This framework provides a flexible way of modeling disease dynamics and generalizes compartmental model, such as the SEIR model \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, roberts2007model, champredon2018equivalence}.
-
-The integral of the kernel $\RR_x(t) = \int K_x(t, \tau) \dtau$ is referred to as the instantaneous reproduction number \citep{fraser2007estimating}.
-The instantaneous reproduction number is a particular kind of weighted average of infectiousness of previously infected individuals at time $t$---in particular, it is weighted by the total relative infectiousness at time $t$, rather than by the actual number of infected individuals present.
-The normalized kernel $g_x(t, \tau) = K_x(t, \tau)/\RR_x(t)$---which we refer to as the instantaneous generation-interval distribution---describes the relative contribution of previously infected individuals to current incidence $i_x(t)$ and provides information about the time scale of disease transmission.
-Both the reproduction number and the generation-interval distribution depend on many factors, including intrinsic infectiousness of an infected individual, non-pharmaceutical interventions, awareness-driven behavior, and population-level susceptibility \citep{fraser2007estimating}.
-
-Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$.
-In this case, the incidence of each strain will exhibit exponential growth (or decay) at rate $r_x$, satisfying the Euler-Lotka equation \citep{wallinga2007generation}:
-\begin{equation}
-\frac{1}{\RR_x} = \int_0^\infty \exp(- r_x \tau) g_x(\tau) \dtau.
-\end{equation}
-We can approximate this $r$--$\RR$ relationship by assuming that the generation-interval distribution is Gamma-distributed, and summarizing it using the mean generation interval $\bar{G}_x$ and the squared coefficient of variation $\kappa_x$ \citep{park2019practical}:
-\begin{equation}
-\RR_x \approx (1 + \kappa_x r_x \bar{G}_x)^{1/\kappa_x}.
-\end{equation}
-The Gamma assumption includes as a special case models that assume exponentially distributed generation intervals (when $\kappa=1$), corresponding to the SIR model \citep{anderson1991infectious}; various Gamma assumptions are widely used in epidemic modeling, including for models of SARS-CoV-2 \citep{doi:10.1098/rsif.2020.0144}.
-We use this framework to investigate how inferences about strength and speed of the variant depend on our assumptions about the underlying generation-interval distributions.
-For simplicity we neglect differences in the squared coefficient of variation and assume $\kappa_{\mathrm{wt}} = \kappa_{\mathrm{var}} = \kappa$; instead, we focus on the effect of potential differences in the mean generation intervals.
-
-\section{Inferring relative strength from relative speed}
-
-Epidemic speed $r_x$ can often be estimated directly from incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}.
-However, estimating $r_x$ can be challenging when case counts are low (as when a new variant begins to spread) and is sensitive to temporal changes in the distribution of testing effort in the population and the overall testing intensity.
-Studies of new SARS-CoV-2 variants have mostly focused on characterizing changes in the \emph{proportion} of a new variant \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
-Focusing on proportions has the advantage, because changes in proportions are less sensitive to changes in testing and to other transient effects that would affect variants and wild type viruses similarly.
-When incidence is changing exponentially ($i_x(t) = i_x(t_0) \exp(r_x t)$), the proportion of the new variant $p(t)$ follows a logistic growth curve:
-\begin{align}
-p(t) &= \frac{\iv(t_0) \exp(\rv t)}{\iw(t_0) \exp(\rw t) + \iv(t_0) \exp(\rv t)},
-\\ &= \frac{1}{1 + \left(\iw(t_0)/\iv(t_0)\right) \exp(-\delta t)},
-\end{align}
-where the logistic growth rate $\delta = \rv - \rw$ corresponds to the relative speed of the epidemic.
-
-\begin{figure}[!t]
-\includegraphics[width=\textwidth]{relstrength.pdf}
-\caption{
-\textbf{Relative strength of the new variant assuming a fixed speed advantage $\delta$ under five epidemiological conditions.}
-The relative strength of the new variant $\rho$ conditional on the speed of the wild type $\rw$; the ratio between the mean generation interval of the new variant $\Gv$ and that of the wild type $\Gw$; and the squared coefficient of variation in generation intervals $\kappa$.
-The relative strength of the new variant $\rho$ is calculated using $\delta=0.1\pday$, $\Gw = 5\days$, and $\kappa = 1/5$.
-Assumed values of $\rw$ (and therefore $\rv$) are shown in the top right corners of each panel.
-}
-\label{fig:relstrength}
-\end{figure}
-
-We thus ask: what factors affect the relative strength $\rho = \Rv/\Rw$ of a new variant, conditional on an observed relative speed $\delta$?
-Inference of $\rho$ from $\delta$ will depend on assumptions about the generation-interval distributions of both strains.
-Given the mean generation interval of the variant $\Gv$ and the wild type $\Gw$, the relative strength $\rho = \Rv/\Rw$ under the Gamma assumption \citep{park2019practical} is given by:
-\begin{equation}
-\rho = \left(\frac{1 + \kappa (\rw + \delta) \Gv}{1 + \kappa \rw \Gw}\right)^{1/\kappa}.
-\end{equation}
-Therefore, the relative strength $\rho$ depends not only on the relative speed $\delta$ and the generation-interval distributions but also on how fast the wild type is spreading in the population (\rw)---
-some analyses have implicitly or explicitly neglected this factor by either assuming $\rw = 0$ \citep{switzerland2021variant} or $\kappa = 0$ \citep{davies2021estimated} (in the latter case $\hat{\rho} = \exp(\delta \bar{G})$).
-
-Based on the observed relative growth rate of B.1.1.7 in the UK, we start by  taking the relative speed of the variant to be $\delta = 0.1\pday$  \citep{davies2021estimated}; the mean generation interval of the wild type to be $\Gw = 5\days$ \citep{ferretti2020quantifying}; and the squared coefficient of variation of generation intervals to be $\kappa=0.2$ \citep{ferretti2020quantifying}.
-We consider the spread of B.1.1.7 as an example but our qualitative conclusions should hold for other VoCs.
-As noted above, we assume that the variant and the wild type have equal $\kappa$ throughout, and only consider differences in the mean.
-We evaluate the estimates of relative strength $\rho$ across a wide range of $\kappa$ from 0 (fixed-length generation intervals) to 1 (exponential distribution).
-To further explore how inference depends on underlying epidemiological conditions (i.e., whether the incidences of the variant and the wild type are growing or shrinking), we consider five epidemiological conditions: (1) $\rw < \rv < 0$, (2) $\rw < \rv = 0$, (3) $\rw < 0 < \rv$, (4) $0 = \rw < \rv$, and (5) $0 < \rw < \rv$.
-
-\fref{relstrength} illustrates how the relative strength depends on the shape of the generation-interval distributions and on underlying epidemiological conditions.
-In general, longer mean generation intervals of the new variant translate to higher values of $\rho$ (and vice versa), except when $\rv \leq 0$ (recall, we always assume $\rw<\rv$).
-When $\rv = 0$, we always have $\Rv = 1$ and so the inferred value of $\rho$ is independent of the generation-interval distribution of the new variant.
-When $\rv < 0$, we see that longer generation intervals decrease $\rho$ because longer generation intervals actually lead to slower decay (higher $r$).
-Assuming a narrower distribution (lower $\kappa$) has qualitatively similar effects as assuming longer generation intervals because both reduce the amount of early transmission; 
-therefore, narrower distributions lead to higher values of $\rho$ when $\rv > 0$ and lower values of $\rho$ when $\rv \leq 0$.
-When $\rw < 0 < \rv$, inference of $\rho$ is relatively insensitive to values of $\kappa$.
-
-Unsurprisingly, we find that an increased speed of $\delta=0.1\pday$ for the variant is consistent with higher strength than the wild type across the range of epidemiological conditions considered \citep{switzerland2021variant, davies2021estimated, di2021impact, ,graham2021changes, leung2021early, volz2021transmission,zhao2021}.
-However, the magnitude of relative strength $\rho$ is sensitive to assumptions about generation intervals.
-For realistic values of $\kappa$ (excluding 0 and 1), we find that the inferred relative strength $\rho$ ranges between 1.1--2.3.
-Even if we restrict the difference in generation interval to less than 20\% and only consider scenarios with $\rv>0$ (bottom panels of \fref{relstrength}), we find that $\rho$ ranges between 1.4--1.8.
-
-\section{Inferring relative speed from relative strength}
-
-We do not generally expect the relative speed $\delta$ to remain constant if other factors governing epidemic spread are changing.
-Instead, many biological mechanisms appear compatible with assuming a constant value of relative strength $\rho$ over changing conditions.
-For example, if the proportion of the population susceptible declines, or the average contact rate changes, while other factors remain constant, the relative strength $\rho$ is expected to remain constant;
-in general, this will imply a change in relative speed $\delta$.
-
-We thus investigate how $\delta$ is expected to change with \Rw if $\rho$ remains constant, and how this expectation changes with the ratio of the generation intervals. 
-Once again, we rely on the Gamma assumption to find the relative speed $\delta$ given the mean generation interval of the variant $\Gv$ and the wild type $\Gw$,:
-\begin{equation}
-\delta = \frac{(\rho \Rw)^{\kappa} - 1}{\kappa \Gv} - \frac{\Rw^{\kappa} - 1}{\kappa \Gw}.
-\end{equation}
-As our baseline scenario we assume $\rho = 1.61$, which is the value we obtain for $\delta=0.1\pday$, $\rw=0\pday$, $\Gw = \Gv = 5\,\textrm{days}$, and $\kappa = 1/5$.
-We evaluate $\delta$ across five different epidemiological conditions as before: (1) $\Rw < \Rv < 1$, (2) $\Rw < \Rv = 1$, (3) $\Rw < 1 < \Rv$, (4) $1 = \Rw < \Rv$, and (5) $1 < \Rw < \Rv$.
-
-\begin{figure}[!th]
-\includegraphics[width=\textwidth]{relspeed.pdf}
-\caption{
-\textbf{Relative speed of the new variant assuming a fixed strength advantage $\rho$ under five epidemiological conditions.}
-The relative speed of the new variant $\delta$ conditional on the strength of the wild type $\Rw$; ratio between the mean generation interval of the new variant $\Gv$ and that of the wild type $\Gw$; and squared coefficient of variation in generation intervals $\kappa$.
-Relative speed of the new variant $\delta$ is calculated using $\rho=1.61$, $\Gw = 5\days$, and $\kappa = 1/5$.
-Assumed values of $\Rw$ (and therefore $\Rv$) are shown in the top right corners of each panel.
-}
-\label{fig:relspeed}
-\end{figure}
-
-In general, longer generation intervals lead to slower relative speed of the variant when the incidence of both strains is increasing (\fref{relspeed}, bottom panels) because slower growth of the variant reduces the differences in absolute speed.
-When $\Rv=1$, the relative speed is insensitive to the generation-interval distribution of the variant because we always have $\rv=0$.
-When $\Rv<1$, longer generation intervals of the variant lead to slower decay ($\rv$ closer to 0), and therefore, greater relative speed.
-Once again, we see that assuming a narrower distribution (lower $\kappa$) has qualitatively similar effects as assuming a longer mean.
-
-\fref{relspeed} also shows that when $\rho$ is fixed relative speed depends on underlying epidemiological conditions. 
-For example, even when there are no differences in the generation-interval distributions ($\Gv=\Gw$, in this case), the relative speed $\delta$ can range between 0.08--0.11 when $\kappa=0.2$ and 0.06--0.14 when $\kappa=0.5$.
-Therefore, characterizing the spread of variants assuming constant relative speed (e.g., by fitting a standard logistic growth curve) without considering how epidemiological conditions change over time should be used with care.
-
-\section{Inferring relative strength from incidence data}
-
-Instead of estimating relative strength from speed, one can estimate time-varying or instantaneous reproduction numbers $\RR(t)$ of the variant and the wild type from incidence data \citep{fraser2007estimating}, and directly compare their ratios;
-such methods have been used in previous analyses of the B.1.1.7 strain by \cite{volz2021transmission}.
-Assuming that the generation-interval distribution remains constant, the instantaneous reproduction numbers of the new variant and of the wild type can be estimated from their corresponding incidence curves---an approach popularized by \cite{cori2013new}:
-\begin{equation}
-\RR_x(t) = \frac{i_x(t)}{\int_0^\infty i_x(t-\tau) g_x(\tau) \dtau}.
-\label{eq:rt}
-\end{equation}
-Under strength-like intervention measures that reduce transmission rates by a constant amount, we expect ratios between reproduction numbers to remain constant and correspond to the true relative strength: $\Rv(t)/\Rw(t) = \rho$.
-However, if the assumed generation-interval distribution $\hat{g}(\tau)$ is different from the true distribution, then the ratio between the estimated reproduction numbers $\hat{\rho}(t) = \hat{\RR}_{\textrm{var}}(t)/\hat{\RR}_{\textrm{wt}}(t)$ may change, even if the true ratio does not.
-
-Here, we investigate how misspecification of the generation-interval distribution of the variant affects our inference of relative strength from inference data under the assumption that the true generation-interval distribution of the wild type is known.
-We use a two-strain renewal equation that assumes perfect cross-immunity to simulate three different scenarios (see Methods):  
-(1) the wild type and the variant have the same generation-interval distributions (which match the known distribution, shown in \fref{Rtbias}A--C);
-(2) the variant has a shorter mean generation interval (\fref{Rtbias}D--F); and
-(3) the variant has a longer mean generation interval (\fref{Rtbias}G--I).
-Then, we compare the estimated ratio $\hat{\rho}(t) = \hat{\RR}_{\textrm{var}}(t)/\hat{\RR}_{\textrm{wt}}(t)$ with the true ratio $\rho = \Rv(t)/\Rw(t)$.
-In order to allow for changes in $\RR$ driven by the introduction and lifting of non-pharmaceutical interventions, we let $\Rw(t)$ decrease from 2 to 0.4 around day 30 and increase back up to 1 around day 60 and assume $\Rv(t) = \rho \Rw(t)$.
-Although previous studies have modeled the impact of non-pharmaceutical interventions as a step function \citep{flaxman2020Rt}, we choose to use a smooth function to model $\Rw(t)$ (\fref{Rtbias}; see Methods) given the possibility that behavioral changes may affect transmission before and after interventions take place.
-We reach similar conclusions if we use a step function instead (Supplementary Figure S1).
-
-\begin{figure}[!pht]
-\begin{center}
-\includegraphics[width=0.9\textwidth]{Rtbias_smooth.pdf}
-\caption{
-\textbf{Estimates of relative strength over time under different scenarios.}
-(A, D, G) True (solid lines) and estimated (dashed lines) reproduction numbers of the new variant and the wild type over time.
-(B, E, H) True (purple, solid) and estimated (orange, dashed) ratios between reproduction numbers of the new variant and the wild type over time.
-(C, F, I) Phase planes (time is implicit) showing true (purple, solid) and estimated (orange, dashed) relationships between estimated reproduction numbers.
-Blue dotted lines represent the regression lines of the estimated variant reproduction numbers against the estimated wild type reproduction numbers.
-Gray lines represent the one-to-one line.
-Top row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type and the variant.
-Middle row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type but is longer than that of the variant (4 days).
-Bottom row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type but is longer than that of the variant (6 days).
-For all simulations, squared coefficient of variation in generation intervals is assumed to equal $\kappa = 1/5$.
-}
-\end{center}
-\label{fig:Rtbias}
-\end{figure}
-
-When the assumed distribution matches the true distribution, the estimated reproduction numbers match the true values (\fref{Rtbias}A); thus, their ratio remains constant and $\hat{\RR}_{\textrm{var}}(t)=\Rv(t)$ (\fref{Rtbias}B).
-However, if the generation-interval distribution of the variant differs from the assumed distribution, the ratio changes over time (\fref{Rtbias}E,H).
-In particular, if the true generation intervals of the variant have a shorter mean than the assumed distribution, we over-estimate $\Rv(t)$ during the growth phase and under-estimate during the decay phase (and conversely, \fref{Rtbias}D,G), which further translates to biases in the estimated relative strength (\fref{Rtbias}E,H).
-
-In practice, estimates of instantaneous reproduction numbers $\RR(t)$ (and therefore, their ratios) can be noisy due to limited data availability or model assumptions;
-instead, we might want to estimate a single value of relative strength $\rho$.
-For example, we can estimate $\rho$ by plotting the estimated strength of the variant $\hat{\RR}_{\textrm{var}}(t)$ against the estimated strength of the wild type $\hat{\RR}_{\textrm{wt}}(t)$---as presented in Figure 2 of \cite{volz2021transmission}---and performing a linear regression (\fref{Rtbias}C,F,I).
-If $\rho$ is constant, and generation-interval distributions are correctly specified, we obtain a straight line with a slope of $\rho$ and intercept at zero (\fref{Rtbias}C).
-However, when the assumed mean generation interval is longer than the that of the variant, we over-estimate the slope (and conversely, \fref{Rtbias}F,I).
-Biases in slopes further translate to biases in intercepts; in theory, we would expect the regression line to go through the origin (because $\Rv = 0$ when $\Rw = 0$).
-
-\section{Implications for intervention strategies}
-
-While relative speed $\delta$ and strength $\rho$ are useful for characterizing the spread of the variant in an epidemiological context with a previously dominant wild type, the \emph{absolute} speed $\rv$ and strength $\Rv$ of the variant are the main variables that determine the spread and conditions for control of the variant over the long term.
-In particular, at any given point in the epidemic, we can measure the speed of the variant $\rv$ (or infer $\rv$ from $\rw$ and $\delta$) and ask how much more intervention is required to control the spread of both strains (since $\Rv < 1$ implies $\Rw < 1$).
-As a baseline scenario, we assume $\rw=0$ and $\delta=0.1\pday$ (and therefore $\rv=0.1\pday$), in which case additional intervention is required to reduce $\rv$ below 0 (or, equivalently, $\Rv$ below 1).
-
-We consider two types of intervention:
-an intervention of constant strength, which reduces transmission by a constant factor $\theta$ regardless of age of infection ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau)/\theta$); and an intervention of constant speed, which reduces transmission after infection by a constant rate $\phi$ ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau) \exp(-\phi \tau)$).
-In this case, we can control the spread of the variant when $\theta > \Rv$ or $\phi > \rv$, respectively \citep{doi:10.1098/rspb.2020.1556}.
-We consider constant-strength and speed interventions that reduce $\Rv$ to 0.9 when both the variant and the wild type have equal mean generation intervals (\fref{strengthspeed}A--B).
-While both interventions are equally effective on the strength scale (that is, $\Ry{post}=\int  K_{\mathrm{post}}(\tau) \dtau = 0.9$), they have different dynamical implications.
-The constant-strength intervention affects transmission equally throughout the course of infection, whereas the constant-speed intervention has greater impact on transmission that occurs later in infection;
-as a result, the constant-speed intervention reduces the post-intervention mean generation interval (\fref{strengthspeed}B) and leads to (slightly) faster exponential decay (therefore, lower $\ry{post}$).
-
-\begin{figure}[!th]
-\includegraphics[width=\textwidth]{control.pdf}
-\caption{
-\textbf{Effects of constant-strength and constant-speed interventions on the spread of a new variant with known speed \rv.}
-(A--B) Pre-intervention (black) and post-intervention (colored) kernel of the new variant under constant-strength (A) and -speed (B) interventions assuming that the variant has equal generation intervals as the wild type.
-(C--D) Pre-intervention (black) and post-intervention kernel of the new variant assuming that the variant has longer generation intervals than the wild type.
-(E) Pre-intervention (black) and post-intervention strength conditional on the mean generation interval of the variant (colored).
-(F) Pre-intervention (black) and post-intervention speed conditional on the mean generation interval of the variant (colored).
-Strength and speed are calculated assuming $\rw=0\pday$, $\delta=0.1\pday$, $\rv=0.1\pday$, $\Gy{wt}=5\,\textrm{days}$, and $\kappa=1/5$ for pre-intervention conditions.
-Intervention strength and speed are chosen so that post intervention strength of the new variant is 0.9 when its mean generation interval is 5 days.
-}
-\label{fig:strengthspeed}
-\end{figure}
-
-However, if the variant has longer generation intervals than the wild type (\fref{strengthspeed}C--D),  then the strength of the variant will be higher conditional on the observed speed (\fref{strengthspeed}E).
-In this case, the same constant-strength intervention can fail to control the epidemic (i.e., $\Ry{post} > 1$; \fref{strengthspeed}E) because this intervention reduces the transmission by a constant amount regardless of age of infection (\fref{strengthspeed}C).
-On the other hand, the same constant-speed intervention will prevent a larger proportion of transmission, leading to lower $\Ry{post}$ (\fref{strengthspeed}E), because it is more effective against late-stage transmission (\fref{strengthspeed}D).
-The constant-speed intervention also reduces the mean generation interval by a larger factor (\fref{strengthspeed}D).
-
-The speed-based paradigm gives the same results regarding control but provides important insights (\fref{strengthspeed}F).
-The observed speed of the variant $\rv$ at a given moment is independent of our estimates of its mean generation interval.
-Likewise, the post-intervention speed of the variant under the constant-speed intervention is also independent of the mean generation interval;
-therefore, the constant-speed intervention will always be equally effective (on the speed scale) in preventing the spread.
-
-The constant post-intervention speed can be understood by considering a speed-based decomposition of the kernel: $K_{\mathrm{pre}}(\tau) = \exp(\rv) b(\tau)$, where $b(\tau)$ is the initial backward generation-interval distribution and therefore integrates to 1 \citep{doi:10.1098/rspb.2020.1556}.
-This decomposition allows us to re-write the post-intervention kernel (under the constant-speed intervention) as follows:
-\begin{equation}
-K_{\mathrm{post}}(\tau) = \exp(\rv-\phi) b(\tau).
-\end{equation}
-Since the post-intervention speed $\ry{post}$ is determined by the following implicit equation,
-\begin{equation}
-1 = \int_0^\infty \exp(-\ry{post} \tau) K_{\mathrm{post}}(\tau) \dtau = \int_0^\infty \exp(\rv-\phi-\ry{post}) b(\tau)  \dtau,
-\end{equation}
-we have $\ry{post} = \phi - \rv$. 
-In particular, if speed of intervention is faster than the observed speed of spread (i.e., if $\phi > \rv$), we can control the epidemic (i.e., $\ry{post} < 0$) regardless of the underlying generation-interval distribution.
-
-\section{Discussion}
-
-We explored how the generation-interval distribution shapes the link between relative strength and speed of an invading disease variant.
-Longer generation intervals generally lead to higher relative strength for a given relative speed (and conversely, lower relative speed for a given relative strength); these relationships are reversed when incidence is decreasing.
-Neglecting potential differences in the mean generation intervals between the variant and the wild type can bias estimates of the relative strength from incidence data;
-these biases may be assessed by considering whether relative strength appears to vary systematically with underlying epidemiological dynamics.
-Finally, differences in generation intervals can also lead to different conclusions about the effectiveness of interventions:
-if the variant has longer generation intervals than the wild type, speed-like interventions, will be relatively more effective than naive estimates would suggest (and conversely, strength-like intervention will be relatively more effective if the variant has shorter generation intervals).
-
-As new variants, in particular the B.1.1.7 variant, of SARS-CoV-2 are spreading and becoming dominant in many countries, it is clear that the variants are more transmissible (have higher strength) than the wild type \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
-Our analysis supports estimates of a higher strength of the new variant across a wide range of assumptions about epidemiological conditions and generation-interval distributions.
-However, our analysis also shows that uncertainty in generation-interval distributions must be taken into account to obtain accurate estimates of the relative strength of variants.
-If new variants has a longer infectious period \citep{kissler2021densely}, and therefore extended generation-interval distributions, current estimates may underestimate the true relative strength.
-If new variants escape vaccinal or natural immunity to some extent \citep{Kupferschmidt329}, their relative strength may increase further.
-
-There is currently a (minor) discrepancy in the estimates of relative strength of new variants, particularly for B.1.1.7.
-Mathematical analyses have typically reported greater than a 1.4 fold increase in reproduction number for the B.1.1.7 variant ($\rho > 1.4$) whereas an independent analysis of secondary attack rates from contact tracing data suggests a 1.25--1.4 fold increase \citep{ukinvest}.
-As shown in \cite{davies2021estimated} and \cite{volz2021transmission}, propagating uncertainty in generation interval estimates (and, in particular, assuming shorter generation intervals) can partially explain these differences.
-
-Using the Gamma-approximation framework \citep{doi:10.1098/rsif.2020.0144} can help us reconcile different estimates more clearly.
-For example, \cite{davies2021estimated} estimated a relative strength of $\rho=1.77$ assuming a fixed generation generation interval at 5.5 days from $\delta=0.104\pday$ in one of their regression analysis.
-Similarly, \cite{volz2021transmission} estimated $\rho=1.79$ by directly comparing the ratio between $\Rv$ and $\Rw$ assuming a mean generation interval of 6.5 days and a squared coefficient of variation of 0.38.
-In this case, the latter study's longer mean generation interval (which increases $\rho$) is counter-balanced by the effects of their wider distribution (which decreases $\rho$); calculations using our approximation framework produce qualitatively similar results ($\rho=1.81$ from $\Gw=\Gv=6.4\days$, $\kappa=0.38$, $\delta=0.104\pday$, and $\rw=0\pday$).
-If we consider short and wide generation interval estimates from Tianjin, China ($\Gw=\Gv=2.57\days$ and $\kappa=1$; \cite{ganyani2020estimating}), we obtain $\rho=1.27$ (from $\delta=0.104\pday$ and $\rw = 0\pday$).
-Although these estimates are more consistent with the attack rate analysis \citep{ukinvest},
-we do not claim that our estimates of $\rho$ are more accurate.
-This calculation simply highlights the importance of propagating uncertainty in generation-interval distributions in assessing the relative strength and speed of SARS-CoV-2 variants.
-
-We further used simulations to show how mis-specification of generation-interval distributions can bias the inference of relative strength from incidence data:
-when the assumed generation-interval distribution does not match the true generation-interval distribution of the variant, estimates of absolute strength $\Rv(t)$ will be biased, causing the estimated relative strength to change over time.
-In our simulations, we assumed that the intervention would reduce transmission caused by the variant and the wild type by equal amounts, thereby preserving the relative strength over time;
-however, this assumption only holds under strength-like interventions, which are insensitive to age of infection, but not under speed-like interventions. 
-As we demonstrated, the effectiveness of speed-like interventions depends on the shape of the generation-interval distribution;
-therefore, if the variant and the wild type have different generation intervals, speed-like interventions such as contact tracing can affect them differently.
-
-We also assumed that the incidence of infection caused by the variant and the wild type is known.
-However, estimating $\RR(t)$ from real data, such as case incidence, is often associated with a variety of practical challenges, including delays between infection and reporting and changes in testing patterns \citep{gostic2020practical}.
-We have chosen to focus here on the underlying dynamical mechanisms that may affect inference.
-Future studies may consider the development of methods based on mechanism outlined here to assess potential biases in their estimates of relative strength.
-
-In moving from mechanism to improved estimates, we argue that both perspectives (i.e., whether we hold relative strength or speed constant) are useful in understanding the dynamics of the new variant.
-Early in the spread of an emerging variant, it is likely more convenient to fix the relative speed, given that speed can be directly observed.
-As more information about the transmission and immunity profiles of the new variant becomes available, we advise instead fixing the relative strength and inferring the speed, as this assumption better matches biological mechanisms for the variants' higher strength (e.g., higher rates of transmission and immune evasion).
-Researchers should be mindful about which quantity they hold constant and how their conclusions follow from their assumptions \citep{doi:10.1098/rspb.2020.1556}.
-
-This study has practical implications for analyzing the epidemiological dynamics of new variants.
-First, models that assume a constant relative speed, such as the standard logistic growth model, should be used with care---it is important to remember that the relative speed is expected to change with epidemiological conditions.
-In some cases, such as early in the spread of new variants, assuming a constant relative speed may be justified if epidemiological conditions remain relatively constant. 
-Such simplifying assumption can also help obtain crude estimates of relative speed. 
-However, epidemiological conditions \textit{will} change over time in response to spreading new variants, causing relative speed to change and invalidating this assumption.
-Second, the absolute strength and speed should not be neglected in favor of relative values; relative and absolute values should be given roughly equal weight when presenting conclusions.
-While the relative strength and speed are useful for describing the spread of new variants, the absolute values determine their spread and control.
-Finally, uncertainty in generation intervals should be carefully considered.
-
-Even though SARS-CoV-2 has been spreading for more than a year, there is still considerable uncertainty about its generation intervals.
-Observational studies typically focus on serial intervals (i.e., time between symptom onset of the infector and the infectee; \cite{svensson2007note}) because they are easier to measure \citep{griffin2020rapid}.
-However, realized serial intervals are subject to dynamical biases that can be difficult to tease apart \citep{park2021forward}.
-A few studies have tried to estimate the generation-interval distribution from serial intervals, with means ranging between 3--6 days and squared coefficients of variation ranging between 0.1--1 \citep{ferretti2020quantifying,Ferretti2020timing,ganyani2020estimating,knight2020estimating};
-as we have shown, this uncertainty in the generation-interval distribution alone can generate large uncertainty in the inference of relative strength.
-Furthermore, most estimates neglect dynamical biases in serial intervals caused by underlying epidemiological dynamics in their estimates of generation intervals \citep{park2021forward}.
-Serial intervals also fail to account for asymptomatic transmission, adding further uncertainty to inferences of speed and strength \citep{park2020time}.
-
-Since the beginning of the SARS-CoV-2 pandemic, a few studies have emphasized the relevance of leveraging generation interval distributions to improve estimates of strength \citep{doi:10.1098/rsif.2020.0144,ali2020serial,gostic2020practical,park2021forward} and the relative importance of modes of transmission (e.g., asymptomatic vs.
-symptomatic, \cite{park2020time}).
-These same lessons should be considered when assessing the spread of variants in a partially susceptible population and improving efforts to control spread.
-Future studies should prioritize detailed assessment of the generation intervals of SARS-CoV-2 and widespread variants, as well as consider how uncertainty in generational intervals might bias conclusions.
-
-The spread of new SARS-CoV-2 variants and the replacement of previously dominant lineages represent ongoing challenges for controlling the SARS-CoV-2 pandemic \citep{abdool2021new,fontanet2021sars,walensky2021sars}.
-By explicitly considering epidemiological context and generation interval differences together, we have shown that improving estimates of the the relative duration of infectiousness at the individual scale may represent a pathway towards more effective interventions.
-Specifically, speed-like interventions, such as contact tracing, will be relatively more effective if variants have longer generation intervals.
-Most intervention strategies throughout the current pandemic have focused on strength-like interventions \citep{flaxman2020Rt}, such as lock-downs, partly because pre-symptomatic transmission of SARS-CoV-2 has limited the effectiveness of contact tracing efforts \citep{hellewell2020feasibility}.
-However, given the possibility that new variants can have different infection characteristics \citep{kissler2021densely}, future studies should consider whether their transmission dynamics also differ (e.g., the amount of pre-symptomatic transmission) and evaluate intervention strategies accordingly.
-
-\section{Methods}
+\section{Supplementary Text}
 
@@
 
-\section*{Data availability}
-
-All data and code are stored in a publicly available GitHub repository (\url{https://github.com/parksw3/newvariant}).
-
-\section*{Competing interests}
-
-We declare no competing interests.
-
-\section*{Funding}
-
-B.M.B. was supported by the Natural Sciences and Engineering Research Council of Canada. 
-J.D. was supported by the Canadian Institutes of Health Research, 
-the Natural Sciences and Engineering Research Council of Canada, 
-and the Michael G. DeGroote Institute for Infectious Disease Research.
-J.S.W. was supported by the National Science Foundation (2032082).
-The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.
-
 \pagebreak
@@
 
-\bibliography{newvariant.bib}
+\bibliography{newvariant_abbv.bib}
 

### 2021-05-14 newvariant.tex
trying to cut down words

@@
 
-Estimating variant epidemic strength and speed remains one of the key questions in controlling the spread and understanding the threat of SARS-CoV-2 variants of concern (VoCs) \citep{switzerland2021variant, davies2021estimated, di2021impact,graham2021changes, leung2021early, volz2021transmission,zhao2021}.
+Estimating variant epidemic strength and speed remains a key question in understanding the threat of SARS-CoV-2 variants of concern (VoCs) \citep{switzerland2021variant, davies2021estimated, di2021impact,graham2021changes, leung2021early, volz2021transmission,zhao2021}.
 Epidemic ``strength'' is measured by the reproduction number $\RR$ -- the average number of new infections caused by a typical infection. 
@@
 Like epidemic strength, epidemic speed also determines conditions for disease elimination: $r=0$ is a threshold equivalent to $\RR=1$.
-Strength and speed are linked by generation intervals---defined as the time between infection and transmission---which characterizes the individual-level time scale of the epidemic \citep{roberts2007model,svensson2007note,wallinga2007generation}.
+Strength and speed are linked by generation intervals---defined as the time between infection and transmission \citep{roberts2007model,svensson2007note,wallinga2007generation}.
 
 Analyses of new variants have typically characterized \emph{relative} strength (i.e., the ratio of reproduction numbers $\Rv/\Rw$) and speed (i.e., the difference between growth rates $\rv-\rw$) of the variants.
-Many studies have tried to estimate the relative strength of the novel variant from the observed relative speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021}.
-Some studies have instead assumed a value of the relative strength of the novel variant and tried to predict its relative speed to determine when a new variant will become the dominant strain \citep{davies2021estimated,di2021impact}.
+Many studies have tried to estimate the relative strength of variants from the observed relative speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021}.
+Some studies have instead assumed a value of the relative strength and tried to predict its relative speed \citep{davies2021estimated,di2021impact}.
 While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the disease and its control \citep{doi:10.1098/rspb.2020.1556}:
@@
 Other studies have considered the possibility that the faster growth rate of new variants may be driven, in part, by shorter generation intervals \citep{davies2021estimated,volz2021transmission}.
-In general, if the generation-interval distribution of a new variant is different from the generation-interval distribution of the wild type, estimates of variant strength may be biased.
+% In general, neglecting differences in generation-interval distributions can lead to biased estimates of variant strength.
 However, linking strength and speed is complicated given that generation intervals depend on many factors including behavior:
@@
 
-Here, we explore how different assumptions affect estimates of the relative strength $\RR$ and speed $r$ for a new variant, such as B.1.1.7.
-We compare the relationship between relative strength (the ratio $\Rv/\Rw$) and relative speed (the difference $\rv-\rw$) under a wide range of assumptions about generation-interval distributions.
-We find that neglecting differences in the generation-interval distributions can lead to biased estimates.
+Here, we explore how different assumptions affect estimates of the relative strength (the ratio $\rho=\Rv/\Rw$) and relative speed (the difference $\delta=\rv-\rw$) for a new variant, such as B.1.1.7.
+% We compare $\rho$--$\delta$ relationship between relative strength and relative speed under a wide range of assumptions about generation-interval distributions.
+We show that neglecting differences in the generation-interval distributions can lead to biased estimates.
 We also discuss how such biases might be assessed in practice and how information on differences in generation interval distributions might influence priorities for controlling the spread of VoCs.
@@
 We use the renewal equation framework to characterize the spread of two pathogen strains---in this case, the wild type SARS-CoV-2 virus and a focal variant of concern.
-In particular, we focus on characterizing the incidence of infection (i.e., the rate at which new infections are generated), which is directly related to $r$ and $\RR$.
-In contrast, observed case reports are subject to reporting delays, which must be taken into account in order to correctly infer $r$ or $\RR$ in practice \citep{goldstein2009reconstructing,gostic2020practical}.
+We focus on characterizing the incidence of infection, which is directly related to $r$ and $\RR$.
+In practice, observed case reports are subject to reporting delays, which must be taken into account in order to correctly infer $r$ or $\RR$ \citep{goldstein2009reconstructing,gostic2020practical}.
 
@@
 Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$.
-In this case, the incidence of each strain will exhibit exponential growth (or decay) at rate $r_x$, satisfying the Euler-Lotka equation \citep{wallinga2007generation}:
+In this case, the incidence of each strain will exponentially grow (or decay) at rate $r_x$, satisfying the Euler-Lotka equation \citep{wallinga2007generation}:
 \begin{equation}
@@
 \end{equation}
-The Gamma assumption includes as a special case models that assume exponentially distributed generation intervals (when $\kappa=1$), corresponding to the SIR model \citep{anderson1991infectious}; various Gamma assumptions are widely used in epidemic modeling, including for models of SARS-CoV-2 \citep{doi:10.1098/rsif.2020.0144}.
+% The Gamma assumption includes as a special case models that assume exponentially distributed generation intervals (when $\kappa=1$), corresponding to the SIR model \citep{anderson1991infectious}; 
+Various Gamma assumptions have been widely used in epidemic modeling, including for models of SARS-CoV-2 \citep{doi:10.1098/rsif.2020.0144}.
 We use this framework to investigate how inferences about strength and speed of the variant depend on our assumptions about the underlying generation-interval distributions.
-For simplicity we neglect differences in the squared coefficient of variation and assume $\kappa_{\mathrm{wt}} = \kappa_{\mathrm{var}} = \kappa$; instead, we focus on the effect of potential differences in the mean generation intervals.
+For simplicity we neglect differences in the squared coefficient of variation and assume $\kappa_{\mathrm{wt}} = \kappa_{\mathrm{var}} = \kappa$ and focus on differences in the mean generation intervals.
 
@@
 
-We thus ask: what factors affect the relative strength $\rho = \Rv/\Rw$ of a new variant, conditional on an observed relative speed $\delta$?
-Inference of $\rho$ from $\delta$ will depend on assumptions about the generation-interval distributions of both strains.
+% We thus ask: what factors affect the relative strength $\rho = \Rv/\Rw$ of a new variant, conditional on an observed relative speed $\delta$?
+Inference of the relative strength $\rho = \Rv/\Rw$ from the observed relative speed $\delta$ depends on assumptions about the generation-interval distributions of both strains.
 Given the mean generation interval of the variant $\Gv$ and the wild type $\Gw$, the relative strength $\rho = \Rv/\Rw$ under the Gamma assumption \citep{park2019practical} is given by:
@@
 Therefore, the relative strength $\rho$ depends not only on the relative speed $\delta$ and the generation-interval distributions but also on how fast the wild type is spreading in the population (\rw)---
-some analyses have implicitly or explicitly neglected this factor by either assuming $\rw = 0$ \citep{switzerland2021variant} or $\kappa = 0$ \citep{davies2021estimated} (in the latter case $\hat{\rho} = \exp(\delta \bar{G})$).
+some analyses have implicitly or explicitly neglected this factor by either assuming $\rw = 0$ \citep{switzerland2021variant} or $\kappa = 0$ \citep{davies2021estimated} (in the latter case, $\rho = \exp(\delta \Gw)$ when $\Gv=\Gw$).
 
-Based on the observed relative growth rate of B.1.1.7 in the UK, we start by  taking the relative speed of the variant to be $\delta = 0.1\pday$  \citep{davies2021estimated}; the mean generation interval of the wild type to be $\Gw = 5\days$ \citep{ferretti2020quantifying}; and the squared coefficient of variation of generation intervals to be $\kappa=0.2$ \citep{ferretti2020quantifying}.
+Based on the observed relative growth rate of B.1.1.7 in the UK, we start by  taking the relative speed of the variant to be $\delta = 0.1\pday$  \citep{davies2021estimated}; the mean generation interval of the wild type to be $\Gw = 5\days$ \citep{ferretti2020quantifying}; and the squared coefficient of variation of generation intervals to be $\kappa=0.2$ \citep{ferretti2020quantifying} for both the variant and the wild type.
+% As noted above, we assume that the variant and the wild type have equal $\kappa$ throughout, and only consider differences in the mean.
+We evaluate the estimates of relative strength $\rho$ across a wide range of $\kappa$ from 0 (fixed-length generation intervals) to 1 (exponential distribution).
+To further explore how inference depends on underlying epidemiological conditions, we consider five scenarios: (1) $\rw < \rv < 0$, (2) $\rw < \rv = 0$, (3) $\rw < 0 < \rv$, (4) $0 = \rw < \rv$, and (5) $0 < \rw < \rv$.
 We consider the spread of B.1.1.7 as an example but our qualitative conclusions should hold for other VoCs.
-As noted above, we assume that the variant and the wild type have equal $\kappa$ throughout, and only consider differences in the mean.
-We evaluate the estimates of relative strength $\rho$ across a wide range of $\kappa$ from 0 (fixed-length generation intervals) to 1 (exponential distribution).
-To further explore how inference depends on underlying epidemiological conditions (i.e., whether the incidences of the variant and the wild type are growing or shrinking), we consider five epidemiological conditions: (1) $\rw < \rv < 0$, (2) $\rw < \rv = 0$, (3) $\rw < 0 < \rv$, (4) $0 = \rw < \rv$, and (5) $0 < \rw < \rv$.
 
@@
 In general, longer mean generation intervals of the new variant translate to higher values of $\rho$ (and vice versa), except when $\rv \leq 0$ (recall, we always assume $\rw<\rv$).
-When $\rv = 0$, we always have $\Rv = 1$ and so the inferred value of $\rho$ is independent of the generation-interval distribution of the new variant.
+When $\rv = 0$, we always have $\Rv = 1$ and so $\rho$ is independent of the generation-interval distribution of the new variant.
 When $\rv < 0$, we see that longer generation intervals decrease $\rho$ because longer generation intervals actually lead to slower decay (higher $r$).
 Assuming a narrower distribution (lower $\kappa$) has qualitatively similar effects as assuming longer generation intervals because both reduce the amount of early transmission; 
-therefore, narrower distributions lead to higher values of $\rho$ when $\rv > 0$ and lower values of $\rho$ when $\rv \leq 0$.
+therefore, narrower distributions lead to higher values of $\rho$ when $\rv > 0$ (and vice versa).
 When $\rw < 0 < \rv$, inference of $\rho$ is relatively insensitive to values of $\kappa$.
@@
 However, the magnitude of relative strength $\rho$ is sensitive to assumptions about generation intervals.
-For realistic values of $\kappa$ (excluding 0 and 1), we find that the inferred relative strength $\rho$ ranges between 1.1--2.3.
-Even if we restrict the difference in generation interval to less than 20\% and only consider scenarios with $\rv>0$ (bottom panels of \fref{relstrength}), we find that $\rho$ ranges between 1.4--1.8.
+% For realistic values of $\kappa$ (excluding 0 and 1), we find that the inferred relative strength $\rho$ ranges between 1.1--2.3.
+Even when we restrict the difference in generation interval to be less than 20\% and only consider scenarios with $\rv>0$ (bottom panels of \fref{relstrength}), $\rho$ ranges between 1.4--1.8.
 
@@
 As our baseline scenario we assume $\rho = 1.61$, which is the value we obtain for $\delta=0.1\pday$, $\rw=0\pday$, $\Gw = \Gv = 5\,\textrm{days}$, and $\kappa = 1/5$.
-We evaluate $\delta$ across five different epidemiological conditions as before: (1) $\Rw < \Rv < 1$, (2) $\Rw < \Rv = 1$, (3) $\Rw < 1 < \Rv$, (4) $1 = \Rw < \Rv$, and (5) $1 < \Rw < \Rv$.
+We evaluate $\delta$ across five scenarios as before: (1) $\Rw < \Rv < 1$, (2) $\Rw < \Rv = 1$, (3) $\Rw < 1 < \Rv$, (4) $1 = \Rw < \Rv$, and (5) $1 < \Rw < \Rv$.
 
@@
 For example, even when there are no differences in the generation-interval distributions ($\Gv=\Gw$, in this case), the relative speed $\delta$ can range between 0.08--0.11 when $\kappa=0.2$ and 0.06--0.14 when $\kappa=0.5$.
-Therefore, characterizing the spread of variants assuming constant relative speed (e.g., by fitting a standard logistic growth curve) without considering how epidemiological conditions change over time should be used with care.
+% Therefore, characterizing the spread of variants assuming constant relative speed (e.g., by fitting a standard logistic growth curve) without considering how epidemiological conditions change over time should be used with care.
 

### 2021-05-14 newvariant.tex
already cut down 873 words

@@
 %% 6635 words
-%% 6040 words
+%% 5762 words
 
@@
 Then, we compare the estimated ratio $\hat{\rho}(t) = \hat{\RR}_{\textrm{var}}(t)/\hat{\RR}_{\textrm{wt}}(t)$ with the true ratio $\rho = \Rv(t)/\Rw(t)$.
-In order to allow for changes in $\RR$ driven by the introduction and lifting of non-pharmaceutical interventions, we let $\Rw(t)$ decrease from 2 to 0.4 around day 30 and increase back up to 1 around day 60 and assume $\Rv(t) = \rho \Rw(t)$.
-Although previous studies have modeled the impact of non-pharmaceutical interventions as a step function \citep{flaxman2020Rt}, we choose to use a smooth function to model $\Rw(t)$ (\fref{Rtbias}; see Supplementary Text) given the possibility that behavioral changes may affect transmission before and after interventions take place.
+In order to simulate introduction and lifting of non-pharmaceutical interventions, we let $\Rw(t)$ to decrease from 2 to 0.4 around day 30 and increase back up to 1 around day 60 and assume $\Rv(t) = \rho \Rw(t)$.
+Previous studies have modeled the impact of non-pharmaceutical interventions as a step function \citep{flaxman2020Rt}, but we use a smooth function to model $\Rw(t)$ (\fref{Rtbias}; see Supplementary Text) given the possibility that behavioral changes may affect transmission before and after interventions take place.
 We reach similar conclusions if we use a step function instead (Supplementary Figure S1).
@@
 When the assumed distribution matches the true distribution, the estimated reproduction numbers match the true values (\fref{Rtbias}A); thus, their ratio remains constant and $\hat{\RR}_{\textrm{var}}(t)=\Rv(t)$ (\fref{Rtbias}B).
-However, if the generation-interval distribution of the variant differs from the assumed distribution, the ratio changes over time (\fref{Rtbias}E,H).
-In particular, if the true generation intervals of the variant have a shorter mean than the assumed distribution, we over-estimate $\Rv(t)$ during the growth phase and under-estimate during the decay phase (and conversely, \fref{Rtbias}D,G), which further translates to biases in the estimated relative strength (\fref{Rtbias}E,H).
+However, when the generation-interval distribution of the variant differs from the assumed distribution, the ratio changes over time (\fref{Rtbias}E,H).
+If the true generation intervals of the variant have a shorter mean than the assumed distribution, we over-estimate $\Rv(t)$ during the growth phase and under-estimate $\Rv(t)$ during the decay phase (and conversely, \fref{Rtbias}D,G), which further translates to biases in the estimated relative strength (\fref{Rtbias}E,H).
 
@@
 
-While relative speed $\delta$ and strength $\rho$ are useful for characterizing the spread of the variant in an epidemiological context with a previously dominant wild type, the \emph{absolute} speed $\rv$ and strength $\Rv$ of the variant are the main variables that determine the spread and conditions for control of the variant over the long term.
+While relative speed $\delta$ and strength $\rho$ are useful for characterizing the spread of the variant in an epidemiological context with a previously dominant wild type, the \emph{absolute} speed $\rv$ and strength $\Rv$ of the variant determine the spread and conditions for control of the variant over the long term.
 In particular, at any given point in the epidemic, we can measure the speed of the variant $\rv$ (or infer $\rv$ from $\rw$ and $\delta$) and ask how much more intervention is required to control the spread of both strains (since $\Rv < 1$ implies $\Rw < 1$).
@@
 these biases may be assessed by considering whether relative strength appears to vary systematically with underlying epidemiological dynamics.
-Finally, differences in generation intervals can also lead to different conclusions about the effectiveness of interventions:
-if the variant has longer generation intervals than the wild type, speed-like interventions, will be relatively more effective than naive estimates would suggest (and conversely, strength-like intervention will be relatively more effective if the variant has shorter generation intervals).
+Finally, differences in generation intervals can also lead to different conclusions about the effectiveness of interventions.
+% if the variant has longer generation intervals than the wild type, speed-like interventions, will be relatively more effective than naive estimates would suggest (and conversely, strength-like intervention will be relatively more effective if the variant has shorter generation intervals).
 
 As new variants, in particular the B.1.1.7 variant, of SARS-CoV-2 are spreading and becoming dominant in many countries, it is clear that the variants are more transmissible (have higher strength) than the wild type \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
-Our analysis supports estimates of a higher strength of the new variant across a wide range of assumptions about epidemiological conditions and generation-interval distributions.
-However, our analysis also shows that uncertainty in generation-interval distributions must be taken into account to obtain accurate estimates of the relative strength of variants.
+While our analysis supports estimates of a higher strength of the new variant across a wide range of assumptions, it also shows that uncertainty in generation-interval distributions must be taken into account to obtain accurate estimates of the relative strength of variants.
 If new variants has a longer infectious period \citep{kissler2021densely}, and therefore extended generation-interval distributions, current estimates may underestimate the true relative strength.
@@
 In our simulations, we assumed that the intervention would reduce transmission caused by the variant and the wild type by equal amounts, thereby preserving the relative strength over time;
-however, this assumption only holds under strength-like interventions, which are insensitive to age of infection, but not under speed-like interventions. 
+however, this assumption only holds under strength-like interventions, which are insensitive to time since infection, but not under speed-like interventions. 
 As we demonstrated, the effectiveness of speed-like interventions depends on the shape of the generation-interval distribution;

### 2021-05-15 newvariant.tex
need to cut 200 more words

@@
 %% 6635 words
-%% 5762 words
+%% 5299 words
 
@@
 Estimating variant epidemic strength and speed remains a key question in understanding the threat of SARS-CoV-2 variants of concern (VoCs) \citep{switzerland2021variant, davies2021estimated, di2021impact,graham2021changes, leung2021early, volz2021transmission,zhao2021}.
-Epidemic ``strength'' is measured by the reproduction number $\RR$ -- the average number of new infections caused by a typical infection. 
+Epidemic ``strength'' is measured by the reproduction number $\RR$---the average number of new infections caused by a typical infection. 
 Disease can spread in a population if $\RR>1$ \citep{diekmann1990definition}.
@@
 Like epidemic strength, epidemic speed also determines conditions for disease elimination: $r=0$ is a threshold equivalent to $\RR=1$.
-Strength and speed are linked by generation intervals---defined as the time between infection and transmission \citep{roberts2007model,svensson2007note,wallinga2007generation}.
+Strength and speed are linked by generation intervals---defined as the time between infection and transmission \citep{svensson2007note,wallinga2007generation}.
 
@@
 Some studies have instead assumed a value of the relative strength and tried to predict its relative speed \citep{davies2021estimated,di2021impact}.
-While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the disease and its control \citep{doi:10.1098/rspb.2020.1556}:
-estimation of speed from strength or strength from speed depends on assumptions about the generation-interval distribution \citep{roberts2007model,svensson2007note,wallinga2007generation}.
+While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the disease and its control \citep{doi:10.1098/rspb.2020.1556}.
 For example, when epidemic strength is fixed, assuming longer generation intervals lead to a slower epidemic (lower $r$), making the epidemic look easier to control with speed-like interventions such as contact tracing.
@@
 \end{equation}
-This framework provides a flexible way of modeling disease dynamics and generalizes compartmental model, such as the SEIR model \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, roberts2007model, champredon2018equivalence}.
+This framework provides a flexible way of modeling disease dynamics and generalizes compartmental model, such as the SEIR model \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, champredon2018equivalence}.
 
@@
 Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$.
-In this case, the incidence of each strain will exponentially grow (or decay) at rate $r_x$, satisfying the Euler-Lotka equation \citep{wallinga2007generation}:
+Then, the incidence of each strain grows (or decays) exponentially at rate $r_x$, satisfying the Euler-Lotka equation \citep{wallinga2007generation}:
 \begin{equation}
@@
 
-Epidemic speed $r_x$ can often be estimated directly from incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}.
-However, estimating $r_x$ can be challenging when case counts are low (as when a new variant begins to spread) and is sensitive to temporal changes in the distribution of testing effort in the population and the overall testing intensity.
-Studies of new SARS-CoV-2 variants have mostly focused on characterizing changes in the \emph{proportion} of a new variant \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
+While epidemic speed $r_x$ can often be estimated directly from incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating},
+studies of new SARS-CoV-2 variants have mostly focused on characterizing changes in the \emph{proportion} of a new variant \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
 Focusing on proportions has the advantage, because changes in proportions are less sensitive to changes in testing and to other transient effects that would affect variants and wild type viruses similarly.
@@
 
-\fref{relstrength} illustrates how the relative strength depends on the shape of the generation-interval distributions and on underlying epidemiological conditions.
+Unsurprisingly, we find that an increased speed of $\delta=0.1\pday$ for the variant is consistent with higher strength than the wild type across the range of epidemiological conditions considered (\fref{relstrength}).
+However, the magnitude of relative strength $\rho$ is sensitive to assumptions about generation intervals:
+For realistic values of $\kappa$ (excluding 0 and 1), the inferred relative strength $\rho$ ranges between 1.1--2.3 when $\Gv$ is allowed to vary between $2\Gw/3$ and $3\Gw/2$.
+
 In general, longer mean generation intervals of the new variant translate to higher values of $\rho$ (and vice versa), except when $\rv \leq 0$ (recall, we always assume $\rw<\rv$).
@@
 When $\rv < 0$, we see that longer generation intervals decrease $\rho$ because longer generation intervals actually lead to slower decay (higher $r$).
-Assuming a narrower distribution (lower $\kappa$) has qualitatively similar effects as assuming longer generation intervals because both reduce the amount of early transmission; 
-therefore, narrower distributions lead to higher values of $\rho$ when $\rv > 0$ (and vice versa).
+Assuming a narrower distribution (lower $\kappa$) has qualitatively similar effects as assuming longer generation intervals (leading to higher values of $\rho$ when $\rv > 0$) because both reduce the amount of early transmission.
 When $\rw < 0 < \rv$, inference of $\rho$ is relatively insensitive to values of $\kappa$.
 
-Unsurprisingly, we find that an increased speed of $\delta=0.1\pday$ for the variant is consistent with higher strength than the wild type across the range of epidemiological conditions considered \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
-However, the magnitude of relative strength $\rho$ is sensitive to assumptions about generation intervals.
-% For realistic values of $\kappa$ (excluding 0 and 1), we find that the inferred relative strength $\rho$ ranges between 1.1--2.3.
-Even when we restrict the difference in generation interval to be less than 20\% and only consider scenarios with $\rv>0$ (bottom panels of \fref{relstrength}), $\rho$ ranges between 1.4--1.8.
+% 
+% Even when we restrict the difference in generation interval to be less than 20\% and only consider scenarios with $\rv>0$ (bottom panels of \fref{relstrength}), $\rho$ ranges between 1.4--1.8.
 
@@
 Gray lines represent the one-to-one line.
-Top row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type and the variant.
-Middle row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type but is longer than that of the variant (4 days).
-Bottom row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type but is longer than that of the variant (6 days).
-For all simulations, squared coefficient of variation in generation intervals is assumed to equal $\kappa = 1/5$.
+For all simulations, the assumed mean generation interval is equal to the mean generation interval of the wild type (5 days), and the squared coefficient of variation in generation intervals is equal to $\kappa = 1/5$.
+The mean generation interval of the variant is equal to 5 days (top row), 4 days (middle row), and 6 days (bottom row).
 }
@@
 The observed speed of the variant $\rv$ at a given moment is independent of our estimates of its mean generation interval.
-Likewise, the post-intervention speed of the variant under the constant-speed intervention is also independent of the mean generation interval;
-therefore, the constant-speed intervention will always be equally effective (on the speed scale) in preventing the spread.
-
-The constant post-intervention speed can be understood by considering a speed-based decomposition of the kernel: $K_{\mathrm{pre}}(\tau) = \exp(\rv) b(\tau)$, where $b(\tau)$ is the initial backward generation-interval distribution and therefore integrates to 1 \citep{doi:10.1098/rspb.2020.1556}.
-This decomposition allows us to re-write the post-intervention kernel (under the constant-speed intervention) as follows:
-\begin{equation}
-K_{\mathrm{post}}(\tau) = \exp(\rv-\phi) b(\tau).
-\end{equation}
-Since the post-intervention speed $\ry{post}$ is determined by the following implicit equation,
-\begin{equation}
-1 = \int_0^\infty \exp(-\ry{post} \tau) K_{\mathrm{post}}(\tau) \dtau = \int_0^\infty \exp(\rv-\phi-\ry{post}) b(\tau)  \dtau,
-\end{equation}
-we have $\ry{post} = \phi - \rv$. 
-In particular, if speed of intervention is faster than the observed speed of spread (i.e., if $\phi > \rv$), we can control the epidemic (i.e., $\ry{post} < 0$) regardless of the underlying generation-interval distribution.
+Likewise, the post-intervention speed of the variant under the constant-speed intervention is also independent of the mean generation interval.
+Therefore, if speed of intervention is faster than the observed speed of spread (i.e., if $\phi > \rv$), we can control the epidemic (i.e., $\ry{post} < 0$) regardless of the underlying generation-interval distribution (see \cite{doi:10.1098/rspb.2020.1556} for mathematical details).
 
@@
 
-As new variants, in particular the B.1.1.7 variant, of SARS-CoV-2 are spreading and becoming dominant in many countries, it is clear that the variants are more transmissible (have higher strength) than the wild type \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
+As new variants of SARS-CoV-2, such as B.1.1.7, are spreading and becoming dominant in many countries, it is clear that the variants are more transmissible (have higher strength) than the wild type \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
 While our analysis supports estimates of a higher strength of the new variant across a wide range of assumptions, it also shows that uncertainty in generation-interval distributions must be taken into account to obtain accurate estimates of the relative strength of variants.
 If new variants has a longer infectious period \citep{kissler2021densely}, and therefore extended generation-interval distributions, current estimates may underestimate the true relative strength.
-If new variants escape vaccinal or natural immunity to some extent \citep{Kupferschmidt329}, their relative strength may increase further.
+% If new variants escape vaccinal or natural immunity to some extent \citep{Kupferschmidt329}, their relative strength may increase further.
 
@@
 Mathematical analyses have typically reported greater than a 1.4 fold increase in reproduction number for the B.1.1.7 variant ($\rho > 1.4$) whereas an independent analysis of secondary attack rates from contact tracing data suggests a 1.25--1.4 fold increase \citep{ukinvest}.
-As shown in \cite{davies2021estimated} and \cite{volz2021transmission}, propagating uncertainty in generation interval estimates (and, in particular, assuming shorter generation intervals) can partially explain these differences.
-
-Using the Gamma-approximation framework \citep{doi:10.1098/rsif.2020.0144} can help us reconcile different estimates more clearly.
-For example, \cite{davies2021estimated} estimated a relative strength of $\rho=1.77$ assuming a fixed generation generation interval at 5.5 days from $\delta=0.104\pday$ in one of their regression analysis.
-Similarly, \cite{volz2021transmission} estimated $\rho=1.79$ by directly comparing the ratio between $\Rv$ and $\Rw$ assuming a mean generation interval of 6.5 days and a squared coefficient of variation of 0.38.
-In this case, the latter study's longer mean generation interval (which increases $\rho$) is counter-balanced by the effects of their wider distribution (which decreases $\rho$); calculations using our approximation framework produce qualitatively similar results ($\rho=1.81$ from $\Gw=\Gv=6.4\days$, $\kappa=0.38$, $\delta=0.104\pday$, and $\rw=0\pday$).
-If we consider short and wide generation interval estimates from Tianjin, China ($\Gw=\Gv=2.57\days$ and $\kappa=1$; \cite{ganyani2020estimating}), we obtain $\rho=1.27$ (from $\delta=0.104\pday$ and $\rw = 0\pday$).
+As shown in \cite{davies2021estimated} and \cite{volz2021transmission}, propagating uncertainty in generation interval estimates (and, in particular, assuming shorter generation intervals) can partially explain these differences:
+For example, if we consider short and wide generation interval estimates from Tianjin, China ($\Gw=\Gv=2.57\days$ and $\kappa=1$; \cite{ganyani2020estimating}), we obtain $\rho=1.26$ (from $\delta=0.1\pday$ and $\rw = 0\pday$).
 Although these estimates are more consistent with the attack rate analysis \citep{ukinvest},
@@
 
+%Using the Gamma-approximation framework \citep{doi:10.1098/rsif.2020.0144} can help us reconcile different estimates more clearly.
+%For example, \cite{davies2021estimated} estimated a relative strength of $\rho=1.77$ assuming a fixed generation generation interval at 5.5 days from $\delta=0.104\pday$ in one of their regression analysis.
+%Similarly, \cite{volz2021transmission} estimated $\rho=1.79$ by directly comparing the ratio between $\Rv$ and $\Rw$ assuming a mean generation interval of 6.5 days and a squared coefficient of variation of 0.38.
+%In this case, the latter study's longer mean generation interval (which increases $\rho$) is counter-balanced by the effects of their wider distribution (which decreases $\rho$); calculations using our approximation framework produce qualitatively similar results ($\rho=1.81$ from $\Gw=\Gv=6.4\days$, $\kappa=0.38$, $\delta=0.104\pday$, and $\rw=0\pday$).
+%If we consider short and wide generation interval estimates from Tianjin, China ($\Gw=\Gv=2.57\days$ and $\kappa=1$; \cite{ganyani2020estimating}), we obtain $\rho=1.27$ (from $\delta=0.104\pday$ and $\rw = 0\pday$).
+%Although these estimates are more consistent with the attack rate analysis \citep{ukinvest},
+%we do not claim that they are more accurate---
+%in particular, contact tracing data can be biased towards particular types of contacts that are easily identified such as household contacts, which could also affect the estimate of $\rho$.
+%Instead, this calculation simply highlights the importance of propagating uncertainty in generation-interval distributions in assessing the relative strength and speed of SARS-CoV-2 variants.
+
 We further used simulations to show how mis-specification of generation-interval distributions can bias the inference of relative strength from incidence data:
@@
 however, this assumption only holds under strength-like interventions, which are insensitive to time since infection, but not under speed-like interventions. 
-As we demonstrated, the effectiveness of speed-like interventions depends on the shape of the generation-interval distribution;
-therefore, if the variant and the wild type have different generation intervals, speed-like interventions such as contact tracing can affect them differently.
+As we demonstrated, the effectiveness of speed-like interventions depends on the shape of the generation-interval distribution.
+Therefore, if the variant and the wild type have different generation intervals, speed-like interventions such as contact tracing can affect them differently.
 
@@
 Even though SARS-CoV-2 has been spreading for more than a year, there is still considerable uncertainty about its generation intervals.
-Observational studies typically focus on serial intervals (i.e., time between symptom onset of the infector and the infectee; \cite{svensson2007note}) because they are easier to measure \citep{griffin2020rapid}.
-However, realized serial intervals are subject to dynamical biases that can be difficult to tease apart \citep{park2021forward}.
-A few studies have tried to estimate the generation-interval distribution from serial intervals, with means ranging between 3--6 days and squared coefficients of variation ranging between 0.1--1 \citep{ferretti2020quantifying,Ferretti2020timing,ganyani2020estimating,knight2020estimating};
-as we have shown, this uncertainty in the generation-interval distribution alone can generate large uncertainty in the inference of relative strength.
-Furthermore, most estimates neglect dynamical biases in serial intervals caused by underlying epidemiological dynamics in their estimates of generation intervals \citep{park2021forward}.
-Serial intervals also fail to account for asymptomatic transmission, adding further uncertainty to inferences of speed and strength \citep{park2020time}.
-
-Since the beginning of the SARS-CoV-2 pandemic, a few studies have emphasized the relevance of leveraging generation interval distributions to improve estimates of strength \citep{doi:10.1098/rsif.2020.0144,ali2020serial,gostic2020practical,park2021forward} and the relative importance of modes of transmission (e.g., asymptomatic vs.
-symptomatic, \cite{park2020time}).
-These same lessons should be considered when assessing the spread of variants in a partially susceptible population and improving efforts to control spread.
-Future studies should prioritize detailed assessment of the generation intervals of SARS-CoV-2 and widespread variants, as well as consider how uncertainty in generational intervals might bias conclusions.
+A few studies have tried to estimate the generation-interval distribution, with means ranging between 3--6 days and squared coefficients of variation ranging between 0.1--1 \citep{ferretti2020quantifying,Ferretti2020timing,ganyani2020estimating,knight2020estimating}.
+However, these estimates are derived from serial intervals (i.e., time between symptom onset of the infector and the infectee; \cite{svensson2007note}), which are subject to dynamical biases \citep{park2021forward} and fail to account for asymptomatic transmission, adding further uncertainty to inferences of speed and strength \citep{park2020time}.
+Future studies should prioritize detailed assessment of the generation intervals of SARS-CoV-2 and widespread variants, as well as consider how uncertainty in generational intervals might bias conclusions \citep{doi:10.1098/rsif.2020.0144,ali2020serial,gostic2020practical}.
 

### 2021-05-15 newvariant.tex
ok i cut down 1477 words. going to submit.

@@
 
-%Using the Gamma-approximation framework \citep{doi:10.1098/rsif.2020.0144} can help us reconcile different estimates more clearly.
-%For example, \cite{davies2021estimated} estimated a relative strength of $\rho=1.77$ assuming a fixed generation generation interval at 5.5 days from $\delta=0.104\pday$ in one of their regression analysis.
-%Similarly, \cite{volz2021transmission} estimated $\rho=1.79$ by directly comparing the ratio between $\Rv$ and $\Rw$ assuming a mean generation interval of 6.5 days and a squared coefficient of variation of 0.38.
-%In this case, the latter study's longer mean generation interval (which increases $\rho$) is counter-balanced by the effects of their wider distribution (which decreases $\rho$); calculations using our approximation framework produce qualitatively similar results ($\rho=1.81$ from $\Gw=\Gv=6.4\days$, $\kappa=0.38$, $\delta=0.104\pday$, and $\rw=0\pday$).
-%If we consider short and wide generation interval estimates from Tianjin, China ($\Gw=\Gv=2.57\days$ and $\kappa=1$; \cite{ganyani2020estimating}), we obtain $\rho=1.27$ (from $\delta=0.104\pday$ and $\rw = 0\pday$).
-%Although these estimates are more consistent with the attack rate analysis \citep{ukinvest},
-%we do not claim that they are more accurate---
-%in particular, contact tracing data can be biased towards particular types of contacts that are easily identified such as household contacts, which could also affect the estimate of $\rho$.
-%Instead, this calculation simply highlights the importance of propagating uncertainty in generation-interval distributions in assessing the relative strength and speed of SARS-CoV-2 variants.
-
-We further used simulations to show how mis-specification of generation-interval distributions can bias the inference of relative strength from incidence data:
-when the assumed generation-interval distribution does not match the true generation-interval distribution of the variant, estimates of absolute strength $\Rv(t)$ will be biased, causing the estimated relative strength to change over time.
-In our simulations, we assumed that the intervention would reduce transmission caused by the variant and the wild type by equal amounts, thereby preserving the relative strength over time;
+We further used simulations to show how mis-specification of generation-interval distributions can bias the inference of relative strength from incidence data.
+In doing so, we assumed that the intervention would reduce transmission caused by the variant and the wild type by equal amounts, thereby preserving the relative strength over time;
 however, this assumption only holds under strength-like interventions, which are insensitive to time since infection, but not under speed-like interventions. 
-As we demonstrated, the effectiveness of speed-like interventions depends on the shape of the generation-interval distribution.
-Therefore, if the variant and the wild type have different generation intervals, speed-like interventions such as contact tracing can affect them differently.
+As we demonstrated, if the variant and the wild type have different generation intervals, speed-like interventions such as contact tracing can affect them differently, causing the relative strength to change over time.
 
 We also assumed that the incidence of infection caused by the variant and the wild type is known.
-However, estimating $\RR(t)$ from real data, such as case incidence, is often associated with a variety of practical challenges, including delays between infection and reporting and changes in testing patterns \citep{gostic2020practical}.
+However, estimating $\RR(t)$ from real data is often associated with a variety of practical challenges, including delays between infection and reporting and changes in testing patterns \citep{gostic2020practical}.
 We have chosen to focus here on the underlying dynamical mechanisms that may affect inference.
@@
 
-In moving from mechanism to improved estimates, we argue that both perspectives (i.e., whether we hold relative strength or speed constant) are useful in understanding the dynamics of the new variant.
-Early in the spread of an emerging variant, it is likely more convenient to fix the relative speed, given that speed can be directly observed.
-As more information about the transmission and immunity profiles of the new variant becomes available, we advise instead fixing the relative strength and inferring the speed, as this assumption better matches biological mechanisms for the variants' higher strength (e.g., higher rates of transmission and immune evasion).
-Researchers should be mindful about which quantity they hold constant and how their conclusions follow from their assumptions \citep{doi:10.1098/rspb.2020.1556}.
+% In moving from mechanism to improved estimates, we argue that both perspectives (i.e., whether we hold relative strength or speed constant) are useful in understanding the dynamics of the new variant.
+% Early in the spread of an emerging variant, it is likely more convenient to fix the relative speed, given that speed can be directly observed.
+% As more information about the transmission and immunity profiles of the new variant becomes available, we advise instead fixing the relative strength and inferring the speed, as this assumption better matches biological mechanisms for the variants' higher strength (e.g., higher rates of transmission and immune evasion).
+% Researchers should be mindful about which quantity they hold constant and how their conclusions follow from their assumptions \citep{doi:10.1098/rspb.2020.1556}.
 
@@
 First, models that assume a constant relative speed, such as the standard logistic growth model, should be used with care---it is important to remember that the relative speed is expected to change with epidemiological conditions.
-In some cases, such as early in the spread of new variants, assuming a constant relative speed may be justified if epidemiological conditions remain relatively constant. 
-Such simplifying assumption can also help obtain crude estimates of relative speed. 
+Early in the spread of an emerging variant, it is likely more convenient to fix the relative speed, given that speed can be directly observed.
 However, epidemiological conditions \textit{will} change over time in response to spreading new variants, causing relative speed to change and invalidating this assumption.
-Second, the absolute strength and speed should not be neglected in favor of relative values; relative and absolute values should be given roughly equal weight when presenting conclusions.
+As more information about the transmission and immunity profiles of the new variant becomes available, we advise instead fixing the relative strength and inferring the speed, as this assumption better matches biological mechanisms for the variants' higher strength (e.g., higher rates of transmission and immune evasion).
+Second, the absolute strength and speed should not be neglected in favor of relative values.
 While the relative strength and speed are useful for describing the spread of new variants, the absolute values determine their spread and control.

### 2021-05-15 newvariant.tex
oops

@@
 Joshua S.\ Weitz\textsuperscript{8,9}
-Bryan T.\ Grenfell\textsuperscript{1,7,10}
+Bryan T.\ Grenfell\textsuperscript{1,7}
 Jonathan Dushoff\textsuperscript{2,3,4}

### 2021-07-05 newvariant.tex
edits

@@
 
-Estimating variant epidemic strength and speed remains a key question in understanding the threat of SARS-CoV-2 variants of concern (VoCs) \citep{switzerland2021variant, davies2021estimated, di2021impact,graham2021changes, leung2021early, volz2021transmission,zhao2021}.
-Epidemic ``strength'' is measured by the reproduction number $\RR$---the average number of new infections caused by a typical infection. 
+Estimating variant epidemic strength and speed remains a key question in understanding the threat of SARS-CoV-2 variants of concern (VoCs) \citep{switzerland2021variant, davies2021estimated, di2021impact,graham2021changes, leung2021early, volz2021transmission,zhao2021,campbell2021increased}.
+Epidemic ``strength'' is measured by the reproduction number $\RR$---a unitless quantity representing the average number of new infections caused by a typical infection. 
 Disease can spread in a population if $\RR>1$ \citep{diekmann1990definition}.
 The epidemic strength also determines the final size of an epidemic in a homogeneous population \citep{anderson1991infectious}.
-Epidemic ``speed'' is characterized by the growth rate $r$, which describes how fast a disease spreads at the population level.
+Epidemic ``speed'' is characterized by the growth rate $r$, which has a unit of $1/\mathrm{time}$ and describes how fast a disease spreads at the population level.
 Like epidemic strength, epidemic speed also determines conditions for disease elimination: $r=0$ is a threshold equivalent to $\RR=1$.
-Strength and speed are linked by generation intervals---defined as the time between infection and transmission \citep{svensson2007note,wallinga2007generation}.
+
+Epidemic speed can be typically estimated from incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}, but epidemic strength is difficult to measure directly.
+Instead, epidemic strength can be inferred from the observed epidemic speed using the generation-interval distribution, an approach popularized by \citep{wallinga2007generation}.
+The generation interval is defined as the time between infection and transmission and provides information about the individual-level transmission \citep{svensson2007note}.
+The exact shape of the distribution depends on several factors---including the shape of latent and infectious periods \citep{lloyd2001realistic,wearing2005appropriate,roberts2007model} as well as more detailed life history of a disease \citep{huber2016quantitative}---and thus can be difficult to estimate.
+While it is possible to consider general forms of generation-interval distributions \citep{miller2010epidemics,svensson2015influence}, summarizing the distribution in terms of its mean and variability---for example, by using the Gamma distribution---can still provide a robust link between epidemic speed and strength for real diseases and yield important biological insights \citep{park2019practical}.
+In particular, several studies have noted, in many contexts, that mechanisms that increase the mean generation interval can increase epidemic strength given observed epidemic speed \citep{eaton2014proportion,powers2014impact,weitz2015modeling,park2020time}.
 
@@
 Some studies have instead assumed a value of the relative strength and tried to predict its relative speed \citep{davies2021estimated,di2021impact}.
-While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the disease and its control \citep{doi:10.1098/rspb.2020.1556}.
+While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the disease and its control because two quantities are linked by the generation-interval distribution \citep{doi:10.1098/rspb.2020.1556}.
 For example, when epidemic strength is fixed, assuming longer generation intervals lead to a slower epidemic (lower $r$), making the epidemic look easier to control with speed-like interventions such as contact tracing.
 When epidemic speed is fixed, assuming longer generation intervals lead to a stronger epidemic (higher $\RR$), making the epidemic look harder to control with strength-like interventions such as social distancing.
+Therefore, it is critical to evaluate which assumption is more appropriate for inferring transmission advantages of new variants.
 
-In practice, most studies have assumed that the previous dominant strain (wild type) and variants have identical generation intervals, but recent evidence suggests that the B.1.1.7 variant may have a longer duration of infection: 13.3 days (90\% CI: 10.1--16.5) for the new variant and 8.2 days (90\% CI: 6.5--9.7) for the wild type \citep{kissler2021densely}.
-Longer duration of infection suggests that the mean generation interval of B.1.1.7 is likely to be longer than that of the wild type.
+In addition, most studies have assumed that the previous dominant strain (wild type) and variants have identical generation-interval distributions, but recent evidence suggests that the B.1.1.7 variant may have a longer duration of infection: 13.3 days (90\% CI: 10.1--16.5) for the new variant and 8.2 days (90\% CI: 6.5--9.7) for the wild type \citep{kissler2021densely}.
+Longer duration of infection suggests that the mean generation interval of B.1.1.7 is likely to be longer than that of the wild type \citep{lloyd2001realistic,wearing2005appropriate,roberts2007model}.
 Other studies have considered the possibility that the faster growth rate of new variants may be driven, in part, by shorter generation intervals \citep{davies2021estimated,volz2021transmission}.
-% In general, neglecting differences in generation-interval distributions can lead to biased estimates of variant strength.
-However, linking strength and speed is complicated given that generation intervals depend on many factors including behavior:
+However, linking strength and speed is complicated given that the generation-interval distribution depend on many factors including behavior: 
 for example, self-isolation after symptom onset will lead to shorter generation intervals.
@@
 
-The integral of the kernel $\RR_x(t) = \int K_x(t, \tau) \dtau$ is referred to as the instantaneous reproduction number \citep{fraser2007estimating}.
+The integral of the kernel $\RR_x(t) = \int K_x(t, \tau) \dtau$ is referred to as the instantaneous reproduction number, which is defined as secondary infections that would be caused by an individual infected at time $t$ if conditions were to remain the same \citep{fraser2007estimating}.
 The instantaneous reproduction number is a particular kind of weighted average of infectiousness of previously infected individuals at time $t$---in particular, it is weighted by the total relative infectiousness at time $t$, rather than by the actual number of infected individuals present.
 The normalized kernel $g_x(t, \tau) = K_x(t, \tau)/\RR_x(t)$---which we refer to as the instantaneous generation-interval distribution---describes the relative contribution of previously infected individuals to current incidence $i_x(t)$ and provides information about the time scale of disease transmission.
+Like the instantaneous reproduction number, the instantaneous generation-interval distribution describes when an individual infected at time $t$ will transmit infections if conditions were to remain the same.
 Both the reproduction number and the generation-interval distribution depend on many factors, including intrinsic infectiousness of an infected individual, non-pharmaceutical interventions, awareness-driven behavior, and population-level susceptibility \citep{fraser2007estimating}.
+
+Under strength-like changes that reduce the hazard of transmission by a constant amount (e.g., susceptible depletion and decreased transmission rates), the instantaneous generation-interval distribution does not across change time \citep{fraser2007estimating}---
+in this case, the instantaneous generation-interval distribution is also often referred to as the intrinsic generation-interval distribution \citep{champredon2015intrinsic,champredon2018two,gostic2020practical,park2020time}.
+For example, the standard SEIR model can be equivalently expressed as a renewal equation with time-invariant $g(\tau)$ as shown in \citep{champredon2018equivalence}.
+The instantaneous generation-interval distribution is also different from the realized generation-interval distribution, which measures time between actual infection events (see \citep{champredon2015intrinsic}).
+Previous studied have noted, in many contexts, that the realized generation intervals can contract due to susceptible depletion---a special case of strength-like changes---due to decreases in transmission even though the instantaneous generation-interval distribution remains unchanged.
+In the context of SARS-CoV-2, assuming a time-invariant instantaneous generation-interval distribution is often appropriate given that the dynamics of its spread have been primarily driven by strength-like interventions, especially social distancing measures and vaccination.
 
@@
 Instead, many biological mechanisms appear compatible with assuming a constant value of relative strength $\rho$ over changing conditions.
-For example, if the proportion of the population susceptible declines, or the average contact rate changes, while other factors remain constant, the relative strength $\rho$ is expected to remain constant;
-in general, this will imply a change in relative speed $\delta$.
+For example, if the proportion of the population susceptible declines, or the average contact rate changes, while other factors remain constant, the relative strength $\rho$ is expected to remain constant \citep{leung2017monitoring,leung2020empirical,di2021impact,leung2021early};
+as we show in this section, this will imply a change in relative speed $\delta$.
 
 We thus investigate how $\delta$ is expected to change with \Rw if $\rho$ remains constant, and how this expectation changes with the ratio of the generation intervals. 
-Once again, we rely on the Gamma assumption to find the relative speed $\delta$ given the mean generation interval of the variant $\Gv$ and the wild type $\Gw$,:
+Once again, we rely on the Gamma assumption \citep{park2019practical} to find the relative speed $\delta$ given the mean generation interval of the variant $\Gv$ and the wild type $\Gw$:
 \begin{equation}
@@
 \end{equation}
-As our baseline scenario we assume $\rho = 1.61$, which is the value we obtain for $\delta=0.1\pday$, $\rw=0\pday$, $\Gw = \Gv = 5\,\textrm{days}$, and $\kappa = 1/5$.
+As our baseline scenario we assume $\rho = 1.61$, which is the value we obtain for $\delta=0.1\pday$ \citep{davies2021estimated}, $\rw=0\pday$, $\Gw = \Gv = 5\,\textrm{days}$, and $\kappa = 1/5$ \citep{ferretti2020quantifying}.
 We evaluate $\delta$ across five scenarios as before: (1) $\Rw < \Rv < 1$, (2) $\Rw < \Rv = 1$, (3) $\Rw < 1 < \Rv$, (4) $1 = \Rw < \Rv$, and (5) $1 < \Rw < \Rv$.
@@
 For example, even when there are no differences in the generation-interval distributions ($\Gv=\Gw$, in this case), the relative speed $\delta$ can range between 0.08--0.11 when $\kappa=0.2$ and 0.06--0.14 when $\kappa=0.5$.
-% Therefore, characterizing the spread of variants assuming constant relative speed (e.g., by fitting a standard logistic growth curve) without considering how epidemiological conditions change over time should be used with care.
+Therefore, characterizing the spread of variants assuming constant relative speed (by fitting a standard logistic growth curve; e.g., \citep{campbell2021increased,switzerland2021variant}) should be used with care.
 
@@
 
-Instead of estimating relative strength from speed, one can estimate time-varying or instantaneous reproduction numbers $\RR(t)$ of the variant and the wild type from incidence data \citep{fraser2007estimating}, and directly compare their ratios;
-such methods have been used in previous analyses of the B.1.1.7 strain by \cite{volz2021transmission}.
+Instead of inferring relative strength from speed, one can directly estimate time-varying reproduction numbers $\RR(t)$ of the variant and the wild type from incidence data and directly compare their ratios when the incidence of infection and the generation-interval distributions are known---such methods have been used in previous analyses of the B.1.1.7 strain by \cite{volz2021transmission}.
+Broadly, there are two types of reproduction numbers: case reproduction number and instantaneous reproduction number.
+The case reproduction number is defined as the average number of secondary infections caused by an individual infected at time $t$ and therefore depends on transmission after time $t$ \citep{wallinga2004different}.
+The instantaneous reproduction number is defined as the average number of secondary infections that would be caused by an individual infected at time $t$ if conditions were to remain the same \citep{fraser2007estimating};
+therefore, the instantaneous reproduction number only depends on transmission at time $t$ and is most appropriate for real-time evaluation of changes in trasmission, as argued by \citep{gostic2020practical}.
+The estimation of the instantaneous reproduction number was popularized by \cite{cori2013new} and has been widely adopted in epidemiological analyses of SARS-CoV-2 \citep{abbott2020estimating,knight2020estimating,flaxman2020Rt,brauner2021inferring,li2021temporal}.
+
+In practice, estimating the instantaneous reproduction number is complicated given that it requires incidence of infection.
+Observed case counts are sensitive to reporting delays \citep{goldstein2009reconstructing} and changes in case definitions \citep{tsang2020effect}, which in turn can affect estimates of the instantaneous reproduction number \citep{gostic2020practical}.
+Here, we choose to focus here on the underlying dynamical mechanisms that may affect inference and assume that the incidence of infection is exactly known.
 Assuming that the generation-interval distribution remains constant, the instantaneous reproduction numbers of the new variant and of the wild type can be estimated from their corresponding incidence curves---an approach popularized by \cite{cori2013new}:
@@
 
-We also assumed that the incidence of infection caused by the variant and the wild type is known.
-However, estimating $\RR(t)$ from real data is often associated with a variety of practical challenges, including delays between infection and reporting and changes in testing patterns \citep{gostic2020practical}.
-We have chosen to focus here on the underlying dynamical mechanisms that may affect inference.
-Future studies may consider the development of methods based on mechanism outlined here to assess potential biases in their estimates of relative strength.
+% SWP: in case you're wondering why this is commented out, I have moved this to section 5.
+% We also assumed that the incidence of infection caused by the variant and the wild type is known.
+% However, estimating $\RR(t)$ from real data is often associated with a variety of practical challenges, including delays between infection and reporting and changes in testing patterns \citep{gostic2020practical}.
+% We have chosen to focus here on the underlying dynamical mechanisms that may affect inference.
+% Future studies may consider the development of methods based on mechanism outlined here to assess potential biases in their estimates of relative strength.
 

### 2021-07-07 response.tex
updating

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}

\usepackage{color}

\usepackage{xspace}

\usepackage{lmodern}
\usepackage{amssymb,amsmath}

\usepackage[pdfencoding=auto, psdextra]{hyperref}

\usepackage{natbib}
\bibliographystyle{chicago}

\newcommand{\eref}[1]{Eq.~(\ref{eq:#1})}
\newcommand{\fref}[1]{Fig.~\ref{fig:#1}}

\newcommand{\rR}{\mbox{$r$--$\cal R$}}
\newcommand{\RR}{\ensuremath{{\cal R}}}
\newcommand{\RRhat}{\ensuremath{{\hat \cal R}}}
\newcommand{\Rx}[1]{\ensuremath{{\cal R}_{#1}}} 
\newcommand{\Ro}{\ensuremath{{\mathcal R}_{0}}\xspace}
\newcommand{\Rs}{\Rx{\mathrm{s}}}
\newcommand{\Rpool}{\ensuremath{{\mathcal R}_{\textrm{\tiny{pool}}}}\xspace}
\newcommand{\Reff}{\Rx{\mathit{eff}}}
\newcommand{\Tc}{\ensuremath{C}}

\newcommand{\dd}[1]{\ensuremath{\, \mathrm{d}#1}}
\newcommand{\dtau}{\dd{\tau}}
\newcommand{\dx}{\dd{x}}
\newcommand{\dsigma}{\dd{\sigma}}

\newcommand{\rev}{\subsection*}
\newcommand{\revtext}{\textsf}
\setlength{\parskip}{\baselineskip}
\setlength{\parindent}{0em}

\newcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textsl{#3}\textbf{]}}}
\newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}
\newcommand{\dc}[1]{\comment{blue}{DC}{#1}}
\newcommand{\jsw}[1]{\comment{green}{JSW}{#1}}
\newcommand{\hotcomment}[1]{\comment{red}{HOT}{#1}}

\newcommand{\psymp}{\ensuremath{p}} %% primary symptom time
\newcommand{\ssymp}{\ensuremath{s}} %% secondary symptom time
\newcommand{\pinf}{\ensuremath{\alpha_1}} %% primary infection time
\newcommand{\sinf}{\ensuremath{\alpha_2}} %% secondary infection time

\newcommand{\psize}{{\mathcal P}} %% primary cohort size
\newcommand{\ssize}{{\mathcal S}} %% secondary cohort size

\newcommand{\gtime}{\tau_{\rm g}} %% generation interval
\newcommand{\gdist}{g} %% generation-interval distribution
\newcommand{\idist}{\ell} %% incubation period distribution

\newcommand{\total}{{\mathcal T}} %% total number of serial intervals

\begin{document}

\noindent Dear Editor:

Thank you for the chance to revise and resubmit our manuscript. 
We have made major revisions to our manuscript to address the reviewers' comments.

\rev{Reviewer \#1}

\revtext{This is a review of ``Roles of generation-interval distributions in shaping relative epidemic strength, speed, and control of new SARS-CoV-2 variants'' by Park and others for Royal Society Interface.  This is the latest in an ongoing series of papers (Champredon 2015, Champredon 2018, Park 2019, Dushoff 2021, ...) in which the last author has considered aspects of incubation, growth, and reproduction in SIR epidemic theory.  In this manuscript, the authors use theory to anticipate the effects of COVID-19 variants where the distribution of infectivity evolves.}

\revtext{The manuscript is excellently written. The figures are clear.  All the mathematics seems satisfactory.  Unfortunately, the manuscript does not make its case for novel results.}

Thank you.

\revtext{The title is overly-broad as it suggests consideration of general forms of generation interval distributions (along the lines of Miller 2010 10.1016/j.jtbi.2009.08.007), when infect the authors study variation around the time of peak infectivity using a (very standard) $\gamma$-chain trick.  The study of generation time and $\gamma$ chains has a long history from the cited-but-unreviewed Svensson 2007, Wallinga 2007, to uncited Kenah 2008, A.  Lloyd 2001, D. Anderson 1980, Bailey 1964 (see Park 2019 and Champredon 2015).  In light of these and other works, the title and abstract fail to identify novel contributions to our understanding of epidemic theory.}

Thank you for the suggestions.

\revtext{Although the authors are aware of the history of research on this topic, they do little to review the relevant material, leaving this manuscript feeling like a superficial re-hashing of obvious material.  The introduction doesn't provide any serious review of the theory}

Done. We do not review gamma chain trick in depth because we're not doing it.

\revtext{Section 4 doesn't give references to prior work, and section 5 ignores almost all of the literature on the topic of estimating R from incidence data.}

Section 4 done. Section 5 done.

\revtext{Other comments:}

\revtext{Topic on which there is a considerable confusion in terminology in the literature.  In addition to "generation interval", terms "generation time", "transmission interval", "serial interval", and others have been used in the literature, all with related by different definitions.  This manuscripts definition "time between infection and transmission" implies a single transmission event, but Kenah pointed out that time-contraction appears when allowing multiple transmission events (see Champredon 2015).  A clearer definition up front will help readers.  Fortunately, the results don't depend on the specific definition.}

Done.

\revtext{The 'speed'-'strength' distinction from the previous paper is seductive, but perhaps leaning into general quantitative illiteracy.  When one pays attention to units -- "speed" L/T,  "Growth rate" 1/T while "strength" has dimensions of force, typically, but reproduction number is a dimensionless ratio of cases per case.  Beyond time-independent final-size relations (which are themselves not robust to effects like behavior change), several researchers have argued that growth rates generally more useful than reproduction numbers (giving equivalent threshold conditions, for example) and not as prone to mis-interpretation.  Other authors in evolutionary biology and economics have argued that the reproduction number's scientific value is best salvaged using the "discounted reproduction success" (McNamara, Houston, and Collins 2001), rather than fighting of an R-r straw-man.}

No.

\revtext{Equation 1 As Kermack and McKendrick introduced it, the compartmental model is a convenient *specialization* of the integral models-- see the nice re-analysis by Breda et al., 2012.}

Thank you for pointing this out. We now cite Breda et al. We acknowledge that ``This framework ... generalizes compartmental model, such as the SEIR model''

\revtext{Equation 4 assumes the wild-type and variant co-dominate the population.  This may not be the cases when multiple variants are co-circulating, particularly during replacement events.}

Done.

\revtext{p 4 l55.  Shape of the distribution matters -- heavy tails and failed central concentration.  This detail will escape the naive reader unless we poke them a little more with it.}

Reworded:

``For simplicity we neglect differences in the squared coefficient of variation and assume $\kappa_{\mathrm{wt}} = \kappa_{\mathrm{var}} = \kappa$.
While the exact shape of the distribution matters, we choose to focus on differences in the mean generation intervals to provide a qualitative understanding of how differences in generation-interval distributions between the wild type and the variant affects the relationship between relative strength and speed.
In general, we expect  differences in the squared coefficient of variation to have qualitatively similar effects as differences in the mean.
For example, a wider distribution allows for more early transmission, which in turn reduces epidemic strength given speed---this effect is qualitatively similar to that of a distribution with a shorter mean.''

\rev{Reviewer \#2}

\revtext{This is a well-written paper about the role of the generation interval distribution in the estimation of the relative strength (reproduction number) and speed
(epidemic growth rate) of a new variant of a disease. Using a renewal equation
framework, they consider inferring relative strength from relative speed and vice
versa as well as inferring relative strength from incidence data. While all of the
analysis in the paper is sound, I am concerned that the strong assumptions made
(mass-action epidemic, constant generation interval distribution, etc.) will limit
the reliability of the results in a practical application. However, they clearly
establish that assumptions about the generation interval distribution have important implications for the analysis of novel variants. While I am generally
skeptical of population-level analyses of infectious disease transmission based
on generation intervals, these are done frequently so this is a useful insight.
More detailed comments are below.}

Thank you. We have tried to make the limitation more clear.

\revtext{1. (page 3, first paragraph) It is important to acknowledge that the renewal equation approach to the relationship between the generation interval
distribution, R, and the growth rate r assumes a mass-action epidemic.
When we assume a fixed generation interval distribution, we must add the
assumption that the rate of depletion of susceptibles over time is negligible and the prevalence of infection is low. One of the most important
findings of Ref 10 is that the mean generation interval contracts when the
rate of depletion of susceptibles increases. While the renewal equation
does allow for the generation interval to change over time, the analyses
in Sections 3–5 assume fixed mean generation intervals. Because variants
are likely to spread when susceptibles are being depleted by both variants
(due to cross-protection) and when the prevalence of infection is not low, these implicit assumptions place significant limitations on the practical
implications of the analysis.}

This is a confusion between the intrinsic and realized. Also done.

\revtext{ (page 2, lines 55–60) It would help to clarify what is meant by “strengthlike” and “speed-like” in interventions. From the examples, it appears that
speed-like interventions work by identifying potential infections quickly
(effectively ending the infectious period early while not necessarily altering
the hazard of infectious contact) and strength-like interventions work by
reducing the hazard of infectious contact (while not necessarily altering
the duration of infectiousness). A good explanation of these ideas is on
lines 228–231 on page 10.}

Done.

\revtext{(page 5, lines 126–128 and 137) It would help to point out that you assume
rwt < rvar when you introduce the five scenarios.}

Done.

\revtext{(page 9, Figure 3) I would order the rows in order of the assumed mean
generation interval of the variant: 4 days at the top, 5 days in the middle,
and 6 days on the bottom.}

Done.

\revtext{(page 13, lines 282–285) While contact tracing and household studies are
susceptible to a number of biases, they are likely to be a far more reliable
source of information about the transmissibility and generation intervals
of variants than population-level analyses of incidence data.}

Thank you for pointing this out. We have re-written this part:

``Although these estimates are more consistent with the attack rate analysis \citep{ukinvest},
we do not claim that they are necessarily more accurate---while individual-level data from contact tracing can provide a more reliable source of information about the transmissibility and time scale of transmission in some cases, they can be also biased towards particular types of contacts that are easily identified such as household contacts, which could also affect the estimate of $\rho$.''

\revtext{(page 15, References) Curly brackets might need to be added to some of
the BibTeX entries to enforce correct capitalization (e.g., Ref. 9).}

Thank you.

\swp{Ref 9 is ``Infectious diseases of humans: dynamics and control.'' Do we capitalize this or no?}

\rev{Reviewer \#3}

\revtext{In this manuscript, Park and colleagues present a thoughtful analysis of how assumptions about the generation interval distribution for wild-type and variant SARS-CoV-2 can bias estimates of the relative growth rate ("speed") and relative reproduction number ("strength") of the variant. The analysis is mathematically sound and offers a thorough investigation of an important concept that has been the subject of much discussion in the epidemiological community but, to my knowledge, hasn't been given the formal treatment it deserves until now. My comments are mostly minor and intended to make the findings more accessible and transferable to epidemiological practice.}

Thank you.

\revtext{Abstract: 37: On first reading, the distinction between 'speed-like' and 'strength-like' interventions wasn't clear to me, nor why contact tracing should target r but not R, and why social distancing should target R but not r. Could you use more recognizable terms here to highlight how these interventions differ from one another? (is the key difference that contact tracing reduces the number of infectious individuals and thus the probability that a susceptible person comes into contact with an infectious person, while social distancing reduces the per-contact chance of transmission? Or something else?) This comes up again in line 57 of the Introduction.}

Thank you for pointing this out. We have removed 'speed-like' and 'strength-like' interventions from the abstract as these ideas can be confusing without further explanation (which is difficult given word limit). Instead, we have tried to make these ideas clearer throughout text, including in the introduction. We have also revised the abstract according to reviewer 1's comments.

\revtext{166: I think it would be worth stating the range of Rwt and/or Rvar that yields this range for delta (I think it's something like 0.49-1.27 for the wild type, 0.78-2.04 for the variant, based on Figure 2 - but would be helpful if the reader didn't have to derive). Is the varying R what you're referring to when you speak of "underlying epidemiological conditions" in 166-167?}

Thank you for this suggestion.

\revtext{184: I think that "inference data" should be "incidence data"}

Done.

\revtext{Figure 4: I found this figure quite difficult to parse and not especially informative. Are the vertical bars the means of these distributions? If so, they should be labeled and referenced as such in the caption. Also in the figure 4 caption: the term ``generation intervals'' (plural) struck me as a bit strange - I think the singular would make more sense and be more consistent with what's used in the literature. Also, it would be helpful to put theta and phi in the vertical axis labels for E and F (e.g., "Strength (theta)" and "Speed (phi; 1/days)"). Unless this is observed speed and strength, meaning Rvar and rvar? In E, I'm also confused why the strength of the "constant-strength" line is increasing - shouldn't it be flat? Or what have I missed? I think some more details to refer back to the variables in the text might be helpful - for example, I think that (F) is the "Pre-intervention (black) and post-intervention speed of variant transmission (rvar) conditional on the mean generation interval of the variant (Gvar, colored)."}

\revtext{Really what I would like to see here (Figure 4) is some simulated epidemic curves and perhaps final epidemic sizes - I think this would go much further to communicate the intuition behind how the speed-based and strength-based interventions differ than what's currently presented. One of the main conclusions of the paper seems to be that speed-based interventions could be more effective than strength-based interventions when the mean generation interval is long; presenting such simulations would be a good way to drive this point home and to give intuition for just how different speed-based vs strength-based interventions actually are.}

\revtext{264: "epidemiological dynamics" is used again here - what specifically do you mean?}

\revtext{265: "can also lead to different conclusions about the effectiveness of interventions" - what sort of different conclusions? In which circumstances?}

\revtext{271: "If new variants has a longer infectious period" (typo)}

Done.

\revtext{Missing equation reference in the Supplement.}

\bibliography{newvariant_abbv.bib}

\end{document}

### 2021-07-07 newvariant.tex
updating

@@
 Inferring the relative strength (i.e., the ratio of reproduction numbers, $\Rv/\Rw$) and relative speed (i.e., the difference between growth rates, $\rv-\rw$) of new SARS-CoV-2 variants compared to their wild types is critical to predicting and controlling the course of the current pandemic.
-Multiple studies have estimated the relative strength of new variants from the observed relative speed, but they typically neglect the possibility that the new variants have different generation intervals (i.e., time between infection and transmission), which determines the relationship between relative strength and speed.
-Notably, the increasingly predominant B.1.1.7 variant may have a longer infectious period (and therefore, a longer generation interval) than prior dominant lineages.
-Here, we explore how differences in generation intervals between a new variant and the wild type affect the relationship between relative strength and speed.
+Although the role of the generation interval (i.e., time between infection and transmission) in linking epidemic strength and speed has been previously well-established, many analyses of new SARS-CoV-2 variants overlook uncertainties in the generation-interval distributions as well as the possibility that their inferences of epidemic dynamics can depend on whether relative strength or speed are held constant.
+Here, we explore how differences in the generation-interval distributions affect the inference of relative strength and speed in light of recent evidence that the B.1.1.7 variant may have a longer infectious period (and therefore, a longer generation interval) than prior dominant lineages.
 We use simulations to show how neglecting these differences can lead to biases in estimates of relative strength in practice and to illustrate how such biases can be assessed.
-Finally, we discuss implications for control: if new variants have longer generation intervals then speed-like interventions such as contact tracing become more effective, whereas strength-like interventions such as social distancing become less effective.
+We also assess which of the two assumptions (i.e., constant relative strength or speed) are more appropriate and further discuss implications for control.
 
@@
 While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the disease and its control because two quantities are linked by the generation-interval distribution \citep{doi:10.1098/rspb.2020.1556}.
-For example, when epidemic strength is fixed, assuming longer generation intervals lead to a slower epidemic (lower $r$), making the epidemic look easier to control with speed-like interventions such as contact tracing.
-When epidemic speed is fixed, assuming longer generation intervals lead to a stronger epidemic (higher $\RR$), making the epidemic look harder to control with strength-like interventions such as social distancing.
+For example, when epidemic strength is fixed, assuming longer generation intervals lead to a slower epidemic (lower $r$), making the epidemic look easier to control with speed-like interventions (e.g., contact tracing), which depend on hazard of isolation and therefore affects late transmission disproportionately.
+When epidemic speed is fixed, assuming longer generation intervals lead to a stronger epidemic (higher $\RR$), making the epidemic look harder to control with strength-like interventions (e.g., social distancing), which reduce the hazard of infectious contact and therefore affecting transmission by an equal amount regardless of age of infection.
 Therefore, it is critical to evaluate which assumption is more appropriate for inferring transmission advantages of new variants.
@@
 \end{equation}
-This framework provides a flexible way of modeling disease dynamics and generalizes compartmental model, such as the SEIR model \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, champredon2018equivalence}.
+This framework provides a flexible way of modeling disease dynamics and generalizes compartmental model, such as the SEIR model \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral,breda2012formulation, champredon2018equivalence}.
 
@@
 
-Under strength-like changes that reduce the hazard of transmission by a constant amount (e.g., susceptible depletion and decreased transmission rates), the instantaneous generation-interval distribution does not across change time \citep{fraser2007estimating}---
-in this case, the instantaneous generation-interval distribution is also often referred to as the intrinsic generation-interval distribution \citep{champredon2015intrinsic,champredon2018two,gostic2020practical,park2020time}.
-For example, the standard SEIR model can be equivalently expressed as a renewal equation with time-invariant $g(\tau)$ as shown in \citep{champredon2018equivalence}.
+Under strength-like changes that reduce the hazard of transmission by a constant amount (e.g., susceptible depletion and decreased transmission rates), the instantaneous generation-interval distribution does not across change time \citep{fraser2007estimating}.
+In this case, the instantaneous generation-interval distribution is also often referred to as the intrinsic generation-interval distribution \citep{champredon2015intrinsic,champredon2018two,gostic2020practical,park2020time}---
+for example, the standard SEIR model can be equivalently expressed as a renewal equation with time-invariant intrinsic generation-interval distribution $g(\tau)$ as shown in \citep{champredon2018equivalence}.
 The instantaneous generation-interval distribution is also different from the realized generation-interval distribution, which measures time between actual infection events (see \citep{champredon2015intrinsic}).
-Previous studied have noted, in many contexts, that the realized generation intervals can contract due to susceptible depletion---a special case of strength-like changes---due to decreases in transmission even though the instantaneous generation-interval distribution remains unchanged.
-In the context of SARS-CoV-2, assuming a time-invariant instantaneous generation-interval distribution is often appropriate given that the dynamics of its spread have been primarily driven by strength-like interventions, especially social distancing measures and vaccination.
+Previous studied have noted, in many contexts, that the realized generation intervals can contract due to susceptible depletion---a special case of strength-like changes---due to decreases in transmission \citep{kenah2008generation,nishiura2010time,champredon2015intrinsic}, even though the instantaneous generation-interval distribution remains unchanged.
+While the instantaneous generation-interval distribution can change under speed-like changes that disproportionately affect late transmissions (see \cite{fraser2007estimating} and Section 6 for discussion), assuming a time-invariant instantaneous generation-interval distribution is often appropriate in the context of SARS-CoV-2, given that the dynamics of its spread have been primarily driven by strength-like interventions, such as social distancing measures \citep{flaxman2020Rt} and vaccination \citep{moore2021vaccination}.
+Therefore, we neglect changes in the intrinsic generation-interval distribution over time for now and return to this idea in Section 6.
 
 Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$.
+Here, we use the term ``epidemiological conditions'' to broadly refer to all factors that affect transmission---mathematically, they are captured by the kernel $K(t, \tau)$.
+In the context of SARS-CoV-2, we are essentially assuming that the changes in susceptible pool, behavior, and contact rates are negligible over a short period of time.
 Then, the incidence of each strain grows (or decays) exponentially at rate $r_x$, satisfying the Euler-Lotka equation \citep{wallinga2007generation}:
@@
 We use this framework to investigate how inferences about strength and speed of the variant depend on our assumptions about the underlying generation-interval distributions.
-For simplicity we neglect differences in the squared coefficient of variation and assume $\kappa_{\mathrm{wt}} = \kappa_{\mathrm{var}} = \kappa$ and focus on differences in the mean generation intervals.
+For simplicity we neglect differences in the squared coefficient of variation and assume $\kappa_{\mathrm{wt}} = \kappa_{\mathrm{var}} = \kappa$.
+While the exact shape of the distribution matters, we choose to focus on differences in the mean generation intervals to provide a qualitative understanding of how differences in generation-interval distributions between the wild type and the variant affects the relationship between relative strength and speed.
+In general, we expect  differences in the squared coefficient of variation to have qualitatively similar effects as differences in the mean.
+For example, a wider distribution allows for more early transmission, which in turn reduces epidemic strength given speed---this effect is qualitatively similar to that of a distribution with a shorter mean.
 
@@
 Focusing on proportions has the advantage, because changes in proportions are less sensitive to changes in testing and to other transient effects that would affect variants and wild type viruses similarly.
-When incidence is changing exponentially ($i_x(t) = i_x(t_0) \exp(r_x t)$), the proportion of the new variant $p(t)$ follows a logistic growth curve:
+When incidence is changing exponentially ($i_x(t) = i_x(t_0) \exp(r_x t)$), the proportion of the new variant $p(t)$ follows a logistic growth curve \citep{switzerland2021variant,davies2021estimated}:
 \begin{align}
@@
 \end{align}
-where the logistic growth rate $\delta = \rv - \rw$ corresponds to the relative speed of the epidemic.
+where the logistic growth rate $\delta = \rv - \rw$ (i.e., the slope of the linear regression at the logit scale) corresponds to the relative speed of the epidemic.
+When more than two strains are co-circulating, the growth rate at a logit scale is expected to change over time and does not correspond to the relative speed even when all strains exhibit exponential changes (see Supplementary Text);
+instead, as shown in \citep{campbell2021increased}, a multinomial logistic regression model is required to infer relative speed.
+Since we are interested in the relationship between the true relative speed and strength, rather than the inference of relative speed from proportion data, our results are insensitive to the number of strains circulating in the population.
 
@@
 To further explore how inference depends on underlying epidemiological conditions, we consider five scenarios: (1) $\rw < \rv < 0$, (2) $\rw < \rv = 0$, (3) $\rw < 0 < \rv$, (4) $0 = \rw < \rv$, and (5) $0 < \rw < \rv$.
+In all of these cases, we always assume $\rw < \rv$ to reflect the observed relative speed of $\delta = 0.1\pday$.
 We consider the spread of B.1.1.7 as an example but our qualitative conclusions should hold for other VoCs.
@@
 \end{equation}
-As our baseline scenario we assume $\rho = 1.61$, which is the value we obtain for $\delta=0.1\pday$ \citep{davies2021estimated}, $\rw=0\pday$, $\Gw = \Gv = 5\,\textrm{days}$, and $\kappa = 1/5$ \citep{ferretti2020quantifying}.
+As our baseline scenario we assume $\rho = 1.61$ (and therefore, we always have $\Rw < \Rv$), which is the value we obtain for $\delta=0.1\pday$ \citep{davies2021estimated}, $\rw=0\pday$, $\Gw = \Gv = 5\,\textrm{days}$, and $\kappa = 1/5$ \citep{ferretti2020quantifying}.
 We evaluate $\delta$ across five scenarios as before: (1) $\Rw < \Rv < 1$, (2) $\Rw < \Rv = 1$, (3) $\Rw < 1 < \Rv$, (4) $1 = \Rw < \Rv$, and (5) $1 < \Rw < \Rv$.
@@
 
-\fref{relspeed} also shows that when $\rho$ is fixed relative speed depends on underlying epidemiological conditions. 
-For example, even when there are no differences in the generation-interval distributions ($\Gv=\Gw$, in this case), the relative speed $\delta$ can range between 0.08--0.11 when $\kappa=0.2$ and 0.06--0.14 when $\kappa=0.5$.
+\fref{relspeed} also shows that when $\rho$ is fixed relative speed depends on underlying epidemiological conditions---that is, whether the incidence of wild type and the variant are increasing or decreasing.
+For example, even when there are no differences in the generation-interval distributions ($\Gv=\Gw$, in this case), changing $\Rw$ from 0.49 to 1.27 (and thus $\Rv$ from 0.79 to 2.04) yields the relative speed $\delta$ between 0.08--0.11 for $\kappa=0.2$ and 0.06--0.14 for $\kappa=0.5$.
 Therefore, characterizing the spread of variants assuming constant relative speed (by fitting a standard logistic growth curve; e.g., \citep{campbell2021increased,switzerland2021variant}) should be used with care.
@@
 
-Here, we investigate how misspecification of the generation-interval distribution of the variant affects our inference of relative strength from inference data under the assumption that the true generation-interval distribution of the wild type is known.
+Here, we investigate how misspecification of the generation-interval distribution of the variant affects our inference of relative strength from incidence data under the assumption that the true generation-interval distribution of the wild type is known.
 We use a two-strain renewal equation that assumes perfect cross-immunity to simulate three different scenarios (see Supplementary Text):
-(1) the wild type and the variant have the same generation-interval distributions (which match the known distribution, shown in \fref{Rtbias}A--C);
-(2) the variant has a shorter mean generation interval (\fref{Rtbias}D--F); and
+(1) the variant has a shorter mean generation interval (\fref{Rtbias}A--C);
+(2) the wild type and the variant have the same generation-interval distributions (which match the known distribution, shown in \fref{Rtbias}D--F); and
 (3) the variant has a longer mean generation interval (\fref{Rtbias}G--I).
@@
 For all simulations, the assumed mean generation interval is equal to the mean generation interval of the wild type (5 days), and the squared coefficient of variation in generation intervals is equal to $\kappa = 1/5$.
-The mean generation interval of the variant is equal to 5 days (top row), 4 days (middle row), and 6 days (bottom row).
+The mean generation interval of the variant is equal to 4 days (top row), 5 days (middle row), and 6 days (bottom row).
 }
@@
 
-When the assumed distribution matches the true distribution, the estimated reproduction numbers match the true values (\fref{Rtbias}A); thus, their ratio remains constant and $\hat{\RR}_{\textrm{var}}(t)=\Rv(t)$ (\fref{Rtbias}B).
-However, when the generation-interval distribution of the variant differs from the assumed distribution, the ratio changes over time (\fref{Rtbias}E,H).
-If the true generation intervals of the variant have a shorter mean than the assumed distribution, we over-estimate $\Rv(t)$ during the growth phase and under-estimate $\Rv(t)$ during the decay phase (and conversely, \fref{Rtbias}D,G), which further translates to biases in the estimated relative strength (\fref{Rtbias}E,H).
+When the assumed distribution matches the true distribution, the estimated reproduction numbers match the true values (\fref{Rtbias}D); thus, their ratio remains constant and $\hat{\RR}_{\textrm{var}}(t)=\Rv(t)$ (\fref{Rtbias}E).
+However, when the generation-interval distribution of the variant differs from the assumed distribution, the ratio changes over time (\fref{Rtbias}B,H).
+If the true generation intervals of the variant have a shorter mean than the assumed distribution, we over-estimate $\Rv(t)$ during the growth phase and under-estimate $\Rv(t)$ during the decay phase (and conversely, \fref{Rtbias}A,G), which further translates to biases in the estimated relative strength (\fref{Rtbias}B,H).
 
@@
 For example, we can estimate $\rho$ by plotting the estimated strength of the variant $\hat{\RR}_{\textrm{var}}(t)$ against the estimated strength of the wild type $\hat{\RR}_{\textrm{wt}}(t)$---as presented in Figure 2 of \cite{volz2021transmission}---and performing a linear regression (\fref{Rtbias}C,F,I).
-If $\rho$ is constant, and generation-interval distributions are correctly specified, we obtain a straight line with a slope of $\rho$ and intercept at zero (\fref{Rtbias}C).
-However, when the assumed mean generation interval is longer than the that of the variant, we over-estimate the slope (and conversely, \fref{Rtbias}F,I).
+If $\rho$ is constant, and generation-interval distributions are correctly specified, we obtain a straight line with a slope of $\rho$ and intercept at zero (\fref{Rtbias}F).
+However, when the assumed mean generation interval is longer than the that of the variant, we over-estimate the slope (and conversely, \fref{Rtbias}C,I).
 Biases in slopes further translate to biases in intercepts; in theory, we would expect the regression line to go through the origin (because $\Rv = 0$ when $\Rw = 0$).
@@
 \textbf{Effects of constant-strength and constant-speed interventions on the spread of a new variant with known speed \rv.}
-(A--B) Pre-intervention (black) and post-intervention (colored) kernel of the new variant under constant-strength (A) and -speed (B) interventions assuming that the variant has equal generation intervals as the wild type.
+(A--B) Pre-intervention (black) and post-intervention (colored) kernel of the new variant under constant-strength (A) and -speed (B) interventions assuming that the variant has equal mean generation interval as the wild type.
 (C--D) Pre-intervention (black) and post-intervention kernel of the new variant assuming that the variant has longer generation intervals than the wild type.
-(E) Pre-intervention (black) and post-intervention strength conditional on the mean generation interval of the variant (colored).
-(F) Pre-intervention (black) and post-intervention speed conditional on the mean generation interval of the variant (colored).
+Vertical lines represent the mean generation interval.
+(E) Pre-intervention $\Rv$ (black) and post-intervention epidemic strength $\Ry{post}$ conditional on the mean generation interval of the variant (colored).
+(F) Pre-intervention $\rv$ (black) and post-intervention epidemic speed $\ry{post}$ conditional on the mean generation interval of the variant (colored).
 Strength and speed are calculated assuming $\rw=0\pday$, $\delta=0.1\pday$, $\rv=0.1\pday$, $\Gy{wt}=5\,\textrm{days}$, and $\kappa=1/5$ for pre-intervention conditions.
+The constant-strength intervention assumes that the ratio $\Rv/\Ry{post} = \theta$ remains constant.
+The constant-speed intervention assumes that the difference $\rv - \ry{post} = \phi$ remains constant.
 Intervention strength and speed are chosen so that post intervention strength of the new variant is 0.9 when its mean generation interval is 5 days.
@@
 While our analysis supports estimates of a higher strength of the new variant across a wide range of assumptions, it also shows that uncertainty in generation-interval distributions must be taken into account to obtain accurate estimates of the relative strength of variants.
-If new variants has a longer infectious period \citep{kissler2021densely}, and therefore extended generation-interval distributions, current estimates may underestimate the true relative strength.
+If new variants have a longer infectious period \citep{kissler2021densely}, and therefore extended generation-interval distributions, current estimates may underestimate the true relative strength.
 % If new variants escape vaccinal or natural immunity to some extent \citep{Kupferschmidt329}, their relative strength may increase further.
@@
 Although these estimates are more consistent with the attack rate analysis \citep{ukinvest},
-we do not claim that they are more accurate---
-in particular, contact tracing data can be biased towards particular types of contacts that are easily identified such as household contacts, which could also affect the estimate of $\rho$.
+we do not claim that they are necessarily more accurate---
+while individual-level data from contact tracing can provide a more reliable source of information about the transmissibility and time scale of transmission in some cases, they can be also biased towards particular types of contacts that are easily identified such as household contacts, which could also affect the estimate of $\rho$.
 Instead, this calculation simply highlights the importance of propagating uncertainty in generation-interval distributions in assessing the relative strength and speed of SARS-CoV-2 variants.

### 2021-07-07 supp.tex
updating

@@
 
+\subsection{Measuring relative growth for multiple co-circulating strains}
+
+In order to model multiple co-circulating strain, we write $i_x(t)$ to represent the incidence of strain $x$ (for $x = 1, \dots, N$).
+When incidence is changing exponentially ($i_x(t) = i_x(t_0) \exp(r_x t)$), the proportion of strain $y$ in the population can be written as:
+\begin{equation}
+p_y(t) = \frac{i_y(t_0) \exp(r_y t)}{\sum_{x=1}^N i_x(t_0) \exp(r_x t)}.
+\end{equation}
+Logit transforming $p_y(t)$ then gives:
+\begin{align}
+\mathrm{logit}(p_y(t)) &= \log\left(\frac{p_y(t)}{1-p_y(t)}\right)\\
+&= \log\left(\frac{i_y(t_0) \exp(r_y t)}{\sum_{x \neq y} i_x(t_0) \exp(r_x t)}\right)\\
+&= \log(i_y(t_0)) + r_y t - \log \left(\sum_{x \neq y} i_x(t_0) \exp(r_x t)\right)
+\end{align}
+which is a nonlinear function of time $t$.
+Therefore, the growth rate measured at the logit scale is expected to change over time when multiple strains are co-circulating. 
+Instead, a multinomial logistic regression model is required to infer relative speed \citep{campbell2021increased}.
+
 \subsection{Two-strain renewal equation}

### 2021-07-07 Rtbias.R

@@ -16,16 +16,16 @@ genfun_assumed <- function(x) dgamma(x, 5, 5/5)
 genfun_short <- function(x) dgamma(x, 5, 5/4)
 genfun_long <- function(x) dgamma(x, 5, 5/6)
 
-svec <- c("Variant GI = Wild type GI = Assumed GI",
-          "Variant GI $<$ Wild type GI = Assumed GI",
+svec <- c("Variant GI $<$ Wild type GI = Assumed GI",
+          "Variant GI = Wild type GI = Assumed GI",
           "Variant GI $>$ Wild type GI = Assumed GI")
 
 slist <- list(
   list(genfun1=genfun_assumed,
-       genfun2=genfun_assumed,
+       genfun2=genfun_short,
        genfun3=genfun_assumed),
   list(genfun1=genfun_assumed,
-       genfun2=genfun_short,
+       genfun2=genfun_assumed,
        genfun3=genfun_assumed),
   list(genfun1=genfun_assumed,
        genfun2=genfun_long,
@@ -70,11 +70,11 @@ for (i in 1:length(svec)) {
       legend.title = element_blank()
     )
   
-  if (i != 1) {
+  if (i != 2) {
     g1 <- g1 + theme(legend.position="none")
   }
   
-  if (i!=1) {
+  if (i!=2) {
     g1 <- ggplot(rr) +
       geom_line(aes(tvec, Rt1, col="Wild Type", lty="Wild Type"), lwd=2) +
       geom_line(aes(tvec, Rt2, col="Variant", lty="Variant"), lwd=2) +
@@ -106,7 +106,7 @@ for (i in 1:length(svec)) {
     coord_fixed() +
     theme(
       panel.grid = element_blank(),
-      legend.position = c(0.3, 0.81),
+      legend.position = c(0.7, 0.81),
       legend.title = element_blank()
     )
   
@@ -145,7 +145,7 @@ for (i in 1:length(svec)) {
                angle=atan(theta)/pi*180, col="purple")
   }
   
-  if (i != 1) {
+  if (i != 2) {
     g3 <- g3 +
       geom_smooth(aes(Rtest1, Rtest2, col="Regression", lty="Regression"), method="lm", fullrange=TRUE, se=FALSE, lwd=2)
     

### 2021-07-07 control.R

@@ -158,7 +158,7 @@ g5 <- ggplot(intervention) +
   scale_x_log10("Generation interval ratio, $\\bar{G}_{\\mathrm{var}}/\\bar{G}_{\\mathrm{wt}}$",
                 breaks=c(2/3, 1, 3/2),
                 labels=c("2/3", 1, "3/2")) +
-  scale_y_log10("Strength", breaks=c(0.5, 1, 2, 4), limits=c(0.5, 4)) +
+  scale_y_log10("Epidemic strength, $\\mathcal{R}_{\\mathrm{var}}$", breaks=c(0.5, 1, 2, 4), limits=c(0.5, 4)) +
   scale_linetype_manual("", values=c(1, 2, 3),
                         breaks=c("Pre-intervention",  "Post-intervention (constant-speed)", "Post-intervention (constant-strength)")) +
   scale_color_manual("", values=(colorblind_pal()(3)),
@@ -179,7 +179,7 @@ g6 <- ggplot(intervention) +
   scale_x_log10("Generation interval ratio, $\\bar{G}_{\\mathrm{var}}/\\bar{G}_{\\mathrm{wt}}$",
                 breaks=c(2/3, 1, 3/2),
                 labels=c("2/3", 1, "3/2")) +
-  scale_y_continuous("Speed (1/days)") +
+  scale_y_continuous("Epidemic speed, $r_{\\mathrm{var}} (1/days)") +
   scale_linetype_manual("", values=1:3) +
   scale_color_manual("", values=(colorblind_pal()(3))) +
   ggtitle("F") +
@@ -190,7 +190,7 @@ g6 <- ggplot(intervention) +
   )
 
 
-gtot <- ggarrange(g1, g2, g3, g4, g5, g6, nrow=3, ncol=2)
+gtot <- ggarrange(g1, g2, g3, g4, g5, g6, nrow=3, ncol=2, draw=FALSE)
 
 tikz(file = "control.tex", width = 12, height = 10, standAlone = T)
 gtot

### 2021-07-08 newvariant.tex
more edits

@@
 
-\begin{figure}[!th]
-\includegraphics[width=\textwidth]{control.pdf}
+\begin{figure}[!pth]
+\begin{center}
+\includegraphics[width=0.95\textwidth]{control.pdf}
 \caption{
 \textbf{Effects of constant-strength and constant-speed interventions on the spread of a new variant with known speed \rv.}
-(A--B) Pre-intervention (black) and post-intervention (colored) kernel of the new variant under constant-strength (A) and -speed (B) interventions assuming that the variant has equal mean generation interval as the wild type.
-(C--D) Pre-intervention (black) and post-intervention kernel of the new variant assuming that the variant has longer generation intervals than the wild type.
+(A--D) Pre-intervention (black) and post-intervention (colored) kernel of the new variant under constant-strength (A,C) and -speed (B,D) interventions when the variant has equal mean generation interval as the wild type (A--B) and when the variant has longer generation intervals than the wild type (C--D).
 Vertical lines represent the mean generation interval.
-(E) Pre-intervention $\Rv$ (black) and post-intervention epidemic strength $\Ry{post}$ conditional on the mean generation interval of the variant (colored).
-(F) Pre-intervention $\rv$ (black) and post-intervention epidemic speed $\ry{post}$ conditional on the mean generation interval of the variant (colored).
-Strength and speed are calculated assuming $\rw=0\pday$, $\delta=0.1\pday$, $\rv=0.1\pday$, $\Gy{wt}=5\,\textrm{days}$, and $\kappa=1/5$ for pre-intervention conditions.
+(E) Pre-intervention $\Rv$ (black) and post-intervention epidemic strength $\Ry{post}$ (colored) conditional on the mean generation interval of the variant.
+(F) Pre-intervention $\rv$ (black) and post-intervention epidemic speed $\ry{post}$ (colored) conditional on the mean generation interval of the variant.
+Epidemic strength and speed are calculated assuming $\rw=0\pday$, $\delta=0.1\pday$, $\rv=0.1\pday$, $\Gy{wt}=5\,\textrm{days}$, and $\kappa=1/5$ for pre-intervention conditions.
 The constant-strength intervention assumes that the ratio $\Rv/\Ry{post} = \theta$ remains constant.
@@
 }
+\end{center}
 \label{fig:strengthspeed}
@@
 Neglecting potential differences in the mean generation intervals between the variant and the wild type can bias estimates of the relative strength from incidence data;
-these biases may be assessed by considering whether relative strength appears to vary systematically with underlying epidemiological dynamics.
+these biases may be assessed by considering whether estimates of relative strength appear to vary systematically with the direction of changes in the incidence of infections caused by the variant.
 Finally, differences in generation intervals can also lead to different conclusions about the effectiveness of interventions.
-% if the variant has longer generation intervals than the wild type, speed-like interventions, will be relatively more effective than naive estimates would suggest (and conversely, strength-like intervention will be relatively more effective if the variant has shorter generation intervals).
+If the variant has longer generation intervals than the wild type, speed-like interventions will be relatively more effective than naive estimates would suggest. 
+Conversely, strength-like intervention will be relatively more effective if the variant has shorter generation intervals.
 

### 2021-07-08 Rtbias.R

@@ -6,7 +6,7 @@ library(gridExtra)
 library(tikzDevice)
 source("renewal_det.R")
 
-## fix theta = 1.5
+## fix theta = 1.61
 ## fix kappa = 1/5
 ## fix assumed GI
 ## assume wild type = var but not necessarily the one we're assuming
@@ -50,14 +50,14 @@ for (i in 1:length(svec)) {
     })
   
   rr <- do.call(renewal_det, tmparg) %>%
-    filter(tvec > 15, tvec < 70)
+    filter(tvec < 70)
   
   g1 <- ggplot(rr) +
     geom_line(aes(tvec, Rt1, color="Wild type", lty="True"), lwd=2) +
     geom_line(aes(tvec, Rt2, color="Variant", lty="True"), lwd=2) +
     geom_line(aes(tvec, Rtest1, color="Wild type", lty="Estimated"), lwd=2) +
     geom_line(aes(tvec, Rtest2, color="Variant", lty="Estimated"), lwd=2) +
-    scale_x_continuous("Time (days)", expand=c(0, 0), limits=c(15, 70)) +
+    scale_x_continuous("Time (days)", expand=c(0, 0), limits=c(0, 70)) +
     scale_y_log10("Reproduction number, $\\mathcal{R}(t)$", limits=c(0.39, 8),
                   breaks=c(0.25, 0.5, 1, 2, 4, 8)) +
     scale_color_manual(values=c("red", "black")) +
@@ -66,7 +66,7 @@ for (i in 1:length(svec)) {
     coord_fixed() +
     theme(
       panel.grid = element_blank(),
-      legend.position = c(0.6, 0.81),
+      legend.position = c(0.7, 0.81),
       legend.title = element_blank()
     )
   
@@ -80,7 +80,7 @@ for (i in 1:length(svec)) {
       geom_line(aes(tvec, Rt2, col="Variant", lty="Variant"), lwd=2) +
       geom_line(aes(tvec, Rtest1, lty="Estimated"), col="black", lwd=2) +
       geom_line(aes(tvec, Rtest2, col="Estimated", lty="Estimated"), lwd=2) +
-      scale_x_continuous("Time (days)", expand=c(0, 0), limits=c(15, 70)) +
+      scale_x_continuous("Time (days)", expand=c(0, 0), limits=c(0, 70)) +
       scale_y_log10("Reproduction number, $\\mathcal{R}(t)$", limits=c(0.39, 8),
[…]

### 2021-07-08 Rtbias_smooth.R

@@ -16,16 +16,16 @@ genfun_assumed <- function(x) dgamma(x, 5, 5/5)
 genfun_short <- function(x) dgamma(x, 5, 5/4)
 genfun_long <- function(x) dgamma(x, 5, 5/6)
 
-svec <- c("Variant GI = Wild type GI = Assumed GI",
-          "Variant GI $<$ Wild type GI = Assumed GI",
+svec <- c("Variant GI $<$ Wild type GI = Assumed GI",
+          "Variant GI = Wild type GI = Assumed GI",
           "Variant GI $>$ Wild type GI = Assumed GI")
 
 slist <- list(
   list(genfun1=genfun_assumed,
-       genfun2=genfun_assumed,
+       genfun2=genfun_short,
        genfun3=genfun_assumed),
   list(genfun1=genfun_assumed,
-       genfun2=genfun_short,
+       genfun2=genfun_assumed,
        genfun3=genfun_assumed),
   list(genfun1=genfun_assumed,
        genfun2=genfun_long,
@@ -45,7 +45,7 @@ for (i in 1:length(svec)) {
     })
   
   rr <- do.call(renewal_det, tmparg) %>%
-    filter(tvec > 15, tvec < 70)
+    filter(tvec < 70)
   
   g1 <- ggplot(rr) +
     geom_line(aes(tvec, Rt1, color="Wild type", lty="True"), lwd=2) +
@@ -61,21 +61,21 @@ for (i in 1:length(svec)) {
     coord_fixed() +
     theme(
       panel.grid = element_blank(),
-      legend.position = c(0.6, 0.81),
+      legend.position = c(0.7, 0.81),
       legend.title = element_blank()
     )
   
-  if (i != 1) {
+  if (i != 2) {
     g1 <- g1 + theme(legend.position="none")
   }
   
-  if (i!=1) {
+  if (i!=2) {
     g1 <- ggplot(rr) +
       geom_line(aes(tvec, Rt1, col="Wild Type", lty="Wild Type"), lwd=2) +
       geom_line(aes(tvec, Rt2, col="Variant", lty="Variant"), lwd=2) +
       geom_line(aes(tvec, Rtest1, lty="Estimated"), col="black", lwd=2) +
       geom_line(aes(tvec, Rtest2, col="Estimated", lty="Estimated"), lwd=2) +
-      scale_x_continuous("Time (days)", expand=c(0, 0), limits=c(15, 70)) +
[…]

### 2021-07-08 control.R

@@ -179,7 +179,7 @@ g6 <- ggplot(intervention) +
   scale_x_log10("Generation interval ratio, $\\bar{G}_{\\mathrm{var}}/\\bar{G}_{\\mathrm{wt}}$",
                 breaks=c(2/3, 1, 3/2),
                 labels=c("2/3", 1, "3/2")) +
-  scale_y_continuous("Epidemic speed, $r_{\\mathrm{var}} (1/days)") +
+  scale_y_continuous("Epidemic speed, $r_{\\mathrm{var}}$ (1/days)") +
   scale_linetype_manual("", values=1:3) +
   scale_color_manual("", values=(colorblind_pal()(3))) +
   ggtitle("F") +

### 2021-07-08 newvariant.tex
lots of edits

@@
 
-The integral of the kernel $\RR_x(t) = \int K_x(t, \tau) \dtau$ is referred to as the instantaneous reproduction number, which is defined as secondary infections that would be caused by an individual infected at time $t$ if conditions were to remain the same \citep{fraser2007estimating}.
+The integral of the kernel is referred to as the instantaneous reproduction number:
+\begin{equation}
+\RR_x(t) = \int K_x(t, \tau) \dtau
+\label{eq:instR}
+\end{equation}
+which is defined as secondary infections that would be caused by an individual infected at time $t$ if conditions were to remain the same \citep{fraser2007estimating}.
 The instantaneous reproduction number is a particular kind of weighted average of infectiousness of previously infected individuals at time $t$---in particular, it is weighted by the total relative infectiousness at time $t$, rather than by the actual number of infected individuals present.
@@
 
+\begin{figure}[!th]
+\begin{center}
+\includegraphics[width=0.95\textwidth]{control_sim.pdf}
+\caption{
+\textbf{Simulations of constant-strength and constant-speed interventions.}
+Constant-strength (A--C) and constant-speed (D--F) interventions are simulated using the renewal equation (see Supplementary Materials for details of simulations).
+(A,D) Daily incidence of infections caused by the wild type ($\iw(t)$, gray) and the variant ($\iv(t)$, colored).
+(B,E) Instantaneous reproduction number of the variant $\Rv(t)$ calculated using \eref{instR}.
+(C,F) Proportion of incidence of infections caused by the variant ($\iv(t)/(\iw(t)+\iv(t))$). 
+Epidemic trajectories are calculated assuming $\rw=0\pday$, $\delta=0.1\pday$, $\rv=0.1\pday$, $\Gy{wt}=5\,\textrm{days}$, and $\kappa=1/5$ for pre-intervention conditions, and then introducing constant-strength and constant-speed interventions on day 30.
+The pre-intervention ratio between the mean generation interval $\Gy{var}/\Gy{wt}$ is assumed to equal 1.5 (longer), 1 (equal), and 1/1.5 (shorter) for three different scenarios.
+The constant-strength and constant-speed interventions are identically modeled as illustrated in \fref{strengthspeed}.
+}
+\end{center}
+\label{fig:control_sim}
+\end{figure}
+
+Finally, we synthesize our findings and illustrate the differences between constant-strength and constant-speed interventions using epidemic simulations using the renewal equation (\fref{control_sim}).
+
 \section{Discussion}

### 2021-07-08 Rtbias.R

@@ -107,7 +107,8 @@ for (i in 1:length(svec)) {
     theme(
       panel.grid = element_blank(),
       legend.position = c(0.7, 0.86),
-      legend.title = element_blank()
+      legend.title = element_blank(),
+      legend.background = element_rect(fill=NA)
     )
   
   if (i != 1) {

### 2021-07-08 control_sim.R

@@ -1,0 +1,243 @@
+library(dplyr)
+library(ggplot2); theme_set(theme_bw(base_size=14, base_family = "Times"))
+library(ggthemes)
+library(egg)
+library(tikzDevice)
+source("renewal_det.R")
+
+delta <- 0.1
+
+rw <- 0
+
+kappa <- 0.2
+
+Gw <- 5
+
+Gratio <- c(1/1.5, 1, 1.5)
+
+Rv_base <- ((1 + kappa * (rw + delta) * Gw)^(1/kappa))
+target <- 0.9
+
+tmax <- 80
+
+constant_strength <- Rv_base/target
+
+constant_speed <- optim(0.1,
+                        function(r) {
+                          (integrate(function(y) Rv_base * dgamma(y, shape=1/kappa, rate=1/kappa/Gw) * exp(-r * y),
+                                     0,
+                                     Inf)[[1]] - target)^2
+                        },
+                        lower=0,
+                        upper=0.2,
+                        method="Brent")[[1]]
+
+thetavec <- ((1 + kappa * (rw + delta) * Gw*Gratio)^(1/kappa))
+
+genfun_short <- function(x) dgamma(x, 5, 5/(Gw*Gratio[1]))
+genfun_medium <- function(x) dgamma(x, 5, 5/(Gw*Gratio[2]))
+genfun_long <- function(x) dgamma(x, 5, 5/(Gw*Gratio[3]))
+
+Rfun <- function(t) {
+  if (t < 30) {
+    return(1) 
+  } else {
+    return(1/constant_strength)
+  }
+}
+  
+hfun <- function(t) {
+  if (t < 30) {
+    return(0) 
+  } else {
+    return(constant_speed)
+  }
+}
+
+simstrength1 <- renewal_det(
+  Rfun=Rfun,
+  theta=thetavec[1],
+  genfun1=genfun_medium,
+  genfun2=genfun_short,
+  tmax=tmax
+) %>%
+  mutate(
+    type="Shorter"
+  )
+
+simstrength2 <- renewal_det(
+  Rfun=Rfun,
+  theta=thetavec[2],
+  genfun1=genfun_medium,
+  genfun2=genfun_medium,
+  tmax=tmax
+) %>%
+  mutate(
+    type="Equal"
+  )
+
+simstrength3 <- renewal_det(
+  Rfun=Rfun,
+  theta=thetavec[3],
+  genfun1=genfun_medium,
[…]

### 2021-07-08 renewal_det.R

@@ -1,4 +1,5 @@
 renewal_det <- function(Rfun=function(t) 2.5,
+                        r1, r2,
                         theta=1.5,
                         N=40000,
                         dt=0.025,
@@ -16,17 +17,40 @@ renewal_det <- function(Rfun=function(t) 2.5,
   gen3 <- genfun3(0:genmax*dt+dt)
   gen3 <- gen3/sum(gen3)
   tvec <- seq(0, tmax, by=dt)
-  Ivec1 <- rep(0, tmax/dt)
-  Ivec2 <- rep(0, tmax/dt)
+  tveclong <- seq(-genmax*dt, tmax, by=dt)
+  Ivec1 <- rep(0, tmax/dt+genmax)
+  Ivec2 <- rep(0, tmax/dt+genmax)
   
-  Ivec1[1] <- I01
-  Ivec2[1] <- I02
+  if (missing(r1)) {
+    r1 <- optim(0, function(x) (1/sum(exp(-x*(0:genmax*dt+dt)) * gen1)-Rfun(0))^2,
+                lower=-0.4,
+                upper=0.4,
+                method="Brent")[[1]]
+    
+  }
+  
+  if (missing(r2)) {
+    r2 <- optim(0, function(x) (1/sum(exp(-x*(0:genmax*dt+dt)) * gen2)-theta*Rfun(0))^2,
+                lower=-0.4,
+                upper=0.4,
+                method="Brent")[[1]]
+    
+  }
+  
+  Ivec1[1:(genmax+1)] <- exp(r1*1:(genmax+1)*dt)
+  Ivec1[1:(genmax+1)] <- I01*Ivec1[1:(genmax+1)]/Ivec1[(genmax+1)]*dt
+  
+  Ivec2[1:(genmax+1)] <- exp(r2*1:(genmax+1)*dt)
+  Ivec2[1:(genmax+1)] <- I02*Ivec2[1:(genmax+1)]/Ivec2[(genmax+1)]*dt
+  
   Svec <- rep(0, tmax/dt)
-  Svec[1] <- N - I01 - I02
+  Svec[1] <- N - sum(Ivec1[1:(genmax+1)]) - sum(Ivec2[1:(genmax+1)])
   
   for (i in 2:length(tvec)) {
[…]

### 2021-07-08 newvariant.tex
one more para

@@
 Finally, we synthesize our findings and illustrate the differences between constant-strength and constant-speed interventions using epidemic simulations using the renewal equation (\fref{control_sim}).
+In particular, we consider a scenario in which a new variant is emerging ($\rv > 0\pday$ and $\Rw > 1$) while the incidence of infections caused by the wild type is constant ($\rw = 0\pday$ and $\Rw =1$) before interventions are reduced.
+The constant-strength intervention then reduces both $\Rw$ and $\Rv$ by a constant factor $\theta$ beginning day 30, 
+whereas the constnat-speed intervention isolates individuals infected with both the wild type and the variant at a constant hazard $\phi$ after day 30.
+If the variant has a longer mean generation interval, the consant-strength intervention fails to suppress the spread of the variant (\fref{control_sim}A) because a stronger intervention ($>\phi$) to reduce a higher pre-intervention reproduction number below 1 (\fref{control_sim}B).
 

### 2021-07-09 newvariant.tex
almost done

@@
 (B,E) Instantaneous reproduction number of the variant $\Rv(t)$ calculated using \eref{instR}.
-(C,F) Proportion of incidence of infections caused by the variant ($\iv(t)/(\iw(t)+\iv(t))$). 
+(C,F) Proportion of incidence of infections caused by the variant ($\iv(t)/(\iw(t)+\iv(t))$) on a logit scale. 
 Epidemic trajectories are calculated assuming $\rw=0\pday$, $\delta=0.1\pday$, $\rv=0.1\pday$, $\Gy{wt}=5\,\textrm{days}$, and $\kappa=1/5$ for pre-intervention conditions, and then introducing constant-strength and constant-speed interventions on day 30.
@@
 If the variant has a longer mean generation interval, the consant-strength intervention fails to suppress the spread of the variant (\fref{control_sim}A) because a stronger intervention ($>\phi$) to reduce a higher pre-intervention reproduction number below 1 (\fref{control_sim}B).
+The relative strength remains constnat under constant-strength intervention (\fref{control_sim}B), but the relative speed changes (\fref{control_sim}C) as discussed earlier in Section 4---in this case, the longer mean generation interval variant increases the relative speed when the intervention is introduced.
+
+In contrast, epidemic trajectories under the constant-speed intervention look identical regardless of the mean generation interval of the variant (\fref{control_sim}D). 
+As shown in \fref{strengthspeed}E, a constant-speed intervention leads to greater reduction in $\Rv(t)$ when the mean generation-interval of the variant is longer (\fref{control_sim}E).
+In this case, the relative speed of the variant also remains unaffected by the constant-speed intervention because the epidemic speed of the wild type $\rw$ and the variant $\rv$ are both reduced by equal amounts $\phi$, resulting in the post-intervention epidemic speed of $\rw-\phi$ and $\rv-\phi$, respectively.
 

### 2021-07-09 newvariant.tex
edits

@@
 
-Inferring the relative strength (i.e., the ratio of reproduction numbers, $\Rv/\Rw$) and relative speed (i.e., the difference between growth rates, $\rv-\rw$) of new SARS-CoV-2 variants compared to their wild types is critical to predicting and controlling the course of the current pandemic.
-Although the role of the generation interval (i.e., time between infection and transmission) in linking epidemic strength and speed has been previously well-established, many analyses of new SARS-CoV-2 variants overlook uncertainties in the generation-interval distributions as well as the possibility that their inferences of epidemic dynamics can depend on whether relative strength or speed are held constant.
-Here, we explore how differences in the generation-interval distributions affect the inference of relative strength and speed in light of recent evidence that the B.1.1.7 variant may have a longer infectious period (and therefore, a longer generation interval) than prior dominant lineages.
-We use simulations to show how neglecting these differences can lead to biases in estimates of relative strength in practice and to illustrate how such biases can be assessed.
-We also assess which of the two assumptions (i.e., constant relative strength or speed) are more appropriate and further discuss implications for control.
+Inferring the relative strength (i.e., the ratio of reproduction numbers, $\rho=\Rv/\Rw$) and relative speed (i.e., the difference between growth rates, $\delta=\rv-\rw$) of new SARS-CoV-2 variants compared to their wild types is critical to predicting and controlling the course of the current pandemic.
+We explore how differences in the generation-interval distributions affect the inference of relative strength $\rho$ and speed $\delta$ in light of recent evidence that the B.1.1.7 variant may have a longer infectious period (and therefore, a longer generation interval) than prior dominant lineages.
+Neglecting these differences gives biased estimates of $\rho$, which can vary systematically with changes in the incidence of infections caused by the variant.
+We further argue that holding $\rho$ constant is generally more appropriate in the context of SARS-CoV-2 and illustrate how $\rho$ and $\delta$ are affected by two idealized interventions.
+A constant-strength intervention, which reduces transmission rates by a constant amount, affects $\delta$ but not $\rho$---this is analogous to vaccination and social distancing.
+A constant-speed intervention, which isolates infected individuals at a constant hazard, affects $\rho$ but not $\delta$---this is analogous to contact tracing.
+Uncertainties in the generation-interval distribution determine the effectiveness of these interventions.
 
@@
 Instead, epidemic strength can be inferred from the observed epidemic speed using the generation-interval distribution, an approach popularized by \citep{wallinga2007generation}.
-The generation interval is defined as the time between infection and transmission and provides information about the individual-level transmission \citep{svensson2007note}.
+The generation interval is defined as the time between infection and transmission and provides information about the time scale of individual-level transmission \citep{svensson2007note}.
 The exact shape of the distribution depends on several factors---including the shape of latent and infectious periods \citep{lloyd2001realistic,wearing2005appropriate,roberts2007model} as well as more detailed life history of a disease \citep{huber2016quantitative}---and thus can be difficult to estimate.
@@
 Analyses of new variants have typically characterized \emph{relative} strength (i.e., the ratio of reproduction numbers $\Rv/\Rw$) and speed (i.e., the difference between growth rates $\rv-\rw$) of the variants.
-Many studies have tried to estimate the relative strength of variants from the observed relative speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021}.
+Many studies have tried to estimate the relative strength of variants from the observed relative speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021,campbell2021increased}.
 Some studies have instead assumed a value of the relative strength and tried to predict its relative speed \citep{davies2021estimated,di2021impact}.
 While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the disease and its control because two quantities are linked by the generation-interval distribution \citep{doi:10.1098/rspb.2020.1556}.
-For example, when epidemic strength is fixed, assuming longer generation intervals lead to a slower epidemic (lower $r$), making the epidemic look easier to control with speed-like interventions (e.g., contact tracing), which depend on hazard of isolation and therefore affects late transmission disproportionately.
-When epidemic speed is fixed, assuming longer generation intervals lead to a stronger epidemic (higher $\RR$), making the epidemic look harder to control with strength-like interventions (e.g., social distancing), which reduce the hazard of infectious contact and therefore affecting transmission by an equal amount regardless of age of infection.
-Therefore, it is critical to evaluate which assumption is more appropriate for inferring transmission advantages of new variants.
+For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals leads to a slower epidemic growth (lower $r$), making the epidemic look easier to control with speed-like interventions (e.g., contact tracing), which depend on isolation of infected individuals and therefore affect late transmission disproportionately---as shown in \cite{doi:10.1098/rspb.2020.1556}, the effective hazard of isolation must be greater than the epidemic speed $r$ to reach the $r=0$ threshold.
+However, since $\RR$ is fixed, the generation-interval distribution is expected to have negligible effects on the effectiveness of strength-like interventions (e.g., social distancing), which reduce the hazard of infectious contacts and therefore affect transmission by an equal amount regardless of age of infection.
+Conversely, when epidemic speed is fixed, assuming longer generation intervals lead to a stronger epidemic (higher $\RR$), making the epidemic look harder to control with strength-like interventions (but not necessarily with speed-like interventions).
 
-In addition, most studies have assumed that the previous dominant strain (wild type) and variants have identical generation-interval distributions, but recent evidence suggests that the B.1.1.7 variant may have a longer duration of infection: 13.3 days (90\% CI: 10.1--16.5) for the new variant and 8.2 days (90\% CI: 6.5--9.7) for the wild type \citep{kissler2021densely}.
+Most studies have assumed that the previous dominant strain (wild type) and variants have identical generation-interval distributions, but recent evidence suggests that the B.1.1.7 variant may have a longer duration of infection: 13.3 days (90\% CI: 10.1--16.5) for the new variant and 8.2 days (90\% CI: 6.5--9.7) for the wild type \citep{kissler2021densely}.
 Longer duration of infection suggests that the mean generation interval of B.1.1.7 is likely to be longer than that of the wild type \citep{lloyd2001realistic,wearing2005appropriate,roberts2007model}.
@@
 Here, we explore how different assumptions affect estimates of the relative strength (the ratio $\rho=\Rv/\Rw$) and relative speed (the difference $\delta=\rv-\rw$) for a new variant, such as B.1.1.7.
+In doing so, we assess 
 % We compare $\rho$--$\delta$ relationship between relative strength and relative speed under a wide range of assumptions about generation-interval distributions.

### 2021-07-11 newvariant.tex
stupid tex fix
edit
i think im mostly done

@@
 However, since $\RR$ is fixed, the generation-interval distribution is expected to have negligible effects on the effectiveness of strength-like interventions (e.g., social distancing), which reduce the hazard of infectious contacts and therefore affect transmission by an equal amount regardless of age of infection.
-Conversely, when epidemic speed is fixed, assuming longer generation intervals lead to a stronger epidemic (higher $\RR$), making the epidemic look harder to control with strength-like interventions (but not necessarily with speed-like interventions).
+Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals lead to a stronger epidemic (higher $\RR$), making the epidemic look harder to control with strength-like interventions (but not necessarily with speed-like interventions).
 
@@
 Here, we explore how different assumptions affect estimates of the relative strength (the ratio $\rho=\Rv/\Rw$) and relative speed (the difference $\delta=\rv-\rw$) for a new variant, such as B.1.1.7.
-In doing so, we assess 
 % We compare $\rho$--$\delta$ relationship between relative strength and relative speed under a wide range of assumptions about generation-interval distributions.
@@
 We focus on characterizing the incidence of infection, which is directly related to $r$ and $\RR$.
-In practice, observed case reports are subject to reporting delays, which must be taken into account in order to correctly infer $r$ or $\RR$ \citep{goldstein2009reconstructing,gostic2020practical}.
+In practice, observed case reports are subject to reporting delays, which must be taken into account in order to correctly infer $r$ and $\RR$ \citep{goldstein2009reconstructing,gostic2020practical}.
 
@@
 Like the instantaneous reproduction number, the instantaneous generation-interval distribution describes when an individual infected at time $t$ will transmit infections if conditions were to remain the same.
-Both the reproduction number and the generation-interval distribution depend on many factors, including intrinsic infectiousness of an infected individual, non-pharmaceutical interventions, awareness-driven behavior, and population-level susceptibility \citep{fraser2007estimating}.
+Both the instantaneous reproduction number and the generation-interval distribution depend on many factors, including intrinsic infectiousness of an infected individual, non-pharmaceutical interventions, awareness-driven behavior, and population-level susceptibility \citep{fraser2007estimating}.
 
-Under strength-like changes that reduce the hazard of transmission by a constant amount (e.g., susceptible depletion and decreased transmission rates), the instantaneous generation-interval distribution does not across change time \citep{fraser2007estimating}.
+Under strength-based changes that reduce the hazard of transmission (e.g., susceptible depletion and decreased transmission rates), the instantaneous generation-interval distribution does not across change time \citep{fraser2007estimating}.
 In this case, the instantaneous generation-interval distribution is also often referred to as the intrinsic generation-interval distribution \citep{champredon2015intrinsic,champredon2018two,gostic2020practical,park2020time}---
@@
 The instantaneous generation-interval distribution is also different from the realized generation-interval distribution, which measures time between actual infection events (see \citep{champredon2015intrinsic}).
-Previous studied have noted, in many contexts, that the realized generation intervals can contract due to susceptible depletion---a special case of strength-like changes---due to decreases in transmission \citep{kenah2008generation,nishiura2010time,champredon2015intrinsic}, even though the instantaneous generation-interval distribution remains unchanged.
-While the instantaneous generation-interval distribution can change under speed-like changes that disproportionately affect late transmissions (see \cite{fraser2007estimating} and Section 6 for discussion), assuming a time-invariant instantaneous generation-interval distribution is often appropriate in the context of SARS-CoV-2, given that the dynamics of its spread have been primarily driven by strength-like interventions, such as social distancing measures \citep{flaxman2020Rt} and vaccination \citep{moore2021vaccination}.
-Therefore, we neglect changes in the intrinsic generation-interval distribution over time for now and return to this idea in Section 6.
+Previous studied have noted, in many contexts, that the realized generation intervals can contract due to susceptible depletion---a special case of strength-based changes---due to decreases in transmission over time \citep{kenah2008generation,nishiura2010time,champredon2015intrinsic}, even though the instantaneous generation-interval distribution remains unchanged in this scenario.
+While the instantaneous generation-interval distribution can change under speed-based changes that disproportionately affect late transmissions (see \cite{fraser2007estimating} and Section 6--7 for detailed discussions), assuming a time-invariant instantaneous generation-interval distribution is often appropriate in the context of SARS-CoV-2, given that the dynamics of its spread have been primarily driven by strength-based changes, such as social distancing measures \citep{flaxman2020Rt} and vaccination \citep{moore2021vaccination}.
+Indeed, most dynamical models of SARS-CoV-2 infections have solely relied on strength-based changes, either implicitly or explicitly assuming a time-invariant intrinsic generation-interval distribution.
+Therefore, we neglect changes in the intrinsic generation-interval distribution over time for now and focus on the impact of strength-based changes on the inference of dynamics of new SARS-CoV-2 variants.
+We revisit these ideas in Section 6 and compare the effects of speed-based interventions with those of strength-based interventions.
 
@@
 Here, we use the term ``epidemiological conditions'' to broadly refer to all factors that affect transmission---mathematically, they are captured by the kernel $K(t, \tau)$.
-In the context of SARS-CoV-2, we are essentially assuming that the changes in susceptible pool, behavior, and contact rates are negligible over a short period of time.
+In the context of SARS-CoV-2 infections, we are essentially assuming that the changes in susceptible pool, behavior, and contact rates are negligible over a short period of time.
 Then, the incidence of each strain grows (or decays) exponentially at rate $r_x$, satisfying the Euler-Lotka equation \citep{wallinga2007generation}:
@@
 \end{equation}
-% The Gamma assumption includes as a special case models that assume exponentially distributed generation intervals (when $\kappa=1$), corresponding to the SIR model \citep{anderson1991infectious}; 
+% The 
 Various Gamma assumptions have been widely used in epidemic modeling, including for models of SARS-CoV-2 \citep{doi:10.1098/rsif.2020.0144}.
+Although the Gamma assumption includes as a special case models that assume exponentially distributed generation intervals (when $\kappa=1$), corresponding to the SIR model \citep{anderson1991infectious}, it is different from the generation-interval distribution we get when infectious periods are Gamma distributed, another standard assumption in epidemic modeling (see \cite{roberts2007model} for discussion).
+
 We use this framework to investigate how inferences about strength and speed of the variant depend on our assumptions about the underlying generation-interval distributions.
@@
 \end{align}
-where the logistic growth rate $\delta = \rv - \rw$ (i.e., the slope of the linear regression at the logit scale) corresponds to the relative speed of the epidemic.
-When more than two strains are co-circulating, the growth rate at a logit scale is expected to change over time and does not correspond to the relative speed even when all strains exhibit exponential changes (see Supplementary Text);
+where the logistic growth rate $\delta = \rv - \rw$ (i.e., the slope of the linear regression of $p(t)$ at the logit scale) corresponds to the relative speed of the epidemic.
+When more than two strains are co-circulating, the growth rate at a logit scale is expected to change over time and does not necessarily correspond to the relative speed even when all strains exhibit exponential changes (see Supplementary Text);
 instead, as shown in \citep{campbell2021increased}, a multinomial logistic regression model is required to infer relative speed.
@@
 Based on the observed relative growth rate of B.1.1.7 in the UK, we start by taking the relative speed of the variant to be $\delta = 0.1\pday$  \citep{davies2021estimated}; the mean generation interval of the wild type to be $\Gw = 5\days$ \citep{ferretti2020quantifying}; and the squared coefficient of variation of generation intervals to be $\kappa=0.2$ \citep{ferretti2020quantifying} for both the variant and the wild type.
-% As noted above, we assume that the variant and the wild type have equal $\kappa$ throughout, and only consider differences in the mean.
+As noted above, we assume that the variant and the wild type have equal $\kappa$ throughout, and only consider differences in the mean.
 We evaluate the estimates of relative strength $\rho$ across a wide range of $\kappa$ from 0 (fixed-length generation intervals) to 1 (exponential distribution).
@@
 
-We thus investigate how $\delta$ is expected to change with \Rw if $\rho$ remains constant, and how this expectation changes with the ratio of the generation intervals. 
+We thus investigate how $\delta$ is expected to change with \Rw when $\rho$ remains constant, and how this expectation changes with the ratio of the generation intervals. 
 Once again, we rely on the Gamma assumption \citep{park2019practical} to find the relative speed $\delta$ given the mean generation interval of the variant $\Gv$ and the wild type $\Gw$:
@@
 For example, even when there are no differences in the generation-interval distributions ($\Gv=\Gw$, in this case), changing $\Rw$ from 0.49 to 1.27 (and thus $\Rv$ from 0.79 to 2.04) yields the relative speed $\delta$ between 0.08--0.11 for $\kappa=0.2$ and 0.06--0.14 for $\kappa=0.5$.
-Therefore, characterizing the spread of variants assuming constant relative speed (by fitting a standard logistic growth curve; e.g., \citep{campbell2021increased,switzerland2021variant}) should be used with care.
+Therefore, characterizing the spread of variants assuming constant relative speed (e.g., by fitting a standard logistic growth curve \citep{switzerland2021variant}) should be used with care.
 
@@
 The instantaneous reproduction number is defined as the average number of secondary infections that would be caused by an individual infected at time $t$ if conditions were to remain the same \citep{fraser2007estimating};
-therefore, the instantaneous reproduction number only depends on transmission at time $t$ and is most appropriate for real-time evaluation of changes in trasmission, as argued by \citep{gostic2020practical}.
+therefore, the instantaneous reproduction number only depends on transmission at time $t$ and is most appropriate for real-time evaluation of changes in transmission, as argued by \citep{gostic2020practical}.
 The estimation of the instantaneous reproduction number was popularized by \cite{cori2013new} and has been widely adopted in epidemiological analyses of SARS-CoV-2 \citep{abbott2020estimating,knight2020estimating,flaxman2020Rt,brauner2021inferring,li2021temporal}.
@@
 Here, we choose to focus here on the underlying dynamical mechanisms that may affect inference and assume that the incidence of infection is exactly known.
-Assuming that the generation-interval distribution remains constant, the instantaneous reproduction numbers of the new variant and of the wild type can be estimated from their corresponding incidence curves---an approach popularized by \cite{cori2013new}:
+Assuming that the instantaneous generation-interval distribution remains constant, the instantaneous reproduction numbers of the new variant and of the wild type can be estimated from their corresponding incidence curves---an approach popularized by \cite{cori2013new}:
 \begin{equation}
@@
 \end{equation}
-Under strength-like intervention measures that reduce transmission rates by a constant amount, we expect ratios between reproduction numbers to remain constant and correspond to the true relative strength: $\Rv(t)/\Rw(t) = \rho$.
+Under strength-like intervention measures that reduce transmission rates of both strains by a constant amount, we expect ratios between reproduction numbers to remain constant and correspond to the true relative strength: $\Rv(t)/\Rw(t) = \rho$ \citep{leung2017monitoring,leung2020empirical,di2021impact,leung2021early}.
 However, if the assumed generation-interval distribution $\hat{g}(\tau)$ is different from the true distribution, then the ratio between the estimated reproduction numbers $\hat{\rho}(t) = \hat{\RR}_{\textrm{var}}(t)/\hat{\RR}_{\textrm{wt}}(t)$ may change, even if the true ratio does not.
@@
 }
+\label{fig:Rtbias}
 \end{center}
-\label{fig:Rtbias}
 \end{figure}
@@
 }
+\label{fig:strengthspeed}
 \end{center}
-\label{fig:strengthspeed}
 \end{figure}
@@
 }
+\label{fig:control_sim}
 \end{center}
-\label{fig:control_sim}
 \end{figure}
 
-Finally, we synthesize our findings and illustrate the differences between constant-strength and constant-speed interventions using epidemic simulations using the renewal equation (\fref{control_sim}).
-In particular, we consider a scenario in which a new variant is emerging ($\rv > 0\pday$ and $\Rw > 1$) while the incidence of infections caused by the wild type is constant ($\rw = 0\pday$ and $\Rw =1$) before interventions are reduced.
+Finally, we synthesize our findings and illustrate the differences between constant-strength and constant-speed interventions using epidemic simulations (\fref{control_sim}).
+As before, we consider a scenario in which a new variant is emerging ($\rv > 0\pday$ and $\Rw > 1$) while the incidence of infections caused by the wild type is constant ($\rw = 0\pday$ and $\Rw =1$) before interventions are reduced.
 The constant-strength intervention then reduces both $\Rw$ and $\Rv$ by a constant factor $\theta$ beginning day 30, 
-whereas the constnat-speed intervention isolates individuals infected with both the wild type and the variant at a constant hazard $\phi$ after day 30.
-If the variant has a longer mean generation interval, the consant-strength intervention fails to suppress the spread of the variant (\fref{control_sim}A) because a stronger intervention ($>\phi$) to reduce a higher pre-intervention reproduction number below 1 (\fref{control_sim}B).
-The relative strength remains constnat under constant-strength intervention (\fref{control_sim}B), but the relative speed changes (\fref{control_sim}C) as discussed earlier in Section 4---in this case, the longer mean generation interval variant increases the relative speed when the intervention is introduced.
+whereas the constant-speed intervention isolates individuals infected with both the wild type and the variant at a constant hazard $\phi$ after day 30.
+If the variant has a longer mean generation interval, the constant-strength intervention fails to suppress the spread of the variant (\fref{control_sim}A) because a stronger intervention ($>\phi$) to reduce a higher pre-intervention reproduction number below 1 (\fref{control_sim}B).
+The relative strength remains constant under constant-strength intervention (\fref{control_sim}B), but the relative speed changes (\fref{control_sim}C) as discussed earlier in Section 4---in this case, the longer mean generation interval variant increases the relative speed when the intervention is introduced.
 
@@
 however, this assumption only holds under strength-like interventions, which are insensitive to time since infection, but not under speed-like interventions. 
-As we demonstrated, if the variant and the wild type have different generation intervals, speed-like interventions such as contact tracing can affect them differently, causing the relative strength to change over time.
+As we demonstrated, if the variant and the wild type have different generation intervals, speed-like interventions such as contact tracing can affect them differently, causing the relative strength to change over time (but not necessarily the relative speed).
 
-% SWP: in case you're wondering why this is commented out, I have moved this to section 5.
-% We also assumed that the incidence of infection caused by the variant and the wild type is known.
-% However, estimating $\RR(t)$ from real data is often associated with a variety of practical challenges, including delays between infection and reporting and changes in testing patterns \citep{gostic2020practical}.
-% We have chosen to focus here on the underlying dynamical mechanisms that may affect inference.
-% Future studies may consider the development of methods based on mechanism outlined here to assess potential biases in their estimates of relative strength.
-
-% In moving from mechanism to improved estimates, we argue that both perspectives (i.e., whether we hold relative strength or speed constant) are useful in understanding the dynamics of the new variant.
-% Early in the spread of an emerging variant, it is likely more convenient to fix the relative speed, given that speed can be directly observed.
-% As more information about the transmission and immunity profiles of the new variant becomes available, we advise instead fixing the relative strength and inferring the speed, as this assumption better matches biological mechanisms for the variants' higher strength (e.g., higher rates of transmission and immune evasion).
-% Researchers should be mindful about which quantity they hold constant and how their conclusions follow from their assumptions \citep{doi:10.1098/rspb.2020.1556}.
+We considered two idealized interventions of constant strength and speed, but epidemiological impacts of real interventions are likely more complex.
+Interventions that reduce contact rates, such as social distancing measures and vaccination, are generally strength-like, but they can also elicit awareness-based behavioral responses that are speed-like---for example, infected individuals may self-isolate faster after symptom onset.
+On the other hand, the hazard of isolation-based interventions (and thus, speed-like), such as contact tracing, can vary over the course of infection depending on delays in tracing infected individuals \citep{fraser2004factors,ferretti2020quantifying,scarabel2021renewal}.
+Their effectiveness can further depend on the ``coverage'' of isolation, which is strength-like---for example, there may be a group of individuals that cannot be isolated by certain interventions due to asymptomaticity or the lack of participation in the contact tracing program.
+However, analyses of SARS-CoV-2 dynamics have primarily relied on the constant-strength framework \citep{gostic2020practical,unwin2020state,brauner2021inferring}, even when modeling speed-like interventions such as self-isolation \citep{flaxman2020Rt,brett2020transmission}.
+While the constant-strength framework provides a convenient tool for modeling and understanding epidemic dynamics, we encourage researchers to consider both strength- and speed-based perspectives as they can lead to different conclusions.
 
@@
 Early in the spread of an emerging variant, it is likely more convenient to fix the relative speed, given that speed can be directly observed.
-However, epidemiological conditions \textit{will} change over time in response to spreading new variants, causing relative speed to change and invalidating this assumption.
+However, epidemiological conditions \textit{will} change over time in response to spreading new variants---in the context of SARS-CoV-2 infections, most of these responses will be strength-like, causing relative speed to change and invalidating this assumption.
 As more information about the transmission and immunity profiles of the new variant becomes available, we advise instead fixing the relative strength and inferring the speed, as this assumption better matches biological mechanisms for the variants' higher strength (e.g., higher rates of transmission and immune evasion).
+However, as interventions can often have both strength- and speed-like aspects, both the relative strength and speed can vary at the same time.
 Second, the absolute strength and speed should not be neglected in favor of relative values.

### 2021-07-12 response.tex
lots of edits
i think im mostly done
done

@@
 \usepackage{lmodern}
-\usepackage{amssymb,amsmath}
+
+\usepackage{amsthm}
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{amsfonts}
 
@@
 
-\newcommand{\rR}{\mbox{$r$--$\cal R$}}
-\newcommand{\RR}{\ensuremath{{\cal R}}}
-\newcommand{\RRhat}{\ensuremath{{\hat \cal R}}}
-\newcommand{\Rx}[1]{\ensuremath{{\cal R}_{#1}}} 
-\newcommand{\Ro}{\ensuremath{{\mathcal R}_{0}}\xspace}
-\newcommand{\Rs}{\Rx{\mathrm{s}}}
-\newcommand{\Rpool}{\ensuremath{{\mathcal R}_{\textrm{\tiny{pool}}}}\xspace}
-\newcommand{\Reff}{\Rx{\mathit{eff}}}
-\newcommand{\Tc}{\ensuremath{C}}
+%% Consistent, changeable style for subscripts
+\newcommand{\vvvar}{\mathrm{var}}
+\newcommand{\wwwt}{\mathrm{wt}}
+
+\newcommand{\rx}[1]{\ensuremath{{r}_{#1}}\xspace} 
+\newcommand{\ry}[1]{\rx{\mathrm{#1}}} 
+\newcommand{\rw}{\rx{\wwwt}}
+\newcommand{\rv}{\rx{\vvvar}}
+
+\newcommand{\Rx}[1]{\ensuremath{{\mathcal R}_{#1}}\xspace} 
+\newcommand{\Ry}[1]{\Rx{\mathrm{#1}}}
+\newcommand{\Ro}{\Rx{0}}
+\newcommand{\RR}{\ensuremath{{\mathcal R}}\xspace}
+\newcommand{\Rw}{\Rx{\wwwt}}
+\newcommand{\Rv}{\Rx{\vvvar}}
 
@@
 Thank you for the chance to revise and resubmit our manuscript. 
-We have made major revisions to our manuscript to address the reviewers' comments.
+We have made revisions to our manuscript to address the reviewers' comments.
+Overall, we have tried to provide a more thorough review of the literature and highlight the novelty of our work while acknowledging its limitations.
+We have also added a new figure as suggested by Reviewer 3.
+Below please find our detailed responses to reviewers.
 
@@
 
-Thank you.
+Thank you for the review.
+We have tried to provide a more thorough review of previous works.
+As suggested by Reviewer 3, many of these concepts have been discussed within the epidemiology community with the recent emergence of new SARS-CoV-2 variants but have not been formally discussed.
+We believe that the addition of a new simulations (Figure 5) and the revised text better highlight the novelty of our work.
 
@@
 
-Thank you for the suggestions.
+\swp{JD, we need a new title.}
+Thank you for the suggestions. We have changed the title and the abstract to highlight the novelty of our work.
 
@@
 
-Done. We do not review gamma chain trick in depth because we're not doing it.
+Thank you for the suggestions. We now provide a review of this topic in the introduction along with a more detailed justification of the assumptions of our work:
+
+``Analyses of new variants have typically characterized \emph{relative} strength (i.e., the ratio of reproduction numbers $\Rv/\Rw$) and speed (i.e., the difference between growth rates $\rv-\rw$) of the variants.
+Many studies have tried to estimate the relative strength of variants from the observed relative speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021,campbell2021increased}.
+Some studies have instead assumed a value of the relative strength and tried to predict its relative speed \citep{davies2021estimated,di2021impact}.
+While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the disease and its control because two quantities are linked by the generation-interval distribution \citep{doi:10.1098/rspb.2020.1556}.
+For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals leads to a slower epidemic growth (lower $r$), making the epidemic look easier to control with speed-like interventions (e.g., contact tracing), which depend on isolation of infected individuals and therefore affect late transmission disproportionately---as shown in \cite{doi:10.1098/rspb.2020.1556}, the effective hazard of isolation must be greater than the epidemic speed $r$ to reach the $r=0$ threshold.
+However, since $\RR$ is fixed, the generation-interval distribution is expected to have negligible effects on the effectiveness of strength-like interventions (e.g., social distancing), which reduce the hazard of infectious contacts and therefore affect transmission by an equal amount regardless of age of infection.
+Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals lead to a stronger epidemic (higher $\RR$), making the epidemic look harder to control with strength-like interventions (but not necessarily with speed-like interventions).''
+
+We note that assuming a Gamma-distributed generation intervals is different from the standard Gamma-chain trick. We note this later in Secion 2:
+
+``Various Gamma assumptions have been widely used in epidemic modeling, including for models of SARS-CoV-2 \citep{doi:10.1098/rsif.2020.0144}.
+Although the Gamma assumption includes as a special case models that assume exponentially distributed generation intervals (when $\kappa=1$), corresponding to the SIR model \citep{anderson1991infectious}, it is different from the generation-interval distribution we get when infectious periods are Gamma distributed, another standard assumption in epidemic modeling (see \cite{roberts2007model} for discussion).''
 
@@
 
-Section 4 done. Section 5 done.
+Thank you for the suggestions. We have added references in Section 4 and a paragraph in Section 5 reviewing previous work on estimating R from incidence data.
 
@@
 
-Done.
+Thank you for the suggestions.
+We have tried to clarify what we mean by the generation interval.
+In particular, often-menetioned time contractions of generation intervals apply to the forward realized generation interval, which measure time between actual infection events, rather than the intrinsic generation interval, which measure the intrinsic infectiousness of infected individuals.
+Under strength-like interventions, the intrinsic generation interval does not vary across time, even though the realized generation interval contracts.
+These distinctions are clearly made in Champredon 2015, and we have tried to clarify these ideas again here (see Section 2 for a detailed discussion).
 
@@
 
-No.
+\swp{I think you would be able to answer this better.}
+Thank you for the suggestions. 
+We agree the strict definition of ``strength'' from physics has a unit of force, and the reproduction number is unitless.
+We now mention in the first paragraph that R is unitless: ``Epidemic ``strength'' is measured by the reproduction number $\RR$---a unitless quantity representing the average number of new infections caused by a typical infection.''
+
+However, we respectfully disagree with the comment that the ``speed''-``strength'' distinction leans into general quantitative illiteracy.
+As we have argued recently (https://doi.org/10.1098/rspb.2020.1556), there is a clear over-emphasis on the strength-based (R-based) perspective in the epidemiology and modeling community. 
+Although the terms  ``speed'' and ``strength'' may be imprecise as noted, we believe that they can help the community move towards thinking about both r- and R-based perspectives.
+We do not claim that either one is necessarily better, but rather they should be viewed as complementary approaches (since both r and R give equivalent threshold conditions).
+In some cases, however, there are clear advantanges in relying on either strength- or speed-based perspectives.
+For example, in the beginning of an epidemic, it is more appropriate to fix epidemic speed $r$ when evaluating uncertainties in the generation-interval distribution.
+On the other hand, as we argue here, it is more appropriate to fix relative strength $\rho$, rather than relative speed $\delta$, when conditions are changing due to interventions.
+We believe that our revisions, including the addition of a new figure as suggested by reviewer 3, have made these points clearer and is important to address the value of considering both perspectives to a wider audience.
+We thus decided to keep the terms ``speed'' and ``strength'' in this manuscript.
 
@@
 
-Done.
+Thank you for pointing this out. We now address this point:
+
+``When more than two strains are co-circulating, the growth rate at a logit scale is expected to change over time and does not necessarily correspond to the relative speed even when all strains exhibit exponential changes (see Supplementary Text);
+instead, as shown in \citep{campbell2021increased}, a multinomial logistic regression model is required to infer relative speed.''
 
@@
 
-Reworded:
+Thank you for the suggestion. We have reworded this section:
 
@@
 
-Thank you. We have tried to make the limitation more clear.
+Thank you for the review. We have tried to make the limitation and assumptions clearer throughout text.
 
@@
 
-This is a confusion between the intrinsic and realized. Also done.
+Thank you for pointing this out.
+We acknowledge that we are assuming mass action.
+However, a fixed generation-interval distribution does not rely on the assumption that the rate of depletion of susceptibles over time is negligible.
+This is due to an ongoing confusion between the intrinsic and realized generation interval as we explained above.
+For example, as shown in Champredon et al 2015 and 2018, the standard SEIR model can be equivalently written in terms of the renewal equation using a fixed \emph{intrinsic} generation-interval distribution, and we can still observe the contraction of the \emph{realized} generation interval for this model.
+We have added a paragraph in Section 2 that makes this distiction clearly:
+
+``Under strength-based changes that reduce the hazard of transmission (e.g., susceptible depletion and decreased transmission rates), the instantaneous generation-interval distribution does not across change time \citep{fraser2007estimating}.
+In this case, the instantaneous generation-interval distribution is also often referred to as the intrinsic generation-interval distribution \citep{champredon2015intrinsic,champredon2018two,gostic2020practical,park2020time}---
+for example, the standard SEIR model can be equivalently expressed as a renewal equation with time-invariant intrinsic generation-interval distribution $g(\tau)$ as shown in \citep{champredon2018equivalence}.
+The instantaneous generation-interval distribution is also different from the realized generation-interval distribution, which measures time between actual infection events (see \citep{champredon2015intrinsic}).
+Previous studied have noted, in many contexts, that the realized generation intervals can contract due to susceptible depletion---a special case of strength-based changes---due to decreases in transmission over time \citep{kenah2008generation,nishiura2010time,champredon2015intrinsic}, even though the instantaneous generation-interval distribution remains unchanged in this scenario.
+While the instantaneous generation-interval distribution can change under speed-based changes that disproportionately affect late transmissions (see \cite{fraser2007estimating} and Section 6--7 for detailed discussions), assuming a time-invariant instantaneous generation-interval distribution is often appropriate in the context of SARS-CoV-2, given that the dynamics of its spread have been primarily driven by strength-based changes, such as social distancing measures \citep{flaxman2020Rt} and vaccination \citep{moore2021vaccination}.
+Indeed, most dynamical models of SARS-CoV-2 infections have solely relied on strength-based changes, either implicitly or explicitly assuming a time-invariant intrinsic generation-interval distribution.
+Therefore, we neglect changes in the intrinsic generation-interval distribution over time for now and focus on the impact of strength-based changes on the inference of dynamics of new SARS-CoV-2 variants.
+We revisit these ideas in Section 6 and compare the effects of speed-based interventions with those of strength-based interventions.''
 
@@
 
-Done.
+Thank you for the suggestion. We now define them more clearly throughout text.
 
@@
 
-Thank you.
-
-\swp{Ref 9 is ``Infectious diseases of humans: dynamics and control.'' Do we capitalize this or no?}
+Done.
 
@@
 
-Thank you.
+Thank you for the review.
 
@@
 
-Thank you for pointing this out.
-We have removed 'speed-like' and 'strength-like' interventions from the abstract as these ideas can be confusing without further explanation (which is difficult given word limit).
-Instead, we have tried to make these ideas clearer throughout text, including in the introduction.
-We have also revised the abstract according to reviewer 1's comments.
+Thank you for pointing this out. We have tried to make these distinctions clearer throughout text.
 
@@
 
-Thank you for this suggestion.
+Thank you for this suggestion. We clarify this when it is mentioned for the first time:
+
+``Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$.
+Here, we use the term ``epidemiological conditions'' to broadly refer to all factors that affect transmission---mathematically, they are captured by the kernel $K(t, \tau)$.
+In the context of SARS-CoV-2 infections, we are essentially assuming that the changes in susceptible pool, behavior, and contact rates are negligible over a short period of time.''
 
@@
 
+Thank you for the suggestions. We have tried to make the figure caption clearer.
+
 \revtext{Really what I would like to see here (Figure 4) is some simulated epidemic curves and perhaps final epidemic sizes - I think this would go much further to communicate the intuition behind how the speed-based and strength-based interventions differ than what's currently presented.
@@
 
+Thank you for the suggestions. We have added Figure 5 and some additional text to explain the differences between strength- and speed-based interventions.
+
 \revtext{264: "epidemiological dynamics" is used again here - what specifically do you mean?}
 
+Thank you for the suggestions. We have rewritten this part:
+
+``these biases may be assessed by considering whether estimates of relative strength appear to vary systematically with the direction of changes in the incidence of infections caused by the variant''
+
 \revtext{265: "can also lead to different conclusions about the effectiveness of interventions" - what sort of different conclusions? In which circumstances?}
+
+We have added further explanations:
+
+``Finally, differences in generation intervals can also lead to different conclusions about the effectiveness of interventions.
+If the variant has longer generation intervals than the wild type, speed-like interventions will be relatively more effective than naive estimates would suggest. 
+Conversely, strength-like intervention will be relatively more effective if the variant has shorter generation intervals.''
 
@@
 
+Done.
+
 \bibliography{newvariant_abbv.bib}

### 2021-07-12 newvariant.tex
done

@@
 
-Neglecting the (relatively slow) rate of new mutations, the current incidence of infection $i_x(t)$ caused by each strain $x$---either the wild type (``wt'') or the variant (``var'')---can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary infections are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
+Neglecting the (relatively slow) rate of new mutations and assuming a mass-action epidemic, the current incidence of infection $i_x(t)$ caused by each strain $x$---either the wild type (``wt'') or the variant (``var'')---can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary infections are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
 \begin{equation}

### 2021-07-12 response.tex
moving

@@
 
-Thank you for the suggestions. We have added references in Section 4 and a paragraph in Section 5 reviewing previous work on estimating R from incidence data.
+Thank you for the suggestions. 
+We have added references in Section 4.
+We have also added a new paragraph in the beginning of Section 5 reviewing previous work on estimating R from incidence data:
+
+``Instead of inferring relative strength from speed, one can directly estimate time-varying reproduction numbers $\RR(t)$ of the variant and the wild type from incidence data and directly compare their ratios when the incidence of infection and the generation-interval distributions are known---such methods have been used in previous analyses of the B.1.1.7 strain by \cite{volz2021transmission}.
+Broadly, there are two types of reproduction numbers: case reproduction number and instantaneous reproduction number.
+The case reproduction number is defined as the average number of secondary infections caused by an individual infected at time $t$ and therefore depends on transmission after time $t$ \citep{wallinga2004different}.
+The instantaneous reproduction number is defined as the average number of secondary infections that would be caused by an individual infected at time $t$ if conditions were to remain the same \citep{fraser2007estimating};
+therefore, the instantaneous reproduction number only depends on transmission at time $t$ and is most appropriate for real-time evaluation of changes in transmission, as argued by \citep{gostic2020practical}.
+The estimation of the instantaneous reproduction number was popularized by \cite{cori2013new} and has been widely adopted in epidemiological analyses of SARS-CoV-2 \citep{abbott2020estimating,knight2020estimating,flaxman2020Rt,brauner2021inferring,li2021temporal}.''
 
@@
 Under strength-like interventions, the intrinsic generation interval does not vary across time, even though the realized generation interval contracts.
-These distinctions are clearly made in Champredon 2015, and we have tried to clarify these ideas again here (see Section 2 for a detailed discussion).
+These distinctions are clearly made in Champredon 2015, and we have tried to clarify these ideas again here in Section 2:
+
+``Under strength-based changes that reduce the hazard of transmission (e.g., susceptible depletion and decreased transmission rates), the instantaneous generation-interval distribution does not across change time \citep{fraser2007estimating}.
+In this case, the instantaneous generation-interval distribution is also often referred to as the intrinsic generation-interval distribution \citep{champredon2015intrinsic,champredon2018two,gostic2020practical,park2020time}---
+for example, the standard SEIR model can be equivalently expressed as a renewal equation with time-invariant intrinsic generation-interval distribution $g(\tau)$ as shown in \citep{champredon2018equivalence}.
+The instantaneous generation-interval distribution is also different from the realized generation-interval distribution, which measures time between actual infection events (see \citep{champredon2015intrinsic}).
+Previous studied have noted, in many contexts, that the realized generation intervals can contract due to susceptible depletion---a special case of strength-based changes---due to decreases in transmission over time \citep{kenah2008generation,nishiura2010time,champredon2015intrinsic}, even though the instantaneous generation-interval distribution remains unchanged in this scenario.
+While the instantaneous generation-interval distribution can change under speed-based changes that disproportionately affect late transmissions (see \cite{fraser2007estimating} and Section 6--7 for detailed discussions), assuming a time-invariant instantaneous generation-interval distribution is often appropriate in the context of SARS-CoV-2, given that the dynamics of its spread have been primarily driven by strength-based changes, such as social distancing measures \citep{flaxman2020Rt} and vaccination \citep{moore2021vaccination}.
+Indeed, most dynamical models of SARS-CoV-2 infections have solely relied on strength-based changes, either implicitly or explicitly assuming a time-invariant intrinsic generation-interval distribution.
+Therefore, we neglect changes in the intrinsic generation-interval distribution over time for now and focus on the impact of strength-based changes on the inference of dynamics of new SARS-CoV-2 variants.
+We revisit these ideas in Section 6 and compare the effects of speed-based interventions with those of strength-based interventions.''
 
@@
 For example, as shown in Champredon et al 2015 and 2018, the standard SEIR model can be equivalently written in terms of the renewal equation using a fixed \emph{intrinsic} generation-interval distribution, and we can still observe the contraction of the \emph{realized} generation interval for this model.
-We have added a paragraph in Section 2 that makes this distiction clearly:
+We have added a paragraph in Section 2 that makes this distiction clearly (please see response to reviewer 1 for the relevant text).
 
-``Under strength-based changes that reduce the hazard of transmission (e.g., susceptible depletion and decreased transmission rates), the instantaneous generation-interval distribution does not across change time \citep{fraser2007estimating}.
-In this case, the instantaneous generation-interval distribution is also often referred to as the intrinsic generation-interval distribution \citep{champredon2015intrinsic,champredon2018two,gostic2020practical,park2020time}---
-for example, the standard SEIR model can be equivalently expressed as a renewal equation with time-invariant intrinsic generation-interval distribution $g(\tau)$ as shown in \citep{champredon2018equivalence}.
-The instantaneous generation-interval distribution is also different from the realized generation-interval distribution, which measures time between actual infection events (see \citep{champredon2015intrinsic}).
-Previous studied have noted, in many contexts, that the realized generation intervals can contract due to susceptible depletion---a special case of strength-based changes---due to decreases in transmission over time \citep{kenah2008generation,nishiura2010time,champredon2015intrinsic}, even though the instantaneous generation-interval distribution remains unchanged in this scenario.
-While the instantaneous generation-interval distribution can change under speed-based changes that disproportionately affect late transmissions (see \cite{fraser2007estimating} and Section 6--7 for detailed discussions), assuming a time-invariant instantaneous generation-interval distribution is often appropriate in the context of SARS-CoV-2, given that the dynamics of its spread have been primarily driven by strength-based changes, such as social distancing measures \citep{flaxman2020Rt} and vaccination \citep{moore2021vaccination}.
-Indeed, most dynamical models of SARS-CoV-2 infections have solely relied on strength-based changes, either implicitly or explicitly assuming a time-invariant intrinsic generation-interval distribution.
-Therefore, we neglect changes in the intrinsic generation-interval distribution over time for now and focus on the impact of strength-based changes on the inference of dynamics of new SARS-CoV-2 variants.
-We revisit these ideas in Section 6 and compare the effects of speed-based interventions with those of strength-based interventions.''
-
-\revtext{ (page 2, lines 55–60) It would help to clarify what is meant by “strengthlike” and “speed-like” in interventions. From the examples, it appears that
+\revtext{(page 2, lines 55–60) It would help to clarify what is meant by “strengthlike” and “speed-like” in interventions. From the examples, it appears that
 speed-like interventions work by identifying potential infections quickly

### 2021-07-13 supp.tex
lots of edits
done for now, i think

@@
 \newcommand{\Rv}{\Rx{\vvvar}}
+\newcommand{\hRw}{\hat{\RR}_{\wwwt}}
+\newcommand{\hRv}{\hat{\RR}_{\vvvar}}
 
@@
 
-In order to model multiple co-circulating strain, we write $i_x(t)$ to represent the incidence of strain $x$ (for $x = 1, \dots, N$).
-When incidence is changing exponentially ($i_x(t) = i_x(t_0) \exp(r_x t)$), the proportion of strain $y$ in the population can be written as:
+In order to model $N$ co-circulating strains, we write $i_x(t)$ to represent the incidence of strain $x$ (for $x = 1, \dots, N$).
+When incidence is changing exponentially ($i_x(t) = i_x(t_0) \exp(r_x t)$), the proportion of infections caused by strain $y$ can be written as:
 \begin{equation}
@@
 Therefore, the growth rate measured at the logit scale is expected to change over time when multiple strains are co-circulating. 
-Instead, a multinomial logistic regression model is required to infer relative speed \citep{campbell2021increased}.
+Instead, a multinomial logistic regression model is required to infer relative speed (see Supplementary Materials in \citep{campbell2021increased} for derivation).
 
@@
 
-We use a two-strain renewal equation to simulate the spread of the variant and the wild type.
-Ignoring birth and death, the incidence of infection caused by the variant $\iv$ and the wild type $\iw$ assuming perfect cross immunity is given by:
+We use a two-strain renewal equation to simulate the spread of the variant and the wild type in Sections 5--6.
+Throughout this section, we assume that the intrinsic infectiousness of the variant is $\rho$ times higher than that of the wild type and that both strains have perfect cross immunity against each other.
+
+Ignoring birth and death, the incidence of infection caused by the variant $\iv$ and the wild type $\iw$ under constant-strength intervention is given by:
 \begin{align}
 \frac{dS}{dt} &= - \iv(t) - \iw(t)\\
-\iv(t) &= \Rv(t) \int_{0}^\infty \iv(t-\tau) g_{\mathrm{var}}(\tau) \dtau\\
-\iw(t) &= \Rw(t) \int_{0}^\infty \iw(t-\tau) g_{\mathrm{wt}}(\tau) \dtau
+\iv(t) &= \rho  \hRw(t) \frac{S(t)}{N} \int_{0}^\infty \iv(t-\tau) g_{\mathrm{var}}(\tau) \dtau\\
+\iw(t) &= \hRw(t) \frac{S(t)}{N}  \int_{0}^\infty \iw(t-\tau) g_{\mathrm{wt}}(\tau) \dtau
 \end{align}
-where $S$ represents the proportion of susceptible individuals.
-We discretize the model at the time scale of 0.025 days as described in \cite{park2021forward} and simulated the model with $\iv(0) = 0.001$ and $\iw(0) = 0.1$.
-To allow for smooth changes in $\Rw(t)$ around day 30 and 60, we let:
+where $S$ represents the number of susceptible individuals and $N$ represents the population size.
+For this model, the instantaneous reproduction numbers are defined as:
+\begin{align}
+\Rv(t) &= \rho  \hRw(t)S(t)/N\\
+\Rw(t) &=\hRw(t) S(t) /N
+\end{align}
+In this case, the ratio between the instantaneous reproduction numbers (i.e., the relative strength) remain constant over time:
+\begin{equation}
+\frac{\Rv(t)}{\Rw(t)} = \rho.
+\end{equation}
+
+Likewise, the incidence of infection caused by the variant $\iv$ and the wild type $\iw$ under constant-speed intervention is given by:
+\begin{align}
+\frac{dS}{dt} &= - \iv(t) - \iw(t)\\
+\iv(t) &= \rho  \Rw(0) \frac{S(t)}{N} \int_{0}^\infty \left[\iv(t-\tau) g_{\mathrm{var}}(\tau) \exp \left(- \int_0^\tau h(t-\sigma) \dsigma \right)\right] \dtau\\
+\iw(t) &= \Rw(0) \frac{S(t)}{N} \int_{0}^\infty \left[\iw(t-\tau) g_{\mathrm{wt}}(\tau) \exp \left(- \int_0^\tau h(t-\sigma) \dsigma \right)\right] \dtau
+\end{align}
+where $h$ represents the time-varying hazard of isolation.
+Following the definition in Equation~(2) of the main text, the instantaneous reproduction numbers are defined as:
+\begin{align}
+\Rv(t) &= \rho \Rw(0)  \frac{S(t)}{N} \int_{0}^\infty g_{\mathrm{var}}(\tau) \exp \left(- \int_0^\tau h(t-\sigma) \dsigma \right) \dtau \\
+\Rw(t) &= \Rw(0)  \frac{S(t)}{N} \int_{0}^\infty g_{\mathrm{wt}}(\tau) \exp \left(- \int_0^\tau h(t-\sigma) \dsigma \right) \dtau
+\end{align}
+In this case, the ratio between the instantaneous reproduction numbers (i.e., the relative strength) will not remain constant unless both the wild type and the variant have identical generation-interval distributions.
+
+In Section 5 (Figure 3 in the main text), we rely on the constant-strength model for simplicity.
+To allow for smooth decrease and increase in $\Rw(t)$ (and $\Rv(t)$) around day 30 and 60, we let:
 \begin{equation}
@@
 In this case, $\Rw(t)$ remains around 2 and slowly decreases to $0.4 = 2 \times (1-0.8)$ around day 30, and slowly increases back up to $1 = 0.4 \times (1 + 1.5)$ around day 60.
-Finally, we assume $\Rv(t) = \rho \Rw(t)$.
-From simulated incidence curves between day 15 and 70, we estimate the instantaneous reproduction number using \eref{rt}.
-We ignore incidence before day 15 to remove any potential transient effects.
+We also present the analogous simulation with sharp changes in $\RR$ in Figure S1 using:
+\begin{equation}
+\Rw(t) = \begin{cases}
+2 & t < 30\\
+0.4 & 30 \leq t < 60\\
+1 & 60 \leq t 
+\end{cases}
+\end{equation}
+
+In Section 6 (Figure 5 in the main text), we compare the impact of constant-strength and constant-speed models in response to an emerging variant.
+For simplicity, we only consider the introduction of interventions and not the lifting of interventions.
+In both scenarios, we assume that the infections caused by the wild type is stable at the introduction of interventions (i.e., $\RR=1$ and $r=0$).
+In the constant-strength scenario, we have: 
+\begin{equation}
+\Rw(t) = \begin{cases}
+1 & t < 30\\
+\theta & 30 \leq t
+\end{cases}.
+\end{equation}
+In the constant-speed scenario, we have: 
+\begin{align}
+\Rw(0) &= 1\\
+h(t) &= \begin{cases}
+0 & t < 30\\
+\phi & 30 \leq t
+\end{cases}.
+\end{align}
+Both $\theta$ and $\phi$ are chosen such that the post-intervention instantaneous reproduction number of the variant is equal to 0.9 when its mean generation interval is 5 days, as in Figure 4 in the main text.
+In both cases, we assume $\rho = 1.61$, which is the value we obtain for $\delta=0.1\pday$ \citep{davies2021estimated}, $\rw=0\pday$, $\Gw = \Gv = 5\,\textrm{days}$, and $\kappa = 1/5$ \citep{ferretti2020quantifying}.
+
+We discretize the model at the time scale of $\Delta t =0.025\ \textrm{days}$ as described in \cite{park2021forward}.
+The initial conditions are set such that infections caused by the variant and the wild type can exhibit exponential changes from $\iv(0) = 0.001$ and $\iw(0) = 0.1$ since day 0.
+The population size is set to $N=40,000$
 

### 2021-07-14 response.tex
Once through the response, and some related changes in the MS.

@@
 
+\jd{If we said Thank you a little less, it might mean a little more.}
+
 Thank you for the review.
-We have tried to provide a more thorough review of previous works.
-As suggested by Reviewer 3, many of these concepts have been discussed within the epidemiology community with the recent emergence of new SARS-CoV-2 variants but have not been formally discussed.
+We have tried to frame our results better, and to clarify both the novelty and importance of this study.
+We agree with Reviewer 3 that, while many of these concepts have been raised in the context of the recent emergence of new SARS-CoV-2 variants, careful treatment has been lacking. 
 We believe that the addition of a new simulations (Figure 5) and the revised text better highlight the novelty of our work.
@@
 
-\swp{JD, we need a new title.}
-Thank you for the suggestions. We have changed the title and the abstract to highlight the novelty of our work.
+\swp{JD, we need a new title.} \jd{I think it helps a lot to just not say ``distributions'' since we're mostly focused on changes in the mean.}
+
+\jd{Have we looked at the Miller paper?}
+
+Thank you for the suggestions. We have changed the title and the abstract to highlight the novelty of our work. See also below.
 
@@
 
-We note that assuming a Gamma-distributed generation intervals is different from the standard Gamma-chain trick. We note this later in Secion 2:
+Assuming Gamma-distributed generation intervals is different from the standard Gamma-chain trick. We [now?] explain this later in Secion 2:
 
@@
 
-Thank you for the suggestions. 
 We have added references in Section 4.
@@
 Under strength-like interventions, the intrinsic generation interval does not vary across time, even though the realized generation interval contracts.
-These distinctions are clearly made in Champredon 2015, and we have tried to clarify these ideas again here in Section 2:
+These distinctions are explained in Champredon 2015, and we have tried to clarify these ideas again here in Section 2:
 
@@
 
-\swp{I think you would be able to answer this better.}
-Thank you for the suggestions. 
-We agree the strict definition of ``strength'' from physics has a unit of force, and the reproduction number is unitless.
-We now mention in the first paragraph that R is unitless: ``Epidemic ``strength'' is measured by the reproduction number $\RR$---a unitless quantity representing the average number of new infections caused by a typical infection.''
+We are certainly not trying to lean in to innumeracy.
+We agree that ``strength'' and ``speed'' are not used in the classical physics sense -- instead, we try to be clear how we define them (and have tried to clarify further).
 
-However, we respectfully disagree with the comment that the ``speed''-``strength'' distinction leans into general quantitative illiteracy.
-As we have argued recently (https://doi.org/10.1098/rspb.2020.1556), there is a clear over-emphasis on the strength-based (R-based) perspective in the epidemiology and modeling community. 
-Although the terms  ``speed'' and ``strength'' may be imprecise as noted, we believe that they can help the community move towards thinking about both r- and R-based perspectives.
-We do not claim that either one is necessarily better, but rather they should be viewed as complementary approaches (since both r and R give equivalent threshold conditions).
-In some cases, however, there are clear advantanges in relying on either strength- or speed-based perspectives.
+We have argued (https://doi.org/10.1098/rspb.2020.1556) that large segments of the epidemiology and modeling community have an over-emphasis on R, and have tried to use the terms speed and strength to underline our contention that these are better seen as complementary perspectives (and to link them to complementary perspectives on control).
+As we argue here and elsewhere, we believe that sometimes one perspective ro another gives a clearer picture.
 For example, in the beginning of an epidemic, it is more appropriate to fix epidemic speed $r$ when evaluating uncertainties in the generation-interval distribution.
-On the other hand, as we argue here, it is more appropriate to fix relative strength $\rho$, rather than relative speed $\delta$, when conditions are changing due to interventions.
-We believe that our revisions, including the addition of a new figure as suggested by reviewer 3, have made these points clearer and is important to address the value of considering both perspectives to a wider audience.
+On the other hand, it is more often appropriate to fix relative strength $\rho$, rather than relative speed $\delta$, when conditions are changing due to interventions.
+We hope that our revisions, including the addition of a new figure (see response to reviewer 3), have made these points clearer.
 We thus decided to keep the terms ``speed'' and ``strength'' in this manuscript.
@@
 
-Thank you for pointing this out. We now cite Breda et al. We acknowledge that ``This framework ... generalizes compartmental model, such as the SEIR model''
+Thank you for pointing this out. We now cite Breda et al., and acknowledge that ``This framework ... generalizes compartmental model, such as the SEIR model''
 
@@
 
+\jd{We should deal separately with $\kappa$ and with the gamma-generation assumption.
 ``For simplicity we neglect differences in the squared coefficient of variation and assume $\kappa_{\mathrm{wt}} = \kappa_{\mathrm{var}} = \kappa$.
@@
 
-Thank you for the review. We have tried to make the limitation and assumptions clearer throughout text.
+Thank you for the review.
+We have tried to make the limitation and assumptions clearer throughout the text.
+\jd{Anything about the generality?
+I don't feel we rely much on mass-action or constant distributions.}
 
@@
 However, a fixed generation-interval distribution does not rely on the assumption that the rate of depletion of susceptibles over time is negligible.
+\jd{Can we say something about the rR link at any particular time? Doesn't that also work?}
 This is due to an ongoing confusion between the intrinsic and realized generation interval as we explained above.
 For example, as shown in Champredon et al 2015 and 2018, the standard SEIR model can be equivalently written in terms of the renewal equation using a fixed \emph{intrinsic} generation-interval distribution, and we can still observe the contraction of the \emph{realized} generation interval for this model.
-We have added a paragraph in Section 2 that makes this distiction clearly (please see response to reviewer 1 for the relevant text).
+We have added a paragraph in Section 2 to clarify this distinction (please see response to reviewer 1 for text).
 
@@
 
-Thank you for the suggestion. We now define them more clearly throughout text.
+Thank you for the suggestion. We now define them more clearly throughout the text. \jd{Presumably there's a particular place where we lay out the definitions.}
 
@@
 
-Thank you for this suggestion. We clarify this when it is mentioned for the first time:
+Thank you for this suggestion. We now clarify this when it is mentioned for the first time:
 
+\jd{This is the sort of text I was hinting at above about mass action.}
 ``Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$.

### 2021-07-14 newvariant.tex
Once through the response, and some related changes in the MS.

@@
 	\textbf\newline{
-	  Roles of generation-interval distributions in shaping relative epidemic strength, speed, and control of new SARS-CoV-2 variants
-	}
+	  The importance of generation intervals in investigating dynamics and control of new SARS-CoV-2 variants }
 }
@@
 Inferring the relative strength (i.e., the ratio of reproduction numbers, $\rho=\Rv/\Rw$) and relative speed (i.e., the difference between growth rates, $\delta=\rv-\rw$) of new SARS-CoV-2 variants compared to their wild types is critical to predicting and controlling the course of the current pandemic.
-We explore how differences in the generation-interval distributions affect the inference of relative strength $\rho$ and speed $\delta$ in light of recent evidence that the B.1.1.7 variant may have a longer infectious period (and therefore, a longer generation interval) than prior dominant lineages.
+We explore how differences in the generation-interval distributions affect the inference of relative strength $\rho$ and speed $\delta$ in light of recent evidence that the alpha \jd{suggesting that we use spelled-out greek names for variants, to reduce confusion with rendered greek letters for quantities} variant may have a longer infectious period (and therefore, a longer generation interval) than prior dominant lineages.
 Neglecting these differences gives biased estimates of $\rho$, which can vary systematically with changes in the incidence of infections caused by the variant.
@@
 The epidemic strength also determines the final size of an epidemic in a homogeneous population \citep{anderson1991infectious}.
-Epidemic ``speed'' is characterized by the growth rate $r$, which has a unit of $1/\mathrm{time}$ and describes how fast a disease spreads at the population level.
-Like epidemic strength, epidemic speed also determines conditions for disease elimination: $r=0$ is a threshold equivalent to $\RR=1$.
+Epidemic ``speed'' is characterized by the growth rate $r$, which has units of $1/\mathrm{time}$ and describes the exponential rate of disease spreads at the population level.
+Like epidemic strength, epidemic speed also determines conditions for disease elimination: $r=0$ is a threshold equivalent to $\RR=1$ under constant conditions.
 
 Epidemic speed can be typically estimated from incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}, but epidemic strength is difficult to measure directly.
-Instead, epidemic strength can be inferred from the observed epidemic speed using the generation-interval distribution, an approach popularized by \citep{wallinga2007generation}.
+Instead, epidemic strength is often inferred from the observed epidemic speed using the generation-interval distribution, an approach popularized by \citep{wallinga2007generation}.
 The generation interval is defined as the time between infection and transmission and provides information about the time scale of individual-level transmission \citep{svensson2007note}.
@@
 
-Analyses of new variants have typically characterized \emph{relative} strength (i.e., the ratio of reproduction numbers $\Rv/\Rw$) and speed (i.e., the difference between growth rates $\rv-\rw$) of the variants.
+Analyses of new variants have typically characterized \emph{relative} strength (i.e., the ratio of reproduction numbers $\Rv/\Rw$, here called $\rho$) and speed (i.e., the difference between growth rates $\rv-\rw$, here called $\delta$) of the variants.
 Many studies have tried to estimate the relative strength of variants from the observed relative speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021,campbell2021increased}.
 Some studies have instead assumed a value of the relative strength and tried to predict its relative speed \citep{davies2021estimated,di2021impact}.
+\jd{``instead'' seems wrong here, since we cite davies2021estimated in both lists. What's going on?}
 While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the disease and its control because two quantities are linked by the generation-interval distribution \citep{doi:10.1098/rspb.2020.1556}.
+
 For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals leads to a slower epidemic growth (lower $r$), making the epidemic look easier to control with speed-like interventions (e.g., contact tracing), which depend on isolation of infected individuals and therefore affect late transmission disproportionately---as shown in \cite{doi:10.1098/rspb.2020.1556}, the effective hazard of isolation must be greater than the epidemic speed $r$ to reach the $r=0$ threshold.
-However, since $\RR$ is fixed, the generation-interval distribution is expected to have negligible effects on the effectiveness of strength-like interventions (e.g., social distancing), which reduce the hazard of infectious contacts and therefore affect transmission by an equal amount regardless of age of infection.
-Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals lead to a stronger epidemic (higher $\RR$), making the epidemic look harder to control with strength-like interventions (but not necessarily with speed-like interventions).
+In this same case, however, the generation-interval distribution is expected to have negligible effects on the effectiveness of strength-like interventions (e.g., social distancing), which reduce the hazard of infectious contacts and therefore affect transmission by an equal amount regardless of age of infection.
+
+Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals leads to a higher estimate of epidemic strength ($\RR$), making the epidemic look harder to control with strength-like interventions (but not with speed-like interventions, whose effectiveness is fixed if speed is fixed).
 
@@
 
-Here, we explore how different assumptions affect estimates of the relative strength (the ratio $\rho=\Rv/\Rw$) and relative speed (the difference $\delta=\rv-\rw$) for a new variant, such as B.1.1.7.
+Here, we explore how different assumptions affect estimates of the relative strength ($\rho=\Rv/\Rw$) and relative speed ($\delta=\rv-\rw$) for a new variant, such as B.1.1.7 \jd{search and replace variant names if you agree}.
 % We compare $\rho$--$\delta$ relationship between relative strength and relative speed under a wide range of assumptions about generation-interval distributions.
@@
 \end{equation}
-This framework provides a flexible way of modeling disease dynamics and generalizes compartmental model, such as the SEIR model \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral,breda2012formulation, champredon2018equivalence}.
+This framework provides a flexible way of modeling disease dynamics and generalizes a wide range of compartmental models, including the SEIR model \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral,breda2012formulation, champredon2018equivalence}.
 
@@
 
+\jd{Where did -based come from? Do we like it?}
 Under strength-based changes that reduce the hazard of transmission (e.g., susceptible depletion and decreased transmission rates), the instantaneous generation-interval distribution does not across change time \citep{fraser2007estimating}.
@@
 The instantaneous generation-interval distribution is also different from the realized generation-interval distribution, which measures time between actual infection events (see \citep{champredon2015intrinsic}).
-Previous studied have noted, in many contexts, that the realized generation intervals can contract due to susceptible depletion---a special case of strength-based changes---due to decreases in transmission over time \citep{kenah2008generation,nishiura2010time,champredon2015intrinsic}, even though the instantaneous generation-interval distribution remains unchanged in this scenario.
-While the instantaneous generation-interval distribution can change under speed-based changes that disproportionately affect late transmissions (see \cite{fraser2007estimating} and Section 6--7 for detailed discussions), assuming a time-invariant instantaneous generation-interval distribution is often appropriate in the context of SARS-CoV-2, given that the dynamics of its spread have been primarily driven by strength-based changes, such as social distancing measures \citep{flaxman2020Rt} and vaccination \citep{moore2021vaccination}.
+Previous studies have noted, in many contexts, that the realized generation intervals can contract due to susceptible depletion---a special case of strength-based changes \citep{kenah2008generation,nishiura2010time,champredon2015intrinsic}--- even though the instantaneous generation-interval distribution remains unchanged in this scenario.
+While the instantaneous generation-interval distribution can change under speed-based changes (see \cite{fraser2007estimating} and Section 6--7 for detailed discussions), assuming a time-invariant instantaneous generation-interval distribution is often appropriate in the context of SARS-CoV-2, given that the dynamics of its spread have been primarily driven by strength-based changes, such as social distancing measures \citep{flaxman2020Rt} and vaccination \citep{moore2021vaccination}.
 Indeed, most dynamical models of SARS-CoV-2 infections have solely relied on strength-based changes, either implicitly or explicitly assuming a time-invariant intrinsic generation-interval distribution.
@@
 Here, we use the term ``epidemiological conditions'' to broadly refer to all factors that affect transmission---mathematically, they are captured by the kernel $K(t, \tau)$.
-In the context of SARS-CoV-2 infections, we are essentially assuming that the changes in susceptible pool, behavior, and contact rates are negligible over a short period of time.
+In the context of SARS-CoV-2 infections, we are essentially assuming that the changes in susceptible pool, behavior, and contact rates are usually small over a short period of time.
 Then, the incidence of each strain grows (or decays) exponentially at rate $r_x$, satisfying the Euler-Lotka equation \citep{wallinga2007generation}:
@@
 Instead of inferring relative strength from speed, one can directly estimate time-varying reproduction numbers $\RR(t)$ of the variant and the wild type from incidence data and directly compare their ratios when the incidence of infection and the generation-interval distributions are known---such methods have been used in previous analyses of the B.1.1.7 strain by \cite{volz2021transmission}.
-Broadly, there are two types of reproduction numbers: case reproduction number and instantaneous reproduction number.
+Broadly, there are two types of time-varying reproduction numbers: case reproduction number and instantaneous reproduction number.
 The case reproduction number is defined as the average number of secondary infections caused by an individual infected at time $t$ and therefore depends on transmission after time $t$ \citep{wallinga2004different}.
 The instantaneous reproduction number is defined as the average number of secondary infections that would be caused by an individual infected at time $t$ if conditions were to remain the same \citep{fraser2007estimating};
-therefore, the instantaneous reproduction number only depends on transmission at time $t$ and is most appropriate for real-time evaluation of changes in transmission, as argued by \citep{gostic2020practical}.
+therefore, the instantaneous reproduction number only depends on transmission at time $t$ and is most appropriate for real-time evaluation of changes in transmission \citep{gostic2020practical}.
 The estimation of the instantaneous reproduction number was popularized by \cite{cori2013new} and has been widely adopted in epidemiological analyses of SARS-CoV-2 \citep{abbott2020estimating,knight2020estimating,flaxman2020Rt,brauner2021inferring,li2021temporal}.
@@
 Observed case counts are sensitive to reporting delays \citep{goldstein2009reconstructing} and changes in case definitions \citep{tsang2020effect}, which in turn can affect estimates of the instantaneous reproduction number \citep{gostic2020practical}.
-Here, we choose to focus here on the underlying dynamical mechanisms that may affect inference and assume that the incidence of infection is exactly known.
+Here, we choose to focus on the underlying dynamical mechanisms that may affect inference and thus assume that the incidence of infection is exactly known.
 Assuming that the instantaneous generation-interval distribution remains constant, the instantaneous reproduction numbers of the new variant and of the wild type can be estimated from their corresponding incidence curves---an approach popularized by \cite{cori2013new}:
@@
 we do not claim that they are necessarily more accurate---
-while individual-level data from contact tracing can provide a more reliable source of information about the transmissibility and time scale of transmission in some cases, they can be also biased towards particular types of contacts that are easily identified such as household contacts, which could also affect the estimate of $\rho$.
+while individual-level data from contact tracing can provide a more reliable source of information about the transmissibility and time scale of transmission in some cases, they can be also biased towards particular types of contacts---for example, household contacts are probably more likely to be identified---which could also affect the estimate of $\rho$.
 Instead, this calculation simply highlights the importance of propagating uncertainty in generation-interval distributions in assessing the relative strength and speed of SARS-CoV-2 variants.

### 2021-07-15 response.tex
Once through

@@
 
-\jd{We should deal separately with $\kappa$ and with the gamma-generation assumption.
+\jd{We should deal separately with $\kappa$ and with the gamma-generation assumption.}
 ``For simplicity we neglect differences in the squared coefficient of variation and assume $\kappa_{\mathrm{wt}} = \kappa_{\mathrm{var}} = \kappa$.

### 2021-07-15 newvariant.tex
Once through

@@
 For simplicity we neglect differences in the squared coefficient of variation and assume $\kappa_{\mathrm{wt}} = \kappa_{\mathrm{var}} = \kappa$.
-While the exact shape of the distribution matters, we choose to focus on differences in the mean generation intervals to provide a qualitative understanding of how differences in generation-interval distributions between the wild type and the variant affects the relationship between relative strength and speed.
-In general, we expect  differences in the squared coefficient of variation to have qualitatively similar effects as differences in the mean.
-For example, a wider distribution allows for more early transmission, which in turn reduces epidemic strength given speed---this effect is qualitatively similar to that of a distribution with a shorter mean.
+Here we focus on differences in mean generation intervals, assuming that both the squared coefficient of variation, and distribution shape of the variant are otherwise the same as for the wild type.
+Changes in shape can be important and we do not investigate them here.
+We do not that, loosely, we expect a distribution with higher coefficient of variation to allow for more early transmission, and thus to have qualitatively similar effects to a distribution with a shorter mean in many cases.
 
@@
 \end{align}
-where the logistic growth rate $\delta = \rv - \rw$ (i.e., the slope of the linear regression of $p(t)$ at the logit scale) corresponds to the relative speed of the epidemic.
-When more than two strains are co-circulating, the growth rate at a logit scale is expected to change over time and does not necessarily correspond to the relative speed even when all strains exhibit exponential changes (see Supplementary Text);
-instead, as shown in \citep{campbell2021increased}, a multinomial logistic regression model is required to infer relative speed.
-Since we are interested in the relationship between the true relative speed and strength, rather than the inference of relative speed from proportion data, our results are insensitive to the number of strains circulating in the population.
+where the relative speed of the variant $\delta = \rv - \rw$ can be estimated as the slope of the log odds of $p$  vs.~time.
+When more than two strains are co-circulating, the picture is more complicated \citep{campbell2021increased}; we focus here on comparing strains two at a time.
+\jd{Simplified drastically here.
+Not sure what you added to supp or whether we need it.}
 
@@
 Based on the observed relative growth rate of B.1.1.7 in the UK, we start by taking the relative speed of the variant to be $\delta = 0.1\pday$  \citep{davies2021estimated}; the mean generation interval of the wild type to be $\Gw = 5\days$ \citep{ferretti2020quantifying}; and the squared coefficient of variation of generation intervals to be $\kappa=0.2$ \citep{ferretti2020quantifying} for both the variant and the wild type.
-As noted above, we assume that the variant and the wild type have equal $\kappa$ throughout, and only consider differences in the mean.
+As noted above, we assume throughout that the variant and the wild type can be approximated with gamma distributions with equal $\kappa$, and only consider differences in the mean.
 We evaluate the estimates of relative strength $\rho$ across a wide range of $\kappa$ from 0 (fixed-length generation intervals) to 1 (exponential distribution).
 To further explore how inference depends on underlying epidemiological conditions, we consider five scenarios: (1) $\rw < \rv < 0$, (2) $\rw < \rv = 0$, (3) $\rw < 0 < \rv$, (4) $0 = \rw < \rv$, and (5) $0 < \rw < \rv$.
-In all of these cases, we always assume $\rw < \rv$ to reflect the observed relative speed of $\delta = 0.1\pday$.
-We consider the spread of B.1.1.7 as an example but our qualitative conclusions should hold for other VoCs.
+In all of these cases, we assume $\delta = \rv - \rw = 0.1\pday$.
 
@@
 
-\fref{relspeed} also shows that when $\rho$ is fixed relative speed depends on underlying epidemiological conditions---that is, whether the incidence of wild type and the variant are increasing or decreasing.
+\fref{relspeed} also shows that when $\rho$ is fixed relative speed depends on underlying epidemiological conditions---specifically, the absolute strength of the two strains. 
 For example, even when there are no differences in the generation-interval distributions ($\Gv=\Gw$, in this case), changing $\Rw$ from 0.49 to 1.27 (and thus $\Rv$ from 0.79 to 2.04) yields the relative speed $\delta$ between 0.08--0.11 for $\kappa=0.2$ and 0.06--0.14 for $\kappa=0.5$.
-Therefore, characterizing the spread of variants assuming constant relative speed (e.g., by fitting a standard logistic growth curve \citep{switzerland2021variant}) should be used with care.
+Therefore, characterizing the spread of variants by assuming constant relative speed (e.g., by fitting a standard logistic growth curve \citep{switzerland2021variant} and using the resulting value of $\delta$) should be done with care.
 
@@
 
-In practice, estimating the instantaneous reproduction number is complicated given that it requires incidence of infection.
+In practice, estimating the instantaneous reproduction number is complicated because it requires estimating incidence of infection.
 Observed case counts are sensitive to reporting delays \citep{goldstein2009reconstructing} and changes in case definitions \citep{tsang2020effect}, which in turn can affect estimates of the instantaneous reproduction number \citep{gostic2020practical}.
@@
 \end{equation}
+\jd{Isn't here where we should say pure strength or constant strength?:}
 Under strength-like intervention measures that reduce transmission rates of both strains by a constant amount, we expect ratios between reproduction numbers to remain constant and correspond to the true relative strength: $\Rv(t)/\Rw(t) = \rho$ \citep{leung2017monitoring,leung2020empirical,di2021impact,leung2021early}.
@@
 (1) the variant has a shorter mean generation interval (\fref{Rtbias}A--C);
-(2) the wild type and the variant have the same generation-interval distributions (which match the known distribution, shown in \fref{Rtbias}D--F); and
+(2) the wild type and the variant have the same (known) generation-interval distributions (\fref{Rtbias}D--F); and
 (3) the variant has a longer mean generation interval (\fref{Rtbias}G--I).
@@
 In this case, we can control the spread of the variant when $\theta > \Rv$ or $\phi > \rv$, respectively \citep{doi:10.1098/rspb.2020.1556}.
-We consider constant-strength and speed interventions that reduce $\Rv$ to 0.9 when both the variant and the wild type have equal mean generation intervals (\fref{strengthspeed}A--B).
+We consider constant-strength and -speed interventions calibrated to reduce $\Rv$ to 0.9 under the assumption that the variant generation intervals match those of the wild type(\fref{strengthspeed}A--B).
 While both interventions are equally effective on the strength scale (that is, $\Ry{post}=\int  K_{\mathrm{post}}(\tau) \dtau = 0.9$), they have different dynamical implications.
@@
 
-However, if the variant has longer generation intervals than the wild type (\fref{strengthspeed}C--D), then the strength of the variant will be higher conditional on the observed speed (\fref{strengthspeed}E).
+However, if the variant has longer generation intervals than the wild type (\fref{strengthspeed}C--D), then the strength of the variant will be higher (because we hold the observed speed  constant)(\fref{strengthspeed}E).
 In this case, the same constant-strength intervention can fail to control the epidemic (i.e., $\Ry{post} > 1$; \fref{strengthspeed}E) because this intervention reduces the transmission by a constant amount regardless of age of infection (\fref{strengthspeed}C).
@@
 Finally, we synthesize our findings and illustrate the differences between constant-strength and constant-speed interventions using epidemic simulations (\fref{control_sim}).
-As before, we consider a scenario in which a new variant is emerging ($\rv > 0\pday$ and $\Rw > 1$) while the incidence of infections caused by the wild type is constant ($\rw = 0\pday$ and $\Rw =1$) before interventions are reduced.
+As before, we consider a scenario in which a new variant is emerging with known relative speed($\delta=\rv > 0.1\pday$ and $\Rw > 1$) while the incidence of infections caused by the wild type is constant ($\rw = 0\pday$ and $\Rw =1$) before interventions are reduced.
 The constant-strength intervention then reduces both $\Rw$ and $\Rv$ by a constant factor $\theta$ beginning day 30, 
 whereas the constant-speed intervention isolates individuals infected with both the wild type and the variant at a constant hazard $\phi$ after day 30.
-If the variant has a longer mean generation interval, the constant-strength intervention fails to suppress the spread of the variant (\fref{control_sim}A) because a stronger intervention ($>\phi$) to reduce a higher pre-intervention reproduction number below 1 (\fref{control_sim}B).
+If the variant has a longer mean generation interval, the constant-strength intervention fails to suppress the spread of the variant (\fref{control_sim}A) because the longer generations imply a higher initial $\Rw$, and thus a stronger intervention is required to reduce $\Rw$ ber below 1 (\fref{control_sim}B).
 The relative strength remains constant under constant-strength intervention (\fref{control_sim}B), but the relative speed changes (\fref{control_sim}C) as discussed earlier in Section 4---in this case, the longer mean generation interval variant increases the relative speed when the intervention is introduced.
@@
 we do not claim that they are necessarily more accurate---
-while individual-level data from contact tracing can provide a more reliable source of information about the transmissibility and time scale of transmission in some cases, they can be also biased towards particular types of contacts---for example, household contacts are probably more likely to be identified---which could also affect the estimate of $\rho$.
+while individual-level data from contact tracing can provide a more reliable source of information about the transmissibility and time scale of transmission in some cases, they can also be biased towards particular types of contacts---for example, household contacts are probably more likely to be identified---which could also affect the estimate of $\rho$.
 Instead, this calculation simply highlights the importance of propagating uncertainty in generation-interval distributions in assessing the relative strength and speed of SARS-CoV-2 variants.
@@
 
-We considered two idealized interventions of constant strength and speed, but epidemiological impacts of real interventions are likely more complex.
+We mostly considered idealized interventions of constant strength  or constant speed. Real interventions are likely more complex.
 Interventions that reduce contact rates, such as social distancing measures and vaccination, are generally strength-like, but they can also elicit awareness-based behavioral responses that are speed-like---for example, infected individuals may self-isolate faster after symptom onset.
@@
 Their effectiveness can further depend on the ``coverage'' of isolation, which is strength-like---for example, there may be a group of individuals that cannot be isolated by certain interventions due to asymptomaticity or the lack of participation in the contact tracing program.
-However, analyses of SARS-CoV-2 dynamics have primarily relied on the constant-strength framework \citep{gostic2020practical,unwin2020state,brauner2021inferring}, even when modeling speed-like interventions such as self-isolation \citep{flaxman2020Rt,brett2020transmission}.
+Analyses of SARS-CoV-2 dynamics have primarily relied on the constant-strength framework \citep{gostic2020practical,unwin2020state,brauner2021inferring}, even when modeling speed-like interventions such as self-isolation \citep{flaxman2020Rt,brett2020transmission}.
 While the constant-strength framework provides a convenient tool for modeling and understanding epidemic dynamics, we encourage researchers to consider both strength- and speed-based perspectives as they can lead to different conclusions.

### 2021-07-16 response.tex
some more edits

@@
 
-\jd{If we said Thank you a little less, it might mean a little more.}
-
 Thank you for the review.
@@
 
-\swp{JD, we need a new title.} \jd{I think it helps a lot to just not say ``distributions'' since we're mostly focused on changes in the mean.}
-
-\jd{Have we looked at the Miller paper?}
-
-Thank you for the suggestions. We have changed the title and the abstract to highlight the novelty of our work. See also below.
+We have changed the title and the abstract to highlight the novelty of our work. 
+See also below.
 
@@
 
-Thank you for the suggestions. We now provide a review of this topic in the introduction along with a more detailed justification of the assumptions of our work:
+We now provide a review of this topic in the introduction along with a more detailed justification of the assumptions of our work:
 
-``Analyses of new variants have typically characterized \emph{relative} strength (i.e., the ratio of reproduction numbers $\Rv/\Rw$) and speed (i.e., the difference between growth rates $\rv-\rw$) of the variants.
-Many studies have tried to estimate the relative strength of variants from the observed relative speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021,campbell2021increased}.
-Some studies have instead assumed a value of the relative strength and tried to predict its relative speed \citep{davies2021estimated,di2021impact}.
-While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the disease and its control because two quantities are linked by the generation-interval distribution \citep{doi:10.1098/rspb.2020.1556}.
-For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals leads to a slower epidemic growth (lower $r$), making the epidemic look easier to control with speed-like interventions (e.g., contact tracing), which depend on isolation of infected individuals and therefore affect late transmission disproportionately---as shown in \cite{doi:10.1098/rspb.2020.1556}, the effective hazard of isolation must be greater than the epidemic speed $r$ to reach the $r=0$ threshold.
-However, since $\RR$ is fixed, the generation-interval distribution is expected to have negligible effects on the effectiveness of strength-like interventions (e.g., social distancing), which reduce the hazard of infectious contacts and therefore affect transmission by an equal amount regardless of age of infection.
-Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals lead to a stronger epidemic (higher $\RR$), making the epidemic look harder to control with strength-like interventions (but not necessarily with speed-like interventions).''
+``Epidemic speed can be typically estimated from incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}, but epidemic strength is difficult to measure directly.
+Instead, epidemic strength is often inferred from the observed epidemic speed using the generation-interval distribution, an approach popularized by \citep{wallinga2007generation}.
+The generation interval is defined as the time between infection and transmission and provides information about the time scale of individual-level transmission \citep{svensson2007note}.
+The exact shape of the distribution depends on several factors---including the shape of latent and infectious periods \citep{lloyd2001realistic,wearing2005appropriate,roberts2007model} as well as more detailed life history of a disease \citep{huber2016quantitative}---and thus can be difficult to estimate.
+While it is possible to consider general forms of generation-interval distributions \citep{miller2010epidemics,svensson2015influence}, summarizing the distribution in terms of its mean and variability---for example, by using the Gamma distribution---can still provide a robust link between epidemic speed and strength for real diseases and yield important biological insights \citep{park2019practical}.
+In particular, several studies have noted, in many contexts, that mechanisms that increase the mean generation interval can increase epidemic strength given observed epidemic speed \citep{eaton2014proportion,powers2014impact,weitz2015modeling,park2020time}.''
 
-Assuming Gamma-distributed generation intervals is different from the standard Gamma-chain trick. We [now?] explain this later in Secion 2:
+Assuming Gamma-distributed generation intervals is different from the standard Gamma-chain trick. We now explain this later in Section 2:
 
@@
 We have argued (https://doi.org/10.1098/rspb.2020.1556) that large segments of the epidemiology and modeling community have an over-emphasis on R, and have tried to use the terms speed and strength to underline our contention that these are better seen as complementary perspectives (and to link them to complementary perspectives on control).
-As we argue here and elsewhere, we believe that sometimes one perspective ro another gives a clearer picture.
+As we argue here and elsewhere, we believe that sometimes one perspective over another gives a clearer picture.
 For example, in the beginning of an epidemic, it is more appropriate to fix epidemic speed $r$ when evaluating uncertainties in the generation-interval distribution.
@@
 
-Thank you for pointing this out. We now cite Breda et al., and acknowledge that ``This framework ... generalizes compartmental model, such as the SEIR model''
+We now cite Breda et al., and acknowledge that ``This framework ... generalizes compartmental model, such as the SEIR model''
 
@@
 
-Thank you for pointing this out. We now address this point:
+We now address this point:
 
@@
 We have tried to make the limitation and assumptions clearer throughout the text. 
-\jd{Anything about the generality?
-I don't feel we rely much on mass-action or constant distributions.}
+\jd{Anything about the generality? I don't feel we rely much on mass-action or constant distributions.}
 
@@
 
-Thank you for pointing this out.
 We acknowledge that we are assuming mass action.
-However, a fixed generation-interval distribution does not rely on the assumption that the rate of depletion of susceptibles over time is negligible.
-\jd{Can we say something about the rR link at any particular time? Doesn't that also work?}
+We note that we primarily focus on short-term dynamics until Section 4.
+In this case, we can approximate the dynamics using exponential growth/decay, and therefore the generation-interval distribution is expected to remain fixed:
+
+``Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$, in which case the incidence of infections changes exponentially.
+Here, we use the term ``epidemiological conditions'' to broadly refer to all factors that affect transmission---mathematically, they are captured by the kernel $K(t, \tau)$.
+In the context of SARS-CoV-2 infections, we are essentially assuming that the changes in susceptible pool, behavior, and contact rates are usually small over a short period of time.
+Then, the incidence of each strain grows (or decays) exponentially at rate $r_x$, satisfying the Euler-Lotka equation \citep{wallinga2007generation}:
+\begin{equation}
+\frac{1}{\RR_x} = \int_0^\infty \exp(- r_x \tau) g_x(\tau) \dtau.
+\end{equation}''
+
+We further note that a fixed generation-interval distribution does not rely on the assumption that the rate of depletion of susceptibles over time is negligible.
 This is due to an ongoing confusion between the intrinsic and realized generation interval as we explained above.
@@
 
-Thank you for the suggestion. We now define them more clearly throughout the text. \jd{Presumably there's a particular place where we lay out the definitions.}
+We now define them more clearly throughout the text, including in the introduction:
+
+``For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals leads to a slower epidemic growth (lower $r$), making the epidemic look easier to control with speed-like interventions (e.g., contact tracing), which depend on isolation of infected individuals and therefore affect late transmission disproportionately---as shown in \cite{doi:10.1098/rspb.2020.1556}, the effective hazard of isolation must be greater than the epidemic speed $r$ to reach the $r=0$ threshold.
+In this same case, however, the generation-interval distribution is expected to have negligible effects on the effectiveness of strength-like interventions (e.g., social distancing), which reduce the hazard of infectious contacts and therefore affect transmission by an equal amount regardless of age of infection.
+Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals leads to a higher estimate of epidemic strength ($\RR$), making the epidemic look harder to control with strength-like interventions (but not with speed-like interventions, whose effectiveness is fixed if speed is fixed).''
 
@@
 
-Thank you for pointing this out. We have re-written this part:
+We have now re-written this part:
 
@@
 
-Thank you for pointing this out. We have tried to make these distinctions clearer throughout text.
+We have tried to make these distinctions clearer throughout text, including in the introduction (see response to reviewer 2).
 
@@
 
-Thank you for this suggestion. We now clarify this when it is mentioned for the first time:
+We now clarify this when it is mentioned for the first time:
 
-\jd{This is the sort of text I was hinting at above about mass action.}
 ``Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$.

### 2021-07-16 newvariant.tex
some more edits

@@
 	\textbf\newline{
-	  The importance of generation intervals in investigating dynamics and control of new SARS-CoV-2 variants }
+	  The importance of the generation interval in investigating dynamics and control of new SARS-CoV-2 variants }
 }
@@
 In this same case, however, the generation-interval distribution is expected to have negligible effects on the effectiveness of strength-like interventions (e.g., social distancing), which reduce the hazard of infectious contacts and therefore affect transmission by an equal amount regardless of age of infection.
-
 Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals leads to a higher estimate of epidemic strength ($\RR$), making the epidemic look harder to control with strength-like interventions (but not with speed-like interventions, whose effectiveness is fixed if speed is fixed).
@@
 
-Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$.
+Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$, in which case the incidence of infections changes exponentially.
 Here, we use the term ``epidemiological conditions'' to broadly refer to all factors that affect transmission---mathematically, they are captured by the kernel $K(t, \tau)$.

### 2021-07-16 newvariant.tex
getting through

@@
 Many studies have tried to estimate the relative strength of variants from the observed relative speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021,campbell2021increased}.
-Some studies have instead assumed a value of the relative strength and tried to predict its relative speed \citep{davies2021estimated,di2021impact}.
+Instead (or additionally), some studies have instead assumed a value of the relative strength and tried to predict its relative speed \citep{davies2021estimated,di2021impact}.
 \jd{``instead'' seems wrong here, since we cite davies2021estimated in both lists. What's going on?}
@@
 
-Neglecting the (relatively slow) rate of new mutations and assuming a mass-action epidemic, the current incidence of infection $i_x(t)$ caused by each strain $x$---either the wild type (``wt'') or the variant (``var'')---can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary infections are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
+Neglecting the (relatively slow) rate of new mutations and assuming a homogeneous mixing, the current incidence of infection $i_x(t)$ caused by each strain $x$---either the wild type (``wt'') or the variant (``var'')---can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary infections are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
 \begin{equation}

### 2021-07-19 response.tex
get back after dinner

@@
 The generation interval is defined as the time between infection and transmission and provides information about the time scale of individual-level transmission \citep{svensson2007note}.
+The generation interval is also distinctly different from other so-called ``transmission intervals'' that measure time between successive infections---this includes the serial interval, which is defined as the time between symptom onsets in an infector-infectee pair \citep{fine2003interval,grassly2008mathematical,britton2019estimation,ali2020serial,park2021forward}.
+
 The exact shape of the distribution depends on several factors---including the shape of latent and infectious periods \citep{lloyd2001realistic,wearing2005appropriate,roberts2007model} as well as more detailed life history of a disease \citep{huber2016quantitative}---and thus can be difficult to estimate.
 While it is possible to consider general forms of generation-interval distributions \citep{miller2010epidemics,svensson2015influence}, summarizing the distribution in terms of its mean and variability---for example, by using the Gamma distribution---can still provide a robust link between epidemic speed and strength for real diseases and yield important biological insights \citep{park2019practical}.
-In particular, several studies have noted, in many contexts, that mechanisms that increase the mean generation interval can increase epidemic strength given observed epidemic speed \citep{eaton2014proportion,powers2014impact,weitz2015modeling,park2020time}.''
+In particular, several studies have noted, in many contexts, that mechanisms that increase the mean generation interval increase epidemic strength $\RR$ given observed epidemic speed $r$ \citep{eaton2014proportion,powers2014impact,weitz2015modeling,park2020time}.''
 
@@
 
-``Instead of inferring relative strength from speed, one can directly estimate time-varying reproduction numbers $\RR(t)$ of the variant and the wild type from incidence data and directly compare their ratios when the incidence of infection and the generation-interval distributions are known---such methods have been used in previous analyses of the B.1.1.7 strain by \cite{volz2021transmission}.
-Broadly, there are two types of reproduction numbers: case reproduction number and instantaneous reproduction number.
+``Instead of inferring relative strength from speed, one can directly estimate time-varying reproduction numbers $\RR(t)$ of the variant and the wild type from incidence data and directly compare their ratios when the incidence of infection and the generation-interval distributions are known---such methods have been used in previous analyses of the alpha variant by \cite{volz2021transmission}.
+Broadly, there are two types of time-varying reproduction numbers: case reproduction number and instantaneous reproduction number.
 The case reproduction number is defined as the average number of secondary infections caused by an individual infected at time $t$ and therefore depends on transmission after time $t$ \citep{wallinga2004different}.
 The instantaneous reproduction number is defined as the average number of secondary infections that would be caused by an individual infected at time $t$ if conditions were to remain the same \citep{fraser2007estimating};
-therefore, the instantaneous reproduction number only depends on transmission at time $t$ and is most appropriate for real-time evaluation of changes in transmission, as argued by \citep{gostic2020practical}.
+therefore, the instantaneous reproduction number only depends on transmission at time $t$ and is most appropriate for real-time evaluation of changes in transmission \citep{gostic2020practical}.
 The estimation of the instantaneous reproduction number was popularized by \cite{cori2013new} and has been widely adopted in epidemiological analyses of SARS-CoV-2 \citep{abbott2020estimating,knight2020estimating,flaxman2020Rt,brauner2021inferring,li2021temporal}.''
@@
 
-``Under strength-based changes that reduce the hazard of transmission (e.g., susceptible depletion and decreased transmission rates), the instantaneous generation-interval distribution does not across change time \citep{fraser2007estimating}.
+``Under constant-strength changes that reduce the transmission rate by a constant amount regardless of the age of infection, the instantaneous generation-interval distribution does not across change time \citep{fraser2007estimating}.
 In this case, the instantaneous generation-interval distribution is also often referred to as the intrinsic generation-interval distribution \citep{champredon2015intrinsic,champredon2018two,gostic2020practical,park2020time}---
@@
 The instantaneous generation-interval distribution is also different from the realized generation-interval distribution, which measures time between actual infection events (see \citep{champredon2015intrinsic}).
-Previous studied have noted, in many contexts, that the realized generation intervals can contract due to susceptible depletion---a special case of strength-based changes---due to decreases in transmission over time \citep{kenah2008generation,nishiura2010time,champredon2015intrinsic}, even though the instantaneous generation-interval distribution remains unchanged in this scenario.
-While the instantaneous generation-interval distribution can change under speed-based changes that disproportionately affect late transmissions (see \cite{fraser2007estimating} and Section 6--7 for detailed discussions), assuming a time-invariant instantaneous generation-interval distribution is often appropriate in the context of SARS-CoV-2, given that the dynamics of its spread have been primarily driven by strength-based changes, such as social distancing measures \citep{flaxman2020Rt} and vaccination \citep{moore2021vaccination}.
-Indeed, most dynamical models of SARS-CoV-2 infections have solely relied on strength-based changes, either implicitly or explicitly assuming a time-invariant intrinsic generation-interval distribution.
-Therefore, we neglect changes in the intrinsic generation-interval distribution over time for now and focus on the impact of strength-based changes on the inference of dynamics of new SARS-CoV-2 variants.
-We revisit these ideas in Section 6 and compare the effects of speed-based interventions with those of strength-based interventions.''
+Previous studies have noted, in many contexts, that the realized generation intervals can contract due to susceptible depletion---a special case of constant-strength changes \citep{kenah2008generation,nishiura2010time,champredon2015intrinsic}---even though the instantaneous generation-interval distribution remains unchanged in this scenario.
+While the instantaneous generation-interval distribution can change under speed-like changes (see \cite{fraser2007estimating} and Section 6--7 for detailed discussions), assuming a time-invariant instantaneous generation-interval distribution is often appropriate in the context of SARS-CoV-2, given that the dynamics of its spread have been primarily driven by strength-like intervention, such as social distancing measures \citep{flaxman2020Rt} and vaccination \citep{moore2021vaccination}.
+Indeed, many dynamical models of SARS-CoV-2 infections have solely relied on constant-strength changes, either implicitly or explicitly assuming a time-invariant intrinsic generation-interval distribution (e.g., \citep{flaxman2020Rt,gostic2020practical,brauner2021inferring}).
+Therefore, we neglect changes in the intrinsic generation-interval distribution over time for now and focus on the impact of constant-strength changes on the inference of dynamics of new SARS-CoV-2 variants.
+We revisit these ideas in Section 6 and compare the effects of constant-speed interventions with those of constant-strength interventions.''
 
@@
 We are certainly not trying to lean in to innumeracy.
-We agree that ``strength'' and ``speed'' are not used in the classical physics sense -- instead, we try to be clear how we define them (and have tried to clarify further).
+We agree that ``strength'' and ``speed'' are not used in the classical physics sense---instead, we try to be clear how we define them (and have tried to clarify further).
 
-We have argued (https://doi.org/10.1098/rspb.2020.1556) that large segments of the epidemiology and modeling community have an over-emphasis on R, and have tried to use the terms speed and strength to underline our contention that these are better seen as complementary perspectives (and to link them to complementary perspectives on control).
+We (the first and the last author) have argued (\url{https://doi.org/10.1098/rspb.2020.1556}) that large segments of the epidemiology and modeling community have an over-emphasis on R, and have tried to use the terms speed and strength to underline our contention that these are better seen as complementary perspectives (and to link them to complementary perspectives on control).
 As we argue here and elsewhere, we believe that sometimes one perspective over another gives a clearer picture.
@@
 
-We now cite Breda et al., and acknowledge that ``This framework ... generalizes compartmental model, such as the SEIR model''
+We now cite Breda et al., and acknowledge that ``This framework provides a flexible way of modeling disease dynamics and generalizes a wide range of compartmental models, including the SEIR model \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral,breda2012formulation, champredon2018equivalence}.''
 
@@
 
-We now address this point:
+We now acknowledge this point and provide a citation for a study that discusses this point:
 
-``When more than two strains are co-circulating, the growth rate at a logit scale is expected to change over time and does not necessarily correspond to the relative speed even when all strains exhibit exponential changes (see Supplementary Text);
-instead, as shown in \citep{campbell2021increased}, a multinomial logistic regression model is required to infer relative speed.''
+``When more than two strains are co-circulating, the picture is more complicated \citep{campbell2021increased}; we focus here on comparing two strains at a time.''
 
@@
 
-Thank you for the suggestion. We have reworded this section:
+We have reworded this section:
 
-\jd{We should deal separately with $\kappa$ and with the gamma-generation assumption.}
-``For simplicity we neglect differences in the squared coefficient of variation and assume $\kappa_{\mathrm{wt}} = \kappa_{\mathrm{var}} = \kappa$.
-While the exact shape of the distribution matters, we choose to focus on differences in the mean generation intervals to provide a qualitative understanding of how differences in generation-interval distributions between the wild type and the variant affects the relationship between relative strength and speed.
-In general, we expect  differences in the squared coefficient of variation to have qualitatively similar effects as differences in the mean.
-For example, a wider distribution allows for more early transmission, which in turn reduces epidemic strength given speed---this effect is qualitatively similar to that of a distribution with a shorter mean.''
+``Here, we focus on differences in mean generation intervals, assuming that both the squared coefficient of variation and distribution shape of the variant are otherwise the same as for the wild type: $\kappa_{\mathrm{wt}} = \kappa_{\mathrm{var}} = \kappa$.
+Changes in shape can be important \citep{miller2010epidemics,svensson2015influence}, and we do not investigate them here. 
+We do note that, loosely, we expect a distribution with higher coefficient of variation to allow for more early transmission, and thus to have qualitatively similar effects to a distribution with a shorter mean in many cases \citep{park2019practical}.''
 
@@
 We have tried to make the limitation and assumptions clearer throughout the text.
-\jd{Anything about the generality? I don't feel we rely much on mass-action or constant distributions.}
 
@@
 
-We acknowledge that we are assuming mass action.
+We acknowledge that we are assuming homogeneous mixing---we feel that this is a clearer explanation because mass-action is often used to refer to how the transmission rate, beta, scales with the population size.
+
 We note that we primarily focus on short-term dynamics until Section 4.
@@
 
-``For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals leads to a slower epidemic growth (lower $r$), making the epidemic look easier to control with speed-like interventions (e.g., contact tracing), which depend on isolation of infected individuals and therefore affect late transmission disproportionately---as shown in \cite{doi:10.1098/rspb.2020.1556}, the effective hazard of isolation must be greater than the epidemic speed $r$ to reach the $r=0$ threshold.
-In this same case, however, the generation-interval distribution is expected to have negligible effects on the effectiveness of strength-like interventions (e.g., social distancing), which reduce the hazard of infectious contacts and therefore affect transmission by an equal amount regardless of age of infection.
-Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals leads to a higher estimate of epidemic strength ($\RR$), making the epidemic look harder to control with strength-like interventions (but not with speed-like interventions, whose effectiveness is fixed if speed is fixed).''
+``For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals leads to a slower epidemic growth (lower $r$), making the epidemic look easier to control with a constant-speed intervention, which isolates infected individuals at a constant hazard and therefore affect late transmission disproportionately---as shown in \cite{doi:10.1098/rspb.2020.1556}, the hazard of isolation must be greater than the epidemic speed $r$ to reach the $r=0$ threshold.
+In this same case, however, the uncertainty in the generation-interval distribution is expected to have no effects on the effectiveness of a constant-strength intervention, which reduce the hazard of infectious contacts and therefore affect transmission by an equal amount regardless of age of infection.
+Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals leads to a higher estimate of epidemic strength ($\RR$), making the epidemic look harder to control with a constant-strength interventions (but not with a constant-speed interventions, whose effectiveness is fixed if speed is fixed).
+Constant strength and speed are idealized representations of real-life interventions, which can be either strength-like (e.g., vaccination and social distancing) or speed-like (e.g., contact tracing and isolation) depending on how their effectiveness varies across the generation interval (see \cite{doi:10.1098/rspb.2020.1556} and Discussion).''
 
@@
 ``Although these estimates are more consistent with the attack rate analysis \citep{ukinvest},
-we do not claim that they are necessarily more accurate---while individual-level data from contact tracing can provide a more reliable source of information about the transmissibility and time scale of transmission in some cases, they can be also biased towards particular types of contacts that are easily identified such as household contacts, which could also affect the estimate of $\rho$.''
+we do not claim that they are necessarily more accurate---
+while individual-level data from contact tracing can provide a more reliable source of information about the transmissibility and time scale of transmission in some cases, they can also be biased towards particular types of contacts---for example, household contacts are probably more likely to be identified---which could also affect the estimate of $\rho$.''
 
@@
 We have tried to make these distinctions clearer throughout text, including in the introduction (see response to reviewer 2).
+We now begin by introducing constant-strength and constant-speed interventions, which have clear definitions, and generalize them (conceptually) to explain speed- and strength-like interventions.
 
@@
 
-``Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$.
+``Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$, in which case the incidence of infections changes exponentially.
 Here, we use the term ``epidemiological conditions'' to broadly refer to all factors that affect transmission---mathematically, they are captured by the kernel $K(t, \tau)$.
-In the context of SARS-CoV-2 infections, we are essentially assuming that the changes in susceptible pool, behavior, and contact rates are negligible over a short period of time.''
+In the context of SARS-CoV-2 infections, we are essentially assuming that the changes in susceptible pool, behavior, and contact rates are usually small over a short period of time.''
 
@@
 
-Thank you for the suggestions. We have tried to make the figure caption clearer.
+We have tried to make the figure caption clearer.
 
@@
 
-Thank you for the suggestions. We have added Figure 5 and some additional text to explain the differences between strength- and speed-based interventions.
+We have added Figure 5 and some additional text to explain the differences between strength- and speed-based interventions.
 
@@
 
-Thank you for the suggestions. We have rewritten this part:
+We have rewritten this part:
 
-``these biases may be assessed by considering whether estimates of relative strength appear to vary systematically with the direction of changes in the incidence of infections caused by the variant''
+``these biases may be assessed by considering whether estimates of relative strength appear to vary systematically with the direction of changes in the incidence of infections caused by the variant.''
 

### 2021-07-19 newvariant.tex
get back after dinner

@@
 	\textbf\newline{
-	  The importance of the generation interval in investigating dynamics and control of new SARS-CoV-2 variants }
+	  The importance of the generation interval in investigating dynamics and control of new SARS-CoV-2 variants}
 }
@@
 Inferring the relative strength (i.e., the ratio of reproduction numbers, $\rho=\Rv/\Rw$) and relative speed (i.e., the difference between growth rates, $\delta=\rv-\rw$) of new SARS-CoV-2 variants compared to their wild types is critical to predicting and controlling the course of the current pandemic.
-We explore how differences in the generation-interval distributions affect the inference of relative strength $\rho$ and speed $\delta$ in light of recent evidence that the alpha \jd{suggesting that we use spelled-out greek names for variants, to reduce confusion with rendered greek letters for quantities} variant may have a longer infectious period (and therefore, a longer generation interval) than prior dominant lineages.
+We explore how differences in the generation-interval distributions affect the inference of relative strength $\rho$ and speed $\delta$ in light of recent evidence that the alpha variant may have a longer infectious period (and therefore, a longer generation interval) than prior dominant lineages.
 Neglecting these differences gives biased estimates of $\rho$, which can vary systematically with changes in the incidence of infections caused by the variant.
@@
 The generation interval is defined as the time between infection and transmission and provides information about the time scale of individual-level transmission \citep{svensson2007note}.
+The generation interval is also distinctly different from other so-called ``transmission intervals'' that measure time between successive infections---this includes the serial interval, which is defined as the time between symptom onsets in an infector-infectee pair \citep{fine2003interval,grassly2008mathematical,britton2019estimation,ali2020serial,park2021forward}.
+
 The exact shape of the distribution depends on several factors---including the shape of latent and infectious periods \citep{lloyd2001realistic,wearing2005appropriate,roberts2007model} as well as more detailed life history of a disease \citep{huber2016quantitative}---and thus can be difficult to estimate.
 While it is possible to consider general forms of generation-interval distributions \citep{miller2010epidemics,svensson2015influence}, summarizing the distribution in terms of its mean and variability---for example, by using the Gamma distribution---can still provide a robust link between epidemic speed and strength for real diseases and yield important biological insights \citep{park2019practical}.
-In particular, several studies have noted, in many contexts, that mechanisms that increase the mean generation interval can increase epidemic strength given observed epidemic speed \citep{eaton2014proportion,powers2014impact,weitz2015modeling,park2020time}.
+In particular, several studies have noted, in many contexts, that mechanisms that increase the mean generation interval increase epidemic strength $\RR$ given observed epidemic speed $r$ \citep{eaton2014proportion,powers2014impact,weitz2015modeling,park2020time}.
 
@@
 Many studies have tried to estimate the relative strength of variants from the observed relative speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021,campbell2021increased}.
-Instead (or additionally), some studies have instead assumed a value of the relative strength and tried to predict its relative speed \citep{davies2021estimated,di2021impact}.
-\jd{``instead'' seems wrong here, since we cite davies2021estimated in both lists. What's going on?}
+Instead (or additionally), some studies have assumed a value of the relative strength and tried to predict its relative speed \citep{davies2021estimated,di2021impact}.
 While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the disease and its control because two quantities are linked by the generation-interval distribution \citep{doi:10.1098/rspb.2020.1556}.
 
-For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals leads to a slower epidemic growth (lower $r$), making the epidemic look easier to control with speed-like interventions (e.g., contact tracing), which depend on isolation of infected individuals and therefore affect late transmission disproportionately---as shown in \cite{doi:10.1098/rspb.2020.1556}, the effective hazard of isolation must be greater than the epidemic speed $r$ to reach the $r=0$ threshold.
-In this same case, however, the generation-interval distribution is expected to have negligible effects on the effectiveness of strength-like interventions (e.g., social distancing), which reduce the hazard of infectious contacts and therefore affect transmission by an equal amount regardless of age of infection.
-Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals leads to a higher estimate of epidemic strength ($\RR$), making the epidemic look harder to control with strength-like interventions (but not with speed-like interventions, whose effectiveness is fixed if speed is fixed).
+For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals leads to a slower epidemic growth (lower $r$), making the epidemic look easier to control with a constant-speed intervention, which isolates infected individuals at a constant hazard and therefore affect late transmission disproportionately---as shown in \cite{doi:10.1098/rspb.2020.1556}, the hazard of isolation must be greater than the epidemic speed $r$ to reach the $r=0$ threshold.
+In this same case, however, the uncertainty in the generation-interval distribution is expected to have no effects on the effectiveness of a constant-strength intervention, which reduce the hazard of infectious contacts and therefore affect transmission by an equal amount regardless of age of infection.
+Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals leads to a higher estimate of epidemic strength ($\RR$), making the epidemic look harder to control with a constant-strength interventions (but not with a constant-speed interventions, whose effectiveness is fixed if speed is fixed).
+Constant strength and speed are idealized representations of real-life interventions, which can be either strength-like (e.g., vaccination and social distancing) or speed-like (e.g., contact tracing and isolation) depending on how their effectiveness varies across the generation interval (see \cite{doi:10.1098/rspb.2020.1556} and Discussion).
 
-Most studies have assumed that the previous dominant strain (wild type) and variants have identical generation-interval distributions, but recent evidence suggests that the B.1.1.7 variant may have a longer duration of infection: 13.3 days (90\% CI: 10.1--16.5) for the new variant and 8.2 days (90\% CI: 6.5--9.7) for the wild type \citep{kissler2021densely}.
-Longer duration of infection suggests that the mean generation interval of B.1.1.7 is likely to be longer than that of the wild type \citep{lloyd2001realistic,wearing2005appropriate,roberts2007model}.
+Most studies have assumed that the previous dominant strain (wild type) and variants have identical generation-interval distributions, but recent evidence suggests that the alpha variant may have a longer duration of infection: 13.3 days (90\% CI: 10.1--16.5) for the new variant and 8.2 days (90\% CI: 6.5--9.7) for the wild type \citep{kissler2021densely}.
+Longer duration of infection suggests that the mean generation interval of the alpha variant is likely to be longer than that of the wild type \citep{lloyd2001realistic,wearing2005appropriate,roberts2007model}.
 Other studies have considered the possibility that the faster growth rate of new variants may be driven, in part, by shorter generation intervals \citep{davies2021estimated,volz2021transmission}.
@@
 
-Here, we explore how different assumptions affect estimates of the relative strength ($\rho=\Rv/\Rw$) and relative speed ($\delta=\rv-\rw$) for a new variant, such as B.1.1.7 \jd{search and replace variant names if you agree}.
-% We compare $\rho$--$\delta$ relationship between relative strength and relative speed under a wide range of assumptions about generation-interval distributions.
+Here, we explore how different assumptions about the generation interval affect estimates of the relative strength ($\rho=\Rv/\Rw$) and relative speed ($\delta=\rv-\rw$) for new variants.
 We show that neglecting differences in the generation-interval distributions can lead to biased estimates.
@@
 
-\jd{Where did -based come from? Do we like it?}
-Under strength-based changes that reduce the hazard of transmission (e.g., susceptible depletion and decreased transmission rates), the instantaneous generation-interval distribution does not across change time \citep{fraser2007estimating}.
+Under constant-strength changes that reduce the transmission rate by a constant amount regardless of the age of infection, the instantaneous generation-interval distribution does not across change time \citep{fraser2007estimating}.
 In this case, the instantaneous generation-interval distribution is also often referred to as the intrinsic generation-interval distribution \citep{champredon2015intrinsic,champredon2018two,gostic2020practical,park2020time}---
@@
 The instantaneous generation-interval distribution is also different from the realized generation-interval distribution, which measures time between actual infection events (see \citep{champredon2015intrinsic}).
-Previous studies have noted, in many contexts, that the realized generation intervals can contract due to susceptible depletion---a special case of strength-based changes \citep{kenah2008generation,nishiura2010time,champredon2015intrinsic}--- even though the instantaneous generation-interval distribution remains unchanged in this scenario.
-While the instantaneous generation-interval distribution can change under speed-based changes (see \cite{fraser2007estimating} and Section 6--7 for detailed discussions), assuming a time-invariant instantaneous generation-interval distribution is often appropriate in the context of SARS-CoV-2, given that the dynamics of its spread have been primarily driven by strength-based changes, such as social distancing measures \citep{flaxman2020Rt} and vaccination \citep{moore2021vaccination}.
-Indeed, most dynamical models of SARS-CoV-2 infections have solely relied on strength-based changes, either implicitly or explicitly assuming a time-invariant intrinsic generation-interval distribution.
-Therefore, we neglect changes in the intrinsic generation-interval distribution over time for now and focus on the impact of strength-based changes on the inference of dynamics of new SARS-CoV-2 variants.
-We revisit these ideas in Section 6 and compare the effects of speed-based interventions with those of strength-based interventions.
+Previous studies have noted, in many contexts, that the realized generation intervals can contract due to susceptible depletion---a special case of constant-strength changes \citep{kenah2008generation,nishiura2010time,champredon2015intrinsic}---even though the instantaneous generation-interval distribution remains unchanged in this scenario.
+While the instantaneous generation-interval distribution can change under speed-like changes (see \cite{fraser2007estimating} and Section 6--7 for detailed discussions), assuming a time-invariant instantaneous generation-interval distribution is often appropriate in the context of SARS-CoV-2, given that the dynamics of its spread have been primarily driven by strength-like intervention, such as social distancing measures \citep{flaxman2020Rt} and vaccination \citep{moore2021vaccination}.
+Indeed, many dynamical models of SARS-CoV-2 infections have solely relied on constant-strength changes, either implicitly or explicitly assuming a time-invariant intrinsic generation-interval distribution (e.g., \citep{flaxman2020Rt,gostic2020practical,brauner2021inferring}).
+Therefore, we neglect changes in the intrinsic generation-interval distribution over time for now and focus on the impact of constant-strength changes on the inference of dynamics of new SARS-CoV-2 variants.
+We revisit these ideas in Section 6 and compare the effects of constant-speed interventions with those of constant-strength interventions.
 
@@
 \end{equation}
-% The 
 Various Gamma assumptions have been widely used in epidemic modeling, including for models of SARS-CoV-2 \citep{doi:10.1098/rsif.2020.0144}.
@@
 We use this framework to investigate how inferences about strength and speed of the variant depend on our assumptions about the underlying generation-interval distributions.
-For simplicity we neglect differences in the squared coefficient of variation and assume $\kappa_{\mathrm{wt}} = \kappa_{\mathrm{var}} = \kappa$.
-Here we focus on differences in mean generation intervals, assuming that both the squared coefficient of variation, and distribution shape of the variant are otherwise the same as for the wild type.
-Changes in shape can be important and we do not investigate them here.
-We do not that, loosely, we expect a distribution with higher coefficient of variation to allow for more early transmission, and thus to have qualitatively similar effects to a distribution with a shorter mean in many cases.
+Here, we focus on differences in mean generation intervals, assuming that both the squared coefficient of variation and distribution shape of the variant are otherwise the same as for the wild type: $\kappa_{\mathrm{wt}} = \kappa_{\mathrm{var}} = \kappa$.
+Changes in shape can be important \citep{miller2010epidemics,svensson2015influence}, and we do not investigate them here. 
+We do note that, loosely, we expect a distribution with higher coefficient of variation to allow for more early transmission, and thus to have qualitatively similar effects to a distribution with a shorter mean in many cases \citep{park2019practical}.
 
@@
 where the relative speed of the variant $\delta = \rv - \rw$ can be estimated as the slope of the log odds of $p$  vs.~time.
-When more than two strains are co-circulating, the picture is more complicated \citep{campbell2021increased}; we focus here on comparing strains two at a time.
-\jd{Simplified drastically here.
-Not sure what you added to supp or whether we need it.}
+When more than two strains are co-circulating, the picture is more complicated \citep{campbell2021increased}; we focus here on comparing two strains at a time.
 
@@
 
-Based on the observed relative growth rate of B.1.1.7 in the UK, we start by taking the relative speed of the variant to be $\delta = 0.1\pday$  \citep{davies2021estimated}; the mean generation interval of the wild type to be $\Gw = 5\days$ \citep{ferretti2020quantifying}; and the squared coefficient of variation of generation intervals to be $\kappa=0.2$ \citep{ferretti2020quantifying} for both the variant and the wild type.
+We start by taking the mean generation interval of the wild type to be $\Gw = 5\days$ \citep{ferretti2020quantifying}; and the squared coefficient of variation of generation intervals to be $\kappa=0.2$ \citep{ferretti2020quantifying} for both the variant and the wild type.
 As noted above, we assume throughout that the variant and the wild type can be approximated with gamma distributions with equal $\kappa$, and only consider differences in the mean.
@@
 To further explore how inference depends on underlying epidemiological conditions, we consider five scenarios: (1) $\rw < \rv < 0$, (2) $\rw < \rv = 0$, (3) $\rw < 0 < \rv$, (4) $0 = \rw < \rv$, and (5) $0 < \rw < \rv$.
-In all of these cases, we assume $\delta = \rv - \rw = 0.1\pday$.
+In all of these cases, we assume $\delta = \rv - \rw = 0.1\pday$ based on the observed relative growth rate of the alpha variant in the UK \citep{davies2021estimated}.
 
@@
 When $\rw < 0 < \rv$, inference of $\rho$ is relatively insensitive to values of $\kappa$.
-
-% 
-% Even when we restrict the difference in generation interval to be less than 20\% and only consider scenarios with $\rv>0$ (bottom panels of \fref{relstrength}), $\rho$ ranges between 1.4--1.8.
 
@@
 
-Instead of inferring relative strength from speed, one can directly estimate time-varying reproduction numbers $\RR(t)$ of the variant and the wild type from incidence data and directly compare their ratios when the incidence of infection and the generation-interval distributions are known---such methods have been used in previous analyses of the B.1.1.7 strain by \cite{volz2021transmission}.
+Instead of inferring relative strength from speed, one can directly estimate time-varying reproduction numbers $\RR(t)$ of the variant and the wild type from incidence data and directly compare their ratios when the incidence of infection and the generation-interval distributions are known---such methods have been used in previous analyses of the alpha variant by \cite{volz2021transmission}.
 Broadly, there are two types of time-varying reproduction numbers: case reproduction number and instantaneous reproduction number.
@@
 \end{equation}
-\jd{Isn't here where we should say pure strength or constant strength?:}
-Under strength-like intervention measures that reduce transmission rates of both strains by a constant amount, we expect ratios between reproduction numbers to remain constant and correspond to the true relative strength: $\Rv(t)/\Rw(t) = \rho$ \citep{leung2017monitoring,leung2020empirical,di2021impact,leung2021early}.
+Under constant-strength intervention measures that reduce transmission rates of both strains by a constant amount, we expect ratios between reproduction numbers to remain constant and correspond to the true relative strength: $\Rv(t)/\Rw(t) = \rho$ \citep{leung2017monitoring,leung2020empirical,di2021impact,leung2021early}.
 However, if the assumed generation-interval distribution $\hat{g}(\tau)$ is different from the true distribution, then the ratio between the estimated reproduction numbers $\hat{\rho}(t) = \hat{\RR}_{\textrm{var}}(t)/\hat{\RR}_{\textrm{wt}}(t)$ may change, even if the true ratio does not.
@@
 
-However, if the variant has longer generation intervals than the wild type (\fref{strengthspeed}C--D), then the strength of the variant will be higher (because we hold the observed speed  constant)(\fref{strengthspeed}E).
+However, if the variant has longer generation intervals than the wild type (\fref{strengthspeed}C--D), then the strength of the variant will be higher because we are holding the observed speed constant (\fref{strengthspeed}E).
 In this case, the same constant-strength intervention can fail to control the epidemic (i.e., $\Ry{post} > 1$; \fref{strengthspeed}E) because this intervention reduces the transmission by a constant amount regardless of age of infection (\fref{strengthspeed}C).
@@
 whereas the constant-speed intervention isolates individuals infected with both the wild type and the variant at a constant hazard $\phi$ after day 30.
-If the variant has a longer mean generation interval, the constant-strength intervention fails to suppress the spread of the variant (\fref{control_sim}A) because the longer generations imply a higher initial $\Rw$, and thus a stronger intervention is required to reduce $\Rw$ ber below 1 (\fref{control_sim}B).
+If the variant has a longer mean generation interval, the constant-strength intervention fails to suppress the spread of the variant (\fref{control_sim}A) because the longer generations imply a higher initial $\Rw$, and thus a stronger intervention is required to reduce $\Rw$ below 1 (\fref{control_sim}B).
 The relative strength remains constant under constant-strength intervention (\fref{control_sim}B), but the relative speed changes (\fref{control_sim}C) as discussed earlier in Section 4---in this case, the longer mean generation interval variant increases the relative speed when the intervention is introduced.
@@
 
-As new variants of SARS-CoV-2, such as B.1.1.7, are spreading and becoming dominant in many countries, it is clear that the variants are more transmissible (have higher strength) than the wild type \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
+As new variants of SARS-CoV-2 are spreading and becoming dominant in many countries, it is clear that the variants are more transmissible (have higher strength) than the wild type \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
 While our analysis supports estimates of a higher strength of the new variant across a wide range of assumptions, it also shows that uncertainty in generation-interval distributions must be taken into account to obtain accurate estimates of the relative strength of variants.
@@
 
-There is currently a (minor) discrepancy in the estimates of relative strength of new variants, particularly for B.1.1.7.
-Mathematical analyses have typically reported greater than a 1.4 fold increase in reproduction number for the B.1.1.7 variant ($\rho > 1.4$) whereas an independent analysis of secondary attack rates from contact tracing data suggests a 1.25--1.4 fold increase \citep{ukinvest}.
+There is currently a (minor) discrepancy in the estimates of relative strength of new variants, particularly for the alpha variant.
+Mathematical analyses have typically reported greater than a 1.4 fold increase in reproduction number for the alpha variant ($\rho > 1.4$) whereas an independent analysis of secondary attack rates from contact tracing data suggests a 1.25--1.4 fold increase \citep{ukinvest}.
 As shown in \cite{davies2021estimated} and \cite{volz2021transmission}, propagating uncertainty in generation interval estimates (and, in particular, assuming shorter generation intervals) can partially explain these differences:

### 2021-07-19 supp.tex
get back after dinner

@@
 
-\subsection{Measuring relative growth for multiple co-circulating strains}
-
-In order to model $N$ co-circulating strains, we write $i_x(t)$ to represent the incidence of strain $x$ (for $x = 1, \dots, N$).
-When incidence is changing exponentially ($i_x(t) = i_x(t_0) \exp(r_x t)$), the proportion of infections caused by strain $y$ can be written as:
-\begin{equation}
-p_y(t) = \frac{i_y(t_0) \exp(r_y t)}{\sum_{x=1}^N i_x(t_0) \exp(r_x t)}.
-\end{equation}
-Logit transforming $p_y(t)$ then gives:
-\begin{align}
-\mathrm{logit}(p_y(t)) &= \log\left(\frac{p_y(t)}{1-p_y(t)}\right)\\
-&= \log\left(\frac{i_y(t_0) \exp(r_y t)}{\sum_{x \neq y} i_x(t_0) \exp(r_x t)}\right)\\
-&= \log(i_y(t_0)) + r_y t - \log \left(\sum_{x \neq y} i_x(t_0) \exp(r_x t)\right)
-\end{align}
-which is a nonlinear function of time $t$.
-Therefore, the growth rate measured at the logit scale is expected to change over time when multiple strains are co-circulating. 
-Instead, a multinomial logistic regression model is required to infer relative speed (see Supplementary Materials in \citep{campbell2021increased} for derivation).
-
 \subsection{Two-strain renewal equation}

### 2021-07-20 newvariant.tex
done

@@
 (B,E) Instantaneous reproduction number of the variant $\Rv(t)$ calculated using \eref{instR}.
-(C,F) Proportion of incidence of infections caused by the variant ($\iv(t)/(\iw(t)+\iv(t))$) on a logit scale. 
+(C,F) Proportion of incidence of infections caused by the variant on a logit scale. 
 Epidemic trajectories are calculated assuming $\rw=0\pday$, $\delta=0.1\pday$, $\rv=0.1\pday$, $\Gy{wt}=5\,\textrm{days}$, and $\kappa=1/5$ for pre-intervention conditions, and then introducing constant-strength and constant-speed interventions on day 30.
@@
 In contrast, epidemic trajectories under the constant-speed intervention look identical regardless of the mean generation interval of the variant (\fref{control_sim}D). 
-As shown in \fref{strengthspeed}E, a constant-speed intervention leads to greater reduction in $\Rv(t)$ when the mean generation-interval of the variant is longer (\fref{control_sim}E).
-In this case, the relative speed of the variant also remains unaffected by the constant-speed intervention because the epidemic speed of the wild type $\rw$ and the variant $\rv$ are both reduced by equal amounts $\phi$, resulting in the post-intervention epidemic speed of $\rw-\phi$ and $\rv-\phi$, respectively.
+As shown in \fref{strengthspeed}E, a constant-speed intervention leads to greater reduction in $\Rv(t)$ when the mean generation-interval of the variant is longer (\fref{control_sim}E)---
+this implies a change in relative strength $\rho$ because the effectiveness of a constant-speed intervention is sensitive to the generation-interval distribution.
+In this case, the relative speed of the variant remains unaffected by the constant-speed intervention because the epidemic speed of the wild type $\rw$ and the variant $\rv$ are both reduced by equal amounts $\phi$, resulting in the post-intervention epidemic speed of $\rw-\phi$ and $\rv-\phi$, respectively.
 

### 2021-07-20 control_sim.R

@@ -3,6 +3,7 @@ library(ggplot2); theme_set(theme_bw(base_size=14, base_family = "Times"))
 library(ggthemes)
 library(egg)
 library(tikzDevice)
+library(ggpubr)
 source("renewal_det.R")
 
 delta <- 0.1
@@ -152,13 +153,11 @@ g1 <- ggplot(simstrengthall) +
   scale_y_log10("Daily incidence", limits=c(1e-6, 0.11), expand=c(0, 0)) +
   scale_color_viridis_d("Variant GI", end=0.9) +
   scale_linetype_manual("Variant GI", values=1:3) +
-  ggtitle("A. Constant strength") +
+  ggtitle("A") +
   theme(
     panel.grid = element_blank(),
     legend.position = c(0.5, 0.3),
-    legend.background = element_rect(fill=NA),
-    axis.title.x = element_blank(),
-    axis.text.x = element_blank()
+    legend.background = element_rect(fill=NA)
   )
 
 g2 <- ggplot(simstrengthall) +
@@ -170,9 +169,7 @@ g2 <- ggplot(simstrengthall) +
   ggtitle("B") +
   theme(
     panel.grid = element_blank(),
-    legend.position = "none",
-    axis.title.x = element_blank(),
-    axis.text.x = element_blank()
+    legend.position = "none"
   )
 
 g3 <- ggplot(simstrengthall) +
@@ -186,9 +183,7 @@ g3 <- ggplot(simstrengthall) +
   ggtitle("C") +
   theme(
     panel.grid = element_blank(),
-    legend.position = "none",
-    axis.title.x = element_blank(),
-    axis.text.x = element_blank()
+    legend.position = "none"
   )
 
 g4 <- ggplot(simspeedall) +
@@ -200,7 +195,7 @@ g4 <- ggplot(simspeedall) +
   scale_y_log10("Daily incidence", limits=c(1e-6, 0.11), expand=c(0, 0)) +
   scale_color_viridis_d("Variant GI", end=0.9) +
   scale_linetype_manual("Variant GI", values=1:3) +
-  ggtitle("D. Constant speed") +
+  ggtitle("D") +
   theme(
     panel.grid = element_blank(),
     legend.position = "none"
@@ -232,12 +227,16 @@ g6 <- ggplot(simspeedall) +
     legend.position = "none"
   )
 
-gtot <- egg::ggarrange(g1, g2, g3, 
-                  g4, g5, g6,
[…]

### 2021-07-26 response.tex
minor edits

@@
 We have tried to frame our results better, and to clarify both the novelty and importance of this study.
-We agree with Reviewer 3 that, while many of these concepts have been raised in the context of the recent emergence of new SARS-CoV-2 variants, careful treatment has been lacking. 
+The primary purpose of our study is two-fold: (1) to illustrate the impact of uncertainty in the generation interval in shaping the population-level transmission advantage of new variants and (2) to evaluate when it is approprite to hold relative strength or speed constant.
+While several studies of new variants have assumed constant relative speed (e.g., by fitting a logistic growth model), we argue that holding relative strength fixed is generally more appropriate given that the dynamics of SARS-CoV-2 have been driven primarily by strength-like interventions.
+We also show that this relationship changes under speed-like interventions---that is, relative strength changes but relative speed remains constant.
 We believe that the addition of a new simulations (Figure 5) and the revised text better highlight the novelty of our work.

### 2021-07-30 response.tex
minor edits

@@
 We have tried to frame our results better, and to clarify both the novelty and importance of this study.
-The primary purpose of our study is two-fold: (1) to illustrate the impact of uncertainty in the generation interval in shaping the population-level transmission advantage of new variants and (2) to evaluate when it is approprite to hold relative strength or speed constant.
+The primary purpose of our study is two-fold: (1) to illustrate the impact of uncertainty in the generation interval in shaping the population-level transmission advantage of new variants and (2) to evaluate when it is appropriate to hold relative strength or speed constant.
 While several studies of new variants have assumed constant relative speed (e.g., by fitting a logistic growth model), we argue that holding relative strength fixed is generally more appropriate given that the dynamics of SARS-CoV-2 have been driven primarily by strength-like interventions.
@@
 We have tried to clarify what we mean by the generation interval.
-In particular, often-menetioned time contractions of generation intervals apply to the forward realized generation interval, which measure time between actual infection events, rather than the intrinsic generation interval, which measure the intrinsic infectiousness of infected individuals.
+In particular, often-mentioned time contractions of generation intervals apply to the forward realized generation interval, which measure time between actual infection events, rather than the intrinsic generation interval, which measure the intrinsic infectiousness of infected individuals.
 Under strength-like interventions, the intrinsic generation interval does not vary across time, even though the realized generation interval contracts.
@@
 
-``For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals leads to a slower epidemic growth (lower $r$), making the epidemic look easier to control with a constant-speed intervention, which isolates infected individuals at a constant hazard and therefore affect late transmission disproportionately---as shown in \cite{doi:10.1098/rspb.2020.1556}, the hazard of isolation must be greater than the epidemic speed $r$ to reach the $r=0$ threshold.
-In this same case, however, the uncertainty in the generation-interval distribution is expected to have no effects on the effectiveness of a constant-strength intervention, which reduce the hazard of infectious contacts and therefore affect transmission by an equal amount regardless of age of infection.
+``For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals leads to a slower epidemic growth (lower $r$), making the epidemic look easier to control with a constant-speed intervention, which isolates infected individuals at a constant hazard and therefore affects late transmission disproportionately---as shown in \cite{doi:10.1098/rspb.2020.1556}, the hazard of isolation must be greater than the epidemic speed $r$ to reach the $r=0$ threshold.
+In this same case, however, the uncertainty in the generation-interval distribution is expected to have no effects on the effectiveness of a constant-strength intervention, which reduces the hazard of infectious contacts and therefore affects transmission by an equal amount regardless of age of infection.
 Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals leads to a higher estimate of epidemic strength ($\RR$), making the epidemic look harder to control with a constant-strength interventions (but not with a constant-speed interventions, whose effectiveness is fixed if speed is fixed).

### 2021-07-30 newvariant.tex
minor edits

@@
 
-Inferring the relative strength (i.e., the ratio of reproduction numbers, $\rho=\Rv/\Rw$) and relative speed (i.e., the difference between growth rates, $\delta=\rv-\rw$) of new SARS-CoV-2 variants compared to their wild types is critical to predicting and controlling the course of the current pandemic.
-We explore how differences in the generation-interval distributions affect the inference of relative strength $\rho$ and speed $\delta$ in light of recent evidence that the alpha variant may have a longer infectious period (and therefore, a longer generation interval) than prior dominant lineages.
-Neglecting these differences gives biased estimates of $\rho$, which can vary systematically with changes in the incidence of infections caused by the variant.
-We further argue that holding $\rho$ constant is generally more appropriate in the context of SARS-CoV-2 and illustrate how $\rho$ and $\delta$ are affected by two idealized interventions.
-A constant-strength intervention, which reduces transmission rates by a constant amount, affects $\delta$ but not $\rho$---this is analogous to vaccination and social distancing.
-A constant-speed intervention, which isolates infected individuals at a constant hazard, affects $\rho$ but not $\delta$---this is analogous to contact tracing.
-Uncertainties in the generation-interval distribution determine the effectiveness of these interventions.
+Inferring the relative strength (i.e., the ratio of reproduction numbers) and relative speed (i.e., the difference between growth rates) of new SARS-CoV-2 variants compared to their wild types is critical to predicting and controlling the course of the current pandemic.
+We explore how differences in the generation-interval distributions affect the inference of relative strength and speed in light of recent evidence that the Alpha and Delta variants may have different infectiousness profiles (and therefore, the generation interval) from prior dominant lineages.
+Neglecting these differences gives biased estimates of relative strength, which can vary systematically with changes in the incidence of infections caused by the variant.
+We further argue that holding relative strength constant is generally more appropriate in the context of SARS-CoV-2 and illustrate how relative strength and speed are affected by two idealized interventions.
+A constant-strength intervention, which reduces transmission rates by a constant amount, affects relative speed but not relative strength---this is analogous to vaccination and social distancing.
+A constant-speed intervention, which isolates infected individuals at a constant hazard, affects relative strength but not relative speed---this is analogous to contact tracing.
+Changes in the generation-interval distribution can alter the effectiveness of these interventions.
 
@@
 
-For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals leads to a slower epidemic growth (lower $r$), making the epidemic look easier to control with a constant-speed intervention, which isolates infected individuals at a constant hazard and therefore affect late transmission disproportionately---as shown in \cite{doi:10.1098/rspb.2020.1556}, the hazard of isolation must be greater than the epidemic speed $r$ to reach the $r=0$ threshold.
-In this same case, however, the uncertainty in the generation-interval distribution is expected to have no effects on the effectiveness of a constant-strength intervention, which reduce the hazard of infectious contacts and therefore affect transmission by an equal amount regardless of age of infection.
-Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals leads to a higher estimate of epidemic strength ($\RR$), making the epidemic look harder to control with a constant-strength interventions (but not with a constant-speed interventions, whose effectiveness is fixed if speed is fixed).
+For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals leads to a slower epidemic growth (lower $r$), making the epidemic look easier to control with a constant-speed intervention, which isolates infected individuals at a constant hazard and therefore affects late transmission disproportionately.
+As shown in \cite{doi:10.1098/rspb.2020.1556}, the hazard of isolation must be greater than the epidemic speed $r$ to reach the $r=0$ threshold.
+In this same case, however, the uncertainty in the generation-interval distribution is expected to have no effects on the effectiveness of a constant-strength intervention, which reduces the hazard of infectious contacts and therefore affects transmission by an equal amount regardless of age of infection.
+Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals leads to a higher estimate of epidemic strength ($\RR$), making the epidemic look harder to control with a constant-strength interventions.
 Constant strength and speed are idealized representations of real-life interventions, which can be either strength-like (e.g., vaccination and social distancing) or speed-like (e.g., contact tracing and isolation) depending on how their effectiveness varies across the generation interval (see \cite{doi:10.1098/rspb.2020.1556} and Discussion).
 
-Most studies have assumed that the previous dominant strain (wild type) and variants have identical generation-interval distributions, but recent evidence suggests that the alpha variant may have a longer duration of infection: 13.3 days (90\% CI: 10.1--16.5) for the new variant and 8.2 days (90\% CI: 6.5--9.7) for the wild type \citep{kissler2021densely}.
-Longer duration of infection suggests that the mean generation interval of the alpha variant is likely to be longer than that of the wild type \citep{lloyd2001realistic,wearing2005appropriate,roberts2007model}.
-Other studies have considered the possibility that the faster growth rate of new variants may be driven, in part, by shorter generation intervals \citep{davies2021estimated,volz2021transmission}.
+Most studies have assumed that the previous dominant strain (wild type) and variants have identical generation-interval distributions, but Kissler \textit{et al.} \citep{kissler2021densely} suggested that the Alpha variant may have a longer duration of infection: 13.3 days (90\% CI: 10.1--16.5) for the new variant and 8.2 days (90\% CI: 6.5--9.7) for the wild type.
+Longer duration of infection suggests that the mean generation interval of the Alpha variant is likely to be longer than that of the wild type \citep{lloyd2001realistic,wearing2005appropriate,roberts2007model}.
+In contrast, some studies have suggested that the more recent Delta variant may have a shorter generation interval \citep{li2021viral, zhang2021transmission}.
+Modeling studies have also considered the possibility that the faster growth rate of new variants may be driven, in part, by shorter generation intervals \citep{davies2021estimated,volz2021transmission}.
 However, linking strength and speed is complicated given that the generation-interval distribution depend on many factors including behavior: 
@@
 
-Neglecting the (relatively slow) rate of new mutations and assuming a homogeneous mixing, the current incidence of infection $i_x(t)$ caused by each strain $x$---either the wild type (``wt'') or the variant (``var'')---can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary infections are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
+Neglecting the (relatively slow) rate of new mutations and assuming homogeneous mixing, the current incidence of infection $i_x(t)$ caused by each strain $x$---either the wild type (``wt'') or the variant (``var'')---can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary infections are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
 \begin{equation}
@@
 
-Under constant-strength changes that reduce the transmission rate by a constant amount regardless of the age of infection, the instantaneous generation-interval distribution does not across change time \citep{fraser2007estimating}.
+Under constant-strength changes that reduce the transmission rate by a constant amount regardless of the age of infection, the instantaneous generation-interval distribution does not change across time \citep{fraser2007estimating}.
 In this case, the instantaneous generation-interval distribution is also often referred to as the intrinsic generation-interval distribution \citep{champredon2015intrinsic,champredon2018two,gostic2020practical,park2020time}---
@@
 \end{equation}
-Various Gamma assumptions have been widely used in epidemic modeling, including for models of SARS-CoV-2 \citep{doi:10.1098/rsif.2020.0144}.
-Although the Gamma assumption includes as a special case models that assume exponentially distributed generation intervals (when $\kappa=1$), corresponding to the SIR model \citep{anderson1991infectious}, it is different from the generation-interval distribution we get when infectious periods are Gamma distributed, another standard assumption in epidemic modeling (see \cite{roberts2007model} for discussion).
+Various Gamma-generation-interval assumptions have been widely used in epidemic modeling, including for models of SARS-CoV-2 \citep{doi:10.1098/rsif.2020.0144}.
+The Gamma-generation-interval assumption includes as a special case models that assume exponentially distributed generation intervals (when $\kappa=1$), corresponding to the SIR model \citep{anderson1991infectious}.
+We note that when the infectious periods are Gamma distributed---another standard assumption in epidemic modeling---the resulting generation interval does not follow the Gamma distribution;
+in particular, the mean generation interval can be shorter than the mean infectious period because transmission occurs throughout the infectious period until recovery (see \cite{roberts2007model} for detailed discussion).
 
@@
 Changes in shape can be important \citep{miller2010epidemics,svensson2015influence}, and we do not investigate them here. 
-We do note that, loosely, we expect a distribution with higher coefficient of variation to allow for more early transmission, and thus to have qualitatively similar effects to a distribution with a shorter mean in many cases \citep{park2019practical}.
+We do note that we expect a distribution with higher coefficient of variation to allow for more early transmission, and thus to have qualitatively similar effects to a distribution with a shorter mean in many cases \citep{park2019practical}.
 
@@
 To further explore how inference depends on underlying epidemiological conditions, we consider five scenarios: (1) $\rw < \rv < 0$, (2) $\rw < \rv = 0$, (3) $\rw < 0 < \rv$, (4) $0 = \rw < \rv$, and (5) $0 < \rw < \rv$.
-In all of these cases, we assume $\delta = \rv - \rw = 0.1\pday$ based on the observed relative growth rate of the alpha variant in the UK \citep{davies2021estimated}.
+In all of these cases, we assume $\delta = \rv - \rw = 0.1\pday$ based on the observed relative growth rate of the Alpha variant in the UK \citep{davies2021estimated}.
 
@@
 
-Instead of inferring relative strength from speed, one can directly estimate time-varying reproduction numbers $\RR(t)$ of the variant and the wild type from incidence data and directly compare their ratios when the incidence of infection and the generation-interval distributions are known---such methods have been used in previous analyses of the alpha variant by \cite{volz2021transmission}.
+Instead of inferring relative strength from speed, one can directly estimate time-varying reproduction numbers $\RR(t)$ of the variant and the wild type from incidence data and directly compare their ratios when the incidence of infection and the generation-interval distributions are known---such methods have been used in previous analyses of the Alpha variant by \cite{volz2021transmission}.
 Broadly, there are two types of time-varying reproduction numbers: case reproduction number and instantaneous reproduction number.
@@
 We consider two types of intervention:
-an intervention of constant strength, which reduces transmission by a constant factor $\theta$ regardless of age of infection ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau)/\theta$); and an intervention of constant speed, which reduces transmission after infection by a constant rate $\phi$ ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau) \exp(-\phi \tau)$).
+an intervention of constant strength (\fref{strengthspeed}A,C,E), which reduces transmission by a constant factor $\theta$ regardless of age of infection ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau)/\theta$); and an intervention of constant speed (\fref{strengthspeed}B,D,F), which reduces transmission after infection by a constant rate $\phi$ ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau) \exp(-\phi \tau)$).
 In this case, we can control the spread of the variant when $\theta > \Rv$ or $\phi > \rv$, respectively \citep{doi:10.1098/rspb.2020.1556}.
-We consider constant-strength and -speed interventions calibrated to reduce $\Rv$ to 0.9 under the assumption that the variant generation intervals match those of the wild type(\fref{strengthspeed}A--B).
+We consider constant-strength and -speed interventions calibrated to reduce $\Rv$ to 0.9 under the assumption that the variant generation intervals match those of the wild type (\fref{strengthspeed}C--D).
 While both interventions are equally effective on the strength scale (that is, $\Ry{post}=\int  K_{\mathrm{post}}(\tau) \dtau = 0.9$), they have different dynamical implications.
@@
 \textbf{Effects of constant-strength and constant-speed interventions on the spread of a new variant with known speed \rv.}
-(A--D) Pre-intervention (black) and post-intervention (colored) kernel of the new variant under constant-strength (A,C) and -speed (B,D) interventions when the variant has equal mean generation interval as the wild type (A--B) and when the variant has longer generation intervals than the wild type (C--D).
+(A--F) Pre-intervention (black) and post-intervention (colored) kernel of the new variant under constant-strength (A,C,E) and -speed (B,D,F) interventions when the variant has shorter (A--B), equal (C--D), or longer (E--F) mean generation interval than the wild type.
 Vertical lines represent the mean generation interval.
-(E) Pre-intervention $\Rv$ (black) and post-intervention epidemic strength $\Ry{post}$ (colored) conditional on the mean generation interval of the variant.
-(F) Pre-intervention $\rv$ (black) and post-intervention epidemic speed $\ry{post}$ (colored) conditional on the mean generation interval of the variant.
+(G) Pre-intervention $\Rv$ (black) and post-intervention epidemic strength $\Ry{post}$ (colored) conditional on the mean generation interval of the variant.
+(H) Pre-intervention $\rv$ (black) and post-intervention epidemic speed $\ry{post}$ (colored) conditional on the mean generation interval of the variant.
 Epidemic strength and speed are calculated assuming $\rw=0\pday$, $\delta=0.1\pday$, $\rv=0.1\pday$, $\Gy{wt}=5\,\textrm{days}$, and $\kappa=1/5$ for pre-intervention conditions.
@@
 
-However, if the variant has longer generation intervals than the wild type (\fref{strengthspeed}C--D), then the strength of the variant will be higher because we are holding the observed speed constant (\fref{strengthspeed}E).
-In this case, the same constant-strength intervention can fail to control the epidemic (i.e., $\Ry{post} > 1$; \fref{strengthspeed}E) because this intervention reduces the transmission by a constant amount regardless of age of infection (\fref{strengthspeed}C).
-On the other hand, the same constant-speed intervention will prevent a larger proportion of transmission, leading to lower $\Ry{post}$ (\fref{strengthspeed}E), because it is more effective against late-stage transmission (\fref{strengthspeed}D).
-The constant-speed intervention also reduces the mean generation interval by a larger factor (\fref{strengthspeed}D).
+However, if the variant has longer generation intervals than the wild type (\fref{strengthspeed}E--F), then the strength of the variant will be higher because we are holding the observed speed constant (\fref{strengthspeed}G).
+In this case, the same constant-strength intervention can fail to control the epidemic (i.e., $\Ry{post} > 1$; \fref{strengthspeed}H) because this intervention reduces the transmission by a constant amount regardless of age of infection (\fref{strengthspeed}E).
+On the other hand, the same constant-speed intervention will prevent a larger proportion of transmission, leading to lower $\Ry{post}$ (\fref{strengthspeed}G), because it is more effective against late-stage transmission (\fref{strengthspeed}F).
+The constant-speed intervention also reduces the mean generation interval by a larger factor (\fref{strengthspeed}F).
+Conversely, if the variant has shorter generation intervals than the wild type (\fref{strengthspeed}A--B), the given constant-strength intervention will be relative more effective (\fref{strengthspeed}A) because the given constant-speed intervention prevents less transmission (\fref{strengthspeed}B).
 
@@
 whereas the constant-speed intervention isolates individuals infected with both the wild type and the variant at a constant hazard $\phi$ after day 30.
-If the variant has a longer mean generation interval, the constant-strength intervention fails to suppress the spread of the variant (\fref{control_sim}A) because the longer generations imply a higher initial $\Rw$, and thus a stronger intervention is required to reduce $\Rw$ below 1 (\fref{control_sim}B).
+If the variant has a longer mean generation interval, this specific constant-strength intervention fails to suppress the spread of the variant (\fref{control_sim}A) because the longer generations imply a higher initial $\Rw$, and thus a stronger intervention is required to reduce $\Rw$ below 1 (\fref{control_sim}B).
+Conversely, if the variant has a shorter mean generation interval, its initial $\Rw$ will be lower, and therefore the same constant-strength intervention will be more effective, leading to lower post-intervention $\Rw$ (\fref{control_sim}B).
 The relative strength remains constant under constant-strength intervention (\fref{control_sim}B), but the relative speed changes (\fref{control_sim}C) as discussed earlier in Section 4---in this case, the longer mean generation interval variant increases the relative speed when the intervention is introduced.
@@
 As shown in \fref{strengthspeed}E, a constant-speed intervention leads to greater reduction in $\Rv(t)$ when the mean generation-interval of the variant is longer (\fref{control_sim}E)---
-this implies a change in relative strength $\rho$ because the effectiveness of a constant-speed intervention is sensitive to the generation-interval distribution.
+this implies a change in relative strength $\rho$ because the resulting post-intervention strength under a constant-speed intervention is sensitive to the generation-interval distribution as shown in \fref{strengthspeed}.
 In this case, the relative speed of the variant remains unaffected by the constant-speed intervention because the epidemic speed of the wild type $\rw$ and the variant $\rv$ are both reduced by equal amounts $\phi$, resulting in the post-intervention epidemic speed of $\rw-\phi$ and $\rv-\phi$, respectively.
@@
 
-There is currently a (minor) discrepancy in the estimates of relative strength of new variants, particularly for the alpha variant.
-Mathematical analyses have typically reported greater than a 1.4 fold increase in reproduction number for the alpha variant ($\rho > 1.4$) whereas an independent analysis of secondary attack rates from contact tracing data suggests a 1.25--1.4 fold increase \citep{ukinvest}.
+There is currently a (minor) discrepancy in the estimates of relative strength of new variants, particularly for the Alpha variant.
+Mathematical analyses have typically reported greater than a 1.4 fold increase in reproduction number for the Alpha variant ($\rho > 1.4$) whereas an independent analysis of secondary attack rates from contact tracing data suggests a 1.25--1.4 fold increase \citep{ukinvest}.
 As shown in \cite{davies2021estimated} and \cite{volz2021transmission}, propagating uncertainty in generation interval estimates (and, in particular, assuming shorter generation intervals) can partially explain these differences:
@@
 Instead, this calculation simply highlights the importance of propagating uncertainty in generation-interval distributions in assessing the relative strength and speed of SARS-CoV-2 variants.
+While we relied on parameter estimates for the Alpha variant throughout our analysis, our qualitative conclusions can be applied to studying other SARS-CoV-2 variants, including the Delta variant.
 

### 2021-07-30 control.R

@@ -85,11 +85,60 @@ gendata_long <- data_frame(
   speed=density*exp(-constant_speed*tvec)
 )
 
-Gpost_long <- integrate(function(x) x*Rv_base*dgamma(x, shape=1/kappa, rate=1/kappa/(Gw*tail(Gratio,1)))*exp(-constant_speed*x),
+Gpost_long <- integrate(function(x) x*filter(intervention, Gratio==tail(Gratio,1))$Rv*dgamma(x, shape=1/kappa, rate=1/kappa/(Gw*tail(Gratio,1)))*exp(-constant_speed*x),
                         0,
                         Inf)[[1]]/filter(intervention, Gratio==tail(Gratio,1))$speed_post_strength
 
-g1 <- ggplot(gendata_base) +
+gendata_short <- data_frame(
+  tvec=seq(0, 21, by=0.1),
+  density=filter(intervention, Gratio==head(Gratio,1))$Rv*dgamma(tvec, shape=1/kappa, rate=1/kappa/(Gw*head(Gratio,1))),
+  strength=density/constant_strength,
+  speed=density*exp(-constant_speed*tvec)
+)
+
+Gpost_short <- integrate(function(x) x*filter(intervention, Gratio==head(Gratio,1))$Rv*dgamma(x, shape=1/kappa, rate=1/kappa/(Gw*head(Gratio,1)))*exp(-constant_speed*x),
+                        0,
+                        Inf)[[1]]/filter(intervention, Gratio==head(Gratio,1))$speed_post_strength
+
+g1 <- ggplot(gendata_short) +
+  geom_line(aes(tvec, density, col="Pre-intervention", lty="Pre-intervention"), lwd=3) +
+  geom_line(aes(tvec, strength, col="Post-intervention (constant-strength)", lty="Post-intervention (constant-strength)"), lwd=3) +
+  geom_vline(xintercept = (Gw*head(Gratio,1)), lwd=2) +
+  geom_vline(xintercept = (Gw*head(Gratio,1)), lwd=2, col=colorblind_pal()(3)[3], lty=3) +
+  scale_x_continuous("Generation intervals (days)") +
+  scale_y_continuous("Kernel density", limits=c(0, 0.45)) +
+  scale_color_manual("", values=(colorblind_pal()(3)[c(1, 3)]),
+                     breaks=c("Pre-intervention", "Post-intervention (constant-strength)")) +
+  scale_linetype_manual("", values=c(1, 3),
[…]

### 2021-07-31 response.tex
done

@@
 
-``Various Gamma assumptions have been widely used in epidemic modeling, including for models of SARS-CoV-2 \citep{doi:10.1098/rsif.2020.0144}.
-Although the Gamma assumption includes as a special case models that assume exponentially distributed generation intervals (when $\kappa=1$), corresponding to the SIR model \citep{anderson1991infectious}, it is different from the generation-interval distribution we get when infectious periods are Gamma distributed, another standard assumption in epidemic modeling (see \cite{roberts2007model} for discussion).''
+``Various Gamma-generation-interval assumptions have been widely used in epidemic modeling, including for models of SARS-CoV-2 \citep{doi:10.1098/rsif.2020.0144}.
+The Gamma-generation-interval assumption includes as a special case models that assume exponentially distributed generation intervals (when $\kappa=1$), corresponding to the SIR model \citep{anderson1991infectious}.
+We note that when the infectious periods are Gamma distributed---another standard assumption in epidemic modeling---the resulting generation interval does not follow the Gamma distribution;
+in particular, the mean generation interval can be shorter than the mean infectious period because transmission occurs throughout the infectious period until recovery (see \cite{roberts2007model} for detailed discussion).''
 
@@
 
-``Under constant-strength changes that reduce the transmission rate by a constant amount regardless of the age of infection, the instantaneous generation-interval distribution does not across change time \citep{fraser2007estimating}.
+``Under constant-strength changes that reduce the transmission rate by a constant amount regardless of the age of infection, the instantaneous generation-interval distribution does not change across time \citep{fraser2007estimating}.
 In this case, the instantaneous generation-interval distribution is also often referred to as the intrinsic generation-interval distribution \citep{champredon2015intrinsic,champredon2018two,gostic2020practical,park2020time}---
@@
 
-``For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals leads to a slower epidemic growth (lower $r$), making the epidemic look easier to control with a constant-speed intervention, which isolates infected individuals at a constant hazard and therefore affects late transmission disproportionately---as shown in \cite{doi:10.1098/rspb.2020.1556}, the hazard of isolation must be greater than the epidemic speed $r$ to reach the $r=0$ threshold.
+``For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals leads to a slower epidemic growth (lower $r$), making the epidemic look easier to control with a constant-speed intervention, which isolates infected individuals at a constant hazard and therefore affects late transmission disproportionately.
+As shown in \cite{doi:10.1098/rspb.2020.1556}, the hazard of isolation must be greater than the epidemic speed $r$ to reach the $r=0$ threshold.
 In this same case, however, the uncertainty in the generation-interval distribution is expected to have no effects on the effectiveness of a constant-strength intervention, which reduces the hazard of infectious contacts and therefore affects transmission by an equal amount regardless of age of infection.
-Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals leads to a higher estimate of epidemic strength ($\RR$), making the epidemic look harder to control with a constant-strength interventions (but not with a constant-speed interventions, whose effectiveness is fixed if speed is fixed).
+Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals leads to a higher estimate of epidemic strength ($\RR$), making the epidemic look harder to control with a constant-strength interventions.
 Constant strength and speed are idealized representations of real-life interventions, which can be either strength-like (e.g., vaccination and social distancing) or speed-like (e.g., contact tracing and isolation) depending on how their effectiveness varies across the generation interval (see \cite{doi:10.1098/rspb.2020.1556} and Discussion).''

### 2021-08-03 response.tex
done for now?

@@
 
-``Epidemic speed can be typically estimated from incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}, but epidemic strength is difficult to measure directly.
+``Epidemic speed can be typically estimated from time series of incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}, but epidemic strength is difficult to measure directly.
 Instead, epidemic strength is often inferred from the observed epidemic speed using the generation-interval distribution, an approach popularized by \citep{wallinga2007generation}.
-The generation interval is defined as the time between infection and transmission and provides information about the time scale of individual-level transmission \citep{svensson2007note}.
-The generation interval is also distinctly different from other so-called ``transmission intervals'' that measure time between successive infections---this includes the serial interval, which is defined as the time between symptom onsets in an infector-infectee pair \citep{fine2003interval,grassly2008mathematical,britton2019estimation,ali2020serial,park2021forward}.
+The generation interval, defined as the time between infection and transmission, provides information about the time scale of individual-level transmission \citep{svensson2007note}.
+The generation interval is also distinctly different from other so-called ``transmission intervals'' that measure time between successive infections---including the serial interval, which is defined as the time between symptom onsets in an infector-infectee pair \citep{fine2003interval,grassly2008mathematical,britton2019estimation,ali2020serial,park2021forward}.
 
@@
 The Gamma-generation-interval assumption includes as a special case models that assume exponentially distributed generation intervals (when $\kappa=1$), corresponding to the SIR model \citep{anderson1991infectious}.
-We note that when the infectious periods are Gamma distributed---another standard assumption in epidemic modeling---the resulting generation interval does not follow the Gamma distribution;
+We note that when the infectious periods are Gamma distributed---another standard assumption in epidemic modeling---the resulting generation interval does not necessarily follow the Gamma distribution;
 in particular, the mean generation interval can be shorter than the mean infectious period because transmission occurs throughout the infectious period until recovery (see \cite{roberts2007model} for detailed discussion).''
@@
 
-``Instead of inferring relative strength from speed, one can directly estimate time-varying reproduction numbers $\RR(t)$ of the variant and the wild type from incidence data and directly compare their ratios when the incidence of infection and the generation-interval distributions are known---such methods have been used in previous analyses of the alpha variant by \cite{volz2021transmission}.
+``Instead of inferring relative strength from speed, one can directly estimate time-varying reproduction numbers $\RR(t)$ of the variant and the wild type from incidence data and directly compare their ratios when the incidence of infection and the generation-interval distributions are known---such methods have been used in previous analyses of the Alpha variant by \cite{volz2021transmission}.
 Broadly, there are two types of time-varying reproduction numbers: case reproduction number and instantaneous reproduction number.
@@
 
-``Under constant-strength changes that reduce the transmission rate by a constant amount regardless of the age of infection, the instantaneous generation-interval distribution does not change across time \citep{fraser2007estimating}.
+``Under constant-strength changes that reduce the transmission rate by a constant amount regardless of the age of infection, the instantaneous generation-interval distribution does not change through time \citep{fraser2007estimating}.
 In this case, the instantaneous generation-interval distribution is also often referred to as the intrinsic generation-interval distribution \citep{champredon2015intrinsic,champredon2018two,gostic2020practical,park2020time}---
 for example, the standard SEIR model can be equivalently expressed as a renewal equation with time-invariant intrinsic generation-interval distribution $g(\tau)$ as shown in \citep{champredon2018equivalence}.
-The instantaneous generation-interval distribution is also different from the realized generation-interval distribution, which measures time between actual infection events (see \citep{champredon2015intrinsic}).
+The instantaneous generation-interval distribution is also different from the realized generation-interval distribution, which measures time between actual infection events \citep{champredon2015intrinsic}.
 Previous studies have noted, in many contexts, that the realized generation intervals can contract due to susceptible depletion---a special case of constant-strength changes \citep{kenah2008generation,nishiura2010time,champredon2015intrinsic}---even though the instantaneous generation-interval distribution remains unchanged in this scenario.
-While the instantaneous generation-interval distribution can change under speed-like changes (see \cite{fraser2007estimating} and Section 6--7 for detailed discussions), assuming a time-invariant instantaneous generation-interval distribution is often appropriate in the context of SARS-CoV-2, given that the dynamics of its spread have been primarily driven by strength-like intervention, such as social distancing measures \citep{flaxman2020Rt} and vaccination \citep{moore2021vaccination}.
+While the instantaneous generation-interval distribution can change under speed-like changes (see \cite{fraser2007estimating} and Section 6--7 for detailed discussions), assuming a time-invariant instantaneous generation-interval distribution is often appropriate in the context of SARS-CoV-2, given that control strategies against its spread have been primarily strength-like, including social distancing measures \citep{flaxman2020Rt} and vaccination \citep{moore2021vaccination}.
 Indeed, many dynamical models of SARS-CoV-2 infections have solely relied on constant-strength changes, either implicitly or explicitly assuming a time-invariant intrinsic generation-interval distribution (e.g., \citep{flaxman2020Rt,gostic2020practical,brauner2021inferring}).
@@
 ``Here, we focus on differences in mean generation intervals, assuming that both the squared coefficient of variation and distribution shape of the variant are otherwise the same as for the wild type: $\kappa_{\mathrm{wt}} = \kappa_{\mathrm{var}} = \kappa$.
-Changes in shape can be important \citep{miller2010epidemics,svensson2015influence}, and we do not investigate them here. 
-We do note that, loosely, we expect a distribution with higher coefficient of variation to allow for more early transmission, and thus to have qualitatively similar effects to a distribution with a shorter mean in many cases \citep{park2019practical}.''
+Changes in shape can be important \citep{miller2010epidemics,svensson2015influence}, but we do not investigate them here. 
+We do note that we expect a distribution with higher coefficient of variation to allow for more early transmission, and thus to have qualitatively similar effects to a distribution with a shorter mean \citep{park2019practical}.''
 
@@
 
-``For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals leads to a slower epidemic growth (lower $r$), making the epidemic look easier to control with a constant-speed intervention, which isolates infected individuals at a constant hazard and therefore affects late transmission disproportionately.
+``For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals leads to a slower estimated epidemic growth rate (lower $r$), making the epidemic look easier to control with a constant-speed intervention, which isolates infected individuals at a constant hazard and therefore affects late transmission disproportionately.
 As shown in \cite{doi:10.1098/rspb.2020.1556}, the hazard of isolation must be greater than the epidemic speed $r$ to reach the $r=0$ threshold.
-In this same case, however, the uncertainty in the generation-interval distribution is expected to have no effects on the effectiveness of a constant-strength intervention, which reduces the hazard of infectious contacts and therefore affects transmission by an equal amount regardless of age of infection.
+In this same case, however, variation in the generation-interval distribution has no effects on the effectiveness of a constant-strength intervention, which reduces the hazard of infectious contacts and therefore affects transmission by an equal amount regardless of age of infection.
 Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals leads to a higher estimate of epidemic strength ($\RR$), making the epidemic look harder to control with a constant-strength interventions.
@@
 ``Although these estimates are more consistent with the attack rate analysis \citep{ukinvest},
-we do not claim that they are necessarily more accurate---
-while individual-level data from contact tracing can provide a more reliable source of information about the transmissibility and time scale of transmission in some cases, they can also be biased towards particular types of contacts---for example, household contacts are probably more likely to be identified---which could also affect the estimate of $\rho$.''
+we do not claim that they are necessarily more accurate.
+While individual-level data from contact tracing can provide a more reliable source of information about the transmissibility and time scale of transmission in some cases, they can also be biased towards particular types of contacts---for example, household contacts are probably more likely to be identified---which could also affect the estimate of $\rho$.
+Instead, this calculation simply highlights the importance of carefully considering generation-interval distributions in assessing the relative strength and speed of SARS-CoV-2 variants.''
 

### 2021-08-03 newvariant.tex
done for now?

@@
 
-\usepackage{natbib}
+\usepackage[sort&compress]{natbib}
 \setcitestyle{numbers} 
@@
 
-Inferring the relative strength (i.e., the ratio of reproduction numbers) and relative speed (i.e., the difference between growth rates) of new SARS-CoV-2 variants compared to their wild types is critical to predicting and controlling the course of the current pandemic.
-We explore how differences in the generation-interval distributions affect the inference of relative strength and speed in light of recent evidence that the Alpha and Delta variants may have different infectiousness profiles (and therefore, the generation interval) from prior dominant lineages.
+Inferring the relative strength (i.e., the ratio of reproduction numbers) and relative speed (i.e., the difference between growth rates) of new SARS-CoV-2 variants compared to the wild types, from which they emerged, is critical to predicting and controlling the course of the current pandemic.
+We explore how differences in the generation-interval distributions affect the inference of relative strength and speed in light of recent evidence that the Alpha and Delta variants may have different generation intervals from prior dominant lineages.
 Neglecting these differences gives biased estimates of relative strength, which can vary systematically with changes in the incidence of infections caused by the variant.
-We further argue that holding relative strength constant is generally more appropriate in the context of SARS-CoV-2 and illustrate how relative strength and speed are affected by two idealized interventions.
-A constant-strength intervention, which reduces transmission rates by a constant amount, affects relative speed but not relative strength---this is analogous to vaccination and social distancing.
-A constant-speed intervention, which isolates infected individuals at a constant hazard, affects relative strength but not relative speed---this is analogous to contact tracing.
+We further argue that holding relative strength constant is generally more appropriate for estimating the transmission advantage of new SARS-CoV-2 variants by comparing how relative strength and speed change over time under two idealized interventions of constant strength and speed.
+Constant-strength interventions like idealized vaccination and social distancing, which reduce transmission rates by a constant amount, affect relative speed but not relative strength.
+Constant-speed interventions like idealized contact tracing, which isolate infected individuals at a constant hazard, affect relative strength but not relative speed.
 Changes in the generation-interval distribution can alter the effectiveness of these interventions.
@@
 
-Estimating variant epidemic strength and speed remains a key question in understanding the threat of SARS-CoV-2 variants of concern (VoCs) \citep{switzerland2021variant, davies2021estimated, di2021impact,graham2021changes, leung2021early, volz2021transmission,zhao2021,campbell2021increased}.
+Estimating variant epidemic strength and speed remains a key question in understanding the threat of SARS-CoV-2 variants of concern (VoCs) \cite{switzerland2021variant,davies2021estimated,di2021impact,graham2021changes,leung2021early,volz2021transmission,zhao2021,campbell2021increased}.
 Epidemic ``strength'' is measured by the reproduction number $\RR$---a unitless quantity representing the average number of new infections caused by a typical infection. 
@@
 The epidemic strength also determines the final size of an epidemic in a homogeneous population \citep{anderson1991infectious}.
-Epidemic ``speed'' is characterized by the growth rate $r$, which has units of $1/\mathrm{time}$ and describes the exponential rate of disease spreads at the population level.
+Epidemic ``speed'' is characterized by the growth rate $r$, which has units of $1/\mathrm{time}$ and describes the exponential rate of disease spread at the population level.
 Like epidemic strength, epidemic speed also determines conditions for disease elimination: $r=0$ is a threshold equivalent to $\RR=1$ under constant conditions.
 
-Epidemic speed can be typically estimated from incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}, but epidemic strength is difficult to measure directly.
+Epidemic speed can be typically estimated from time series of incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}, but epidemic strength is difficult to measure directly.
 Instead, epidemic strength is often inferred from the observed epidemic speed using the generation-interval distribution, an approach popularized by \citep{wallinga2007generation}.
-The generation interval is defined as the time between infection and transmission and provides information about the time scale of individual-level transmission \citep{svensson2007note}.
-The generation interval is also distinctly different from other so-called ``transmission intervals'' that measure time between successive infections---this includes the serial interval, which is defined as the time between symptom onsets in an infector-infectee pair \citep{fine2003interval,grassly2008mathematical,britton2019estimation,ali2020serial,park2021forward}.
+The generation interval, defined as the time between infection and transmission, provides information about the time scale of individual-level transmission \citep{svensson2007note}.
+The generation interval is also distinctly different from other so-called ``transmission intervals'' that measure time between successive infections---including the serial interval, which is defined as the time between symptom onsets in an infector-infectee pair \citep{fine2003interval,grassly2008mathematical,britton2019estimation,ali2020serial,park2021forward}.
 
@@
 Instead (or additionally), some studies have assumed a value of the relative strength and tried to predict its relative speed \citep{davies2021estimated,di2021impact}.
-While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the disease and its control because two quantities are linked by the generation-interval distribution \citep{doi:10.1098/rspb.2020.1556}.
+While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the disease and its control because the two quantities are linked by the generation-interval distribution \citep{doi:10.1098/rspb.2020.1556}.
 
-For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals leads to a slower epidemic growth (lower $r$), making the epidemic look easier to control with a constant-speed intervention, which isolates infected individuals at a constant hazard and therefore affects late transmission disproportionately.
+For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals leads to a slower estimated epidemic growth rate (lower $r$), making the epidemic look easier to control with a constant-speed intervention, which isolates infected individuals at a constant hazard and therefore affects late transmission disproportionately.
 As shown in \cite{doi:10.1098/rspb.2020.1556}, the hazard of isolation must be greater than the epidemic speed $r$ to reach the $r=0$ threshold.
-In this same case, however, the uncertainty in the generation-interval distribution is expected to have no effects on the effectiveness of a constant-strength intervention, which reduces the hazard of infectious contacts and therefore affects transmission by an equal amount regardless of age of infection.
+In this same case, however, variation in the generation-interval distribution has no effects on the effectiveness of a constant-strength intervention, which reduces the hazard of infectious contacts and therefore affects transmission by an equal amount regardless of age of infection.
 Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals leads to a higher estimate of epidemic strength ($\RR$), making the epidemic look harder to control with a constant-strength interventions.
@@
 Longer duration of infection suggests that the mean generation interval of the Alpha variant is likely to be longer than that of the wild type \citep{lloyd2001realistic,wearing2005appropriate,roberts2007model}.
-In contrast, some studies have suggested that the more recent Delta variant may have a shorter generation interval \citep{li2021viral, zhang2021transmission}.
+In contrast, some studies have suggested that the more recent Delta variant may have a shorter generation interval \citep{li2021viral,zhang2021transmission}.
 Modeling studies have also considered the possibility that the faster growth rate of new variants may be driven, in part, by shorter generation intervals \citep{davies2021estimated,volz2021transmission}.
-However, linking strength and speed is complicated given that the generation-interval distribution depend on many factors including behavior: 
+However, linking strength and speed is complicated given that the generation-interval distribution depends on many factors including behavior: 
 for example, self-isolation after symptom onset will lead to shorter generation intervals.
@@
 
-Neglecting the (relatively slow) rate of new mutations and assuming homogeneous mixing, the current incidence of infection $i_x(t)$ caused by each strain $x$---either the wild type (``wt'') or the variant (``var'')---can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary infections are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
+Neglecting the (relatively slow) rate of new mutations and assuming homogeneous mixing, the current incidence of infection $i_x(t)$ caused by each strain $x$---either the wild type (``wt'') or the variant (``var'')---can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate $K_x(t, \tau)$ at which secondary infections are generated at time $t$ by individuals infected $\tau$ time units ago:
 \begin{equation}
@@
 \end{equation}
-which is defined as secondary infections that would be caused by an individual infected at time $t$ if conditions were to remain the same \citep{fraser2007estimating}.
+which is defined as the expected number of secondary infections that would be caused by an individual infected at time $t$ if conditions were to remain the same \citep{fraser2007estimating}.
 The instantaneous reproduction number is a particular kind of weighted average of infectiousness of previously infected individuals at time $t$---in particular, it is weighted by the total relative infectiousness at time $t$, rather than by the actual number of infected individuals present.
@@
 
-Under constant-strength changes that reduce the transmission rate by a constant amount regardless of the age of infection, the instantaneous generation-interval distribution does not change across time \citep{fraser2007estimating}.
+Under constant-strength changes that reduce the transmission rate by a constant amount regardless of the age of infection, the instantaneous generation-interval distribution does not change through time \citep{fraser2007estimating}.
 In this case, the instantaneous generation-interval distribution is also often referred to as the intrinsic generation-interval distribution \citep{champredon2015intrinsic,champredon2018two,gostic2020practical,park2020time}---
 for example, the standard SEIR model can be equivalently expressed as a renewal equation with time-invariant intrinsic generation-interval distribution $g(\tau)$ as shown in \citep{champredon2018equivalence}.
-The instantaneous generation-interval distribution is also different from the realized generation-interval distribution, which measures time between actual infection events (see \citep{champredon2015intrinsic}).
+The instantaneous generation-interval distribution is also different from the realized generation-interval distribution, which measures time between actual infection events \citep{champredon2015intrinsic}.
 Previous studies have noted, in many contexts, that the realized generation intervals can contract due to susceptible depletion---a special case of constant-strength changes \citep{kenah2008generation,nishiura2010time,champredon2015intrinsic}---even though the instantaneous generation-interval distribution remains unchanged in this scenario.
-While the instantaneous generation-interval distribution can change under speed-like changes (see \cite{fraser2007estimating} and Section 6--7 for detailed discussions), assuming a time-invariant instantaneous generation-interval distribution is often appropriate in the context of SARS-CoV-2, given that the dynamics of its spread have been primarily driven by strength-like intervention, such as social distancing measures \citep{flaxman2020Rt} and vaccination \citep{moore2021vaccination}.
+While the instantaneous generation-interval distribution can change under speed-like changes (see \cite{fraser2007estimating} and Section 6--7 for detailed discussions), assuming a time-invariant instantaneous generation-interval distribution is often appropriate in the context of SARS-CoV-2, given that control strategies against its spread have been primarily strength-like, including social distancing measures \citep{flaxman2020Rt} and vaccination \citep{moore2021vaccination}.
 Indeed, many dynamical models of SARS-CoV-2 infections have solely relied on constant-strength changes, either implicitly or explicitly assuming a time-invariant intrinsic generation-interval distribution (e.g., \citep{flaxman2020Rt,gostic2020practical,brauner2021inferring}).
@@
 The Gamma-generation-interval assumption includes as a special case models that assume exponentially distributed generation intervals (when $\kappa=1$), corresponding to the SIR model \citep{anderson1991infectious}.
-We note that when the infectious periods are Gamma distributed---another standard assumption in epidemic modeling---the resulting generation interval does not follow the Gamma distribution;
+We note that when the infectious periods are Gamma distributed---another standard assumption in epidemic modeling---the resulting generation interval does not necessarily follow the Gamma distribution;
 in particular, the mean generation interval can be shorter than the mean infectious period because transmission occurs throughout the infectious period until recovery (see \cite{roberts2007model} for detailed discussion).
@@
 Here, we focus on differences in mean generation intervals, assuming that both the squared coefficient of variation and distribution shape of the variant are otherwise the same as for the wild type: $\kappa_{\mathrm{wt}} = \kappa_{\mathrm{var}} = \kappa$.
-Changes in shape can be important \citep{miller2010epidemics,svensson2015influence}, and we do not investigate them here. 
-We do note that we expect a distribution with higher coefficient of variation to allow for more early transmission, and thus to have qualitatively similar effects to a distribution with a shorter mean in many cases \citep{park2019practical}.
+Changes in shape can be important \citep{miller2010epidemics,svensson2015influence}, but we do not investigate them here. 
+We do note that we expect a distribution with higher coefficient of variation to allow for more early transmission, and thus to have qualitatively similar effects to a distribution with a shorter mean \citep{park2019practical}.
 
@@
 studies of new SARS-CoV-2 variants have mostly focused on characterizing changes in the \emph{proportion} of a new variant \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
-Focusing on proportions has the advantage, because changes in proportions are less sensitive to changes in testing and to other transient effects that would affect variants and wild type viruses similarly.
+Focusing on proportions is advantageous, because changes in proportions are less sensitive to changes in testing and to other transient effects that would affect variants and wild type viruses similarly.
 When incidence is changing exponentially ($i_x(t) = i_x(t_0) \exp(r_x t)$), the proportion of the new variant $p(t)$ follows a logistic growth curve \citep{switzerland2021variant,davies2021estimated}:
@@
 \end{align}
-where the relative speed of the variant $\delta = \rv - \rw$ can be estimated as the slope of the log odds of $p$  vs.~time.
+where the relative speed of the variant $\delta$ can be estimated as the slope of the log odds of $p$  vs.~time.
 When more than two strains are co-circulating, the picture is more complicated \citep{campbell2021increased}; we focus here on comparing two strains at a time.
@@
 \textbf{Relative strength of the new variant assuming a fixed speed advantage $\delta$ under five epidemiological conditions.}
-The relative strength of the new variant $\rho$ conditional on the speed of the wild type $\rw$; the ratio between the mean generation interval of the new variant $\Gv$ and that of the wild type $\Gw$; and the squared coefficient of variation in generation intervals $\kappa$.
+The relative strength of the new variant $\rho$ (shown on a log scale) conditional on the speed of the wild type $\rw$; the ratio between the mean generation interval of the new variant $\Gv$ and that of the wild type $\Gw$; and the squared coefficient of variation in generation intervals $\kappa$.
 The relative strength of the new variant $\rho$ is calculated using $\delta=0.1\pday$, $\Gw = 5\days$, and $\kappa = 1/5$.
-Assumed values of $\rw$ (and therefore $\rv$) are shown in the top right corners of each panel.
+Assumed values of $\rw$ and $\rv$ are shown in the top right corners of each panel.
 }
@@
 
-We start by taking the mean generation interval of the wild type to be $\Gw = 5\days$ \citep{ferretti2020quantifying}; and the squared coefficient of variation of generation intervals to be $\kappa=0.2$ \citep{ferretti2020quantifying} for both the variant and the wild type.
+We start by taking the mean generation interval of the wild type to be $\Gw = 5\days$ \citep{ferretti2020quantifying} and the squared coefficient of variation of generation intervals to be $\kappa=0.2$ \citep{ferretti2020quantifying} for both the variant and the wild type.
 As noted above, we assume throughout that the variant and the wild type can be approximated with gamma distributions with equal $\kappa$, and only consider differences in the mean.
-We evaluate the estimates of relative strength $\rho$ across a wide range of $\kappa$ from 0 (fixed-length generation intervals) to 1 (exponential distribution).
+We evaluate the estimates of relative strength $\rho$ across a wide range of $\kappa$ from 0 (fixed-length generation intervals) to 1 (exponential distributions).
 To further explore how inference depends on underlying epidemiological conditions, we consider five scenarios: (1) $\rw < \rv < 0$, (2) $\rw < \rv = 0$, (3) $\rw < 0 < \rv$, (4) $0 = \rw < \rv$, and (5) $0 < \rw < \rv$.
@@
 However, the magnitude of relative strength $\rho$ is sensitive to assumptions about generation intervals:
-For realistic values of $\kappa$ (excluding 0 and 1), the inferred relative strength $\rho$ ranges between 1.1--2.3 when $\Gv$ is allowed to vary between $2\Gw/3$ and $3\Gw/2$.
+For realistic values of $\kappa$ (excluding 0 and 1), the inferred relative strength $\rho$ ranges between 1.1 and 2.3 when $\Gv$ is allowed to vary between $2\Gw/3$ and $3\Gw/2$.
 
@@
 For example, if the proportion of the population susceptible declines, or the average contact rate changes, while other factors remain constant, the relative strength $\rho$ is expected to remain constant \citep{leung2017monitoring,leung2020empirical,di2021impact,leung2021early};
-as we show in this section, this will imply a change in relative speed $\delta$.
+as we show in this section, these changes will imply a change in relative speed $\delta$.
 
@@
 
-\fref{relspeed} also shows that when $\rho$ is fixed relative speed depends on underlying epidemiological conditions---specifically, the absolute strength of the two strains. 
-For example, even when there are no differences in the generation-interval distributions ($\Gv=\Gw$, in this case), changing $\Rw$ from 0.49 to 1.27 (and thus $\Rv$ from 0.79 to 2.04) yields the relative speed $\delta$ between 0.08--0.11 for $\kappa=0.2$ and 0.06--0.14 for $\kappa=0.5$.
+\fref{relspeed} also shows that when $\rho$ is fixed, relative speed depends on underlying epidemiological conditions---specifically, the absolute strength of the two strains. 
+For example, even when the generation-interval distributions are identical ($\Gv=\Gw$, in this case), changing $\Rw$ from 0.49 to 1.27 (and thus $\Rv$ from 0.79 to 2.04) yields a relative speed $\delta$ between 0.08--0.11 for $\kappa=0.2$ and 0.06--0.14 for $\kappa=0.5$.
 Therefore, characterizing the spread of variants by assuming constant relative speed (e.g., by fitting a standard logistic growth curve \citep{switzerland2021variant} and using the resulting value of $\delta$) should be done with care.
@@
 Observed case counts are sensitive to reporting delays \citep{goldstein2009reconstructing} and changes in case definitions \citep{tsang2020effect}, which in turn can affect estimates of the instantaneous reproduction number \citep{gostic2020practical}.
-Here, we choose to focus on the underlying dynamical mechanisms that may affect inference and thus assume that the incidence of infection is exactly known.
+Here, we choose to focus on the underlying dynamical mechanisms that may affect inference and thus assume that the incidence of infection is known exactly.
 Assuming that the instantaneous generation-interval distribution remains constant, the instantaneous reproduction numbers of the new variant and of the wild type can be estimated from their corresponding incidence curves---an approach popularized by \cite{cori2013new}:
@@
 Under constant-strength intervention measures that reduce transmission rates of both strains by a constant amount, we expect ratios between reproduction numbers to remain constant and correspond to the true relative strength: $\Rv(t)/\Rw(t) = \rho$ \citep{leung2017monitoring,leung2020empirical,di2021impact,leung2021early}.
-However, if the assumed generation-interval distribution $\hat{g}(\tau)$ is different from the true distribution, then the ratio between the estimated reproduction numbers $\hat{\rho}(t) = \hat{\RR}_{\textrm{var}}(t)/\hat{\RR}_{\textrm{wt}}(t)$ may change, even if the true ratio does not.
+However, if the assumed generation-interval distribution $\hat{g}(\tau)$ differs from the true distribution, then the ratio between the estimated reproduction numbers $\hat{\rho}(t) = \hat{\RR}_{\textrm{var}}(t)/\hat{\RR}_{\textrm{wt}}(t)$ may change, even if the true ratio does not.
 
@@
 Then, we compare the estimated ratio $\hat{\rho}(t) = \hat{\RR}_{\textrm{var}}(t)/\hat{\RR}_{\textrm{wt}}(t)$ with the true ratio $\rho = \Rv(t)/\Rw(t)$.
-In order to simulate introduction and lifting of non-pharmaceutical interventions, we let $\Rw(t)$ to decrease from 2 to 0.4 around day 30 and increase back up to 1 around day 60 and assume $\Rv(t) = \rho \Rw(t)$.
+In order to simulate introduction and lifting of non-pharmaceutical interventions, we let $\Rw(t)$ decrease from 2 to 0.4 around day 30 and increase back up to 1 around day 60 and assume $\Rv(t) = \rho \Rw(t)$.
 Previous studies have modeled the impact of non-pharmaceutical interventions as a step function \citep{flaxman2020Rt}, but we use a smooth function to model $\Rw(t)$ (\fref{Rtbias}; see Supplementary Text) given the possibility that behavioral changes may affect transmission before and after interventions take place.
@@
 However, when the assumed mean generation interval is longer than the that of the variant, we over-estimate the slope (and conversely, \fref{Rtbias}C,I).
-Biases in slopes further translate to biases in intercepts; in theory, we would expect the regression line to go through the origin (because $\Rv = 0$ when $\Rw = 0$).
+Biases in slopes further translate to biases in intercepts; in theory, the regression line should go through the origin (because $\Rv = 0$ when $\Rw = 0$).
 
@@
 The constant-speed intervention also reduces the mean generation interval by a larger factor (\fref{strengthspeed}F).
-Conversely, if the variant has shorter generation intervals than the wild type (\fref{strengthspeed}A--B), the given constant-strength intervention will be relative more effective (\fref{strengthspeed}A) because the given constant-speed intervention prevents less transmission (\fref{strengthspeed}B).
+Conversely, if the variant has shorter generation intervals than the wild type (\fref{strengthspeed}A--B), the given constant-strength intervention will be relatively more effective (\fref{strengthspeed}A) because the given constant-speed intervention prevents less transmission (\fref{strengthspeed}B).
 
@@
 As before, we consider a scenario in which a new variant is emerging with known relative speed($\delta=\rv > 0.1\pday$ and $\Rw > 1$) while the incidence of infections caused by the wild type is constant ($\rw = 0\pday$ and $\Rw =1$) before interventions are reduced.
-The constant-strength intervention then reduces both $\Rw$ and $\Rv$ by a constant factor $\theta$ beginning day 30, 
+The constant-strength intervention then reduces both $\Rw$ and $\Rv$ by a constant factor $\theta$ beginning on day 30, 
 whereas the constant-speed intervention isolates individuals infected with both the wild type and the variant at a constant hazard $\phi$ after day 30.
@@
 
-In contrast, epidemic trajectories under the constant-speed intervention look identical regardless of the mean generation interval of the variant (\fref{control_sim}D). 
+In contrast, epidemic trajectories under the constant-speed intervention behave identically regardless of the mean generation interval of the variant (\fref{control_sim}D). 
 As shown in \fref{strengthspeed}E, a constant-speed intervention leads to greater reduction in $\Rv(t)$ when the mean generation-interval of the variant is longer (\fref{control_sim}E)---
@@
 As new variants of SARS-CoV-2 are spreading and becoming dominant in many countries, it is clear that the variants are more transmissible (have higher strength) than the wild type \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
-While our analysis supports estimates of a higher strength of the new variant across a wide range of assumptions, it also shows that uncertainty in generation-interval distributions must be taken into account to obtain accurate estimates of the relative strength of variants.
-If new variants have a longer infectious period \citep{kissler2021densely}, and therefore extended generation-interval distributions, current estimates may underestimate the true relative strength.
+While our analysis supports estimates of a higher strength of the new variant across a wide range of assumptions, it also shows that uncertainty in generation-interval distributions must be taken into account to obtain accurate estimates of the relative strength of variants (and the corresponding uncertainty).
+If new variants have different infectiousness profiles \citep{kissler2021densely,li2021viral,zhang2021transmission}, and therefore generation-interval distributions, current estimates that do not account for different generation-interval distributions may be systematically biased.
 % If new variants escape vaccinal or natural immunity to some extent \citep{Kupferschmidt329}, their relative strength may increase further.
@@
 Although these estimates are more consistent with the attack rate analysis \citep{ukinvest},
-we do not claim that they are necessarily more accurate---
-while individual-level data from contact tracing can provide a more reliable source of information about the transmissibility and time scale of transmission in some cases, they can also be biased towards particular types of contacts---for example, household contacts are probably more likely to be identified---which could also affect the estimate of $\rho$.
-Instead, this calculation simply highlights the importance of propagating uncertainty in generation-interval distributions in assessing the relative strength and speed of SARS-CoV-2 variants.
+we do not claim that they are necessarily more accurate.
+While individual-level data from contact tracing can provide a more reliable source of information about the transmissibility and time scale of transmission in some cases, they can also be biased towards particular types of contacts---for example, household contacts are probably more likely to be identified---which could also affect the estimate of $\rho$.
+Instead, this calculation simply highlights the importance of carefully considering generation-interval distributions in assessing the relative strength and speed of SARS-CoV-2 variants.
 While we relied on parameter estimates for the Alpha variant throughout our analysis, our qualitative conclusions can be applied to studying other SARS-CoV-2 variants, including the Delta variant.
@@
 however, this assumption only holds under strength-like interventions, which are insensitive to time since infection, but not under speed-like interventions. 
-As we demonstrated, if the variant and the wild type have different generation intervals, speed-like interventions such as contact tracing can affect them differently, causing the relative strength to change over time (but not necessarily the relative speed).
+As we demonstrated, if the variant and the wild type have different generation intervals, speed-like interventions such as contact tracing can affect them differently, causing their relative strength to change over time (but not necessarily their relative speed).
 
@@
 Analyses of SARS-CoV-2 dynamics have primarily relied on the constant-strength framework \citep{gostic2020practical,unwin2020state,brauner2021inferring}, even when modeling speed-like interventions such as self-isolation \citep{flaxman2020Rt,brett2020transmission}.
-While the constant-strength framework provides a convenient tool for modeling and understanding epidemic dynamics, we encourage researchers to consider both strength- and speed-based perspectives as they can lead to different conclusions.
+While the constant-strength framework provides a convenient tool for modeling and understanding epidemic dynamics, we encourage researchers to consider both strength- and speed-based perspectives, as they can lead to different conclusions.
 
@@
 Early in the spread of an emerging variant, it is likely more convenient to fix the relative speed, given that speed can be directly observed.
-However, epidemiological conditions \textit{will} change over time in response to spreading new variants---in the context of SARS-CoV-2 infections, most of these responses will be strength-like, causing relative speed to change and invalidating this assumption.
+However, epidemiological conditions \textit{will} change over time in response to spreading new variants---in the context of SARS-CoV-2 infections, most of these responses (e.g., vaccination and social distancing) have been strength-like, causing relative speed to change and invalidating this assumption.
 As more information about the transmission and immunity profiles of the new variant becomes available, we advise instead fixing the relative strength and inferring the speed, as this assumption better matches biological mechanisms for the variants' higher strength (e.g., higher rates of transmission and immune evasion).
@@
 Even though SARS-CoV-2 has been spreading for more than a year, there is still considerable uncertainty about its generation intervals.
-A few studies have tried to estimate the generation-interval distribution, with means ranging between 3--6 days and squared coefficients of variation ranging between 0.1--1 \citep{ferretti2020quantifying,Ferretti2020timing,ganyani2020estimating,knight2020estimating}.
+Several studies have tried to estimate the generation-interval distribution, with means ranging between 3--6 days and squared coefficients of variation ranging between 0.1--1 \citep{ferretti2020quantifying,Ferretti2020timing,ganyani2020estimating,knight2020estimating}.
 However, these estimates are derived from serial intervals (i.e., time between symptom onset of the infector and the infectee; \cite{svensson2007note}), which are subject to dynamical biases \citep{park2021forward} and fail to account for asymptomatic transmission, adding further uncertainty to inferences of speed and strength \citep{park2020time}.
@@
 The spread of new SARS-CoV-2 variants and the replacement of previously dominant lineages represent ongoing challenges for controlling the SARS-CoV-2 pandemic \citep{abdool2021new,fontanet2021sars,walensky2021sars}.
-By explicitly considering epidemiological context and generation interval differences together, we have shown that improving estimates of the the relative duration of infectiousness at the individual scale may represent a pathway towards more effective interventions.
+By explicitly considering epidemiological context and generation interval differences together, we have shown that improving estimates of the the relative duration of infectiousness at the individual scale may help guide more effective interventions.
 Specifically, speed-like interventions, such as contact tracing, will be relatively more effective if variants have longer generation intervals.

### 2021-08-03 newvariant.tex
edits

@@
 Constant-strength interventions like idealized vaccination and social distancing, which reduce transmission rates by a constant amount, affect relative speed but not relative strength.
-Constant-speed interventions like idealized contact tracing, which isolate infected individuals at a constant hazard, affect relative strength but not relative speed.
+Constant-speed interventions like idealized contact tracing, which isolate infected individuals at a constant rate, affect relative strength but not relative speed.
 Changes in the generation-interval distribution can alter the effectiveness of these interventions.
@@
 The epidemic strength also determines the final size of an epidemic in a homogeneous population \citep{anderson1991infectious}.
-Epidemic ``speed'' is characterized by the growth rate $r$, which has units of $1/\mathrm{time}$ and describes the exponential rate of disease spread at the population level.
+Epidemic ``speed'' is characterized by the growth rate $r$, which has units of $1/\mathrm{time}$ and describes the exponential rate of pathogen spread at the population level.
 Like epidemic strength, epidemic speed also determines conditions for disease elimination: $r=0$ is a threshold equivalent to $\RR=1$ under constant conditions.
 
-Epidemic speed can be typically estimated from time series of incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}, but epidemic strength is difficult to measure directly.
+Epidemic speed can be typically estimated from time series of incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}, but epidemic strength is difficult to measure from incidence time series.
 Instead, epidemic strength is often inferred from the observed epidemic speed using the generation-interval distribution, an approach popularized by \citep{wallinga2007generation}.
@@
 
-The exact shape of the distribution depends on several factors---including the shape of latent and infectious periods \citep{lloyd2001realistic,wearing2005appropriate,roberts2007model} as well as more detailed life history of a disease \citep{huber2016quantitative}---and thus can be difficult to estimate.
-While it is possible to consider general forms of generation-interval distributions \citep{miller2010epidemics,svensson2015influence}, summarizing the distribution in terms of its mean and variability---for example, by using the Gamma distribution---can still provide a robust link between epidemic speed and strength for real diseases and yield important biological insights \citep{park2019practical}.
+The exact shape of the distribution depends on several factors---including the shape of latent and infectious period distributions \citep{lloyd2001realistic,wearing2005appropriate,roberts2007model} as well as more detailed life history of a disease \citep{huber2016quantitative}---and thus can be difficult to estimate.
+While it is possible to consider general forms of generation-interval distributions \citep{miller2010epidemics,svensson2015influence}, summarizing the distribution in terms of its mean and variability---for example, by assuming they are Gamma distributed---can still provide a robust link between epidemic speed and strength for real diseases and yield important biological insights \citep{park2019practical}.
 In particular, several studies have noted, in many contexts, that mechanisms that increase the mean generation interval increase epidemic strength $\RR$ given observed epidemic speed $r$ \citep{eaton2014proportion,powers2014impact,weitz2015modeling,park2020time}.
@@
 
-For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals leads to a slower estimated epidemic growth rate (lower $r$), making the epidemic look easier to control with a constant-speed intervention, which isolates infected individuals at a constant hazard and therefore affects late transmission disproportionately.
+For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals results in a slower estimated epidemic growth rate (lower $r$), making an epidemic look easier to control with a constant-speed intervention, which isolates infected individuals at a constant hazard---such intervention affects late transmission disproportionately and further affect the post-intervention generation-interval distribution.
 As shown in \cite{doi:10.1098/rspb.2020.1556}, the hazard of isolation must be greater than the epidemic speed $r$ to reach the $r=0$ threshold.
 In this same case, however, variation in the generation-interval distribution has no effects on the effectiveness of a constant-strength intervention, which reduces the hazard of infectious contacts and therefore affects transmission by an equal amount regardless of age of infection.
-Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals leads to a higher estimate of epidemic strength ($\RR$), making the epidemic look harder to control with a constant-strength interventions.
-Constant strength and speed are idealized representations of real-life interventions, which can be either strength-like (e.g., vaccination and social distancing) or speed-like (e.g., contact tracing and isolation) depending on how their effectiveness varies across the generation interval (see \cite{doi:10.1098/rspb.2020.1556} and Discussion).
+Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals leads to a higher estimate of epidemic strength ($\RR$), making the epidemic look harder to control with a constant-strength intervention.
+Constant strength and speed interventions are idealized representations of real-life interventions, which can be either strength-like (e.g., vaccination and social distancing) or speed-like (e.g., contact tracing and isolation) depending on how their effectiveness varies across the generation interval (see \cite{doi:10.1098/rspb.2020.1556} and Discussion).
 
@@
 We focus on characterizing the incidence of infection, which is directly related to $r$ and $\RR$.
-In practice, observed case reports are subject to reporting delays, which must be taken into account in order to correctly infer $r$ and $\RR$ \citep{goldstein2009reconstructing,gostic2020practical}.
+In practice, observed case reports are subject to reporting delays as well as changes in testing behaviors or capacity, which must be taken into account in order to correctly infer $r$ and $\RR$ \citep{goldstein2009reconstructing,gostic2020practical}.
 
@@
 Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$, in which case the incidence of infections changes exponentially.
-Here, we use the term ``epidemiological conditions'' to broadly refer to all factors that affect transmission---mathematically, they are captured by the kernel $K(t, \tau)$.
-In the context of SARS-CoV-2 infections, we are essentially assuming that the changes in susceptible pool, behavior, and contact rates are usually small over a short period of time.
+Here, we use the term ``epidemiological conditions'' to broadly refer to all factors that affect transmission at the population-level---mathematically, they are captured by the kernel $K(t, \tau)$.
+In the context of SARS-CoV-2 infections, we are essentially assuming that the changes in susceptible pool, behavior, and contact rates are usually small over a short period of time---for example, this assumption would not apply at the times of drastic policy changes but can still work before and after policy changes have been made.
 Then, the incidence of each strain grows (or decays) exponentially at rate $r_x$, satisfying the Euler-Lotka equation \citep{wallinga2007generation}:
@@
 We start by taking the mean generation interval of the wild type to be $\Gw = 5\days$ \citep{ferretti2020quantifying} and the squared coefficient of variation of generation intervals to be $\kappa=0.2$ \citep{ferretti2020quantifying} for both the variant and the wild type.
-As noted above, we assume throughout that the variant and the wild type can be approximated with gamma distributions with equal $\kappa$, and only consider differences in the mean.
+As noted above, we assume throughout that the variant and the wild type can be approximated with Gamma distributions with equal $\kappa$, and only consider differences in the mean.
 We evaluate the estimates of relative strength $\rho$ across a wide range of $\kappa$ from 0 (fixed-length generation intervals) to 1 (exponential distributions).
@@
 Here, we choose to focus on the underlying dynamical mechanisms that may affect inference and thus assume that the incidence of infection is known exactly.
-Assuming that the instantaneous generation-interval distribution remains constant, the instantaneous reproduction numbers of the new variant and of the wild type can be estimated from their corresponding incidence curves---an approach popularized by \cite{cori2013new}:
+Assuming that the instantaneous generation-interval distribution remains constant, the instantaneous reproduction numbers of the new variant and of the wild type can be estimated from their corresponding incidence curves \cite{cori2013new}:
 \begin{equation}
@@
 \textbf{Effects of constant-strength and constant-speed interventions on the spread of a new variant with known speed \rv.}
-(A--F) Pre-intervention (black) and post-intervention (colored) kernel of the new variant under constant-strength (A,C,E) and -speed (B,D,F) interventions when the variant has shorter (A--B), equal (C--D), or longer (E--F) mean generation interval than the wild type.
+(A--F) Pre-intervention (black) and post-intervention (colored) kernel of the new variant under constant-strength (A,C,E) and -speed (B,D,F) interventions when the variant has shorter (A--B), equal (C--D), or longer (E--F) mean generation interval (GI) than the wild type.
 Vertical lines represent the mean generation interval.
@@
 Finally, we synthesize our findings and illustrate the differences between constant-strength and constant-speed interventions using epidemic simulations (\fref{control_sim}).
-As before, we consider a scenario in which a new variant is emerging with known relative speed($\delta=\rv > 0.1\pday$ and $\Rw > 1$) while the incidence of infections caused by the wild type is constant ($\rw = 0\pday$ and $\Rw =1$) before interventions are reduced.
+As before, we consider a scenario in which a new variant is emerging with known relative speed ($\delta=\rv > 0.1\pday$ and $\Rw > 1$) while the incidence of infections caused by the wild type is constant ($\rw = 0\pday$ and $\Rw =1$) before interventions are reduced.
 The constant-strength intervention then reduces both $\Rw$ and $\Rv$ by a constant factor $\theta$ beginning on day 30, 

### 2021-08-03 Rtbias.R

@@ -57,12 +57,14 @@ for (i in 1:length(svec)) {
     geom_line(aes(tvec, Rt2, color="Variant", lty="True"), lwd=2) +
     geom_line(aes(tvec, Rtest1, color="Wild type", lty="Estimated"), lwd=2) +
     geom_line(aes(tvec, Rtest2, color="Variant", lty="Estimated"), lwd=2) +
+    annotate(geom="text",label=LETTERS[(i-1)*3+1],x=-Inf,y=Inf,
+             size=5,
+             vjust=1.5,hjust=-0.5) +
     scale_x_continuous("Time (days)", expand=c(0, 0), limits=c(0, 70)) +
     scale_y_log10("Reproduction number, $\\mathcal{R}(t)$", limits=c(0.39, 8),
                   breaks=c(0.25, 0.5, 1, 2, 4, 8)) +
     scale_color_manual(values=c("red", "black")) +
     scale_linetype_manual(values=c(2, 1), guide=FALSE) +
-    ggtitle(LETTERS[(i-1)*3+1]) +
     coord_fixed() +
     theme(
       panel.grid = element_blank(),
@@ -80,12 +82,14 @@ for (i in 1:length(svec)) {
       geom_line(aes(tvec, Rt2, col="Variant", lty="Variant"), lwd=2) +
       geom_line(aes(tvec, Rtest1, lty="Estimated"), col="black", lwd=2) +
       geom_line(aes(tvec, Rtest2, col="Estimated", lty="Estimated"), lwd=2) +
+      annotate(geom="text",label=LETTERS[(i-1)*3+1],x=-Inf,y=Inf,
+               size=5,
+               vjust=1.5,hjust=-0.5) +
       scale_x_continuous("Time (days)", expand=c(0, 0), limits=c(0, 70)) +
       scale_y_log10("Reproduction number, $\\mathcal{R}(t)$", limits=c(0.39, 8),
                     breaks=c(0.25, 0.5, 1, 2, 4, 8)) +
       scale_color_manual("a", values=c("red", "red", "black")) +
       scale_linetype_manual("a", values=c(2, 1, 1)) +
-      ggtitle(LETTERS[(i-1)*3+1]) +
       coord_fixed() +
       theme(
         panel.grid = element_blank(),
@@ -98,11 +102,13 @@ for (i in 1:length(svec)) {
   g2 <- ggplot(rr) +
[…]

### 2021-08-03 Rtbias_smooth.R

@@ -52,12 +52,14 @@ for (i in 1:length(svec)) {
     geom_line(aes(tvec, Rt2, color="Variant", lty="True"), lwd=2) +
     geom_line(aes(tvec, Rtest1, color="Wild type", lty="Estimated"), lwd=2) +
     geom_line(aes(tvec, Rtest2, color="Variant", lty="Estimated"), lwd=2) +
-    scale_x_continuous("Time (days)", expand=c(0, 0), limits=c(15, 70)) +
+    annotate(geom="text",label=LETTERS[(i-1)*3+1],x=-Inf,y=Inf,
+             size=5,
+             vjust=1.5,hjust=-0.5) +
+    scale_x_continuous("Time (days)", expand=c(0, 0), limits=c(0, 70)) +
     scale_y_log10("Reproduction number, $\\mathcal{R}(t)$", limits=c(0.4, 8),
                   breaks=c(0.25, 0.5, 1, 2, 4, 8)) +
     scale_color_manual(values=c("red", "black")) +
     scale_linetype_manual(values=c(2, 1), guide=FALSE) +
-    ggtitle(LETTERS[(i-1)*3+1]) +
     coord_fixed() +
     theme(
       panel.grid = element_blank(),
@@ -75,12 +77,14 @@ for (i in 1:length(svec)) {
       geom_line(aes(tvec, Rt2, col="Variant", lty="Variant"), lwd=2) +
       geom_line(aes(tvec, Rtest1, lty="Estimated"), col="black", lwd=2) +
       geom_line(aes(tvec, Rtest2, col="Estimated", lty="Estimated"), lwd=2) +
+      annotate(geom="text",label=LETTERS[(i-1)*3+1],x=-Inf,y=Inf,
+               size=5,
+               vjust=1.5,hjust=-0.5) +
       scale_x_continuous("Time (days)", expand=c(0, 0), limits=c(0, 70)) +
       scale_y_log10("Reproduction number, $\\mathcal{R}(t)$", limits=c(0.4, 8),
                     breaks=c(0.25, 0.5, 1, 2, 4, 8)) +
       scale_color_manual("a", values=c("red", "red", "black")) +
       scale_linetype_manual("a", values=c(2, 1, 1)) +
-      ggtitle(LETTERS[(i-1)*3+1]) +
       coord_fixed() +
       theme(
         panel.grid = element_blank(),
[…]

### 2021-08-03 control.R

@@ -1,5 +1,5 @@
 library(dplyr)
-library(ggplot2); theme_set(theme_bw(base_size=18, base_family = "Times"))
+library(ggplot2); theme_set(theme_bw(base_size=16, base_family = "Times"))
 library(ggthemes)
 library(egg)
 library(tikzDevice)
@@ -102,16 +102,16 @@ Gpost_short <- integrate(function(x) x*filter(intervention, Gratio==head(Gratio,
 
 g1 <- ggplot(gendata_short) +
   geom_line(aes(tvec, density, col="Pre-intervention", lty="Pre-intervention"), lwd=3) +
-  geom_line(aes(tvec, strength, col="Post-intervention (constant-strength)", lty="Post-intervention (constant-strength)"), lwd=3) +
+  geom_line(aes(tvec, strength, col="Post-intervention", lty="Post-intervention"), lwd=3) +
   geom_vline(xintercept = (Gw*head(Gratio,1)), lwd=2) +
   geom_vline(xintercept = (Gw*head(Gratio,1)), lwd=2, col=colorblind_pal()(3)[3], lty=3) +
   scale_x_continuous("Generation intervals (days)") +
   scale_y_continuous("Kernel density", limits=c(0, 0.45)) +
   scale_color_manual("", values=(colorblind_pal()(3)[c(1, 3)]),
-                     breaks=c("Pre-intervention", "Post-intervention (constant-strength)")) +
+                     breaks=c("Pre-intervention", "Post-intervention")) +
   scale_linetype_manual("", values=c(1, 3),
-                        breaks=c("Pre-intervention", "Post-intervention (constant-strength)")) +
-  ggtitle("A. Shorter generation interval") +
+                        breaks=c("Pre-intervention", "Post-intervention")) +
+  ggtitle("A. Shorter GI, Constant-strength intervention") +
   theme(
     panel.grid = element_blank(),
     legend.position = c(0.6, 0.95),
@@ -121,16 +121,16 @@ g1 <- ggplot(gendata_short) +
 
 g2 <- ggplot(gendata_short) +
   geom_line(aes(tvec, density, col="Pre-intervention", lty="Pre-intervention"), lwd=3) +
-  geom_line(aes(tvec, speed, col="Post-intervention (constant-speed)", lty="Post-intervention (constant-speed)"), lwd=3) +
[…]

### 2021-08-04 response.tex
Quick response changes

@@
 
-We acknowledge that we are assuming homogeneous mixing---we feel that this is a clearer explanation because mass-action is often used to refer to how the transmission rate, beta, scales with the population size.
+We now acknowledge that we are assuming homogeneous mixing---we feel that this is a clearer explanation because mass-action is often used to refer to how the transmission rate, beta, scales with the population size.
 
@@
 
-We further note that a fixed generation-interval distribution does not rely on the assumption that the rate of depletion of susceptibles over time is negligible.
-This is due to an ongoing confusion between the intrinsic and realized generation interval as we explained above.
-For example, as shown in Champredon et al 2015 and 2018, the standard SEIR model can be equivalently written in terms of the renewal equation using a fixed \emph{intrinsic} generation-interval distribution, and we can still observe the contraction of the \emph{realized} generation interval for this model.
+We further note that our fixed generation-interval distribution does not rely on the assumption that the rate of depletion of susceptibles over time is negligible.
+This is because we are assuming that intrinsic, not realized, generation intervals are fixed.
+As shown in Champredon et al 2015 and 2018, the standard SEIR model can be written as a renewal equation using a fixed \emph{intrinsic} generation-interval distribution, and we can still observe the contraction of the \emph{realized} generation interval for this model.
 We have added a paragraph in Section 2 to clarify this distinction (please see response to reviewer 1 for text).

### 2021-08-05 newvariant.tex
downing

@@
 
-Inferring the relative strength (i.e., the ratio of reproduction numbers) and relative speed (i.e., the difference between growth rates) of new SARS-CoV-2 variants compared to the wild types, from which they emerged, is critical to predicting and controlling the course of the current pandemic.
+Inferring the relative strength (i.e., the ratio of reproduction numbers) and relative speed (i.e., the difference between growth rates) of new SARS-CoV-2 variants compared to earlier strains is critical to predicting and controlling the course of the current pandemic.
 We explore how differences in the generation-interval distributions affect the inference of relative strength and speed in light of recent evidence that the Alpha and Delta variants may have different generation intervals from prior dominant lineages.
@@
 The generation interval, defined as the time between infection and transmission, provides information about the time scale of individual-level transmission \citep{svensson2007note}.
-The generation interval is also distinctly different from other so-called ``transmission intervals'' that measure time between successive infections---including the serial interval, which is defined as the time between symptom onsets in an infector-infectee pair \citep{fine2003interval,grassly2008mathematical,britton2019estimation,ali2020serial,park2021forward}.
+The generation interval is also distinctly different from other ``transmission intervals'' that measure time between successive infections---including the serial interval, which is defined as the time between symptom onsets in an infector-infectee pair \citep{fine2003interval,grassly2008mathematical,britton2019estimation,ali2020serial,park2021forward}.
 
@@
 While it is possible to consider general forms of generation-interval distributions \citep{miller2010epidemics,svensson2015influence}, summarizing the distribution in terms of its mean and variability---for example, by assuming they are Gamma distributed---can still provide a robust link between epidemic speed and strength for real diseases and yield important biological insights \citep{park2019practical}.
-In particular, several studies have noted, in many contexts, that mechanisms that increase the mean generation interval increase epidemic strength $\RR$ given observed epidemic speed $r$ \citep{eaton2014proportion,powers2014impact,weitz2015modeling,park2020time}.
+In particular, several studies have noted, in many contexts, that mechanisms that increase the mean generation interval increase the epidemic strength $\RR$ that corresponds to an observed epidemic speed $r$ \citep{eaton2014proportion,powers2014impact,weitz2015modeling,park2020time}.
 
@@
 Many studies have tried to estimate the relative strength of variants from the observed relative speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021,campbell2021increased}.
-Instead (or additionally), some studies have assumed a value of the relative strength and tried to predict its relative speed \citep{davies2021estimated,di2021impact}.
-While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the disease and its control because the two quantities are linked by the generation-interval distribution \citep{doi:10.1098/rspb.2020.1556}.
+Instead (or additionally), some studies have assumed a value of the relative strength and tried to predict relative speed \citep{davies2021estimated,di2021impact}.
+While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the disease and its control\citep{doi:10.1098/rspb.2020.1556}.
 
@@
 Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals leads to a higher estimate of epidemic strength ($\RR$), making the epidemic look harder to control with a constant-strength intervention.
-Constant strength and speed interventions are idealized representations of real-life interventions, which can be either strength-like (e.g., vaccination and social distancing) or speed-like (e.g., contact tracing and isolation) depending on how their effectiveness varies across the generation interval (see \cite{doi:10.1098/rspb.2020.1556} and Discussion).
+Constant-strength and -speed interventions are idealized representations of real-life interventions, which can be either strength-like (e.g., vaccination and social distancing) or speed-like (e.g., contact tracing and isolation) depending on how their effectiveness varies across the generation interval (see \cite{doi:10.1098/rspb.2020.1556} and Discussion).
 
@@
 However, the magnitude of relative strength $\rho$ is sensitive to assumptions about generation intervals:
-For realistic values of $\kappa$ (excluding 0 and 1), the inferred relative strength $\rho$ ranges between 1.1 and 2.3 when $\Gv$ is allowed to vary between $2\Gw/3$ and $3\Gw/2$.
+For realistic values of $\kappa$ (excluding 0 and 1), the inferred relative strength $\rho$ ranges between 1.1 and 2.3 when $\Gv$ is allowed to vary between $2/3$ and $3/2$ of \Gw.
 

### 2021-08-08 response.tex
disease or pathogen?

@@
 
-``Epidemic speed can be typically estimated from time series of incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}, but epidemic strength is difficult to measure directly.
+``Epidemic speed can be typically estimated from time series of incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}, but epidemic strength is difficult to measure from incidence time series.
 Instead, epidemic strength is often inferred from the observed epidemic speed using the generation-interval distribution, an approach popularized by \citep{wallinga2007generation}.
 The generation interval, defined as the time between infection and transmission, provides information about the time scale of individual-level transmission \citep{svensson2007note}.
-The generation interval is also distinctly different from other so-called ``transmission intervals'' that measure time between successive infections---including the serial interval, which is defined as the time between symptom onsets in an infector-infectee pair \citep{fine2003interval,grassly2008mathematical,britton2019estimation,ali2020serial,park2021forward}.
+The generation interval is also distinctly different from other ``transmission intervals'' that measure time between successive infections---including the serial interval, which is defined as the time between symptom onsets in an infector-infectee pair \citep{fine2003interval,grassly2008mathematical,britton2019estimation,ali2020serial,park2021forward}.
 
-The exact shape of the distribution depends on several factors---including the shape of latent and infectious periods \citep{lloyd2001realistic,wearing2005appropriate,roberts2007model} as well as more detailed life history of a disease \citep{huber2016quantitative}---and thus can be difficult to estimate.
-While it is possible to consider general forms of generation-interval distributions \citep{miller2010epidemics,svensson2015influence}, summarizing the distribution in terms of its mean and variability---for example, by using the Gamma distribution---can still provide a robust link between epidemic speed and strength for real diseases and yield important biological insights \citep{park2019practical}.
-In particular, several studies have noted, in many contexts, that mechanisms that increase the mean generation interval increase epidemic strength $\RR$ given observed epidemic speed $r$ \citep{eaton2014proportion,powers2014impact,weitz2015modeling,park2020time}.''
+The exact shape of the distribution depends on several factors---including the shape of latent and infectious period distributions \citep{lloyd2001realistic,wearing2005appropriate,roberts2007model} as well as more detailed life history of a disease \citep{huber2016quantitative}---and thus can be difficult to estimate.
+While it is possible to consider general forms of generation-interval distributions \citep{miller2010epidemics,svensson2015influence}, summarizing the distribution in terms of its mean and variability---for example, by assuming they are Gamma distributed---can still provide a robust link between epidemic speed and strength for real diseases and yield important biological insights \citep{park2019practical}.
+In particular, several studies have noted, in many contexts, that mechanisms that increase the mean generation interval increase the epidemic strength $\RR$ that corresponds to an observed epidemic speed $r$ \citep{eaton2014proportion,powers2014impact,weitz2015modeling,park2020time}.''
 
@@
 ``Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$, in which case the incidence of infections changes exponentially.
-Here, we use the term ``epidemiological conditions'' to broadly refer to all factors that affect transmission---mathematically, they are captured by the kernel $K(t, \tau)$.
-In the context of SARS-CoV-2 infections, we are essentially assuming that the changes in susceptible pool, behavior, and contact rates are usually small over a short period of time.
+Here, we use the term ``epidemiological conditions'' to broadly refer to all factors that affect transmission at the population-level---mathematically, they are captured by the kernel $K(t, \tau)$.
+In the context of SARS-CoV-2 infections, we are essentially assuming that the changes in susceptible pool, behavior, and contact rates are usually small over a short period of time---for example, this assumption would not apply at the times of drastic policy changes but can still work before and after policy changes have been made.
 Then, the incidence of each strain grows (or decays) exponentially at rate $r_x$, satisfying the Euler-Lotka equation \citep{wallinga2007generation}:
@@
 
-``For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals leads to a slower estimated epidemic growth rate (lower $r$), making the epidemic look easier to control with a constant-speed intervention, which isolates infected individuals at a constant hazard and therefore affects late transmission disproportionately.
+``For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals results in a slower estimated epidemic growth rate (lower $r$), making an epidemic look easier to control with a constant-speed intervention, which isolates infected individuals at a constant hazard---such intervention affects late transmission disproportionately and further affect the post-intervention generation-interval distribution.
 As shown in \cite{doi:10.1098/rspb.2020.1556}, the hazard of isolation must be greater than the epidemic speed $r$ to reach the $r=0$ threshold.
 In this same case, however, variation in the generation-interval distribution has no effects on the effectiveness of a constant-strength intervention, which reduces the hazard of infectious contacts and therefore affects transmission by an equal amount regardless of age of infection.
-Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals leads to a higher estimate of epidemic strength ($\RR$), making the epidemic look harder to control with a constant-strength interventions.
-Constant strength and speed are idealized representations of real-life interventions, which can be either strength-like (e.g., vaccination and social distancing) or speed-like (e.g., contact tracing and isolation) depending on how their effectiveness varies across the generation interval (see \cite{doi:10.1098/rspb.2020.1556} and Discussion).''
+Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals leads to a higher estimate of epidemic strength ($\RR$), making the epidemic look harder to control with a constant-strength intervention.
+Constant-strength and -speed interventions are idealized representations of real-life interventions, which can be either strength-like (e.g., vaccination and social distancing) or speed-like (e.g., contact tracing and isolation) depending on how their effectiveness varies across the generation interval (see \cite{doi:10.1098/rspb.2020.1556} and Discussion).''
 
@@
 ``Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$, in which case the incidence of infections changes exponentially.
-Here, we use the term ``epidemiological conditions'' to broadly refer to all factors that affect transmission---mathematically, they are captured by the kernel $K(t, \tau)$.
-In the context of SARS-CoV-2 infections, we are essentially assuming that the changes in susceptible pool, behavior, and contact rates are usually small over a short period of time.''
+Here, we use the term ``epidemiological conditions'' to broadly refer to all factors that affect transmission at the population-level---mathematically, they are captured by the kernel $K(t, \tau)$.
+In the context of SARS-CoV-2 infections, we are essentially assuming that the changes in susceptible pool, behavior, and contact rates are usually small over a short period of time---for example, this assumption would not apply at the times of drastic policy changes but can still work before and after policy changes have been made.''
 

### 2021-08-08 newvariant.tex
disease or pathogen?

@@
 Instead (or additionally), some studies have assumed a value of the relative strength and tried to predict relative speed \citep{davies2021estimated,di2021impact}.
-While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the disease and its control\citep{doi:10.1098/rspb.2020.1556}.
+While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the pathogen and its control\citep{doi:10.1098/rspb.2020.1556}.
 

### 2021-08-09 response.tex
time to submit

@@
 
-The exact shape of the distribution depends on several factors---including the shape of latent and infectious period distributions \citep{lloyd2001realistic,wearing2005appropriate,roberts2007model} as well as more detailed life history of a disease \citep{huber2016quantitative}---and thus can be difficult to estimate.
-While it is possible to consider general forms of generation-interval distributions \citep{miller2010epidemics,svensson2015influence}, summarizing the distribution in terms of its mean and variability---for example, by assuming they are Gamma distributed---can still provide a robust link between epidemic speed and strength for real diseases and yield important biological insights \citep{park2019practical}.
+The exact shape of the distribution depends on several factors---including the shape of latent and infectious period distributions \citep{lloyd2001realistic,wearing2005appropriate,roberts2007model} as well as more detailed life history of a pathogen \citep{huber2016quantitative}---and thus can be difficult to estimate.
+While it is possible to consider general forms of generation-interval distributions \citep{miller2010epidemics,svensson2015influence}, summarizing the distribution in terms of its mean and variability---for example, by assuming they are Gamma distributed---can still provide a robust link between epidemic speed and strength for real pathogens and yield important biological insights \citep{park2019practical}.
 In particular, several studies have noted, in many contexts, that mechanisms that increase the mean generation interval increase the epidemic strength $\RR$ that corresponds to an observed epidemic speed $r$ \citep{eaton2014proportion,powers2014impact,weitz2015modeling,park2020time}.''
@@
 
-We now cite Breda et al., and acknowledge that ``This framework provides a flexible way of modeling disease dynamics and generalizes a wide range of compartmental models, including the SEIR model \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral,breda2012formulation, champredon2018equivalence}.''
+We now cite Breda et al., and acknowledge that ``This framework provides a flexible way of modeling pathogen dynamics and generalizes a wide range of compartmental models, including the SEIR model \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral,breda2012formulation, champredon2018equivalence}.''
 

### 2021-08-09 newvariant.tex
time to submit

@@
 Epidemic ``strength'' is measured by the reproduction number $\RR$---a unitless quantity representing the average number of new infections caused by a typical infection. 
-Disease can spread in a population if $\RR>1$ \citep{diekmann1990definition}.
+A pathogen can spread in a population if $\RR>1$ \citep{diekmann1990definition}.
 The epidemic strength also determines the final size of an epidemic in a homogeneous population \citep{anderson1991infectious}.
 Epidemic ``speed'' is characterized by the growth rate $r$, which has units of $1/\mathrm{time}$ and describes the exponential rate of pathogen spread at the population level.
-Like epidemic strength, epidemic speed also determines conditions for disease elimination: $r=0$ is a threshold equivalent to $\RR=1$ under constant conditions.
+Like epidemic strength, epidemic speed also determines conditions for pathogen elimination: $r=0$ is a threshold equivalent to $\RR=1$ under constant conditions.
 
@@
 
-The exact shape of the distribution depends on several factors---including the shape of latent and infectious period distributions \citep{lloyd2001realistic,wearing2005appropriate,roberts2007model} as well as more detailed life history of a disease \citep{huber2016quantitative}---and thus can be difficult to estimate.
-While it is possible to consider general forms of generation-interval distributions \citep{miller2010epidemics,svensson2015influence}, summarizing the distribution in terms of its mean and variability---for example, by assuming they are Gamma distributed---can still provide a robust link between epidemic speed and strength for real diseases and yield important biological insights \citep{park2019practical}.
+The exact shape of the distribution depends on several factors---including the shape of latent and infectious period distributions \citep{lloyd2001realistic,wearing2005appropriate,roberts2007model} as well as more detailed life history of a pathogen \citep{huber2016quantitative}---and thus can be difficult to estimate.
+While it is possible to consider general forms of generation-interval distributions \citep{miller2010epidemics,svensson2015influence}, summarizing the distribution in terms of its mean and variability---for example, by assuming they are Gamma distributed---can still provide a robust link between epidemic speed and strength for real pathogens and yield important biological insights \citep{park2019practical}.
 In particular, several studies have noted, in many contexts, that mechanisms that increase the mean generation interval increase the epidemic strength $\RR$ that corresponds to an observed epidemic speed $r$ \citep{eaton2014proportion,powers2014impact,weitz2015modeling,park2020time}.
@@
 \end{equation}
-This framework provides a flexible way of modeling disease dynamics and generalizes a wide range of compartmental models, including the SEIR model \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral,breda2012formulation, champredon2018equivalence}.
+This framework provides a flexible way of modeling pathogen dynamics and generalizes a wide range of compartmental models, including the SEIR model \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral,breda2012formulation, champredon2018equivalence}.
 
@@
 The instantaneous reproduction number is a particular kind of weighted average of infectiousness of previously infected individuals at time $t$---in particular, it is weighted by the total relative infectiousness at time $t$, rather than by the actual number of infected individuals present.
-The normalized kernel $g_x(t, \tau) = K_x(t, \tau)/\RR_x(t)$---which we refer to as the instantaneous generation-interval distribution---describes the relative contribution of previously infected individuals to current incidence $i_x(t)$ and provides information about the time scale of disease transmission.
+The normalized kernel $g_x(t, \tau) = K_x(t, \tau)/\RR_x(t)$---which we refer to as the instantaneous generation-interval distribution---describes the relative contribution of previously infected individuals to current incidence $i_x(t)$ and provides information about the time scale of pathogen transmission.
 Like the instantaneous reproduction number, the instantaneous generation-interval distribution describes when an individual infected at time $t$ will transmit infections if conditions were to remain the same.
@@
 
-We explored how the generation-interval distribution shapes the link between relative strength and speed of an invading disease variant.
+We explored how the generation-interval distribution shapes the link between relative strength and speed of an invading pathogen variant.
 Longer generation intervals generally lead to higher relative strength for a given relative speed (and conversely, lower relative speed for a given relative strength); these relationships are reversed when incidence is decreasing.

### 2021-08-10 newvariant.tex
bib

@@
 
-\bibliography{newvariant_abbv.bib}
+\begin{thebibliography}{10}
+
+\bibitem{switzerland2021variant}
+Althaus CL, Reichmuth M, Hodcroft E, Riou J, Schibler M, Eckerle I, Kaiser L,
+  Suter F, Huber M, Trkola A, and others .
+\newblock 2021 {Transmission of SARS-CoV-2 variants in Switzerland}.
+\newblock \url{https://ispmbern.github.io/covid-19/variants/}.
+
+\bibitem{davies2021estimated}
+Davies NG, Abbott S, Barnard RC, Jarvis CI, Kucharski AJ, Munday JD, Pearson
+  CAB, Russell TW, Tully DC, Washburne AD, and others .
+\newblock 2021 {Estimated transmissibility and impact of SARS-CoV-2 lineage
+  B.1.1.7 in England}.
+\newblock {\em Science}.
+
+\bibitem{di2021impact}
+Di~Domenico L, Pullano G, Sabbatini CE, L{\'e}vy-Bruhl D, and Colizza V.
+\newblock 2021 {Impact of January 2021 social distancing measures on SARS-CoV-2
+  B. 1.1. 7 circulation in France}.
+\newblock {\em medRxiv}.
+\newblock \url{https://www.medrxiv.org/content/10.1101/2021.02.14.21251708v1}.
+
+\bibitem{graham2021changes}
+Graham MS, Sudre CH, May A, Antonelli M, Murray B, Varsavsky T, Kl{\"a}ser K,
+  Canas LS, Molteni E, and others .
+\newblock 2021/05/03 2021 {Changes in symptomatology, reinfection, and
+  transmissibility associated with the SARS-CoV-2 variant B.1.1.7: an
+  ecological study}.
+\newblock {\em Lancet Public Health}.
+\newblock {\bfseries 6}, e335--e345.
+
+\bibitem{leung2021early}
+Leung K, Shum MH, Leung GM, Lam TT, and Wu~JT.
+\newblock 2021 {Early transmissibility assessment of the N501Y mutant strains
+  of SARS-CoV-2 in the United Kingdom, October to November 2020}.
+\newblock {\em Euro Surveill.}
+\newblock {\bfseries 26}, 2002106.
+
+\bibitem{volz2021transmission}
+Volz E, Mishra S, Chand M, Barrett JC, Johnson R, Geidelberg L, Hinsley WR,
+  Laydon DJ, Dabrera G, O'Toole {\'A}, and others .
+\newblock 2021 {Assessing transmissibility of SARS-CoV-2 lineage B.1.1.7 in
+  England}.
+\newblock {\em Nature}.
+
+\bibitem{zhao2021}
+Zhao S, Lou J, Cao L, Zheng H, Chong MKC, Chen Z, Chan RWY, Zee BCY, Chan PKS,
+  and Wang MH.
+\newblock 01 2021 {Quantifying the transmission advantage associated with N501Y
+  substitution of SARS-CoV-2 in the UK: an early data-driven analysis}.
+\newblock {\em J. Travel Med}.
+\newblock taab011.
+
+\bibitem{campbell2021increased}
+Campbell F, Archer B, Laurenson-Schafer H, Jinnai Y, Konings F, Batra N, Pavlin
+  B, Vandemaele K, Van~Kerkhove MD, Jombart T, and others .
+\newblock 2021 Increased transmissibility and global spread of {SARS-CoV-2}
+  variants of concern as at {June}, 2021.
+\newblock {\em Euro Surveill.}
+\newblock {\bfseries 26}, 2100509.
+
+\bibitem{diekmann1990definition}
+Diekmann O, Heesterbeek JAP, and Metz JA.
+\newblock 1990 On the definition and the computation of the basic reproduction
+  ratio $\mathcal{R}_0$ in models for infectious diseases in heterogeneous
+  populations.
+\newblock {\em J. Math. Biol.}
+\newblock {\bfseries 28}, 365--382.
+
+\bibitem{anderson1991infectious}
+Anderson RM and May RM.
+\newblock 1991 {\em {Infectious diseases of humans: dynamics and control}}.
+\newblock Oxford University Press.
+
+\bibitem{mills2004transmissibility}
+Mills CE, Robins JM, and Lipsitch M.
+\newblock 2004 Transmissibility of 1918 pandemic influenza.
+\newblock {\em Nature}.
+\newblock {\bfseries 432}, 904--906.
+
+\bibitem{nishiura2009transmission}
+Nishiura H, Castillo-Chavez C, Safan M, and Chowell G.
+\newblock 2009 {Transmission potential of the new influenza A (H1N1) virus and
+  its age-specificity in Japan}.
+\newblock {\em Euro Surveill.}
+\newblock {\bfseries 14}, 19227.
+
+\bibitem{ma2014estimating}
+Ma~J, Dushoff J, Bolker BM, and Earn DJD.
+\newblock 2014 Estimating initial epidemic growth rates.
+\newblock {\em Bull. Math. Biol.}
+\newblock {\bfseries 76}, 245--260.
+
+\bibitem{wallinga2007generation}
+Wallinga J and Lipsitch M.
+\newblock 2007 How generation intervals shape the relationship between growth
+  rates and reproductive numbers.
+\newblock {\em Proc. R. Soc. B}.
+\newblock {\bfseries 274}, 599--604.
+
+\bibitem{svensson2007note}
+Svensson {\AA}.
+\newblock 2007 A note on generation times in epidemic models.
+\newblock {\em Math. Biosci.}
+\newblock {\bfseries 208}, 300--311.
+
+\bibitem{fine2003interval}
+Fine PE.
+\newblock 2003 The interval between successive cases of an infectious disease.
+\newblock {\em Am. J. Epidemiol.}
+\newblock {\bfseries 158}, 1039--1047.
+
+\bibitem{grassly2008mathematical}
+Grassly NC and Fraser C.
+\newblock 2008 Mathematical models of infectious disease transmission.
+\newblock {\em Nat. Rev. Microbiol.}
+\newblock {\bfseries 6}, 477--487.
+
+\bibitem{britton2019estimation}
+Britton T and Scalia~Tomba G.
+\newblock 2019 {Estimation in emerging epidemics: Biases and remedies}.
+\newblock {\em J. R. Soc. Interface}.
+\newblock {\bfseries 16}, 20180670.
+
+\bibitem{ali2020serial}
+Ali ST, Wang L, Lau EH, Xu~XK, Du~Z, Wu~Y, Leung GM, and Cowling BJ.
+\newblock 2020 {Serial interval of SARS-CoV-2 was shortened over time by
+  nonpharmaceutical interventions}.
+\newblock {\em Science}.
+\newblock {\bfseries 369}, 1106--1109.
+
+\bibitem{park2021forward}
+Park SW, Sun K, Champredon D, Li~M, Bolker BM, Earn DJ, Weitz JS, Grenfell BT,
+  and Dushoff J.
+\newblock 2021 Forward-looking serial intervals correctly link epidemic growth
+  to reproduction numbers.
+\newblock {\em Proc. Natl. Acad. Sci. U.S.A.}
+\newblock {\bfseries 118}, e2011548118.
+
+\bibitem{lloyd2001realistic}
+Lloyd AL.
+\newblock 2001 Realistic distributions of infectious periods in epidemic
+  models: changing patterns of persistence and dynamics.
+\newblock {\em Theor. Popul. Biol.}
+\newblock {\bfseries 60}, 59--71.
+
+\bibitem{wearing2005appropriate}
+Wearing HJ, Rohani P, and Keeling MJ.
+\newblock 2005 Appropriate models for the management of infectious diseases.
+\newblock {\em PLoS medicine}.
+\newblock {\bfseries 2}, e174.
+
+\bibitem{roberts2007model}
+Roberts M and Heesterbeek J.
+\newblock 2007 Model-consistent estimation of the basic reproduction number
+  from the incidence of an emerging infection.
+\newblock {\em J. Math. Biol.}
+\newblock {\bfseries 55}, 803.
+
+\bibitem{huber2016quantitative}
+Huber JH, Johnston GL, Greenhouse B, Smith DL, and Perkins TA.
+\newblock 2016 Quantitative, model-based estimates of variability in the
+  generation and serial intervals of \textit{Plasmodium falciparum} malaria.
+\newblock {\em Malaria journal}.
+\newblock {\bfseries 15}, 1--12.
+
+\bibitem{miller2010epidemics}
+Miller JC, Davoudi B, Meza R, Slim AC, and Pourbohloul B.
+\newblock 2010 Epidemics with general generation interval distributions.
+\newblock {\em J. Theor. Biol.}
+\newblock {\bfseries 262}, 107--115.
+
+\bibitem{svensson2015influence}
+Svensson {\AA}.
+\newblock 2015 The influence of assumptions on generation time distributions in
+  epidemic models.
+\newblock {\em Math. Biosci.}
+\newblock 270, 81--89.
+
+\bibitem{park2019practical}
+Park SW, Champredon D, Weitz JS, and Dushoff J.
+\newblock 2019 A practical generation-interval-based approach to inferring the
+  strength of epidemics from their speed.
+\newblock {\em Epidemics}.
+\newblock 27, 12--18.
+
+\bibitem{eaton2014proportion}
+Eaton JW and Hallett TB.
+\newblock 2014 Why the proportion of transmission during early-stage {HIV}
+  infection does not predict the long-term impact of treatment on {HIV}
+  incidence.
+\newblock {\em Proc. Natl. Acad. Sci. U.S.A.}
+\newblock {\bfseries 111}, 16202--16207.
+
+\bibitem{powers2014impact}
+Powers KA, Kretzschmar ME, Miller WC, and Cohen MS.
+\newblock 2014 Impact of early-stage {HIV} transmission on treatment as
+  prevention.
+\newblock {\em Proc. Natl. Acad. Sci. U.S.A.}
+\newblock {\bfseries 111}, 15867--15868.
+
+\bibitem{weitz2015modeling}
+Weitz JS and Dushoff J.
+\newblock 2015 Modeling post-death transmission of {Ebola}: challenges for
+  inference and opportunities for control.
+\newblock {\em Sci. Rep.}
+\newblock {\bfseries 5}, 1--7.
+
+\bibitem{park2020time}
+Park SW, Cornforth DM, Dushoff J, and Weitz JS.
+\newblock 2020 The time scale of asymptomatic transmission affects estimates of
+  epidemic potential in the {COVID-19} outbreak.
+\newblock {\em Epidemics}.
+\newblock 31, 100392.
+
+\bibitem{doi:10.1098/rspb.2020.1556}
+Dushoff J and Park SW.
+\newblock 2021 Speed and strength of an epidemic intervention.
+\newblock {\em Proc. R. Soc. B}.
+\newblock {\bfseries 288}, 20201556.
+
+\bibitem{kissler2021densely}
+Kissler S, Fauver JR, Mack C, Tai CG, Breban MI, Watkins AE, Samant RM,
+  Anderson DJ, Ho~DD, Grubaugh ND, and Grad YH.
+\newblock 2021 {Densely sampled viral trajectories suggest longer duration of
+  acute infection with B.1.1.7 variant relative to non-B.1.1.7 SARS-CoV-2}.
+\newblock {\em medRxiv}.
+\newblock \url{https://www.medrxiv.org/content/10.1101/2021.02.16.21251535v1}.
+
+\bibitem{li2021viral}
+Li~B, Deng A, Li~K, Hu~Y, Li~Z, Xiong Q, Liu Z, Guo Q, Zou L, Zhang H, and
+  others .
+\newblock 2021 {Viral infection and Transmission in a large well-traced
+  outbreak caused by the Delta SARS-CoV-2 variant}.
+\newblock {\em medRxiv}.
+\newblock \url{https://www.medrxiv.org/content/10.1101/2021.07.07.21260122v1}.
+
+\bibitem{zhang2021transmission}
+Zhang M, Xiao J, Deng A, Zhang Y, Zhuang Y, Hu~T, Li~J, Tu~H, Li~B, Zhou Y, and
+  others .
+\newblock 2021 Transmission dynamics of an outbreak of the {COVID-19} {Delta}
+  variant {B.1.617.2}--{Guangdong Province, China, May--June 2021}.
+\newblock {\em China CDC Weekly}.
+\newblock 3, 584--586.
+
+\bibitem{goldstein2009reconstructing}
+Goldstein E, Dushoff J, Ma~J, Plotkin JB, Earn DJ, and Lipsitch M.
+\newblock 2009 Reconstructing influenza incidence by deconvolution of daily
+  mortality time series.
+\newblock {\em Proc. Natl. Acad. Sci. U.S.A.}
+\newblock {\bfseries 106}, 21825--21829.
+
+\bibitem{gostic2020practical}
+Gostic KM, McGough L, Baskerville EB, Abbott S, Joshi K, Tedijanto C, Kahn R,
+  Niehus R, Hay JA, De~Salazar PM, and others .
+\newblock 12 2020 Practical considerations for measuring the effective
+  reproductive number, $\mathcal{R}_t$.
+\newblock {\em PLOS Comp. Biol.}
+\newblock {\bfseries 16}, 1--21.
+
+\bibitem{heesterbeek1996concept}
+Heesterbeek J and Dietz K.
+\newblock 1996 The concept of $\mathcal{R}_0$ in epidemic theory.
+\newblock {\em Stat. Neerl.}
+\newblock {\bfseries 50}, 89--110.
+
+\bibitem{diekmann2000mathematical}
+Diekmann O and Heesterbeek JAP.
+\newblock 2000 {\em Mathematical epidemiology of infectious diseases: model
+  building, analysis and interpretation}, volume~5.
+\newblock John Wiley \& Sons.
+
+\bibitem{roberts2004modelling}
+Roberts M.
+\newblock 2004 Modelling strategies for minimizing the impact of an imported
+  exotic infection.
+\newblock {\em Proc. R. Soc. B}.
+\newblock {\bfseries 271}, 2411--2415.
+
+\bibitem{aldis2005integral}
+Aldis G and Roberts M.
+\newblock 2005 An integral equation model for the control of a smallpox
+  outbreak.
+\newblock {\em Math. Biosci.}
+\newblock {\bfseries 195}, 1--22.
+
+\bibitem{breda2012formulation}
+Breda D, Diekmann O, De~Graaf W, Pugliese A, and Vermiglio R.
+\newblock 2012 {On the formulation of epidemic models (an appraisal of Kermack
+  and McKendrick)}.
+\newblock {\em J. Biol. Dyn.}
+\newblock {\bfseries 6}, 103--117.
+
+\bibitem{champredon2018equivalence}
+Champredon D, Dushoff J, and Earn DJ.
+\newblock 2018 Equivalence of the {Erlang}-distributed {SEIR} epidemic model
+  and the renewal equation.
+\newblock {\em SIAM J. Appl. Math}.
+\newblock {\bfseries 78}, 3258--3278.
+
+\bibitem{fraser2007estimating}
+Fraser C.
+\newblock 2007 Estimating individual and household reproduction numbers in an
+  emerging epidemic.
+\newblock {\em PloS one}.
+\newblock {\bfseries 2}.
+
+\bibitem{champredon2015intrinsic}
+Champredon D and Dushoff J.
+\newblock 2015 Intrinsic and realized generation intervals in
+  infectious-disease transmission.
+\newblock {\em Proc. R. Soc. B}.
+\newblock {\bfseries 282}, 20152026.
+
+\bibitem{champredon2018two}
+Champredon D, Li~M, Bolker BM, and Dushoff J.
+\newblock 2018 Two approaches to forecast {Ebola} synthetic epidemics.
+\newblock {\em Epidemics}.
+\newblock 22, 36--42.
+
+\bibitem{kenah2008generation}
+Kenah E, Lipsitch M, and Robins JM.
+\newblock 2008 Generation interval contraction and epidemic data analysis.
+\newblock {\em Math. Biosci.}
+\newblock {\bfseries 213}, 71--79.
+
+\bibitem{nishiura2010time}
+Nishiura H.
+\newblock 2010 Time variations in the generation time of an infectious disease:
+  implications for sampling to appropriately quantify transmission potential.
+\newblock {\em Math. Biosci. Eng.}
+\newblock {\bfseries 7}, 851.
+
+\bibitem{flaxman2020Rt}
+Flaxman S, Mishra S, Gandy A, Unwin HJT, Mellan TA, Coupland H, Whittaker C,
+  Zhu H, Berah T, Eaton JW, Monod M, and others .
+\newblock 2020 {Estimating the effects of non-pharmaceutical interventions on
+  COVID-19 in Europe}.
+\newblock {\em Nature}.
+\newblock {\bfseries 584}, 257--261.
+
+\bibitem{moore2021vaccination}
+Moore S, Hill EM, Tildesley MJ, Dyson L, and Keeling MJ.
+\newblock 2021 {Vaccination and non-pharmaceutical interventions for COVID-19:
+  a mathematical modelling study}.
+\newblock {\em Lancet Infect. Dis.}
+\newblock {\bfseries 21}, 793--802.
+
+\bibitem{brauner2021inferring}
+Brauner JM, Mindermann S, Sharma M, Johnston D, Salvatier J, Gaven{\v{c}}iak T,
+  Stephenson AB, Leech G, Altman G, Mikulik V, and others .
+\newblock 2021 {Inferring the effectiveness of government interventions against
+  COVID-19}.
+\newblock {\em Science}.
+\newblock {\bfseries 371}.
+
+\bibitem{doi:10.1098/rsif.2020.0144}
+Park SW, Bolker BM, Champredon D, Earn DJD, Li~M, Weitz JS, Grenfell BT, and
+  Dushoff J.
+\newblock 2020 {Reconciling early-outbreak estimates of the basic reproductive
+  number and its uncertainty: framework and applications to the novel
+  coronavirus (SARS-CoV-2) outbreak}.
+\newblock {\em J. R. Soc. Interface}.
+\newblock {\bfseries 17}, 20200144.
+
+\bibitem{ferretti2020quantifying}
+Ferretti L, Wymant C, Kendall M, Zhao L, Nurtay A, Abeler-D{\"o}rner L, Parker
+  M, Bonsall D, and Fraser C.
+\newblock 2020 {Quantifying SARS-CoV-2 transmission suggests epidemic control
+  with digital contact tracing}.
+\newblock {\em Science}.
+\newblock {\bfseries 368}.
+
+\bibitem{leung2017monitoring}
+Leung K, Lipsitch M, Yuen KY, and Wu~JT.
+\newblock 2017 Monitoring the fitness of antiviral-resistant influenza strains
+  during an epidemic: a mathematical modelling study.
+\newblock {\em Lancet Infect. Dis.}
+\newblock {\bfseries 17}, 339--347.
+
+\bibitem{leung2020empirical}
+Leung K, Pei Y, Leung GM, Lam TT, and Wu~JT.
+\newblock 2020 {Empirical transmission advantage of the D614G mutant strain of
+  SARS-CoV-2}.
+\newblock {\em medRxiv}.
+\newblock \url{https://www.medrxiv.org/content/10.1101/2020.09.22.20199810v1}.
+
+\bibitem{wallinga2004different}
+Wallinga J and Teunis P.
+\newblock 2004 Different epidemic curves for severe acute respiratory syndrome
+  reveal similar impacts of control measures.
+\newblock {\em Am. J. Epidemiol.}
+\newblock {\bfseries 160}, 509--516.
+
+\bibitem{cori2013new}
+Cori A, Ferguson NM, Fraser C, and Cauchemez S.
+\newblock 2013 A new framework and software to estimate time-varying
+  reproduction numbers during epidemics.
+\newblock {\em Am. J. Epidemiol.}
+\newblock {\bfseries 178}, 1505--1512.
+
+\bibitem{abbott2020estimating}
+Abbott S, Hellewell J, Thompson RN, Sherratt K, Gibbs HP, Bosse NI, Munday JD,
+  Meakin S, Doughty EL, Chun JY, and others .
+\newblock 2020 {Estimating the time-varying reproduction number of SARS-CoV-2
+  using national and subnational case counts}.
+\newblock {\em Wellcome Open Research}.
+\newblock {\bfseries 5}, 112.
+
+\bibitem{knight2020estimating}
+Knight J and Mishra S.
+\newblock 2020 {Estimating effective reproduction number using generation time
+  versus serial interval, with application to COVID-19 in the Greater Toronto
+  Area, Canada}.
+\newblock {\em Infect. Dis. Model.}
+\newblock 5, 889--896.
+
+\bibitem{li2021temporal}
+Li~Y, Campbell H, Kulkarni D, Harpur A, Nundy M, Wang X, Nair H, for COVID UN,
+  and others .
+\newblock 2021 {The temporal association of introducing and lifting
+  non-pharmaceutical interventions with the time-varying reproduction number
+  (R) of SARS-CoV-2: a modelling study across 131 countries}.
+\newblock {\em Lancet Infect. Dis.}
+\newblock {\bfseries 21}, 193--202.
+
+\bibitem{tsang2020effect}
+Tsang TK, Wu~P, Lin Y, Lau EH, Leung GM, and Cowling BJ.
+\newblock 2020 {Effect of changing case definitions for COVID-19 on the
+  epidemic curve and transmission parameters in mainland China: a modelling
+  study}.
+\newblock {\em The Lancet Public Health}.
+\newblock {\bfseries 5}, e289--e296.
+
+\bibitem{ukinvest}
+{Public Health England} .
+\newblock 2020 {Investigation of novel SARS-CoV-2 variant: Variant of Concern
+  202012/01}.
+\newblock
+  \url{https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/959426/Variant_of_Concern_VOC_202012_01_Technical_Briefing_5.pdf}.
+
+\bibitem{ganyani2020estimating}
+Ganyani T, Kremer C, Chen D, Torneri A, Faes C, Wallinga J, and Hens N.
+\newblock 2020 {Estimating the generation interval for coronavirus disease
+  (COVID-19) based on symptom onset data, March 2020}.
+\newblock {\em Euro Surveill.}
+\newblock {\bfseries 25}, 2000257.
+
+\bibitem{fraser2004factors}
+Fraser C, Riley S, Anderson RM, and Ferguson NM.
+\newblock 2004 Factors that make an infectious disease outbreak controllable.
+\newblock {\em Proc. Natl. Acad. Sci. U.S.A.}
+\newblock {\bfseries 101}, 6146--6151.
+
+\bibitem{scarabel2021renewal}
+Scarabel F, Pellis L, Ogden NH, and Wu~J.
+\newblock 2021 A renewal equation model to assess roles and limitations of
+  contact tracing for disease outbreak control.
+\newblock {\em R. Soc. Open Sci.}
+\newblock {\bfseries 8}, 202091.
+
+\bibitem{unwin2020state}
+Unwin HJT, Mishra S, Bradley VC, Gandy A, Mellan TA, Coupland H, Ish-Horowicz
+  J, Vollmer MA, Whittaker C, Filippi SL, and others .
+\newblock 2020 {State-level tracking of COVID-19 in the United States}.
+\newblock {\em Nat. Commun.}
+\newblock {\bfseries 11}, 1--9.
+
+\bibitem{brett2020transmission}
+Brett TS and Rohani P.
+\newblock 2020 Transmission dynamics reveal the impracticality of {COVID-19}
+  herd immunity strategies.
+\newblock {\em Proc. Natl. Acad. Sci. U.S.A.}
+\newblock {\bfseries 117}, 25897--25903.
+
+\bibitem{Ferretti2020timing}
+Ferretti L, Ledda A, Wymant C, Zhao L, Ledda V, Abeler-D{\"o}rner L, Kendall M,
+  Nurtay A, Cheng HY, Ng~TC, and others .
+\newblock 2020 The timing of {COVID-19} transmission.
+\newblock {\em medRxiv}.
+\newblock \url{https://www.medrxiv.org/content/10.1101/2020.09.04.20188516v2}.
+
+\bibitem{abdool2021new}
+Abdool~Karim SS and de~Oliveira T.
+\newblock 2021 {New SARS-CoV-2 variants—clinical, public health, and vaccine
+  implications}.
+\newblock {\em N. Engl. J. Med.}
+
+\bibitem{fontanet2021sars}
+Fontanet A, Autran B, Lina B, Kieny MP, Karim SSA, and Sridhar D.
+\newblock 2021 {SARS-CoV-2 variants and ending the COVID-19 pandemic}.
+\newblock {\em Lancet}.
+\newblock {\bfseries 397}, 952--954.
+
+\bibitem{walensky2021sars}
+Walensky RP, Walke HT, and Fauci AS.
+\newblock 2021 {SARS-CoV-2 variants of concern in the United
+  States---Challenges and opportunities}.
+\newblock {\em JAMA}.
+\newblock {\bfseries 325}, 1037--1038.
+
+\bibitem{hellewell2020feasibility}
+Hellewell J, Abbott S, Gimma A, Bosse NI, Jarvis CI, Russell TW, Munday JD,
+  Kucharski AJ, Edmunds WJ, Sun F, and others .
+\newblock 2020 {Feasibility of controlling COVID-19 outbreaks by isolation of
+  cases and contacts}.
+\newblock {\em Lancet Glob. Health}.
+\newblock {\bfseries 8}, e488--e496.
+
+\end{thebibliography}
 

### 2021-11-15 newvariant.tex
edits edits edits

@@
 Inferring the relative strength (i.e., the ratio of reproduction numbers) and relative speed (i.e., the difference between growth rates) of new SARS-CoV-2 variants compared to earlier strains is critical to predicting and controlling the course of the current pandemic.
-We explore how differences in the generation-interval distributions affect the inference of relative strength and speed in light of recent evidence that the Alpha and Delta variants may have different generation intervals from prior dominant lineages.
+Here, we challenge a common assumption (constant relative speed) in estimating the transmission advantage of SARS-CoV-2 variants.
+We argue that holding relative strength constant is generally more appropriate by comparing how relative strength and speed change over time under two idealized interventions:
+a constant-strength intervention like idealized vaccination and social distancing, which reduces transmission rates by a constant amount, and a constant-speed intervention like idealized contact tracing, which isolate infected individuals at a constant rate.
+The impact of these interventions on the spread of new variants is evaluated using a generation-interval-based framework in light of recent evidence that the Alpha and Delta variants may have different generation intervals from prior dominant lineages.
+Differences in the generation-interval distributions affect not only the effectiveness of these interventions but also the realized transmission advantages of new variants.
 Neglecting these differences gives biased estimates of relative strength, which can vary systematically with changes in the incidence of infections caused by the variant.
-We further argue that holding relative strength constant is generally more appropriate for estimating the transmission advantage of new SARS-CoV-2 variants by comparing how relative strength and speed change over time under two idealized interventions of constant strength and speed.
-Constant-strength interventions like idealized vaccination and social distancing, which reduce transmission rates by a constant amount, affect relative speed but not relative strength.
-Constant-speed interventions like idealized contact tracing, which isolate infected individuals at a constant rate, affect relative strength but not relative speed.
-Changes in the generation-interval distribution can alter the effectiveness of these interventions.
 
@@
 Like epidemic strength, epidemic speed also determines conditions for pathogen elimination: $r=0$ is a threshold equivalent to $\RR=1$ under constant conditions.
+Dushoff and Park \citep{doi:10.1098/rspb.2020.1556} argued previously that large segments of the epidemiology and modeling community have over-emphasized $\RR$ at the expense of $r$; 
+we thus use the terms ``strength'' and ``speed''  here to underline our contention that these are better seen as complementary perspectives (and to link them to complementary perspectives on measuring the transmission advantage of new variants).
 
@@
 Instead (or additionally), some studies have assumed a value of the relative strength and tried to predict relative speed \citep{davies2021estimated,di2021impact}.
-While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the pathogen and its control\citep{doi:10.1098/rspb.2020.1556}.
-
-For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals results in a slower estimated epidemic growth rate (lower $r$), making an epidemic look easier to control with a constant-speed intervention, which isolates infected individuals at a constant hazard---such intervention affects late transmission disproportionately and further affect the post-intervention generation-interval distribution.
-As shown in \cite{doi:10.1098/rspb.2020.1556}, the hazard of isolation must be greater than the epidemic speed $r$ to reach the $r=0$ threshold.
-In this same case, however, variation in the generation-interval distribution has no effects on the effectiveness of a constant-strength intervention, which reduces the hazard of infectious contacts and therefore affects transmission by an equal amount regardless of age of infection.
+
+While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the pathogen and its control \citep{doi:10.1098/rspb.2020.1556}.
+To illustrate the differences in implications for holding $\RR$ or $r$ fixed, we consider two idealized interventions of constant strength and constant speed in this paper.
+Given a pre-intervention kernel $K(\tau)$, which represents the rate at which an infected individual generates secondary infections $\tau$ time units after infection, a constant-strength intervention reduces transmission by a constant factor $\theta$ regardless of age of infection ($K(\tau)/\theta$)---in this case, the intervention strength $\theta$ must be greater than the epidemic strength $\RR$ to prevent onward transmission.
+On the other hand, a constant-speed intervention reduces transmission after infection by a constant rate $\phi$: $K(\tau) \exp(-\phi \tau)$---in this case, the intervention speed $\phi$ must be greater than the epidemic speed $r$ to prevent onward transmission \citep{doi:10.1098/rspb.2020.1556}.
+
+For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals results in a slower estimated epidemic growth rate (lower $r$), making an epidemic look easier to control with a constant-speed intervention.
+In this same case, however, variation in the generation-interval distribution has no effects on the effectiveness of a constant-strength intervention because $\RR$ is fixed.
 Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals leads to a higher estimate of epidemic strength ($\RR$), making the epidemic look harder to control with a constant-strength intervention.
@@
 
-Here, we explore how different assumptions about the generation interval affect estimates of the relative strength ($\rho=\Rv/\Rw$) and relative speed ($\delta=\rv-\rw$) for new variants.
-We show that neglecting differences in the generation-interval distributions can lead to biased estimates.
-We also discuss how such biases might be assessed in practice and how information on differences in generation interval distributions might influence priorities for controlling the spread of VoCs.
+Here, we use the generation-interval-based framework to compare two measures of relative transmissibility of new variants: the relative strength ($\rho=\Rv/\Rw$) and relative speed ($\delta=\rv-\rw$).
+We assess how relative strength and speed depend on underlying epidemiologicaldynamics of prior dominant lineages and argue that assuming a constant relative strength (rather than a constant relative speed) is more appropriate for estimating relative transmissibility of new variants.
+We also show how neglecting differences in the generation-interval distributions of new variants can lead to biased estimates of their relative transmissibility and how such biases might be assessed in practice.
+Finally, we discuss how information on differences in generation interval distributions might influence priorities for controlling the spread of VoCs.
 
@@
 Anderson RM and May RM.
-\newblock 1991 {\em {Infectious diseases of humans: dynamics and control}}.
+\newblock 1991 {\em {Infectious Diseases of Humans: Dynamics and Control}}.
 \newblock Oxford University Press.
@@
 Diekmann O and Heesterbeek JAP.
-\newblock 2000 {\em Mathematical epidemiology of infectious diseases: model
-  building, analysis and interpretation}, volume~5.
+\newblock 2000 {\em Mathematical Epidemiology of Infectious Diseases: Model
+  Building, Analysis and Interpretation}, volume~5.
 \newblock John Wiley \& Sons.

### 2021-11-16 newvariant.tex
almost there?

@@
 
-We explored how the generation-interval distribution shapes the link between relative strength and speed of an invading pathogen variant.
-Longer generation intervals generally lead to higher relative strength for a given relative speed (and conversely, lower relative speed for a given relative strength); these relationships are reversed when incidence is decreasing.
+We explored how the generation-interval distribution shapes the link between relative strength and speed.
+While the role of generation-interval distributions in linking epidemic strength and speed has been well-established, using the generation-interval-based framework allowed us to obtain imporant insight: the relative strength and speed of new variants also depend on epidemic dynamics of the prior dominant lineages as well as the type of control measures (i.e., strength- or speed-like) that limit their spread.
+For example, under a constant-strength intervention that reduces the transmission rate of the wild type and a variant by the same amount (and therefore, keeping the relative strength constant), the relative speed of the variant is sensitive to
+
+Longer generation intervals generally lead to higher relative strength for a given relative speed (and conversely, lower relative speed for a given relative strength), but these relationships are reversed when incidence is decreasing.
+
 Neglecting potential differences in the mean generation intervals between the variant and the wild type can bias estimates of the relative strength from incidence data;

### 2021-11-29 newvariant.tex
am I done?

--- 2021-11-15 newvariant.tex
@@
 Inferring the relative strength (i.e., the ratio of reproduction numbers) and relative speed (i.e., the difference between growth rates) of new SARS-CoV-2 variants compared to earlier strains is critical to predicting and controlling the course of the current pandemic.
-Here, we challenge a common assumption (constant relative speed) in estimating the transmission advantage of SARS-CoV-2 variants.
+Here, we challenge a commonly-made assumption (i.e., a constant relative speed) in estimating the transmission advantage of SARS-CoV-2 variants.
 We argue that holding relative strength constant is generally more appropriate by comparing how relative strength and speed change over time under two idealized interventions:
@@
 
-Here, we use the generation-interval-based framework to compare two measures of relative transmissibility of new variants: the relative strength ($\rho=\Rv/\Rw$) and relative speed ($\delta=\rv-\rw$).
-We assess how relative strength and speed depend on underlying epidemiologicaldynamics of prior dominant lineages and argue that assuming a constant relative strength (rather than a constant relative speed) is more appropriate for estimating relative transmissibility of new variants.
+Here, we use the generation-interval-based framework to compare two measures of transmission advantage of new variants: the relative strength ($\rho=\Rv/\Rw$) and relative speed ($\delta=\rv-\rw$).
+We assess how relative strength and speed depend on underlying epidemiological dynamics of prior dominant lineages and argue that assuming a constant relative strength (rather than a constant relative speed) is more appropriate for estimating relative transmissibility of new variants.
 We also show how neglecting differences in the generation-interval distributions of new variants can lead to biased estimates of their relative transmissibility and how such biases might be assessed in practice.
@@
 
-We explored how the generation-interval distribution shapes the link between relative strength and speed of an invading pathogen variant.
-Longer generation intervals generally lead to higher relative strength for a given relative speed (and conversely, lower relative speed for a given relative strength); these relationships are reversed when incidence is decreasing.
-Neglecting potential differences in the mean generation intervals between the variant and the wild type can bias estimates of the relative strength from incidence data;
-these biases may be assessed by considering whether estimates of relative strength appear to vary systematically with the direction of changes in the incidence of infections caused by the variant.
-Finally, differences in generation intervals can also lead to different conclusions about the effectiveness of interventions.
+We explored how the generation-interval distribution shapes the link between relative strength and speed.
+While the role of generation-interval distributions in linking epidemic strength and speed has been well-established, using the generation-interval-based framework provides clear expanations for an imporant, but under-appreciated concept: the relative strength and speed of new variants also depend on epidemic dynamics of the prior dominant lineages as well as the type of control measures (i.e., strength- or speed-like) that limit their spread.
+For example, under a constant-strength intervention that reduces the transmission rate of the wild type and a variant by the same amount (and therefore, keeping the relative strength constant), the relative speed of the variant is sensitive to the epidemic growth rate of the wild type strain---this challenges the commonly-made assumption that relative speed of new variants remains constant over time.
+Differences in the generation-interval distributions further exaggerate this effect.
+
+Therefore, neglecting potential differences in the mean generation intervals between the variant and the wild type can bias estimates of the relative strength from incidence data.
+These biases may be assessed by considering whether estimates of relative strength appear to vary systematically with the direction of changes in the incidence of infections caused by the variant.
+Differences in generation intervals can also lead to different conclusions about the effectiveness of interventions.
 If the variant has longer generation intervals than the wild type, speed-like interventions will be relatively more effective than naive estimates would suggest. 
@@
 Several studies have tried to estimate the generation-interval distribution, with means ranging between 3--6 days and squared coefficients of variation ranging between 0.1--1 \citep{ferretti2020quantifying,Ferretti2020timing,ganyani2020estimating,knight2020estimating}.
-However, these estimates are derived from serial intervals (i.e., time between symptom onset of the infector and the infectee; \cite{svensson2007note}), which are subject to dynamical biases \citep{park2021forward} and fail to account for asymptomatic transmission, adding further uncertainty to inferences of speed and strength \citep{park2020time}.
+However, these estimates are derived from serial intervals (i.e., time between symptom onset of the infector and the infectee; \cite{svensson2007note}), which are subject to dynamical biases \citep{park2021forward} and fail to account for asymptomatic transmission \citep{park2020time}.
+Recently, it has been further suggested that the generation intervals of the original SARS-CoV-2 strain may be considerably longer than previously thought \citep{sender2021unmitigated}, adding further uncertainty.
+Although a few studies have tried to quantify both the generation- and serial-interval distributions of new SARS-CoV-2 variants, their conclusions on whether transmission time scale of new variants differs are conflicting \citep{pung2021serial, hart2021generation}---we note that these comparisons are inherently difficult especially due to temporal changes in generation- and serial-interval distributions due to dynamical and intervention effects \citep{ali2020serial,park2021forward}.
 Future studies should prioritize detailed assessment of the generation intervals of SARS-CoV-2 and widespread variants, as well as consider how uncertainty in generational intervals might bias conclusions \citep{doi:10.1098/rsif.2020.0144,ali2020serial,gostic2020practical}.
@@
 
-\begin{thebibliography}{10}
-
-\bibitem{switzerland2021variant}
-Althaus CL, Reichmuth M, Hodcroft E, Riou J, Schibler M, Eckerle I, Kaiser L,
-  Suter F, Huber M, Trkola A, and others .
-\newblock 2021 {Transmission of SARS-CoV-2 variants in Switzerland}.
-\newblock \url{https://ispmbern.github.io/covid-19/variants/}.
-
-\bibitem{davies2021estimated}
-Davies NG, Abbott S, Barnard RC, Jarvis CI, Kucharski AJ, Munday JD, Pearson
-  CAB, Russell TW, Tully DC, Washburne AD, and others .
-\newblock 2021 {Estimated transmissibility and impact of SARS-CoV-2 lineage
-  B.1.1.7 in England}.
-\newblock {\em Science}.
-
-\bibitem{di2021impact}
-Di~Domenico L, Pullano G, Sabbatini CE, L{\'e}vy-Bruhl D, and Colizza V.
-\newblock 2021 {Impact of January 2021 social distancing measures on SARS-CoV-2
-  B. 1.1. 7 circulation in France}.
-\newblock {\em medRxiv}.
-\newblock \url{https://www.medrxiv.org/content/10.1101/2021.02.14.21251708v1}.
-
-\bibitem{graham2021changes}
-Graham MS, Sudre CH, May A, Antonelli M, Murray B, Varsavsky T, Kl{\"a}ser K,
-  Canas LS, Molteni E, and others .
-\newblock 2021/05/03 2021 {Changes in symptomatology, reinfection, and
-  transmissibility associated with the SARS-CoV-2 variant B.1.1.7: an
-  ecological study}.
-\newblock {\em Lancet Public Health}.
-\newblock {\bfseries 6}, e335--e345.
-
-\bibitem{leung2021early}
-Leung K, Shum MH, Leung GM, Lam TT, and Wu~JT.
-\newblock 2021 {Early transmissibility assessment of the N501Y mutant strains
-  of SARS-CoV-2 in the United Kingdom, October to November 2020}.
-\newblock {\em Euro Surveill.}
-\newblock {\bfseries 26}, 2002106.
-
-\bibitem{volz2021transmission}
-Volz E, Mishra S, Chand M, Barrett JC, Johnson R, Geidelberg L, Hinsley WR,
-  Laydon DJ, Dabrera G, O'Toole {\'A}, and others .
-\newblock 2021 {Assessing transmissibility of SARS-CoV-2 lineage B.1.1.7 in
-  England}.
-\newblock {\em Nature}.
-
-\bibitem{zhao2021}
-Zhao S, Lou J, Cao L, Zheng H, Chong MKC, Chen Z, Chan RWY, Zee BCY, Chan PKS,
-  and Wang MH.
-\newblock 01 2021 {Quantifying the transmission advantage associated with N501Y
-  substitution of SARS-CoV-2 in the UK: an early data-driven analysis}.
-\newblock {\em J. Travel Med}.
-\newblock taab011.
-
-\bibitem{campbell2021increased}
-Campbell F, Archer B, Laurenson-Schafer H, Jinnai Y, Konings F, Batra N, Pavlin
-  B, Vandemaele K, Van~Kerkhove MD, Jombart T, and others .
-\newblock 2021 Increased transmissibility and global spread of {SARS-CoV-2}
-  variants of concern as at {June}, 2021.
-\newblock {\em Euro Surveill.}
-\newblock {\bfseries 26}, 2100509.
-
-\bibitem{diekmann1990definition}
-Diekmann O, Heesterbeek JAP, and Metz JA.
-\newblock 1990 On the definition and the computation of the basic reproduction
-  ratio $\mathcal{R}_0$ in models for infectious diseases in heterogeneous
-  populations.
-\newblock {\em J. Math. Biol.}
-\newblock {\bfseries 28}, 365--382.
-
-\bibitem{anderson1991infectious}
-Anderson RM and May RM.
-\newblock 1991 {\em {Infectious Diseases of Humans: Dynamics and Control}}.
-\newblock Oxford University Press.
-
-\bibitem{mills2004transmissibility}
-Mills CE, Robins JM, and Lipsitch M.
-\newblock 2004 Transmissibility of 1918 pandemic influenza.
-\newblock {\em Nature}.
-\newblock {\bfseries 432}, 904--906.
-
-\bibitem{nishiura2009transmission}
-Nishiura H, Castillo-Chavez C, Safan M, and Chowell G.
-\newblock 2009 {Transmission potential of the new influenza A (H1N1) virus and
-  its age-specificity in Japan}.
-\newblock {\em Euro Surveill.}
-\newblock {\bfseries 14}, 19227.
-
-\bibitem{ma2014estimating}
-Ma~J, Dushoff J, Bolker BM, and Earn DJD.
-\newblock 2014 Estimating initial epidemic growth rates.
-\newblock {\em Bull. Math. Biol.}
-\newblock {\bfseries 76}, 245--260.
-
-\bibitem{wallinga2007generation}
-Wallinga J and Lipsitch M.
-\newblock 2007 How generation intervals shape the relationship between growth
-  rates and reproductive numbers.
-\newblock {\em Proc. R. Soc. B}.
-\newblock {\bfseries 274}, 599--604.
-
-\bibitem{svensson2007note}
-Svensson {\AA}.
-\newblock 2007 A note on generation times in epidemic models.
-\newblock {\em Math. Biosci.}
-\newblock {\bfseries 208}, 300--311.
-
-\bibitem{fine2003interval}
-Fine PE.
-\newblock 2003 The interval between successive cases of an infectious disease.
-\newblock {\em Am. J. Epidemiol.}
-\newblock {\bfseries 158}, 1039--1047.
-
-\bibitem{grassly2008mathematical}
-Grassly NC and Fraser C.
-\newblock 2008 Mathematical models of infectious disease transmission.
-\newblock {\em Nat. Rev. Microbiol.}
-\newblock {\bfseries 6}, 477--487.
-
-\bibitem{britton2019estimation}
-Britton T and Scalia~Tomba G.
-\newblock 2019 {Estimation in emerging epidemics: Biases and remedies}.
-\newblock {\em J. R. Soc. Interface}.
-\newblock {\bfseries 16}, 20180670.
-
-\bibitem{ali2020serial}
-Ali ST, Wang L, Lau EH, Xu~XK, Du~Z, Wu~Y, Leung GM, and Cowling BJ.
-\newblock 2020 {Serial interval of SARS-CoV-2 was shortened over time by
-  nonpharmaceutical interventions}.
-\newblock {\em Science}.
-\newblock {\bfseries 369}, 1106--1109.
-
-\bibitem{park2021forward}
-Park SW, Sun K, Champredon D, Li~M, Bolker BM, Earn DJ, Weitz JS, Grenfell BT,
-  and Dushoff J.
-\newblock 2021 Forward-looking serial intervals correctly link epidemic growth
-  to reproduction numbers.
-\newblock {\em Proc. Natl. Acad. Sci. U.S.A.}
-\newblock {\bfseries 118}, e2011548118.
-
-\bibitem{lloyd2001realistic}
-Lloyd AL.
-\newblock 2001 Realistic distributions of infectious periods in epidemic
-  models: changing patterns of persistence and dynamics.
-\newblock {\em Theor. Popul. Biol.}
-\newblock {\bfseries 60}, 59--71.
-
-\bibitem{wearing2005appropriate}
-Wearing HJ, Rohani P, and Keeling MJ.
-\newblock 2005 Appropriate models for the management of infectious diseases.
-\newblock {\em PLoS medicine}.
-\newblock {\bfseries 2}, e174.
-
-\bibitem{roberts2007model}
-Roberts M and Heesterbeek J.
-\newblock 2007 Model-consistent estimation of the basic reproduction number
-  from the incidence of an emerging infection.
-\newblock {\em J. Math. Biol.}
-\newblock {\bfseries 55}, 803.
-
-\bibitem{huber2016quantitative}
-Huber JH, Johnston GL, Greenhouse B, Smith DL, and Perkins TA.
-\newblock 2016 Quantitative, model-based estimates of variability in the
-  generation and serial intervals of \textit{Plasmodium falciparum} malaria.
-\newblock {\em Malaria journal}.
-\newblock {\bfseries 15}, 1--12.
-
-\bibitem{miller2010epidemics}
-Miller JC, Davoudi B, Meza R, Slim AC, and Pourbohloul B.
-\newblock 2010 Epidemics with general generation interval distributions.
-\newblock {\em J. Theor. Biol.}
-\newblock {\bfseries 262}, 107--115.
-
-\bibitem{svensson2015influence}
-Svensson {\AA}.
-\newblock 2015 The influence of assumptions on generation time distributions in
-  epidemic models.
-\newblock {\em Math. Biosci.}
-\newblock 270, 81--89.
-
-\bibitem{park2019practical}
-Park SW, Champredon D, Weitz JS, and Dushoff J.
-\newblock 2019 A practical generation-interval-based approach to inferring the
-  strength of epidemics from their speed.
-\newblock {\em Epidemics}.
-\newblock 27, 12--18.
-
-\bibitem{eaton2014proportion}
-Eaton JW and Hallett TB.
-\newblock 2014 Why the proportion of transmission during early-stage {HIV}
-  infection does not predict the long-term impact of treatment on {HIV}
-  incidence.
-\newblock {\em Proc. Natl. Acad. Sci. U.S.A.}
-\newblock {\bfseries 111}, 16202--16207.
-
-\bibitem{powers2014impact}
-Powers KA, Kretzschmar ME, Miller WC, and Cohen MS.
-\newblock 2014 Impact of early-stage {HIV} transmission on treatment as
-  prevention.
-\newblock {\em Proc. Natl. Acad. Sci. U.S.A.}
-\newblock {\bfseries 111}, 15867--15868.
-
-\bibitem{weitz2015modeling}
-Weitz JS and Dushoff J.
-\newblock 2015 Modeling post-death transmission of {Ebola}: challenges for
-  inference and opportunities for control.
-\newblock {\em Sci. Rep.}
-\newblock {\bfseries 5}, 1--7.
-
-\bibitem{park2020time}
-Park SW, Cornforth DM, Dushoff J, and Weitz JS.
-\newblock 2020 The time scale of asymptomatic transmission affects estimates of
-  epidemic potential in the {COVID-19} outbreak.
-\newblock {\em Epidemics}.
-\newblock 31, 100392.
-
-\bibitem{doi:10.1098/rspb.2020.1556}
-Dushoff J and Park SW.
-\newblock 2021 Speed and strength of an epidemic intervention.
-\newblock {\em Proc. R. Soc. B}.
-\newblock {\bfseries 288}, 20201556.
-
-\bibitem{kissler2021densely}
-Kissler S, Fauver JR, Mack C, Tai CG, Breban MI, Watkins AE, Samant RM,
-  Anderson DJ, Ho~DD, Grubaugh ND, and Grad YH.
-\newblock 2021 {Densely sampled viral trajectories suggest longer duration of
-  acute infection with B.1.1.7 variant relative to non-B.1.1.7 SARS-CoV-2}.
-\newblock {\em medRxiv}.
-\newblock \url{https://www.medrxiv.org/content/10.1101/2021.02.16.21251535v1}.
-
-\bibitem{li2021viral}
-Li~B, Deng A, Li~K, Hu~Y, Li~Z, Xiong Q, Liu Z, Guo Q, Zou L, Zhang H, and
-  others .
-\newblock 2021 {Viral infection and Transmission in a large well-traced
-  outbreak caused by the Delta SARS-CoV-2 variant}.
-\newblock {\em medRxiv}.
-\newblock \url{https://www.medrxiv.org/content/10.1101/2021.07.07.21260122v1}.
-
-\bibitem{zhang2021transmission}
-Zhang M, Xiao J, Deng A, Zhang Y, Zhuang Y, Hu~T, Li~J, Tu~H, Li~B, Zhou Y, and
-  others .
-\newblock 2021 Transmission dynamics of an outbreak of the {COVID-19} {Delta}
-  variant {B.1.617.2}--{Guangdong Province, China, May--June 2021}.
-\newblock {\em China CDC Weekly}.
-\newblock 3, 584--586.
-
-\bibitem{goldstein2009reconstructing}
-Goldstein E, Dushoff J, Ma~J, Plotkin JB, Earn DJ, and Lipsitch M.
-\newblock 2009 Reconstructing influenza incidence by deconvolution of daily
-  mortality time series.
-\newblock {\em Proc. Natl. Acad. Sci. U.S.A.}
-\newblock {\bfseries 106}, 21825--21829.
-
-\bibitem{gostic2020practical}
-Gostic KM, McGough L, Baskerville EB, Abbott S, Joshi K, Tedijanto C, Kahn R,
-  Niehus R, Hay JA, De~Salazar PM, and others .
-\newblock 12 2020 Practical considerations for measuring the effective
-  reproductive number, $\mathcal{R}_t$.
-\newblock {\em PLOS Comp. Biol.}
-\newblock {\bfseries 16}, 1--21.
-
-\bibitem{heesterbeek1996concept}
-Heesterbeek J and Dietz K.
-\newblock 1996 The concept of $\mathcal{R}_0$ in epidemic theory.
-\newblock {\em Stat. Neerl.}
-\newblock {\bfseries 50}, 89--110.
-
-\bibitem{diekmann2000mathematical}
-Diekmann O and Heesterbeek JAP.
-\newblock 2000 {\em Mathematical Epidemiology of Infectious Diseases: Model
-  Building, Analysis and Interpretation}, volume~5.
-\newblock John Wiley \& Sons.
-
-\bibitem{roberts2004modelling}
-Roberts M.
-\newblock 2004 Modelling strategies for minimizing the impact of an imported
-  exotic infection.
-\newblock {\em Proc. R. Soc. B}.
-\newblock {\bfseries 271}, 2411--2415.
-
-\bibitem{aldis2005integral}
-Aldis G and Roberts M.
-\newblock 2005 An integral equation model for the control of a smallpox
-  outbreak.
-\newblock {\em Math. Biosci.}
-\newblock {\bfseries 195}, 1--22.
-
-\bibitem{breda2012formulation}
-Breda D, Diekmann O, De~Graaf W, Pugliese A, and Vermiglio R.
-\newblock 2012 {On the formulation of epidemic models (an appraisal of Kermack
-  and McKendrick)}.
-\newblock {\em J. Biol. Dyn.}
-\newblock {\bfseries 6}, 103--117.
-
-\bibitem{champredon2018equivalence}
-Champredon D, Dushoff J, and Earn DJ.
-\newblock 2018 Equivalence of the {Erlang}-distributed {SEIR} epidemic model
-  and the renewal equation.
-\newblock {\em SIAM J. Appl. Math}.
-\newblock {\bfseries 78}, 3258--3278.
-
-\bibitem{fraser2007estimating}
-Fraser C.
-\newblock 2007 Estimating individual and household reproduction numbers in an
-  emerging epidemic.
-\newblock {\em PloS one}.
-\newblock {\bfseries 2}.
-
-\bibitem{champredon2015intrinsic}
-Champredon D and Dushoff J.
-\newblock 2015 Intrinsic and realized generation intervals in
-  infectious-disease transmission.
-\newblock {\em Proc. R. Soc. B}.
-\newblock {\bfseries 282}, 20152026.
-
-\bibitem{champredon2018two}
-Champredon D, Li~M, Bolker BM, and Dushoff J.
-\newblock 2018 Two approaches to forecast {Ebola} synthetic epidemics.
-\newblock {\em Epidemics}.
-\newblock 22, 36--42.
-
-\bibitem{kenah2008generation}
-Kenah E, Lipsitch M, and Robins JM.
-\newblock 2008 Generation interval contraction and epidemic data analysis.
-\newblock {\em Math. Biosci.}
-\newblock {\bfseries 213}, 71--79.
-
-\bibitem{nishiura2010time}
-Nishiura H.
-\newblock 2010 Time variations in the generation time of an infectious disease:
-  implications for sampling to appropriately quantify transmission potential.
-\newblock {\em Math. Biosci. Eng.}
-\newblock {\bfseries 7}, 851.
-
-\bibitem{flaxman2020Rt}
-Flaxman S, Mishra S, Gandy A, Unwin HJT, Mellan TA, Coupland H, Whittaker C,
-  Zhu H, Berah T, Eaton JW, Monod M, and others .
-\newblock 2020 {Estimating the effects of non-pharmaceutical interventions on
-  COVID-19 in Europe}.
-\newblock {\em Nature}.
-\newblock {\bfseries 584}, 257--261.
-
-\bibitem{moore2021vaccination}
-Moore S, Hill EM, Tildesley MJ, Dyson L, and Keeling MJ.
-\newblock 2021 {Vaccination and non-pharmaceutical interventions for COVID-19:
-  a mathematical modelling study}.
-\newblock {\em Lancet Infect. Dis.}
-\newblock {\bfseries 21}, 793--802.
-
-\bibitem{brauner2021inferring}
-Brauner JM, Mindermann S, Sharma M, Johnston D, Salvatier J, Gaven{\v{c}}iak T,
-  Stephenson AB, Leech G, Altman G, Mikulik V, and others .
-\newblock 2021 {Inferring the effectiveness of government interventions against
-  COVID-19}.
-\newblock {\em Science}.
-\newblock {\bfseries 371}.
-
-\bibitem{doi:10.1098/rsif.2020.0144}
-Park SW, Bolker BM, Champredon D, Earn DJD, Li~M, Weitz JS, Grenfell BT, and
-  Dushoff J.
-\newblock 2020 {Reconciling early-outbreak estimates of the basic reproductive
-  number and its uncertainty: framework and applications to the novel
-  coronavirus (SARS-CoV-2) outbreak}.
-\newblock {\em J. R. Soc. Interface}.
-\newblock {\bfseries 17}, 20200144.
-
-\bibitem{ferretti2020quantifying}
-Ferretti L, Wymant C, Kendall M, Zhao L, Nurtay A, Abeler-D{\"o}rner L, Parker
-  M, Bonsall D, and Fraser C.
-\newblock 2020 {Quantifying SARS-CoV-2 transmission suggests epidemic control
-  with digital contact tracing}.
-\newblock {\em Science}.
-\newblock {\bfseries 368}.
-
-\bibitem{leung2017monitoring}
-Leung K, Lipsitch M, Yuen KY, and Wu~JT.
-\newblock 2017 Monitoring the fitness of antiviral-resistant influenza strains
-  during an epidemic: a mathematical modelling study.
-\newblock {\em Lancet Infect. Dis.}
-\newblock {\bfseries 17}, 339--347.
-
-\bibitem{leung2020empirical}
-Leung K, Pei Y, Leung GM, Lam TT, and Wu~JT.
-\newblock 2020 {Empirical transmission advantage of the D614G mutant strain of
-  SARS-CoV-2}.
-\newblock {\em medRxiv}.
-\newblock \url{https://www.medrxiv.org/content/10.1101/2020.09.22.20199810v1}.
-
-\bibitem{wallinga2004different}
-Wallinga J and Teunis P.
-\newblock 2004 Different epidemic curves for severe acute respiratory syndrome
-  reveal similar impacts of control measures.
-\newblock {\em Am. J. Epidemiol.}
-\newblock {\bfseries 160}, 509--516.
-
-\bibitem{cori2013new}
-Cori A, Ferguson NM, Fraser C, and Cauchemez S.
-\newblock 2013 A new framework and software to estimate time-varying
-  reproduction numbers during epidemics.
-\newblock {\em Am. J. Epidemiol.}
-\newblock {\bfseries 178}, 1505--1512.
-
-\bibitem{abbott2020estimating}
-Abbott S, Hellewell J, Thompson RN, Sherratt K, Gibbs HP, Bosse NI, Munday JD,
-  Meakin S, Doughty EL, Chun JY, and others .
-\newblock 2020 {Estimating the time-varying reproduction number of SARS-CoV-2
-  using national and subnational case counts}.
-\newblock {\em Wellcome Open Research}.
-\newblock {\bfseries 5}, 112.
-
-\bibitem{knight2020estimating}
-Knight J and Mishra S.
-\newblock 2020 {Estimating effective reproduction number using generation time
-  versus serial interval, with application to COVID-19 in the Greater Toronto
-  Area, Canada}.
-\newblock {\em Infect. Dis. Model.}
-\newblock 5, 889--896.
-
-\bibitem{li2021temporal}
-Li~Y, Campbell H, Kulkarni D, Harpur A, Nundy M, Wang X, Nair H, for COVID UN,
-  and others .
-\newblock 2021 {The temporal association of introducing and lifting
-  non-pharmaceutical interventions with the time-varying reproduction number
-  (R) of SARS-CoV-2: a modelling study across 131 countries}.
-\newblock {\em Lancet Infect. Dis.}
-\newblock {\bfseries 21}, 193--202.
-
-\bibitem{tsang2020effect}
-Tsang TK, Wu~P, Lin Y, Lau EH, Leung GM, and Cowling BJ.
-\newblock 2020 {Effect of changing case definitions for COVID-19 on the
-  epidemic curve and transmission parameters in mainland China: a modelling
-  study}.
-\newblock {\em The Lancet Public Health}.
-\newblock {\bfseries 5}, e289--e296.
-
-\bibitem{ukinvest}
-{Public Health England} .
-\newblock 2020 {Investigation of novel SARS-CoV-2 variant: Variant of Concern
-  202012/01}.
-\newblock
-  \url{https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/959426/Variant_of_Concern_VOC_202012_01_Technical_Briefing_5.pdf}.
-
-\bibitem{ganyani2020estimating}
-Ganyani T, Kremer C, Chen D, Torneri A, Faes C, Wallinga J, and Hens N.
-\newblock 2020 {Estimating the generation interval for coronavirus disease
-  (COVID-19) based on symptom onset data, March 2020}.
-\newblock {\em Euro Surveill.}
-\newblock {\bfseries 25}, 2000257.
-
-\bibitem{fraser2004factors}
-Fraser C, Riley S, Anderson RM, and Ferguson NM.
-\newblock 2004 Factors that make an infectious disease outbreak controllable.
-\newblock {\em Proc. Natl. Acad. Sci. U.S.A.}
-\newblock {\bfseries 101}, 6146--6151.
-
-\bibitem{scarabel2021renewal}
-Scarabel F, Pellis L, Ogden NH, and Wu~J.
-\newblock 2021 A renewal equation model to assess roles and limitations of
-  contact tracing for disease outbreak control.
-\newblock {\em R. Soc. Open Sci.}
-\newblock {\bfseries 8}, 202091.
-
-\bibitem{unwin2020state}
-Unwin HJT, Mishra S, Bradley VC, Gandy A, Mellan TA, Coupland H, Ish-Horowicz
-  J, Vollmer MA, Whittaker C, Filippi SL, and others .
-\newblock 2020 {State-level tracking of COVID-19 in the United States}.
-\newblock {\em Nat. Commun.}
-\newblock {\bfseries 11}, 1--9.
-
-\bibitem{brett2020transmission}
-Brett TS and Rohani P.
-\newblock 2020 Transmission dynamics reveal the impracticality of {COVID-19}
-  herd immunity strategies.
-\newblock {\em Proc. Natl. Acad. Sci. U.S.A.}
-\newblock {\bfseries 117}, 25897--25903.
-
-\bibitem{Ferretti2020timing}
-Ferretti L, Ledda A, Wymant C, Zhao L, Ledda V, Abeler-D{\"o}rner L, Kendall M,
-  Nurtay A, Cheng HY, Ng~TC, and others .
-\newblock 2020 The timing of {COVID-19} transmission.
-\newblock {\em medRxiv}.
-\newblock \url{https://www.medrxiv.org/content/10.1101/2020.09.04.20188516v2}.
-
-\bibitem{abdool2021new}
-Abdool~Karim SS and de~Oliveira T.
-\newblock 2021 {New SARS-CoV-2 variants—clinical, public health, and vaccine
-  implications}.
-\newblock {\em N. Engl. J. Med.}
-
-\bibitem{fontanet2021sars}
-Fontanet A, Autran B, Lina B, Kieny MP, Karim SSA, and Sridhar D.
-\newblock 2021 {SARS-CoV-2 variants and ending the COVID-19 pandemic}.
-\newblock {\em Lancet}.
-\newblock {\bfseries 397}, 952--954.
-
-\bibitem{walensky2021sars}
-Walensky RP, Walke HT, and Fauci AS.
-\newblock 2021 {SARS-CoV-2 variants of concern in the United
-  States---Challenges and opportunities}.
-\newblock {\em JAMA}.
-\newblock {\bfseries 325}, 1037--1038.
-
-\bibitem{hellewell2020feasibility}
-Hellewell J, Abbott S, Gimma A, Bosse NI, Jarvis CI, Russell TW, Munday JD,
-  Kucharski AJ, Edmunds WJ, Sun F, and others .
-\newblock 2020 {Feasibility of controlling COVID-19 outbreaks by isolation of
-  cases and contacts}.
-\newblock {\em Lancet Glob. Health}.
-\newblock {\bfseries 8}, e488--e496.
-
-\end{thebibliography}
+\bibliography{newvariant}
 

### 2021-11-30 newvariant.tex
done?

@@
 Inferring the relative strength (i.e., the ratio of reproduction numbers) and relative speed (i.e., the difference between growth rates) of new SARS-CoV-2 variants compared to earlier strains is critical to predicting and controlling the course of the current pandemic.
-Here, we challenge a commonly-made assumption (i.e., a constant relative speed) in estimating the transmission advantage of SARS-CoV-2 variants.
-We argue that holding relative strength constant is generally more appropriate by comparing how relative strength and speed change over time under two idealized interventions:
-a constant-strength intervention like idealized vaccination and social distancing, which reduces transmission rates by a constant amount, and a constant-speed intervention like idealized contact tracing, which isolate infected individuals at a constant rate.
-The impact of these interventions on the spread of new variants is evaluated using a generation-interval-based framework in light of recent evidence that the Alpha and Delta variants may have different generation intervals from prior dominant lineages.
-Differences in the generation-interval distributions affect not only the effectiveness of these interventions but also the realized transmission advantages of new variants.
+Here, we challenge a commonly-made assumption in estimating the transmission advantage of SARS-CoV-2 variants and argue that holding relative strength (rather than relative speed) constant is generally more appropriate.
+In doing so, we use the generation-interval-based framework to compare how relative strength and speed change over time under two idealized interventions:
+a constant-strength intervention like idealized vaccination and social distancing, which reduces transmission rates by a constant amount, and a constant-speed intervention like idealized contact tracing, which isolates infected individuals at a constant rate.
+Using the generation-interval-based framework further allows us to account for the possibility that new SARS-CoV-2 variants, such as the Alpha and Delta variants, may have different generation intervals from prior dominant lineages.
+Differences in the generation-interval distributions can affect the effectiveness of these interventions as well as the realized transmission advantages of new variants.
 Neglecting these differences gives biased estimates of relative strength, which can vary systematically with changes in the incidence of infections caused by the variant.
@@
 Like epidemic strength, epidemic speed also determines conditions for pathogen elimination: $r=0$ is a threshold equivalent to $\RR=1$ under constant conditions.
-Dushoff and Park \citep{doi:10.1098/rspb.2020.1556} argued previously that large segments of the epidemiology and modeling community have over-emphasized $\RR$ at the expense of $r$; 
+However, large segments of the epidemiology and modeling community have over-emphasized $\RR$ at the expense of $r$ \citep{doi:10.1098/rspb.2020.1556};
 we thus use the terms ``strength'' and ``speed''  here to underline our contention that these are better seen as complementary perspectives (and to link them to complementary perspectives on measuring the transmission advantage of new variants).

### 2021-11-30 newvariant.tex
done?

@@
 Instead, this calculation simply highlights the importance of carefully considering generation-interval distributions in assessing the relative strength and speed of SARS-CoV-2 variants.
-While we relied on parameter estimates for the Alpha variant throughout our analysis, our qualitative conclusions can be applied to studying other SARS-CoV-2 variants, including the Delta variant.
+We relied on parameter estimates for the Alpha variant throughout our analysis, but our qualitative conclusions can be applied to studying other SARS-CoV-2 variants, including the Delta variant.
 
@@
 
-We mostly considered idealized interventions of constant strength  or constant speed. Real interventions are likely more complex.
-Interventions that reduce contact rates, such as social distancing measures and vaccination, are generally strength-like, but they can also elicit awareness-based behavioral responses that are speed-like---for example, infected individuals may self-isolate faster after symptom onset.
+Other studies have also tried to explain apparent changes in their estimates of relative strength using generation-interval-based arguemtns but sometimes gave inaccurate explanations. 
+For example, Volz \textit{et al.} estimated that the relative strength of the Alpha variant declined in England between December 2020 and January 2021 \cite{volz2021transmission};
+they hypothesized that a shorter generation interval of the Alpha variant could explain this phenomenon by reducing the relative speed (and therefore, the relative strength) under intervention measures.
+However, our analysis suggests that a shorter generation interval of a new variant cannot explain the decline in relative strength.
+Under a constant-strength intervention, the relative speed decreases (as predicted by  \cite{volz2021transmission}) but the relative strength remains constant (\fref{control_sim}B). 
+Under a constant-speed intervention, the relative speed remains constant (\fref{control_sim}F), but the relative strength increases because the intervention will have a smaller effect on the variant with a shorter generation interval (\fref{control_sim}E).
+
+We mostly considered idealized interventions of constant strength or constant speed.
+Real interventions are likely more complex.
+Interventions that reduce contact rates, such as social distancing measures, are generally strength-like, but they can also elicit awareness-based behavioral responses that are speed-like---for example, infected individuals may self-isolate faster after symptom onset.
+Other strength-like interventions, such as vaccination, can also have disproportionately smaller effects on new variants depending on their antigenic variability.
 On the other hand, the hazard of isolation-based interventions (and thus, speed-like), such as contact tracing, can vary over the course of infection depending on delays in tracing infected individuals \citep{fraser2004factors,ferretti2020quantifying,scarabel2021renewal}.

### 2021-12-01 response_new.tex
new variant and new response

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}

\usepackage{color}

\usepackage{xspace}

\usepackage{lmodern}

\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}

\usepackage[pdfencoding=auto, psdextra]{hyperref}

\usepackage{natbib}
\bibliographystyle{chicago}

\newcommand{\eref}[1]{Eq.~(\ref{eq:#1})}
\newcommand{\fref}[1]{Fig.~\ref{fig:#1}}

%% Consistent, changeable style for subscripts
\newcommand{\vvvar}{\mathrm{var}}
\newcommand{\wwwt}{\mathrm{wt}}

\newcommand{\rx}[1]{\ensuremath{{r}_{#1}}\xspace} 
\newcommand{\ry}[1]{\rx{\mathrm{#1}}} 
\newcommand{\rw}{\rx{\wwwt}}
\newcommand{\rv}{\rx{\vvvar}}

\newcommand{\Rx}[1]{\ensuremath{{\mathcal R}_{#1}}\xspace} 
\newcommand{\Ry}[1]{\Rx{\mathrm{#1}}}
\newcommand{\Ro}{\Rx{0}}
\newcommand{\RR}{\ensuremath{{\mathcal R}}\xspace}
\newcommand{\Rw}{\Rx{\wwwt}}
\newcommand{\Rv}{\Rx{\vvvar}}

\newcommand{\dd}[1]{\ensuremath{\, \mathrm{d}#1}}
\newcommand{\dtau}{\dd{\tau}}
\newcommand{\dx}{\dd{x}}
\newcommand{\dsigma}{\dd{\sigma}}

\newcommand{\rev}{\subsection*}
\newcommand{\revtext}{\textsf}
\setlength{\parskip}{\baselineskip}
\setlength{\parindent}{0em}

\newcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textsl{#3}\textbf{]}}}
\newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}
\newcommand{\dc}[1]{\comment{blue}{DC}{#1}}
\newcommand{\jsw}[1]{\comment{green}{JSW}{#1}}
\newcommand{\hotcomment}[1]{\comment{red}{HOT}{#1}}

\newcommand{\psymp}{\ensuremath{p}} %% primary symptom time
\newcommand{\ssymp}{\ensuremath{s}} %% secondary symptom time
\newcommand{\pinf}{\ensuremath{\alpha_1}} %% primary infection time
\newcommand{\sinf}{\ensuremath{\alpha_2}} %% secondary infection time

\newcommand{\psize}{{\mathcal P}} %% primary cohort size
\newcommand{\ssize}{{\mathcal S}} %% secondary cohort size

\newcommand{\gtime}{\tau_{\rm g}} %% generation interval
\newcommand{\gdist}{g} %% generation-interval distribution
\newcommand{\idist}{\ell} %% incubation period distribution

\newcommand{\total}{{\mathcal T}} %% total number of serial intervals

\begin{document}

\noindent Dear Editor:

We are re-submitting our manuscript, “The importance of the generation interval in investigating dynamics and control of new SARS-CoV-2 variants”, for consideration for publication in JRSI. This material has not previously been published elsewhere.

We had previously submitted the manuscript to JRSI (rsif-2021-0438) and were asked to revise with positive comments---for example, reviewer 3 said “The analysis is mathematically sound and offers a thorough investigation of an important concept that has been the subject of much discussion in the epidemiological community but, to my knowledge, hasn't been given the formal treatment it deserves until now.” We had submitted the revised version of the manuscript but the paper was rejected despite having addressed all comments due to reviewer 2's comment that “On a second reading of this manuscript, I am much less optimistic about its value to infectious disease epidemiology.” We emailed the editor since then and were provided the opportunity to resubmit.  
We have made revisions to our manuscript to address the reviewer's comments and to better highlight the novelty of our work.

The ongoing COVID-19 pandemic underlines the importance of characterizing the spread of new variants. Previous studies have particularly prioritized estimating the relative strength (i.e., the ratio of reproduction numbers) from relative speed (i.e., the differences between growth rates).

In this study, we challenge a commonly made assumption (i.e., a constant relative speed) in estimating transmission advantage of new variants and argue that holding relative strength is generally more appropriate.
In doing so, we consider the possibility that new variants can have different generation intervals from the original variant and explore how such differences affect the dynamics and control of new variants. 
We also show that neglecting these differences can lead to biased inference and provide a practical guidance for how such biases can be assessed in practice.
Below please find our detailed responses to reviewers.

The manuscript and its findings should be of interest to a broad range of readers, particularly given its theoretical and practical implications for characterizing the spread of new variants. All code is available as open-source for use, re-use, and adaptation by the community. Please do not hesitate to contact us if you have any questions.

Thank you for your consideration of our submission.

Sincerely,

Sang Woo Park

\rev{Reviewer \#2}

\revtext{On a second reading of this manuscript, I am much less optimistic about its value to infectious disease epidemiology. The analysis is sound, and the revision does a better job of articulating the assumptions that are made along the way---addressing one of the main concerns in my previous review. However, the paper does not seem to make any particularly useful novel contribution to the field. The importance of assumptions about the generation interval distribution for estimates of reproduction numbers (based on growth rates) and growth rates (based on reproduction numbers) is well-established. The authors are right that this is often neglected in practice, and it is interesting to see how this plays out in their simulation studies.}

Thank you for your review.
We agree that the role of generation-interval distribution for estimates of reproduction numbers is well-established. We also agree that this is often neglected in practice. 
While we emphasize the role of generation-interval distribution in estimating transmission advantage of new variants, this conclusion is secondary to the main point of our paper. 
The goal is to challenge the idea of holding relative speed (differences in growth rates) constant.
In particular, we show that the relative speed changes under a constant-strength intervention, which reduces transmission rate by a constant amount and therefore keeps the relative strength constant---this effect is present even when the generation-interval distributions are identical and is exaggerated in the presence of differences in the generation-interval distributions.
We believe that using the generation-interval-based framework allows us to make this argument clearly.
We have re-written large segments of the abstract, introduction, and discussion to make the contribution of our work clear.

\revtext{It is also well-established that the generation interval distribution is an important determinant of the effectiveness of interventions like contact tracing and isolation, which is also borne out in these simulations.}

We also agree that the role of generation-interval distribution in determining the effectiveness of interventions has been well established.
However, we believe that there is still confusion in the literature that need to be clarified---in particular, the effects of different interventions (speed- vs strength-like interventions) as well as differences in the generation-interval distributions on the transmission advantage of new variants has not yet been formally investigated.
We have tried to highlight this throughout the text.
We have also added a paragraph in the Discussion section to illustrate the confusion in the literature and how our framework provides the correct explanation:

``
Other studies have also tried to explain apparent changes in their estimates of relative strength using generation-interval-based arguments but sometimes gave inaccurate explanations. 
For example, Volz \textit{et al.} estimated that the relative strength of the Alpha variant declined in England between December 2020 and January 2021 \cite{volz2021transmission};
they hypothesized that a shorter generation interval of the Alpha variant could explain this phenomenon by reducing the relative speed (and therefore, the relative strength) under intervention measures.
However, our analysis suggests that a shorter generation interval of a new variant cannot explain the decline in relative strength.
Under a constant-strength intervention, the relative speed decreases (as predicted by  \cite{volz2021transmission}) but the relative strength remains constant (Figure 5B). 
Under a constant-speed intervention, the relative speed remains constant (Figure 5F), but the relative strength increases because the intervention will have a smaller effect on the variant with a shorter generation interval (Figure 5E).''

\revtext{The use of ``strength'' to refer to reproduction numbers and ``speed'' to refer to growth rates is more confusing than helpful, and the related concepts of ``strength-like''/``constant-strength'' and ``speed-like''/``constant-speed'' interventions are not defined clearly until late in the manuscript. In my opinion, the manuscript does not make a compelling case that these are useful novel concepts for the theory or practice of infectious disease epidemiology.}

This concern was brought up during the first round of reviews: ``From the examples, it appears that
speed-like interventions work by identifying potential infections quickly
(effectively ending the infectious period early while not necessarily altering
the hazard of infectious contact) and strength-like interventions work by
reducing the hazard of infectious contact (while not necessarily altering
the duration of infectiousness). A good explanation of these ideas is on
lines 228–231 on page 10.''

We had included these exact verbal definitions to the introduction, but these revisions seem to have been neglected. Instead, we now provide a clear mathematical definitions of constant-strength and constant-speed interventions in the introduction. We have also added the following two sentences in the first paragraph of the introduction to justify our uses of ``strength'' and ``speed'':

``However, large segments of the epidemiology and modeling community have over-emphasized $\RR$ at the expense of $r$ \citep{doi:10.1098/rspb.2020.1556};
we thus use the terms ``strength'' and ``speed''  here to underline our contention that these are better seen as complementary perspectives (and to link them to complementary perspectives on measuring the transmission advantage of new variants).''

\rev{Reviewer \#3}

\revtext{The authors have sufficiently addressed my concerns; I have no further comments.}

Thank you for your review.

\bibliography{newvariant_abbv.bib}

\end{document}

### 2021-12-01 newvariant.tex
new variant and new response

@@
 We explored how the generation-interval distribution shapes the link between relative strength and speed.
-While the role of generation-interval distributions in linking epidemic strength and speed has been well-established, using the generation-interval-based framework provides clear expanations for an imporant, but under-appreciated concept: the relative strength and speed of new variants also depend on epidemic dynamics of the prior dominant lineages as well as the type of control measures (i.e., strength- or speed-like) that limit their spread.
+While the role of generation-interval distributions in linking epidemic strength and speed has been well-established, using the generation-interval-based framework provides clear explanations for an important, but under-appreciated concept: the relative strength and speed of new variants also depend on epidemic dynamics of the prior dominant lineages as well as the type of control measures (i.e., strength- or speed-like) that limit their spread.
 For example, under a constant-strength intervention that reduces the transmission rate of the wild type and a variant by the same amount (and therefore, keeping the relative strength constant), the relative speed of the variant is sensitive to the epidemic growth rate of the wild type strain---this challenges the commonly-made assumption that relative speed of new variants remains constant over time.
@@
 
-Other studies have also tried to explain apparent changes in their estimates of relative strength using generation-interval-based arguemtns but sometimes gave inaccurate explanations. 
+Other studies have also tried to explain apparent changes in their estimates of relative strength using generation-interval-based arguments but sometimes gave inaccurate explanations. 
 For example, Volz \textit{et al.} estimated that the relative strength of the Alpha variant declined in England between December 2020 and January 2021 \cite{volz2021transmission};

### 2021-12-19 newvariant.tex
some rewrites

@@
 Inferring the relative strength (i.e., the ratio of reproduction numbers) and relative speed (i.e., the difference between growth rates) of new SARS-CoV-2 variants compared to earlier strains is critical to predicting and controlling the course of the current pandemic.
-Here, we challenge a commonly-made assumption in estimating the transmission advantage of SARS-CoV-2 variants and argue that holding relative strength (rather than relative speed) constant is generally more appropriate.
-In doing so, we use the generation-interval-based framework to compare how relative strength and speed change over time under two idealized interventions:
+Analyses of new variants have primarily focused on characterizing changes in the \emph{proportion} of new variants, implicitly assuming that the relative speed remains fixed over the course of an invasion.
+We use the generation-interval-based framework to challenge this assumption and illustrate how relative strength and speed change over time under two idealized interventions:
 a constant-strength intervention like idealized vaccination and social distancing, which reduces transmission rates by a constant amount, and a constant-speed intervention like idealized contact tracing, which isolates infected individuals at a constant rate.
-Using the generation-interval-based framework further allows us to account for the possibility that new SARS-CoV-2 variants, such as the Alpha and Delta variants, may have different generation intervals from prior dominant lineages.
-Differences in the generation-interval distributions can affect the effectiveness of these interventions as well as the realized transmission advantages of new variants.
-Neglecting these differences gives biased estimates of relative strength, which can vary systematically with changes in the incidence of infections caused by the variant.
+A constant-strength intervention causes relative speed to change, whereas a constant-speed intervention causes relative strength to change.
+Differences in the generation-interval distributions between variants can exaggerate these changes and affect the effectiveness of interventions.
+Neglecting the differences in the generation-interval distributions differences gives biased estimates of relative strength, which can vary systematically with changes in the incidence of infections caused by the variant.
 
@@
 Epidemic speed can be typically estimated from time series of incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}, but epidemic strength is difficult to measure from incidence time series.
-Instead, epidemic strength is often inferred from the observed epidemic speed using the generation-interval distribution, an approach popularized by \citep{wallinga2007generation}.
+Instead, epidemic strength is often inferred from the observed epidemic speed using the generation-interval distribution $g(\tau)$, an approach popularized by \citep{wallinga2007generation}.
 The generation interval, defined as the time between infection and transmission, provides information about the time scale of individual-level transmission \citep{svensson2007note}.
@@
 
-Analyses of new variants have typically characterized \emph{relative} strength (i.e., the ratio of reproduction numbers $\Rv/\Rw$, here called $\rho$) and speed (i.e., the difference between growth rates $\rv-\rw$, here called $\delta$) of the variants.
-Many studies have tried to estimate the relative strength of variants from the observed relative speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021,campbell2021increased}.
-Instead (or additionally), some studies have assumed a value of the relative strength and tried to predict relative speed \citep{davies2021estimated,di2021impact}.
+Analyses of new variants have characterized \emph{relative} strength (i.e., the ratio of reproduction numbers $\Rv/\Rw$, here called $\rho$) and speed (i.e., the difference between growth rates $\rv-\rw$, here called $\delta$) of the variants, where $\Rv$, $\Rw$ and $\rv$, $\rw$ represent reproduction numbers and growth rates at a given time.
+Many analyses have particularly focused on changes in the \emph{proportion} of a new variant to estimate the relative speed \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}, instead of taking the difference between their observed speed.
+Focusing on proportions can be advantageous, because changes in proportions may be less sensitive to changes in testing and to other transient effects that would affect variants and wild type viruses
+similarly;
+however, estimates of relative speed from changes in the proportion of a new variant have typically relied on the assumption that the relative speed remains fixed over the time scale of an invasion.
+Instead (or additionally), some studies have assumed a fixed value of the relative strength and tried to predict relative speed \citep{davies2021estimated,di2021impact}.
 
@@
 To illustrate the differences in implications for holding $\RR$ or $r$ fixed, we consider two idealized interventions of constant strength and constant speed in this paper.
-Given a pre-intervention kernel $K(\tau)$, which represents the rate at which an infected individual generates secondary infections $\tau$ time units after infection, a constant-strength intervention reduces transmission by a constant factor $\theta$ regardless of age of infection ($K(\tau)/\theta$)---in this case, the intervention strength $\theta$ must be greater than the epidemic strength $\RR$ to prevent onward transmission.
-On the other hand, a constant-speed intervention reduces transmission after infection by a constant rate $\phi$: $K(\tau) \exp(-\phi \tau)$---in this case, the intervention speed $\phi$ must be greater than the epidemic speed $r$ to prevent onward transmission \citep{doi:10.1098/rspb.2020.1556}.
+Given a pre-intervention kernel $K(\tau) = \RR g(\tau)$, which represents the rate at which an infected individual generates secondary infections $\tau$ time units after infection, a constant-strength intervention reduces transmission by a constant factor $\theta$ across generation intervals ($K(\tau)/\theta$)---in this case, the intervention strength $\theta$ must be greater than the epidemic strength $\RR$ to prevent onward transmission.
+On the other hand, a constant-speed intervention reduces transmission after infection by a constant rate $\phi$ across generation intervals: $K(\tau) \exp(-\phi \tau)$---in this case, the intervention speed $\phi$ must be greater than the epidemic speed $r$ to prevent onward transmission \citep{doi:10.1098/rspb.2020.1556}.
 
@@
 
-Most studies have assumed that the previous dominant strain (wild type) and variants have identical generation-interval distributions, but Kissler \textit{et al.} \citep{kissler2021densely} suggested that the Alpha variant may have a longer duration of infection: 13.3 days (90\% CI: 10.1--16.5) for the new variant and 8.2 days (90\% CI: 6.5--9.7) for the wild type.
+Recent studies have further suggested the possibility that new variants may have different generation-interval distributions.
+For example, Kissler \textit{et al.} \citep{kissler2021densely} suggested that the Alpha variant may have a longer duration of infection: 13.3 days (90\% CI: 10.1--16.5) for the Alpha variant and 8.2 days (90\% CI: 6.5--9.7) for the wild type.
 Longer duration of infection suggests that the mean generation interval of the Alpha variant is likely to be longer than that of the wild type \citep{lloyd2001realistic,wearing2005appropriate,roberts2007model}.
-In contrast, some studies have suggested that the more recent Delta variant may have a shorter generation interval \citep{li2021viral,zhang2021transmission}.
+In contrast, some studies have suggested that the more recent Delta variant may have a shorter generation interval due to faster within-host viral replication \citep{li2021viral,zhang2021transmission}.
 Modeling studies have also considered the possibility that the faster growth rate of new variants may be driven, in part, by shorter generation intervals \citep{davies2021estimated,volz2021transmission}.
@@
 for example, self-isolation after symptom onset will lead to shorter generation intervals.
+The recent emergence of a new Omicron variant, and its breakthrough infections in previously immune individuals, adds further uncertainty to individual-level transmission dynamics and therefore the generation-interval of SARS-CoV-2 variants \citep{pulliam2021increased}.
 
@@
 
-Under constant-strength changes that reduce the transmission rate by a constant amount regardless of the age of infection, the instantaneous generation-interval distribution does not change through time \citep{fraser2007estimating}.
+Under constant-strength changes that reduce the transmission rate by a constant amount across generation intervals, the instantaneous generation-interval distribution does not change through time \citep{fraser2007estimating}.
 In this case, the instantaneous generation-interval distribution is also often referred to as the intrinsic generation-interval distribution \citep{champredon2015intrinsic,champredon2018two,gostic2020practical,park2020time}---
@@
 While epidemic speed $r_x$ can often be estimated directly from incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating},
-studies of new SARS-CoV-2 variants have mostly focused on characterizing changes in the \emph{proportion} of a new variant \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
-Focusing on proportions is advantageous, because changes in proportions are less sensitive to changes in testing and to other transient effects that would affect variants and wild type viruses similarly.
+studies of new SARS-CoV-2 variants have mostly focused on characterizing changes in the proportion of a new variant \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
 When incidence is changing exponentially ($i_x(t) = i_x(t_0) \exp(r_x t)$), the proportion of the new variant $p(t)$ follows a logistic growth curve \citep{switzerland2021variant,davies2021estimated}:
@@
 % We thus ask: what factors affect the relative strength $\rho = \Rv/\Rw$ of a new variant, conditional on an observed relative speed $\delta$?
-Inference of the relative strength $\rho = \Rv/\Rw$ from the observed relative speed $\delta$ depends on assumptions about the generation-interval distributions of both strains.
+Inference of the relative strength $\rho = \Rv/\Rw$ from the relative speed $\delta$ depends on assumptions about the generation-interval distributions of both strains.
 Given the mean generation interval of the variant $\Gv$ and the wild type $\Gw$, the relative strength $\rho = \Rv/\Rw$ under the Gamma assumption \citep{park2019practical} is given by:
@@
 
-We do not generally expect the relative speed $\delta$ to remain constant if other factors governing epidemic spread are changing.
-Instead, many biological mechanisms appear compatible with assuming a constant value of relative strength $\rho$ over changing conditions.
-For example, if the proportion of the population susceptible declines, or the average contact rate changes, while other factors remain constant, the relative strength $\rho$ is expected to remain constant \citep{leung2017monitoring,leung2020empirical,di2021impact,leung2021early};
+We do not generally expect the relative speed $\delta$ to remain fixed if other factors governing epidemic spread are changing on the time scale of the invasion of a new variant relative to that of a wild type.
+Instead, many biological mechanisms appear compatible with assuming a fixed value of relative strength $\rho$ over changing conditions.
+For example, if the proportion of the population susceptible declines, or the average contact rate changes, while other factors remain unchanging, the relative strength $\rho$ is expected to remain fixed \citep{leung2017monitoring,leung2020empirical,di2021impact,leung2021early};
 as we show in this section, these changes will imply a change in relative speed $\delta$.
 
-We thus investigate how $\delta$ is expected to change with \Rw when $\rho$ remains constant, and how this expectation changes with the ratio of the generation intervals. 
+We thus investigate how $\delta$ is expected to change with \Rw when $\rho$ remains fixed, and how this expectation changes with the ratio of the generation intervals. 
 Once again, we rely on the Gamma assumption \citep{park2019practical} to find the relative speed $\delta$ given the mean generation interval of the variant $\Gv$ and the wild type $\Gw$:
@@
 For example, even when the generation-interval distributions are identical ($\Gv=\Gw$, in this case), changing $\Rw$ from 0.49 to 1.27 (and thus $\Rv$ from 0.79 to 2.04) yields a relative speed $\delta$ between 0.08--0.11 for $\kappa=0.2$ and 0.06--0.14 for $\kappa=0.5$.
-Therefore, characterizing the spread of variants by assuming constant relative speed (e.g., by fitting a standard logistic growth curve \citep{switzerland2021variant} and using the resulting value of $\delta$) should be done with care.
+Differences in the generation-interval distributions exaggerate these changes.
+Therefore, characterizing changes in the proportion of variants by assuming a fixed relative speed (e.g., by fitting a standard logistic growth curve \citep{switzerland2021variant} and using the resulting value of $\delta$) should be done with care.
 
@@
 
-Instead of inferring relative strength from speed, one can directly estimate time-varying reproduction numbers $\RR(t)$ of the variant and the wild type from incidence data and directly compare their ratios when the incidence of infection and the generation-interval distributions are known---such methods have been used in previous analyses of the Alpha variant by \cite{volz2021transmission}.
+Instead of inferring relative strength from speed, one can directly estimate time-varying reproduction numbers $\RR(t)$ of the variant and the wild type from incidence data and directly compare their ratios when the incidence of infection and the generation-interval distributions are known---such methods have been used in previous analyses of the Alpha \citep{volz2021transmission} and Delta variants \citep{abbott2021estimating}.
 Broadly, there are two types of time-varying reproduction numbers: case reproduction number and instantaneous reproduction number.
@@
 We consider two types of intervention:
-an intervention of constant strength (\fref{strengthspeed}A,C,E), which reduces transmission by a constant factor $\theta$ regardless of age of infection ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau)/\theta$); and an intervention of constant speed (\fref{strengthspeed}B,D,F), which reduces transmission after infection by a constant rate $\phi$ ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau) \exp(-\phi \tau)$).
+an intervention of constant strength (\fref{strengthspeed}A,C,E), which reduces transmission by a constant factor $\theta$ regardless of age of infection ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau)/\theta$); and an intervention of constant speed (\fref{strengthspeed}B,D,F), which reduces transmission after infection by a constant rate $\phi$ ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau) \exp(-\phi \tau)$);
+both of these interventions are constant across generation intervals, but not necessarily across calendar time. 
 In this case, we can control the spread of the variant when $\theta > \Rv$ or $\phi > \rv$, respectively \citep{doi:10.1098/rspb.2020.1556}.
@@
 While the role of generation-interval distributions in linking epidemic strength and speed has been well-established, using the generation-interval-based framework provides clear explanations for an important, but under-appreciated concept: the relative strength and speed of new variants also depend on epidemic dynamics of the prior dominant lineages as well as the type of control measures (i.e., strength- or speed-like) that limit their spread.
-For example, under a constant-strength intervention that reduces the transmission rate of the wild type and a variant by the same amount (and therefore, keeping the relative strength constant), the relative speed of the variant is sensitive to the epidemic growth rate of the wild type strain---this challenges the commonly-made assumption that relative speed of new variants remains constant over time.
+For example, under a constant-strength intervention that reduces the transmission rate of the wild type and a variant by the same amount (and therefore, keeping the relative strength constant), the relative speed of the variant is sensitive to the epidemic growth rate of the wild type strain.
+Our results challenge the commonly-made assumption that relative speed of new variants remains constant over time when characterizing changes in the proportion of new variants.
 Differences in the generation-interval distributions further exaggerate this effect.
 
-Therefore, neglecting potential differences in the mean generation intervals between the variant and the wild type can bias estimates of the relative strength from incidence data.
+Therefore, neglecting potential differences in the mean generation intervals between the variant and the wild type can bias estimates of the its transmission advantage.
 These biases may be assessed by considering whether estimates of relative strength appear to vary systematically with the direction of changes in the incidence of infections caused by the variant.
@@
 
-As new variants of SARS-CoV-2 are spreading and becoming dominant in many countries, it is clear that the variants are more transmissible (have higher strength) than the wild type \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
-While our analysis supports estimates of a higher strength of the new variant across a wide range of assumptions, it also shows that uncertainty in generation-interval distributions must be taken into account to obtain accurate estimates of the relative strength of variants (and the corresponding uncertainty).
-If new variants have different infectiousness profiles \citep{kissler2021densely,li2021viral,zhang2021transmission}, and therefore generation-interval distributions, current estimates that do not account for different generation-interval distributions may be systematically biased.
-% If new variants escape vaccinal or natural immunity to some extent \citep{Kupferschmidt329}, their relative strength may increase further.
-
-There is currently a (minor) discrepancy in the estimates of relative strength of new variants, particularly for the Alpha variant.
-Mathematical analyses have typically reported greater than a 1.4 fold increase in reproduction number for the Alpha variant ($\rho > 1.4$) whereas an independent analysis of secondary attack rates from contact tracing data suggests a 1.25--1.4 fold increase \citep{ukinvest}.
+There were a (minor) discrepancy in the estimates of relative strength of the Alpha variant.
+Mathematical analyses typically reported greater than a 1.4 fold increase in reproduction number for the Alpha variant ($\rho > 1.4$) whereas an independent analysis of secondary attack rates from contact tracing data suggested a 1.25--1.4 fold increase \citep{ukinvest}.
 As shown in \cite{davies2021estimated} and \cite{volz2021transmission}, propagating uncertainty in generation interval estimates (and, in particular, assuming shorter generation intervals) can partially explain these differences:
@@
 Instead, this calculation simply highlights the importance of carefully considering generation-interval distributions in assessing the relative strength and speed of SARS-CoV-2 variants.
-We relied on parameter estimates for the Alpha variant throughout our analysis, but our qualitative conclusions can be applied to studying other SARS-CoV-2 variants, including the Delta variant.
+We relied on parameter estimates for the Alpha variant throughout our analysis, but our qualitative conclusions can be applied to studying other SARS-CoV-2 variants, including the Delta and Omicron  variant.
+
+Other studies have also tried to explain apparent changes in their estimates of relative strength using generation-interval-based arguments. 
+For example, Volz \textit{et al.} estimated that the relative strength of the Alpha variant declined in England between December 2020 and January 2021 \cite{volz2021transmission};
+they hypothesized that a shorter generation interval of the Alpha variant could explain this phenomenon by reducing the relative speed (and therefore, the relative strength) under intervention measures.
+However, our analysis suggests that a shorter generation interval of a new variant cannot explain the decline in relative strength.
+Under a constant-strength intervention, the relative speed decreases (as predicted by  \cite{volz2021transmission}) but the relative strength remains constant (\fref{control_sim}B). 
+Under a constant-speed intervention, the relative speed remains constant (\fref{control_sim}F), but the relative strength increases because the intervention will have a smaller effect on the variant with a shorter generation interval (\fref{control_sim}E).
+
+We have primarily focused on the differeces in the mean generation interval but the differences in the amount of variability, characeterized by the squared coefficient of variation $\kappa$, can also have important, but different implications \citep{blanquart2021selection}.
+For example, if two variants have identical reproduction numbers, the variant with a wider generation-interval distribution will always have a faster growth rate and outcompete the other variant (except when both variants have $\RR=1$).
+This is because a variant with a wider generation-interval distribution can take advantage of more early transmission during the growth phase and more late transmission during the decay phase.s
 
@@
 As we demonstrated, if the variant and the wild type have different generation intervals, speed-like interventions such as contact tracing can affect them differently, causing their relative strength to change over time (but not necessarily their relative speed).
-
-Other studies have also tried to explain apparent changes in their estimates of relative strength using generation-interval-based arguments but sometimes gave inaccurate explanations. 
-For example, Volz \textit{et al.} estimated that the relative strength of the Alpha variant declined in England between December 2020 and January 2021 \cite{volz2021transmission};
-they hypothesized that a shorter generation interval of the Alpha variant could explain this phenomenon by reducing the relative speed (and therefore, the relative strength) under intervention measures.
-However, our analysis suggests that a shorter generation interval of a new variant cannot explain the decline in relative strength.
-Under a constant-strength intervention, the relative speed decreases (as predicted by  \cite{volz2021transmission}) but the relative strength remains constant (\fref{control_sim}B). 
-Under a constant-speed intervention, the relative speed remains constant (\fref{control_sim}F), but the relative strength increases because the intervention will have a smaller effect on the variant with a shorter generation interval (\fref{control_sim}E).
 
@@
 This study has practical implications for analyzing the epidemiological dynamics of new variants.
-First, models that assume a constant relative speed, such as the standard logistic growth model, should be used with care---it is important to remember that the relative speed is expected to change with epidemiological conditions.
+First, models that assume time-invariant relative speed, such as the standard logistic growth model, should be used with care---it is important to remember that the relative speed is expected to change with epidemiological conditions.
 Early in the spread of an emerging variant, it is likely more convenient to fix the relative speed, given that speed can be directly observed.
@@
 
-The spread of new SARS-CoV-2 variants and the replacement of previously dominant lineages represent ongoing challenges for controlling the SARS-CoV-2 pandemic \citep{abdool2021new,fontanet2021sars,walensky2021sars}.
+The spread of new SARS-CoV-2 variants and the replacement of previously dominant lineages represent ongoing challenges for controlling the SARS-CoV-2 pandemic \citep{abdool2021new,fontanet2021sars,walensky2021sars,pulliam2021increased}.
 By explicitly considering epidemiological context and generation interval differences together, we have shown that improving estimates of the the relative duration of infectiousness at the individual scale may help guide more effective interventions.
@@
 Most intervention strategies throughout the current pandemic have focused on strength-like interventions \citep{flaxman2020Rt}, such as lock-downs, partly because pre-symptomatic transmission of SARS-CoV-2 has limited the effectiveness of contact tracing efforts \citep{hellewell2020feasibility}.
-However, given the possibility that new variants can have different infection characteristics \citep{kissler2021densely}, future studies should consider whether their transmission dynamics also differ (e.g., the amount of pre-symptomatic transmission) and evaluate intervention strategies accordingly.
+However, given the possibility that new variants can have different infection characteristics, future studies should consider whether their transmission dynamics also differ (e.g., the amount of pre-symptomatic transmission) and evaluate intervention strategies accordingly.
 
@@
 
-\bibliography{newvariant}
+\bibliography{newvariant_abbv}
 

### 2021-12-19 response_new.tex
forgot to commit this
changes changes changes

@@
 Previous studies have particularly prioritized estimating the relative strength (i.e., the ratio of reproduction numbers) from relative speed (i.e., the differences between growth rates).
-
-In this study, we challenge a commonly made assumption (i.e., a constant relative speed) in estimating transmission advantage of new variants and argue that holding relative strength is generally more appropriate.
+In doing so, analyses of new variants have focused on estimating the relative speed from the observed changes in the proportion of new variants while assuming that relative speed does not change over the course of invasion.
+Here, we challenge this assumption by showing how relative strength and speed can change in response to different kinds of epidemic interventions and argue that holding relative strength fixed is generally more appropriate in the context of modeling the spread of SARS-CoV-2 variants.
 In doing so, we consider the possibility that new variants can have different generation intervals from the original variant and explore how such differences affect the dynamics and control of new variants.
@@
 We believe that using the generation-interval-based framework allows us to make this argument clearly.
-We have re-written large segments of the abstract, introduction, and discussion to make the contribution of our work clear.
+We have re-written segments of the abstract, introduction, and discussion to make the contribution of our work clear.
+We have also included brief reviews of more recent papers.
 

### 2021-12-19 newvariant.tex
changes changes changes

@@
 
-We have primarily focused on the differeces in the mean generation interval but the differences in the amount of variability, characeterized by the squared coefficient of variation $\kappa$, can also have important, but different implications \citep{blanquart2021selection}.
-For example, if two variants have identical reproduction numbers, the variant with a wider generation-interval distribution will always have a faster growth rate and outcompete the other variant (except when both variants have $\RR=1$).
+We have primarily focused on the differences in the mean generation interval but the differences in the amount of variability, characterized by the squared coefficient of variation $\kappa$, can also have important, but different implications \citep{blanquart2021selection}.
+For example, if two variants have identical reproduction numbers, the variant with a wider generation-interval distribution will always have a faster growth rate and out-compete the other variant (except when both variants have $\RR=1$).
 This is because a variant with a wider generation-interval distribution can take advantage of more early transmission during the growth phase and more late transmission during the decay phase.s

### 2021-12-19 newvariant.tex
tweak

@@
 Inferring the relative strength (i.e., the ratio of reproduction numbers) and relative speed (i.e., the difference between growth rates) of new SARS-CoV-2 variants compared to earlier strains is critical to predicting and controlling the course of the current pandemic.
-Analyses of new variants have primarily focused on characterizing changes in the \emph{proportion} of new variants, implicitly assuming that the relative speed remains fixed over the course of an invasion.
+Analyses of new variants have primarily focused on characterizing changes in the \emph{proportion} of new variants, implicitly or explicitly assuming that the relative speed remains fixed over the course of an invasion.
 We use the generation-interval-based framework to challenge this assumption and illustrate how relative strength and speed change over time under two idealized interventions:

### 2022-01-12 newvariant.tex
some thoughts, comments and suggestions

@@
 \newcommand{\djde}[1]{\comment{red}{DJDE}{#1}}
+\newcommand{\old}{\sffamily}
+\newcommand{\told}{\textsf}
 
@@
 We use the generation-interval-based framework to challenge this assumption and illustrate how relative strength and speed change over time under two idealized interventions:
-a constant-strength intervention like idealized vaccination and social distancing, which reduces transmission rates by a constant amount, and a constant-speed intervention like idealized contact tracing, which isolates infected individuals at a constant rate.
-A constant-strength intervention causes relative speed to change, whereas a constant-speed intervention causes relative strength to change.
+a constant-strength intervention like idealized vaccination or social distancing, which reduces transmission rates by a constant proportion, and a constant-speed intervention like idealized contact tracing, which isolates infected individuals at a constant rate.
+In general, a constant-strength intervention causes relative speed to change, while a constant-speed intervention causes relative strength to change.
 Differences in the generation-interval distributions between variants can exaggerate these changes and affect the effectiveness of interventions.
-Neglecting the differences in the generation-interval distributions differences gives biased estimates of relative strength, which can vary systematically with changes in the incidence of infections caused by the variant.
+Neglecting differences in generation-interval distributions can bias estimates of relative strength.
+\jd{Can we delete this?
+Is there a a good way to clarify?: \old\ which can vary systematically with changes in the incidence of infections caused by the variant.}
 
@@
 
-Epidemic speed can be typically estimated from time series of incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}, but epidemic strength is difficult to measure from incidence time series.
+Epidemic speed is typically estimated from time series of incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}, but epidemic strength is difficult to measure from incidence time series.
 Instead, epidemic strength is often inferred from the observed epidemic speed using the generation-interval distribution $g(\tau)$, an approach popularized by \citep{wallinga2007generation}.
@@
 While it is possible to consider general forms of generation-interval distributions \citep{miller2010epidemics,svensson2015influence}, summarizing the distribution in terms of its mean and variability---for example, by assuming they are Gamma distributed---can still provide a robust link between epidemic speed and strength for real pathogens and yield important biological insights \citep{park2019practical}.
-In particular, several studies have noted, in many contexts, that mechanisms that increase the mean generation interval increase the epidemic strength $\RR$ that corresponds to an observed epidemic speed $r$ \citep{eaton2014proportion,powers2014impact,weitz2015modeling,park2020time}.
+In particular, several studies have noted, in various contexts, that mechanisms that increase the mean generation interval increase the epidemic strength $\RR$ that would be estimated for a given epidemic speed $r$ \citep{eaton2014proportion,powers2014impact,weitz2015modeling,park2020time}.
 
 Analyses of new variants have characterized \emph{relative} strength (i.e., the ratio of reproduction numbers $\Rv/\Rw$, here called $\rho$) and speed (i.e., the difference between growth rates $\rv-\rw$, here called $\delta$) of the variants, where $\Rv$, $\Rw$ and $\rv$, $\rw$ represent reproduction numbers and growth rates at a given time.
-Many analyses have particularly focused on changes in the \emph{proportion} of a new variant to estimate the relative speed \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}, instead of taking the difference between their observed speed.
+Many analyses have particularly focused on changes in the \emph{proportion} of a new variant to estimate the relative speed \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}\jd{Can we drop this?: \told{instead of taking the difference between their observed speed}--if we're looking at the same time period, these are probably equivalent, and otherwise it's not clear what we mean}.
 Focusing on proportions can be advantageous, because changes in proportions may be less sensitive to changes in testing and to other transient effects that would affect variants and wild type viruses
@@
 To illustrate the differences in implications for holding $\RR$ or $r$ fixed, we consider two idealized interventions of constant strength and constant speed in this paper.
+\jd{I feel this explanation becomes more straightforward if we introduce $\hat K$ for the new kernel.}
 Given a pre-intervention kernel $K(\tau) = \RR g(\tau)$, which represents the rate at which an infected individual generates secondary infections $\tau$ time units after infection, a constant-strength intervention reduces transmission by a constant factor $\theta$ across generation intervals ($K(\tau)/\theta$)---in this case, the intervention strength $\theta$ must be greater than the epidemic strength $\RR$ to prevent onward transmission.
@@
 
-For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals results in a slower estimated epidemic growth rate (lower $r$), making an epidemic look easier to control with a constant-speed intervention.
-In this same case, however, variation in the generation-interval distribution has no effects on the effectiveness of a constant-strength intervention because $\RR$ is fixed.
+For example, if we assume epidemic strength $\RR$ is known, then variation in the generation-interval distribution has no effects on the estimated effectiveness of a constant-strength intervention.
+In this same case, however, assuming longer generation intervals results in a slower estimated epidemic growth rate (lower $r$), making an epidemic look easier to control with a constant-speed intervention.
 Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals leads to a higher estimate of epidemic strength ($\RR$), making the epidemic look harder to control with a constant-strength intervention.
-Constant-strength and -speed interventions are idealized representations of real-life interventions, which can be either strength-like (e.g., vaccination and social distancing) or speed-like (e.g., contact tracing and isolation) depending on how their effectiveness varies across the generation interval (see \cite{doi:10.1098/rspb.2020.1556} and Discussion).
+Constant-strength and -speed interventions are idealized representations of real-life interventions, which can vary from strength-like (e.g., vaccination and social distancing) or speed-like (e.g., contact tracing and isolation) depending on how their effectiveness varies across the generation interval (see \cite{doi:10.1098/rspb.2020.1556} and Discussion).
 
-Recent studies have further suggested the possibility that new variants may have different generation-interval distributions.
-For example, Kissler \textit{et al.} \citep{kissler2021densely} suggested that the Alpha variant may have a longer duration of infection: 13.3 days (90\% CI: 10.1--16.5) for the Alpha variant and 8.2 days (90\% CI: 6.5--9.7) for the wild type.
-Longer duration of infection suggests that the mean generation interval of the Alpha variant is likely to be longer than that of the wild type \citep{lloyd2001realistic,wearing2005appropriate,roberts2007model}.
-In contrast, some studies have suggested that the more recent Delta variant may have a shorter generation interval due to faster within-host viral replication \citep{li2021viral,zhang2021transmission}.
-Modeling studies have also considered the possibility that the faster growth rate of new variants may be driven, in part, by shorter generation intervals \citep{davies2021estimated,volz2021transmission}.
+Recent studies have suggested the possibility that new variants may have different generation-interval distributions.
+For example, Kissler \textit{et al.} \citep{kissler2021densely} suggested that the Alpha variant may have a longer duration of infection: 13.3 days (90\% CI: 10.1--16.5), compared to 8.2 days (90\% CI: 6.5--9.7) for the wild type, thus suggesting that the mean generation interval of the Alpha variant is likely to be longer than that of the wild type \citep{lloyd2001realistic,wearing2005appropriate,roberts2007model}.
+In contrast, some studies have suggested that the Delta variant may have a shorter generation interval due to faster within-host viral replication \citep{li2021viral,zhang2021transmission}.
+Modeling studies have also considered the possibility that the fast replacement seen in some variants may be driven, in part, by shorter generation intervals \citep{davies2021estimated,volz2021transmission}.
 However, linking strength and speed is complicated given that the generation-interval distribution depends on many factors including behavior:
 for example, self-isolation after symptom onset will lead to shorter generation intervals.
-The recent emergence of a new Omicron variant, and its breakthrough infections in previously immune individuals, adds further uncertainty to individual-level transmission dynamics and therefore the generation-interval of SARS-CoV-2 variants \citep{pulliam2021increased}.
+The recent emergence of the variant, and its breakthrough infections in previously immune individuals, adds further uncertainty to individual-level transmission dynamics and therefore the generation-interval of SARS-CoV-2 variants \citep{pulliam2021increased}.
 
@@
 We use the renewal equation framework to characterize the spread of two pathogen strains---in this case, the wild type SARS-CoV-2 virus and a focal variant of concern.
+\jd{We should probably change wild type to resident (res) by now.}
+\jd{And then do we need to also say invader (inv)?. Bleah.}
 We focus on characterizing the incidence of infection, which is directly related to $r$ and $\RR$.
@@
 
-The integral of the kernel is referred to as the instantaneous reproduction number:
+Integrating the kernel at a fixed calendar time gives the instantaneous reproduction number:
 \begin{equation}
@@
 The normalized kernel $g_x(t, \tau) = K_x(t, \tau)/\RR_x(t)$---which we refer to as the instantaneous generation-interval distribution---describes the relative contribution of previously infected individuals to current incidence $i_x(t)$ and provides information about the time scale of pathogen transmission.
-Like the instantaneous reproduction number, the instantaneous generation-interval distribution describes when an individual infected at time $t$ will transmit infections if conditions were to remain the same.
-Both the instantaneous reproduction number and the generation-interval distribution depend on many factors, including intrinsic infectiousness of an infected individual, non-pharmaceutical interventions, awareness-driven behavior, and population-level susceptibility \citep{fraser2007estimating}.
+Like the instantaneous reproduction number, the instantaneous generation-interval distribution describes contributions to the epidemic under the counter-factual of conditions remaining constant at the values characteristic of a particular time.
+Both the instantaneous reproduction number and the instantaneous generation-interval distribution depend on many factors, including intrinsic infectiousness of an infected individual, non-pharmaceutical interventions, awareness-driven behavior, and population-level susceptibility \citep{fraser2007estimating}.
 
-Under constant-strength changes that reduce the transmission rate by a constant amount across generation intervals, the instantaneous generation-interval distribution does not change through time \citep{fraser2007estimating}.
+Constant-strength changes, which reduce transmission rate of infectious individuals independent of age of infection, do not change the instantaneous generation-interval distribution \citep{fraser2007estimating}.
 In this case, the instantaneous generation-interval distribution is also often referred to as the intrinsic generation-interval distribution \citep{champredon2015intrinsic,champredon2018two,gostic2020practical,park2020time}---
 for example, the standard SEIR model can be equivalently expressed as a renewal equation with time-invariant intrinsic generation-interval distribution $g(\tau)$ as shown in \citep{champredon2018equivalence}.
-The instantaneous generation-interval distribution is also different from the realized generation-interval distribution, which measures time between actual infection events \citep{champredon2015intrinsic}.
+The instantaneous generation-interval distribution is different from the realized generation-interval distribution, which measures time between actual infection events \citep{champredon2015intrinsic}.
 Previous studies have noted, in many contexts, that the realized generation intervals can contract due to susceptible depletion---a special case of constant-strength changes \citep{kenah2008generation,nishiura2010time,champredon2015intrinsic}---even though the instantaneous generation-interval distribution remains unchanged in this scenario.
@@
 Here, we use the term ``epidemiological conditions'' to broadly refer to all factors that affect transmission at the population-level---mathematically, they are captured by the kernel $K(t, \tau)$.
-In the context of SARS-CoV-2 infections, we are essentially assuming that the changes in susceptible pool, behavior, and contact rates are usually small over a short period of time---for example, this assumption would not apply at the times of drastic policy changes but can still work before and after policy changes have been made.
+In the context of SARS-CoV-2 infections, we are essentially assuming that the changes in susceptible pool, behavior, and contact rates are usually small over a short period of time---this assumption would not apply at the moment of a drastic policy change, but could be applied to the periods before and after.
 Then, the incidence of each strain grows (or decays) exponentially at rate $r_x$, satisfying the Euler-Lotka equation \citep{wallinga2007generation}:
@@
 in particular, the mean generation interval can be shorter than the mean infectious period because transmission occurs throughout the infectious period until recovery (see \cite{roberts2007model} for detailed discussion).
+\jd{This is unbalanced; it can also be longer, because longer infections are generally assumed to seed proportionally more infections -- maybe leave it out here.}
 
 We use this framework to investigate how inferences about strength and speed of the variant depend on our assumptions about the underlying generation-interval distributions.
-Here, we focus on differences in mean generation intervals, assuming that both the squared coefficient of variation and distribution shape of the variant are otherwise the same as for the wild type: $\kappa_{\mathrm{wt}} = \kappa_{\mathrm{var}} = \kappa$.
+Here, we focus on differences in mean generation intervals, assuming that both the variant and wild type are Gamma-distributed with squared coefficient of variation $\kappa_{\mathrm{wt}} = \kappa_{\mathrm{var}} = \kappa$.
 Changes in shape can be important \citep{miller2010epidemics,svensson2015influence}, but we do not investigate them here. 
-We do note that we expect a distribution with higher coefficient of variation to allow for more early transmission, and thus to have qualitatively similar effects to a distribution with a shorter mean \citep{park2019practical}.
+We do note that we expect a distribution with higher coefficient of variation to allow for more early transmission, and thus to have qualitatively similar effects to a distribution with a shorter mean \citep{park2019practical} \jd{add: during periods of growth, when early events are more important than later events--or drop the sentence.}
 
@@
 When incidence is changing exponentially ($i_x(t) = i_x(t_0) \exp(r_x t)$), the proportion of the new variant $p(t)$ follows a logistic growth curve \citep{switzerland2021variant,davies2021estimated}:
+\jd{Do we need the second equation?}
 \begin{align}

### 2022-01-14 newvariant.tex
Incremental progress

@@
 We evaluate the estimates of relative strength $\rho$ across a wide range of $\kappa$ from 0 (fixed-length generation intervals) to 1 (exponential distributions).
-To further explore how inference depends on underlying epidemiological conditions, we consider five scenarios: (1) $\rw < \rv < 0$, (2) $\rw < \rv = 0$, (3) $\rw < 0 < \rv$, (4) $0 = \rw < \rv$, and (5) $0 < \rw < \rv$.
-In all of these cases, we assume $\delta = \rv - \rw = 0.1\pday$ based on the observed relative growth rate of the Alpha variant in the UK \citep{davies2021estimated}.
+To further explore how inference depends on underlying epidemiological conditions, we consider five scenarios, all with $\delta = \rv - \rw = 0.1\pday$ (based on observations of the Alpha variant in the UK \citep{davies2021estimated}), and with increasing underlying $r$: (1) $\rw < \rv < 0$, (2) $\rw < \rv = 0$, (3) $\rw < 0 < \rv$, (4) $0 = \rw < \rv$, and (5) $0 < \rw < \rv$.
 

### 2022-01-14 newvariant.tex
A few thoughts.

@@
 \jd{And then do we need to also say invader (inv)?. Bleah.}
+\jd{Maybe just say resident and invader once or twice in the beginning, but tell the metaphorical story with $v$ and $w$ -- it's good notation, and easier.}
 We focus on characterizing the incidence of infection, which is directly related to $r$ and $\RR$.
@@
 
-We do not generally expect the relative speed $\delta$ to remain fixed if other factors governing epidemic spread are changing on the time scale of the invasion of a new variant relative to that of a wild type.
+We do not generally expect the relative speed $\delta$ to remain fixed if other factors governing epidemic spread are changing \jd{Drop?
+\old on the time scale of the invasion of a new variant relative to that of a wild type}.
 Instead, many biological mechanisms appear compatible with assuming a fixed value of relative strength $\rho$ over changing conditions.
-For example, if the proportion of the population susceptible declines, or the average contact rate changes, while other factors remain unchanging, the relative strength $\rho$ is expected to remain fixed \citep{leung2017monitoring,leung2020empirical,di2021impact,leung2021early};
-as we show in this section, these changes will imply a change in relative speed $\delta$.
+For example, if the proportion of the population susceptible declines, or the average contact rate changes, while other factors remain unchanging, the relative strength $\rho$ is expected to remain fixed \citep{leung2017monitoring,leung2020empirical,di2021impact,leung2021early}; \jd{Drop?
+\old as we show in this section, these changes will imply a change in relative speed $\delta$}.
 
@@
 \end{equation}
-As our baseline scenario we assume $\rho = 1.61$ (and therefore, we always have $\Rw < \Rv$), which is the value we obtain for $\delta=0.1\pday$ \citep{davies2021estimated}, $\rw=0\pday$, $\Gw = \Gv = 5\,\textrm{days}$, and $\kappa = 1/5$ \citep{ferretti2020quantifying}.
+As our baseline scenario we assume $\rho = 1.61$ (i.e., $\Rv = 1.61 \Rw$), which is the value we obtain for $\delta=0.1\pday$ \citep{davies2021estimated}, $\rw=0\pday$, $\Gw = \Gv = 5\,\textrm{days}$, and $\kappa = 1/5$ \citep{ferretti2020quantifying}.
 We evaluate $\delta$ across five scenarios as before: (1) $\Rw < \Rv < 1$, (2) $\Rw < \Rv = 1$, (3) $\Rw < 1 < \Rv$, (4) $1 = \Rw < \Rv$, and (5) $1 < \Rw < \Rv$.

### 2022-01-16 newvariant.tex
Thinking about kappa

@@
 Once again, we see that assuming a narrower distribution (lower $\kappa$) has qualitatively similar effects as assuming a longer mean.
+\jd{This is very deep, and I don't think there's any ``once again'' here -- meaning, we haven't explained this one yet.
+This is the effect of lower $\kappa$ for both being compared to the effect of longer GI for just one.
+In the non-comparative version, the effect of high $\kappa$ is on the influential end: it's like longer (shorter) GIs when disease is decreasing (increasing).}
 
 \fref{relspeed} also shows that when $\rho$ is fixed, relative speed depends on underlying epidemiological conditions---specifically, the absolute strength of the two strains. 
-For example, even when the generation-interval distributions are identical ($\Gv=\Gw$, in this case), changing $\Rw$ from 0.49 to 1.27 (and thus $\Rv$ from 0.79 to 2.04) yields a relative speed $\delta$ between 0.08--0.11 for $\kappa=0.2$ and 0.06--0.14 for $\kappa=0.5$.
+For example, even when the generation-interval distributions are identical ($\Gv=\Gw$, in this case), changing $\Rw$ from 0.49 to 1.27 (and thus $\Rv$ from 0.79 to 2.04) changes the relative speed $\delta$ from 0.08 to 0.11 when $\kappa=0.2$---and from 0.06 to 0.14 when $\kappa=0.5$.
 Differences in the generation-interval distributions exaggerate these changes.

### 2022-01-16 newvariant.tex
S5, quickly.

@@
 
-Instead of inferring relative strength from speed, one can directly estimate time-varying reproduction numbers $\RR(t)$ of the variant and the wild type from incidence data and directly compare their ratios when the incidence of infection and the generation-interval distributions are known---such methods have been used in previous analyses of the Alpha \citep{volz2021transmission} and Delta variants \citep{abbott2021estimating}.
+\jd{Instead here doesn't quite work -- it's still about strength from speed because you're still observing $r(t)$ more directly and making assumptions about GI. \old
+Instead of inferring relative strength from speed, one can directly estimate time-varying reproduction numbers $\RR(t)$ of the variant and the wild type from incidence data and directly compare their ratios when the incidence of infection and the generation-interval distributions are known}
+Instead of inferring relative strength from relative speed, one can separately estimate time-varying reproduction numbers $\RR(t)$ of the variant and the wild type from observed incidence and assumptions about the generation intervals, and then calculate the ratio---such methods have been used in previous analyses of the Alpha \citep{volz2021transmission} and Delta variants \citep{abbott2021estimating}.
 Broadly, there are two types of time-varying reproduction numbers: case reproduction number and instantaneous reproduction number.
@@
 Biases in slopes further translate to biases in intercepts; in theory, the regression line should go through the origin (because $\Rv = 0$ when $\Rw = 0$).
+\jd{Did you fit lines constrained through the origin or not?
+Need to be clear; what would a bias in intercept even mean?}
 
@@
 While relative speed $\delta$ and strength $\rho$ are useful for characterizing the spread of the variant in an epidemiological context with a previously dominant wild type, the \emph{absolute} speed $\rv$ and strength $\Rv$ of the variant determine the spread and conditions for control of the variant over the long term.
-In particular, at any given point in the epidemic, we can measure the speed of the variant $\rv$ (or infer $\rv$ from $\rw$ and $\delta$) and ask how much more intervention is required to control the spread of both strains (since $\Rv < 1$ implies $\Rw < 1$).
+In particular, at any given point in the epidemic, we can measure the speed of the variant $\rv$ (or infer $\rv$ from $\rw$ and $\delta$) and ask how much more intervention is required to control the spread of the variant (and thus also the wild-type, which is assumed to be easier to control).
 As a baseline scenario, we assume $\rw=0$ and $\delta=0.1\pday$ (and therefore $\rv=0.1\pday$), in which case additional intervention is required to reduce $\rv$ below 0 (or, equivalently, $\Rv$ below 1).

### 2022-01-16 newvariant.tex
pass back to Daniel

@@
 
-The speed-based paradigm gives the same results regarding control but provides important insights (\fref{strengthspeed}F).
+The speed-based paradigm gives the same results regarding control but provides additional insight (\fref{strengthspeed}F).
 The observed speed of the variant $\rv$ at a given moment is independent of our estimates of its mean generation interval.
@@
 Finally, we synthesize our findings and illustrate the differences between constant-strength and constant-speed interventions using epidemic simulations (\fref{control_sim}).
-As before, we consider a scenario in which a new variant is emerging with known relative speed ($\delta=\rv > 0.1\pday$ and $\Rw > 1$) while the incidence of infections caused by the wild type is constant ($\rw = 0\pday$ and $\Rw =1$) before interventions are reduced.
-The constant-strength intervention then reduces both $\Rw$ and $\Rv$ by a constant factor $\theta$ beginning on day 30, 
+As before, we consider a scenario in which a new variant is emerging with known relative speed ($\delta=\rv > 0.1\pday$ $\implies$ $\Rw > 1$) while the incidence of infections caused by the wild type is constant ($\rw = 0\pday$ $\implies$ $\Rw =1$) before interventions are introduced.
+The constant-strength intervention then reduces both $\Rw$ and $\Rv$ by a factor $\theta$ beginning on day 30, 
 whereas the constant-speed intervention isolates individuals infected with both the wild type and the variant at a constant hazard $\phi$ after day 30.
-If the variant has a longer mean generation interval, this specific constant-strength intervention fails to suppress the spread of the variant (\fref{control_sim}A) because the longer generations imply a higher initial $\Rw$, and thus a stronger intervention is required to reduce $\Rw$ below 1 (\fref{control_sim}B).
+If the variant has a longer mean generation interval, this particular constant-strength intervention fails to suppress the spread of the variant (\fref{control_sim}A) because the longer generations imply a higher initial $\Rw$, and thus a stronger intervention is required to reduce $\Rw$ below 1 (\fref{control_sim}B).
 Conversely, if the variant has a shorter mean generation interval, its initial $\Rw$ will be lower, and therefore the same constant-strength intervention will be more effective, leading to lower post-intervention $\Rw$ (\fref{control_sim}B).
@@
 We explored how the generation-interval distribution shapes the link between relative strength and speed.
-While the role of generation-interval distributions in linking epidemic strength and speed has been well-established, using the generation-interval-based framework provides clear explanations for an important, but under-appreciated concept: the relative strength and speed of new variants also depend on epidemic dynamics of the prior dominant lineages as well as the type of control measures (i.e., strength- or speed-like) that limit their spread.
+While the role of generation-interval distributions in linking epidemic strength and speed has been well-established, this framework can also provide provides insight into an ut under-appreciated concept: the relative strength and speed of new variants also depend on both current epidemic conditions and the nature of control measures (e.g., strength- or speed-like) to be introduced.
 For example, under a constant-strength intervention that reduces the transmission rate of the wild type and a variant by the same amount (and therefore, keeping the relative strength constant), the relative speed of the variant is sensitive to the epidemic growth rate of the wild type strain.
-Our results challenge the commonly-made assumption that relative speed of new variants remains constant over time when characterizing changes in the proportion of new variants.
+Our results challenge the commonly made assumption that relative speed of new variants remains constant over time when characterizing changes in the proportion of new variants.
+
 Differences in the generation-interval distributions further exaggerate this effect.
-
 Therefore, neglecting potential differences in the mean generation intervals between the variant and the wild type can bias estimates of the its transmission advantage.
@@
 
-There were a (minor) discrepancy in the estimates of relative strength of the Alpha variant.
-Mathematical analyses typically reported greater than a 1.4 fold increase in reproduction number for the Alpha variant ($\rho > 1.4$) whereas an independent analysis of secondary attack rates from contact tracing data suggested a 1.25--1.4 fold increase \citep{ukinvest}.
+This perspective can shed light on earlier estimates of relative strength of the Alpha variant.
+Mathematical analyses typically reported greater than a 1.4 fold increase in reproduction number for the Alpha variant ($\rho > 1.4$) whereas an independent analysis of secondary attack rates from contact tracing data suggested a somewhat smaller 1.25--1.4 fold increase \citep{ukinvest}.
 As shown in \cite{davies2021estimated} and \cite{volz2021transmission}, propagating uncertainty in generation interval estimates (and, in particular, assuming shorter generation intervals) can partially explain these differences:
@@
 
-Other studies have also tried to explain apparent changes in their estimates of relative strength using generation-interval-based arguments. 
+Other studies have also used generation-interval-based arguments to explore changes in estimates of relative strength. 
 For example, Volz \textit{et al.} estimated that the relative strength of the Alpha variant declined in England between December 2020 and January 2021 \cite{volz2021transmission};
@@
 
-We have primarily focused on the differences in the mean generation interval but the differences in the amount of variability, characterized by the squared coefficient of variation $\kappa$, can also have important, but different implications \citep{blanquart2021selection}.
+We have primarily focused on the differences in the mean generation interval, but differences in the amount of variability, characterized by the squared coefficient of variation $\kappa$, could also have important, but different implications \citep{blanquart2021selection}.
 For example, if two variants have identical reproduction numbers, the variant with a wider generation-interval distribution will always have a faster growth rate and out-compete the other variant (except when both variants have $\RR=1$).
-This is because a variant with a wider generation-interval distribution can take advantage of more early transmission during the growth phase and more late transmission during the decay phase.s
+This is because a variant with a wider generation-interval distribution can take advantage of more early transmission during the growth phase and more late transmission during the decay phase.
 
@@
 Interventions that reduce contact rates, such as social distancing measures, are generally strength-like, but they can also elicit awareness-based behavioral responses that are speed-like---for example, infected individuals may self-isolate faster after symptom onset.
-Other strength-like interventions, such as vaccination, can also have disproportionately smaller effects on new variants depending on their antigenic variability.
 On the other hand, the hazard of isolation-based interventions (and thus, speed-like), such as contact tracing, can vary over the course of infection depending on delays in tracing infected individuals \citep{fraser2004factors,ferretti2020quantifying,scarabel2021renewal}.
 Their effectiveness can further depend on the ``coverage'' of isolation, which is strength-like---for example, there may be a group of individuals that cannot be isolated by certain interventions due to asymptomaticity or the lack of participation in the contact tracing program.
+Further, some interventions, such as vaccination, can also have disproportionate different effects on different variants---another complication not considered in our analysis.
+
 Analyses of SARS-CoV-2 dynamics have primarily relied on the constant-strength framework \citep{gostic2020practical,unwin2020state,brauner2021inferring}, even when modeling speed-like interventions such as self-isolation \citep{flaxman2020Rt,brett2020transmission}.

### 2022-01-16 newvariant.tex
minor changes

@@
 Neglecting differences in generation-interval distributions can bias estimates of relative strength.
-\jd{Can we delete this?
-Is there a a good way to clarify?: \old\ which can vary systematically with changes in the incidence of infections caused by the variant.}
 
@@
 
-Analyses of new variants have characterized \emph{relative} strength (i.e., the ratio of reproduction numbers $\Rv/\Rw$, here called $\rho$) and speed (i.e., the difference between growth rates $\rv-\rw$, here called $\delta$) of the variants, where $\Rv$, $\Rw$ and $\rv$, $\rw$ represent reproduction numbers and growth rates at a given time.
-Many analyses have particularly focused on changes in the \emph{proportion} of a new variant to estimate the relative speed \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}\jd{Can we drop this?: \told{instead of taking the difference between their observed speed}--if we're looking at the same time period, these are probably equivalent, and otherwise it's not clear what we mean}.
+Analyses of new variants have characterized \emph{relative} strength (i.e., the ratio of reproduction numbers of the invading and resident strains, here called $\rho$) and speed (i.e., the difference between growth rates of the invading and resident strains, here called $\delta$).
+As an example, we consider a new variant invading a wild type strain in this paper and use $\Rv$ and $\Rw$ to denote their reproduction numbers and $\rv$ and $\rw$ to denote their growth rates at a given time (and therefore $\rho = \Rv/\Rw$ and $\delta = \rv-\rw$).
+Many analyses have particularly focused on changes in the \emph{proportion} of a new variant to estimate the relative speed \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
 Focusing on proportions can be advantageous, because changes in proportions may be less sensitive to changes in testing and to other transient effects that would affect variants and wild type viruses
@@
 To illustrate the differences in implications for holding $\RR$ or $r$ fixed, we consider two idealized interventions of constant strength and constant speed in this paper.
-\jd{I feel this explanation becomes more straightforward if we introduce $\hat K$ for the new kernel.}
-Given a pre-intervention kernel $K(\tau) = \RR g(\tau)$, which represents the rate at which an infected individual generates secondary infections $\tau$ time units after infection, a constant-strength intervention reduces transmission by a constant factor $\theta$ across generation intervals ($K(\tau)/\theta$)---in this case, the intervention strength $\theta$ must be greater than the epidemic strength $\RR$ to prevent onward transmission.
-On the other hand, a constant-speed intervention reduces transmission after infection by a constant rate $\phi$ across generation intervals: $K(\tau) \exp(-\phi \tau)$---in this case, the intervention speed $\phi$ must be greater than the epidemic speed $r$ to prevent onward transmission \citep{doi:10.1098/rspb.2020.1556}.
+Given a pre-intervention kernel $K_{\mathrm{pre}}(\tau) = \RR g(\tau)$, which represents the rate at which an infected individual generates secondary infections $\tau$ time units after infection, a constant-strength intervention reduces transmission by a constant factor $\theta$ across generation intervals such that the post-intervention kernel corresponds to: $K_{\mathrm{post}} = K_{\mathrm{pre}}(\tau)/\theta$.
+In this case, the intervention strength $\theta$ must be greater than the epidemic strength $\RR$ to prevent onward transmission.
+On the other hand, a constant-speed intervention reduces transmission after infection by a constant rate $\phi$ across generation intervals: $K_{\mathrm{post}} = K_{\mathrm{pre}}(\tau) \exp(-\phi \tau)$---in this case, the intervention speed $\phi$ must be greater than the epidemic speed $r$ to prevent onward transmission \citep{doi:10.1098/rspb.2020.1556}.
 
@@
 We use the renewal equation framework to characterize the spread of two pathogen strains---in this case, the wild type SARS-CoV-2 virus and a focal variant of concern.
-\jd{We should probably change wild type to resident (res) by now.}
-\jd{And then do we need to also say invader (inv)?. Bleah.}
-\jd{Maybe just say resident and invader once or twice in the beginning, but tell the metaphorical story with $v$ and $w$ -- it's good notation, and easier.}
 We focus on characterizing the incidence of infection, which is directly related to $r$ and $\RR$.
@@
 The Gamma-generation-interval assumption includes as a special case models that assume exponentially distributed generation intervals (when $\kappa=1$), corresponding to the SIR model \citep{anderson1991infectious}.
-We note that when the infectious periods are Gamma distributed---another standard assumption in epidemic modeling---the resulting generation interval does not necessarily follow the Gamma distribution;
-in particular, the mean generation interval can be shorter than the mean infectious period because transmission occurs throughout the infectious period until recovery (see \cite{roberts2007model} for detailed discussion).
-\jd{This is unbalanced; it can also be longer, because longer infections are generally assumed to seed proportionally more infections -- maybe leave it out here.}
+We note that when the infectious periods are Gamma distributed---another standard assumption in epidemic modeling---the resulting generation interval does not necessarily follow the Gamma distribution (see \cite{roberts2007model} for detailed discussion).
 
@@
 Changes in shape can be important \citep{miller2010epidemics,svensson2015influence}, but we do not investigate them here. 
-We do note that we expect a distribution with higher coefficient of variation to allow for more early transmission, and thus to have qualitatively similar effects to a distribution with a shorter mean \citep{park2019practical} \jd{add: during periods of growth, when early events are more important than later events--or drop the sentence.}
+We do note that we expect a distribution with higher coefficient of variation to allow for more early transmission, and thus to have qualitatively similar effects to a distribution with a shorter mean during periods of growth, when early events are more important than later events \citep{park2019practical}.
 
@@
 When incidence is changing exponentially ($i_x(t) = i_x(t_0) \exp(r_x t)$), the proportion of the new variant $p(t)$ follows a logistic growth curve \citep{switzerland2021variant,davies2021estimated}:
-\jd{Do we need the second equation?}
 \begin{align}
@@
 When $\rv < 0$, we see that longer generation intervals decrease $\rho$ because longer generation intervals actually lead to slower decay (higher $r$).
-Assuming a narrower distribution (lower $\kappa$) has qualitatively similar effects as assuming longer generation intervals (leading to higher values of $\rho$ when $\rv > 0$) because both reduce the amount of early transmission.
+Assuming a narrower distribution for both the variant and the wild type strain (lower $\kappa$) has qualitatively similar effects as assuming longer generation intervals (leading to higher values of $\rho$ when $\rv > 0$ and lower values of $\rho$ when $\rv < 0$) because both reduce the amount of early transmission.
 When $\rw < 0 < \rv$, inference of $\rho$ is relatively insensitive to values of $\kappa$.
@@
 
-We do not generally expect the relative speed $\delta$ to remain fixed if other factors governing epidemic spread are changing \jd{Drop?
-\old on the time scale of the invasion of a new variant relative to that of a wild type}.
+We do not generally expect the relative speed $\delta$ to remain fixed if other factors governing epidemic spread are changing.
 Instead, many biological mechanisms appear compatible with assuming a fixed value of relative strength $\rho$ over changing conditions.
-For example, if the proportion of the population susceptible declines, or the average contact rate changes, while other factors remain unchanging, the relative strength $\rho$ is expected to remain fixed \citep{leung2017monitoring,leung2020empirical,di2021impact,leung2021early}; \jd{Drop?
-\old as we show in this section, these changes will imply a change in relative speed $\delta$}.
+For example, if the proportion of the population susceptible declines, or the average contact rate changes, while other factors remain unchanging, the relative strength $\rho$ is expected to remain fixed \citep{leung2017monitoring,leung2020empirical,di2021impact,leung2021early}.
 
@@
 When $\Rv<1$, longer generation intervals of the variant lead to slower decay ($\rv$ closer to 0), and therefore, greater relative speed.
-Once again, we see that assuming a narrower distribution (lower $\kappa$) has qualitatively similar effects as assuming a longer mean.
-\jd{This is very deep, and I don't think there's any ``once again'' here -- meaning, we haven't explained this one yet.
-This is the effect of lower $\kappa$ for both being compared to the effect of longer GI for just one.
-In the non-comparative version, the effect of high $\kappa$ is on the influential end: it's like longer (shorter) GIs when disease is decreasing (increasing).}
+We see that assuming a narrower distribution for both the variant and the wild type strain (lower $\kappa$) has qualitatively similar effects as assuming a longer mean (leading to lower values of $\delta$ when $\Rv > 0$ and higher values of $\delta$ when $\Rv < 0$).
 
@@
 
-\jd{Instead here doesn't quite work -- it's still about strength from speed because you're still observing $r(t)$ more directly and making assumptions about GI. \old
-Instead of inferring relative strength from speed, one can directly estimate time-varying reproduction numbers $\RR(t)$ of the variant and the wild type from incidence data and directly compare their ratios when the incidence of infection and the generation-interval distributions are known}
 Instead of inferring relative strength from relative speed, one can separately estimate time-varying reproduction numbers $\RR(t)$ of the variant and the wild type from observed incidence and assumptions about the generation intervals, and then calculate the ratio---such methods have been used in previous analyses of the Alpha \citep{volz2021transmission} and Delta variants \citep{abbott2021estimating}.
@@
 For example, we can estimate $\rho$ by plotting the estimated strength of the variant $\hat{\RR}_{\textrm{var}}(t)$ against the estimated strength of the wild type $\hat{\RR}_{\textrm{wt}}(t)$---as presented in Figure 2 of \cite{volz2021transmission}---and performing a linear regression (\fref{Rtbias}C,F,I).
+Here, we estimate the slope while fixing the intercept: in theory, the regression line should go through the origin in theory because $\Rv = 0$ when $\Rw = 0$.
 If $\rho$ is constant, and generation-interval distributions are correctly specified, we obtain a straight line with a slope of $\rho$ and intercept at zero (\fref{Rtbias}F).
 However, when the assumed mean generation interval is longer than the that of the variant, we over-estimate the slope (and conversely, \fref{Rtbias}C,I).
-Biases in slopes further translate to biases in intercepts; in theory, the regression line should go through the origin (because $\Rv = 0$ when $\Rw = 0$).
-\jd{Did you fit lines constrained through the origin or not?
-Need to be clear; what would a bias in intercept even mean?}
 
@@
 Recently, it has been further suggested that the generation intervals of the original SARS-CoV-2 strain may be considerably longer than previously thought \citep{sender2021unmitigated}, adding further uncertainty.
-Although a few studies have tried to quantify both the generation- and serial-interval distributions of new SARS-CoV-2 variants, their conclusions on whether transmission time scale of new variants differs are conflicting \citep{pung2021serial, hart2021generation}---we note that these comparisons are inherently difficult especially due to temporal changes in generation- and serial-interval distributions due to dynamical and intervention effects \citep{ali2020serial,park2021forward}.
+A few studies have tried to compare generation- and serial-interval distributions of new SARS-CoV-2 variants with those of the original wild type strain \citep{pung2021serial, hart2021generation}, these comparisons are inherently difficult especially due to temporal changes in generation- and serial-interval distributions caused by dynamical and intervention effects \citep{ali2020serial,park2021forward}.
 Future studies should prioritize detailed assessment of the generation intervals of SARS-CoV-2 and widespread variants, as well as consider how uncertainty in generational intervals might bias conclusions \citep{doi:10.1098/rsif.2020.0144,ali2020serial,gostic2020practical}.

### 2022-01-16 Rtbias.R

@@ -156,7 +156,7 @@ for (i in 1:length(svec)) {
   
   if (i != 2) {
     g3 <- g3 +
-      geom_smooth(aes(Rtest1, Rtest2, col="Regression", lty="Regression"), method="lm", fullrange=TRUE, se=FALSE, lwd=2)
+      geom_smooth(aes(Rtest1, Rtest2, col="Regression", lty="Regression"), method="lm", formula=y~-1+x, fullrange=TRUE, se=FALSE, lwd=2)
     
   }
   

### 2022-01-16 Rtbias_smooth.R

@@ -6,7 +6,7 @@ library(gridExtra)
 library(tikzDevice)
 source("renewal_det.R")
 
-## fix theta = 1.5
+## fix theta = 1.61
 ## fix kappa = 1/5
 ## fix assumed GI
 ## assume wild type = var but not necessarily the one we're assuming
@@ -151,7 +151,7 @@ for (i in 1:length(svec)) {
   
   if (i != 2) {
     g3 <- g3 +
-      geom_smooth(aes(Rtest1, Rtest2, col="Regression", lty="Regression"), method="lm", fullrange=TRUE, se=FALSE, lwd=2)
+      geom_smooth(aes(Rtest1, Rtest2, col="Regression", lty="Regression"), method="lm", formula=y~-1+x, fullrange=TRUE, se=FALSE, lwd=2)
       
   }
   

### 2022-01-16 newvariant.tex
more checks

@@
 for example, self-isolation after symptom onset will lead to shorter generation intervals.
-The recent emergence of the variant, and its breakthrough infections in previously immune individuals, adds further uncertainty to individual-level transmission dynamics and therefore the generation-interval of SARS-CoV-2 variants \citep{pulliam2021increased}.
+The recent emergence of the Omicron variant, and its breakthrough infections in previously immune individuals, adds further uncertainty to individual-level transmission dynamics and therefore the generation-interval of SARS-CoV-2 variants \citep{pulliam2021increased}.
 
@@
 We explored how the generation-interval distribution shapes the link between relative strength and speed.
-While the role of generation-interval distributions in linking epidemic strength and speed has been well-established, this framework can also provide provides insight into an ut under-appreciated concept: the relative strength and speed of new variants also depend on both current epidemic conditions and the nature of control measures (e.g., strength- or speed-like) to be introduced.
+While the role of generation-interval distributions in linking epidemic strength and speed has been well-established, this framework can also provide provides insight into an under-appreciated concept: the relative strength and speed of new variants also depend on both current epidemic conditions and the nature of control measures (e.g., strength- or speed-like) to be introduced.
 For example, under a constant-strength intervention that reduces the transmission rate of the wild type and a variant by the same amount (and therefore, keeping the relative strength constant), the relative speed of the variant is sensitive to the epidemic growth rate of the wild type strain.

### 2022-01-17 newvariant.tex
done

@@
 Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals leads to a higher estimate of epidemic strength ($\RR$), making the epidemic look harder to control with a constant-strength intervention.
-Constant-strength and -speed interventions are idealized representations of real-life interventions, which can vary from strength-like (e.g., vaccination and social distancing) or speed-like (e.g., contact tracing and isolation) depending on how their effectiveness varies across the generation interval (see \cite{doi:10.1098/rspb.2020.1556} and Discussion).
+Constant-strength and -speed interventions are idealized representations of real-life interventions, which can vary from strength-like (e.g., vaccination and social distancing) to speed-like (e.g., contact tracing and isolation) depending on how their effectiveness varies across the generation interval (see \cite{doi:10.1098/rspb.2020.1556} and Discussion).
 

### 2022-02-07 newvariant.tex
done

--- 2022-01-16 newvariant.tex
@@
 
-Inferring the relative strength (i.e., the ratio of reproduction numbers) and relative speed (i.e., the difference between growth rates) of new SARS-CoV-2 variants compared to earlier strains is critical to predicting and controlling the course of the current pandemic.
-Analyses of new variants have primarily focused on characterizing changes in the \emph{proportion} of new variants, implicitly or explicitly assuming that the relative speed remains fixed over the course of an invasion.
-We use the generation-interval-based framework to challenge this assumption and illustrate how relative strength and speed change over time under two idealized interventions:
+Inferring the relative strength (i.e., the ratio of reproduction numbers) and relative speed (i.e., the difference between growth rates) of new SARS-CoV-2 variants is critical to predicting and controlling the course of the current pandemic.
+Analyses of new variants have primarily focused on characterizing changes in the proportion of new variants, implicitly or explicitly assuming that the relative speed remains fixed over the course of an invasion.
+We use a generation-interval-based framework to challenge this assumption and illustrate how relative strength and speed change over time under two idealized interventions:
 a constant-strength intervention like idealized vaccination or social distancing, which reduces transmission rates by a constant proportion, and a constant-speed intervention like idealized contact tracing, which isolates infected individuals at a constant rate.
-In general, a constant-strength intervention causes relative speed to change, while a constant-speed intervention causes relative strength to change.
-Differences in the generation-interval distributions between variants can exaggerate these changes and affect the effectiveness of interventions.
-Neglecting differences in generation-interval distributions can bias estimates of relative strength.
+In general, constant-strength interventions change the relative speed of a new variant, while constant-speed interventions change its relative strength.
+Differences in the generation-interval distributions between variants can exaggerate these changes and modify the effectiveness of interventions.
+Finally, neglecting differences in generation-interval distributions can bias estimates of relative strength.
 
@@
 Like epidemic strength, epidemic speed also determines conditions for pathogen elimination: $r=0$ is a threshold equivalent to $\RR=1$ under constant conditions.
-However, large segments of the epidemiology and modeling community have over-emphasized $\RR$ at the expense of $r$ \citep{doi:10.1098/rspb.2020.1556};
-we thus use the terms ``strength'' and ``speed''  here to underline our contention that these are better seen as complementary perspectives (and to link them to complementary perspectives on measuring the transmission advantage of new variants).
+However, epidemiological modelers have often over-emphasized $\RR$ at the expense of $r$ \citep{doi:10.1098/rspb.2020.1556};
+we thus use the terms ``strength'' and ``speed''  here to underline our contention that these metrics are better seen as complementary perspectives (and to link them to complementary perspectives on measuring the transmission advantage of new variants).
 
@@
 As an example, we consider a new variant invading a wild type strain in this paper and use $\Rv$ and $\Rw$ to denote their reproduction numbers and $\rv$ and $\rw$ to denote their growth rates at a given time (and therefore $\rho = \Rv/\Rw$ and $\delta = \rv-\rw$).
-Many analyses have particularly focused on changes in the \emph{proportion} of a new variant to estimate the relative speed \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
-Focusing on proportions can be advantageous, because changes in proportions may be less sensitive to changes in testing and to other transient effects that would affect variants and wild type viruses
+Many analyses have focused on changes in the \emph{proportion} of a new variant to estimate its relative speed \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
+Focusing on proportions can be advantageous, because changes in proportions are less sensitive to changes in testing and to other transient phenomena that would affect variants and wild type viruses
 similarly;
@@
 While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the pathogen and its control \citep{doi:10.1098/rspb.2020.1556}.
-To illustrate the differences in implications for holding $\RR$ or $r$ fixed, we consider two idealized interventions of constant strength and constant speed in this paper.
-Given a pre-intervention kernel $K_{\mathrm{pre}}(\tau) = \RR g(\tau)$, which represents the rate at which an infected individual generates secondary infections $\tau$ time units after infection, a constant-strength intervention reduces transmission by a constant factor $\theta$ across generation intervals such that the post-intervention kernel corresponds to: $K_{\mathrm{post}} = K_{\mathrm{pre}}(\tau)/\theta$.
-In this case, the intervention strength $\theta$ must be greater than the epidemic strength $\RR$ to prevent onward transmission.
-On the other hand, a constant-speed intervention reduces transmission after infection by a constant rate $\phi$ across generation intervals: $K_{\mathrm{post}} = K_{\mathrm{pre}}(\tau) \exp(-\phi \tau)$---in this case, the intervention speed $\phi$ must be greater than the epidemic speed $r$ to prevent onward transmission \citep{doi:10.1098/rspb.2020.1556}.
+To illustrate the differences in conclusions when holding $\RR$ or $r$ fixed, we consider two idealized interventions of constant strength and constant speed.
+Given a pre-intervention kernel $K_{\mathrm{pre}}(\tau) = \RR g(\tau)$, which represents the rate at which an infected individual generates secondary infections $\tau$ time units after infection, a constant-strength intervention reduces transmission by a constant factor $\theta$ across generation intervals such that the post-intervention kernel is $K_{\mathrm{post}} = K_{\mathrm{pre}}(\tau)/\theta$.
+In this case, the intervention strength $\theta$ must be greater than the epidemic strength $\RR$ to control the epidemic.
+In contrast, a constant-speed intervention reduces transmission after infection by a constant rate $\phi$ across generation intervals: $K_{\mathrm{post}} = K_{\mathrm{pre}}(\tau) \exp(-\phi \tau)$.
+In this case, the intervention speed $\phi$ must be greater than the epidemic speed $r$ to control the epidemic \citep{doi:10.1098/rspb.2020.1556}.
 
-For example, if we assume epidemic strength $\RR$ is known, then variation in the generation-interval distribution has no effects on the estimated effectiveness of a constant-strength intervention.
-In this same case, however, assuming longer generation intervals results in a slower estimated epidemic growth rate (lower $r$), making an epidemic look easier to control with a constant-speed intervention.
-Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals leads to a higher estimate of epidemic strength ($\RR$), making the epidemic look harder to control with a constant-strength intervention.
-Constant-strength and -speed interventions are idealized representations of real-life interventions, which can vary from strength-like (e.g., vaccination and social distancing) or speed-like (e.g., contact tracing and isolation) depending on how their effectiveness varies across the generation interval (see \cite{doi:10.1098/rspb.2020.1556} and Discussion).
+For example, if we assume epidemic strength $\RR$ is known, then variation in the generation-interval distribution does not change the estimated effectiveness of a constant-strength intervention.
+In this same case, however, assuming longer generation intervals decreases the estimated epidemic growth rate (lower $r$), making an epidemic look easier to control with a constant-speed intervention.
+Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals increases the estimated epidemic strength ($\RR$), making the epidemic look harder to control with a constant-strength intervention.
+Constant-strength and constant-speed interventions are idealized representations of real-life interventions, which can range from strength-like (e.g., vaccination and social distancing) to speed-like (e.g., contact tracing and isolation) depending on how their effectiveness varies across the generation interval (see \cite{doi:10.1098/rspb.2020.1556} and Discussion).
 
@@
 In contrast, some studies have suggested that the Delta variant may have a shorter generation interval due to faster within-host viral replication \citep{li2021viral,zhang2021transmission}.
-Modeling studies have also considered the possibility that the fast replacement seen in some variants may be driven, in part, by shorter generation intervals \citep{davies2021estimated,volz2021transmission}.
+Modeling studies have also considered the possibility that the observed fast replacement of some variants may be driven, in part, by shorter generation intervals \citep{davies2021estimated,volz2021transmission}.
 However, linking strength and speed is complicated given that the generation-interval distribution depends on many factors including behavior:
 for example, self-isolation after symptom onset will lead to shorter generation intervals.
-The recent emergence of the variant, and its breakthrough infections in previously immune individuals, adds further uncertainty to individual-level transmission dynamics and therefore the generation-interval of SARS-CoV-2 variants \citep{pulliam2021increased}.
+The emergence of the Omicron variant, and its breakthrough infections in previously immune individuals, adds further uncertainty to individual-level transmission dynamics and therefore the generation-interval of SARS-CoV-2 variants \citep{pulliam2021increased}.
 
-Here, we use the generation-interval-based framework to compare two measures of transmission advantage of new variants: the relative strength ($\rho=\Rv/\Rw$) and relative speed ($\delta=\rv-\rw$).
-We assess how relative strength and speed depend on underlying epidemiological dynamics of prior dominant lineages and argue that assuming a constant relative strength (rather than a constant relative speed) is more appropriate for estimating relative transmissibility of new variants.
+Here, we use the generation-interval-based framework to compare two measures of transmission advantage of new variants: the relative strength and relative speed.
+We assess how relative strength and speed depend on underlying epidemiological dynamics of previously dominant lineages and argue that assuming a constant relative strength (rather than a constant relative speed) is more appropriate for estimating relative transmissibility of new variants.
 We also show how neglecting differences in the generation-interval distributions of new variants can lead to biased estimates of their relative transmissibility and how such biases might be assessed in practice.
@@
 The normalized kernel $g_x(t, \tau) = K_x(t, \tau)/\RR_x(t)$---which we refer to as the instantaneous generation-interval distribution---describes the relative contribution of previously infected individuals to current incidence $i_x(t)$ and provides information about the time scale of pathogen transmission.
-Like the instantaneous reproduction number, the instantaneous generation-interval distribution describes contributions to the epidemic under the counter-factual of conditions remaining constant at the values characteristic of a particular time.
+Like the instantaneous reproduction number, the instantaneous generation-interval distribution describes contributions to the epidemic under the counterfactual case where conditions remain constant at a particular set of values.
 Both the instantaneous reproduction number and the instantaneous generation-interval distribution depend on many factors, including intrinsic infectiousness of an infected individual, non-pharmaceutical interventions, awareness-driven behavior, and population-level susceptibility \citep{fraser2007estimating}.
@@
 Here, we use the term ``epidemiological conditions'' to broadly refer to all factors that affect transmission at the population-level---mathematically, they are captured by the kernel $K(t, \tau)$.
-In the context of SARS-CoV-2 infections, we are essentially assuming that the changes in susceptible pool, behavior, and contact rates are usually small over a short period of time---this assumption would not apply at the moment of a drastic policy change, but could be applied to the periods before and after.
+In the context of SARS-CoV-2 infections, we are essentially assuming that the changes in susceptible pool, behavior, and contact rates are usually small over a short period of time---this assumption would not apply at the moment of a drastic policy change, but could be applied to the periods immediately before and after.
 Then, the incidence of each strain grows (or decays) exponentially at rate $r_x$, satisfying the Euler-Lotka equation \citep{wallinga2007generation}:
@@
 We use this framework to investigate how inferences about strength and speed of the variant depend on our assumptions about the underlying generation-interval distributions.
-Here, we focus on differences in mean generation intervals, assuming that both the variant and wild type are Gamma-distributed with squared coefficient of variation $\kappa_{\mathrm{wt}} = \kappa_{\mathrm{var}} = \kappa$.
+Here, we focus on differences in mean generation intervals, assuming that both the variant and wild type intervals are Gamma-distributed with squared coefficient of variation $\kappa_{\mathrm{wt}} = \kappa_{\mathrm{var}} = \kappa$.
 Changes in shape can be important \citep{miller2010epidemics,svensson2015influence}, but we do not investigate them here. 
@@
 
-While epidemic speed $r_x$ can often be estimated directly from incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating},
-studies of new SARS-CoV-2 variants have mostly focused on characterizing changes in the proportion of a new variant \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
 When incidence is changing exponentially ($i_x(t) = i_x(t_0) \exp(r_x t)$), the proportion of the new variant $p(t)$ follows a logistic growth curve \citep{switzerland2021variant,davies2021estimated}:
@@
 For example, we can estimate $\rho$ by plotting the estimated strength of the variant $\hat{\RR}_{\textrm{var}}(t)$ against the estimated strength of the wild type $\hat{\RR}_{\textrm{wt}}(t)$---as presented in Figure 2 of \cite{volz2021transmission}---and performing a linear regression (\fref{Rtbias}C,F,I).
-Here, we estimate the slope while fixing the intercept: in theory, the regression line should go through the origin in theory because $\Rv = 0$ when $\Rw = 0$.
+Here, we estimate the slope while fixing the intercept to zero: the regression line should go through the origin in theory because $\Rv = 0$ when $\Rw = 0$.
 If $\rho$ is constant, and generation-interval distributions are correctly specified, we obtain a straight line with a slope of $\rho$ and intercept at zero (\fref{Rtbias}F).
@@
 We explored how the generation-interval distribution shapes the link between relative strength and speed.
-While the role of generation-interval distributions in linking epidemic strength and speed has been well-established, this framework can also provide provides insight into an ut under-appreciated concept: the relative strength and speed of new variants also depend on both current epidemic conditions and the nature of control measures (e.g., strength- or speed-like) to be introduced.
-For example, under a constant-strength intervention that reduces the transmission rate of the wild type and a variant by the same amount (and therefore, keeping the relative strength constant), the relative speed of the variant is sensitive to the epidemic growth rate of the wild type strain.
-Our results challenge the commonly made assumption that relative speed of new variants remains constant over time when characterizing changes in the proportion of new variants.
+While the role of generation-interval distributions in linking epidemic strength and speed is well-established, this framework also provides insight into an under-appreciated phenomenon: the relative strength and speed of new variants also depend on both current epidemic conditions and the nature of proposed control measures (e.g., strength- or speed-like) to be introduced.
+For example, under a constant-strength intervention that reduces the transmission rate of the wild type and a variant by the same amount (and therefore, keeps the relative strength constant), the relative speed of the variant depends on the epidemic growth rate of the wild type strain.
+Our results challenge the assumption commonly made when characterizing changes in the proportion of new variants that the relative speed of new variants remains constant over time.
 
 Differences in the generation-interval distributions further exaggerate this effect.
-Therefore, neglecting potential differences in the mean generation intervals between the variant and the wild type can bias estimates of the its transmission advantage.
+Therefore, neglecting potential differences in the mean generation intervals between the variant and the wild type can bias estimates of the variant's transmission advantage.
 These biases may be assessed by considering whether estimates of relative strength appear to vary systematically with the direction of changes in the incidence of infections caused by the variant.
@@
 
-This perspective can shed light on earlier estimates of relative strength of the Alpha variant.
+This perspective sheds light on earlier estimates of relative strength of the Alpha variant.
 Mathematical analyses typically reported greater than a 1.4 fold increase in reproduction number for the Alpha variant ($\rho > 1.4$) whereas an independent analysis of secondary attack rates from contact tracing data suggested a somewhat smaller 1.25--1.4 fold increase \citep{ukinvest}.
-As shown in \cite{davies2021estimated} and \cite{volz2021transmission}, propagating uncertainty in generation interval estimates (and, in particular, assuming shorter generation intervals) can partially explain these differences:
-For example, if we consider short and wide generation interval estimates from Tianjin, China ($\Gw=\Gv=2.57\days$ and $\kappa=1$; \cite{ganyani2020estimating}), we obtain $\rho=1.26$ (from $\delta=0.1\pday$ and $\rw = 0\pday$).
-Although these estimates are more consistent with the attack rate analysis \citep{ukinvest},
-we do not claim that they are necessarily more accurate.
-While individual-level data from contact tracing can provide a more reliable source of information about the transmissibility and time scale of transmission in some cases, they can also be biased towards particular types of contacts---for example, household contacts are probably more likely to be identified---which could also affect the estimate of $\rho$.
+As shown in \cite{davies2021estimated} and \cite{volz2021transmission}, propagating uncertainty in generation interval estimates (and, in particular, assuming shorter generation intervals) partially explains these differences:
+For example, if we consider a generation-interval distribution estimate with a short mean and wide variability from Tianjin, China ($\Gw=\Gv=2.57\days$ and $\kappa=1$; \cite{ganyani2020estimating}), we obtain $\rho=1.26$ (from $\delta=0.1\pday$ and $\rw = 0\pday$).
+Although these estimates are more consistent with the attack rate analysis \citep{ukinvest}, they are necessarily more accurate.
+While individual-level data based on contact tracing can provide a more reliable source of information about the transmissibility and time scale of transmission in some cases, they can also be biased towards particular types of contacts---for example, household contacts are more likely to be identified---which could also affect the estimate of $\rho$.
 Instead, this calculation simply highlights the importance of carefully considering generation-interval distributions in assessing the relative strength and speed of SARS-CoV-2 variants.
@@
 Other studies have also used generation-interval-based arguments to explore changes in estimates of relative strength. 
-For example, Volz \textit{et al.} estimated that the relative strength of the Alpha variant declined in England between December 2020 and January 2021 \cite{volz2021transmission};
-they hypothesized that a shorter generation interval of the Alpha variant could explain this phenomenon by reducing the relative speed (and therefore, the relative strength) under intervention measures.
+For example, Volz \textit{et al.} estimated that the relative strength of the Alpha variant declined in England between December 2020 and January 2021 \cite{volz2021transmission}.
+They hypothesized that a shorter generation interval of the Alpha variant could explain this phenomenon by reducing the Alpha variant's relative speed (and therefore its relative strength) under intervention measures.
 However, our analysis suggests that a shorter generation interval of a new variant cannot explain the decline in relative strength.
 Under a constant-strength intervention, the relative speed decreases (as predicted by  \cite{volz2021transmission}) but the relative strength remains constant (\fref{control_sim}B). 
-Under a constant-speed intervention, the relative speed remains constant (\fref{control_sim}F), but the relative strength increases because the intervention will have a smaller effect on the variant with a shorter generation interval (\fref{control_sim}E).
+Under a constant-speed intervention, the relative speed remains constant (\fref{control_sim}F), but the relative strength increases because the intervention will have less effect on the variant with a shorter generation interval (\fref{control_sim}E).
 
-We have primarily focused on the differences in the mean generation interval, but differences in the amount of variability, characterized by the squared coefficient of variation $\kappa$, could also have important, but different implications \citep{blanquart2021selection}.
-For example, if two variants have identical reproduction numbers, the variant with a wider generation-interval distribution will always have a faster growth rate and out-compete the other variant (except when both variants have $\RR=1$).
-This is because a variant with a wider generation-interval distribution can take advantage of more early transmission during the growth phase and more late transmission during the decay phase.
+We have focused on the differences in the mean generation interval, but differences in the amount of variability, characterized by the squared coefficient of variation $\kappa$, could also have important, but different implications \citep{blanquart2021selection}.
+For example, if two variants have identical reproduction numbers, the variant with a wider generation-interval distribution will always out-compete the other variant (except when both variants have $\RR=1$) because a variant with a wider generation-interval distribution can take advantage of more early transmission during the growth phase and more late transmission during the decay phase.
 
@@
 However, these estimates are derived from serial intervals (i.e., time between symptom onset of the infector and the infectee; \cite{svensson2007note}), which are subject to dynamical biases \citep{park2021forward} and fail to account for asymptomatic transmission \citep{park2020time}.
-Recently, it has been further suggested that the generation intervals of the original SARS-CoV-2 strain may be considerably longer than previously thought \citep{sender2021unmitigated}, adding further uncertainty.
-A few studies have tried to compare generation- and serial-interval distributions of new SARS-CoV-2 variants with those of the original wild type strain \citep{pung2021serial, hart2021generation}, these comparisons are inherently difficult especially due to temporal changes in generation- and serial-interval distributions caused by dynamical and intervention effects \citep{ali2020serial,park2021forward}.
+A recent study has further suggested that the generation intervals of the original SARS-CoV-2 strain may be considerably longer than previously thought \citep{sender2021unmitigated}, adding further uncertainty.
+A few studies have tried to compare generation- and serial-interval distributions of new SARS-CoV-2 variants with those of the original wild type strain \citep{pung2021serial, hart2021generation};
+these comparisons are inherently difficult due to temporal changes in generation- and serial-interval distributions caused by dynamical and intervention effects \citep{ali2020serial,park2021forward}.
 Future studies should prioritize detailed assessment of the generation intervals of SARS-CoV-2 and widespread variants, as well as consider how uncertainty in generational intervals might bias conclusions \citep{doi:10.1098/rsif.2020.0144,ali2020serial,gostic2020practical}.

### 2022-03-04 newvariant.tex
done

@@
 Analyses of new variants have characterized \emph{relative} strength (i.e., the ratio of reproduction numbers of the invading and resident strains, here called $\rho$) and speed (i.e., the difference between growth rates of the invading and resident strains, here called $\delta$).
-As an example, we consider a new variant invading a wild type strain in this paper and use $\Rv$ and $\Rw$ to denote their reproduction numbers and $\rv$ and $\rw$ to denote their growth rates at a given time (and therefore $\rho = \Rv/\Rw$ and $\delta = \rv-\rw$).
+As an example, we consider a new variant invading a wild type strain in this paper and use $\Rv$ and $\Rw$ to denote their respective reproduction numbers and $\rv$ and $\rw$ to denote their respective growth rates at a given time (and therefore $\rho = \Rv/\Rw$ and $\delta = \rv-\rw$).
 Many analyses have focused on changes in the \emph{proportion} of a new variant to estimate its relative speed \citep{switzerland2021variant, davies2021estimated, di2021impact, graham2021changes, leung2021early, volz2021transmission,zhao2021}.
@@
 In contrast, a constant-speed intervention reduces transmission after infection by a constant rate $\phi$ across generation intervals: $K_{\mathrm{post}} = K_{\mathrm{pre}}(\tau) \exp(-\phi \tau)$.
-In this case, the intervention speed $\phi$ must be greater than the epidemic speed $r$ to control the epidemic \citep{doi:10.1098/rspb.2020.1556}.
+In this case, the intervention speed $\phi$ must be greater than the epidemic speed $r$ to control the epidemic.
 
@@
 Future studies should prioritize detailed assessment of the generation intervals of SARS-CoV-2 and widespread variants, as well as consider how uncertainty in generational intervals might bias conclusions \citep{doi:10.1098/rsif.2020.0144,ali2020serial,gostic2020practical}.
+Combining sequencing or viral load trajectory data can also help narrow estimates of generation- and serial-interval distributions as well as time-varying reproduction numbers even when contact tracing data are limited \citep{lin2020viral,Stockdale2022.02.23.22271355}.
 

### 2022-05-03 response_final.tex
really done

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}

\usepackage{color}

\usepackage{xspace}

\usepackage{lmodern}

\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}

\usepackage[pdfencoding=auto, psdextra]{hyperref}

\usepackage{natbib}
\bibliographystyle{chicago}

\newcommand{\eref}[1]{Eq.~(\ref{eq:#1})}
\newcommand{\fref}[1]{Fig.~\ref{fig:#1}}

%% Consistent, changeable style for subscripts
\newcommand{\vvvar}{\mathrm{var}}
\newcommand{\wwwt}{\mathrm{wt}}

\newcommand{\rx}[1]{\ensuremath{{r}_{#1}}\xspace} 
\newcommand{\ry}[1]{\rx{\mathrm{#1}}} 
\newcommand{\rw}{\rx{\wwwt}}
\newcommand{\rv}{\rx{\vvvar}}

\newcommand{\Rx}[1]{\ensuremath{{\mathcal R}_{#1}}\xspace} 
\newcommand{\Ry}[1]{\Rx{\mathrm{#1}}}
\newcommand{\Ro}{\Rx{0}}
\newcommand{\RR}{\ensuremath{{\mathcal R}}\xspace}
\newcommand{\Rw}{\Rx{\wwwt}}
\newcommand{\Rv}{\Rx{\vvvar}}

\newcommand{\dd}[1]{\ensuremath{\, \mathrm{d}#1}}
\newcommand{\dtau}{\dd{\tau}}
\newcommand{\dx}{\dd{x}}
\newcommand{\dsigma}{\dd{\sigma}}

\newcommand{\rev}{\subsection*}
\newcommand{\revtext}{\textsf}
\setlength{\parskip}{\baselineskip}
\setlength{\parindent}{0em}

\newcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textsl{#3}\textbf{]}}}
\newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}
\newcommand{\dc}[1]{\comment{blue}{DC}{#1}}
\newcommand{\jsw}[1]{\comment{green}{JSW}{#1}}
\newcommand{\hotcomment}[1]{\comment{red}{HOT}{#1}}

\newcommand{\psymp}{\ensuremath{p}} %% primary symptom time
\newcommand{\ssymp}{\ensuremath{s}} %% secondary symptom time
\newcommand{\pinf}{\ensuremath{\alpha_1}} %% primary infection time
\newcommand{\sinf}{\ensuremath{\alpha_2}} %% secondary infection time

\newcommand{\psize}{{\mathcal P}} %% primary cohort size
\newcommand{\ssize}{{\mathcal S}} %% secondary cohort size

\newcommand{\gtime}{\tau_{\rm g}} %% generation interval
\newcommand{\gdist}{g} %% generation-interval distribution
\newcommand{\idist}{\ell} %% incubation period distribution

\newcommand{\total}{{\mathcal T}} %% total number of serial intervals

\begin{document}

\noindent Dear Editors:

We are submitting the final version our manuscript, “The importance of the generation interval in investigating dynamics and control of new SARS-CoV-2 variants”.

Thank you for your consideration of our submission.

Sincerely,

Sang Woo Park

\rev{Reviewer \#1}

\revtext{When I started reviewing this paper, I was highly skeptical of the notions of strength and speed for epidemics and interventions. This paper has convinced me that these are actually very useful concepts, certainly in theory if not necessarily in practice. The assumptions are mostly clearly explained, but the assumption of homogeous mass-action should be made explicit early in the paper (though many of the qualitative insights probably transfer beyond this class of models). While the results are shown in only a special case (homogeneous massaction with a transmission kernal that is constant in t and a gamma-distributed generation interval), it is an important special case because it is similar to many models used in practice. The discussion clearly addresses situations where the strength/speed dichotomy and several assumptions might break down, but it makes a good case that the qualitative insights are still useful. I have reviewed very few papers that changed my mind about something so important, and I expect this will be equally useful to many other readers. My only general comment is that the presentation of the basic insights and the figures sometimes seem scattered; I found myself jumping around quite a bit while reviewing the paper. I have a few more specific comments below.}

Thank you for your review. We believe that a part of the scattered-ness is caused by pdf conversion process from a latex document---for example, figure 3 is described in section 5 (pages 10--11) but appears in page 12 due to its size. Likewise, figures 4 and 5 have been pushed back to the end of section 6. We think that these issues will be less problematic when the paper is properly formatted according to the journal guideline. Otherwise, we have made minor changes throughout the manuscript to try to tie the results closer to the figures whenever appropriate and re-ordered some text for clearer presentation.

\revtext{(page 3, lines 43–44) It is not true that R determines the final size of an epidemic in a homogeneous population. For example, the final size of an epidemic in a homogeneous population on a configuration model network depends on the degree distribution as well as $\RR$, and this final size corresponds to the mass-action final size with the same R only when the degree distribution is Poisson. It would be accurate to say that $\RR$ determines the final size of an epidemic in a homogeneous mass-action model. The entire paper is based on homogeneous mass-action models, so it help to make this explicit.}

We have reworded this sentence as follows:

``The epidemic strength also determines the final size of an epidemic in a homogeneously mixing population under the mass-action assumption''

\revtext{(page 4, lines 90 and 92) ``$K_{\mathrm{post}}$'' should be ``$K_{\mathrm{post}}(\tau)$''}

Done.

\revtext{(page 4, lines 93–94) The ``kernel'' needs to be explained more clearly before the statement that intervention speed ($\phi$) must be greater than epidemic speed ($r$) in order to control the epidemic. The explanation given could be interpreted to mean that the pre-intervention generation interval distribution $g_x(\tau)$ should be replaced by something proportional
to $g_x(\tau) \exp(-\phi \tau)$. If you then imitate the derivation from Wallinga and Lipsitch (American Journal of Epidemiology, 2007) using the pre- and post-intervention kernels, you get $\RR_{\textrm{pre}} = M_g(-r)^{-1}$ and $$\RR_{\textrm{post}} = \frac{M_g(-\phi)}{M_g(-r-\phi)}$$
where $M_g()$ is the moment generating function for the PDF $g(\tau)$. This
does not lead to $\RR_{\textrm{post}} < 1$ if $\phi > r$. However, we do get $\RR_{\textrm{post}} < 1$ if $\phi < -r$, which would seem to correspond to an intervention that makes people more infectious over time. 
This counterintuitive result comes from the need to re-normalize $g(\tau) e^{-\phi \tau}$.
In order to get a PDF that integrates to one, and it shows the importance of clearly maintaining the distinction between the generation interval distribution and the transmission kernel.}

Thank you for pointing this out. We agree that making the distinction between the generation-interval distribution and the transmission kernel is important. We have rephrased this paragraph to try to make the distinction clearer.

``While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the pathogen and its control \citep{doi:10.1098/rspb.2020.1556}.
To illustrate the differences in conclusions when holding $\RR$ or $r$ fixed, we consider two idealized interventions of constant strength and constant speed.
Before these interventions are introduced, the dynamics of pathogen spread can be characterized in terms of the pre-intervention kernel $K_{\mathrm{pre}}(\tau)$, which represents the rate at which an infected individual generates secondary infections $\tau$ time units after infection.
The pre-intervention kernel can be further decomposed in terms of the epidemic strength $\RR$ and the generation-interval distribution $g(\tau)$: $K_{\mathrm{pre}}(\tau) = \RR g(\tau)$;
therefore, the infection knerel integrates to $\RR$ while the generation-interval distribution integrates to 1.
Then, a constant-strength intervention reduces transmission by a constant factor $\theta$ throughout infection such that the post-intervention kernel is $K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau)/\theta$.
In this case, the intervention strength $\theta$ must be greater than the epidemic strength $\RR$ to control the epidemic.
In contrast, a constant-speed intervention reduces transmission after infection by a constant rate $\phi$ throughout infection: $K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau) \exp(-\phi \tau)$.
In this case, the intervention speed $\phi$ must be greater than the pre-intervention epidemic speed $r$ to control the epidemic.
We note that the resulting post-intervention generation-interval distribution under a constant-speed intervention is \emph{not} equal to $g(\tau) \exp(-\phi \tau)$; instead $g(\tau) \exp(-\phi \tau)$ needs to be renormalized to integrate to 1 because the generation-interval distribution is a probability distribution.''

\revtext{(page 6, line 192) It would be helpful to point out that $\kappa^{-1}$ is the gamma shape parameter.}

Done.

\revtext{(page 9, lines 270–277) It would help to explain that Rwt = 0.49 corresponds to the top left of Figure 2 and Rwt = 1.27 corresponds to the bottom right.}

Done.

\revtext{(page 10, lines 282–290) I appreciate the careful distinction made between
the case reproduction number and the instantaneous reproduction number
as well as the explanation of why the instantaneous reproduction is the
more relevant here}

Thank you.

\revtext{(page 14, Figure 4) Subfigures A, C, E and B, D, F should have a label
for the x-axis. To make room, the y-axis labels in G and H could be just
Rvar and rvar, respectively.}

We have added axis titles for the subfigures.

\revtext{(page 19, lines 513–529) In practice, generation intervals can be observed
``forwards'' and ``backwards'' but both are subject to bias (as explained in
Ref. 19). Do the authors have an opinion about which of these would allow
a more reliable estimate of the relative length of the generation intervals
of a variant compared to wild type?}

We have added a new paragraph:

``While detailed contact tracing data can help narrow down uncertainties in the generation-interval distributions, there are additional complexities to comparing generation-interval estimates of different variants.
The generation intervals can be measured in two different ways: backward and forward \citep{champredon2015intrinsic}.
The backward distribution starts from the cohort of infectees who were infected at the same time and looks at when their infectors were infected. 
The backward distribution is subject to dynamical biases because we are more likely to observe recent infections (and therefore shorter generation intervals) when the epidemic is growing.
On the other hand, the forward distribution starts from the cohort of infectors who were infected at the same time and looks at when they transmitted infections.
While the forward distribution is more reliable, changes in epidemic dynamics due to intervention measures and susceptible depletions can affect the shape of the distribution \citep{ali2020serial, nishiura2010time}.
For example, comparing generation-interval estimates from different intervention periods will necessarily bias the true differences in the generation-interval distributions of different variants.''

\bibliography{newvariant_abbv.bib}

\end{document}

### 2022-05-03 response_new.tex
really done

@@
 
-\noindent Dear Editor:
+\noindent Dear Editors:
 
@@
 
-We had previously submitted the manuscript to JRSI (rsif-2021-0438) and were asked to revise with positive comments---for example, reviewer 3 said “The analysis is mathematically sound and offers a thorough investigation of an important concept that has been the subject of much discussion in the epidemiological community but, to my knowledge, hasn't been given the formal treatment it deserves until now.” We had submitted the revised version of the manuscript but the paper was rejected despite having addressed all comments due to reviewer 2's comment that “On a second reading of this manuscript, I am much less optimistic about its value to infectious disease epidemiology.” We emailed the editor since then and were provided the opportunity to resubmit.
-We have made revisions to our manuscript to address the reviewer's comments and to better highlight the novelty of our work.
+We had previously submitted the manuscript to JRSI (rsif-2021-0438) and were asked to revise with positive comments---for example, reviewer 3 said “The analysis is mathematically sound and offers a thorough investigation of an important concept that has been the subject of much discussion in the epidemiological community but, to my knowledge, hasn't been given the formal treatment it deserves until now.”
+
+We revised the manuscript, addressing all comments, but the paper was rejected due to reviewer 2's comment that “On a second reading of this manuscript, I am much less optimistic about its value to infectious disease epidemiology.” We emailed the editor about this and were provided the opportunity to resubmit.
+We have now revised the manuscript to better highlight the novelty of our work.
 
@@
 Previous studies have particularly prioritized estimating the relative strength (i.e., the ratio of reproduction numbers) from relative speed (i.e., the differences between growth rates).
-In doing so, analyses of new variants have focused on estimating the relative speed from the observed changes in the proportion of new variants while assuming that relative speed does not change over the course of invasion.
-Here, we challenge this assumption by showing how relative strength and speed can change in response to different kinds of epidemic interventions and argue that holding relative strength fixed is generally more appropriate in the context of modeling the spread of SARS-CoV-2 variants.
-In doing so, we consider the possibility that new variants can have different generation intervals from the original variant and explore how such differences affect the dynamics and control of new variants.
-We also show that neglecting these differences can lead to biased inference and provide a practical guidance for how such biases can be assessed in practice.
-Below please find our detailed responses to reviewers.
+
+In this study, we challenge a common, sometimes implicit, assumption in estimating transmission advantage of new variants: that relative speed will remain constant.
+We argue that constant relative strength is generally (though not always) more appropriate under changing conditions.
+We also consider the possibility that new variants may have different generation intervals from the original variant and explore how such differences affect the dynamics and control of new variants. 
+We discuss how neglecting these differences can lead to biased inference and provide a practical guidance for how such biases can be assessed in practice.
 

### 2022-05-03 newvariant.tex
really done

@@
 A pathogen can spread in a population if $\RR>1$ \citep{diekmann1990definition}.
-The epidemic strength also determines the final size of an epidemic in a homogeneous population \citep{anderson1991infectious}.
+The epidemic strength also determines the final size of an epidemic in a homogeneously mixing population under the mass-action assumption \citep{anderson1991infectious}.
 Epidemic ``speed'' is characterized by the growth rate $r$, which has units of $1/\mathrm{time}$ and describes the exponential rate of pathogen spread at the population level.
@@
 To illustrate the differences in conclusions when holding $\RR$ or $r$ fixed, we consider two idealized interventions of constant strength and constant speed.
-Given a pre-intervention kernel $K_{\mathrm{pre}}(\tau) = \RR g(\tau)$, which represents the rate at which an infected individual generates secondary infections $\tau$ time units after infection, a constant-strength intervention reduces transmission by a constant factor $\theta$ across generation intervals such that the post-intervention kernel is $K_{\mathrm{post}} = K_{\mathrm{pre}}(\tau)/\theta$.
+Before these interventions are introduced, the dynamics of pathogen spread can be characterized in terms of the pre-intervention kernel $K_{\mathrm{pre}}(\tau)$, which represents the rate at which an infected individual generates secondary infections $\tau$ time units after infection.
+The pre-intervention kernel can be further decomposed in terms of the epidemic strength $\RR$ and the generation-interval distribution $g(\tau)$: $K_{\mathrm{pre}}(\tau) = \RR g(\tau)$;
+therefore, the infection kernel integrates to $\RR$ while the generation-interval distribution integrates to 1.
+Then, a constant-strength intervention reduces transmission by a constant factor $\theta$ throughout infection such that the post-intervention kernel is $K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau)/\theta$.
 In this case, the intervention strength $\theta$ must be greater than the epidemic strength $\RR$ to control the epidemic.
-In contrast, a constant-speed intervention reduces transmission after infection by a constant rate $\phi$ across generation intervals: $K_{\mathrm{post}} = K_{\mathrm{pre}}(\tau) \exp(-\phi \tau)$.
+In contrast, a constant-speed intervention reduces transmission after infection by a constant rate $\phi$ throughout infection: $K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau) \exp(-\phi \tau)$.
+In this case, the intervention speed $\phi$ must be greater than the pre-intervention epidemic speed $r$ to control the epidemic.
+We note that the resulting post-intervention generation-interval distribution under a constant-speed intervention is \emph{not} equal to $g(\tau) \exp(-\phi \tau)$; instead $g(\tau) \exp(-\phi \tau)$ needs to be renormalized to integrate to 1 because the generation-interval distribution is a probability distribution.
+
+Given a pre-intervention kernel $K_{\mathrm{pre}}(\tau) = \RR g(\tau)$, which represents the rate at which an infected individual generates secondary infections $\tau$ time units after infection, a constant-strength intervention reduces transmission by a constant factor $\theta$ across generation intervals such that the post-intervention kernel is $K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau)/\theta$.
+In this case, the intervention strength $\theta$ must be greater than the epidemic strength $\RR$ to control the epidemic.
+In contrast, a constant-speed intervention reduces transmission after infection by a constant rate $\phi$ across generation intervals: $K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau) \exp(-\phi \tau)$.
 In this case, the intervention speed $\phi$ must be greater than the epidemic speed $r$ to control the epidemic.
@@
 \end{equation}
+We note that $\kappa_x^{-1}$ also corresponds to the shape parameter of a Gamma distribution.
 Various Gamma-generation-interval assumptions have been widely used in epidemic modeling, including for models of SARS-CoV-2 \citep{doi:10.1098/rsif.2020.0144}.
@@
 
-% We thus ask: what factors affect the relative strength $\rho = \Rv/\Rw$ of a new variant, conditional on an observed relative speed $\delta$?
+We thus ask: what factors affect the relative strength $\rho = \Rv/\Rw$ of a new variant, conditional on an observed relative speed $\delta$?
 Inference of the relative strength $\rho = \Rv/\Rw$ from the relative speed $\delta$ depends on assumptions about the generation-interval distributions of both strains.
@@
 
-Unsurprisingly, we find that an increased speed of $\delta=0.1\pday$ for the variant is consistent with higher strength than the wild type across the range of epidemiological conditions considered (\fref{relstrength}).
+Unsurprisingly, we find that an increased speed of $\delta=0.1\pday$ for the variant is consistent with higher strength ($\rho > 1$) than the wild type across all five epidemiological scenarios considered (\fref{relstrength}).
 However, the magnitude of relative strength $\rho$ is sensitive to assumptions about generation intervals:
-For realistic values of $\kappa$ (excluding 0 and 1), the inferred relative strength $\rho$ ranges between 1.1 and 2.3 when $\Gv$ is allowed to vary between $2/3$ and $3/2$ of \Gw.
+For realistic values of $\kappa$ (excluding 0 and 1), the inferred relative strength $\rho$ ranges between 1.1 and 2.3 (across all five scenarios) when $\Gv$ is allowed to vary between $2/3$ and $3/2$ of \Gw.
 
-In general, longer mean generation intervals of the new variant translate to higher values of $\rho$ (and vice versa), except when $\rv \leq 0$ (recall, we always assume $\rw<\rv$).
+In general, longer mean generation intervals of the new variant translate to higher values of $\rho$ (and vice versa;  \fref{relstrength}, bottom panels), except when $\rv \leq 0$ (recall, we always assume $\rw<\rv$; \fref{relstrength}, top panels).
 When $\rv = 0$, we always have $\Rv = 1$ and so $\rho$ is independent of the generation-interval distribution of the new variant.
@@
 In general, longer generation intervals lead to slower relative speed of the variant when the incidence of both strains is increasing (\fref{relspeed}, bottom panels) because slower growth of the variant reduces the differences in absolute speed.
-When $\Rv=1$, the relative speed is insensitive to the generation-interval distribution of the variant because we always have $\rv=0$.
-When $\Rv<1$, longer generation intervals of the variant lead to slower decay ($\rv$ closer to 0), and therefore, greater relative speed.
+When $\Rv=1$, the relative speed is insensitive to the generation-interval distribution of the variant because we always have $\rv=0$ (\fref{relspeed}, top-right panel).
+When $\Rv<1$, longer generation intervals of the variant lead to slower decay ($\rv$ closer to 0), and therefore, greater relative speed (\fref{relspeed}, top-left panel).
 We see that assuming a narrower distribution for both the variant and the wild type strain (lower $\kappa$) has qualitatively similar effects as assuming a longer mean (leading to lower values of $\delta$ when $\Rv > 0$ and higher values of $\delta$ when $\Rv < 0$).
@@
 \fref{relspeed} also shows that when $\rho$ is fixed, relative speed depends on underlying epidemiological conditions---specifically, the absolute strength of the two strains. 
-For example, even when the generation-interval distributions are identical ($\Gv=\Gw$, in this case), changing $\Rw$ from 0.49 to 1.27 (and thus $\Rv$ from 0.79 to 2.04) changes the relative speed $\delta$ from 0.08 to 0.11 when $\kappa=0.2$---and from 0.06 to 0.14 when $\kappa=0.5$.
+For example, even when the generation-interval distributions are identical ($\Gv=\Gw$, in this case), changing $\Rw$ from 0.49 (\fref{relspeed}, top-left panel) to 1.27 (\fref{relspeed}, bottom-right panel)---and thus $\Rv$ from 0.79 to 2.04---changes the relative speed $\delta$ from 0.08 to 0.11 when $\kappa=0.2$ (and from 0.06 to 0.14 when $\kappa=0.5$).
 Differences in the generation-interval distributions exaggerate these changes.
@@
 
-Instead of inferring relative strength from relative speed, one can separately estimate time-varying reproduction numbers $\RR(t)$ of the variant and the wild type from observed incidence and assumptions about the generation intervals, and then calculate the ratio---such methods have been used in previous analyses of the Alpha \citep{volz2021transmission} and Delta variants \citep{abbott2021estimating}.
+Instead of inferring relative strength from relative speed, one can separately estimate time-varying reproduction numbers $\RR(t)$ of the variant and the wild type from observed incidence and assumptions about the generation intervals, and then calculate the ratio---such methods have been used in previous analyses of the Alpha \citep{volz2021transmission}, Delta \citep{abbott2021estimating}, and Omicron variants \citep{pearson2021bounding}.
 Broadly, there are two types of time-varying reproduction numbers: case reproduction number and instantaneous reproduction number.
@@
 
-When the assumed distribution matches the true distribution, the estimated reproduction numbers match the true values (\fref{Rtbias}D); thus, their ratio remains constant and $\hat{\RR}_{\textrm{var}}(t)=\Rv(t)$ (\fref{Rtbias}E).
-However, when the generation-interval distribution of the variant differs from the assumed distribution, the ratio changes over time (\fref{Rtbias}B,H).
-If the true generation intervals of the variant have a shorter mean than the assumed distribution, we over-estimate $\Rv(t)$ during the growth phase and under-estimate $\Rv(t)$ during the decay phase (and conversely, \fref{Rtbias}A,G), which further translates to biases in the estimated relative strength (\fref{Rtbias}B,H).
+When the wild type and the variant have the same (known) generation-interval distributions, the assumed generation-interval distribution matches the true distribution (\fref{Rtbias}D--F).
+In this case, the estimated reproduction numbers match the true values (\fref{Rtbias}D); thus, their estimated ratio $\hat{\rho}(t)$ remains constant and $\hat{\RR}_{\textrm{var}}(t)=\Rv(t)$ (\fref{Rtbias}E).
+However, when the true generation intervals of the variant have a shorter mean than the assumed distribution, we over-estimate $\Rv(t)$ during the growth phase and under-estimate $\Rv(t)$ during the decay phase (\fref{Rtbias}A);
+therefore, the estimated ratio $\hat{\rho}(t)$ changes over time (\fref{Rtbias}B).
+Conversely, when the true generation intervals of the variant have a longer mean than the assumed distribution, we find similar biases in opposite directions (\fref{Rtbias}G--H).
 
@@
 If $\rho$ is constant, and generation-interval distributions are correctly specified, we obtain a straight line with a slope of $\rho$ and intercept at zero (\fref{Rtbias}F).
-However, when the assumed mean generation interval is longer than the that of the variant, we over-estimate the slope (and conversely, \fref{Rtbias}C,I).
+However, when the assumed mean generation interval is longer than the that of the variant (\fref{Rtbias}C), we over-estimate the slope (and conversely, \fref{Rtbias}I).
 
@@
 
+While detailed contact tracing data can help narrow down uncertainties in the generation-interval distributions, there are additional complexities to comparing generation-interval estimates of different variants.
+The generation intervals can be measured in two different ways: backward and forward \citep{champredon2015intrinsic}.
+The backward distribution starts from the cohort of infectees who were infected at the same time and looks at when their infectors were infected. 
+The backward distribution is subject to dynamical biases because we are more likely to observe recent infections (and therefore shorter generation intervals) when the epidemic is growing.
+On the other hand, the forward distribution starts from the cohort of infectors who were infected at the same time and looks at when they transmitted infections.
+While the forward distribution is more reliable, changes in epidemic dynamics due to intervention measures and susceptible depletions can affect the shape of the distribution \citep{ali2020serial, nishiura2010time}.
+For example, comparing generation-interval estimates from different intervention periods will necessarily bias the true differences in the generation-interval distributions of different variants.
+
 Even though SARS-CoV-2 has been spreading for more than a year, there is still considerable uncertainty about its generation intervals.

### 2022-05-03 control.R

@@ -115,8 +115,7 @@ g1 <- ggplot(gendata_short) +
   theme(
     panel.grid = element_blank(),
     legend.position = c(0.6, 0.95),
-    legend.background = element_rect(fill=NA),
-    axis.title = element_blank()
+    legend.background = element_rect(fill=NA)
   )
 
 g2 <- ggplot(gendata_short) +
@@ -134,8 +133,7 @@ g2 <- ggplot(gendata_short) +
   theme(
     panel.grid = element_blank(),
     legend.position = c(0.6, 0.95),
-    legend.background = element_rect(fill=NA),
-    axis.title = element_blank()
+    legend.background = element_rect(fill=NA)
   )
 
 g3 <- ggplot(gendata_base) +
@@ -153,8 +151,7 @@ g3 <- ggplot(gendata_base) +
   theme(
     panel.grid = element_blank(),
     legend.position = "none",
-    legend.background = element_rect(fill=NA),
-    axis.title = element_blank()
+    legend.background = element_rect(fill=NA)
   )
 
 g4 <- ggplot(gendata_base) +
@@ -172,8 +169,7 @@ g4 <- ggplot(gendata_base) +
   theme(
     panel.grid = element_blank(),
     legend.position = "none",
-    legend.background = element_rect(fill=NA),
-    axis.title = element_blank()
+    legend.background = element_rect(fill=NA)
   )
 
 g5 <- ggplot(gendata_long) +
@@ -185,8 +181,7 @@ g5 <- ggplot(gendata_long) +
   scale_y_continuous("Kernel density", limits=c(0, 0.4)) +
   ggtitle("E. Longer GI, Constant-strength intervention") +
   theme(
-    panel.grid = element_blank(),
-    axis.title = element_blank()
+    panel.grid = element_blank()
   )
 
 g6 <- ggplot(gendata_long) +
@@ -198,8 +193,7 @@ g6 <- ggplot(gendata_long) +
   scale_y_continuous("Kernel density", limits=c(0, 0.4)) +
   ggtitle("F. Longer GI, Constant-speed intervention") +
   theme(
-    panel.grid = element_blank(),
-    axis.title = element_blank()
+    panel.grid = element_blank()
   )
 
 g7 <- ggplot(intervention) +

### 2022-05-03 newvariant.tex
delete

@@
 We note that the resulting post-intervention generation-interval distribution under a constant-speed intervention is \emph{not} equal to $g(\tau) \exp(-\phi \tau)$; instead $g(\tau) \exp(-\phi \tau)$ needs to be renormalized to integrate to 1 because the generation-interval distribution is a probability distribution.
-
-Given a pre-intervention kernel $K_{\mathrm{pre}}(\tau) = \RR g(\tau)$, which represents the rate at which an infected individual generates secondary infections $\tau$ time units after infection, a constant-strength intervention reduces transmission by a constant factor $\theta$ across generation intervals such that the post-intervention kernel is $K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau)/\theta$.
-In this case, the intervention strength $\theta$ must be greater than the epidemic strength $\RR$ to control the epidemic.
-In contrast, a constant-speed intervention reduces transmission after infection by a constant rate $\phi$ across generation intervals: $K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau) \exp(-\phi \tau)$.
-In this case, the intervention speed $\phi$ must be greater than the epidemic speed $r$ to control the epidemic.
 

### 2022-05-03 newvariant.tex
minor changes

@@
 
+These two idealized interventions, in turn, allow us to understand how holding $\RR$ or $r$ fixed can lead to different conclusions about epidemic control.
 For example, if we assume epidemic strength $\RR$ is known, then variation in the generation-interval distribution does not change the estimated effectiveness of a constant-strength intervention.
@@
 \end{equation}
-We can approximate this $r$--$\RR$ relationship by assuming that the generation-interval distribution is Gamma-distributed, and summarizing it using the mean generation interval $\bar{G}_x$ and the squared coefficient of variation $\kappa_x$ \citep{park2019practical}:
+We can approximate this $r$--$\RR$ relationship by assuming that the generation-interval distribution is Gamma-distributed, and summarizing it using the mean generation interval $\bar{G}_x$ and the squared coefficient of variation $\kappa_x$ (equal to the reciprocal of the Gamma shape parameter) \citep{park2019practical}:
 \begin{equation}
@@
 \end{equation}
-We note that $\kappa_x^{-1}$ also corresponds to the shape parameter of a Gamma distribution.
 Various Gamma-generation-interval assumptions have been widely used in epidemic modeling, including for models of SARS-CoV-2 \citep{doi:10.1098/rsif.2020.0144}.
@@
 We thus ask: what factors affect the relative strength $\rho = \Rv/\Rw$ of a new variant, conditional on an observed relative speed $\delta$?
-Inference of the relative strength $\rho = \Rv/\Rw$ from the relative speed $\delta$ depends on assumptions about the generation-interval distributions of both strains.
+This inference depends on assumptions about the generation-interval distributions of both strains.
 Given the mean generation interval of the variant $\Gv$ and the wild type $\Gw$, the relative strength $\rho = \Rv/\Rw$ under the Gamma assumption \citep{park2019practical} is given by:
@@
 On the other hand, the forward distribution starts from the cohort of infectors who were infected at the same time and looks at when they transmitted infections.
-While the forward distribution is more reliable, changes in epidemic dynamics due to intervention measures and susceptible depletions can affect the shape of the distribution \citep{ali2020serial, nishiura2010time}.
+While the forward distribution is more stable, changes in epidemic dynamics due to intervention measures and susceptible depletions can affect the shape of either distribution \citep{ali2020serial, nishiura2010time}.
 For example, comparing generation-interval estimates from different intervention periods will necessarily bias the true differences in the generation-interval distributions of different variants.

### 2022-05-03 response_final.tex
delete
tiny changes

@@
 
-We are submitting the final version our manuscript, “The importance of the generation interval in investigating dynamics and control of new SARS-CoV-2 variants”.
+We are submitting the final version of our manuscript, “The importance of the generation interval in investigating dynamics and control of new SARS-CoV-2 variants”.
 
@@
 Likewise, figures 4 and 5 have been pushed back to the end of section 6.
-We think that these issues will be less problematic when the paper is properly formatted according to the journal guideline.
+We think that these issues will be less problematic when the paper is properly formatted for publication.
 Otherwise, we have made minor changes throughout the manuscript to try to tie the results closer to the figures whenever appropriate and re-ordered some text for clearer presentation.
@@
 The pre-intervention kernel can be further decomposed in terms of the epidemic strength $\RR$ and the generation-interval distribution $g(\tau)$: $K_{\mathrm{pre}}(\tau) = \RR g(\tau)$;
-therefore, the infection knerel integrates to $\RR$ while the generation-interval distribution integrates to 1.
+therefore, the infection kernel integrates to $\RR$ while the generation-interval distribution integrates to 1.
 Then, a constant-strength intervention reduces transmission by a constant factor $\theta$ throughout infection such that the post-intervention kernel is $K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau)/\theta$.

### 2022-05-03 response_final.tex
edits

@@
 
-We are submitting the final version of our manuscript, “The importance of the generation interval in investigating dynamics and control of new SARS-CoV-2 variants”.
+We are submitting the final version our manuscript, “The importance of the generation interval in investigating dynamics and control of new SARS-CoV-2 variants”.
 
@@
 On the other hand, the forward distribution starts from the cohort of infectors who were infected at the same time and looks at when they transmitted infections.
-While the forward distribution is more reliable, changes in epidemic dynamics due to intervention measures and susceptible depletions can affect the shape of the distribution \citep{ali2020serial, nishiura2010time}.
+While the forward distribution is more stable, changes in epidemic dynamics due to intervention measures and susceptible depletions can affect the shape of either distribution \citep{ali2020serial, nishiura2010time}.
 For example, comparing generation-interval estimates from different intervention periods will necessarily bias the true differences in the generation-interval distributions of different variants.''
